---
title: "Chapter 5 Net Present Value and Other Investment Rules"
subtitle: "Concept Questions"
author: "Ross, Westerfield, Jaffe"
output:
  pdf_document:
    toc: true
---

```{r setup, include=F}
require(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
pretty <- function(x, y=0) format(round(x, y), big.mark=",")
```

```{r Internal Rate of Return function, echo=F}
# I must be a negative initial investment.
# C must be positive cash flows.
irr <- function(I, C) {
  f <- function(r) I + sum(C/(1+r)^(1:length(C)))
  uniroot(f, lower=1e-5, upper=1, tol=1e-8)$root
}
```

```{r Net present value, echo=F}
npv <- function(I, C, r) I + sum(C/(1 + r)^(1:length(C)))
```

# 1. Payback period and net present value

If a project with conventional cash flows has a payback period less than the
project's life, can you definitively state the algebraic sign of the NPV? Why or
why not? If you know that the discounted payback period is less than the
project's life, what can you say about the NPV? Explain.

A project's payback period is not enough to tell if the NPV will be positive
(profitable project). This is because the payback period does not take into
account the interest rate. However, if the discounted payback period is less
than the project's life, then we can say that the NPV will be positive.

---

## 2. Net present value

Suppose a project has conventional cash flows and a positive NPV. What do you
know about its payback? Its discounted payback? Its profitability index? Its
IRR? Explain.

If the project has conventional (positive) cash flows and a positive NPV, then
we know the payback period will be less than the life of the project, its
discounted payback will be less than the life of the project, its profitability
index will be greater than 1, and its IRR will be greater than the discount
rate. All of this means that the project should be accepted.

---

## 3. Comparing investment criteria

Define each of the following investment rules and discuss any potential
shortcomings of each. In your definition, state the criterion for accepting or
rejecting independent projects under each rule.

a. Payback period.
b. Internal rate of return.
c. Profitability index.
d. Net present value.

### Payback period

The payback period defines how an investment should be viewed based on how long
it takes for the investment to pay for itself. The issue with the payback period
is that it does not take into account the time value of money as well as the
cash flows after the payback period has lapsed. This makes it difficult to use
when comparing multiple investment options because it doesn't take into account
cash flows after the payback period.

A modified version of the payback period is the discounted payback period which
does take into account the time value of money. However, it still does not take
into account the cash flows after the payback period. The NPV method is better.

### Internal rate of return

The internal rate of return is a better method of evaluating an investment than
the payback period. It calculates the investments interal rate, which can be
used to compare with the discount rate. If the internal rate is greater than the
discount rate the investment should be made. However, there are problems with
the internal rate of return:

1. Some investments are inverted. Meaning that the investment has an initial
   inflow, followed by a sequence of outflows. In this case the IRR rule should
   be inverted: You should make the investment if the internal rate is less than
   the discount rate.

2. Sometimes there are a number of sign changes in the cash flows (i.e. there
   are additional investments that need to be made beside the initial
   investment). If this is the case there will be multiple internal rates of
   return. To compensate for this problem you can use the Modified IRR, which
   addresses the sign change issue by combining cash flows until only one sign
   change remains. The alternative is to use NPV.

The IRR also has an issue when trying to compare mutually exclusive investments
due to scale and timing. The IRR does not take into account the scale of the
investments; meaning that larger investments may generate larger cash flows even
though the IRR may be lower. This can be compensated using Incremental IRR which
uses the difference between the two investments when comparing. Timing of the
cash flows can also affect the validity of the IRR. If larger cash flows occur
later in the life of the investment, depending on the discount rate, the IRR may
give the wrong decision.

### Profitability index

The profitability index is a useful way of comparing investments, however it
does suffer from the same problems as the internal rate of return in terms of
scaling. It can be compensated using incremental analysis.

### Net present value

The net present value is the most superior method in evaluating an investment.
In all cases it will yield the right result. The net present value takes into
account the time value of money of all cash flows including the initial
investment. If the NPV is positive the investment will pay off. If the NPV is
negative it will not.

---

## 4. Payback and internal rate of return

A project has perpetual cash flows of $C$ per period, a cost of $I$, and a
required return of $R$. What is the relationship between the project's payback
and its IRR? What implications does your answer have for long-lived projects
with relatively constant cash flows?

The payback period takes into account the initial investment and the cash flows
to determine how long it will take to pay back the investment.

$$
\begin{aligned}
\mathbf{Payback\ Period} &= \frac{I}{C}
\end{aligned}
$$

Since the cash flows are perpetual the present value of the investment is:

$$
\begin{aligned}
\mathbf{PV} &= \frac{C}{r}
\end{aligned}
$$

Setting $\mathbf{PV} = I$, the IRR can be calculated:

$$
\begin{aligned}
I &= \frac{C}{r}\\\\
r &= \frac{C}{I}
\end{aligned}
$$

Comparing the payback period and the IRR, we can see there is an inverse
relationship between the two:

$$
\begin{aligned}
\mathbf{Payback\ Period} &= \frac{1}{IRR} \\ \\
\mathbf{IRR} &= \frac{1}{\text{Payback Period}}
\end{aligned}
$$

What this means is that a short payback period implies a high IRR because cash
flows quickly recover the initial investment and continue generating value.
Conversely, a long payback period implies a low IRR because the initial
investment takes more time to be recovered.

---

## 5. International investment projects

In October 2010, BMW announced plans to spend \$1 billion to expand production
at its plant in South Carolina. The plant produced the second generation BMW X3
as well as the company's X5 and X6 models. BMW apparently felt it would be
better able to compete and create value with a U.S.-based facility. In fact, BMW
actually expected to export 70 percent of the X3s produced in South Carolina.
About the same time, Samsung Electronics announced plans to build a \$3.6
billion plant in Taxes, and Novartis announced plans to open a \$1 billion-plus
plant in North Carolina. What are some of the reasons that foreign manufacturers
of products as diverse as automobiles, electronics, and pharmaceuticals might
arrive at the same conclusion to build plants in the U.S.?

There are many reasons why foreign manufacturers would want to build plants in
the U.S.:

1. The U.S. is one of the largest consumer markets in the world. To have a
   manufacturing plant in the U.S. gives them direct access to this market.

2. Producing goods in the U.S. may offer the manufacturer export opportunity
   benefits they wouldn't have in their host country.

3. Producing goods in the U.S. allows the manufacturer to avoid tariffs and
   trade barriers.

4. The U.S. has one of the most skilled and educated workforce in the world.
   Building a plant in the U.S. gives them access to this workforce.

This is by no means an exhaustive list but are some examples as to why foreign
manufacturers would build plants in the U.S. to produce their products.

---

## 6. Capital budgeting problems

What are some of the difficulties that might come up in actual applications of
the various criteria we discussed in this chapter? Which would be the easiest to
implement in actual applications? The most difficult?

### Difficulties

#### Payback period

1. Ignores the time value of money.

2. Has an arbitrary cutoff period.

3. Ignores cash flows beyond the cutoff period.

4. Not a measure of profitability.

#### Discounted payback period

1. Because we now discount the relevant cash flows, discounted payback period is
   more complex than the original payback period.

2. Still ignores cash flows beyond the cutoff period.

3. Still has an arbitrary cutoff period.

#### Internal rate of return (IRR)

1. For projects with non-conventional cash flows, such as multiple outflows,
   there could be multiple IRRs or no IRR at all.

2. IRR assumes that interim cash flows are reinvested at the IRR, which may be
   unrealistic. NPV, by contrast, assumes reinvestment at the cost of capital.

3. IRR may give incorrect rankings when comparing mutually exclusive projects
   when the scale of the projects, or the timing of cash flows differ.

4. Because it is a ratio, IRR does not take into account the scale of the
   project.

#### Profitability index (PI)

1. Because the PI is a ratio as well, it too does not take into account the
   scale of the project.

#### Net present value (NPV)

1. Requires accurate estimates to be effective.

2. The net present value method is highly complex, and difficult to explain its
   conclusions.

3. The net present value method has a bias toward larger projects because it
   deals with absolute, rather than relative cash flows. Using this method may
   miss smaller projects with better rates of return.

#### General difficulty issues when using capital budgeting tools

1. Capital budgeting tools require accurate estimates of future cash flows,
   discount rates, and project life. All of these are difficult to forecast.

2. When trying to compare mutually exclusive projects, it is important to take
   note of the size of the projects because the ratio based capital budgeting
   tools do not take into account the scale of the projects. The net present
   value method, however, does overcome this issue.

3. Because of the issue described in (2), comparing projects in general is
   difficult.

4. Capital budgeting tools only focus on financial metrics. Although financial
   metrics should be the main focus, other factors, such as strategic alignment,
   environmental impact, or regulatory considerations may be overlooked.

The advantages of each of the capital budgeting tools discussed:

#### Payback period

This method is good for quick assessments, or for projects with high risk where
a fast payback is important to reduce uncertainty.

#### Discounted payback period

Similar to the payback period, but for projects that need to take into account
the time value of money this method should be used.

#### Net present value

This is the most accurate method and always leads to the correct decision if the
estimates are accurate. Net present value is very useful for long and complex
projects with predictable cash flows and where maximizing value is important.
Net present value is also useful when comparing mutually exclusive projects.

#### Internal rate of return (IRR)

Very useful when comparing projects where the relative rate of return is
important because it returns this rate as an easily communicable percentage.

#### Profitability index (PI)

An excellent choice when comparing mutually exclusive projects of similar scale
to determine the most efficient use of resources.

---

## 7. Capital budgeting in not-for-profit entities

Are the capital budgeting criteria discussed applicable to not-for-profit
corporations? How should such entities make capital budgeting decisions? What
about the U.S. government? Should it evaluate spending proposals using these
techniques?

Although not-for-profit corporations, or government organizations like the U.S.
government, have different goals than for-profit corporations, they still need
to determine the most efficient use of limited resources. Capital budgeting
techniques such as the profitability index (PI), the internal rate of return
(IRR), and the net present value (NPV) are excellent choices for these tasks.

---

## 8. Net present value

The investment in project A is \$1 million, and the investment in project B is
\$2 million. Both projects have unique internal rate of return of 20 percent. Is
the following statement true or false?

> For any discount rate from 0 percent to 20 percent, project B has an NPV twice
> as great as that of project A.

This statement is true.

$$
\begin{aligned}
\mathbf{NPV} &= \sum^{T}_{t=1}{\frac{C_t}{(1 + r)^{t}}} - I
\end{aligned}
$$

where:

* $C_t =$ cash flow at period $t$.
* $r =$ the discount rate.
* $I =$ the initial investment.

Since both have the same IRR, their cash flow structure are proportional to
their respective initial investments. Since the cash flows are proportional to
the investments, then the cash inflows of project B are twice the cash inflows
of project A.

---

## 9. Net present value versus profitability index

Consider the following two mutually exclusive projects available to Global
Investments, Inc.:

```{r Table for concepts 9, echo=F}
profitability.index <- function(C, r) {
  I <- head(C,1)
  C <- tail(C,-1)
  n <- length(C)
  pv <- sum(C/(1 + r)^(1:n))
  -pv/I
}
npv <- function(C, r){
  n <- length(C)-1
  sum(C/(1 + r)^(0:n))
}
A <- c(-1000, 1000, 500)
B <- c(-500, 500, 400)
c9.tb <- as.data.frame(
  matrix(c(A, round(profitability.index(A, .1), 2), round(npv(A, .1)),
           B, round(profitability.index(B, .1), 2), round(npv(B, .1))),
         nrow=2, byrow=T))
rownames(c9.tb) <- c("A", "B")
colnames(c9.tb) <- c("C0", "C1", "C2", "PI", "NPV")
knitr::kable(c9.tb)
```

The appropriate rate for the projects is 10 percent. Global Investments chose to
undertake Project A. At a luncheon for shareholders, the manager of a pension
fund that owns a substantial amount of the firm's stock asks you why the firm
chose Project A instead of Project B when Project B has a higher profitability
index.

How would you, the CFO, justify your firm's action? Are there any circumstances
under which Global Investments should choose Project B?

Although project B has a higher profitability index rating, Project A has a
higher net present value. Net present value evaluates the absolute value of the
projects where as the profitability index evaluates the financial efficiency of
a project (how much "bang" for your buck). The profitability index is not useful
when comparing two projects of different scale because it doesn't take into
account the absolute values of the projects value.

---

## 10. Internal rate of return

Projects $A$ and $B$ have the following cash flows:

```{r Table for concept 10, echo=F}
Year <- c(0,1,2,3)
A <- c(-1000, "C1A", "C2A", "C3A")
B <- c(-2000, "C1B", "C2B", "C3B")
c10.df <- as.data.frame(matrix(c(Year,A,B),ncol=3))
colnames(c10.df) <- c("Year", "Project A", "Project B")
knitr::kable(c10.df)
```

a. If the cash flows from the projects are identical, which of the two projects
   would have a higher IRR? Why?

b. If C1B = 2C1A, C2B = 2C2A, and C3B = 2C3A, then is
   $\text{IRR}_A = \text{IRR}_B$?

IRR analysis:

$$
\begin{aligned}
\mathbf{NPV} &= -I + \sum^T_{t=1}{\frac{C}{(1 + R)^{t}}}
\end{aligned}
$$

To find the internal rate of return we choose an arbitrary rate and solve for
$\text{NPV}$. If the value is $>0$ than we choose a new lower rate. If the rate
is $<0$ then we choose a higher rate until we get $\text{NPV}=0$.

This process can be done via `R` using the `uniroot` function.


a. If the two projects have identical cash flows, then project A has the higher
IRR because it has a lower initial investment.

b. If the cash flows for project B are twice the cash flows of project A, then
the two projects would have the same IRR.

The reason for this result is because the IRR is a percentage that is compared
to the discount rate. If the value of IRR is greater than the discount rate than
we choose the project. However, the IRR is a percentage and not an aboslute
value. Therefore, comparing two projects as such would most likely give us an
incorrect result and we need to compare these projects using NPV.

---

## 11. Net present value

You are evaluating Project A and Project B. Project A has a short period of
future cash flows, while project B has relatively long future cash flows. Which
project will be more sensitive to changes in the required return? Why?

Project B would be more sensitive to changes in the required return because its
cash flows extend for a longer time period and the time value of money effects
are heavier on cash flows further out into the future.

---

## 12. Modified internal rate of return

One of the less flattering interpretations of the acronym MIRR is "meaningless
internal rate of return." Why do you think this term is applied to MIRR?

1. Loss of original IRR intuition - The MIRR adjusts cash flows based on
   reinvestment and financing assumptions (reinvestment at a predefined
   borrowing costs). While these adjustments address some of the shortcomings,
   they result in a metric that no longer has the same intuitive connection
   connection to the project's cash flows.

2. Assumptions about the reinvestment rate - The MIRR assumes that all positive
   cash flows are reinvested at a specified reinvestment rate (the discount
   rate). However, this may not be the case.

3. MIRR is not a pure rate of return - The MIRR is not purely derived from the
   project's cash flows since it uses a discount rate.

* Lacks the intuitive appeal of IRR as a breakeven return.

* Relies heavily on assumptions about reinvestment and financing rates.

* Deviates from representing the project's intrinsic profitability.

---

## 13. Net present value

It is sometimes stated that "the net present value approach assumes reinvestment
of the intermediate cash flows at the required return." Is this claim correct?
To answer, suppose you calculate the NPV of a project in the usual way. Next,
suppose you do the following:

a. Calculate the future value (as of the end of the project) of all the cash
flows other than the initial outlay assuming they are reinvested at the required
return, producing a single future value figure for the project.

b. Calculate the NPV of the project using the single future value calculated in
the previous step and the initial outlay. It is easy to verify that you will get
the same NPV as in your original calculation only if you use the required return
as the reinvestment rate in the previous step.

The statement that the "net present value (NPV) approach assumes reinvestment of
intermediate cash flows at the required return" is not entirely accurate. The
NPV method does not explicitly assume any reinvestment rate for intermediate
cash flows. Instead, it discounts all cash flows back to the present at the
required return (discount rate) and evaluates the net value created.

### Calculate the future value of intermediate cash flows

Suppose the project's cash flows are $C_0, C_1, C_2, \ldots, C_n$, where $C_0$
is the initial outlay (negative cash flow), and $C_1, C_2, \ldots, C_n$ are
positive cash inflows.

1. To calculate the future value (FV) of all cash flows other than the initial
   outlay:

   * Assume each intermediate cash flow($C_t$) is reinvested at a specific rate
     $r$ until the end of the project.

   * The future value at time $n$ is:

$$
\begin{aligned}
\mathbf{NPV} &= - C_0 + \frac{\text{FV}}{(1 + R)^{n}}
\end{aligned}
$$

2. For this calculation to match the original NPV calculated using the standard
   formula:

$$
\begin{aligned}
\mathbf{NPV} &= - C_0 + \sum^n_{t=1}{\frac{C_t}{(1 + R)^t}}\text{,}
\end{aligned}
$$

> the reinvestment rate $r$ in Step(a) must equal the required return $R$.

### Key Insights

* The NPV method does not explicitly assume a reinvestment rate for intermediate
  cash flows. It merely discounts cash flows to the present using the required
  return $R$.

* The equivalence between the two approaches described in the question arises
  only if the reinvestment rate $r$ used in Step (a) equals the required areturn
  $R$. This is a mathematical coincidence rather than an inherent assumption of
  the NPV method.

### Conclusion

The claim that the NPV approach assumes reinvestment of intermediate cash flows
at the required return is misleading. The NPV method:

1. Focuses on discounting all cash flows to the present at the required return.

2. Does not explicitly account for the reinvestment of intermediate cash flows.

The equivalence described in the question arises only when the reinvestment rate
matches the required return, but this is not an assumption of the NPV approach
itself. Instead, the NPV method evaluates the project's value based on its cash
flows and the required return, without considering reinvestment explicitly.

---

## 14. Internal rate of return

It is sometimes stated that "the internal rate of return approach assumes
reinvestment of the intermediate cash flows at the internal rate of return." Is
this claim correct? To answer, suppose you calculate the IRR of a project in the
usual way. Next, suppose you do the following:

a. Calculate the future value (as of the end of the project) of all the cash
flows other than the initial outlay assuming they are reinvested at the IRR,
producing a single future value figure for the project.

b. Calculate the IRR of the project using the single future value calculated in
the previous step and the initial outlay. It is easy to verify that you will get
the same IRR as in your original calculation only if you use the IRR as the
reinvestment rate in the previous step.

The statement that the "internal rate of return (IRR) approach assumes
reinvestment of the intermediate cash flows at the internal rate of return" is
correct in a practical sense. The IRR implicitly assumes that intermediate cash
flows are reinvested at the IRR itself. This is a key limitation of the IRR
method and one of the reasons it is often criticized.

### Step a: Calculate the future value of intermediate cash flows

Suppose a project has cash flows $C_0, C_1, C_2, \ldots, C_n,$ where $C_0$ is
the initial outlay (negative cash flow), and $C_1, C_2, \ldots, C_n$ are
positive inflows. The IRR, denoted as $r$, satisfies:

$$
\begin{aligned}
\mathbf{NPV} &= - C_0 + \sum^{n}_{t=1}{\frac{C_t}{(1 + r)^{t}}} = 0
\end{aligned}
$$

1. Assume each intermediate cash flow ($C_t$) is reinvested at the IRR $r$ until
   the end of the project. The future value(FV) at time $n$ is:

$$
\begin{aligned}
\mathbf{FV} &= \sum^{n}_{t=1}{C_t (1 + r)^{n-t}}
\end{aligned}
$$

2. This future value represents the total value of the cash flows at the end of
   the project, assuming reinvestment at the IRR.

### Step b: Calculate the IRR using the future value

1. Use the future value $\text{FV}$ calculated in Step (a) and determine the IRR
   $r$ by solving:

$$
C_0 (1 + r)^{n} = \text{FV}
$$

2. This equation is equivalent to solving for $r$ in the original IRR
   calculation:

$$
0 = - C_0 + \sum^{n}_{t=1}{\frac{C_t}{(1 + r)^{t}}}
$$

3. The two IRR values will be identical only if the reinvestment rate used in
   Step (a) equals the IRR itself.

### Key Insights

* The IRR implicitly assumes that all intermediate cash flows are reinvested at
  the IRR. This is why the future value described in Step (a) aligns with the
  IRR calculation only when the reinvestment rate is the IRR.

* If the reinvestment rate differes from the IRR, the results of the two
  calculations will diverge.

### Limitations of the IRR assumption

1. Unrealistic reinvestment rate - In practice, it is unlikely that intermediate
   cash flows can consistently be reinvested at the project's IRR, especially if
   the IRR is high. This makes the IRR's implicit assumption unrealistic in many
   cases.

2. Alternative methods - The modified internal rate of return (MIRR) addresses
   this limitation by allowing for a more realistic reinvestment rate, such as
   the cost of capital.

### Conclusion

The claim that the IRR approach assumes reinvestment of intermediate cash flows
at the IRR is correct. The equivalence in the steps provided confirms this
assumption. However, this assumption is a major limitation of the IRR method, as
it may not reflect reinvestment opportunities. For this reason, decision-makers
often rely on NPV or MIRR to provide a more accurate assessment of a project's
value.

