---
title: "Chapter 5 Net Present Value and Other Investment Rules"
subtitle: "Problems"
author: "Ross, Westerfield, Jaffe"
output:
  pdf_document:
    toc: true
---

```{r setup, include=F}
require(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
pretty <- function(x, y=2) {
    ifelse(x == Inf, Inf, format(round(x, y), big.mark = ","))
}
```

```{r present value, echo=F}
# Input, present value type vector with investment (as negative)
# at the beginning [i.e. cash.flows = c(25, 25, 35, 50)]
# The discount rate r also is specified
pv <- function(C, r){
    sum(C / (1 + r)^(1:length(C)))
}
```

```{r Payback period, echo=F}
# Input, NPV type vector with investment (as negative)
# at the beginning [i.e. cash.flows = c(-100, 25, 25, 35, 50)]
payback.period <- function(C, fractional=F){
    cum_C = cumsum(C)
    idx = which(cum_C >= 0)[1]
    if(is.na(idx)) return(Inf)
    if(!fractional){
        return(idx-1)
    } else {
        prev_cum = cum_C[idx - 1]
        current_C = C[idx]
        frac = -prev_cum / current_C
        return((idx-2) + frac)
    }
}
```

```{r Internal Rate of Return function, echo=F}
# Input, NPV type vector with investment (as negative)
# at the beginning [i.e. cash.flows = c(-100, 25, 25, 35, 50)]
irr <- function(C) {
    uniroot(function(r) sum(C / (1 + r)^(0:(length(C)-1))), c(0, 1))$root
}

mirr <- function(C, r) {
    n = length(C) - 1
    t = 0:n
    pv_neg = sum(ifelse(C < 0, C / (1 + r)^t, 0))
    fv_pos = sum(ifelse(C > 0, C * (1 + r)^(n-t), 0))
    mirr_value = (fv_pos / -pv_neg)^(1/n) - 1
    return(mirr_value)
}
```

```{r Net present value, echo=F}
# Input, NPV type vector with investment (as negative)
# at the beginning [i.e. cash.flows = c(-100, 25, 25, 35, 50)]
# The discount rate r also is specified
npv <- function(C, r) sum(C / (1 + r)^(0:(length(C)-1)))
```

```{r profitability index, echo=F}
# Input, NPV type vector with investment (as negative)
# at the beginning [i.e. cash.flows = c(-100, 25, 25, 35, 50)]
# The discount rate r also is specified
profitability.index <- function(C, r) {
  I = -head(C, 1)
  C = tail(C, -1)
  pv = sum(C / (1 + r)^(1:length(C)))
  pv / I
}
```

# 1. Calculating payback period and NPV

Fuji Software, Inc., has the following mutually exclusive projects.

```{r problem 1 Table, echo=F}
p1 = new.env()
p1$tbl = tibble(
    Year = c(0, 1, 2, 3),
    `Project A` = c(-15000, 9500, 6000, 2400),
    `Project B` = c(-18000, 10500, 7000, 6000)
)
p1$payback.A = payback.period(p1$tbl$`Project A`)
p1$npv.A = npv(p1$tbl$`Project A`, .15)
p1$payback.B = payback.period(p1$tbl$`Project B`)
p1$npv.B = npv(p1$tbl$`Project B`, .15)
knitr::kable(p1$tbl)
```

a. Suppose Fuji's payback period cutoff is two years. Which of these two
   projects should be chosen?

b. Suppose Fuji uses the NPV rule to rank these two projects. Which project
   should be chosen if the appropriate discount rate is 15 percent?

## A. Payback cutoff is two years

For project A, the payback period is `r p1$payback.A`

For project B, the payback period is `r p1$payback.B`

If we use a payback cutoff of two years, we should select Project A.

##  B. NPV at a discount of 15 percent

If we use npv with a discount rate of 15 percent.

\[
\text{NPV} = C_0 + \sum_{t=1}^{T}{\frac{C_t}{(1 + r)^t}}
\]

Project A, the NPV is \$ `r pretty(p1$npv.A)`

Project B, the NPV is \$ `r pretty(p1$npv.B)`

Project A has a negative NPV, Project B is positive. You should choose project
B.

---

# 2. Calculating payback

An investment project provides cash inflows of \$840 per year for eight years.
What is the project payback period if the initial cost is \$3,200? What if the
initial cost is \$4,800? What if it is \$7,300?

```{r problem 2, echo=F}
p2 = new.env()
p2$C.1 = c(-3.2e3, rep(840, 8))
p2$C.2 = c(-4.8e3, rep(840, 8))
p2$C.3 = c(-7.3e3, rep(840, 8))
p2$payback.1 = payback.period(p2$C.1, fractional=T)
p2$payback.2 = payback.period(p2$C.2, fractional=T)
p2$payback.3 = payback.period(p2$C.3, fractional=T)
```
To calculate the payback period you divide the initial cost by the cash flow.

\[
\text{Payback} = \frac{3,200}{840} = `r pretty(p2$payback.1)`\ \text{years}
\]

\[
\text{Payback} = \frac{4,800}{840} = `r pretty(p2$payback.2)`\ \text{years}
\]

\[
\text{Payback} = \frac{7,300}{840} = `r pretty(p2$payback.3)`\ \text{years}
\]

The last investment cost would not payback in eight years.

Book answer: 3.81 years, 5.71 years, Never

# 3. Calculating discounted payback

An investment project has annual cash inflows of \$5,000, \$5,500, \$6,000,
\$7,000, and a discount rate of 14 percent. What is the discounted payback
period for these cash flows if the initial cost is \$8,000? What if the initial
cost is \$12,000? What if it is \$16,000?

```{r problem 3; calculating discounted payback, echo=F}
p3 = new.env()
p3$C = c(5e3, 5.5e3, 6e3, 7e3)
p3$r = .14
p3$C.discounted = p3$C * (1 + p3$r)^(-(1:length(p3$C)))
p3$payback.1 = payback.period(c(-8e3, p3$C.discounted), fractional=T)
p3$payback.2 = payback.period(c(-12e3, p3$C.discounted), fractional=T)
p3$payback.3 = payback.period(c(-16e3, p3$C.discounted), fractional=T)
```

To calculate the discounted payback period, you have to first discount each of
the cash flows:

\[
[5,000/(1.14), 5,500/(1.14)^2, 6,000/(1.14)^3, 7,000/(1.14)^4]
    = (`r pretty(p3$C.discounted)`)
\]

We can now calculate the payback period using the discounted cash flows:


The discounted payback is calculated:

For an initial cost of \$8,000, the payback period is:
`r pretty(p3$payback.1)` years.

For an initial cost of \$12,000, the payback period is:
`r pretty(p3$payback.2)` years.

For an initial cost of \$16,000, the payback period is:
`r pretty(p3$payback.3)` years.

---

# 4. Calculating discounted payback

An investment project costs \$15,000 and has annual cash flows of \$3,800 for
six years. What is the discounted payback period if the discount rate is 0
percent? What if the discount rate is 10 percent? If it is 15 percent?

```{r problem 4; calculating discounted payback, echo=F}
p4 = new.env()
p4$C = rep(3.8e3, 6)
p4$C.discounted.1 = p4$C * (1)^(-(1:length(p4$C)))
p4$C.discounted.2 = p4$C * (1.1)^(-(1:length(p4$C)))
p4$C.discounted.3 = p4$C * (1.15)^(-(1:length(p4$C)))
p4$payback.1 = payback.period(c(-15e3, p4$C.discounted.1), fractional=T)
p4$payback.2 = payback.period(c(-15e3, p4$C.discounted.2), fractional=T)
p4$payback.3 = payback.period(c(-15e3, p4$C.discoutned.3), fractional=T)
```

For a discount rate of 0\%: `r pretty(p4$payback.1)` Years

The discounted cash flows are:

\[
(`r p4$C.discounted.1`)
\]

For a discount rate of 10\%: `r pretty(p4$payback.2)` Years

The discounted cash flows are:

\[
(`r pretty(p4$C.discounted.2)`)
\]

For a discount rate of 15\%: `r pretty(p4$payback.3)` Years

The discounted cash flows are:

\[
(`r pretty(p4$C.discounted.3)`)
\]

This final discount rate means that the project will never be profitable.

---

# 5. Calculating IRR

Stone Sour, Inc., has a project with the following cash flows:

```{r problem 5 table, echo=F}
p5 = new.env()
p5$tbl = tibble(
    Year = c(0, 1, 2, 3),
    `Cash Flows` = c(-20e3, 8.5e3, 10.2e3, 6.2e3)
)
knitr::kable(p5$tbl)
p5$irr = irr(p5$tbl$`Cash Flows`)
```

The company evaluates all projects by applying the IRR rule. If the appropriate
interest rate is 9 percent, should the company accept the project?

We use the following model to find the internal rate of return.

\[
0 = \text{Initial investment} + \sum^T_{t=1}{\frac{\text{CF}}{(1 + R)^t}}
\]

In this equation the knowns are $\text{Initial investment}$, and
$\text{CF}$. We solve for $R$:

\[
R = `r pretty(p5$irr*100)`\%
\]

---

# 6. Calculating IRR

Compute the internal rate of return for the cash flows of the following two projects:

```{r question 6 table, echo=F}
p6 = new.env()
p6$tbl = tibble(
    Year = c(0, 1, 2, 3),
    `Project A` = c(-5.3e3, 2e3, 2.8e3, 1.6e3),
    `Project B` = c(-2.9e3, 1.1e3, 1.8e3, 1.2e3)
)
knitr::kable(p6$tbl)
p6$irr.A = irr(p6$tbl$`Project A`)
p6$irr.B = irr(p6$tbl$`Project B`)
```

Again, we use the following model to find the internal rate of return for the
two projects.

\[
0 = \text{Initial investment} + \sum^T_{t=1}{\frac{\text{CF}}{(1 + R)^t}}
\]

In this equation the knowns are $\text{Initial investment}$, and
$\text{CF}$. We solve for $R$:

\[
R_{\text{Project A}} = `r pretty(p6$irr.A*100)`\%
\]

\[
R_{\text{Project B}} = `r pretty(p6$irr.B*100)`\%
\]

---

# 7. Calculating profitability index

Bill plans to open a self-serve grooming center in a storefront. The grooming
equipment will cost \$385,000, to be paid immediately. Bill expects aftertax
cash inflows of \$84,000 annually for seven years, after which he plans to scrap
the equipment and retire to the beaches of Nevis. The first cash inflow occurs
at the end of the first year. Assume the required return is 13 percent. What is
the project's PI? Should it be accepted.

The profitability index is calculated:

```{r problem 7  calculating profitability index, echo=F}
p7 = new.env()
p7$cf <- c(-385e3, rep(84e3, 7))
p7$r = .13
p7$pv = pv(tail(p7$cf, -1), p7$r)
p7$pi = profitability.index(p7$cf, p7$r)
```

$$
\begin{aligned}
\text{PV}_{\text{Cash flows}} &= \sum^{T}_{t=1}{\frac{C_t}{(1 + r)^t}}\\
    &= `r pretty(p7$pv, 0)`\\\\
\text{Profitability\ Index} &= \frac{\text{PV}_{\text{Cash flows}}}
    {\text{Initial investment}}\\
    &= \frac{`r pretty(p7$pv, 0)`}{`r pretty(-head(p7$cf, 1), 0)`}\\
    &= `r pretty(p7$pi)`
\end{aligned}
$$

The profitability index is 0.965 which is less than 1. The project should not be accpeted.

---

# 8. Calculating profitability index

Suppose the following two independent investment opportunities are available to
Greenplain, Inc. The appropriate discount rate is 10 percent.

```{r problem 8 table, echo=F}
p8 = new.env()
p8$tbl = tibble(
    Year = c(0, 1, 2, 3),
    Alpha = c(-2.3e3, 1.2e3, 1.1e3, 900),
    Beta = c(-3.9e3, 800, 2.3e3, 2.9e3)
)
p8$r = .1
knitr::kable(p8$tbl)
p8$pi.alpha = profitability.index(p8$tbl$Alpha, p8$r)
p8$pi.beta  = profitability.index(p8$tbl$Beta,  p8$r)
```

a. Compute the profitability index for each of the two projects.

b. Which project(s) should Greenplain accept based on the profitability index rule?

## A. The profitability index

The profitability index of both projects are:

* Project Alpha: `r pretty(p8$pi.alpha)`
* Project Beta:  `r pretty(p8$pi.beta)`

## B. Which projects to accept

These projects are not mutually exclusive so Greenplain should accept both
projects since their profitability indexes are above 1.

---

# 9. Cash flow intuition

A project has an initial cost of $I$, has a required return of $R$, and pays $C$ annually for $N$ years.

a. Find $C$ in terms of $I$ and $N$ such that the project has a payback period just equal to its life.
b. Find $C$ in terms of $I$, $N$, and $R$ such that this is a profitable project according to the NPV decision rule.
c. Find $C$ in terms of $I$, $N$, and $R$ such that the project has a benefit--cost ratio of 2.

Setting the present value of the cash flows to the initial investment, we
describe the model below:

## A. $C$ in terms of $I$ and $N$ for a payback period equal to its life.

With payback period we disregard discounting. For a payback period equal to the
life of the project, the model is:

\[
\begin{aligned}
N &= \frac{I}{C}\\\\
C &= \frac{I}{N}
\end{aligned}
\]

In words, this says the cash flows should equal the initial investment divided
by the life of the project.

## B. An NPV profitable project, find $C$

For a profitable project the NPV model is:

\[
\text{NPV} = -I + \sum^N_{n=1}{\frac{C}{(1 + R)^{n}}} > 0
\]

The geometric series can be rewritten as

\[
\sum^N_{n=1}{\frac{1}{(1 + R)^n}} = \frac{1 - (1 + r)^{-N}}{r}
\]

Substituting back into the equation

\[
\begin{aligned}
I &< C \times \frac{1 - (1 + r)^{-N}}{r}\\
C &> \frac{Ir}{1 - (1 + r)^{-N}}
\end{aligned}
\]

## C. Benefit---cost ratio of 2

To model a benefit/cost ratio of 2 we use the following equation

\[
\begin{aligned}
\text{BC} &= \frac{\text{PV}_{\text{cash flows}}}{\text{initial investment}}\\\\
   2  &= \frac{C}{I} \frac{1 - (1 + r)^{-N}}{r}\\\\
   C &= \frac{2Ir}{1 - (1 + r)^{-N}}
\end{aligned}
\]

---

# 10. Problems with IRR

Suppose you are offered \$7,000 today but must make the following payments:

```{r problem 10 table, echo=F}
p10 = new.env()
p10$tbl = tibble(
    Year = c(0, 1, 2, 3, 4),
    `Cash Flows` = c(7e3, -3.7e3, -2.4e3, -1.5e3, -1.2e3)
)
year <- c(0,1,2,3,4)
cf <- c(7000, -3700, -2400, -1500, -1200)
knitr::kable(p10$tbl)
p10$irr = irr(p10$tbl$`Cash Flows`)
p10$npv.10 = npv(p10$tbl$`Cash Flows`, .10)
p10$npv.20 = npv(p10$tbl$`Cash Flows`, .20)
```

a. What is the IRR of this offer?
b. If the appropriate discount rate is 10 percent, should you accept this offer?
c. If the appropriate discount rate is 20 percent, should you accept this offer?
d. What is the NPV of the offer if the appropriate discount rate is 10 percent? 20 percent?
e. Are the decisions under the NPV rule in part (d) consistent with those of the IRR?

Notice that the cash flows are opposite from the normal cash flows (i.e. There is an initial inflow followed by a series of outflows.) This means that the IRR rule is inverted: (accept the project if the IRR < discount rate).

## A. What is the IRR

We solve the following equation for $r$

$$
\begin{aligned}
I = \sum^{4}_{n=1}{\frac{C_n}{(1 + r)^{n}}}
\end{aligned}
$$

The internal rate of return is `r pretty(p10$irr*100)`\%.

## B. Discount rate is 10\%

The IRR $>$ the discount rate (10\%). Since the cash flows are inverted, we
invert the IRR rule so we reject this project.

## C. Discount rate is 20\%

With a 20\% discount rate we accept the project since this is less than the
`r pretty(p10$irr*100)`\% IRR rate.

## D. NPV for discount rates of 10\% and 20\%

The NPV for both 10% and 20% discount rates can be computed:

For a 10\% discount rate the NPV is `r pretty(p10$npv.10)`.

For a 20\% discount rate the NPV is `r pretty(p10$npv.20)`.

This is consistent with the IRR rule.

---

# 11. NPV versus IRR

Consider the following cash flows on two mutually exclusive projects for the
Bahamas Recreation Corporation (BRC). Both projects require an annual return of
14 percent.

```{r problem 11 table, echo=F}
p11 = new.env()
p11$year = c(0, 1, 2, 3)
p11$cf.dwf = c(-950e3, 370e3, 510e3, 420e3)
p11$cf.nsr = c(-1.850e6, 900e3, 800e3, 750e3)
p11$tbl = tibble(
    Year = p11$year,
    `Deepwater Fishing` = pretty(p11$cf.dwf),
    `New Submarine Ride` = pretty(p11$cf.nsr)
)
p11$irr.dwf = irr(p11$cf.dwf)
p11$irr.nsr = irr(p11$cf.nsr)
p11$cf.inc = p11$cf.nsr - p11$cf.dwf
p11$irr.inc = irr(p11$cf.inc)
p11$npv.dwf = npv(p11$cf.dwf, .14)
p11$npv.nsr = npv(p11$cf.nsr, .14)
knitr::kable(p11$tbl)
```

As a financial analyst for BRC, you are asked the following questions:

a. If your decision rule is to accept the project with the greater IRR, which
   project should you choose?

b. Because you are fully aware of the IRR rule's scale problem, you calculate
   the incremental IRR for the cash flows, Based on your computation, which
   project should you choose?

c. To be prudent, you compute the NPV for both projects. Which project should
   you choose? Is it consistent with the incremental IRR rule?

## A. Greater IRR

We calculate the IRR for both projects.

\[
\begin{aligned}
\text{IRR} = \sum^N_{n=1}{\frac{\text{CF}}{(1 + r)^{n}}}
\end{aligned}
\]

Solving for $r$ for each project we have:

* Deepwater Fishing: `r pretty(p11$irr.dwf*100)`\%

* New Submarine Ride: `r pretty(p11$irr.nsr*100)`\%

Both are above the required annual return. If we were to choose one based on the
higher IRR than Deepwater fishing would be the choice. However, the two projects
are not the same size. The IRR method has a problem when comparing projects of
different scale. In order to fix this, we should use the incremental IRR to
determine which project to choose.

## B. Incremental IRR


Using incremental IRR, we first find the incremental cash flows from Deepwater
Fishing to New Submarine Ride

```{r problem 11; incremental IRR table, echo=F}
p11$inc.irr.tbl = cbind(p11$tbl, pretty(p11$cf.inc))
colnames(p11$inc.irr.tbl) = c(colnames(p11$tbl), 'Incremental')
knitr::kable(p11$inc.irr.tbl)
```

The incremental IRR is `r pretty(p11$irr.inc*100)`\%. This is larger than the
required annual return of 14\% so we choose the New Submarine Ride project and
not the Deepwater Fishing project. Notice this is entirely different from the
previous choice where the larger IRR rate would have had us choose Deepwater
Fishing.

## C. NPV of both projects

To be sure we are making the right choice, we will also use NPV for each
project:

\[
\text{NPV} = \sum^N_{n=1}{\frac{C}{(1 + r)^n}}
\]

* Deepwater Fishing: \$`r pretty(p11$npv.dwf, 0)`

* New Submarine Ride: \$`r pretty(p11$npv.nsr, 0)`

Both projects have positive NPVs but, since they are mutually exclusive, we want
to choose the New Submarine Ride project, which confirms our choice using the
incremental IRR and confirms that comparing projects of different scales using
IRR is problematic.

---

# 12. Problems with profitability index

The Robb Computer Corporation is trying to choose between the following two
mutually exclusive design projects:

```{r problem 12 table, echo=F}
p12 = new.env()
p12$year = c(0,1,2,3)
p12$cf.I = c(-30000, rep(18000, 3))
p12$cf.II = c(-12000, rep(7500, 3))
p12$tbl = tibble(p12$year, p12$cf.I, p12$cf.II)
colnames(p12$tbl) = c("Year", "Cash Flow (I)", "Cash Flow (II)")
knitr::kable(p12$tbl)
p12$pv.I = pv(head(p12$cf.I, 1), .1)
p12$pv.II = pv(head(p12$cf.II, 1), .1)
p12$pi.I = profitability.index(p12$cf.I, .1)
p12$pi.II = profitability.index(p12$cf.II, .1)
p12$npv.I = npv(p12$cf.I, .1)
p12$npv.II = npv(p12$cf.II, .1)
```

a. If the required return is 10 percent and Robb Computer applies the
   profitability index decision rule, which project should the firm accept?

b. If the company applies the NPV decision rule, which project should it take?

c. Explain why your answers in (a) and (b) are different.

## A. Using the profitability index

\[
\begin{aligned}
\text{PI} &= \frac{\text{PV}}{\text{Initial investment}}\\\\
\text{PI}_{\text{I}} &= \frac{`r pretty(p12$pv.I)`}
    {`r pretty(head(p12$cf.I, 1))`}\\
    &= `r pretty(p12$pi.I)`\\
\text{PI}_{\text{II}} &= \frac{`r pretty(p12$pv.II)`}
    {`r pretty(head(p12$cf.II, 1))`}\\
    &= `r pretty(p12$pi.II)`
\end{aligned}
\]

Bsed on the profitability index alone, Robb Computer should choose project II.
However, there are scaling issues using the profitability index to compare the
two projects. Let's use NPV to understand which project to choose.

## B. NPV

\[
\begin{aligned}
\text{NPV} &= -I + \sum^N_{n=1}{\frac{C}{(1 + r)^n}}\\\\
\text{NPV}_{\text{I}} &= \$`r pretty(p12$npv.I, 0)`\\
\text{NPV}_{\text{II}} &= \$`r pretty(p12$npv.II, 0)`
\end{aligned}
\]

Using NPV, we see that project I is clearly the project to select.

## C. Why the difference between Profitability Index and NPV

This is an issue with scale. (c) Notice that cash flow I is more twice the size
of cash flow II in terms of initial investment and cash flows. The profitability
index is a ratio of the present value of the cash flows over its initial
investment. Because it is a ratio it doesn't consider scale. Net present value,
on the other hand, is an absolute value and therefore does consider scale.

---

# 13. Problems with IRR

Cutler Petroleum, Inc., is trying to evaluate a generation project with the
following cash flows:

```{r problem 13 table, echo=F}
p13 = new.env()
p13$year = c(0,1,2)
p13$cf = c(-85000000, 125000000, -15000000)
p13$tbl = tibble(p13$year, pretty(p13$cf, 0))
colnames(p13$tbl) = c("Year", "Cash Flow")
knitr::kable(p13$tbl)
p13$npv = npv(p13$cf, .1)
```

a. If the company requires a 10 percent return on its investments, should it
   accept this project? Why?

b. Compute the IRR for this project. How many IRRs are there? If you apply the
   IRR decision rule, should you accept the project or not? What's going on
   here?

## A. Return rate of 10\% should we accept this project

We should accept this project if the NPV is positive. This is the NPV rule and
using NPV is the best method to make such decisions.

* NPV: `r pretty(p13$npv)`

The net present value is positive so we should accept the project.

## B. Computing the IRR

There are two signs here so now we will have two IRRs.

---

# 14. Comparing investment criteria

Mario Brothers, a game manufacturer, has a new idea for an adventure game. It
can market the game either as a traditional board game or as an interactive DVD,
but not both. Consider the following cash flows of the two mutually exclusive
projects for Mario Brothers. Assume the discount rate for Mario Brothers is 10
percent.

```{r problem 14 table, echo=F}
p14 = new.env()
p14$year <- c(0,1,2,3)
p14$bg <- c(-750, 600, 450, 120)
p14$dvd <- c(-1800, 1300, 850, 350)
p14$tbl <- tibble(p14$year, p14$bg, p14$dvd)
colnames(p14$tbl) <- c("Year", "Board Game", "DVD")
knitr::kable(p14$tbl)
p14$pp.bg  = payback.period(p14$bg, fractional=T)
p14$pp.dvd = payback.period(p14$dvd, fractional=T)
p14$npv.bg = npv(p14$bg, .1)
p14$npv.dvd = npv(p14$dvd, .1)
p14$irr.bg = irr(p14$bg)
p14$irr.dvd = irr(p14$dvd)
p14$inc = p14$dvd - p14$bg
p14$irr.inc = irr(p14$inc)
```

a. Based on the payback period rule, which project should be chosen?
b. Based on the NPV, which project should be chosen?
c. Based on the IRR, which project should be chosen?
d. Based on the incremental IRR,w hich porject should be chosen?

## A. Calculating the payback period

* Board game payback period: `r pretty(p14$pp.bg)`

* DVD: `r pretty(p14$pp.dvd)`

Based on the payback period the board game should be chosen.

## B. Calculating the NPV

* Board game NPV: `r pretty(p14$npv.bg)`

* DVD: `r pretty(p14$npv.dvd)`

Based on the NPV rule, the DVD should be chosen.

## C. Calculating the IRR

* Board game NPV: `r pretty(p14$irr.bg, 4)`

* DVD: `r pretty(p14$irr.dvd, 4)`

Based on the IRR rule, the board game should be chosen.

## D. Calculating the incremental IRR

* Incremental IRR: `r pretty(p14$irr.inc, 4)`

Based on the incremental IRR rule, The DVD should be chosen.

Book answer:

* a. 1.33 years; 1.59 years
* b. \$257.51; \$347.26
* c. 33.79\%; 23.31\%
* d. 15.86\%

---

# 15. Profitability index versus NPV

Hanmi Group, a consumer electronics conglomerate, is reviewing its annual budget
in wireless technology. It is considering investments in three different
technologies to develop wireless communication devices. Consider the following
cash flows of the three independent projects for Hanmi. Assume the discount rate
for Hanmi is 10 percent. Further, Hanmi Group has only \$20 million to invest in
new projects this year.

```{r problem 15 table, echo=F}
p15 = new.env()
p15$year = c(0, 1, 2, 3)
p15$cdma = c(-8, 11, 7.5, 2.5)
p15$g4 = c(-12,10, 25, 20)
p15$wifi = c(-20, 18, 32, 20)
p15$tbl = tibble(p15$year, p15$cdma, p15$g4, p15$wifi)
colnames(p15$tbl) <- c("Year", "CDMA", "G4", "Wi-Fi")
knitr::kable(p15$tbl)
p15$cdma.pv = pv(p15$cdma, .1)
p15$g4.pv = pv(p15$g4, .1)
p15$wifi.pv = pv(p15$wifi, .1)
p15$cdma.pi = profitability.index(p15$cdma, .1)
p15$g4.pi = profitability.index(p15$g4, .1)
p15$wifi.pi = profitability.index(p15$wifi, .1)
p15$cdma.npv = npv(p15$cdma, .1)
p15$g4.npv = npv(p15$g4, .1)
p15$wifi.npv = npv(p15$wifi, .1)
```

a. Based on the profitability index decision rule, rank these investments.
b. Based on the NPV, rank these investments.
c. Based on your findings in (a) and (b), what would you recommend to the CEO of Hanmi Group and why?

## A. The profitability index

$$
\begin{aligned}
\text{PV} &= \sum^{N}_{n=1}{\frac{C_{n}}{(1 + r)^{n}}} \\
\text{PV}_{\text{CDMA}} &= `r pretty(p15$cdma.pv)`\\
\text{PV}_{\text{G4}} &= `r pretty(p15$g4.pv)`\\
\text{PV}_{\text{WIFI}} &= `r pretty(p15$wifi.pv)`\\\\
\text{Profitability Index} &= \frac{\text{PV}}{I}\\
\text{Profitability Index}_{\text{CDMA}} &= `r pretty(p15$cdma.pi)`\\
\text{Profitability Index}_{\text{G4}} &= `r pretty(p15$g4.pi)`\\
\text{Profitability Index}_{\text{WIFI}} &= `r pretty(p15$wifi.pi)`
\end{aligned}
$$

Based on the profitability index, we should choose G4.

## B. The net present value

$$
\begin{aligned}
\text{NPV} &= C_{0} + \sum^{N}_{n=1}{\frac{C_{n}}{(1 + r)^{n}}}\\
\text{NPV}_{\text{CDMA}} &= `r pretty(p15$cdma.npv)`\\
\text{NPV}_{\text{G4}} &= `r pretty(p15$g4.npv)`\\
\text{NPV}_{\text{WIFI}} &= `r pretty(p15$wifi.npv)`
\end{aligned}
$$

Based on the net present value, we should choose Wi-Fi.

## C. Recomendation

Because profitability index does not take into account scale, (c) we should rely
on the net present value. Since we have \$20 million to invest, we should choose
Wi-Fi.

---

# 16. Comparing investment criteria

Consider the following cash flows of two mutually exclusive projects for
AZ-Motorcars. Assume the discount rate for AZ-Motorcars is 10 percent.

```{r problem 16 table, echo=F}
p16 = new.env()
p16$year = c(0,1,2,3)
p16$azm = c(-450000, 320000, 180000, 150000)
p16$azf = c(-800000, 350000, 420000, 290000)
p16$tbl = tibble(p16$year, pretty(p16$azm, 0), pretty(p16$azf, 0))
colnames(p16$tbl) <- c("Year", "AZM Mini-SUV", "AZF Full-SUV")
knitr::kable(p16$tbl)
p16$azm.pb = payback.period(p16$azm, fractional=T)
p16$azf.pb = payback.period(p16$azf, fractional=T)
p16$azm.npv = npv(p16$azm, .1)
p16$azf.npv = npv(p16$azf, .1)
FinCal::irr(p16$azm)
p16$azm.irr = irr(p16$azm)
p16$azf.irr = irr(p16$azf)
p16$inc = p16$azf - p16$azm
p16$inc.irr = irr(p16$inc)
```

a. Based on the payback period, which project should be accepted?
b. Based on the NPV, which project should be accepted?
c. Based on the IRR, which project should be accepted?
d. Based on this analysis, is incremental IRR analysis necessary? If yes, please conduct the analysis.

## A. Payback period

* AZM Mini-SUV: `r pretty(p16$azm.pb)` years.

* AZF Full-SUV: `r pretty(p16$azf.pb)` years.

The AZM Mini-SUV has a lower payback period. Based on payback period, we should
choose the AZM Mini-SUV.

## B. NPV

* AZM Mini-SUV: \$`r pretty(p16$azm.npv)`

* AZF Full-SUV: \$`r pretty(p16$azf.npv)`

The NPV method is the most comprehensive method to use. At a 10\% discount rate,
we should choose the AZM Mini-SUV.

## C. IRR

* AZM Mini-SUV: `r pretty(p16$azm.irr*100)`\%

* AZF Full-SUV: `r pretty(p16$azf.irr*100)`\%

Based on IRR, we should choose AZM Mini-SUV. However, it should be noted that
there is an issue using the IRR method when it comes to scale.

## D. Incremental IRR

* Incremental (AZF Full-SUV - AZM Mini-SUV): `r pretty(p16$inc.irr*100)`\%

Based on the incremental IRR, we should choose AZF Full-SUV.

---

# 17. Comparing investment criteria

The treasurer of Amaro Canned Fruits, Inc., has projected the cash flows of
projects $A$, $B$, and $C$ as follows:

```{r problem 17 table, echo=F}
p17 = new.env()
p17$year = c(0,1,2)
p17$project.A = c(-150000, 110000, 110000)
p17$project.B = c(-300000, 200000, 200000)
p17$project.C = c(-150000, 120000,  90000)
p17$tbl = tibble(p17$year, pretty(p17$project.A, 0),
                 pretty(p17$project.B, 2), pretty(p17$project.C, 0))
colnames(p17$tbl) <- c("Year", "Project A", "Project B", "Project C")
knitr::kable(p17$tbl)
p17$project.A.pi = profitability.index(p17$project.A, .12)
p17$project.B.pi = profitability.index(p17$project.B, .12)
p17$project.C.pi = profitability.index(p17$project.C, .12)
p17$project.A.npv = npv(p17$project.A, .12)
p17$project.B.npv = npv(p17$project.B, .12)
p17$project.C.npv = npv(p17$project.C, .12)
```

Suppose the relevant discount rate is 12 percent a year.

a. Compute the profitability index for each of the three projects.

b. Compute the NPV for each of the three projects.

c. Suppose these three projects are independent. Which project(s) should Amaro
   accept based on the profitability index rule?

d. Suppose these three projects are mutually exclusive. Which project(s) should
   Amaro accept based on the profitability index rule?

e. Suppose Amaro's budget for these projects is \$450,000. The projects are not
   divisible. Which project(s) should Amaro accept?

## A. Profitability Index

$$
\begin{aligned}
\text{I} &= \text{initial investment}\\
\text{PV} &= \sum^{N}_{n=1}{\frac{C_{n}}{(1 + r)^{n}}}\\
\text{Profitability Index} &= \frac{C}{I}\\\\
\text{PI}_{A} &= `r pretty(p17$project.A.pi)`\\
\text{PI}_{B} &= `r pretty(p17$project.B.pi)`\\
\text{PI}_{C} &= `r pretty(p17$project.C.pi)`
\end{aligned}
$$

## B. NPV

$$
\begin{aligned}
\text{NPV} &= C_0 + \sum^{N}_{n=1}{\frac{C_n}{(1 + r)^{n}}}\\\\
\text{NPV}_A &= \$`r pretty(p17$project.A.npv)`\\
\text{NPV}_B &= \$`r pretty(p17$project.B.npv)`\\
\text{NPV}_C &= \$`r pretty(p17$project.C.npv)`
\end{aligned}
$$

## C. Independent projects and the profitability index

All three projects are profitable (> 1), so all three projects should be
accepted.

## D. Mutually exclusive and the profitability index

If the projects are mutually exclusive and you can only choose one based on
profitability index alone, then you would choose project A. Keep in mind that
the profitability index does not take into account the differences in the
projects scale. This can be shown in the more comprehensive method of capital
decision making, NPV, which shows that project B should be chosen. Project B is
the correct answer.

## E. Project budget of \$450,000

We can only choose two of the three projects with this budget. Basing the
decision on the NPV of the projects you should choose both projects A and B.
All three projects have positive NPVs meaning they are all profitable. But
project A and project B yield higher returns.

---

# 18. Comparing investment criteria

Consider the following cash flows of two mutually exclusive projects for Tokyo
Rubber Company. Assume the discount rate for Tokyo Rubber Company is 10 percent.

```{r problem 18 table, echo=F}
p18 = new.env()
p18$year = c(0, 1, 2, 3)
p18$dry = c(-1700000, 1100000, 900000, 750000)
p18$solvent = c(-750000, 375000, 600000, 390000)
p18$tbl = tibble(p18$year, pretty(p18$dry), pretty(p18$solvent))
colnames(p18$tbl) = c("Year", "Dry Prepreg", "Solvent Prepreg")
knitr::kable(p18$tbl)
p18$dry.pb = payback.period(p18$dry, fractional=T)
p18$solvent.pb = payback.period(p18$solvent, fractional=T)
p18$dry.npv = npv(p18$dry, .1)
p18$solvent.npv = npv(p18$solvent, .1)
p18$dry.irr = irr(p18$dry)
p18$solvent.irr = irr(p18$solvent)
p18$inc = p18$dry - p18$solvent
p18$inc.irr = irr(p18$inc)
```

a. Based on the payback period, which project should be taken?
b. Based on the NPV, which project should be taken?
c. Based on the IRR, which project should be taken?
d. Based on this analysis, is incremental IRR analysis necessary? If yes, please conduct the analysis.

## A. Payback period

* Dry Prepreg: `r pretty(p18$dry.pb)`

* Solvent Prepreg: `r pretty(p18$solvent.pb)`

For the two mutually exclusive projects, the Solvent Prepreg project should be
chosen based on the payback period. However, the payback period has issues
because it doesn't account for the present value of the cash flows.

## B. NPV

* Dry Prepreg: \$`r pretty(p18$dry.npv, 0)`

* Solvent Prepreg: \$`r pretty(p18$solvent.npv, 0)`

The NPV method is the most comprehensive method and should be used when making
capital decisions. Based on the NPV method, we should choose the Dry Prepreg
project.

## C. IRR

* Dry Prepreg: `r pretty(p18$dry.irr*100)`\%

* Solvent Prepreg: `r pretty(p18$solvent.irr*100)`\%

Based on the IRR method we should choose the Solvent Prepreg project which has
the larger IRR. However, the IRR method does not take into account the scale of
the project. This means that we should explore this further with the incremental
IRR method.

## D. Incremental IRR

* Incremental (Dry Prepreg - Solvent Prepreg): `r pretty(p18$inc.irr*100)`\%

Based on the incremental IRR, We should choose the Dry Prepreg project because
this larger project has a positive incremental IRR. 

---

# 19. Comparing investment criteria

Consider two mutually exclusive new product launch projects that Nagano Golf is
considering. Assume the discount rate for Nagano Golf is 15 percent.

## Project A:

* Nagano NP-30.

* Professional clubs that will take an initial investment of \$550,000 at
  time 0.

* Next five years (Years 1--5) of sales will generate a consistent cash flow of
  \$185,000 per year.

* Introduction of new product at Year 6 will terminate further cash flows from
  this project.

## Project B:

* Nagano NX-20.

* High-end amateur clubs that will take an initial investment of \$350,000 at
  time 0.

* Cash flows at Year 1 is \$100,000. In each subsequent year cash flow will grow
  at 10 percent per year.

* Introduction of new product at Year 6 will terminate further cash flows from
  this project.

```{r question 19 table, echo=F}
p19 = new.env()
p19$year = c(0, 1, 2, 3, 4, 5)
p19$np.30 = c(-550000, rep(185000, 5))
p19$nx.20 = c(-350000, 100000*1.1^(0:4))
p19$tbl = tibble(p19$year, pretty(p19$np.30), pretty(p19$nx.20))
colnames(p19$tbl) <- c("Year", "NP-30", "NX-20")
knitr::kable(p19$tbl)
p19$np.30.pb = payback.period(p19$np.30)
p19$nx.20.pb = payback.period(p19$nx.20)
p19$np.30.irr = irr(p19$np.30)
p19$nx.20.irr = irr(p19$nx.20)
p19$np.30.pi = profitability.index(p19$np.30, .15)
p19$nx.20.pi = profitability.index(p19$nx.20, .15)
p19$np.30.npv = npv(p19$np.30, .15)
p19$nx.20.npv = npv(p19$nx.20, .15)
```

Please fill in the following table:

| | NP-30 | NX-20 | Implications |
|:---|:---:|:---:|:---|
| Payback | `r pretty(p19$np.30.pb)` | `r pretty(p19$nx.20.pb)` | Choose NP-30 |
| IRR | `r pretty(p19$np.30.irr*100)`\% | `r pretty(p19$nx.20.irr*100)`\% | Choose NX-20 |
| PI | `r pretty(p19$np.30.pi)` | `r pretty(p19$nx.20.pi)` | Choose NX-20 |
| NPV | \$`r pretty(p19$np.30.npv)` | `r pretty(p19$nx.20.npv)` | Choose NP-30 (Correct choice) |

---

# 20. NPV and Multiple IRRs

You are evaluating a project that costs \$75,000 today. The project has an
inflow of \$155,000 in one year and an outflow of \$65,000 in two years. What
are the IRRs for the project? What discount rate results in the maximum NPV for
this project? How can you determine that this is the maximum NPV?

```{r problem 20, echo=F}
p20 = new.env()
p20$f = function(r) -75 + 155/(1 + r) - 65 / (1 + r)^2
p20$df = function(r) 130/(1 + r)^3 - 155/(1 + r)^2
p20$tbl = tibble(
    r = seq(-.5, .6, by=.01),
    v = p20$f(r)
)
p20$x.intercept = c(
    uniroot(p20$f, c(-.3, -.5))$root,
    uniroot(p20$f, c(.4, .5))$root
)
p20$vmax = uniroot(p20$df, c(-.2, 0))$root
ggplot(p20$tbl, aes(x=r, y=v)) +
    geom_line() +
    geom_hline(yintercept = 0, col='red') +
    geom_vline(xintercept = p20$x.intercept, col='blue') +
    geom_hline(yintercept = p20$f(p20$vmax))
```

Because of the sign changes, there are two IRRs.

## Set up the equation

\[
\begin{aligned}
\text{CF} &= (-75,000, 155,000, -65,000)\\\\
\text{NPV}(r) &= -75,000 + \frac{155,000}{(1 + r)} - \frac{65,000}{(1 + r)^2}
\end{aligned}
\]

## We solve for the IRR. Let $x = \frac{1}{1 + r}$.

\[
\begin{aligned}
-75,000 + 155,000x - 65,000x^2 = 0\\
65x^2 - 155x + 75 = 0
\end{aligned}
\]

The roots are:

\[
x_1 \approx 1.70975477,\ \ \ \ x_2 \approx 0.67486061.
\]

We convert:

\[
\begin{aligned}
r_1 &= \frac{1}{x_1} - 1 \approx `r p20$x.intercept[1]`\\
r_2 &= \frac{1}{x_2} - 1 \approx  `r p20$x.intercept[2]`
\end{aligned}
\]

## Find the discount rate that gives the maximum NPV

\[
\begin{aligned}
\text{NPV}(r) &= -75,000 + 155,000(1 + r)^{-1} - 65,000(1 + r)^{-2}\\
\frac{d\text{NPV}}{dr} &= -155,000(1 + r)^{-2} + 130,000(1 + r)^{-3}
\end{aligned}
\]

Equate $d/dr = 0$

\[
\begin{aligned}
-155,000(1 + r)^{-2} + 130,000(1 + r)^{-3} &= 0\\
-155,000(1 + r) + 130,000 &= 0\\
155,000 + 155,000r &= 130,000\\
155,000r &= -25,000\\
r &= \frac{-25}{155} \approx `r p20$vmax`
\end{aligned}
\]

---

# 21. Payback and NPV

An investment under consideration has a payback of six years and a cost of
\$434,000. If the required return is 12 percent, what is the worst-case NPV? The
best-case NPV? Explain. Assume the cash flows are conventional.

```{r problem 21, echo=F}
p21 = new.env()
p21$npv.worst.case = npv(c(-434e3, rep(0, 5), 434e3), .12)
p21$npv.best.case = npv(c(-434e3, 434e3, rep(0, 5)), .12)
```

## Wort-case scenario

Under the worst-case scenario all of the cash flow occurs as late as possible
  (i.e. year 6).

\[
\text{NPV} = -434,000 + \frac{434,000}{(1.12)^6}
    = `r pretty(p21$npv.worst.case, 0)`
\]

## Best-case scenario

The best-case scenario is that all cash flows occur as early as possible (i.e.
year 1). The cash flows are _front-loaded_.

\[
\text{NPV} = -434,000 + \frac{434,000}{(1.12)}
    = `r pretty(p21$npv.best.case, 0)`
\]

---

# 22. Multiple IRRs

This problem is useful for testing the ability of financial calculators and
computer software. Consider the following cash flows. How many different IRRs
are there? (_Hint:_ Search between 20 percent and 70 percent.) When should we
take this project.

```{r problem 22, echo=F}
p22 = new.env()
p22$year = c(0, 1, 2, 3, 4)
p22$cf = c(-1008, 5724, -12140, 11400, -4000)
p22$tbl = tibble(
    Year = p22$year,
    `Cash Flow` = pretty(p22$cf, 0)
)
knitr::kable(p22$tbl)
p22$f <- function(r) sum(p22$cf / (1 + r)^(0:(length(p22$cf)-1)))
p22$v = c()
for(r in (p22$r <- seq(.2, .7, by=.01))){
  p22$v = c(p22$v, p22$f(r))
}
p22$res = tibble(x=p22$r, y=p22$v)
p22$cf.discounted = p22$cf / (1 + r)^(0:(length(p22$cf)-1))
p22$zeros = c(
    uniroot(p22$f, c(.2, .3))$root,
    uniroot(p22$f, c(.3, .35))$root,
    uniroot(p22$f, c(.4, .45))$root,
    uniroot(p22$f, c(.65, .7))$root
)
p22$npv.1 = npv(p22$cf, p22$zeros[1])
```

We can calculate the number of IRRs from the number of sign changes in the cash
flows. There are four sign changes therefore there are four IRRs. This can be
represented graphically.

```{r problem 22; graph, echo=F}
ggplot(p22$res, aes(x=x, y=y)) +
    geom_line() +
    geom_hline(yintercept=0, col='red')
```

Because of the multiple IRRs, we can't use the IRR method for capital budgeting.
We can use the modified IRR. However, we need a discount rate to do the
calculations.

The graph shows two regions where the rate is positive. These are the regions
where we should take this project. These regions are between
[`r pretty(p22$zeros[1])`, `r pretty(p22$zeros[2])`] and
[`r pretty(p22$zeros[3])`, `r pretty(p22$zeros[4])`].

We can find the local maximas for each region.

\[
-1008 + 5724 \times (1 + r)^{-1} - 12,140 \times (1 + r)^{-2}
    + 11,400 \times (1 + r)^{-3} - 4000 \times (1 + r)^{-4} = 0
\]

---

# 23. NPV Valuation

The Yurdone Corporation wants to set up a private cemetery business. According
to the CFO, Barry M. Deep, business is "looking up." As a result, the cemetery
project will provide a net cash inflow of \$290,000 for the firm during the
first year, and the cash flows are projected to grow at a rate of 5 percent per
year forever. The project requires an initial investment of \$3,900,000.

a. If Yurdone requires a return of 11 percent on such undertakings, should the
   cemertery business be started?

b. The company is somewhat unsure about the assumption of a growth rate of 5
   percent in its cash flows. At what constant growth rate would the company
   just break even if it still required a return of 11 percent on investment?

```{r problem 23, echo=F}
p23 = new.env()
p23$pv.perpetuity = 290e3 / (.11 - .05)
p23$investment = 3.9e6
p23$npv = -p23$investment + p23$pv.perpetuity
p23$f = function(g) 290e3 / (.11 - g) - 3.9e6
p23$g = uniroot(p23$f, c(0, .05))$root
```

## A. Calculate NPV for a growing perpetuity investment at 11\%

We are asked to calculate the NPV of a growing perpetuity investment

\[
\begin{aligned}
\text{NPV} &= -\text{Investment} + \text{PV}_{\text{Perpetuity}}\\
    &= -\$3,900,000 + \frac{\$290,000}{(.11 - .05)}\\
    &= \$`r pretty(p23$npv, 0)`
\end{aligned}
\]

Yes, Yurdone Corporation should invest in this project. It pays very well.

## B. Break even growth rate for a required return of 11\%

If Yurdone wants to know the growth rate that it need to break even for this
project we need to find $g$ in the following equation:

\[
-3,900,000 + \frac{290,000}{(.11 - g)} = 0
\]

Solving for $g$ we get: `r pretty(p23$g*100)`\%

---

# 24. Calculating IRR

The Utah Mining Corporation is set to open a gold mine near Provo, Utah.
According to the treasurer, Monty Goldstein, "This is a golden opportunity." The
mine will cost \$2,400,000 to open and will have an economic life of 11 years.
It will generate a cash inflow of \$345,000 at the end of the first year, and
the cash inflows are projected to grow at 8 percent per year for the next 10
years. After 11 years, the mine will be abandoned. Abandonment costs will be
\$400,000 at the end of year 11.

a. What is the IRR for the gold mine?

b. The Utah Mining Corporation requires a return of 10 percent on such
   undertakings. Should the mine be opened?

```{r problem 24, echo=F}
p24 = new.env()
p24$I = 2.4e6
p24$f <- function(r)
    -2.4e6 + 345e3 * sum((1.08)^(0:10)/(1 + r)^(1:11)) - 400e3 / (1 + r)^11
p24$r = uniroot(p24$f, c(0,.5))$root
p24$v = c()
for(r in (p24$r.s <- seq(0, .5, by=.01))){
    p24$v = c(p24$v, p24$f(r))
}
p24$res = tibble(x=p24$r.s, y=p24$v)
gplot <- ggplot(p24$res, aes(x=x, y=y)) +
    geom_line() +
    geom_hline(yintercept=0, col='red')
```

## A. Calculating the IRR

We need to solve the following equation

\[
\text{NPV} = -2,400,000 + 345,000 \times \sum^{11}_{t=1}{
    \frac{(1.08)^{t-1}}{(1 + r)^{t}}} - \frac{400,000}{(1 + r)^{11}}
\]

Solving for r we get: `r pretty(p24$r*100)`\%.

```{r problem 24; graph, echo=F}
gplot
```

## B. Required return is 10\% should the project be undertaken?

Yes. With an IRR of `r pretty(p24$r*100)`\% > 10\%, the project should be
undertaken.

---

# 25. NPV and IRR

Butler International Limited is evaluating a project in Erewhon. The project
will create the following cash flows:

```{r problem 25, echo=F}
p25 = new.env()
p25$year = seq(0, 4, by=1)
p25$cf = c(-950e3, 285e3, 345e3, 415e3, 255e3)
p25$tbl = tibble(
    Year = p25$year,
    `Cash Flow` = pretty(p25$cf)
)
knitr::kable(p25$tbl)
```

All cash flows will occur in Erewhon and are expressed in dollars. In an attempt
to improve its economy, the Erewhonian government has declared that all cash
flows created by a foreign company are "blocked" and must be reinvested with the
government for one year. The reinvestment rate for these funds is 4 percent. If
Butler uses a required return of 11 percent on this project, what are the NPV
and IRR of the project? Is the IRR you calculated the MIRR of the project? Why
or why not?

---
