---
title: "Discounted Cash Flow Valuation"
subtitle: "Text"
author: "Ross, Westerfield, Jaffe"
output: 
  pdf_document:
   toc: true 
---

```{r setup, echo=F, message=F}
require(data.table)
require(kableExtra)
pretty <- function(x, n = 2) format(round(x, n), big.mark=",", nsmall = n)
percent <- function(x, a = 0.01) scales::percent(x, accuracy = a)
table <- function(df, caption=NULL) {
  knitr::kable(df, caption=caption, format = "latex", booktabs=T) %>%
    kable_styling(latex_options = c("HOLD_position"))
options(scipen = 999)
}
```

# Chapter introduction

The signing of big-name athletes is often accompanied by great fanfare, but the
numbers are often misleading. For example, in late 2010, catcher Victor Martinez
reached a deal with the Detroit Tigers, signing a contract with a reported value
of \$50 million. Not bad, especially for someone who makes a living using the
"tools of ignorance" (jock jargon for a catcher's equipment). Another example is
the contract signed by Jayson Werth of the Washington Nationals, which had a
stated value of \$126 million.

It looks like Victor and Jayson did pretty well, but then there was Carl
Crawford, who signed to play in fron of Boston's Red Sox nation. Carl's contract
has a stated value of \$142 million, but this amount was actually payable over
several years. The contract consisted of a \$6 million signing bonus, along with
\$14 million in the first year plus \$122 million in future salary to be paid in
the years 2011 through 2017. Victor's and Jayson's payments were similarly
spread over time. Because all three contracts called for payments that are made
at future dates, we must consider the time value of money, which means none of
these players received the quoted amounts. How much did they really get? This
chapter gives you the "tools of knowledge" to answer this question.

# List of topics

1. Valuation: The one-period case

2. The multiperiod case

   * Future value and compounding
   * The power of compounding: A digression
   * Present value and discounting
   * Finding the number of periods
   * The algebraic formula

3. Compounding periods

   * Distinction between stated annual interest
   * Rate and effective annual rate
   * Compounding over many years
   * Continuous compounding

4. Simplifications

   * Perpetuity
   * Growing perpetuity
   * Annuity
   * Growing annuity

5. Loan amortization

6. What is a firm worth?

# Notes

## Valuation: The one-period case

* __Future value__ or __compound value__ is the value of a sum after investing
over one or more periods.

* __Present value__, the current value of a sum that will be received in the
future.

\begin{center}
\textbf{Present value of investment}
\end{center}

$$
\text{PV} = \frac{C_1}{1 + r}\tag{4.1}
$$

This is called _present value analysis_ and it will show the present value of a
cash flow ($C_1$) one period in the future. The variable $r$ is the
_discount rate_.

The net present value (NPV) is the present value of future cash flows minus the
present value of the cost of the investment.

\begin{center}
\textbf{Net present value of investment}
\end{center}

$$
\text{NPV} = - \text{Cost} + \text{PV}\tag{4.2}
$$

When determining the risk rate ($r$), you would choose a return that is
available in the market on other investments of the same risk.

## The multiperiod case

### Future value and compounding

After the first period in investing, you have a choice: (1) you can choose to
take the profits of the investment, or (2) you can reinvest both the principle
and interest and accumulate compounding.

The compounding interest is computed for the number of periods

$$
(1 + r)^t
$$

where $t$ is the number of compounding periods.

Simple interest does not take account the interest and thus is computed:

$$
r^t
$$

When cash is invested at compound interest, each interest payment is reinvested.
With simple interest, the interest is not reinvested.

The general formula for an investment over many periods can be written as:

\begin{center}
\textbf{Future value of an investment}
\end{center}

$$
\text{FV} = C_0 \times (1 + r)^T\tag{4.3}
$$

where $C_0$ is the cash to be invested at Date 0 (i.e., today), $r$ is the
itnerest rate per period, and $T$ is the number of periods over which the cash
is invested.

### The power of compounding: A digression

The real affect of compounding can be seen at large intervals. For example if
you invested one dollar in the U.S. stock market in 1929, it would be worth
\$2,982.24.24 in 2010. This is 9.87 percent compounded annually for 85
years---that is, $(1.0987)^{85} = \$2,982.24$, ignoring a small rounding error.

### Present value and discounting

The multiperiod equation for the present value of future cash flows is

$$
\text{PV} = \frac{C_T}{(1 + r)^T}\tag{4.4}
$$

The process of calculating the present value of a future cash flow is called
__discounting__.

### Finding the number of periods

When solving for the number of periods you can use regression:

$$
\frac{\text{PV}}{\text{FV}} = (1 + r)^T
$$

### The algebraic formula

$$
\text{NPV} = - C_0 + \frac{C_1}{1 + r} + \frac{C_2}{(1 + r)^2}
  + \cdots + \frac{C_T}{(1 + r)^T}
  = - C_0 + \sum^{T}_{i=1}{\frac{C_i}{(1 + r)^i}}\tag{4.5}
$$

## Compounding Periods

Sometime compounding occurs more frequently than just one year.

Other types of compounding:

* Semi-annual compounding

* Quarterly compounding

* Monthly compounding

The general formula for the present value factor:

$$
C_0 \left( 1 + \frac{r}{m} \right)^m
$$

In this formula $r$ is the __stated annual interest rate (SAIR)__ and $m$ is the
number of compounding periods in the year.

The stated annual interest rate is the annual interest rate without
consideration of compounding. Banks and other financial institutions may use
other names for the stated annual interest rate. __Annual percentage rate
(APR)__ is perhaps the most common synonym.

### Distinction betwen stated annual interest rate and effective annual rate

The stated annual interest rate (SAIR) is only meaningful when the compounding
interval is given. However the effective annual rate (EAR) is meaningful without
the compounding interval.

When interest rates are large the difference between the APR and the EAR can be
big.

### Compounding over many years

When applying compounding over many years the general form of the present value
factor has the form:

$$
\left( 1 + \frac{r}{m} \right)^{mT}\tag{4.8}
$$

### Continuous compounding

$$
C_0 \times e^{rT}\tag{4.9}
$$

## Simplifications

The following sections show simplifications to the basic formulas.

### Perpetuity

For a perpetuity the equation is:

$$
\text{PV} = \frac{C}{1 + r} + \frac{C}{(1 + r)^2}
  + \frac{C}{(1 + r)^3} + \cdots
$$

We will state, without proof that the formula reduces to:

$$
\text{PV} = \frac{C}{r}\tag{4.10}
$$

### Growing perpetuities

For a growing perpetuity the equation is

$$
\text{PV} = \frac{C}{1 + r} + \frac{C \times (1 + g)}{(1 + r)^2}
  + \frac{C \times (1 + g)^2}{(1 + r)^3} + \cdots
  + \frac{C \times (1 + g)^{N - 1}}{(1 + r)^N} + \cdots
$$

For the case when $r > g$, the formula reduces to:

$$
\text{PV} = \frac{C}{r - g}\tag{4.12}
$$

A few important points:

1. _The numerator_: The numerator is the cash flow one period hence, not at
   date 0.

2. _The discount rate and the growth rate_: The discount rate $r$ must be
   greater than the growth rate $g$ for the growing perpetuity formula to work.
   Consider the case in which the growth rate approaches the interest rate in
   magnitude. Then, the denominator in the growing perpetuity formula gets
   infinitesimally small and the present value grows infinitely large. The
   present value is in fact undefined when $r$ is less than $g$.

3. _The timing assumption_: Cash generally flows into and out of real-world
   firms both randomly and nearly continuously. However, the equation assumes
   that cash flows are received and disbursed at regular and discrete points in
   time.

### Annuity

You can derive the formula for an annuity from the formula of two perpetuities.
Mainly:

> For $C_1$ which starts at $t=0$ and $C_2$ which starts at $t=T$. The formula
> for an annuity is:

$$
\text{PV}_{\text{annuity}} = \frac{C_1}{r} - \frac{C_2}{r}
  \left[ \frac{1}{(1 + r)^T} \right]\tag{4.14}
$$

This simplifies to:

\begin{center}
\textbf{Formula for present value of annuity}
\end{center}

$$
\text{PV}_{\text{annuity}} = C
  \frac{\left( 1 - (1 + r)^{-T} \right)}{r}\tag{4.15}
$$

The term used to compute the present value of the stream of level payments, $C$,
for $T$ years is called a __present value interest factor for annuitites__.

### Trick 1: A delayed annuity

One of the tricks in working with annuities and perpetuities is getting the
timing right.

#### Trick 2: Annuity due

The annuity formula of equation 4.15 assumes that the first annuity  payment
begins a full period hence. This type of annuity is sometimes called an
_annuity arrears_ or _ordinary annuity_. What happens if the annuity begins
today---in other words, at Date 0?

#### Trick 3: The infrequent annuity

Here the trick is to determine the interest rate over the period.

#### Trick 4: Equating present value of two annuities

### Growing annuity

\begin{center}
\textbf{Formula for present value of growing annuity}
\end{center}

$$
\text{PV} = C \left[\frac{1}{r - g} - \frac{1}{r - g}
  \times \left(\frac{1 + g}{1 + r}\right)^T\right]
  = C \left[\frac{1 - \left(\frac{1 + g}{1 + r}\right)^T}{r - g}\right]
  \tag{4.17}
$$

---

# Examples

---

## Example 4.1: Present value

Diane Badame, a financial analyst at Kaufman & Broad, a leading real estate
firm, is thinking about recommending that Kaufman & Broad invest in a piece of
land that costs \$85,000. She is certain that next year the land will be worth
\$91,000, a sure \$6,000 gain. Given that the interest rate in similar
alternative investments is 10 percent, should Kaufman & Broad undertake the
investment in land? Diane's choice is described below with the cash flow time
chart.

A moment's thought should be all it takes to convince her that this is not an
attractive business deal. By investing \$85,000 in the land, she will have
\$91,000 available next year. Suppose, instead, that Kaufman & Broad puts the
same \$85,000 into similar alternative investments. At the interest rate of 10
percent, this \$85,000 would grow to:

$$
(1 + .10) \times \$85,000 = \$93,500
$$

next year.

It would be foolish to buy the land when investing the same \$85,000 in similar
alternative investments would produce an extra \$2,500 (that is, \$93,500 from
the bank minus \$91,000 from the land investment). This is a future value
calculation.

Alternatively, she could calculate the present value of the sale price next year
as:

$$
\text{Present value} = \frac{\$91,000}{1.10} = \$82,727.27
$$

Because the present value of next year's sales price is less than this year's
purchase price of \$85,000, present value analysis also indicates that she
should not recommend purchasing th property.

---

## Example 4.2: Uncertainty and valuation

Professional Artworks, Inc. is a firm that speculates in modern paintings. The
manager is thinking of buying an original Picasso for \$400,000 with the
intention of selling it at the end of one year. The manager expects that the
painting will be worth \$480,000 in one year. The relavent cash flows are
depicted below.

<<FIGURE 4.3>>

Of course, this is only an expectation---the painting could be worth more or
less than \$480,000. Suppose the guaranteed interest rate granted by banks is 10
percent. Should the firm purchase the piece of art?

Our first thought might be to discount at the interest rate, yielding:

$$
\frac{\$480,000}{1.10} = \$436,364
$$

Because \$436,364 is greater than \$400,000, it looks at first glance as if the
painting should be purchased. However, 10 percent is the return one can earn on
a low risk investment. Because the painting is quite risky, a higher discount
rate is called for: The manager chooses a rate of 25 percent to reflect this
risk. In other words, he argues that a 25 percent expected return is fair
compensation for an investment as risky as this painting.

The present value of the painting becomes:

$$
\frac{\$480,000}{1.25} = \$384,000
$$

Thus, the manager believes that the painting is currently overpriced at
\$400,000 and does not make the purchase.

---

## Example 4.3: Interest on interest

Suh-Pyng Ku has put \$500 in a savings account at the First National Bank of
Kent. The account earns 7 percent, compounded annually. How much will ms. Ku
have at the end of three years? The answer is:

$$
\$500 \times 1.07 \times 1.07 \times 1.07 = \$500 \times (1.07)^3 = \$612.52
$$

---

## Example 4.4: Compound growth

Jay Ritter invested \$1,000 in the stock of the SDH Company. The company pays a
current dividend of \$2, which is expected to grow by 20 percent per year for
hte next two years. What will the dividend of the SDH Company be after two
years? A simple calculation gives:

$$
\$2 \times (1.20)^2 = \$2.88
$$

---

## Example 4.5: Finding the rate

Carl Voigt, who recently won \$10,000 in the lottery, wants to buy a car in five
years. Carl estimates that the car will cost \$16,105 at that time. His cash
flows are displayed in the figure below.

What interest rate must he earn to be able to afford the car?

The ratio of purchase price to initial cash is:

$$
\frac{\$16,105}{\$10,000} = 1.6105
$$

Thus, he must earn an interest rate that allows \$1 to become \$1.6105 in five
years. Table A.3 tells us that an interest rate of 10 percent will allow him to
purchase the car:

We can express the problem algebraically as:

$$
\$10,000 \times (1 + r)^5 = \$16,105
$$

where $r$ is the interest rate needed to purchase the car. Because
\$16,105/\$10,000 = 1.6105, we have:

$$
\begin{aligned}
(1 + r)^5 &= 1.6105\\
r &= 10\%
\end{aligned}
$$

Either the table, a spreadsheet, or a hand calculator lets us solve for $r$.

---

## Example 4.6: How much for that island?

Some people have said that it was the best real estate deal in history. Peter
Minuit, director general of New Netherlands, the Dutch West India Company's
colony in North America, in 1626 allegedly bought Manhattan Island for 60
guilders' worth of trinkets from native Americans. By 1667, the Dutch were
forced by the Britsh to exchnage it for Suriname
(perhaps the worst real estate deal ever). This sounds cheap; but did the Dutch
really get the better end of the deal? It is reported that 60 guilders was worth
about \$24 at the prevailing exchange rate. If the native Americans had sold the
trinkets at a fair market value and invested the \$24 at 5 percent (tax free),
it would now, about 385 years later, be worth more than \$3.45 billion. Today,
Manhattan is undoutedly worth more than \$3.45 billion, so at a 5 percent rate
of return the native Americans got th eworst of the deal. However, if invested
at 10 percent, the amount of money they received would be worth about:

$$
\$24(1 + r)^T = 24 \times 1.1^{385} \cong \$207\ \text{quadrillion}
$$

This is a lot of money. In fact, \$207 quadrillion is more than all the real
estate in the world is worth today. Note that no one in the history of the world
has ever been able to find an investment yielding 10 percent every year for 385
years.

---

## Example 4.7: Multiperiod discounting

Bernard Dumas will receive \$10,000 three years from now. Bernard can earn 8
percent on his investments, so the appropriate discount rate is 8 percent. What
is the present value of his future cash flow? The answer is:

$$
\begin{aligned}
\text{PV} &= \$10,000 \times \left(\frac{1}{1.08}\right)^3\\
  &= \$10,000 \times .7938\\
  &= \$7,938
\end{aligned}
$$

When his investments grow at an 8 percent rate of interest, Bernard Dumas is
equally inclined toward receivng \$7,938 now and receiving \$10,000 in three
years' time. After all, he could convert the \$7,938 he receives today into
\$10,000 in three years by lending it at an interest rate of 8 percent.

Bernard Dumas could have reached his present value calculation in one of several
ways. The computation could have been done by hand, by calculator, with a
spreadsheet, or with the help of Table A.1, which appears in the back of the
text. This table presents the _present value of_ \$1 to be received after $T$
periods. We use the table by locating the appropriate interest rate on the
horizontal and the appropriate number of periods on the vertical. For example,
Bernard Dumas would look at the following portion of Table A.1.

The appropriate present value factor is .7938.

---

### Example 4.8: Finding the rate

A customer of the Chaffkin Corp. wants to buy a tugboat today. Rather than
paying immediately, he will pay \$50,000 in three years. It will cost the
Chaffkin Corp. \$38,610 to build the tugboat immediately. The relevant cash
flows to Chaffkin Corp. are displayed in the figure below. What interest rate
would the Chaffkin Corp. charge to neither gain nor lose on the sale?

The ratio of construction cost (present value) to sales price (future value) is:

$$
\frac{\$38,000}{\$50,000} = .7722
$$

We must determine the interest rate that allows \$1 to be received in three
years to have a present value of \$.7722. Table A.1 tells us that 9 percent is
that interest rate.

---

## Example 4.9: Waiting for Godot

You've been saving up to buy the Godot Company. The total cost will be \$10
million. You currently have about \$2.3 million. If you can earn 5 percent on
your money, how long will you have to wait? At 16 percent, how long must you
wait?

At 5 percent, you'll have to wait a long time. From the basic present value
equation:

$$
\begin{aligned}
\$2.3 \text{million} &= \$10 \text{million} / 1.05^t\\
1.05^t &= 4.35\\
t &= 30 \text{years}
\end{aligned}
$$

At 16 percent, things are a little better. Verify for your self that it will
take about 10 years.

---

## Example 4.10: Cash flow valuation

Kyle Mayer has won the Kentucky State Lottery and will receive the following set
of cash flows over the next two years:

| Year | Cash flow |
|:---:|---:|
| 1 | \$20,000 |
| 2 | 50,000 |

Mr. Mayer can currently earn 6 percent in his money market account, so the
appropriate discount rate is 6 percent. The present value of the cash flows is:

$$
\begin{aligned}
\text{CF}_{\text{year 1}} &= \$20,000 \times \frac{1}{1.06}
  = \$18,867.9\\
\text{CF}_{\text{year 2}} &= \$50,000 \times \left( \frac{1}{1.06} \right)^2
  = \$44,499.8\\\\
\text{Total} &= \$18,867.9 + \$44,499.8 = \$63,367.7
\end{aligned}
$$

---

## Example 4.11: NPV

Finance.com has an opportunity to invest in a new high-speed computer that costs
\$50,000. The computer will generate cash flows (from cost savings) of \$25,000
one year from now, \$20,000 two years from now, and \$15,000 three years from
now. The computer will be worthless after three years, and no additional cash
flows will occur. Finance.com has determined that the appropriate discount rate
is 7 percent for this investment. Should Finance.com make this investment in a
new high-speed computer? What is the net present value of the investment?

The cash flows and present value factors of the proposed computer are as
follows:

```{r example 4.11, echo=F}
ex4.11 <- new.env()
ex4.11$cf <- c(-50000, 25000, 20000, 15000)
ex4.11$discount.rate <- .07
ex4.11$pv.factor <- round(
        (1 + ex4.11$discount.rate)^(-(0:(length(ex4.11$cf)-1))), 4)
ex4.11$pv <- ex4.11$cf * ex4.11$pv.factor
ex4.11$npv <- sum(ex4.11$pv)
ex4.11$df.1 <- data.frame(Cash.flows=ex4.11$cf,
        Present.Value.Factor=ex4.11$pv.factor)
ex4.11$df.2 <- data.frame(Year=0:3, Present.Value=ex4.11$pv)
knitr::kable(ex4.11$df.1)
knitr::kable(ex4.11$df.2)
knitr::kable(pretty(ex4.11$npv), col.names="Total")
```

---

## Example 4.12: EARs

What is the end-of-year wealth if Jane Christine receives a stated annual
interest rate of 24 percent compounded monthly on a \$1 investment?

Using Equation 4.6, her wealth is:

$$
\begin{aligned}
\$1 \left(1 + \frac{.24}{12} \right)^{12} &= \$1 \times (1.02)^{12}\\
  &= \$1.2682\\
\end{aligned}
$$

The annual rate of return is 26.82 percent. This annual rate of return is called
either the __effective annual rate (EAR)__ or the
__effective annual yield (EAY)__. Due to compounding, the effective annual
interest rate is greater than the stated annual interest rate of 24 percent.
Algebraically, we can rewrite the effective annual interest rate as follows:

\begin{center}
\textbf{Effective Annual Rate}
\end{center}

$$
\left( 1 + \frac{r}{m} \right)^M - 1\tag{4.7}
$$

Students are often bothered by the subtraction of 1 in equation 4.7. Note that
end-of-year wealth is composed of both the interest earned over the year and the
original principal. We remove the original principal by subtracting 1 in
equation 4.7.

---

## Example 4.13: Compounding frequencies

If the stated annual rate of interest, 8 percent, is compounded quarterly, what
is the effective annual rate?

Using equation 4.7, we have:

$$
\left( 1 + \frac{r}{m} \right)^m - 1 = \left( 1 + \frac{.08}{4} \right)^4 - 1
    = .0824 = 8.24\%
$$

Referring back to our original example where $C_0 = \$1,000$ and $r = 10\%$, we
can generate the following table:

```{r example 4.13, echo=F}
e4.13 <- new.env()
e4.13$C.0 <- rep(1000, 4)
e4.13$r <- .1
e4.13$m <- c(1, 2, 4, 365)
e4.13$ear <- round((1 + (e4.13$r/e4.13$m))^e4.13$m - 1, 4)
e4.13$C.1 <- e4.13$C.0 * (1 + e4.13$ear)
e4.13$df <- data.frame(C0=e4.13$C.0, m=e4.13$m, C1=e4.13$C.1, EAR=e4.13$ear)
knitr::kable(pretty(e4.13$df))
```

---

## Example 4.14: Multiyear compounding

Harry DeAngelo is investing \$5,000 at a stated annual interest rate of 12
percent per year, compounded quarterly, for five years. What is his wealth at
the end of five years?

Using equation 4.8, his wealth is:

$$
\$5,000 \times \left(1 + \frac{.12}{4} \right)^{4 \times 5}
  = \$5,000 \times (1.03)^{20} = \$5,000 \times 1.8061 = \$9,030.50
$$

---

## Example 4.15: Continuous compounding

Linda DeFond invested \$1,000 at a continuously compounded rate of 10 percent
for one year. What is the value of her wealth at the end of one year?

From equation 4.9 we have:

$$
\$1,000 \times e^{.10} = \$1,000 \times 1.1052 = \$1,105.20
$$

This number can easily be read from Table A.5. We merely set $r$, the value on
the horizontal dimension, to 10 percent and $T$, the value on the vertical
dimension, to 1.

Note that a continuously compounded rate of 10 percent is equivalent to an
annually compounded rate of 10.52 percent. In other words, Linda DeFond would
not care whether her bank quoted a continuously compounded rate of 10 percent or
a 10.52 percent rate, compounded annually.

---

## Example 4.16: Continuous compounding, continued

Linda DeFond's brother, Mark, invested \$1,000 at a continuously compounded rate
of 10 percent for two years.

The appropriate formula here is:

$$
\$1,000 \times e^{.10 \times 2} = \$1,000 \times e^{.20} = \$1,221.40
$$

Using the portion of the table of continuously compounded rates shown in the
previous example, we find the value to be 1.2214.

---

## Example 4.17: Present value with continuously compounding

The Michigan State Lottery is going to pay you \$100,000 at the end of four
years. If the annual continuously compounded rate of interest is 8 percent, what
is the present value of this payment?

$$
\$100,000 \times \frac{1}{e^{.08 \times 4}}
  = \$100,000 \times \frac{1}{1.3771} = \$72,616.37
$$

---

## Example 4.18: Perpetuities

Consider a perpetuity paying \$100 a year. If the relevant interest rate is 8
percent, what is the value of the consol?

Using equation 4.10 we have:

$$
\text{PV} = \frac{\$100}{.08} = \$1,250
$$

Now suppose that interest rates fall to 6 percent. Using Equation 4.10 the value
of the perpetuity is:

$$
\text{PV} = \frac{\$100}{.06} = \$1,666.67
$$

Note that the value of the perpetuity rises with a drop in the interest rate.
Conversely, the value of the perpetuity falls with a rise in the interest rate.

---

## Example 4.19: Paying dividends

Popovich Corporation is just about to pay a dividend of \$3.00 per share.
Investors anticipate that the annual dividend will rise by 6 percent a year
forever. The applicable discount rate is 11 percent. What is th eprice of the
stock today?

The numerator in equation 4.12 is the cash flow to be received next period.
Since the growth rate is 6 percent, the dividend next year is \$3.18
 ($=$3.00 \times 1.06$). The price of the stock today is:

$$
\$66.60 = \$3.00 + \frac{\$3.18}{.11 - .06}
$$

The price of \$66.60 includes both the dividend to be received immediately and
the present value of all dividends beginning a year from now. Equation 4.12
makes it possible to calculate only the present value of all dividends beginning
a year from now. Be sure you understand this eample; test questions on this
subject always seem to trip up a few of our students.

---

## Example 4.20: Lottery valuation

Mark Young has just won the state lottery, paying \$50,000 a year for 20 years.
He is to receive his first payment a year from now. THe state advertises this as
the Million Dollar Lottery because \$1,000,000 = \$50,000 $\times$ 20. If the
interest rate is 8 percent, what is the present value of the lottery?

$$
\begin{aligned}
\text{PV} &= \$50,000 \times
  \left[\frac{1 - \frac{1}{(1.08)^{20}}}{.08} \right]\\
  &= \text{periodic payment} \times \text{annuity factor} \\
  &= \$50,000 \times 9.8181\\
  &= \$490,905
\end{aligned}
$$

## Example 4.21: Retirement investing

Suppose you put \$3,000 per year into a Roth IRA. The account pays 6 percent
interest per year. How much will you have when you retire in 30 years?

This question asks for the future value of an annuity of \$3,000 per year for 30
years at 6 percent, which we can calculate as follows:

$$
\begin{aligned}
\text{FV} &= C \left[\frac{(1 + r)^T - 1}{r}\right]\\
  &= \$3,000 \times \left[\frac{1.06^{30} - 1}{.06}\right]\\
  &= \$3,000 \times 79.0582\\
  &= \$237,174.56
\end{aligned}
$$

So, you'll have close to a quarter million dollars in the account.

## Example 4.22: Delayed annuities

Danielle Caravello will receive a four-year annuity of \$500 per year, beginning
at date 6. If the interest rate is 10 percent, what is the present value of her
annuity?

1. Calculate the annuity:

$$
500 \times (1 - (1.10)^4)/.10 = 1,584.93
$$

2. Discount the present value of the annuity back 5 years (date 6 - date 1):

$$
\frac{1,584.93}{(1.10)^5} = 984.12
$$

---

## Example 4.23: Annuity due

In a previous example, Mark Young received \$50,000 a year for 20 years from the
state lottery. In that example, he was to receive the first payment a year from
the winning date. Let us now assume that the first payment occurs immediately.
The total number of payments remains 20.

Under this new assumption, we have a 19-date annuity with the first payment
occurring at Date 1---plus an extra payment at Date 0. The present value is:

$$
\begin{aligned}
\$50,000\ \ \ \ \ \ \ &+ &\$50,000 \times \text{PVIA}(.08, 19)\\
\text{Payment at date 0}\ &\ \ \ &\text{19-year annuity}\\
\ &= \$50,000 + &(\$50,000 \times 9.6036)\\
  &= \$530,180\ &\ 
\end{aligned}
$$

\$530,180, the present value in this example, is greater than \$490,905, the
present value in the earlier lottery example. This is to be expected because the
annuity of the current example begins earlier. An annuity with an immediate
initial payment is called an _annuity in advance_ or, more commonly, an _annuity
due_. Always remember that equation 4.15 and Table A.2 in this book refer to an
_ordinary annuity_.

## Example 4.24: Infrequent annuities

Ann Chen receives an annuity of \$450, payable once every two years. The annuity
stretches out over 20 years. The first payment occurs at Date 2---that is, two
years from today. The annual interest rate is 6 percent.

The trick is to determine the interest rate over a two-year period. The interest
rate over two years is:

$$
(1.06 \times 1.06) - 1 = 12.36\%
$$

That is, \$100 invested over two years will yield \$112.36.

What we want is the present value of a \$450 annuity over 10 periods, with an
interest rate of 12.36 percent per period:

$$
\begin{aligned}
\text{PV} &= \$450 \left[\frac{1 - \frac{1}{(1 + .1236)^{10}}}{.1236}\right]\\
  &= \$450 \times \text{PVIA}(.1236, 10)\\
  &= \$2,505.57
\end{aligned}
$$

---

## Example 4.25 Working with annuities

Harold and Helen Nash are saving for the college education of their newborn
daughter, Susan. The Nashes estimate that college expenses will run \$30,000 per
year when their daughter reaches college in 18 years. The annual interest rate
over the next few decades will be 14 percent. How much money must they deposit
in the bank each year so that their daughter will be completely supported
through four years of college?

To simplify the calculations, we assume that Susan is born today. Her parents
will make the first of her four annual tuition payments on her 18th birthday.
They will make equal bank deposits on each of her first 17 birthdays, but no
deposit at Date 0.

Mr. and Ms. Nash will be making deposits to the bank over the next 17 years.
They will be withdrawing \$30,000 per year over the following four years. We can
be sure they will be able to withdraw fully \$30,000 per year if the present
value of the deposits is equal to the present value of the four \$30,000
withdrawals.

This calculation requires three steps. The first two determine the present value
fo the withdrawals. The final step determines yearly deposits that will have a
present value equal to that of the withdrawals.

1. We calculate the present value of the four years at college using the annuity
   formula.

$$
\begin{aligned}
\text{PV} &= \$30,000 \times \left[\frac{1 - (1.14)^{-4}}{.14}\right]\\
  &= \$30,000 \times \text{PVIA}(.14, 4)\\
  &= \$30,000 \times 2.9137\\
  &= \$87,411
\end{aligned}
$$

We assume that Susan enters college on her 18th birthday. Given our discussion
in Trick 1, \$87,411 represents the present value at Date 17.

2. We calculate the present value of the college education at Date 0 as:

$$
\begin{aligned}
\text{PV} &= \frac{\$87,411}{(1.14)^{17}}\\
  &= \$9,422.91
\end{aligned}
$$

3. Assuming that Harold and Helen Nash make deposits to the bank at the end of
   each of the 17 years, we calculate the annual deposit that will yield a
   present value of all deposits of \$9,422.91. This is calculated as:

$$
C \times \text{PVIA}(.14, 17) = \$9,422.91
$$

Because $\text{PVIA}(.14, 17) = 6.3729$,

$$
C = \frac{\$9,422.91}{6.3729} = \$1,478.59
$$

Thus deposits of \$1,478.59 are made at the end of each of the first 17 years
and invested at 14 percent will provide enough money to make tuition payments of
\$30,000 over the following four years.

## Example 4.26: Growing annuities

Stuart Gabriel, a second-year MBA student, has just been offered a job at
\$80,000 a year. He anticipates his salary increasing by 9 percent a year until
his retirement in 40 years. Given an interest rate of 20  percent, what is the
present value of his lifetime salary?

We simplify by assuming he will be paid his \$80,000 salary exactly one year
from now, and that his salary will continue to be paid in annual installments.
The appropriate discount rate is 20 percent.

$$
\begin{aligned}
\text{PV} &= \$80,000 \times \left[
    \frac{1 - \left(\frac{1.09}{1.20}\right)^{40}}{.20 - .09}
  \right]\\
  &= \$711,730.71
\end{aligned}
$$

Though the growing annuity formula is quite useful, it is more tedious than the
other simplifying formulas. Whereas most sophisticated calculators have special
programs for perpetuity, growing perpetuity, and annuity, there is no special
program for a growing annuity. Hence, we must calculate all the terms in
equation 4.17 directly.

---

## Example 4.27: More growing annuities

In a previous example, Helen and Harold Nash planned to make 17 identical
payments to fund the college education of their daughter, Susan. Alternatively,
imagine that they planned to increase their payments at 4 percent per year.What
would their first payment be?

The first two steps of the previous Nash family example showed that the present
value of the college costs was \$9,422.91. These two steps would be the same
here. However, the third step must be altered. Now we must ask, How much should
their first payement be so that, if payments increase by 4 percent per year, the
present value of all payments will be \$9,422.91?

We set the growing annuity formula equal to \$9,422.91 and solve for $C$.

---

# Summary

1. Two basic concepts, _future value_ and _present value_, were introduced in
   the beginning of this chapter. With a 10 percent interest rate, an investor
   with \$1 today can generate a future value of \$1.10 in a year, \$1.21 [$=\$1
   \times (1.10)^2$] in two years, and so on. Conversely, present value analysis
   places a current value on a future cash flow. With the same 10 percent
   interest rate, a dollar to be received in one year has a present value of
   \$.909 ($=\$1/1.10$) in Year 0. A dollar to be received in two years has a
   present value of \$.826 [$=\$1/(1.10)^2$].

2. We commonly express an interest rate as, say, 12 percent per year. However,
   we can speak of the interest rate as 3 percent per quarter. Although the
   stated annual interest rate remains 12 percent ($=3 \text{ percent } \times
   4$), the effective annual interest rate is 12.55 percent [$=(1.03)^4 - 1$].
   In other words, the compounding process increases the future value of an
   investment. The limiting case is continuous compounding, where funds are
   assumed to be reinvested every infinitesimal instant.

3. A basic quantitative technique for financial decision making is net present
   value analysis. The net present value formula for an investment that
   generates cash flows ($C_i$) in future periods is:

$$
\mathbf{NPV} = - C_0 + \frac{C_1}{(1 + r)} + \frac{C_2}{(1 + r)^2} + \cdots
  + \frac{C_T}{(1 + r)^T} = -C_0 + \sum^T_{i=1} \frac{C_i}{(1 + r)^i}
$$

The formula assumes that the cash flow at date 0 is the initial investment (a
cash outflow). $r$ is the appropriate interest rate reflecting time and risk.

4. Frequently, the actual calculation of present value is long and tedious. The
   computation of the present value of a long-term mortgage with monthly
   payments is a good example of this. We presented four simplifying formulas:

$$
\begin{aligned}
\mathbf{Perpetuity}: &\ &\ &\ & \text{PV} &= \frac{C}{r} \\ \\
\mathbf{Growing\ perpetuity}: &\ &\ &\ & \text{PV} &= \frac{C}{r - g} \\ \\
\mathbf{Annuity}: &\ &\ &\ & \text{PV} &= C \left[ \frac{1
  - \frac{1}{(1 + r)^T}}{r} \right] \\ \\
\mathbf{Growing\ annuity}: &\ &\ &\ & \text{PV} &= C \left[
    \frac{1 - \left( \frac{1 + g}{1 + r} \right)^T}{r - g}
  \right]
\end{aligned}
$$

5. We stressed a few practical considerations in the application of these
   formulas:

> a. The numerator in each of the formulas, $C$, is the cash flow to be received
> _one full period hence_.

> b. Cash flows are generally irregular in practice. To avoid unwieldy problems,
> assumptions to create more regular cash flows are made both in this textbook
> and in the real world.

> c. A number of present value problems involve annuities (or perpetuities)
> beginning a few periods hence. Students should practice combining the annuity
> (or perpetuity) formula with the discounting formula to solve these problems.

> d. Annuities and perpetuities may have periods of every two or every $n$
> years, rather than once a year. The annuity and perpetuity formulas can easily
> handle such circumstances.

> e. We frequently encounter problems where the present value of one annuity
> must be equated with the present value of another annuity.

