---
title: "Discounted Cash Flow Valuation"
subtitle: "Concepts"
author: "Ross, Westerfield, Jaffe"
output: 
  pdf_document:
   toc: true 
---

```{r setup, echo=F, message=F}
require(data.table)
require(kableExtra)
pretty <- function(x, n = 2) format(round(x, n), big.mark=",", nsmall = n)
percent <- function(x, a = 0.01) scales::percent(x, accuracy = a)
table <- function(df, caption=NULL) {
  knitr::kable(df, caption=caption, format = "latex", booktabs=T) %>%
    kable_styling(latex_options = c("HOLD_position"))
options(scipen = 999)
}
```

# 1. Compounding and period

As you increase the length of time involved, what happens to future values? What
happens to present values?

---

Using the formulas for future value and present value of money:

$$
\begin{aligned}
\mathbf{FV} &= \text{PV} \times (1 + r)^t\\\\
\mathbf{PV} &=  \frac{\text{FV}}{(1 + r)^t}
\end{aligned}
$$

Where $r$ is the rate and $t$ is the time period.

As $t$ increases:

* __Future value (FV)__ increases because there's more time for compounding to
  grow the initial amount.

* __Present value (PV)__ decreases because the future cash flow is discounted
  more heavily.

This holds true as long as the interest rate $r$ is positive. If $r$ is negative
then the trends reverse.

---

# 2. Interest rates

What happens to the future value of an annuity if you increase the rate $r$?
What happens to the present value?

The future value and present value of an annuity as defined as:

$$
\begin{aligned}
\mathbf{FV\ Annuity} &= C \times \frac{(1 + r)^t - 1}{r}\\\\
\mathbf{PV\ Annuity} &= C \times \frac{1 - (1 + r)^{-t}}{r}
\end{aligned}
$$

As $r$ (the interest rate) increases, $(1 + r)$ increases. The future value of
an annuity will increase with the interest rate and the present value decreases.

---

# 3. Present value

Suppose two athletes sign 10-year contracts for \$80 million. In one case, we're
told that the \$80 million will be paid in 10 equal installments. In the other
case, we're told that the \$80 million will be paid in 10 installments, but the
installments will increase by 5 percent per year. Who got the better deal?

---

To evaluate which athlete got the better deal we need to compare the two
annuities. The differences between the annuities are in the structure of the
payments.

## Annuity 1: Equal installments.

$$
\begin{aligned}
\text{PV} &= C \times \frac{1 - (1 + r)^{-t}}{r}\\
  &= \$8,000,000 \times \frac{1 - (1 + r)^{-10}}{r}
\end{aligned}
$$

## Annuity 2: 5% Growing installments

The payments will grow 5% annually to 80 million. We need to find the starting
payment:

* Total nominal sum

$$
C \times \frac{1.05^{10} - 1}{.05} = 80\ \text{million}
$$

Solving for C:

$$
C \times 12.578 = 80 \rightarrow C \approx 6.36\ \text{million}
$$

Here is a plot of differences between the two annuities for discount rates from
.06 to 1:

```{r concept 3, echo=F}
growth.annuity <- function(c, r, g, t){
    c * ((1 + g)^(0:(t-1))) / (1 + r)^(1:t)
}
cf.1 <- 8
cf.2 <- 6.36

f <- function(r){
  ret <- c()
  for(i in r){
    lhs <- sum(growth.annuity(cf.1, i, .05, 10))
    rhs <- sum(growth.annuity(cf.2, i, 0, 10))
    ret <- c(ret, lhs - rhs)
  }
  return(ret)
}
x <- seq(.06, 1, by=.01)
y <- f(x)
plot(x, y, type='l', ylab='difference in millions of dollars',
    xlab='discount rate', main='Difference between Annuity 1 and Annuity 2')
```

As the plot shows, as the discount rate increase, the difference between the two
annuities diminishes. However, the plot also shows that the discount rate needs
to be extremely high for the difference between the two to become neglible.
Therefore, the growth annuity is the better deal.

---

# 4. APR and EAR

Should lending laws be changed to require lenders to report EARs instead of
APRs? Why or why not?

---

There are a few benefits in requiring lenders to report EARs instead of APRs.
APRs are effectively meaningless without the compounding interval. Thus a 10
percent APR is only meaningful if you know it will be compouned semi-annually,
or quarterly, etc... EARs are meaningful without the compounding interval
because it is the effective annual interest rate. This becomes especially useful
to know when the interest rate is large because the difference between the APR
and the EAR can be dramatic.

---

# 5. Time value

On subsidized Stafford loans, a common source of financial aid for college
students, interest does not begin to accrue until repayment begins. Who receives
a bigger subsidy, a freshman or a senior? Explain.

---

A freshman receives a bigger subsidy compared to a senior because of the longer
period before repayment begins. This period is interest-free while the student
is attending school. A freshman has four years before repayment begins, while a
senior only has one year of school left. The interest that is saved during this
period, compounded over the time attending school, makes the subsidy more
valuable to the freshman than the senior. Thus, if both the freshman and the
senior borrow the same amount each year, the freshman accrues a greater
cumulative benefit across the four years, while the senior only benefits for the
final year.

---

Use the following information to answer the next five question:

Toyota Motor Credit Corporation (TMCC), a subsidiary of Toyota Motor
Corporation, offered some securities for sale to the public on March 28, 2008.
Under the terms of the deal, TMCC promised to repay the owner of one of these
securities \$100,000 on March 28, 2038, but investors would receive nothing
until then. Investors paid TMCC \$24,099 for each of these securities; so they
gave up \$24,099 on March 28, 2008, for the promise of a \$100,000 payment 30
years later.

---

# 6. Time value of money

Why would TMCC be willing to accept such a small amount today (\$24,099) in
exchange for a promise to repay about four times that amount (\$100,000) in the
future?

---

TMCC is willing to accept a smaller amount today (\$24,099) in exchange for a
promise to repay \$100,000 in 30 years because of the time value of money and
the concept of discounting future cash flows.

For a future value of \$100,000 and a present value of \$24,099 and a term of 30
years, we can calculate the interest rate TMCC is expecting to pay.

$$
\begin{aligned}
\text{FV} &= \text{PV} \times (1 + r)^{t}\\\\
\frac{\text{FV}}{\text{PV}} &= (1 + r)^t\\\\
\left( \frac{\text{FV}}{\text{PV}} \right)^{1/t} &= 1 + r\\\\
r &= \left( \frac{\text{FV}}{\text{PV}} \right)^{1/t} - 1\\\\
r &= 0.04857629 \approx 5\%
\end{aligned}
$$

This discount rate of 5% reflects the rate investors see as the effects of
inflation, opportunity cost, and the perceived risk of TMCC's ability to fulfill
the obligation. The investors will find this deal attractive because they are
effectively locking in a specific return over 30 years. The discount rate
(yield) aligns with what they consider fair compensation for lending their money
to TMCC for three decades. TMCC uses this willingness to raise capital cheaply
in the present.

Japans own interest rate in March 2008 was 0.5%. This indicates an extremely low
cost of borrowing for domestic markets. However, there is reason to believe that
TMCC wanted to gain access to a global base of investors as a funding source.
With Japanese securities yielding only 0.5%, TMCC's securities yielding 5% look
extremely attractive.

---

# 7. Call provisions

TMCC has the right to buy back the securities on the anniversary date at a price
established when the securities were issued (this feature is a term of this
particular deal). What impact does this feature have on the desirability of this
security as an investment?

---

This feature has a negative impact on the desirability of the security. It
limits the upside potential of the security because if rates fall or if the
value of the security increases (due to falling discount rates or inflation
expectations), TMCC is likely to exercise its option to buy back the securities.
Also, because the security is callable, the benefit of the investor to lock in
an investment for 30 years is not guaranteed.

---

# 8. Time value of money

Would you be willing to pay \$24,099 today in exchange for \$100,000 in 30
years? What would be the key considerations in answering yes or no? Would your
answer depend on who is making the promise to repay?

---

The implied rate of return for this security was calculated as about 5%. You
would compare this with other investment options with similar risk profiles. You
would also consider inflation which can erode the real value of the investment.
Finally, you would need to evaluate the risk and creditworthiness of TMCC. This
means it does matter who is making the promise to repay.

---

# 9. Investment comparison

Suppose that when TMCC offered the security for \$24,099 the U.S. Treasury had
offered an essentially identical security. Do you think it would have had a
higher or lower price? Why?

---

It would have a higher price because the U.S. Treasury is less risky than the
TMCC security. The security is also more liquid because there are more buyers
for the security than for the TMCC security.

---

# 10. Length of investment

The TMCC security is bought and sold on the New York Stock Exchange. If you
looked at the price today, do you think the price would exceed the \$24,099
original price? Why? If you looked in the year 2019, do you think the price
would be higher or lower than today's price? Why?

---

Yes the price of the TMCC security is likely to have increased.

