---
title: "Discounted Cash Flow Valuation"
author: "Ross, Westerfield, Jaffe"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  pdf_document:
   toc: true 
---

```{r setup, echo=F, message=F}
require(data.table)
require(kableExtra)
pretty <- function(x, n = 2) format(round(x, n), big.mark=",", nsmall = n)
percent <- function(x, a = 0.01) scales::percent(x, accuracy = a)
table <- function(df, caption=NULL) {
  knitr::kable(df, caption=caption, format = "latex", booktabs=T) %>%
    kable_styling(latex_options = c("HOLD_position"))
options(scipen = 999)
}
```

# Chapter introduction

The signing of big-name athletes is often accompanied by great fanfare, but the
numbers are often misleading. For example, in late 2010, catcher Victor Martinez
reached a deal with the Detroit Tigers, signing a contract with a reported value
of \$50 million. Not bad, especially for someone who makes a living using the
"tools of ignorance" (jock jargon for a catcher's equipment). Another example is
the contract signed by Jayson Werth of the Washington Nationals, which had a
stated value of \$126 million.

It looks like Victor and Jayson did pretty well, but then there was Carl
Crawford, who signed to play in fron of Boston's Red Sox nation. Carl's contract
has a stated value of \$142 million, but this amount was actually payable over
several years. The contract consisted of a \$6 million signing bonus, along with
\$14 million in the first year plus \$122 million in future salary to be paid in
the years 2011 through 2017. Victor's and Jayson's payments were similarly
spread over time. Because all three contracts called for payments that are made
at future dates, we must consider the time value of money, which means none of
these players received the quoted amounts. How much did they really get? This
chapter gives you the "tools of knowledge" to answer this question.

# List of topics

1. Valuation: The one-period case

2. The multiperiod case

   * Future value and compounding
   * The power of compounding: A digression
   * Present value and discounting
   * Finding the number of periods
   * The algebraic formula

3. Compounding periods

   * Distinction between stated annual interest
   * Rate and effective annual rate
   * Compounding over many years
   * Continuous compounding

4. Simplifications

   * Perpetuity
   * Growing perpetuity
   * Annuity
   * Growing annuity

5. Loan amortization

6. What is a firm worth?

# Notes

## Valuation: The one-period case

* __Future value__ or __compound value__ is the value of a sum after investing
over one or more periods.

* __Present value__, the current value of a sum that will be received in the
future.

\begin{center}
\textbf{Present value of investment}
\end{center}

$$
\text{PV} = \frac{C_1}{1 + r}\tag{4.1}
$$

This is called _present value analysis_ and it will show the present value of a
cash flow ($C_1$) one period in the future. The variable $r$ is the
_discount rate_.

The net present value (NPV) is the present value of future cash flows minus the
present value of the cost of the investment.

\begin{center}
\textbf{Net present value of investment}
\end{center}

$$
\text{NPV} = - \text{Cost} + \text{PV}\tag{4.2}
$$

When determining the risk rate ($r$), you would choose a return that is
available in the market on other investments of the same risk.

## The multiperiod case

### Future value and compounding

After the first period in investing, you have a choice: (1) you can choose to
take the profits of the investment, or (2) you can reinvest both the principle
and interest and accumulate compounding.

The compounding interest is computed for the number of periods

$$
(1 + r)^t
$$

where $t$ is the number of compounding periods.

Simple interest does not take account the interest and thus is computed:

$$
r^t
$$

When cash is invested at compound interest, each interest payment is reinvested.
With simple interest, the interest is not reinvested.

The general formula for an investment over many periods can be written as:

\begin{center}
\textbf{Future value of an investment}
\end{center}

$$
\text{FV} = C_0 \times (1 + r)^T\tag{4.3}
$$

where $C_0$ is the cash to be invested at Date 0 (i.e., today), $r$ is the
itnerest rate per period, and $T$ is the number of periods over which the cash
is invested.

### The power of compounding: A digression

The real affect of compounding can be seen at large intervals. For example if
you invested one dollar in the U.S. stock market in 1929, it would be worth
\$2,982.24.24 in 2010. This is 9.87 percent compounded annually for 85
years---that is, $(1.0987)^{85} = \$2,982.24$, ignoring a small rounding error.

### Present value and discounting

The multiperiod equation for the present value of future cash flows is

$$
\text{PV} = \frac{C_T}{(1 + r)^T}\tag{4.4}
$$

The process of calculating the present value of a future cash flow is called
__discounting__.

### Finding the number of periods

When solving for the number of periods you can use regression:

$$
\frac{\text{PV}}{\text{FV}} = (1 + r)^T
$$

### The algebraic formula

$$
\text{NPV} = - C_0 + \frac{C_1}{1 + r} + \frac{C_2}{(1 + r)^2}
  + \cdots + \frac{C_T}{(1 + r)^T}
  = - C_0 + \sum^{T}_{i=1}{\frac{C_i}{(1 + r)^i}}\tag{4.5}
$$

## Compounding Periods

Sometime compounding occurs more frequently than just one year.

Other types of compounding:

* Semi-annual compounding

* Quarterly compounding

* Monthly compounding

The general formula for the present value factor:

$$
C_0 \left( 1 + \frac{r}{m} \right)^m
$$

In this formula $r$ is the __stated annual interest rate (SAIR)__ and $m$ is the
number of compounding periods in the year.

The stated annual interest rate is the annual interest rate without
consideration of compounding. Banks and other financial institutions may use
other names for the stated annual interest rate. __Annual percentage rate
(APR)__ is perhaps the most common synonym.

### Distinction betwen stated annual interest rate and effective annual rate

The stated annual interest rate (SAIR) is only meaningful when the compounding
interval is given. However the effective annual rate (EAR) is meaningful without
the compounding interval.

When interest rates are large the difference between the APR and the EAR can be
big.

### Compounding over many years

When applying compounding over many years the general form of the present value
factor has the form:

$$
\left( 1 + \frac{r}{m} \right)^{mT}\tag{4.8}
$$

### Continuous compounding

$$
C_0 \times e^{rT}\tag{4.9}
$$

## Simplifications

The following sections show simplifications to the basic formulas.

### Perpetuity

For a perpetuity the equation is:

$$
\text{PV} = \frac{C}{1 + r} + \frac{C}{(1 + r)^2}
  + \frac{C}{(1 + r)^3} + \cdots
$$

We will state, without proof that the formula reduces to:

$$
\text{PV} = \frac{C}{r}\tag{4.10}
$$

### Growing perpetuities

For a growing perpetuity the equation is

$$
\text{PV} = \frac{C}{1 + r} + \frac{C \times (1 + g)}{(1 + r)^2}
  + \frac{C \times (1 + g)^2}{(1 + r)^3} + \cdots
  + \frac{C \times (1 + g)^{N - 1}}{(1 + r)^N} + \cdots
$$

For the case when $r > g$, the formula reduces to:

$$
\text{PV} = \frac{C}{r - g}\tag{4.12}
$$

A few important points:

1. _The numerator_: The numerator is the cash flow one period hence, not at
   date 0.

2. _The discount rate and the growth rate_: The discount rate $r$ must be
   greater than the growth rate $g$ for the growing perpetuity formula to work.
   Consider the case in which the growth rate approaches the interest rate in
   magnitude. Then, the denominator in the growing perpetuity formula gets
   infinitesimally small and the present value grows infinitely large. The
   present value is in fact undefined when $r$ is less than $g$.

3. _The timing assumption_: Cash generally flows into and out of real-world
   firms both randomly and nearly continuously. However, the equation assumes
   that cash flows are received and disbursed at regular and discrete points in
   time.

### Annuity

You can derive the formula for an annuity from the formula of two perpetuities.
Mainly:

> For $C_1$ which starts at $t=0$ and $C_2$ which starts at $t=T$. The formula
> for an annuity is:

$$
\text{PV}_{\text{annuity}} = \frac{C_1}{r} - \frac{C_2}{r}
  \left[ \frac{1}{(1 + r)^T} \right]\tag{4.14}
$$

This simplifies to:

\begin{center}
\textbf{Formula for present value of annuity}
\end{center}

$$
\text{PV}_{\text{annuity}} = C
  \frac{\left( 1 - (1 + r)^{-T} \right)}{r}\tag{4.15}
$$

The term used to compute the present value of the stream of level payments, $C$,
for $T$ years is called a __present value interest factor for annuitites__.

### Trick 1: A delayed annuity

One of the tricks in working with annuities and perpetuities is getting the
timing right.

#### Trick 2: Annuity due

The annuity formula of equation 4.15 assumes that the first annuity  payment
begins a full period hence. This type of annuity is sometimes called an
_annuity arrears_ or _ordinary annuity_. What happens if the annuity begins
today---in other words, at Date 0?

#### Trick 3: The infrequent annuity

Here the trick is to determine the interest rate over the period.

#### Trick 4: Equating present value of two annuities

### Growing annuity

\begin{center}
\textbf{Formula for present value of growing annuity}
\end{center}

$$
\text{PV} = C \left[\frac{1}{r - g} - \frac{1}{r - g}
  \times \left(\frac{1 + g}{1 + r}\right)^T\right]
  = C \left[\frac{1 - \left(\frac{1 + g}{1 + r}\right)^T}{r - g}\right]
  \tag{4.17}
$$

---

# Examples

---

## Example 4.1: Present value

Diane Badame, a financial analyst at Kaufman & Broad, a leading real estate
firm, is thinking about recommending that Kaufman & Broad invest in a piece of
land that costs \$85,000. She is certain that next year the land will be worth
\$91,000, a sure \$6,000 gain. Given that the interest rate in similar
alternative investments is 10 percent, should Kaufman & Broad undertake the
investment in land? Diane's choice is described below with the cash flow time
chart.

A moment's thought should be all it takes to convince her that this is not an
attractive business deal. By investing \$85,000 in the land, she will have
\$91,000 available next year. Suppose, instead, that Kaufman & Broad puts the
same \$85,000 into similar alternative investments. At the interest rate of 10
percent, this \$85,000 would grow to:

$$
(1 + .10) \times \$85,000 = \$93,500
$$

next year.

It would be foolish to buy the land when investing the same \$85,000 in similar
alternative investments would produce an extra \$2,500 (that is, \$93,500 from
the bank minus \$91,000 from the land investment). This is a future value
calculation.

Alternatively, she could calculate the present value of the sale price next year
as:

$$
\text{Present value} = \frac{\$91,000}{1.10} = \$82,727.27
$$

Because the present value of next year's sales price is less than this year's
purchase price of \$85,000, present value analysis also indicates that she
should not recommend purchasing th property.

---

## Example 4.2: Uncertainty and valuation

Professional Artworks, Inc. is a firm that speculates in modern paintings. The
manager is thinking of buying an original Picasso for \$400,000 with the
intention of selling it at the end of one year. The manager expects that the
painting will be worth \$480,000 in one year. The relavent cash flows are
depicted below.

<<FIGURE 4.3>>

Of course, this is only an expectation---the painting could be worth more or
less than \$480,000. Suppose the guaranteed interest rate granted by banks is 10
percent. Should the firm purchase the piece of art?

Our first thought might be to discount at the interest rate, yielding:

$$
\frac{\$480,000}{1.10} = \$436,364
$$

Because \$436,364 is greater than \$400,000, it looks at first glance as if the
painting should be purchased. However, 10 percent is the return one can earn on
a low risk investment. Because the painting is quite risky, a higher discount
rate is called for: The manager chooses a rate of 25 percent to reflect this
risk. In other words, he argues that a 25 percent expected return is fair
compensation for an investment as risky as this painting.

The present value of the painting becomes:

$$
\frac{\$480,000}{1.25} = \$384,000
$$

Thus, the manager believes that the painting is currently overpriced at
\$400,000 and does not make the purchase.

---

## Example 4.3: Interest on interest

Suh-Pyng Ku has put \$500 in a savings account at the First National Bank of
Kent. The account earns 7 percent, compounded annually. How much will ms. Ku
have at the end of three years? The answer is:

$$
\$500 \times 1.07 \times 1.07 \times 1.07 = \$500 \times (1.07)^3 = \$612.52
$$

---

## Example 4.4: Compound growth

Jay Ritter invested \$1,000 in the stock of the SDH Company. The company pays a
current dividend of \$2, which is expected to grow by 20 percent per year for
hte next two years. What will the dividend of the SDH Company be after two
years? A simple calculation gives:

$$
\$2 \times (1.20)^2 = \$2.88
$$

---

## Example 4.5: Finding the rate

Carl Voigt, who recently won \$10,000 in the lottery, wants to buy a car in five
years. Carl estimates that the car will cost \$16,105 at that time. His cash
flows are displayed in the figure below.

What interest rate must he earn to be able to afford the car?

The ratio of purchase price to initial cash is:

$$
\frac{\$16,105}{\$10,000} = 1.6105
$$

Thus, he must earn an interest rate that allows \$1 to become \$1.6105 in five
years. Table A.3 tells us that an interest rate of 10 percent will allow him to
purchase the car:

We can express the problem algebraically as:

$$
\$10,000 \times (1 + r)^5 = \$16,105
$$

where $r$ is the interest rate needed to purchase the car. Because
\$16,105/\$10,000 = 1.6105, we have:

$$
\begin{aligned}
(1 + r)^5 &= 1.6105\\
r &= 10\%
\end{aligned}
$$

Either the table, a spreadsheet, or a hand calculator lets us solve for $r$.

---

## Example 4.6: How much for that island?

Some people have said that it was the best real estate deal in history. Peter
Minuit, director general of New Netherlands, the Dutch West India Company's
colony in North America, in 1626 allegedly bought Manhattan Island for 60
guilders' worth of trinkets from native Americans. By 1667, the Dutch were
forced by the Britsh to exchnage it for Suriname
(perhaps the worst real estate deal ever). This sounds cheap; but did the Dutch
really get the better end of the deal? It is reported that 60 guilders was worth
about \$24 at the prevailing exchange rate. If the native Americans had sold the
trinkets at a fair market value and invested the \$24 at 5 percent (tax free),
it would now, about 385 years later, be worth more than \$3.45 billion. Today,
Manhattan is undoutedly worth more than \$3.45 billion, so at a 5 percent rate
of return the native Americans got th eworst of the deal. However, if invested
at 10 percent, the amount of money they received would be worth about:

$$
\$24(1 + r)^T = 24 \times 1.1^{385} \cong \$207\ \text{quadrillion}
$$

This is a lot of money. In fact, \$207 quadrillion is more than all the real
estate in the world is worth today. Note that no one in the history of the world
has ever been able to find an investment yielding 10 percent every year for 385
years.

---

## Example 4.7: Multiperiod discounting

Bernard Dumas will receive \$10,000 three years from now. Bernard can earn 8
percent on his investments, so the appropriate discount rate is 8 percent. What
is the present value of his future cash flow? The answer is:

$$
\begin{aligned}
\text{PV} &= \$10,000 \times \left(\frac{1}{1.08}\right)^3\\
  &= \$10,000 \times .7938\\
  &= \$7,938
\end{aligned}
$$

When his investments grow at an 8 percent rate of interest, Bernard Dumas is
equally inclined toward receivng \$7,938 now and receiving \$10,000 in three
years' time. After all, he could convert the \$7,938 he receives today into
\$10,000 in three years by lending it at an interest rate of 8 percent.

Bernard Dumas could have reached his present value calculation in one of several
ways. The computation could have been done by hand, by calculator, with a
spreadsheet, or with the help of Table A.1, which appears in the back of the
text. This table presents the _present value of_ \$1 to be received after $T$
periods. We use the table by locating the appropriate interest rate on the
horizontal and the appropriate number of periods on the vertical. For example,
Bernard Dumas would look at the following portion of Table A.1.

The appropriate present value factor is .7938.

---

### Example 4.8: Finding the rate

A customer of the Chaffkin Corp. wants to buy a tugboat today. Rather than
paying immediately, he will pay \$50,000 in three years. It will cost the
Chaffkin Corp. \$38,610 to build the tugboat immediately. The relevant cash
flows to Chaffkin Corp. are displayed in the figure below. What interest rate
would the Chaffkin Corp. charge to neither gain nor lose on the sale?

The ratio of construction cost (present value) to sales price (future value) is:

$$
\frac{\$38,000}{\$50,000} = .7722
$$

We must determine the interest rate that allows \$1 to be received in three
years to have a present value of \$.7722. Table A.1 tells us that 9 percent is
that interest rate.

---

## Example 4.9: Waiting for Godot

You've been saving up to buy the Godot Company. The total cost will be \$10
million. You currently have about \$2.3 million. If you can earn 5 percent on
your money, how long will you have to wait? At 16 percent, how long must you
wait?

At 5 percent, you'll have to wait a long time. From the basic present value
equation:

$$
\begin{aligned}
\$2.3 \text{million} &= \$10 \text{million} / 1.05^t\\
1.05^t &= 4.35\\
t &= 30 \text{years}
\end{aligned}
$$

At 16 percent, things are a little better. Verify for your self that it will
take about 10 years.

---

## Example 4.10: Cash flow valuation

Kyle Mayer has won the Kentucky State Lottery and will receive the following set
of cash flows over the next two years:

| Year | Cash flow |
|:---:|---:|
| 1 | \$20,000 |
| 2 | 50,000 |

Mr. Mayer can currently earn 6 percent in his money market account, so the
appropriate discount rate is 6 percent. The present value of the cash flows is:

$$
\begin{aligned}
\text{CF}_{\text{year 1}} &= \$20,000 \times \frac{1}{1.06}
  = \$18,867.9\\
\text{CF}_{\text{year 2}} &= \$50,000 \times \left( \frac{1}{1.06} \right)^2
  = \$44,499.8\\\\
\text{Total} &= \$18,867.9 + \$44,499.8 = \$63,367.7
\end{aligned}
$$

---

## Example 4.11: NPV

Finance.com has an opportunity to invest in a new high-speed computer that costs
\$50,000. The computer will generate cash flows (from cost savings) of \$25,000
one year from now, \$20,000 two years from now, and \$15,000 three years from
now. The computer will be worthless after three years, and no additional cash
flows will occur. Finance.com has determined that the appropriate discount rate
is 7 percent for this investment. Should Finance.com make this investment in a
new high-speed computer? What is the net present value of the investment?

The cash flows and present value factors of the proposed computer are as
follows:

```{r example 4.11, echo=F}
ex4.11 <- new.env()
ex4.11$cf <- c(-50000, 25000, 20000, 15000)
ex4.11$discount.rate <- .07
ex4.11$pv.factor <- round(
        (1 + ex4.11$discount.rate)^(-(0:(length(ex4.11$cf)-1))), 4)
ex4.11$pv <- ex4.11$cf * ex4.11$pv.factor
ex4.11$npv <- sum(ex4.11$pv)
ex4.11$df.1 <- data.frame(Cash.flows=ex4.11$cf,
        Present.Value.Factor=ex4.11$pv.factor)
ex4.11$df.2 <- data.frame(Year=0:3, Present.Value=ex4.11$pv)
knitr::kable(ex4.11$df.1)
knitr::kable(ex4.11$df.2)
knitr::kable(pretty(ex4.11$npv), col.names="Total")
```

---

## Example 4.12: EARs

What is the end-of-year wealth if Jane Christine receives a stated annual
interest rate of 24 percent compounded monthly on a \$1 investment?

Using Equation 4.6, her wealth is:

$$
\begin{aligned}
\$1 \left(1 + \frac{.24}{12} \right)^{12} &= \$1 \times (1.02)^{12}\\
  &= \$1.2682\\
\end{aligned}
$$

The annual rate of return is 26.82 percent. This annual rate of return is called
either the __effective annual rate (EAR)__ or the
__effective annual yield (EAY)__. Due to compounding, the effective annual
interest rate is greater than the stated annual interest rate of 24 percent.
Algebraically, we can rewrite the effective annual interest rate as follows:

\begin{center}
\textbf{Effective Annual Rate}
\end{center}

$$
\left( 1 + \frac{r}{m} \right)^M - 1\tag{4.7}
$$

Students are often bothered by the subtraction of 1 in equation 4.7. Note that
end-of-year wealth is composed of both the interest earned over the year and the
original principal. We remove the original principal by subtracting 1 in
equation 4.7.

---

## Example 4.13: Compounding frequencies

If the stated annual rate of interest, 8 percent, is compounded quarterly, what
is the effective annual rate?

Using equation 4.7, we have:

$$
\left( 1 + \frac{r}{m} \right)^m - 1 = \left( 1 + \frac{.08}{4} \right)^4 - 1
    = .0824 = 8.24\%
$$

Referring back to our original example where $C_0 = \$1,000$ and $r = 10\%$, we
can generate the following table:

```{r example 4.13, echo=F}
e4.13 <- new.env()
e4.13$C.0 <- rep(1000, 4)
e4.13$r <- .1
e4.13$m <- c(1, 2, 4, 365)
e4.13$ear <- round((1 + (e4.13$r/e4.13$m))^e4.13$m - 1, 4)
e4.13$C.1 <- e4.13$C.0 * (1 + e4.13$ear)
e4.13$df <- data.frame(C0=e4.13$C.0, m=e4.13$m, C1=e4.13$C.1, EAR=e4.13$ear)
knitr::kable(pretty(e4.13$df))
```

---

## Example 4.14: Multiyear compounding

Harry DeAngelo is investing \$5,000 at a stated annual interest rate of 12
percent per year, compounded quarterly, for five years. What is his wealth at
the end of five years?

Using equation 4.8, his wealth is:

$$
\$5,000 \times \left(1 + \frac{.12}{4} \right)^{4 \times 5}
  = \$5,000 \times (1.03)^{20} = \$5,000 \times 1.8061 = \$9,030.50
$$

---

## Example 4.15: Continuous compounding

Linda DeFond invested \$1,000 at a continuously compounded rate of 10 percent
for one year. What is the value of her wealth at the end of one year?

From equation 4.9 we have:

$$
\$1,000 \times e^{.10} = \$1,000 \times 1.1052 = \$1,105.20
$$

This number can easily be read from Table A.5. We merely set $r$, the value on
the horizontal dimension, to 10 percent and $T$, the value on the vertical
dimension, to 1.

Note that a continuously compounded rate of 10 percent is equivalent to an
annually compounded rate of 10.52 percent. In other words, Linda DeFond would
not care whether her bank quoted a continuously compounded rate of 10 percent or
a 10.52 percent rate, compounded annually.

---

## Example 4.16: Continuous compounding, continued

Linda DeFond's brother, Mark, invested \$1,000 at a continuously compounded rate
of 10 percent for two years.

The appropriate formula here is:

$$
\$1,000 \times e^{.10 \times 2} = \$1,000 \times e^{.20} = \$1,221.40
$$

Using the portion of the table of continuously compounded rates shown in the
previous example, we find the value to be 1.2214.

---

## Example 4.17: Present value with continuously compounding

The Michigan State Lottery is going to pay you \$100,000 at the end of four
years. If the annual continuously compounded rate of interest is 8 percent, what
is the present value of this payment?

$$
\$100,000 \times \frac{1}{e^{.08 \times 4}}
  = \$100,000 \times \frac{1}{1.3771} = \$72,616.37
$$

---

## Example 4.18: Perpetuities

Consider a perpetuity paying \$100 a year. If the relevant interest rate is 8
percent, what is the value of the consol?

Using equation 4.10 we have:

$$
\text{PV} = \frac{\$100}{.08} = \$1,250
$$

Now suppose that interest rates fall to 6 percent. Using Equation 4.10 the value
of the perpetuity is:

$$
\text{PV} = \frac{\$100}{.06} = \$1,666.67
$$

Note that the value of the perpetuity rises with a drop in the interest rate.
Conversely, the value of the perpetuity falls with a rise in the interest rate.

---

## Example 4.19: Paying dividends

Popovich Corporation is just about to pay a dividend of \$3.00 per share.
Investors anticipate that the annual dividend will rise by 6 percent a year
forever. The applicable discount rate is 11 percent. What is th eprice of the
stock today?

The numerator in equation 4.12 is the cash flow to be received next period.
Since the growth rate is 6 percent, the dividend next year is \$3.18
 ($=$3.00 \times 1.06$). The price of the stock today is:

$$
\$66.60 = \$3.00 + \frac{\$3.18}{.11 - .06}
$$

The price of \$66.60 includes both the dividend to be received immediately and
the present value of all dividends beginning a year from now. Equation 4.12
makes it possible to calculate only the present value of all dividends beginning
a year from now. Be sure you understand this eample; test questions on this
subject always seem to trip up a few of our students.

---

## Example 4.20: Lottery valuation

Mark Young has just won the state lottery, paying \$50,000 a year for 20 years.
He is to receive his first payment a year from now. THe state advertises this as
the Million Dollar Lottery because \$1,000,000 = \$50,000 $\times$ 20. If the
interest rate is 8 percent, what is the present value of the lottery?

$$
\begin{aligned}
\text{PV} &= \$50,000 \times
  \left[\frac{1 - \frac{1}{(1.08)^{20}}}{.08} \right]\\
  &= \text{periodic payment} \times \text{annuity factor} \\
  &= \$50,000 \times 9.8181\\
  &= \$490,905
\end{aligned}
$$

## Example 4.21: Retirement investing

Suppose you put \$3,000 per year into a Roth IRA. The account pays 6 percent
interest per year. How much will you have when you retire in 30 years?

This question asks for the future value of an annuity of \$3,000 per year for 30
years at 6 percent, which we can calculate as follows:

$$
\begin{aligned}
\text{FV} &= C \left[\frac{(1 + r)^T - 1}{r}\right]\\
  &= \$3,000 \times \left[\frac{1.06^{30} - 1}{.06}\right]\\
  &= \$3,000 \times 79.0582\\
  &= \$237,174.56
\end{aligned}
$$

So, you'll have close to a quarter million dollars in the account.

## Example 4.22: Delayed annuities

Danielle Caravello will receive a four-year annuity of \$500 per year, beginning
at date 6. If the interest rate is 10 percent, what is the present value of her
annuity?

1. Calculate the annuity:

$$
500 \times (1 - (1.10)^4)/.10 = 1,584.93
$$

2. Discount the present value of the annuity back 5 years (date 6 - date 1):

$$
\frac{1,584.93}{(1.10)^5} = 984.12
$$

---

## Example 4.23: Annuity due

In a previous example, Mark Young received \$50,000 a year for 20 years from the
state lottery. In that example, he was to receive the first payment a year from
the winning date. Let us now assume that the first payment occurs immediately.
The total number of payments remains 20.

Under this new assumption, we have a 19-date annuity with the first payment
occurring at Date 1---plus an extra payment at Date 0. The present value is:

$$
\begin{aligned}
\$50,000\ \ \ \ \ \ \ &+ &\$50,000 \times \text{PVIA}(.08, 19)\\
\text{Payment at date 0}\ &\ \ \ &\text{19-year annuity}\\
\ &= \$50,000 + &(\$50,000 \times 9.6036)\\
  &= \$530,180\ &\ 
\end{aligned}
$$

\$530,180, the present value in this example, is greater than \$490,905, the
present value in the earlier lottery example. This is to be expected because the
annuity of the current example begins earlier. An annuity with an immediate
initial payment is called an _annuity in advance_ or, more commonly, an _annuity
due_. Always remember that equation 4.15 and Table A.2 in this book refer to an
_ordinary annuity_.

## Example 4.24: Infrequent annuities

Ann Chen receives an annuity of \$450, payable once every two years. The annuity
stretches out over 20 years. The first payment occurs at Date 2---that is, two
years from today. The annual interest rate is 6 percent.

The trick is to determine the interest rate over a two-year period. The interest
rate over two years is:

$$
(1.06 \times 1.06) - 1 = 12.36\%
$$

That is, \$100 invested over two years will yield \$112.36.

What we want is the present value of a \$450 annuity over 10 periods, with an
interest rate of 12.36 percent per period:

$$
\begin{aligned}
\text{PV} &= \$450 \left[\frac{1 - \frac{1}{(1 + .1236)^{10}}}{.1236}\right]\\
  &= \$450 \times \text{PVIA}(.1236, 10)\\
  &= \$2,505.57
\end{aligned}
$$

---

## Example 4.25 Working with annuities

Harold and Helen Nash are saving for the college education of their newborn
daughter, Susan. The Nashes estimate that college expenses will run \$30,000 per
year when their daughter reaches college in 18 years. The annual interest rate
over the next few decades will be 14 percent. How much money must they deposit
in the bank each year so that their daughter will be completely supported
through four years of college?

To simplify the calculations, we assume that Susan is born today. Her parents
will make the first of her four annual tuition payments on her 18th birthday.
They will make equal bank deposits on each of her first 17 birthdays, but no
deposit at Date 0.

Mr. and Ms. Nash will be making deposits to the bank over the next 17 years.
They will be withdrawing \$30,000 per year over the following four years. We can
be sure they will be able to withdraw fully \$30,000 per year if the present
value of the deposits is equal to the present value of the four \$30,000
withdrawals.

This calculation requires three steps. The first two determine the present value
fo the withdrawals. The final step determines yearly deposits that will have a
present value equal to that of the withdrawals.

1. We calculate the present value of the four years at college using the annuity
   formula.

$$
\begin{aligned}
\text{PV} &= \$30,000 \times \left[\frac{1 - (1.14)^{-4}}{.14}\right]\\
  &= \$30,000 \times \text{PVIA}(.14, 4)\\
  &= \$30,000 \times 2.9137\\
  &= \$87,411
\end{aligned}
$$

We assume that Susan enters college on her 18th birthday. Given our discussion
in Trick 1, \$87,411 represents the present value at Date 17.

2. We calculate the present value of the college education at Date 0 as:

$$
\begin{aligned}
\text{PV} &= \frac{\$87,411}{(1.14)^{17}}\\
  &= \$9,422.91
\end{aligned}
$$

3. Assuming that Harold and Helen Nash make deposits to the bank at the end of
   each of the 17 years, we calculate the annual deposit that will yield a
   present value of all deposits of \$9,422.91. This is calculated as:

$$
C \times \text{PVIA}(.14, 17) = \$9,422.91
$$

Because $\text{PVIA}(.14, 17) = 6.3729$,

$$
C = \frac{\$9,422.91}{6.3729} = \$1,478.59
$$

Thus deposits of \$1,478.59 are made at the end of each of the first 17 years
and invested at 14 percent will provide enough money to make tuition payments of
\$30,000 over the following four years.

## Example 4.26: Growing annuities

Stuart Gabriel, a second-year MBA student, has just been offered a job at
\$80,000 a year. He anticipates his salary increasing by 9 percent a year until
his retirement in 40 years. Given an interest rate of 20  percent, what is the
present value of his lifetime salary?

We simplify by assuming he will be paid his \$80,000 salary exactly one year
from now, and that his salary will continue to be paid in annual installments.
The appropriate discount rate is 20 percent.

$$
\begin{aligned}
\text{PV} &= \$80,000 \times \left[
    \frac{1 - \left(\frac{1.09}{1.20}\right)^{40}}{.20 - .09}
  \right]\\
  &= \$711,730.71
\end{aligned}
$$

Though the growing annuity formula is quite useful, it is more tedious than the
other simplifying formulas. Whereas most sophisticated calculators have special
programs for perpetuity, growing perpetuity, and annuity, there is no special
program for a growing annuity. Hence, we must calculate all the terms in
equation 4.17 directly.

---

## Example 4.27: More growing annuities

In a previous example, Helen and Harold Nash planned to make 17 identical
payments to fund the college education of their daughter, Susan. Alternatively,
imagine that they planned to increase their payments at 4 percent per year.What
would their first payment be?

The first two steps of the previous Nash family example showed that the present
value of the college costs was \$9,422.91. These two steps would be the same
here. However, the third step must be altered. Now we must ask, How much should
their first payement be so that, if payments increase by 4 percent per year, the
present value of all payments will be \$9,422.91?

We set the growing annuity formula equal to \$9,422.91 and solve for $C$.

---

# Summary

1. Two basic concepts, _future value_ and _present value_, were introduced in
   the beginning of this chapter. With a 10 percent interest rate, an investor
   with \$1 today can generate a future value of \$1.10 in a year, \$1.21 [$=\$1
   \times (1.10)^2$] in two years, and so on. Conversely, present value analysis
   places a current value on a future cash flow. With the same 10 percent
   interest rate, a dollar to be received in one year has a present value of
   \$.909 ($=\$1/1.10$) in Year 0. A dollar to be received in two years has a
   present value of \$.826 [$=\$1/(1.10)^2$].

2. We commonly express an interest rate as, say, 12 percent per year. However,
   we can speak of the interest rate as 3 percent per quarter. Although the
   stated annual interest rate remains 12 percent ($=3 \text{ percent } \times
   4$), the effective annual interest rate is 12.55 percent [$=(1.03)^4 - 1$].
   In other words, the compounding process increases the future value of an
   investment. The limiting case is continuous compounding, where funds are
   assumed to be reinvested every infinitesimal instant.

3. A basic quantitative technique for financial decision making is net present
   value analysis. The net present value formula for an investment that
   generates cash flows ($C_i$) in future periods is:

$$
\mathbf{NPV} = - C_0 + \frac{C_1}{(1 + r)} + \frac{C_2}{(1 + r)^2} + \cdots
  + \frac{C_T}{(1 + r)^T} = -C_0 + \sum^T_{i=1} \frac{C_i}{(1 + r)^i}
$$

The formula assumes that the cash flow at date 0 is the initial investment (a
cash outflow). $r$ is the appropriate interest rate reflecting time and risk.

4. Frequently, the actual calculation of present value is long and tedious. The
   computation of the present value of a long-term mortgage with monthly
   payments is a good example of this. We presented four simplifying formulas:

$$
\begin{aligned}
\mathbf{Perpetuity}: &\ &\ &\ & \text{PV} &= \frac{C}{r} \\ \\
\mathbf{Growing\ perpetuity}: &\ &\ &\ & \text{PV} &= \frac{C}{r - g} \\ \\
\mathbf{Annuity}: &\ &\ &\ & \text{PV} &= C \left[ \frac{1
  - \frac{1}{(1 + r)^T}}{r} \right] \\ \\
\mathbf{Growing\ annuity}: &\ &\ &\ & \text{PV} &= C \left[
    \frac{1 - \left( \frac{1 + g}{1 + r} \right)^T}{r - g}
  \right]
\end{aligned}
$$

5. We stressed a few practical considerations in the application of these
   formulas:

> a. The numerator in each of the formulas, $C$, is the cash flow to be received
> _one full period hence_.

> b. Cash flows are generally irregular in practice. To avoid unwieldy problems,
> assumptions to create more regular cash flows are made both in this textbook
> and in the real world.

> c. A number of present value problems involve annuities (or perpetuities)
> beginning a few periods hence. Students should practice combining the annuity
> (or perpetuity) formula with the discounting formula to solve these problems.

> d. Annuities and perpetuities may have periods of every two or every $n$
> years, rather than once a year. The annuity and perpetuity formulas can easily
> handle such circumstances.

> e. We frequently encounter problems where the present value of one annuity
> must be equated with the present value of another annuity.

---

# Concept Questions

## 1. Compounding and period

As you increase the length of time involved, what happens to future values? What
happens to present values?

---

Using the formulas for future value and present value of money:

$$
\begin{aligned}
\mathbf{FV} &= \text{PV} \times (1 + r)^t\\\\
\mathbf{PV} &=  \frac{\text{FV}}{(1 + r)^t}
\end{aligned}
$$

Where $r$ is the rate and $t$ is the time period.

As $t$ increases:

* __Future value (FV)__ increases because there's more time for compounding to
  grow the initial amount.

* __Present value (PV)__ decreases because the future cash flow is discounted
  more heavily.

This holds true as long as the interest rate $r$ is positive. If $r$ is negative
then the trends reverse.

---

## 2. Interest rates

What happens to the future value of an annuity if you increase the rate $r$?
What happens to the present value?

The future value and present value of an annuity as defined as:

$$
\begin{aligned}
\mathbf{FV\ Annuity} &= C \times \frac{(1 + r)^t - 1}{r}\\\\
\mathbf{PV\ Annuity} &= C \times \frac{1 - (1 + r)^{-t}}{r}
\end{aligned}
$$

As $r$ (the interest rate) increases, $(1 + r)$ increases. The future value of
an annuity will increase with the interest rate and the present value decreases.

---

## 3. Present value

Suppose two athletes sign 10-year contracts for \$80 million. In one case, we're
told that the \$80 million will be paid in 10 equal installments. In the other
case, we're told that the \$80 million will be paid in 10 installments, but the
installments will increase by 5 percent per year. Who got the better deal?

---

To evaluate which athlete got the better deal we need to compare the two
annuities. The differences between the annuities are in the structure of the
payments.

### Annuity 1: Equal installments.

$$
\begin{aligned}
\text{PV} &= C \times \frac{1 - (1 + r)^{-t}}{r}\\
  &= \$8,000,000 \times \frac{1 - (1 + r)^{-10}}{r}
\end{aligned}
$$

### Annuity 2: 5% Growing installments

The payments will grow 5% annually to 80 million. We need to find the starting
payment:

* Total nominal sum

$$
C \times \frac{1.05^{10} - 1}{.05} = 80\ \text{million}
$$

Solving for C:

$$
C \times 12.578 = 80 \rightarrow C \approx 6.36\ \text{million}
$$

Here is a plot of differences between the two annuities for discount rates from
.06 to 1:

```{r concept 3, echo=F}
growth.annuity <- function(c, r, g, t){
    c * ((1 + g)^(0:(t-1))) / (1 + r)^(1:t)
}
cf.1 <- 8
cf.2 <- 6.36

f <- function(r){
  ret <- c()
  for(i in r){
    lhs <- sum(growth.annuity(cf.1, i, .05, 10))
    rhs <- sum(growth.annuity(cf.2, i, 0, 10))
    ret <- c(ret, lhs - rhs)
  }
  return(ret)
}
x <- seq(.06, 1, by=.01)
y <- f(x)
plot(x, y, type='l', ylab='difference in millions of dollars',
    xlab='discount rate', main='Difference between Annuity 1 and Annuity 2')
```

As the plot shows, as the discount rate increase, the difference between the two
annuities diminishes. However, the plot also shows that the discount rate needs
to be extremely high for the difference between the two to become neglible.
Therefore, the growth annuity is the better deal.

---

## 4. APR and EAR

Should lending laws be changed to require lenders to report EARs instead of
APRs? Why or why not?

---

There are a few benefits in requiring lenders to report EARs instead of APRs.
APRs are effectively meaningless without the compounding interval. Thus a 10
percent APR is only meaningful if you know it will be compouned semi-annually,
or quarterly, etc... EARs are meaningful without the compounding interval
because it is the effective annual interest rate. This becomes especially useful
to know when the interest rate is large because the difference between the APR
and the EAR can be dramatic.

---

## 5. Time value

On subsidized Stafford loans, a common source of financial aid for college
students, interest does not begin to accrue until repayment begins. Who receives
a bigger subsidy, a freshman or a senior? Explain.

---

A freshman receives a bigger subsidy compared to a senior because of the longer
period before repayment begins. This period is interest-free while the student
is attending school. A freshman has four years before repayment begins, while a
senior only has one year of school left. The interest that is saved during this
period, compounded over the time attending school, makes the subsidy more
valuable to the freshman than the senior. Thus, if both the freshman and the
senior borrow the same amount each year, the freshman accrues a greater
cumulative benefit across the four years, while the senior only benefits for the
final year.

---

Use the following information to answer the next five question:

Toyota Motor Credit Corporation (TMCC), a subsidiary of Toyota Motor
Corporation, offered some securities for sale to the public on March 28, 2008.
Under the terms of the deal, TMCC promised to repay the owner of one of these
securities \$100,000 on March 28, 2038, but investors would receive nothing
until then. Investors paid TMCC \$24,099 for each of these securities; so they
gave up \$24,099 on March 28, 2008, for the promise of a \$100,000 payment 30
years later.

---

## 6. Time value of money

Why would TMCC be willing to accept such a small amount today (\$24,099) in
exchange for a promise to repay about four times that amount (\$100,000) in the
future?

---

TMCC is willing to accept a smaller amount today (\$24,099) in exchange for a
promise to repay \$100,000 in 30 years because of the time value of money and
the concept of discounting future cash flows.

For a future value of \$100,000 and a present value of \$24,099 and a term of 30
years, we can calculate the interest rate TMCC is expecting to pay.

$$
\begin{aligned}
\text{FV} &= \text{PV} \times (1 + r)^{t}\\\\
\frac{\text{FV}}{\text{PV}} &= (1 + r)^t\\\\
\left( \frac{\text{FV}}{\text{PV}} \right)^{1/t} &= 1 + r\\\\
r &= \left( \frac{\text{FV}}{\text{PV}} \right)^{1/t} - 1\\\\
r &= 0.04857629 \approx 5\%
\end{aligned}
$$

This discount rate of 5% reflects the rate investors see as the effects of
inflation, opportunity cost, and the perceived risk of TMCC's ability to fulfill
the obligation. The investors will find this deal attractive because they are
effectively locking in a specific return over 30 years. The discount rate
(yield) aligns with what they consider fair compensation for lending their money
to TMCC for three decades. TMCC uses this willingness to raise capital cheaply
in the present.

Japans own interest rate in March 2008 was 0.5%. This indicates an extremely low
cost of borrowing for domestic markets. However, there is reason to believe that
TMCC wanted to gain access to a global base of investors as a funding source.
With Japanese securities yielding only 0.5%, TMCC's securities yielding 5% look
extremely attractive.

---

## 7. Call provisions

TMCC has the right to buy back the securities on the anniversary date at a price
established when the securities were issued (this feature is a term of this
particular deal). What impact does this feature have on the desirability of this
security as an investment?

---

This feature has a negative impact on the desirability of the security. It
limits the upside potential of the security because if rates fall or if the
value of the security increases (due to falling discount rates or inflation
expectations), TMCC is likely to exercise its option to buy back the securities.
Also, because the security is callable, the benefit of the investor to lock in
an investment for 30 years is not guaranteed.

---

## 8. Time value of money

Would you be willing to pay \$24,099 today in exchange for \$100,000 in 30
years? What would be the key considerations in answering yes or no? Would your
answer depend on who is making the promise to repay?

---

The implied rate of return for this security was calculated as about 5%. You
would compare this with other investment options with similar risk profiles. You
would also consider inflation which can erode the real value of the investment.
Finally, you would need to evaluate the risk and creditworthiness of TMCC. This
means it does matter who is making the promise to repay.

---

## 9. Investment comparison

Suppose that when TMCC offered the security for \$24,099 the U.S. Treasury had
offered an essentially identical security. Do you think it would have had a
higher or lower price? Why?

---

It would have a higher price because the U.S. Treasury is less risky than the
TMCC security. The security is also more liquid because there are more buyers
for the security than for the TMCC security.

---

## Length of investment

The TMCC security is bought and sold on the New York Stock Exchange. If you
looked at the price today, do you think the price would exceed the \$24,099
original price? Why? If you looked in the year 2019, do you think the price
would be higher or lower than today's price? Why?

---

Yes the price of the TMCC security is likely to have increased.

---

# Questions and problems

## 1. Simple interest versus compound interest

First City Bank pays 8 percent simple interest on its savings account balances,
whereas Second City Bank pays 8 percent interest compounded annually. If you
made a \$5,000 deposit in each bank, how much more money would you earn from
your Second City Bank account at the end of 10 years?

---

```{r question 1; setup, echo=F}
p1 <- new.env()
p1$r <- .08
p1$deposit <- 5000
p1$years <- 10
p1$simple <- p1$deposit * (1 + p1$years * p1$r)
p1$compound <- p1$deposit * (1 + p1$r)^(p1$years)
p1$difference <- abs(p1$simple - p1$compound)
```

For a \$5,000 deposit in First Bank with simple interest you simply calculate
the value of the deposit as such:


$$
\begin{aligned}
\text{Simple interest} &= C \cdot (1 + n \times r)\\
 &= \$`r pretty(p1$deposit, 0)` \times (1 + `r p1$years` \times `r p1$r`)\\
 &= \$`r pretty(p1$simple, 0)`
\end{aligned}
$$

Second Bank uses compound interest.

$$
\begin{aligned}
\text{Compound interest} &= C \cdot (1 + r)^{n}\\
 &= \$`r pretty(p1$deposit, 0)` \times (1 + `r p1$r`)^{`r p1$years`}\\
 &= \$`r pretty(p1$compound, 0)`
\end{aligned}
$$

The difference between the two is:

$$
\$`r pretty(p1$compound, 0)` - \$`r pretty(p1$simple, 0)` = \$`r pretty(p1$difference, 0)`
$$

---

## 2. Calculating future values

Compute the future value of \$1,000 compounded annually for

a. 10 years at 5 percent.

b. 10 years at 10 percent.

c. 20 years at 5 percent.

d. Why is the interest earned in part (c) not twice the amount earned 
   in part (a)?

---

```{r problem 2; setup, echo=F}
p2 <- new.env()
p2$C <- 1000
p2$years <- c(10, 10, 20)
p2$rate <- c(.05, .1, .05)
p2$future.value <- p2$C * (1 + p2$rate)^(p2$years)
```

The future value of an investment can be computed:

$$
\text{FV} = C_0 \times (1 + r)^t
$$

$$
\begin{aligned}
\mathbf{A}:\ & `r pretty(p2$C, 0)`
  \times (1 + `r p2$rate[1]`)^{`r p2$years[1]`}\\
  &= `r pretty(p2$future.value[1])`\\\\
\mathbf{B}:\ & `r pretty(p2$C, 0)`
  \times (1 + `r p2$rate[2]`)^{`r p2$years[2]`}\\
  &= `r pretty(p2$future.value[2])`\\\\
\mathbf{C}:\ & `r pretty(p2$C, 0)`
  \times (1 + `r p2$rate[3]`)^{`r p2$years[3]`}\\
  &= `r pretty(p2$future.value[3])`\\\\
\end{aligned}
$$

The interest in part (c) is not twice the amount earned in part (a) because
compound interest grows exponentially, not linearly. Over longer periods, the
interest itself earns additional interest, which accelerates the growth of the
future value. This effect, called "compounding," results in a higher future
value for part (c) than simply doubling the future value from part (a).  If the
interest were simple (not compounded), then the earnings in part (c) would
indeed be twice those in part (a).

Book: (a) \$1,628.89, (b) \$2,593.74, (c) \$2,653.30

---

## 3. Calculating present values

For each of the following, compute the present value:

```{r problem 3, echo=F}
p3 <- new.env()
p3$years <- c(6, 9, 18, 23)
p3$interest.rate <- c(.07, .15, .11, .18)
p3$future.value <- c(13827, 43852, 725380, 590710)
p3$present.value <- rep('-', 4)

p3$tbl <- data.frame(
  years=p3$years,
  interest.rate=percent(p3$interest.rate, 1),
  future.value=pretty(p3$future.value, 0),
  present.value=p3$present.value
)
table(p3$tbl)
```

```{r problem 3; solution, echo=F}
p3$pv <- round(p3$future.value /
  (1 + p3$interest.rate)^(p3$years))
p3$tbl$present.value <- pretty(p3$pv, 0)
```

The formula for calculating the present value from the future value is:

$$
\begin{aligned}
\text{PV} &= \frac{\text{FV}}{(1 + r)^t}\\\\
\text{PV}_{\text{6 years}} &= \frac{\$`r pretty(p3$future.value[1], 0)`}
  {`r p3$interest.rate[1]`} = \$`r pretty(p3$pv[1], 0)`\\\\
\text{PV}_{\text{9 years}} &= \frac{\$`r pretty(p3$future.value[2], 0)`}
  {`r p3$interest.rate[2]`} = \$`r pretty(p3$pv[2], 0)`\\\\
\text{PV}_{\text{18 years}} &= \frac{\$`r pretty(p3$future.value[3], 0)`}
  {`r p3$interest.rate[3]`} = \$`r pretty(p3$pv[3], 0)`\\\\
\text{PV}_{\text{23 years}} &= \frac{\$`r pretty(p3$future.value[4], 0)`}
  {`r p3$interest.rate[4]`} = \$`r pretty(p3$pv[4], 0)`\\\\
\end{aligned}
$$

```{r problem 3; solution , echo=F}
table(p3$tbl)
```

---

## 4. Calculating interest rates

Solve for the unknown interst rate in each of the following:

```{r problem 4; setup, echo=F}
p4 <- new.env()
p4$years = c(4, 8, 16, 27)
p4$present.value = c(242, 410, 51700, 18750)
p4$future.value = c(307, 896, 162181, 483500)
p4$tbl <- data.frame(
  years=p4$years,
  present.value=pretty(p4$present.value, 0),
  future.value=pretty(p4$future.value, 0)
)
table(p4$tbl)
```

---

To calculate the interest rate we solve for $r$:

$$
\begin{aligned}
\text{FV} &= \text{PV} \times (1 + r)^t\\\\
\frac{\text{FV}}{\text{PV}} &= (1 + r)^t\\\\
\left( \frac{\text{FV}}{\text{PV}} \right)^{1/t} &= 1 + r\\\\
r &= \left( \frac{\text{FV}}{\text{PV}} \right)^{1/t} - 1
\end{aligned}
$$

```{r problem 4; solution, echo=F}
p4$interest.rate <- (p4$future.value / p4$present.value)^(1/(p4$years)) - 1
p4$cols <- colnames(p4$tbl)
p4$tbl <- cbind(p4$tbl, percent(p4$interest.rate, 0.01))
colnames(p4$tbl) <- c(p4$cols, 'interest.rate')
table(p4$tbl)
```

#### NOTE:

The solution to this problem is the same as calculating the compound annual
growth rate (CAGR):

$$
\mathbf{CAGR} = \left(\frac{\text{ending value}}
  {\text{beginning value}}\right)^{1/t} - 1
$$

Book: 6.13\%, 10.27\%, 7.41\%, 12.79\%.

---

## 5. Calculating the number of periods

Solve for the years:

```{r problem 5, echo=F}
p5 <- new.env()
p5$present.value = c(625, 810, 18400, 21500)
p5$interest.rate = c(.09, .11, .17, .08)
p5$future.value = c(1284, 4341, 402662, 173439)
p5$years = rep('-', 4)

p5$tbl <- data.table(
  present.value = pretty(p5$present.value, 0),
  interest.rate = percent(p5$interest.rate, 1),
  future.value  = pretty(p5$future.value, 0),
  years         = p5$years
)
table(p5$tbl)
```

---

We solve for the period:

$$
\begin{aligned}
\mathbf{FV} &= \text{PV} \times (1 + r)^t\\\\
\frac{\text{FV}}{\text{PV}} &= (1 + r)^t\\\\
\ln{\left(\frac{\text{FV}}{\text{PV}} \right)} &= t \ln{(1 + r)}\\\\
t &= \frac{\ln{\left(\frac{\text{FV}}{\text{PV}} \right)}}{\ln{(1 + r)}}
\end{aligned}
$$

```{r solution 5, echo=F}
p5$years <- log(p5$future.value/p5$present.value) /
            log(1 + p5$interest.rate)
p5$tbl$years <- round(p5$years, 2)
table(p5$tbl)
```

---

## 6. Calculating the number of periods

At 8 percent interest, how long does it take to double your money? To quadruple
it?

---

This depends on if we want to use simple or compound interest.

### Simple interest

#### Double

$$
\begin{aligned}
2X &= X \times (1 + t * .08)\\
2 &= 1 + t \times .08\\
\frac{1}{.08} &= t\\
t &= 12.5
\end{aligned}
$$

For simple interest, it will take 12.5 periods to double.

#### Quadruple

$$
t = \frac{4 - 1}{.08} = 37.5
$$

It will take 37.5 periods to quadruple.

### Compound interest

#### Double

$$
\begin{aligned}
2X &= X \times (1.08)^t\\
2 &= (1.08)^t\\
\ln{2} &= t \ln{1.08}\\
t &= \frac{\ln{2}}{\ln{1.08}}\\\\
t &= 9
\end{aligned}
$$

With compound interest it takes about 9 periods to double.

#### Quadruple

$$
t = \frac{\ln{4}}{\ln{1.08}} = 18
$$

It will take 18 periods to quadruple.

---

## 7. Calculating present values

Imprudential, Inc., has an unfunded pension liability of \$630 million that must
be paid in 20 years. To assess the value of the firm's stock, financial analysts
want to discount this liability back to the present. If the relevant discount
rate is 7.1 percent, what is the present value of this liability?

---

```{r problem 7; setup, echo=F}
p7 <- new.env()
p7$pension <- 630
p7$years <- 20
p7$discount <- .071
p7$present.value <- p7$pension/(1 + p7$discount)^(p7$years)
```

$$
\begin{aligned}
\text{PV} &= \frac{\text{FV}}{(1 + r)^t}\\
  &= \frac{\$`r pretty(p7$pension, 0)`}{(1 + `r p7$discount`)^{`r p7$years`}}\\
  &= \$`r pretty(p7$present.value, 0)`
\end{aligned}
$$

The liability is presently worth \$`r pretty(p7$present.value, 0)` million.

---

## 8. Calculating rates of return

Although appealing to more refined tastes, art as a collectible has not always
performed so profitably. During 2010, Deutscher-Menzies sold _Arkies under the
Shower_, a painting by renowned Australian painter Brett Whiteley, at auction
for a price of \$1,100,000. Unfortunately for the previous owner, he had
purchased it three years earlier at a price of \$1,680,000. What was his annual
rate of return on this painting?

---

```{r problem 8; setup, echo=F}
p8 <- new.env()
p8$ending.value <- 1.1e6
p8$beginning.value <- 1.68e6
p8$years <- 3
p8$ratio <- p8$beginning.value / p8$ending.value
```

### Step 1: Calculate the ratio of present value over future value

$$
\frac{\text{PV}}{\text{FV}} = \frac{`r pretty(p8$beginning.value, 0)`}
  {`r pretty(p8$ending.value, 0)`} = `r p8$ratio`
$$

### Step 2: Determine the interest rate that gives this value

We can use `R`'s `uniroot` to find this value:

```{r problem8; uniroot solution, echo=T}
p8$f <- function(r) (1 + r)^(-p8$years) - p8$ratio
percent(p8$r <- uniroot(p8$f, upper=1, lower=-1, tol=1e-4)$root)
```

The annual rate of return is `r percent(p8$r)`.

---

## 9. Perpetuities

An investor purchasing a British consol is entitled to receive annual payments
from the British government forever. What is the price of a consol that pays
\$150 annually if the next payment occurs one year from today? The market
interest rate is 4.6 percent.

---

```{r problem 9; setup, echo=F}
p9 <- new.env()
p9$C <- 150
p9$rate <- .046
p9$perpetuity <- p9$C / p9$rate
p9$perpetuity.next.pay <- p9$C + p9$C / p9$rate
```

$$
\begin{aligned}
\text{PV} &= \frac{C}{r}\\
  &= \frac{\$`r p9$C`}{`r p9$rate`}\\
  &= \$`r pretty(p9$perpetuity)`
\end{aligned}
$$

### NOTE:

The value of the consol will depend on when the next payment is. For example,
instead of the next payment being one year from now, what if the next payment
was tomorrow instead? What is the new present value of the consol? Well, since
the next payment is tomorrow then we have to include the next payment into the
calculation. The present value is calculated:

$$
\begin{aligned}
\text{PV}_{\text{pay tomorrow}} &= \$`r p9$C` + \frac{\$`r p9$C`}{`r p9$rate`}\\
  &= \$`r pretty(p9$perpetuity.next.pay)`
\end{aligned}
$$

You can see this dynamic play out in the stock market as well with stocks that
pay dividends. Just after the ex-dividend date, the stock price will drop
according to the value of the present value of the stock's dividend (its cash
dispursments to shareholders.

---

## 10. Continuous compounding

Compute the future value of \$1,900 continuously compounded for:

a. 7 years at a stated annual interest rate of 12 percent.
b. 5 years at a stated annual interest rate of 10 percent.
c. 12 years at a stated annual interest rate of 5 percent.
d. 10 years at a stated annual interest rate of 7 percent.

---

```{r problem 10; setup, echo=F}
p10 <- new.env()
p10$pv <- 1900
p10$n <- c(7, 5, 12, 10)
p10$r <- c(.12, .1, .05, .07)
p10$fv <- p10$pv * exp(p10$n * p10$r)
```

Continuous compounding

$$
C_0 \times e^{rT}
$$

$$
\begin{aligned}
\mathbf{A}:\ &\ `r pretty(p10$pv, 0)`
  \times e^{`r p10$r[1]` \times `r p10$n[1]`}\\
  &= `r pretty(p10$fv[1])`\\\\
\mathbf{B}:\ &\ `r pretty(p10$pv, 0)`
  \times e^{`r p10$r[2]` \times `r p10$n[2]`}\\
  &= `r pretty(p10$fv[2])`\\\\
\mathbf{C}:\ &\ `r pretty(p10$pv, 0)`
  \times e^{`r p10$r[3]` \times `r p10$n[3]`}\\
  &= `r pretty(p10$fv[3])`\\\\
\mathbf{D}:\ &\ `r pretty(p10$pv, 0)`
  \times e^{`r p10$r[4]` \times `r p10$n[4]`}\\
  &= `r pretty(p10$fv[4])`\\\\
\end{aligned}
$$

Book: (a) \$4,401.10, (b) \$3,132.57, (c) \$3,462.03, (d) \$3,826.13.

---

## 11. Present value and multiple cash flows

Conoly Co. has identified an investment project with the following cash flows.
If the discount rate is 10 percent, what is the present value of these cash
flows? What is the present value at 18 percent? At 24 percent?

```{r problem 11; setup, echo=F}
p11 <- new.env()
p11$year <- c(1, 2, 3, 4)
p11$cf <- c(960, 840, 935, 1350)
p11$r <- c(.1, .18, .24)

p11$tbl <- data.frame(
  year = p11$year,
  cash.flow = p11$cf
)

table(p11$tbl)
p11$pv.f <- function(r) sum(p11$cf / (1 + r)^(p11$year))
p11$pv <- unlist(lapply(p11$r, p11$pv.f))
```

---

The formula for present value:

$$
\text{PV} = \sum{\frac{\text{FV}}{(1 + r)^t}}
$$

For a rate of 10 percent the cash flows are:

$$
\text{PV} =
    \frac{\$`r p11$cf[1]`}{(1 + `r p11$r[1]`)^{`r p11$year[1]`}}
  + \frac{\$`r p11$cf[2]`}{(1 + `r p11$r[1]`)^{`r p11$year[2]`}}
  + \frac{\$`r p11$cf[3]`}{(1 + `r p11$r[1]`)^{`r p11$year[3]`}}
  + \frac{\$`r p11$cf[4]`}{(1 + `r p11$r[1]`)^{`r p11$year[4]`}}
  = \$`r pretty(p11$pv[1], 0)`
$$

For a rate of 18 percent:

$$
\text{PV} =
    \frac{\$`r p11$cf[1]`}{(1 + `r p11$r[2]`)^{`r p11$year[1]`}}
  + \frac{\$`r p11$cf[2]`}{(1 + `r p11$r[2]`)^{`r p11$year[2]`}}
  + \frac{\$`r p11$cf[3]`}{(1 + `r p11$r[2]`)^{`r p11$year[3]`}}
  + \frac{\$`r p11$cf[4]`}{(1 + `r p11$r[2]`)^{`r p11$year[4]`}}
  = \$`r pretty(p11$pv[2], 0)`
$$

For a rate of 24 percent:

$$
\text{PV} =
    \frac{\$`r p11$cf[1]`}{(1 + `r p11$r[3]`)^{`r p11$year[1]`}}
  + \frac{\$`r p11$cf[2]`}{(1 + `r p11$r[3]`)^{`r p11$year[2]`}}
  + \frac{\$`r p11$cf[3]`}{(1 + `r p11$r[3]`)^{`r p11$year[3]`}}
  + \frac{\$`r p11$cf[4]`}{(1 + `r p11$r[3]`)^{`r p11$year[4]`}}
  = \$`r pretty(p11$pv[3], 0)`
$$

We will calculate for each of the different discount rates:

* The present value of the cash flows at 10 percent: \$`r pretty(p11$pv[1])`

* The present value of the cash flows at 18 percent: \$`r pretty(p11$pv[2])`

* The present value of the cash flows at 24 percent: \$`r pretty(p11$pv[3])`

---

## 12. Present value and multiple cash flows

Investment X offers to pay you \$4,500 per year for nine years, whereas
Investment Y offers to pay you \$7,000 per year for five years. Which of these
cash flow streams has the higher present value if discount rate is 5 percent? If
the discount rate is 22 percent?

---

```{r problem 12; setup, echo=F}
p12 <- new.env()
p12$X.cf <- 4500
p12$X.n <- 9
p12$Y.cf <- 7000
p12$Y.n <- 5
p12$r <- c(.05, .22)
p12$pvia <- function(r, n) sum(1 / (1 + r)^(1:n))
p12$X.pv <- p12$X.cf * unlist(lapply(p12$r, p12$pvia, p12$X.n))
p12$Y.pv <- p12$Y.cf * unlist(lapply(p12$r, p12$pvia, p12$Y.n))
```

### Given

* Investment X cash flow: \$`r pretty(p12$X.cf, 0)`

* Term for investment X: `r p12$X.n` years

* Investment Y cash flow: \$`r pretty(p12$Y.cf, 0)`

* Term for investment Y: `r p12$Y.n` years

* Discount rates: 5\% and 22\%.

We solve this with the annuity equation:

$$
\begin{aligned}
\text{PV}_{\text{annuity}} &= C \cdot \left[\frac{1 - (1 + r)^{-T}}{r}\right]
\end{aligned}
$$

### For interest rate 5\%

$$
\begin{aligned}
\text{PV}_{\text{X}} &= `r pretty(p12$X.cf, 0)`
  \times \left[\frac{1 - (`r 1+p12$r[1]`)^{- `r p12$X.n`}}{`r p12$r[1]`}\right]\\
  &= `r pretty(p12$X.pv[1])`\\\\
\text{PV}_{\text{Y}} &= `r pretty(p12$Y.cf, 0)`
  \times \left[\frac{1 - (`r 1+p12$r[1]`)^{- `r p12$Y.n`}}{`r p12$r[1]`}\right]\\
  &= `r pretty(p12$Y.pv[1])`
\end{aligned}
$$

For the lower interest rate, the longer timespan makes investment X more
attractive.

### For interest rate 22\%

$$
\begin{aligned}
\text{PV}_{\text{X}} &= `r pretty(p12$X.cf, 0)`
  \times \left[\frac{1 - (`r 1+p12$r[2]`)^{- `r p12$X.n`}}{`r p12$r[2]`}\right]\\
  &= `r pretty(p12$X.pv[2])`\\\\
\text{PV}_{\text{Y}} &= `r pretty(p12$Y.cf, 0)`
  \times \left[\frac{1 - (`r 1+p12$r[2]`)^{- `r p12$Y.n`}}{`r p12$r[2]`}\right]\\
  &= `r pretty(p12$Y.pv[2])`
\end{aligned}
$$

However, for higher interest rates, the longer timeframe reduces the
contribution of the cash flows further out. Thus, investment Y is a better
choice.

---

## 13. Calculating annuity present value

An investment offers \$4,900 per year for 15 years, with the first payment
occurring one year from now. If the required return is 8 percent, what is the
value of the investment? What would the value be if the payments occurred for 40
years? For 75 years? Forever?

---

```{r problem 13; setup, echo=F}
p13 <- new.env()
p13$cf <- 4900
p13$rate <- .08
p13$n <- c(15, 40, 75)
p13$pv <- unlist(lapply(p13$n, function(n) sum(p13$cf/((1 + p13$rate)^(1:n)))))
p13$pv.perpetuity <- p13$cf/p13$rate
```

$$
\begin{aligned}
\text{PV}_{\text{annuity}} &= \text{CF}
  \times \left[\frac{1 - (1 + r)^{-T}}{r}\right]\\\\
\text{PV}_{\text{15 years}} &= `r pretty(p13$cf, 0)`
  \times \left[
    \frac{1 - (1 + `r p13$rate`)^{- `r p13$n[1]`}}{`r p13$rate`}
  \right] \\
  &= `r pretty(p13$pv[1])`\\\\
\text{PV}_{\text{40 years}} &= `r pretty(p13$cf, 0)`
  \times \left[
    \frac{1 - (1 + `r p13$rate`)^{- `r p13$n[2]`}}{`r p13$rate`}
  \right] \\
  &= `r pretty(p13$pv[2])`\\\\
\text{PV}_{\text{75 years}} &= `r pretty(p13$cf, 0)`
  \times \left[
    \frac{1 - (1 + `r p13$rate`)^{- `r p13$n[3]`}}{`r p13$rate`}
  \right] \\
  &= `r pretty(p13$pv[3], 0)`\\\\
\text{PV}_{\text{perpetuity}} &= \frac{CF}{r}
   = \frac{`r pretty(p13$cf)`}{`r p13$rate`}\\
  &= `r pretty(p13$pv.perpetuity)`
\end{aligned}
$$

---

## 14. Calculating perpetuity value

The Perpetual Life Insurance Co. is trying to sell you an investment policy that
will pay you and your heirs \$15,000 per year forever. If the required return on
this investment is 5.2 percent, how much will you pay for the policy? Suppose
the Perpetual Life Insurance Co. told you the policy costs \$320,000. At what
interest rate would this be a fair deal?

---

```{r problem 14; setup, echo=F}
p14 <- new.env()
p14$cf <- 15000
p14$rate <- .052
p14$pv <- p14$cf / p14$rate
p14$pv.given <- 320000
p14$rate.calculated <- p14$cf / p14$pv.given
```

The perpetuity formula is:

$$
\begin{aligned}
\text{PV}_{\text{perpetuity}} &= \frac{C_0}{r}\\
  &= \frac{`r pretty(p14$cf, 0)`}{`r p14$rate`}\\
  &= \$`r pretty(p14$pv)`\\\\
\end{aligned}
$$

For a policy that costs \$320,000, the interest rate should be:

$$
\begin{aligned}
r &= \frac{C_0}{\text{PV}_{\text{perpetuity}}} \\
  &= \frac{`r pretty(p14$cf, 0)`}{`r pretty(p14$pv.given, 0)`}\\
  &= `r percent(p14$rate.calculated)`
  \%
\end{aligned}
$$

Book: \$88,461.54, 4.69\%

---

## 15. Calculating EAR

Find the EAR in each of the following cases:

```{r problem 15; setup, echo=F}
p15 <- new.env()
p15$apr <- c(.07, .16, .11, .12)
p15$compound <- c(4, 12, 365, NA)
p15$tbl <- data.frame(
  apr = percent(p15$apr, 1),
  compound = c("Quarterly", "Monthly", "Daily", "Infinite")
)
table(p15$tbl)
```

```{r problem 15; solution, echo=F}
# First three are not infinite
p15$ear <- (1 + p15$apr[1:3] / p15$compound[1:3])^(p15$compound[1:3]) - 1
# The infinite compounding uses a special formula
p15$ear <- c(p15$ear, exp(p15$apr[4]) - 1)
p15$cols <- colnames(p15$tbl)
p15$tbl <- cbind(p15$tbl, percent(p15$ear))
colnames(p15$tbl) <- c(p15$cols, "ear")
```
To calculate the EAR from the APR we need to know the compounding period. Then
we can use the following two formulas:

### Set period compounding (Quarterly, Monthly, Daily, etc...)

$$
\begin{aligned}
\text{EAR} &= \left(1 + \frac{r}{m}\right)^{m} - 1\\\\
\text{EAR}_{\text{7\% qtr}} &= \left(1 + \frac{`r p15$apr[1]`}
  {`r p15$compound[1]`}\right)^{`r p15$compound[1]`} - 1\\
  &= `r percent(p15$ear[1])`
  \%\\\\
\text{EAR}_{\text{16\% monthly}} &= \left(1 + \frac{`r p15$apr[2]`}
  {`r p15$compound[2]`}\right)^{`r p15$compound[2]`} - 1\\
  &= `r percent(p15$ear[2])`
  \%\\\\
\text{EAR}_{\text{11\% daily}} &= \left(1 + \frac{`r p15$apr[3]`}
  {`r p15$compound[3]`}\right)^{`r p15$compound[3]`} - 1\\
  &= `r percent(p15$ear[3])`
  \%\\\\
\end{aligned}
$$

### Infinite compounding

$$
\begin{aligned}
\text{EAR} &= e^r - 1\\\\
\text{EAR}_{\text{12\% continuous}} &= e^{`r p15$apr[4]`} - 1\\
  &= `r percent(p15$ear[4])`
  \%
\end{aligned}
$$

### Completed table

```{r problem 15; display  table, echo=F}
table(p15$tbl)
```

---

## 16. Calculating APR

Find the APR, or stated rate, in each of the following cases:

```{r problem 16; setup, echo=F}
p16 <- new.env()
p16$ear <- c(.098, .196, .083, .143)
p16$compound <- c(2, 12, 52, NA)

p16$tbl <- data.frame(
  ear = percent(p16$ear),
  compound = c("Semiannually", "Monthly", "Weekly", "Infinite")
)
table(p16$tbl)
```

---

```{r problem 16; solution, echo=F}
p16$apr <- p16$compound[1:3] * ((p16$ear[1:3] + 1)^(1/p16$compound[1:3]) - 1)
p16$apr <- c(p16$apr, log(p16$ear[4] + 1))
```

To calculate the APR from the EAR we work the EAR equation backwards to:

$$
\begin{aligned}
\text{APR}_{\text{noncontinuous}} &=
  m \times \left[\left(1 + \text{EAR}\right)^{\frac{1}{m}} - 1  \right]\\\\
\text{APR}_{\text{9.8\% semiannual}} &= 
  `r p16$compound[1]` \times
  \left[(1 + `r p16$ear[1]`)^{1/`r p16$compound[1]`} - 1\right]\\
  &= `r percent(p16$apr[1])`
  \%\\\\
\text{APR}_{\text{19.6\% monthly}} &= 
  `r p16$compound[2]` \times
  \left[(1 + `r p16$ear[2]`)^{1/`r p16$compound[2]`} - 1\right]\\
  &= `r percent(p16$apr[2])`
  \%\\\\
\text{APR}_{\text{8.3\% weekly}} &= 
  `r p16$compound[3]` \times
  \left[(1 + `r p16$ear[3]`)^{1/`r p16$compound[3]`} - 1\right]\\
  &= `r percent(p16$apr[3])`
  \%\\\\
\text{APR}_{\text{continous}} &= \ln{(1 + \text{EAR})}\\
  &= \ln{(1 + `r p16$ear[4]`)}\\
  &= `r percent(p16$apr[4])`
  \%\\\\
\end{aligned}
$$

```{r problem 16; display table, echo=F}
p16$tbl <- cbind(p16$tbl, percent(p16$apr))
colnames(p16$tbl)[3] <- "apr"
table(p16$tbl)
```

---

## 17. Calculating EAR

First National Bank charges 11.2 percent compounded monthly on its business
loans. First United Bank charges 11.4 percent compounded semiannually. As a
potential borrower, to which bank would you go for a new loan?

---

```{r problem 17; setup, echo=F}
p17 <- new.env()
p17$rate.national <- .112
p17$m.national <- 12
p17$ear.national <- (1 + p17$rate.national /
  p17$m.national)^(p17$m.national) - 1
p17$rate.united <- .114
p17$m.united <- 2
p17$ear.united <- (1 + p17$rate.united /
  p17$m.united)^(p17$m.united) - 1
```

$$
\begin{aligned}
\text{EAR} &= \left( 1 + \frac{r}{m}  \right)^m - 1\\\\
\text{EAR}_{\text{First National Bank}} &=
  \left(1 + \frac{`r p17$rate.national`}
  {`r p17$m.national`}\right)^{`r p17$m.national`} - 1\\
  &= `r percent(p17$ear.national)`
  \%\\\\
\text{EAR}_{\text{First United Bank}} &=
  \left(1 + \frac{`r p17$rate.united`}
  {`r p17$m.united`}\right)^{`r p17$m.united`} - 1 \\
  &= `r percent(p17$ear.united)`
  \%
\end{aligned}
$$

The EARs are similar for the two banks. However, I would get my loan from First
United Bank since the EAR rate is lower.

---

## 18. Interest rates

Well-known financial writer Andrew Tobias argues that he can earn 177 percent
per year buying wine by the case. Specifically, he assumes that he will consume
one \$10 bottle of fine Bordeaux per week for the next 12 weeks. He can either
pay \$10 per week or buy a case of 12 bottles today. If he buys the case, he
receives a 10 percent discount and, by doing so, earns the 177 percent. Assume
he buys the wine and consumes the first bottle today. Do you agree with his
analysis? Do you see a problem with his numbers?

---

```{r problem 18; setup, echo=F}
p18 <- new.env()
```

### Scenario Summary:

* Cost per bottle if paid weekly: $\$10 \times 12 = \$120$

* Cost for buying a case today (10\% discount): $\$120 \times 0.90 = \$108$

* Savings: \$12

* Wine is consumed weekly over 12 weeks.

* He drinks one bottle immediately (Week 0), then one each week until Week 11.

### Financial framing

The comparison is between two cash flow patterns:

1. Installment plan (implicit): \$10 per week for 12 weeks.

2. Upfront payment: \$108 now.

We can treat this like a loan: Tobias "invests" \$108 today to avoid paying
\$10/week for 12 weeks. The IRR of this investment tells us the effective
return.

### IRR caclulation:

We solve for the interest rate `r` that satisfies:

$$
108 = \sum^{11}_{t=0}{\frac{10}{(1 + r)^t}}
$$

This is the present value of a 12-week annuity of \$10 payments, discounted at
weekly rate `r`.

We can determine the rate using R's `uniroot`

```{r problem 18; show use of uniroot, echo=T}
p18$cost.per.bottle <- 10
p18$weeks <- 12
p18$discount <- .9
p18$savings <- p18$cost.per.bottle * p18$weeks * p18$discount
p18$t <- 0:(p18$weeks-1) # first bottle is consumed today.
f <- function(r) sum(p18$cost.per.bottle / (1 + r)^(p18$t)) - p18$savings
(p18$irr <- uniroot(f, upper=1, lower=.01, tol=1e-4)$root)
```

```{r problem 18; calculate ear, echo=F}
p18$ear <- (1 + p18$irr)^(52) - 1
```

To annualize:

$$
(1 + \text{IRR})^{52} - 1 = (1 + `r p18$irr`)^{52} - 1 = `r percent(p18$ear)`
\%
$$

However, it is misleading to compare this to an investment return---it's a rate
of savings on a fixed set of planned purchases.

Book: EAR = 176.68\%

---

## 19. Calculating number of periods

One of your customers is delinquent on his accounts payable balance. You've
mutually agreed to a repayment schedule of \$700 per month. You will charge 1.3
percent per month interest on the overdue balance. If the current balance is
\$21,500, how long will it take for the account to be paid off?

---

```{r problem 19; setup, echo=F}
p19 <- new.env()
p19$monthly.payment <- 700
p19$rate <- .013
p19$balance <- 21500
p19$n <- - (log(1 - (p19$balance * p19$rate)/(p19$monthly.payment))) /
            log(1 + p19$rate)
```

We are asked to determine how long it will take to pay off a \$21,500 debt if
the customer pays \$700 per month, and interest accrues at 1.3\% per month.

### This is a present value of an annuity problem

You are solving for n, the number of months needed to pay off the loan, given:

* Present value (PV): \$21,500

* Payment (PMT): \$700

* Monthly interest rate (r): .013 (1.3\%)

Solving for $n$:

$$
\begin{aligned}
\text{\text{PV}} &= C \left[\frac{1 - (1 + r)^{-n}}{r}\right]\\
\frac{\text{PV}}{C} &= \frac{1 - (1 + r)^{-n}}{r}\\
\frac{\text{PV} \cdot r}{C} &= 1 - (1 + r)^{-n}\\
(1 + r)^{-n} &= 1 - \frac{\text{PV} \cdot r}{C}\\
-n \log{(1 + r)} &= \log{(1 - \frac{\text{PV} \cdot r}{C})}\\\\
n &= - \frac{\log{\left(1 - \frac{\text{PV} \cdot r}{C}\right)}}
  {\log{(1 + r)}}\\
  &= - \frac{\log{\left(1 - \frac{`r pretty(p19$balance)` \cdot `r p19$rate`}
  {`r p19$monthly.payment`}\right)}}{\log{(1 + `r p19$rate`)}}\\
  &= `r round(p19$n, 2)`\ \text{months}
\end{aligned}
$$

---

## 20. Calculating EAR

Friendly's Quick Loans, Inc., offers you "three for four or I knock on your
door." This means you get \$3 today and repay \$4 when you get your paycheck in
one week (or else). What's the effective annual return Friendly's earns on this
lending business? If you were brave enough to ask, what APR would Friendly's say
you were paying?

---

```{r problem 20; setup, echo=F}
p20 <- new.env()
p20$pv <- 3
p20$fv <- 4
p20$r <- p20$fv / p20$pv - 1
p20$apr <- p20$r * 52
p20$ear <- (1 + p20$r)^(52) - 1
```

Find the weekly interest rate:

$$
\begin{aligned}
\text{PV} &= \frac{\text{FV}}{1 + r}\\
r &= \frac{\text{FV}}{\text{PV}} - 1\\
  &= \frac{`r p20$fv`}{`r p20$pv`} - 1\\
  &= `r round(p20$r, 2)`
\end{aligned}
$$

Convert to APR:

$$
\begin{aligned}
\text{APR} &= r_{\text{weekly}} \times 52\\
  &= `r round(p20$r, 2)` \times 52\\
  &= `r round(p20$apr, 3)`
\end{aligned}
$$

Find the EAR:

$$
\begin{aligned}
\text{EAR} &= (1 + r_{\text{weekly}})^{52} - 1\\
  &= (1 + `r round(p20$r, 2)`)^{52} - 1\\
  &= `r round(p20$ear, 3)`
\end{aligned}
$$

---

## 21. Future value

What is the future value in six years \$1,000 invested in an account with a
stated annual interest rate of 9 percent,

a. Compounded annually?
b. Compounded semiannually?
c. Compounded monthly?
d. Compounded continuously?
e. Why does the future value increase as the compounding period shortens?

---

```{r problem 21; setup, echo=F}
p21 <- new.env()
p21$t <- 6
p21$pv <- 1000
p21$r <- .09
p21$m <- c(1, 2, 12)
p21$fv <- p21$pv * (1 + p21$r / p21$m )^(p21$m * p21$t)
(p21$fv <- c(p21$fv, p21$pv * exp(p21$r * p21$t)))
```

We use the future value formula with different compounding periods:

$$
\begin{aligned}
\text{FV} &= \text{PV} \times (1 + \frac{r}{m})^{mt}\\\\
\text{FV}_{\text{annually}} &= `r pretty(p21$pv)`
  \times (1 + `r p21$r`/ `r p21$m[1]`)^{`r p21$m[1]` \cdot `r p21$t`}\\
  &= `r pretty(p21$fv[1])`\\\\
\text{FV}_{\text{semiannually}} &= `r pretty(p21$pv)`
  \times (1 + `r p21$r`/ `r p21$m[2]`)^{`r p21$m[2]` \cdot `r p21$t`}\\
  &= `r pretty(p21$fv[2])`\\\\
\text{FV}_{\text{monthly}} &= `r pretty(p21$pv)`
  \times (1 + `r p21$r`/ `r p21$m[3]`)^{`r p21$m[3]` \cdot `r p21$t`}\\
  &= `r pretty(p21$fv[3])`\\\\
\text{FV}_{\text{continuous}} &= \text{PV}\  e^{rT}\\
  &= `r pretty(p21$pv)`\  e^{`r p21$r` \cdot `r p21$t`}\\
  &= `r pretty(p21$fv[4])`
\end{aligned}
$$

The future value increases as the compounding period shortens because interest
is being calculated and added to the balance more frequently. This causes the
investment to earn "interest on interest" sooner and more often, resulting in a
higher overall return. Continuous compounding is the theoretical maximum for
compounding, as it assumes interest is added at every possible instant.

---

## 22. Simple interest versus compound interest

First Simple Bank pays 5 percent simple interest on its investment accounts. If
First Complex Bank pays interest on its accounts compounded annually, what rate
should the bank set if it wants to match First Simple Bank over an investment
horizon of 10 years?

---

```{r problem 22; setup, echo=F}
p22 <- new.env()
p22$r.simple <- .05
p22$t <- 10
p22$r.complex <- (1 + p22$r.simple * p22$t)^(1 / p22$t) - 1
```

We simply equate each banks interest rate structure to each other and determine
what First Complex Bank should set their rate to.

### Simple interest (First Simple Bank)

$$
\begin{aligned}
\text{FV}_{\text{simple}} &= \text{PV} \times ( 1 + r_{\text{simple}} \times t)
\end{aligned}
$$

#### Compound interest (First Comples Bank)

$$
\begin{aligned}
\text{FV}_{\text{complex}} &= \text{PV} \times (1 + r_{\text{complex}})^t
\end{aligned}
$$

### Equate both to each other and solve for $r_{\text{complex}}$

$$
\begin{aligned}
(1 + r_{\text{simple}} \times t) &= (1 + r_{\text{complex}})^{t}\\
(1 + r_{\text{simple}} \times t)^{1/t} &= 1 + r_{\text{complex}}\\\\
r_{\text{complex}} &= (1 + r_{\text{simple}} \times t)^{1/t} - 1\\
 &= (1 + `r p22$r.simple` \times `r p22$t`)^{1 / `r p22$t`} - 1\\
 &= `r percent(p22$r.complex)`
 \%
\end{aligned}
$$

Book: 4.14\%

---

## 23. Calculating annuities

You are planning to save for retirement over the next 30 years. To do this, you
will invest \$800 a month in a stock account and \$350 a month in a bond
account. The return of the stock account is expected to be 11 percent, and the
bond account will pay 6 percent. When you retire, you will combine your money
into an account with an 8 percent return. How much can you withdraw each month
from your account assuming a 25-year withdrawal period?

---

```{r problem 23; setup, echo=F}
p23 <- new.env()
p23$cf.stock <- 800
p23$cf.bond <- 350
p23$n.contribution <- 30
p23$n.withdrawal <- 25
p23$r.stock <- .11
p23$r.bond <- .06
p23$r.withdrawal <- .08
p23$fv.stock <- p23$cf.stock *
  ((1 + (p23$r.stock/12))^(12 * p23$n.contribution) - 1) /
  (p23$r.stock / 12)
p23$fv.bond <- p23$cf.bond *
  ((1 + (p23$r.bond/12))^(12 * p23$n.contribution) - 1) /
  (p23$r.bond / 12)
p23$fv.combined <- p23$fv.stock + p23$fv.bond
p23$fv.withdrawal <- p23$fv.combined *
  (p23$r.withdrawal / 12) /
  (1 - (1 + p23$r.withdrawal / 12)^(- 12 * p23$n.withdrawal))
t <- 1:30
```

The future value of all the contributions can be calculated:

$$
\begin{aligned}
\text{FV} &= C \cdot \sum^N_{n=1}{\left(1 + \frac{r}{m}\right)^{(n-1)}}
\end{aligned}
$$

Where $N = mT$ or the number of years multiplied by the number of months
(monthly compounding).

This is a geometric series that can be broken down:

$$
\begin{aligned}
\text{FV} &= C + C(1 + r/m) + C(1 + r/m)^2 + \cdots + C(1 + r/m)^{(n - 1)}\\\\
(1 + r/m)\ \text{FV} &= C(1 + r/m) + C(1 + r/m)^2 + \cdots + C(1 + r/m)^n\\\\
\text{FV} - (1 + r/m)\ \text{FV} &= C + C(1 + r/m)^n\\\\
\text{FV}\ (1 - (1 + r/m)) &= C (1 - (1 + r/m)^n)\\\\
\text{FV} &= C \cdot \frac{(1 - (1 + r/m)^n)}{1 - (1 + r/m)}\\\\
\text{FV} &= C \cdot \left[\frac{(1 + r/m)^n - 1}{r/m}\right]
\end{aligned}
$$


### Saving for retirement

We calculate the contribution, each month, over a 30 year window. The stock
account will return 11 percent a year, and the bond account will return 6
percent a year.

$$
\begin{aligned}
\text{FV}_{\text{stock}} &= \$`r p23$cf.stock` \times
  \left[
    \frac{(1 + \frac{`r p23$r.stock`}{12})^{(12 \times `r p23$n.contribution`)}
    - 1}{\frac{`r p23$r.stock`}{12}}
  \right]\\
  &= \$`r pretty(p23$fv.stock)`\\\\
\text{FV}_{\text{bond}} &= \$`r p23$cf.bond` \times
  \left[
    \frac{(1 + \frac{`r p23$r.bond`}{12})^{(12 \times `r p23$n.contribution`)}
    - 1}{\frac{`r p23$r.bond`}{12}}
  \right]\\
  &=  \$`r pretty(p23$fv.bond)`\\\\
\text{FV}_{\text{combined}} &=  \$`r pretty(p23$fv.stock)`
  + \$`r pretty(p23$fv.bond)`\\
  &= \$`r pretty(p23$fv.combined)`
\end{aligned}
$$

### Withdrawals during retirement

We use the annuity equation and solve for the withdrawal cash flows ($W$).

$$
\begin{aligned}
\text{PV} &= W \times
  \frac{1 - \left(1 + \frac{r}{m}\right)^{-mT}}{\frac{r}{m}}\\\\
\text{W} &= \text{PV} \times
  \frac{\frac{r}{m}}{1 - \left(1 + \frac{r}{m}\right)^{-mT}}\\
  &= \$`r pretty(p23$fv.combined)` \times \frac{\frac{`r p23$r.withdrawal`}{12}}
  {1 - (1 + \frac{`r p23$r.withdrawal`}
  {12})^{-(12 \times `r p23$n.withdrawal`)}}\\
  &= \$`r pretty(p23$fv.withdrawal)`
\end{aligned}
$$

---

## 24. Calculating rates of return

Suppose an investment offers to quadruple your money in 12 months (don't believe
it). What rate of return per quarter are you being offered?

---

```{r problem 24; setup, echo=F}
p24 <- new.env()
```

We need to find the rate of return that satisfies the following equation:

$$
\begin{aligned}
\text{FV} &= \text{PV} \times (1 + r/m)^{mT}\\ 
4 &= (1 + r/12)^{12}
\end{aligned}
$$

Here we can use `R`s `uniroot` to find `r`:

```{r problem 24; show uniroot, echo=T}
p24$n <- 12
p24$fv <- 4
f <- function(r) (1 + r/p24$n)^(p24$n) - p24$fv
p24$r <- uniroot(f, upper=10, lower=.01, tol=1e-4)$root
```

The return rate for an investment that quadruples your money in 12 months is: `r percent(p24$r)`.

---


## 25. Calculating rates of return

You're trying to choose between two different investments, both of which have
up-front costs of \$65,000. Investment G returns \$125,000 in six years.
Investment H returns \$185,000 in 10 years. Which of these investments has the
higher return?

---

```{r problem 25; setup, echo=F}
p25 <- new.env()
p25$pv <- 65000
p25$fv <- c(125000, 185000)
p25$n <- c(6, 10)
p25$r <- (p25$fv / p25$pv)^(1/p25$n) - 1
```

We need to find the rate of return for two investments have an up-front cost of
\$65,000. We can use the future value formula to find `r`:

$$
\begin{aligned}
\text{FV} &= \text{PV} \times (1 + r)^n\\
\frac{\text{FV}}{\text{PV}} &= (1 + r)^n\\
\left(\frac{\text{FV}}{\text{PV}}\right)^{1/n} &= 1 + r\\
r &= \left(\frac{\text{FV}}{\text{PV}}\right)^{1/n} - 1
\end{aligned}
$$

### Investment G: return of \$125,000 in six years.

$$
\begin{aligned}
r_{G} &= \left(\frac{`r pretty(p25$fv[1])`}
  {`r pretty(p25$pv)`}\right)^{1/`r p25$n[1]`}- 1\\
  &= `r percent(p25$r[1])`
  \%
\end{aligned}
$$

### Investment H: return of \$185,000 in 10 years.

$$
\begin{aligned}
r_{H} &= \left( \frac{`r pretty(p25$fv[2])`}
  {`r pretty(p25$pv)`} \right)^{1/`r p25$n[2]`} - 1\\
  &= `r percent(p25$r[2])`
  \%
\end{aligned}
$$

We can also use `R`s `uniroot` function to find the returns:

```{r problem25; uniroot for investment 1, echo=T}
p25$f.G <- function(r) p25$pv * (1 + r)^p25$n[1] - p25$fv[1]
percent(p25$r.G <- uniroot(p25$f.G, upper=1, lower=.01, tol=1e-4)$root)
p25$f.H <- function(r) p25$pv * (1 + r)^p25$n[2] - p25$fv[2]
percent(p25$r.H <- uniroot(p25$f.H, upper=1, lower=.01, tol=1e-4)$root)
```

---

## 26. Growing perpetuities

Mark Weinstein has been working on an advanced technology in laser eye surgery.
His technology will be available in the near term. He anticipates his first
annual cash flow from the technology to be \$175,000, received two years from
today. Subsequent annual cash flows will grow at 3.5 percent in perpetuity. What
is the present value of the technology if the discount rate is 10 percent?

---

```{r problem 26; setup, echo=F}
p26 <- new.env()
p26$cf.0 <- 175000
p26$n.start <- 1
p26$growth <- .035
p26$discount <- .10
p26$pv.perpituity <- p26$cf.0 / (p26$discount - p26$growth)
# Discounted back to today.
p26$pv.discounted <- p26$pv.perpituity / (1 + p26$discount)^(p26$n.start)
```

We first calculate the present value at year 1:

$$
\begin{aligned}
\text{PV}_{\text{year 1}} &= \frac{C_0}{r - g}\\
  &= \frac{`r pretty(p26$cf.0)`}{`r p26$discount` - `r p26$growth`}\\
  &= `r pretty(p26$pv.perpituity)`
\end{aligned}
$$

Now we need to discount year 1 back to today:

$$
\begin{aligned}
\text{PV}_{\text{Today}} &= \frac{\text{PV}_{\text{year 1}}}{(1 + r)}\\
  &= \frac{`r pretty(p26$pv.perpituity)`}{(1 + `r p26$discount`)}\\
  &= `r pretty(p26$pv.discounted)`
\end{aligned}
$$

Book: \$2,477,552

---

## 27. Perpetuities

A prestigious investment bank designed a new security that pays a quarterly
dividend of \$4.50 in perpetuity. The first dividend occurs one quarter from
today. What is the price of the security if the stated annual interest rate
is 6.5 percent, compounded quarterly?

---

```{r problem 27; setup, echo=F}
p27 <- new.env()
p27$dividend <- 4.5
p27$m <- 4
p27$rate <- .065
p27$pv.div <- p27$dividend / (p27$rate / p27$m)
```

$$
\begin{aligned}
\text{PV}_{\text{dividend}} &= \frac{C_0}{r}\\\\
 &= \frac{\$`r pretty(p27$dividend)`}{`r p27$rate`/ `r p27$m`}\\
  &= \$`r pretty(p27$pv.div)`
\end{aligned}
$$

---

## 28. Annuity present values

What is the present value of an annuity of \$6,500 per year with the first cash
flow received three years from today and the last one received 25 years from
today? Use a discount rate of 7 percent.

---

```{r problem 28; setup, echo=F}
p28 <- new.env()
p28$cf <- 6500
p28$start <- 2
p28$end <- 25
p28$n.payments <- p28$end - p28$start
p28$discount <- .07
p28$pv.annuity <- p28$cf * (1 - (1 + p28$discount)^(-p28$n.payments)) /
  p28$discount
p28$pv.discounted <- p28$pv.annuity / (1 + p28$discount)^(p28$start)
```

The annuity begins to pay out three years from now and the last payment is 25
years from now. This means that there are $25 - 2 = 23$ payments.

Find the present value of the annuity:

$$
\begin{aligned}
\text{PV}_{\text{annuity}} &= C_0 \cdot
  \left[\frac{1 - (1 + r)^{-T}}{r}\right]\\\\
  &= \$`r pretty(p28$cf)` \cdot
  \left[
    \frac{1 - (1 + `r p28$discount`)^{-`r p28$n.payments`}}{`r p28$discount`}
  \right]\\
  &= \$`r pretty(p28$pv.annuity)`
\end{aligned}
$$

Now date it back two years to date 0

$$
\begin{aligned}
\frac{\$`r pretty(p28$pv.annuity)`}{(1 + `r p28$discount`)^{`r p28$start`}}
  = \$`r pretty(p28$pv.discounted)`
\end{aligned}
$$

---

## 29. Annuity present values

What is the value today of a 15-year annuity that pays \$650 a year? The
annuity's first payment occurs six years from today. The annual interest rate is
11 percent for Years 1 through 5, and 13 percent thereafter.

---

```{r problem 29; setup, echo=F}
p29 <- new.env()
p29$pv.f <- function(cf, r, n) cf * sum((1 + r)^(-n))
p29$cf <- 650
# Discount rate during annuity payments
p29$r.1 <- .13
p29$n.1 <- 1:15
# Discount rate before annuity payments
p29$r.2 <- .11
p29$n.2 <- 5
p29$pv.annuity <- p29$cf * sum((1 + p29$r.1)^(-p29$n.1))
p29$pv.discounted <- p29$pv.annuity / (1 + p29$r.2)^(p29$n.2)
```

To find the value today of a 15 year annuity that pays \$650 per year, beginning
6 years from today, we need to do the following:

### Step 1: Understand the timing

* Payments: \$650 per year for 15 years, starting at year 6 through year 20.

* Interest rates: Years 1-5: 11\%, years 6-20: 13\%

### Step 2: Calculate the present value of the annuity at year 5

$$
\begin{aligned}
\text{PV}_{\text{annuity}} &= C \times
  \left[\frac{1 - (1 + r)^{-T}}{r}\right]\\\\
\text{PV}_{\text{annuity}} &= `r p29$cf` \times
  \left[\frac{1 - (1 + `r p29$r.1`)^{-`r tail(p29$n.1,1)`}}
  {`r p29$r.1`}\right]\\
  &= `r pretty(p29$pv.annuity)`
\end{aligned}
$$

### Step 3: Discount the annuity back to year 0 (today)

$$
\begin{aligned}
\text{PV}_{\text{at year 0}} &= \frac{\text{PV}_{\text{annuity}}}
  {(1 + r)^T}\\\\
  &= \frac{`r pretty(p29$pv.annuity)`}{(1 + `r p29$r.2`)^{`r p29$n.2`}}\\
  &= `r pretty(p29$pv.discounted)`
\end{aligned}
$$

---

## 30. Balloon payments

Audrey Sanborn has just arranged to purchase a \$550,000 vacation home in the
Bahamas with a 20 percent down payment. The mortgage has a 6.1 percent stated
annual interest rate, compounded monthly, and calls for equal monthly payments
over the next 30 years. Her first payment will be due one month from now.
However, the mortgage has an eight-year ballon payment, meaning that the balance
of the loan must be paid off at the end of Year 8. There were no other
transaction costs or finance charges. How much will Audrey's ballon payment be
in eight years?

---

```{r problem 30; setup, echo=F}
p30 <- new.env()
p30$mortgage <- 550000 * .8
p30$apr <- .061
p30$m <- 12
p30$t <- 30
p30$n <- p30$m * p30$t
p30$n.balloon <- p30$n - p30$m * 8
p30$c <- p30$mortgage / ((1 - (1 + p30$apr/p30$m)^(-p30$n)) / (p30$apr/p30$m))
p30$balloon <- p30$c * (1 - (1 + p30$apr/p30$m)^(-p30$n.balloon)) /
  (p30$apr/p30$m)
```

### Step 1: Calculate the monthly payments


$$
\begin{aligned}
\text{PV}_{\text{mortgage}} &= C \times [1 - (1 + r/m)^{-mT}] / (r/m)\\
C &= \frac{\text{PV}_{\text{mortgage}}}{([1 - (1 + r/m)^{-mT}] / (r/m))}\\\\
C &= \frac{`r pretty(p30$mortgage)`}
  {[1 - (1 + `r p30$apr` / `r p30$m`)^{-`r p30$n`}]}\\
  &= `r pretty(p30$c)`
\end{aligned}
$$

### Step 2: Calculate the balloon payment

Now, there ia an easy and hard way to determine the balloon payment. The hard
way is to actually amortize the loan for `r p30$m * 8` months to see what the
balance is at that time. The easy way is to recognize that after `r p30$m * 8`
months, we have a `r p30$n.balloon`-month loan. The payment is
still \$`r pretty(p30$c)` per month, and the interest rate is
still `r percent(p30$apr/p30$m)` per month. The loan balance is thus the present
value of the remaining payments:

$$
\begin{aligned}
\text{Loan balance} & = `r pretty(p30$c)` \times
  \left[\frac{1 - (1 + (1 + `r p30$apr` / `r p30$m`)^{-`r p30$n.balloon`}}
  {`r p30$apr` / `r p30$m`}\right]\\\\
  &= `r pretty(p30$balloon)`
\end{aligned}
$$

Book: \$386,994.11

---

## 31. Calculating interest expense

You receive a credit card application from Shady Banks Savings and Loan offering
an introductory rate of 2.40 percent per year, compounded monthly for the first
six months, increasing thereafter to 18 percent compounded monthly. Assuming you
transfer the \$7,500 balance from your existing credit card and make no
subsequent payments, how much interest will you owe at the end of the first
year?

---

```{r problem 31; setup, echo=F}
p31 <- new.env()
p31$r.intro <- .024
p31$n <- 6
p31$m <- 12
p31$r <- .18
p31$c <- 7500
p31$pv.intro <- 7500 * (1 + p31$r.intro / p31$m)^p31$n
p31$pv.after <- p31$pv.intro * (1 + p31$r / p31$m)^p31$n
p31$interest <- p31$pv.after - p31$c
```

### Step 1: Calculate the present value for the introductory period

$$
\begin{aligned}
\text{PV}_{\text{introductory}} &= `r pretty(p31$c)` \times
  \left(1 + \frac{`r p31$r.intro`}{`r p31$m`} \right)^{`r p31$n`}\\
  &= `r pretty(p31$pv.intro)`
\end{aligned}
$$

### Step 2: Calculate the present value after the introductory period

$$
\begin{aligned}
\text{PV}_{\text{after introductory}} &= `r pretty(p31$pv.intro)` \times
  \left(1 + \frac{`r p31$r`}{`r p31$m`} \right)^{`r p31$n`}\\
  &= `r pretty(p31$pv.after)`
\end{aligned}
$$

### Step 3: Interest is the amount minus the principal

$$
\begin{aligned}
\text{Interest} &= `r pretty(p31$pv.after)` - `r pretty(p31$c)`\\
  &= `r pretty(p31$interest)`
\end{aligned}
$$

---

## 32. Perpetuities

Barrett Pharmaceuticals is considering a drug project that costs \$2.5 million
today and is expected to generate end-of-year annual cash flows of \$227,000,
forever. At what discount rate would Barrett be indifferent between accepting or
rejecting the project?

---

```{r problem 32; setup, echo=F}
p32 <- new.env()
p32$investment <- 2.5e6
p32$cf <- 227000
p32$r <- p32$cf / p32$investment
```

Using the perpetuity equation we can find the rate:

$$
\begin{aligned}
\text{PV}_{\text{Perpetuity}} &= \frac{C}{r}\\\\
`r pretty(p32$investment)`  &= \frac{`r pretty(p32$cf)`}{r}\\\\
r &= \frac{`r pretty(p32$cf)`}{`r pretty(p32$investment)`}\\
  &= `r percent(p32$r)`
  \%
\end{aligned}
$$

---

## 33. Growing Annuity

Southern California Publishing Company is trying to decide whether to revise its
popular textbook, _Financial Psychoanalysis Made Simple_. The company has
estimated that the revision will cost \$75,000. Cash flows from increased sales
will be \$21,000 the first year. These cash flows will increase by 4 percent per
year. The book will go out of print five years from now. Assume that the initial
cost is paid now and revenues are received at the end of each year. If the
company requires a return of 10 percent for such an investment, should it
undertake the revision?

---

```{r problem 33; setup, echo=F}
p33 <- new.env()
p33$investment <- 75000
p33$cf <- 21000
p33$g <- .04
p33$n <- 5
p33$r <- .1
p33$pv <- p33$cf * sum((1 + p33$g)^(0:(p33$n-1)) /
  (1 + p33$r)^(1:p33$n))
p33$npv <- p33$pv - p33$investment
```

We use the growth formula for an annuity. Which can be derived:

$$
\begin{aligned}
\text{PV}_{\text{annuity}} &= \text{CF} \times
  \sum^N_{n=1}{\frac{(1 + g)^{n-1}}{(1 + r)^{n}}}\\
  &= \text{CF} \times \left[\frac{1}{(1 + r)} + \frac{(1 + g)}{(1 + r)^2}
  + \frac{(1 + g)^2}{(1 + r)^3} + \cdots
  + \frac{(1 + g)^{n-1}}{(1 + r)^{n}}\right]\\\\
  &= \text{CF} \times \left[\frac{1 - \left(\frac{1 + g}{1 + r}\right)^N}
     {r - g}\right]\\
  &= `r pretty(p33$cf)` \times
  \left[
    \frac{1 - \left(\frac{1 + `r p33$g`}{1 + `r p33$r`}\right)^{`r p33$n`}}
    {`r p33$r` - `r p33$g`}
  \right]\\
  &= `r pretty(p33$pv)`
\end{aligned}
$$

We will subtract the initial investment from the calculation of the annuity.

$$
\begin{aligned}
\text{NPV} &= \text{PV}_{\text{annuity}} - \text{Initial investment}\\
  &= `r pretty(p33$pv)` - `r pretty(p33$investment)`\\
  &= `r pretty(p33$npv)`
\end{aligned}
$$

\$`r pretty(p33$pv)` is greater than the initial investment of
\$`r pretty(p33$investment)` by \$`r pretty(p33$npv)`. Yes, Southern California
Publishing Company should undertake the revision.

---

## 34. Growing annuity

Your job pays you only once a year for all the work you did over the previous 12
months. Today, December 31, you just received your salary of \$65,000, and you
plan to spend all of it. However, you want to start saving for retirement
beginning next year. You have decided that one year from today you will begin
depositing 5 percent of your annual salary in an account that will earn 10
percent per year. Your salary will increase at 4 percent per year throughout
your career. How much money will you have on the date of your retirement 40
years from today?

---

```{r problem 34; setup, echo=F}
p34 <- new.env()
p34$salary <- 65000
p34$saving <- p34$salary * .05
p34$r <- .1
p34$g <- .04
p34$n <- 40
p34$fv <- p34$saving * ((1 + p34$r)^p34$n - (1 + p34$g)^p34$n) /
  (p34$r - p34$g)
```

__BONUS__: Derive the growing annuity formula

We are calculating the future value of a series of growing deposits into an
account that will earn interest of 10 percent a year.

### Step 1: Derive the future value formula from the annuity formula

We know the present value of a series of growing payments is:

$$
\text{PV} = C_0
  \left[\frac{1 - \left(\frac{(1 + g)}{(1 + r)}\right)^{t}}{r - g}\right]
$$

The future value is simply the present value times the compound interest:

$$
\text{FV} = \text{PV} (1 + r)^t
$$

Therefore, the formula for the future value of our account is:

$$
\begin{aligned}
\text{FV} &= C_0
  \left[\frac{1 - (\frac{(1 + g)^t}{(1 + r)^t}}{r - g}\right]
  (1 + r)^t\\
  &= C_0
  \left[\frac{(1 + r)^t - (1 + g)^t}{r - g}\right]
\end{aligned}
$$

### Step 2: Calculate the future value

The starting amount we are saving (which will grow at 4\%) is:

$$
\text{Saving} = \$`r p34$salary` \times .05 = \$`r p34$saving`
$$

Now lets plug in the numbers:

$$
\begin{aligned}
\text{FV} &= \$`r p34$saving` \left[\frac{(1 + `r p34$r`)^{`r p34$n`}
  - (1 + `r p34$g`)^{`r p34$n`}}{`r p34$r` - `r p34$g`}\right]\\
  &= \$`r pretty(p34$fv, 0)`
\end{aligned}
$$

The value of the account at the time of retirement will be
\$`r pretty(p34$fv, 0)`.

Book: \$2,279,147

Error: `r percent((2279147 - p34$fv) / (2279147))`

---

## 35. Present value and interest rates

What is the relationship between the value of an annuity and the level of
interest rates? Suppose you just bought a 15-year annuity of \$6,800 per year at
the current interest rate of 10 percent per year. What happens to the value of
your investment if interest rates suddenly drop to 5 percent? What if interest
rates suddenly rise to 15 percent?

---

```{r problem 35; setup, echo=F}
p35 <- new.env()
p35$n <- 15
p35$c <- 6800
p35$r <- c(.1, .05, .15)
p35$pv <- p35$c * unlist(lapply( p35$r,
  function(r) sum((1 + r)^(-(1:p35$n)))))
```

An annuity is a series of payments over a set period of time. The effect of
interest rates on the value of those payments depends on its change. If interest
rates increase than the value of those payments decrease. If interest rates
decrease than the value of those payments increase. The following will
demonstrate this.

### The present value of an annuity

An annuity is a series of payments over a set period of time $t$:

$$
\begin{aligned}
\text{PV} &= \frac{C}{(1 + r)} + \frac{C}{(1 + r)^2} + \cdots
  + \frac{C}{(1 + r)^t} = C \sum^T_{t = 1} \frac{1}{(1 + r)^t}
\end{aligned}
$$

As $t$ increases, the term in the denominator $(1 + r)$ becomes more prominent
and begins to diminish the value of the cash flows. The interest rate $r$ also
has an effect. The larger the interest rate $r$, the more prominent the
denominator becomes, and the effects compound at the long end of this equation.

### Demonstration

Lets calculate the present value of a 15-year annuity paying \$6,800 a year at
an interest rate of 10 percent:

$$
\begin{aligned}
\text{PV} &= C \times \frac{1 - (1 + r)^{-T}}{r}\\\\
  &= \$`r pretty(p35$c, 0)` \left[\frac{1 - (1 + `r p35$r[1]`)^{`r p35$T`}}
  {`r p35$r[1]`}\right]\\
  &= \$`r pretty(p35$pv[1], 0)`
\end{aligned}
$$

With a 10 percent interest rate, the present value of the 15-year annuity is 
\$`r pretty(p35$pv[1], 0)`.

If the interst rate were to drop to 5 percent, then the value would be:

$$
\begin{aligned}
\text{PV} &= C \times \frac{1 - (1 + r)^{-T}}{r}\\\\
  &= \$`r pretty(p35$c, 0)` \left[\frac{1 - (1 + `r p35$r[2]`)^{`r p35$T`}}
  {`r p35$r[2]`}\right]\\
  &= \$`r pretty(p35$pv[2], 0)`
\end{aligned}
$$

This is an increase of \$`r pretty(p35$pv[2] - p35$pv[1], 0)`.

If the interest rate were to increase to 15 percent instead, then the value of
the 15-year annuity would be:

$$
\begin{aligned}
\text{PV} &= C \times \frac{1 - (1 + r)^{-T}}{r}\\\\
  &= \$`r pretty(p35$c, 0)` \left[\frac{1 - (1 + `r p35$r[3]`)^{`r p35$T`}}
  {`r p35$r[3]`}\right]\\
  &= \$`r pretty(p35$pv[3], 0)`
\end{aligned}
$$

This is a decrease of \$`r pretty(p35$pv[1] - p35$pv[3])`.

---

## 36. Calculating the number of payments

You're prepared to make monthly payments of \$350, beginning at the end of this
month, into an account that pays 10 percent interest compounded monthly. How
many payments will you have made when your account balance reaches \$35,000?

---

```{r problem 36; setup, echo=F}
p36 <- new.env()
p36$c <- 350
p36$r <- .1 / 12 # months
p36$fv <- 35000
p36$n <- log(p36$r * (p36$fv / p36$c) + 1) / log(1 + p36$r)
```

We need to derive the future value of an annuity to understand how many months
it will take to reach \$35,000.

The present value of an annuity is derived:

$$
\begin{aligned}
\text{PV} &= \frac{C}{(1 + r)} + \frac{C}{(1 + r)^2} + \cdots
  + \frac{C}{(1 + r)^n}\\\\
(1 + r) \cdot \text{PV} &= C + \frac{C}{(1 + r)}
  + \frac{C}{(1 + r)^2} + \cdots + \frac{C}{(1 + r)^{n-1}}\\\\
(1 + r) \cdot \text{PV} - \text{PV} &= C - \frac{C}{(1 + r)^n}\\\\
\text{PV} \cdot ((1 + r) - 1) &= C \cdot \left[1 - \frac{1}{(1 + r)^n}\right]\\\\
\text{PV} &= C \cdot \left[\frac{1 - (1 + r)^{-n}}{r}\right]
\end{aligned}
$$

The future value is the present value times the compounding interest:

$$
\begin{aligned}
\text{FV} &= \text{PV} \cdot (1 + r)^n\\
  &= C \cdot \left[\frac{1 - (1 + r)^{-n}}{r}\right] \cdot (1 + r)^n\\\\
\text{FV} &= C \cdot \left[\frac{(1 + r)^n - 1}{r}\right]
\end{aligned}
$$

Now we need to solve for the number of months $n$. Keep in mind that $r$ is the
APR and needs to be divided by 12, the number of months in a year.

$$
\begin{aligned}
\frac{\text{FV}}{C} &= \frac{(1 + r)^n + 1}{r}\\\\
r \cdot \frac{\text{FV}}{C} - 1 &= (1 + r)^n\\\\
\ln{\left(r \cdot \frac{\text{FV}}{C} - 1\right)} &= n \ln{(1 + r)}\\\\
n &= \frac{\ln{\left(r \cdot \frac{\text{FV}}{C} - 1\right)}}
  {\ln{(1 + r)}}\\\\
  &= \frac{\ln{\left(`r p36$r`
  \cdot \frac{`r pretty(p36$fv)`}{`r pretty(p36$c)`} - 1\right)}}
  {\ln{(1 + `r p36$r`)}}\\
  &= `r round(p36$n)`
\end{aligned}
$$

Using `R`s `uniroot` function we can also determine $n$

```{r problem 36; show uniroot solution, echo=T}
p36$f <- function(n) p36$c * ((1 + p36$r)^n - 1) / p36$r - p36$fv
uniroot(p36$f, upper=1000, lower=10, tol=.01)$root
```

Here `p36$c`, `p36$r`, and `p36$fv` have already been defined as:
`r pretty(p36$c, 0)`, `r p36$r` (.1 / 12), and `r pretty(p36$fv, 0)`
respectively.

---

## 37. Calculating annuity present values

You want to borrow \$65,000 from your local bank to buy a new sailboat. You can
afford to make monthly payments of \$1,320, but no more. Assuming monthly
compounding, what is the highest APR you can afford on a 60-month loan?

---

```{r problem 37; setup, echo=F}
p37 <- new.env()
```

We are borrowing \$65,000 from the bank for a period of 60-months and will be
making payments of \$1,320. That means we find the APR rate $r$ that satisfies:

$$
\text{PV} = C \times \sum^{60}_{n=1}{\frac{1}{(1 + r/m)^{-n}}}
$$

Where $m$ is the number of months in a year (12) and $C$ is the monthly payments.

The best way to solve this is using `R`s `uniroot` function:

```{r problem 37; show uniroot solution, echo=T}
p37$pv <- 65000
p37$c <- 1320
p37$n <- 60
p37$f <- function(r) p37$c * sum((1 + r/12)^(-(1:p37$n))) - p37$pv
percent(p37$apr <- uniroot(p37$f, upper=1, lower=1e-4, tol=1e-8)$root)
percent(p37$ear <- (1 + p37$apr / 12)^(12) - 1)
```

The highest APR is `r percent(p37$apr)`.

---

## 38. Calculating loan payments

You need a 30-year, fixed-rate mortgage to buy a new home for \$250,000. Your
mortgage bank will lend you the money at a 5.3 percent APR for this 360-month
loan. However, you can only afford monthly payments of \$950, so you offer to
pay off any remaining loan balance at the end of the loan in the form of a
single balloon payment. How large will this balloon payment have to be for you
to keep your monthly payment at \$950?

---

```{r problem 38; setup, echo=F}
p38 <- new.env()
p38$pv <- 250000
p38$r <- .053 / 12 # Monthly
p38$t <- 360
p38$c <- 950
p38$fv <- p38$pv * (1 + p38$r)^p38$t
p38$fv.payments <- p38$c * ((1 + p38$r)^p38$t - 1) / p38$r
p38$balloon <- p38$fv - p38$fv.payments
```

We want to find out what the final balloon payment will be if we want to limit
our payments to \$950 a month for the entire term of the loan.

### Step 1: Calculate the future value of the loan

$$
\begin{aligned}
\text{FV} &= C \cdot \left[\frac{(1 + r)^t - 1}{r}\right]\\
  &= \$`r pretty(p38$pv, 0)` \cdot
  \left[\frac{(1 + `r p38$r`)^{`r p38$t`} - 1}{`r p38$r`}\right]\\
  &= \$`r pretty(p38$fv, 0)`
\end{aligned}
$$

### Step 2: Calculate the future value of the loan using \$950 as payments

$$
\begin{aligned}
\text{FV} &= \$`r pretty(p38$c)` \cdot
  \left[\frac{(1 + `r p38$r`)^{`r p38$t`} - 1}{`r p38$r`}\right]\\
  &= `r pretty(p38$fv.payments, 0)`
\end{aligned}
$$

### Step 3: The monthly payments will be paid for 30 years

The amount of the loan remaining:

$$
\begin{aligned}
\text{Balloon} &= \text{FV}_{\text{loan}} - \text{FV}_{\text{payments}}\\
  &= `r pretty(p38$fv)` - `r pretty(p38$fv.payments)`\\
  &= `r pretty(p38$balloon)`
\end{aligned}
$$

Book: \$385,664.73

---

## 39. Present and future values

The present value of the following cash flow stream is \$7,300 when discounted
at 8 percent annually. What is the value of the missing cash flow?

```{r problem 39; table, echo=F}
p39 <- new.env()
p39$year <- c(1,2,3,4)
p39$cash.flow <- c(1500, NA, 2700, 2900)
p39$pv <- 7300
p39$r <- .08
p39$calcs <- p39$cash.flow / (1 + p39$r)^(p39$year)
p39$sum.calcs <- sum(p39$calcs, na.rm=T)
p39$c.2 <- (p39$pv - p39$sum.calcs) * (1 + p39$r)^2
knitr::kable(data.frame(year=p39$year, cash.flow=p39$cash.flow))
p39$check.cash.flow <- c(1500, p39$c.2, 2700, 2900)
p39$check <- sum(p39$check.cash.flow * (1 + p39$r)^(-p39$year))
```

---

We need to find the missing cash flow given the present value and the discount
rate.

We know the structure of the series, therefore we can find $C_2$:

$$
\begin{aligned}
\text{PV} &= \frac{C_{1}}{1 + r} + \frac{C_{2}}{(1 + r)^{2}} +
  \frac{C_{3}}{(1 + r)^3} + \frac{C_{4}}{(1 + r)^4}\\\\
`r pretty(p39$pv, 0)` &=
    \frac{`r pretty(p39$cash.flow[1], 0)`}{(1 + `r p39$r`)^{`r p39$year[1]`}}
  + \frac{`r pretty(p39$cash.flow[2], 0)`}{(1 + `r p39$r`)^{`r p39$year[2]`}}
  + \frac{`r pretty(p39$cash.flow[3], 0)`}{(1 + `r p39$r`)^{`r p39$year[3]`}}
  + \frac{`r pretty(p39$cash.flow[4], 0)`}{(1 + `r p39$r`)^{`r p39$year[4]`}}\\\\
`r pretty(p39$pv, 0)` &=
    `r pretty(p39$calcs[1], 0)`
  + `r pretty(p39$calcs[2], 0)`
  + `r pretty(p39$calcs[3], 0)`
  + `r pretty(p39$calcs[4], 0)`\\
C_2 &= `r pretty(p39$c.2)`
\end{aligned}
$$

---

## 40. Calculating present values

You just won the TVM Lottery. You will receive \$1 million today plus another 10
annual payments that increase by \$275,000 per year. Thus, in one year you
receive \$1.275 million. In two years, you get \$1.55 million, and so on. If
the appropriate interest rate is 9 percent, what is the present value of your
winnings?

---

```{r problem 40; setup, echo=F}
p40 <- new.env()
p40$c.0 <- 1e6
p40$inc <- 275000
p40$r <- .09
p40$t <- 0:10
p40$g <- p40$inc / p40$c.0
p40$pv <- sum((p40$c.0 + p40$inc*p40$t) / (1 + p40$r)^(p40$t))
```

We need to calculate the value of a growing annuity. However the growth rate
of `r percent(p40$g)` is far larger than the interest rate of 9\%. This means we
cannot use the annuity formula and must calculate the 10 annual payments
manually.

The annuity is as follows:

$$
\begin{aligned}
\text{PV} &= 1 + \frac{1.275}{(1 + .09)} + \frac{1.55}{(1 + .09)^2} + \cdots
  + \frac{1 + .275 \cdot n}{(1 + .09)^n}\\\\
  &= \sum_{n=0}^{10}{\frac{1 + .275 \cdot n}{(1 + .09)^n}}\\
  &= `r pretty(p40$pv, 0)`
\end{aligned}
$$

The present value of the winnings is \$`r pretty(p40$pv, 0)`.

---

## 41. EAR versus APR

You have just purchased a new warehouse. To finance the purchase, you've
arranged for a 30-year mortgage for 80 percent of the \$4,500,000 purchase price.
The monthly payment on this loan will be \$27,500. What is the APR on this loan?
The EAR?

---

```{r problem 41; setup, echo=F}
p41 <- new.env()
p41$t <- 30 * 12
p41$pv <- 4500000 * .8
p41$c <- 27500
```

The amount of the loan is 80 percent of the warehouse purchase price:

$$
\text{Loan} = \$4,500,000 \times .80 = \$`r pretty(p41$pv)`\\
$$

The present value of the loan is equal to the sum of all the monthly payments:

$$
\begin{aligned}
\text{PV} &= \sum^{T}_{t = 1}{\frac{C}{(1 + r)^t}}\\
`r pretty(p41$pv, 0)`  &= \sum^{`r p41$t`}_{t = 1}
  {\frac{`r pretty(p41$c, 0)`}{(1 + r)^t}}\\
0 &= \sum^{`r p41$t`}_{t = 1}{\frac{`r pretty(p41$c, 0)`}{(1 + r)^t}}
  - `r pretty(p41$pv, 0)`
\end{aligned}
$$

We can use `R`s `uniroot` function to find the EAR, and multiply it by 12 months
to find the APR:

```{r problem 41; show uniroot, echo=T}
p41$t <- 30 * 12
p41$pv <- 4500000 * .8
p41$c <- 27500
p41$f <- function(r) p41$c * sum((1 + r/12)^(-(1:p41$t))) - p41$pv
percent(p41$apr <- uniroot(p41$f, lower=1e-5, upper=10, tol=1e-8)$root)
percent(p41$ear <- (1 + p41$apr/12)^12 - 1)
```

The APR is $r \times 12 = 8.429$%. The EAR is:

$$
\begin{aligned}
\mathbf{EAR} &= (1 + r)^{12} - 1 \\
 &= 8.762\%
\end{aligned}
$$

---

## 42. Present value and break-even interest

Consider a firm with a contract to sell an asset for \$115,000 three years from
now. The asset costs \$76,000 to produce today. Given a relevant discount rate
on this asset of 13 percent per year, will the firm make a profit on this asset?
At what rate does the firm just break even?

---

```{r problem 42; setup, echo=F}
p42 <- new.env()
p42$pv <- 76000
p42$fv <- 115000
p42$r <- .13
p42$t <- 3
p42$npv <- p42$fv / (1 + p42$r)^(p42$t) - p42$pv
p42$f <- function(r) p42$fv / (1 + r)^(p42$t) - p42$pv
p42$r.break.even <- uniroot(p42$f, upper=1, lower=1e-5, tol=1e-8)$root
```

### Profit

To find if the firm will make a profit, calculate the present value of \$115,000
to find if it is greater than \$76,000.

$$
\begin{aligned}
\text{PV} &= \frac{\$`r pretty(p42$fv)`}{(1 + `r p42$r`)^{`r p42$t`}}\\\\
 &= \$`r pretty(p42$npv, 2)`
\end{aligned}
$$

So the firm will make a profit of \$`r pretty(p42$npv)`.

### Break even

The break even interest rate is the rate where:

$$
\begin{aligned}
\$`r pretty(p42$pv)` &= \frac{\$`r pretty(p42$fv)`}{(1 + r)^{`r p42$t`}}
\end{aligned}
$$

Using `R`s `uniroot` function we find $r$ as `r percent(p42$r.break.even)`.

Book: \$3,700.77; 14.81\%

---

## 43. Present value and multiple cash flows

What is the present value of \$5,000 per year, at a discount rate of 6 percent,
if the first payment is received 6 years from now and the last payment is
received 25 years from now?

---

```{r problem 43; setup, echo=F}
p43 <- new.env()
p43$c <- 5000
p43$r <- .06
p43$start <- 6
p43$end <- 25
p43$t <- p43$end - p43$start + 1
p43$pv.1 <- p43$c * (1 - (1 + p43$r)^(-p43$t)) / p43$r
p43$pv <- p43$c * sum((1 + p43$r)^(-(p43$start:p43$end)))
```

First calculate the present value of the payments for $5 - 25 = 20$ years.


$$
\begin{aligned}
\text{PV}_{\text{6 years from now}} &=
  \sum_{n=0}^{`r p43$t`}{\frac{`r pretty(p43$c)`}{(1 + `r p43$r`)^{n}}}\\\\
  &= `r pretty(p43$c, 0)` \cdot \frac{1 - (1 + `r p43$r`)^{-20}}{`r p43$r`}\\\\
  &= `r pretty(p43$pv.1)`
\end{aligned}
$$

Now we project this back six years to find its present value today:

$$
\begin{aligned}
\mathbf{PV}_{\text{today}} &= \frac{`r pretty(p43$pv.1)`}
  {(1 + `r p43$r`)^{`r p43$start - 1`}}\\\\
  &= `r pretty(p43$pv)`
\end{aligned}
$$

---

## 44. Variable interest rates

A 15-year annuity pays \$1,500 per month, and payments are made at the end of
each month. If the interest rate is 12 percent compounded monthly for the first
seven years, and 6 percent compounded monthly thereafter, what is the present
value of the annuity?

---

```{r problem 44; setup, echo=F}
p44 <- new.env()
p44$t <- c(7, 15 - 7) * 12
p44$c <- 1500
p44$r <- c(.12, .06) / 12
p44$pv.f <- function(c, r, t) c * sum((1 + r)^(-(1:t)))
p44$pv.1 <- p44$pv.f(p44$c, p44$r[1], p44$t[1])
p44$pv.2 <- p44$pv.f(p44$c, p44$r[2], p44$t[2])
p44$pv.2.discounted <- p44$pv.2 * (1 + p44$r[1])^(-p44$t[1])
p44$pv <- p44$pv.1 + p44$pv.2.discounted
```

---

The given information can be restated as:

$$
\begin{aligned}
r_1 &= \frac{.12}{12} &= .01\\\\
r_2 &= \frac{.06}{12} &= .005\\\\
T_1 &= 12 \times 7 &= 84\\\\
T_2 &= 12 \times 8 &= 96\\\\
C_0 &= 1,500
\end{aligned}
$$

We need to solve for an annuity that pays the same but the interest rate changes
in the middile of the annuity cash flows.

$$
\begin{aligned}
\mathbf{PV}_{\text{annuity}} &= C_0 \times \frac{1 - (1 + r)^{-T}}{r}
\end{aligned}
$$

### Step 1: Calculate the two components of the annuity

$$
\begin{aligned}
\text{PV}_{1} &=
  `r pretty(p44$c)` \cdot \frac{1 - (1 + `r p44$r[1]`)^{-`r p44$t[1]`}}
  {`r p44$r[1]`}\\
  &= `r pretty(p44$pv.1)`\\\\
\text{PV}_{2} &=
  `r pretty(p44$c)` \cdot \frac{1 - (1 + `r p44$r[2]`)^{-`r p44$t[2]`}}
  {`r p44$r[2]`}\\
  &= `r pretty(p44$pv.2)`
\end{aligned}
$$

### Step 2: Project the second annuity back to the present day

We need to use the current interest rate (not the future rate that is used in
calculating the second component of the annuity) when projecting the value back
to the present day.

$$
\begin{aligned}
\text{PV}_{\text{today}} &= \frac{`r pretty(p44$pv.2)`}
  {(1 + `r p44$r[1]`)^{`r p44$t[1]`}}\\\\
  &= `r pretty(p44$pv.2.discounted)`
\end{aligned}
$$

The present value of the annuity is the sum of the two annuity components.

$$
  `r pretty(p44$pv.1)` + `r pretty(p44$pv.2.discounted)` = `r pretty(p44$pv)`
$$


Book: \$134,455.36

---

## 45. Comparing cash flow streams

You have your choice of two investment accounts. Investment A is a 15-year
annuity that features end-of-month \$1,500 payments and has an interest rate
of 8.7 percent compounded monthly. Investment B is an 8 percent continuously
compounded lump-sum investment, also good for 15 years. How much money would you
need to invest in B today for it to be worth as much as Investment A 15 years
from now.

---

```{r problem 45; setup, echo=F}
p45 <- new.env()
p45$t <- 15
p45$m <- 12
p45$n.1 <- p45$t * p45$m
p45$r <- c(.087, .08)
p45$rm <- p45$r / p45$m
p45$c.A <- 1500
# We calculate from 0 to n-1 power series
p45$fv.A <- sum(p45$c.A * (1 + p45$rm[1])^((0:(p45$n.1-1))))
p45$c.B <- p45$fv.A / exp(p45$r[2] * p45$t)
```

### Step 1: Calculate the present value of investment A

$$
\begin{aligned}
\text{FV}_{\text{A}} &= \$`r pretty(p45$c.A, 0)` \times
  \frac{1 - (1 + \frac{`r p45$r[1]`}{12})^{`r p45$n.1`}}{\frac{`r p45$r[1]`}{12}}\\\\
  &= \$`r pretty(p45$fv.A)`
\end{aligned}
$$

### Step 2: Find the initial investment for B

$$
\begin{aligned}
\text{FV}_{\text{B}} &= C_B \times e^{rt}\\\\
\$`r pretty(p45$fv.A, 0)` &= C_B e^{`r p45$r[2]` \times `r p45$t`}\\\\
C_B &= \frac{`r pretty(p45$fv.A, 0)`}{e^{`r p45$r[2] * p45$t`}}\\
  &= \$`r pretty(p45$c.B, 0)`
\end{aligned}
$$

---

## 46. Calculating present value of a perpetuity

Given an interest rate of 6.1 percent per year, what is the value at Date
$t = 7$ of a perpetual stream of \$2,500 annual payments that begins at Date
$t = 15$?

---

```{r problem 46; setup, echo=F}
p46 <- new.env()
p46$c <- 2500
p46$n <- c(7, 15)
p46$t <- diff(p46$n)
p46$r <- .061
p46$pv.perpetuity <- p46$c / p46$r
p46$pv.at.7 <- p46$pv.perpetuity / (1 + p46$r)^(p46$t-1)
```

So we can pretend that $t = 7$ is the present. Therefore, We find the present
value of a perpetuity that begins $15 - 7 = 8$ years from now.

$$
\begin{aligned}
\text{PV} &= \frac{C}{r}\\\\
\text{PV} &= \frac{\$`r pretty(p46$c, 0)`}{`r p46$r`}\\
  &= \$`r pretty(p46$pv.perpetuity, 0)`
\end{aligned}
$$

We now project back 8 years:

$$
\begin{aligned}
\text{PV}_{\text{8 years previous}} &= \frac{\$`r pretty(p46$pv.perpetuity)`}
  {(1 + `r p46$r`)^{`r p46$t-1`}}\\
  &= \$`r pretty(p46$pv.at.7, 0)`
\end{aligned}
$$

---

## 47. Calculating EAR

A local finance company quotes a 16 percent interest rate on one-year loans. So,
if you borrow \$26,000, the interest for the year will be \$4,160. Because you
must repay a total of \$30,160 in one year, the finance company requires you to
pay \$30,160/12, or \$2,513.33, per month over the next 12 months. Is this a 15
percent loan? What rate would legally have to be quoted? What is the effective
annual rate?

---

```{r problem 47; setup, echo=F}
p47 <- new.env()
p47$pmt <- 2513.33
p47$PV <- 26000
p47$n <- 12
```

The loan involves equal monthly payments of \$2,513.33 for 12 months.

$$
\begin{aligned}
\text{P} &= C \times \frac{1 - (1 + r)^{-n}}{n}
\end{aligned}
$$

$$
\begin{aligned}
\text{PMT} &= 2,513.33 \\
\text{P} &= 26,000 \\
\text{n} &= 12
\end{aligned}
$$

Using R we can solve for `r`:

```{r problem 47; uniroot solution, echo=T}
f <- function(r) p47$pmt * (1 - (1 + r)^(-p47$n)) / r - p47$PV
(p47$r <- uniroot(f, lower=.001, upper=1, tol=1e-4)$root)
(p47$apr <- p47$r * 12)
(p47$ear <- (1 + p47$apr / p47$n) ^ (p47$n) - 1)
```

The monthly interest is `r percent(p47$r)`.

The APR is: `r percent(p47$r)` $\times 12 =$ `r percent(p47$apr)`.

The EAR is:
  $(1 + \frac{`r round(p47$apr, 2)`}{`r p47$n`})^{`r p47$n`} - 1 =$
  `r percent(p47$ear)`

Thus, this is not a 15% loan it is significantly higher. The finance company
would have to legally quote the __APR__ as `r percent(p47$apr)`, reflecting the cost based on
monthly compounding.

The __EAR__ of `r percent(p47$ear)` accounts for the full compounding effect over the year.

---

## 48. Calculating present values

A 5-year annuity of ten \$5,300 semiannual payments will begin 9 years from now,
with the first payment coming 9.5 years from now. If the discount rate is 12
percent compounded monthly, what is the value of this annuity five years from
now? What is the value three years from now? What is the current value of the
annuity?

---

```{r problem 48; setup, echo=F}
p48 <- new.env()
p48$c <- 5300
p48$t <- 5
p48$n <- 2
p48$tn <- 5 * 2
p48$r <- .12
p48$m <- 12
p48$rm <- p48$r / p48$m
p48$ear <- (1 + p48$rm)^6 - 1
p48$n.discount <- (9 - c(5, 3, 0))*p48$m
p48$pv.9y <- p48$c * sum((1 + p48$ear)^(-(1:p48$tn)))
p48$pv.discount <- p48$pv.9y / (1 + p48$rm)^(p48$n.discount)
```

This is a time value of money problem that involves a deferred annuity with
semiannual payments and monthly compounding.

### Step 1: Given data

* Payment: \$`r pretty(p48$c, 0)`

* Number of payments: `r p48$tn` (over `r p48$t` years, semiannual payments)

* First payment is 9.5 years from now.

* Discount rate is `r percent(p48$r)` compounded monthly. The monthly rate is
  `r p48$r` / `r p48$m` = `r p48$rm` per month.

### Step 2: Compute the PV of the Annuity at its starting point

$$
\begin{aligned}
\text{r} &= (1 + `r p48$r`)^6 -1 = `r p48$ear`\\
\text{PV}_9 &=
  `r pretty(p48$c, 0)` \times \frac{1 - (1 + `r round(p48$ear, 3)`)^{-`r p48$tn`}}
  {`r round(p48$ear, 3)`}\\
  &= `r pretty(p48$pv.9y)`\\\\
\end{aligned}
$$

### Step 3: Discount the annuity from 9 to 5, 3, and to the present:

$$
\begin{aligned}
\text{PV}_5 &= \frac{`r pretty(p48$pv.9y, 0)`}
  {(1 + `r p48$rm`)^{`r p48$n.discount[1]`}}\\
  &= `r pretty(p48$pv.discount[1], 0)`\\\\
\text{PV}_3 &= \frac{`r pretty(p48$pv.9y, 0)`}
  {(1 + `r p48$rm`)^{`r p48$n.discount[2]`}}\\
  &= `r pretty(p48$pv.discount[2], 0)`\\\\
\text{PV}_0 &= \frac{`r pretty(p48$pv.9y, 0)`}
  {(1 + `r p48$rm`)^{`r p48$n.discount[3]`}}\\
  &= `r pretty(p48$pv.discount[3], 0)`
\end{aligned}
$$

---

## 49. Calculating annuities due

Suppose you are going to receive \$20,000 per year for five years. The
appropriate interest rate is 7 percent.

a. What is the present value of the payments if they are in the form of an
ordinary annuity? What is the present value if the payments are an annuity due?

b. Suppose you plan to invest the payments for five years. What is the future
value if the payments are an ordinary annuity? What if the payments are an
annuity due?

c. Which has the highest present value, the ordinary annuity or annuity due?
Which has the highest future value? Will this always be true?

---

```{r problem 49; setup, echo=F}
p49 <- new.env()
p49$c <- 20000
p49$t <- 5
p49$r <- .07
p49$pv.a <- p49$c * sum((1 + p49$r)^(-(1:p49$t)))
p49$pv.due.a <- p49$c * sum((1 + p49$r)^(-(0:p49$t)))
p49$fv.b <- p49$c * sum((1 + p49$r)^(1:p49$t-1))
p49$fv.due.b <- p49$c * sum((1 + p49$r)^(1:p49$t))
```

### A. Present value

#### Annuity in arrears

An ordinary annuity assumes that the first annuity payment begins a full period
hence.

$$
\begin{aligned}
\text{PV}_{\text{ordinary}}
  &= C \times \left(\frac{1 - (1 + r)^{-T}}{r}\right)\\\\
  &= \$`r pretty(p49$c, 0)` \times \left(\frac{1 - (1 + `r p49$r`)^{-`r p49$t`}}
      {`r p49$r`}\right)\\
  &= \$`r pretty(p49$pv.a, 0)`
\end{aligned}
$$

#### Annuity due

An annuity due starts at date 0. Thus:

$$
\begin{aligned}
\text{PV}_{\text{due}} &= C + C \left(\frac{1 - (1 + r)^{-T}}
  {r}\right)\\\\
  &= \$`r pretty(p49$c, 0)` + \$`r pretty(p49$c, 0)`
  \times \left(\frac{1 - (1 + `r p49$r`)^{-`r p49$t`}}{`r p49$r`}\right)\\
  &= \$`r pretty(p49$pv.due.a)`
\end{aligned}
$$

### B. Future value

#### Ordinary annuity

$$
\begin{aligned}
\text{FV}_{\text{ordinary}} &= C \times \left(\frac{(1 + r)^T - 1}{r}\right)\\\\
  &= \$`r pretty(p49$c, 0)` \left(\frac{(1 + `r p49$r`)^{-`r p49$t`} - 1}
  {`r p49$r`}\right)\\
  &= \$`r pretty(p49$fv.b, 0)`
\end{aligned}
$$

#### Annuity due

$$
\begin{aligned}
\text{FV}_{\text{due}} &= C \times \left(\frac{(1 + r)^T - 1}{r}\right)
  \times (1 + r)\\\\
  &= \$`r pretty(p49$c, 0)` \times \left(\frac{(1 + `r p49$r`)^{`r p49$t`} - 1}
  {`r p49$r`}\right) \times (1 + `r p49$r`)\\
  &= \$`r pretty(p49$fv.due.b, 0)`
\end{aligned}
$$

### C.

For both the __present value__ and the __future value__ have a higher value because of the earlier payment. In the future value it has an additional period to earn interest.

---

## 50. Calculating annuities due

You want to buy a new sports car from Muscle Motors for \$73,000. The contract
is in the form of a 60-month annuity due at a 6.45 percent APR. What will your
monthly payment be?

---

```{r problem 50; setup, echo=F}
p50 <- new.env()
p50$pv <- 73000
p50$t <- 60
p50$apr <- .0645
p50$m <- 12
p50$r <- p50$apr / p50$m
p50$c <- p50$pv * (p50$r) / (1 - (1 + p50$r)^(-p50$t)) * (1 + p50$r)
```

### Given

* Price of the car (Present value): \$73,000

* Loan terms: 60 months

* APR = `r percent(p50$apr)` comounded monthly

* Payments are made at the beginning of each month $\rightarrow$ Annuity due.

### Calculate the monthly rate:

$$
\begin{aligned}
\text{r} &= \frac{\text{APR}}{m}\\
  &= \frac{`r p50$apr`}{`r p50$m`}\\
  &= `r p50$r`\\\\
\end{aligned}
$$

### Calculate the annuity due

$$
\begin{aligned}
\text{C} &= \$`r pretty(p50$pv, 0)` \times \frac{`r p50$r`}
  {1 - (1 + `r p50$r`)^{-`r p50$t`}} \times (1 + `r p50$r`)\\
  &= \$`r pretty(p50$c, 0)`
\end{aligned}
$$

---

## 51. Calculating annuities due

You want to lease a set of golf clubs from Pings Ltd. The lease contract is in
the form of 24 equal monthly payments at a 10.4 percent stated annual interest
rate, compounded monthly. Because the clubs cost \$2,300 retail, Pings wants the
PV of the lease payments to equal \$2,300. Suppose that your first payment is
due immediately. What will your monthly lease payments be?

---

```{r problem 51; setup, echo=F}
p51 <- new.env()
p51$pv <- 2300
p51$apr <- .104
p51$t <- 24
p51$m <- 12
p51$r <- p51$apr / p51$m
p51$c <- p51$pv * (p51$r / (p51$r + 1 - (1 + p51$r)^(-p51$t)))
p51$r.display <- round(p51$r, 5)
```

### Step 1: Calculate `r`

$$
r = \frac{\text{APR}}{m} = \frac{`r p51$apr`}{`r p51$m`} = `r p51$r.display`
$$

### Step 2: Calculate the annuity due

$$
\begin{aligned}
\text{PV} &= C  + C \times \left(\frac{1 - (1 + r)^{-T}}{r}\right)\\\\
\text{PV} &= C \times \left[1 + \left(
  \frac{1 - (1 + r)^{-T}}{r}\right)\right]\\\\
C &= \text{PV} \times \frac{1}{1 + \left(\frac{1 - (1 + r)^{-T}}{r}\right)}\\\\
C &= \text{PV} \times \frac{r}{r + 1 - (1 + r)^{-T}}\\
  &= \$`r pretty(p51$pv, 0)` \times \frac{`r p51$r.display`}
  {`r p51$r.display` + 1 - (1 + `r p51$r.display`)^(-`r p51$t`)}\\
  &= \$`r pretty(p51$c, 2)`
\end{aligned}
$$

---

## 52. Annuities

You are saving for the college education of your two children. They are two
years apart in age; one will begin college 15 years from today and the other
will begin 17 years from today. You estimate your children's college expenses to
be \$45,000 per year per child, payable at the beginning of each school year.
The annual interest rate is 7.5 percent. How much money must you deposit in an
account each year to fund your children's education? Your deposits begin one
year from today. You will make your last deposit when your oldest child enters
college. Assume four years of college.

---

```{r problem 52 annuities; setup, echo=F}
p52 <- new.env()

# Yearly college tuition.
p52$college.tuition <- 45000

# annual interest rate.
p52$r <- .075

# number of years in college.
p52$n.college <- 4

# child one starts college in 15 years.
p52$n.1 <- 15

# child two starts college in 17 years.
p52$n.2 <- 17

# The annuity factor.
p52$pvia <- function(r, t) sum((1 + r)^(-(1:t)))

# We begin payments as soon as the child begins college. This is an
# annuity due, therefore we calculate the annuity factor with one less year.
p52$pv.annuity.due <- p52$college.tuition *
    (1 + p52$pvia(p52$r, (p52$n.college - 1)))

# Present value of the payments today for child one.
p52$pv.1 <- p52$pv.annuity.due / (1 + p52$r)^(p52$n.1)

# Present value of the payments today for child two.
p52$pv.2 <- p52$pv.annuity.due / (1 + p52$r)^(p52$n.2)

# The total price to send both children to college.
p52$pv.total <- p52$pv.1 + p52$pv.2

# Caluclate the payments via `uniroot`.
p52$pv <- function(a) a * p52$pvia(p52$r, p52$n.1) - p52$pv.total
p52$yearly.deposits <- uniroot(p52$pv, upper=40000, lower=100, tol=10)$root
```

We are asked to compute equal annual deposits to fund two children's college
expenses, considering:

* Each child will need \$45,000 per year, for 4 years of college.

* Payments are made at the beginning of each college year (so these are annuity
  due payments).

* The first child begins in 15 years, the second child in 17 years.

* Annual interest rate is 7.5\%.

* You will make equal deposits each year starting one year from today, and the
  last deposit is when the first child starts college (year 15).

### Step 1: Determine present value (PV) of college expenses

We will compute the present value at time 0 of all the future college payments,
tehn solve for the annual deposit required to reach that amount.

For each child, the annuity due calculation will be the same. For the annuity
due payments, the first payment is due at the beginning of the first year. Then
we discound the following three years.

$$
\begin{aligned}
\text{PV Annuity Due} &= \$`r pretty(p52$college.tuition)` \times
    \left( 1 + \left[\frac{1 - (1 + `r p52$r`)^{-`r (p52$n.college - 1)`}}
    {`r p52$r`} \right] \right)\\
    &= \$`r pretty(p52$pv.annuity.due)`
\end{aligned}
$$

#### Child 1 starts in year 15

$$
\begin{aligned}
\text{PV}_{\text{Child 1}} &= \frac{\$`r pretty(p52$pv.annuity.due)`}
    {(1 + `r p52$r`)^{`r p52$n.1`}}\\
    &= \$`r pretty(p52$pv.1)`
\end{aligned}
$$

#### Child 2 starts in year 17

$$
\begin{aligned}
\text{PV}_{\text{Child 2}} &= \frac{\$`r pretty(p52$pv.annuity.due)`}
    {(1 + `r p52$r`)^{`r p52$n.2`}}\\
    &= \$`r pretty(p52$pv.2)`
\end{aligned}
$$

#### Total present value of sending both children to college

$$
\begin{aligned}
\text{PV}_{\text{College}} &= \$`r pretty(p52$pv.1)` + \$`r pretty(p52$pv.2)`\\
    \$`r pretty(p52$pv.total)`
\end{aligned}
$$

#### The payments, starting next year, will be

$$
\begin{aligned}
\text{PV} &= C \times \left[\frac{1 - (1 + r)^{-n}}{r}\right]\\\\
C &= \frac{\text{PV}}{\left[\frac{1 - (1 + r)^{-n}}{r}\right]}\\\\
  &= \frac{\$`r pretty(p52$pv.total)`}
  {\left[\frac{1 - (1 + `r p52$r`)^{-`r p52$n.1`}}{`r p52$r`}\right]}\\
  &= \$`r pretty(p52$yearly.deposits)`
\end{aligned}
$$

We will need to make payments of \$`r pretty(p52$yearly.deposits)` each year for
15 years to save enough to send the two children to college.

Book: \$11,571.48

---

## 53. Growing Annuities

Tom Adams has received a job offer from a large investment bank as a clerk to an
associate banker. His base salary will be \$55,000. He will receive his first
annual salary payment one year from the day he begins to work. In addition, he
will get an immediate \$10,000 bonus for joining the company. His salary will
grow at 3.5 percent each year. Each year he will receive a bonus equal to 10
percent of his salary. Mr. Adams is expected to work for 25 years. What is the
present value of the offer if the discount rate is 9 percent?

---

```{r problem 53 growing annuities; setup, echo=F}
p53 <- new.env()
p53$base.salary <- 55000
p53$immediate.bonus <- 10000
p53$bonus.modified.salary <- 55000*1.1
p53$g <- .035
p53$r <- .09
p53$n <- 25
p53$pv <- p53$immediate.bonus + p53$bonus.modified.salary / (1 + p53$r) *
    (1 + sum(((1 + p53$g) / (1 + p53$r))^(1:(p53$n - 1))))
```

### Given:

* The base salary: \$`r pretty(p53$base.salary)`

* Annual 10\% bonus a year, modifies the base salary:
$\$`r pretty(p53$base.salary, 0)` \times 1.10 =
\$`r pretty(p53$bonus.modified.salary, 0)`$. We will use this value to calculate
the growing annuity.

* Immediate \$`r pretty(p53$immediate.bonus, 0)` bonus for joining the company.
  This bonus is added to the present value, unmodified, because it is immediate.

$$
\begin{aligned}
\text{PV}_{\text{Offer}} &= `r pretty(p53$immediate.bonus, 0)` +
    `r pretty(p53$bonus.modified.salary)` \times \left[
    \frac{1 - \left(\frac{(1 + `r p53$g`)}
    {(1 + `r p53$r`)}\right)^{`r p53$n`}}{`r p53$r` - `r p53$g`}\right]\\
    &= `r pretty(p53$pv, 0)`
\end{aligned}
$$

---

## 54. Calculating Annuities

You have recently won the super jackpot in the Washington State Lottery. On
reading the fine print, you discover that you have the following two options:

a. You will receive 31 annual payments of \$250,000, with the first payment
being delivered today. The income will be taxed at a rate of 28 percent. Taxes
will be withheld when the checks are issued.

b. You will receive \$530,000 now, and you will not have to pay taxes on this
amount. In addition, beginning one year from today, you receive \$200,000 each
year for 30 years. The cash flows from this annuity will be taxed at 28 percent.

Using a discount rate of 7 percent, which option should you select?

---

We are comparing two options. The first option is a 31 year annuity due of
\$250,000. Each payment will have taxes taken out at a rate of 28 percent per
year.

The second option is a tax free upfront payment of \$530,000, with taxed
payments of \$200,000 a year as an ordinary annuity. This annuity will be taxed
at 28 percent.

```{r problem 54 calculating annuities; setup, echo=F}
p54 <- new.env()
p54$c.1 <- 250000
p54$c.2 <- 200000
p54$initial.2 <- 530000
p54$tax.rate <- .28
p54$r <- .07
p54$n.1 <- 31
p54$n.2 <- 30
p54$pv.1 <- p54$c.1 * (1 - p54$tax.rate) *
    (1 + sum((1 + p54$r)^(-(1:(p54$n.1-1)))))
p54$pv.2 <- p54$initial.2 + p54$c.2 * (1 - p54$tax.rate) *
    (sum((1 + p54$r)^(-(1:p54$n.2))))
```

### The first option (A pure annuity due)

$$
\begin{aligned}
\text{PV}_{\text{option 1}} &= \$`r pretty(p54$c.1, 0)`
    \times (1 - `r p54$tax.rate`) \times
    \left[1 + \frac{1 - (1 + `r p54$r`)^{-30}}{`r p54$r`}\right]\\
    &= \$`r pretty(p54$pv.1, 0)`
\end{aligned}
$$

### The second option (Immediate bonus with an ordinary annuity)

$$
\begin{aligned}
\text{PV}_{\text{option 2}} &= \$`r pretty(p54$initial.2, 0)` +
    \$`r pretty(p54$c.2, 0)` \times (1 - `r p54$tax.rate`)
    \times \left[\frac{1 - (1 + `r p54$r`)^{-30}}{`r p54$r`}\right]\\
    &= \$`r pretty(p54$pv.2, 0)`
\end{aligned}
$$

### Conclusion

Clearly the first option is the better option.

---

## 55. Calculating growing annuities

You have 30 years left until retirement and want to retire with \$2 million.
Your salary is paid annually, and you will receive \$70,000 at the end of the
current year. Your salary will increase at 3 percent per year, and you can earn
a return of 9 percent on the money you invest. If you save a constant percentage
of your salary, what percentage of your salary must you save each year?

---

```{r problem 55 calculating growing annuities; setup, echo=F}
p55 <- new.env()
p55$salary <- 70000
p55$r <- .09
p55$g <- .03
p55$n <- 30
p55$fv <- 2e6
p55$pvia <- function(r, g, n) (1 - ((1 + g)/(1 + r))^n) / (r - g)
p55$pvia.2 <- function(r, g, n) sum((1 + g)^(0:(n-1)) / (1 + r)^(1:n))

p55$fvga <- function(r, g, n) ((1 + r)^n - (1 + g)^n) / (r - g)
p55$fvga.2 <- function(r, g, n) sum((1 + g)^(0:(n-1)) * (1 + r)^((n-1):0))
p55$f <- function(p) p55$salary * p * p55$fvga(p55$r, p55$g, p55$n) - p55$fv
uniroot(p55$f, upper=5, lower=.01, tol=1e-2)$root
p55$f <- function(p) p55$salary * p * p55$fvga.2(p55$r, p55$g, p55$n) - p55$fv
uniroot(p55$f, upper=5, lower=.01, tol=1e-2)$root
```

We need to calculate the percentage saved to reach a future value of \$2
million.

We use the future value of a growing annuity formula.

The series formula for a growing annuity can be derived from the following
sequence:

### Deriving the formula for the future value of an annuity

The cash flow sequence:

$$
C, C(1 + g), C(1 + g)^{2}, \ldots, C(1 + g)^{n-1}
$$

### Future value formula

$$
\text{FV} = C(1 + g)^{0}(1 + r)^{n-1} + C(1 + g)^{1}(1 + r)^{n-2}
    + \cdots + C(1 + g)^{n-1}(1 + r)^0
$$

$$
\text{FV} = \sum_{k=0}^{n-1}{C(1 + g)^k (1 + r)^{n-1-k}}
$$

Factor out $C \cdot (1 + r)^{n-1}$:

$$
\text{FV} = C(1 + r)^{n-1} \sum^{n-1}_{k=0}{\left(\frac{1 + g}{1 + r}\right)^k}
$$


This is a geometric series with ratio $\frac{1+g}{1+r}$. The sum of the first
$n$ terms of a geomtric series is:

$$
\sum^{n-1}_{k=0}{x^k} = \frac{1 - x^n}{1 - x}, \text{ for } x \ne 1
$$

So:

$$
\sum^{n-1}_{k=0}\left(\frac{1 + g}{1 + r}\right)^k
  = \frac{1 - \left(\frac{1 + g}{1 + r}\right)^n}
         {1 - \left(\frac{1 + g}{1 + r}\right)}
$$

Doing some algebra:

$$
\begin{aligned}
\text{FV} &= C(1 + r)^{n-1} \cdot \frac{1 - \left({1 + g}{1 + r}\right)^n}
    {1 - \left(\frac{1 + g}{1+r}\right)}\\\\
    &= C (1 + r)^{n-1} \cdot \frac{1 - \left(\frac{1 + g}{1 + r}\right)^n}
        {\frac{r - g}{1 + r}}\\\\
    &= C \frac{(1 + r)^n - (1 + g)^n}{r - g}
\end{aligned}
$$

With this equation we can use `R`s `uniroot` function to find the percentage of
the salary that must be saved:

```{r}
p55$salary <- 70000
p55$r <- .09
p55$g <- .03
p55$n <- 30
p55$fv <- 2e6
# Future value factor depends on the discount rate, the growth rate, and the
# number of years we will be investing.
p55$fvga <- function(r, g, n) ((1 + r)^n - (1 + g)^n) / (r - g)
# Function used to determine the percent (p) of our salary to invest.
p55$f <- function(p) p55$salary * p * p55$fvga(p55$r, p55$g, p55$n) - p55$fv
# Uniroot will iterate over the function until it finds when the value is zero.
uniroot(p55$f, upper=5, lower=.01, tol=1e-2)$root
```

---

## 56. Balloon payments

On September 1, 2009, Susan Chao bought a motorcycle for \$30,000. She paid
\$1,000 down and financed the balance with a five-year loan at a stated annual
interest rate of 7.2 percent, compounded monthly. She started the monthly
payments exactly one month after the purchase (i.e., October 1, 2009). Two years
later, at the end of October 2011, Susan got a new job and decided to pay off
the loan. If the bank charges her a 1 percent prepayment penalty based on the
loan balance, how much must she pay the bank on November 1, 2011?

---

```{r problem 56 balloon payments; setup, echo=F}
p56 <- new.env()
p56$loan <- 30000 - 1000
p56$r <- .072
p56$n <- 5*12
p56$f <- function(c) c * (1 - (1 + p56$r/12)^(-p56$n))/(p56$r/12) - p56$loan
p56$factor <- (1 - (1 + p56$r/12)^(-p56$n))/(p56$r/12)
p56$payments <- round(uniroot(p56$f, upper=1000, lower=100, tol=10)$root, 2)
p56$n.paid <- 2*12
p56$balance <- p56$payments *
    (1 - (1 + p56$r/12)^(-(p56$n - p56$n.paid)))/(p56$r/12)
p56$balloon <- p56$balance * 1.01
```

### Determine the payments

For a 5-year loan of \$`r pretty(p56$loan,0)` at `r percent(p56$r)` APR. The
payments can be calculated:

$$
\begin{aligned}
`r pretty(p56$loan, 0)` &= C
\times \frac{\left[1 - (1 + `r p56$r/12`)^{`r p56$n`}\right]}{`r p56$r/12`}\\\\
C &= \frac{`r pretty(p56$loan, 0)`}{`r p56$factor`}\\
  &= `r pretty(p56$payments, 2)`
\end{aligned}
$$

### Amortize the loan

When Susan decided to pay off the loan she had already paid for `r p56$n.paid`
months.  The remaining term of the loan is `r p56$n - p56$n.paid` months. The bank
charges 1 percent prepayment penalty. The payment on November 1, 2011 will be:

$$
\begin{aligned}
\text{Payment} &= `r pretty(p56$payments, 2)`
    \times \frac{\left[1 - (1 + `r p56$r/12`)^{-`r p56$n - p56$n.paid`}\right]}
    {`r p56$r/12`} \times 1.01\\
    &= `r pretty(p56$balloon, 2)`
\end{aligned}
$$

Book: \$18,451.74

---

## 57. Calculating annuity values

Bilbo Baggins wants to save money to meet three objectives. First, he would like
to be able to retire 30 years from now with a retirement income of \$23,000 per
month for 20 years, with the first payment received 30 years and 1 month from
now. Second, he would like to purchase a cabin in Rivendell in 10 years at an
estimated cost of \$320,000. Third, after he passes on at the end of the 20
years of withdrawals, he would like to leave an inheritance of \$1,000,000 to
his nephew Frodo. He can afford to save \$2,100 per month for the next 10 years.
If he can earn an 11 percent EAR before he retires and an 8 percent EAR after he
retires, how much will he have to save each month in Years 11 through 30?

---

```{r problem 57: calculating annuity values; setup, echo=F}
p57 <- new.env()
p57$r.pre <- round((1 + .11)^(1/12) - 1, 6)
p57$r.post <- round((1 + .08)^(1/12) - 1, 6)
p57$pv.withdrawals <- 23000 * (1 - (1 + p57$r.post)^(-20*12)) / p57$r.post
p57$pv.inheritance <- 1e6 * (1 + p57$r.post)^(-20*12)
p57$pv.retirement <- p57$pv.withdrawals + p57$pv.inheritance
p57$savings.1 <- (2100 * ((1 + p57$r.pre)^(10*12) - 1)/p57$r.pre)
p57$savings.minus.cabin <- p57$savings.1 - 320000
p57$savings.for.20yrs <- p57$savings.minus.cabin * (1 + p57$r.pre)^(20*12)
p57$shortfall <- p57$pv.retirement -  p57$savings.for.20yrs
p57$payment.2 <- (p57$shortfall * p57$r.pre) / ((1 + p57$r.pre)^(20*12) - 1)
```

### Objectives summary

* __Retirement goal__: \$23,000/month for 20 years, starting 30 years + 1 month
  from now (i.e., at $t = 361 months).

* __Cabin purchase__: \$320,000 lump sum at year 10.

* __Inheritance: \$1,000,000 to Frodo at the end of withdrawals (i.e., 50 years
  from now).

* He saves \$21,000/month for 10 years.

* Must compute monthly savings from Year 11 to Year 30 (i.e., months 121 to
  360).

* EARs:

  - Before retirement (Years 1-30): 11\% EAR
  
  - After retirement (Years 31-50): 8\% EAR

### Step 1: Convert EARs to monthly rates.

We need monthly rates because all payments/savings are monthly.

* Before retirement

$$
r_{\text{pre}} = (1 + 0.11)^{1/12} - 1 \approx `r p57$r.pre`
$$

* After retirement:

$$
r_{\text{post}} = (1 + 0.08)^{1/12} - 1 \approx `r p57$r.post`
$$

### Step 2: Present value at retirement of retirement income + inheritance

We want the amount needed at $t = 360$ (30 years from now to:

1. Support 20 years of \$23,000/month withdrawals.

2. Leave \$1,000,000 at end of 50 years (i.e., at t = 600).

The PV at $t = 360$ of the withdrawals:

$$
\begin{aligned}
\text{PV}_{\text{withdrawals}} &= 23,000 \times \left[
    \frac{1 - (1 + r_{\text{post}})^{-240}}{r_{\text{post}}}\right]\\
    &= 23,000 \times \frac{1 - (1 + `r p57$r.post`)^{-240}}{`r p57$r.post`}\\
    \approx `r pretty(p57$pv.withdrawals)`
\end{aligned}
$$

The PV at $t = 360$ of the \$1,000,000 inheritance (i.e., 240 months later)

$$
\begin{aligned}
\text{PV}_{\text{inheritance}} = 1,000,000
  \times (1 + r_{\text{post}})^{-240}\\
  &= `r pretty(p57$pv.inheritance)`
\end{aligned}
$$

Total required at retirement ($t = 360$):

$$
\text{Needed}_{360} = \text{PV}_{\text{withdrawals}} +
  \text{PV}_{\text{inheritance}} \approx `r pretty(p57$pv.retirement)`
$$

### Step 3: Value of existing savings

This is a future value of an annuity with

* PMT = \$2,100
* n = 120
* r = 0.008746

$$
\begin{aligned}
\text{FV}_{10} &= 2,100 \times \frac{(1 + r)^{120} - 1}{r}
  &= `r pretty(p57$savings.1)`
\end{aligned}
$$

BUT---he uses this to buy a \$320,000 cabin at Year 10:

$$
\begin{aligned}
\text{Net FV}_{10} &= `r pretty(p57$savings.1)` - 320,000\\
  &= `r pretty(p57$savings.minus.cabin)`
\end{aligned}
$$

### Step 4: This balance grows for 20 more years (to year 30):

$$
\text{FV}_{30} = `r pretty(p57$savings.minus.cabin)`
  \times (1 + `r p57$r.pre`)^{-240} \approx
  `r pretty(p57$savings.for.20yrs)`
$$

### Step 5: Remaining needed at year 30

$$
\text{Shortfall}_{30} = `r pretty(p57$pv.retirement)`
  - `r pretty(p57$savings.for.20yrs)` = `r pretty(p57$shortfall)`
$$

### Step 6: Compute monthly savings needed in years 11-30

$$
\begin{aligned}
\text{FV} &= \text{PMT} \times \frac{(1 + r)^n - 1}{r} \Leftarrow
  \text{PMT} = \frac{FV \cdot r}{(1 + r)^n - 1}\\\\
\text{PMT} &= \frac{`r pretty(p57$shortfall)` \times `r p57$r.post`}
  {(1 + `r p57$r.post`)^{240} - 1}\\
  &\approx `r pretty(p57$payment.2)`
\end{aligned}
$$

---

## 58. Calculating annuity values

After deciding to buy a new car, you can either lease the car or purchase it
with a three-year loan. The car you wish to buy costs \$31,000. The dealer has a
special leasing arrangement where you pay \$1,500 today and \$405 per month for
the next three years. If you purchase the car, you will pay it off in monthly
payments over the next three years at an APR of 6 percent. You believe that you
will be able to sell the car for \$20,000 in three years. Should you buy or
lease the car?  What break-even resale price in three years would make you
indifferent between buying and leasing?

---

```{r problem 58: calculating annuity values; setup, echo=F}
p58 <- new.env()
p58$cost <- 31000
p58$resale <- 20000
p58$lease.today <- 1500
p58$lease.monthly.payments <- 405
p58$r <- .06 / 12
p58$t <- 3 * 12
p58$pvia <- function(r, t) (1 - (1 + r)^(-t))/r
p58$pv.lease <- 1500 + 405 * (1-(1+p58$r)^(-p58$t))/(p58$r)
p58$f <- function(c) c * p58$pvia(p58$r, p58$t) - p58$cost
p58$c <- uniroot(p58$f, upper=1000, lower=100, tol=1)$root
p58$pv.resale <- 20000 / (1 + p58$r)^(p58$t)
p58$pv.buying <- p58$cost - p58$pv.resale
p58$break.even <- uniroot(
    function(x) p58$cost - x / (1 + p58$r)^(p58$t) - p58$pv.lease,
    upper=31000,
    lower=1000,
    tol=10)$root
```

### Given

* To buy the car, it will cost \$`r pretty(p58$cost,0)`. You will pay off the
  car in monthly payments for three years at an APR of 6\%. You are confident
  you can sell the car for \$20,000 after three years.

* To lease the car, it will cost \$`r pretty(p58$lease.today,0)` upfront, and
  monthly payments of \$`r pretty(p58$lease.monthly.payments,0)` for three
  years.

### Option 1: Lease

* Upfront: \$`r pretty(p58$lease.today, 0)`

* Monthly lease payments: \$`r pretty(p58$lease.monthly.payments, 0)` for
  `r p58$t` months.

The PV of leasing the car is:

$$
\text{PV}_{\text{lease}} = `r pretty(p58$lease.today, 0)` +
  `r pretty(p58$lease.monthly.payments, 0)` \times
  \left[\frac{1 - (1 + `r p58$r`)^{-`r p58$t`}}{`r p58$r`}\right]
  \approx `r pretty(p58$pv.lease)`
$$

### Option 2: Buy

* Car price: \$31,000

* Loan: 6\% APR over 36 months

* Resale: \$20,000

#### Monthly payments

The monthly payments can be calculated:

$$
\text{PMT} = \frac{P \cdot r}{1 - (1 + r)^{-n}}
  = \frac{`r pretty(p58$cost, 0)` \cdot `r p58$r`}
    {1 - (1 + `r p58$r`)^{-`r p58$t`}}
  \approx `r pretty(p58$c, 2)`
$$

The PV of the payments is just the \$31,000---you're buying the car, so that's
the up-front cost (financed, but paid off monthly).

The present value of resale in 3 years:

$$
\text{PV}_{\text{resale}} = \frac{`r pretty(p58$resale, 0)`}
    {(1 + `r p58$r`)^{`r p58$t`}}
  \approx `r pretty(p58$pv.resale, 0)`
$$

The net cost of buying the car:

$$
\text{Total PV}_{\text{buy}} = `r pretty(p58$cost, 0)` -
  `r pretty(p58$pv.resale, 0)` = `r pretty(p58$pv.buying, 0)`
$$

### Comparing leasing and buying:

* Lease: `r pretty(p58$pv.lease, 0)`

* Buy: `r pretty(p58$pv.buying, 0)`

In this demonstration, buying is cheaper than leasing.

### Break even resale value

$$
`r pretty(p58$cost, 0)` - \frac{X}{(1 + `r p58$r`)^{`r p58$t`}}
  = `r pretty(p58$pv.lease, 0)`
$$

Solving for X we get \$`r pretty(p58$break.even, 0)`.

---

## 59. Calculating annuity values

An All-Pro defensive lineman is in contract negotiations. The team has offered
the following salary structure:

```{r problem 59; setup, echo=F}
p59 <- new.env()
p59$time <- seq(0, 6, by=1)
p59$salary <- c(8.5e6, 3.9e6, 4.6e6, 5.3e6, 5.8e6, 6.4e6, 7.3e6)
knitr::kable(data.frame(Time = p59$time, Salary = pretty(p59$salary, 0)))
```

All salaries are to be paid in a lump sum. The player has asked you as his agent
to renegotiate the terms. He wants a \$10 million signing bonus payable today
and a contract value increase of \$1,500,000. He also wants an equal salary paid
every three months, with the first paycheck three months from now. If the
interest rate is 5 percent compounded daily, what is the amount of his quarterly
check? Assume 365 days in a year.

---

```{r problem59; solution, echo=F}
p59$bonus <- 10e6
p59$contract.increase <- 1.5e6
p59$payment.interval <- 4 # Quarterly
p59$r <- .05/365

p59$salary <- p59$salary + p59$contract.increase/7
```
