---
title: "Options and Corporate Finance"
author: "Ross, Westerfield, Jaffe"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Examples from the text

## Call option payoffs

Suppose Mr. Optimist holds a one-year call option on TIX common stock. It is a European call option and can be exercised at \$150. Assume that the expiration date has arrived. What is th evalue of the TIX call option on the expiration date? If TIX is selling for \$200 per share, Mr. Optimist can exercise the option---purchase TIX at \$150---and then immediately sell the share at \$200. Mr. Optimist will have made \$50 ($=\$200 - \$150$). Thus, the price of this call option must e \$50 at expiration.

Instead assume that TIX is selling for \$100 per share on the expiration date. If Mr. Optimist sill holds the call option, he will throw it out. The value of the TIX call option on the expiration date will be zero in this case.

---

## Put option payoffs

Ms. Pessimist believes that BMI will fall from its current \$160 per-share price. She buys a put. Her put option contract gives her the right to sell a share of BMI stock at \$150 one year from now. If the  price of BMI is \$200 on the expiration date, she will tear up the put option contract because it is worthless. That is, she will not want to sell stock worth \$200 for the exercise price of \$150.

On the other hand, if BMI is selling for \$100 on the expiration date, she will exercise the option. In this case she can buy a share of BMI in the market for \$100 per share and turn around and sell the share at the exercise price of \$150. Her profit will be \$50 ($=\$150 - \$100$). Therfore, the value of the put option on the expiration date will be \$50.

---

## A Synthetic T-Bill

Suppose shares of stock in Smolira Corp. are selling for \$110. A call option on Smolira with one year to maturity and a \$110 strike price sells for \$15. A put with the same terms sells for \$5. What's the risk-free rate?

To answer, we need to use put-call parity to determine the price of a risk-free, zero coupon bond:

$$
\text{Price of underlying stock} + \text{Price of put}
  - \text{Price of call} = \text{Present value of exercise price}
$$

Plugging in the numbers, we get:

$$
\$110 + \$5 - \$15 = \$100
$$

Because the present value of the \$110 strike price is \$100, the implied risk-free rate is 10 percent.

$$
X = \frac{110}{100} = 1.1
$$

---

## Black-Scholes

Consider Private Equipment Company (PEC). On October 4 of Year 0, the PEC April 29 call option had a closing value of \$4. The stock itself was selling at \$50. On October 4, the option had 199 days to expiration (maturity date = April 21, Year 1). The annual risk-free interest rate, continuously compounded, was 7 percent.

This information determines three variables directly:

1. The stock price, $S$, is \$50.
2. The exercise price, $E$, is \$49.
3. The risk-free rate $R$, is .07.
4. The time to maturity $t$, is $199/365$.

In the real world, an option trader would know $S$ and $E$ exactly. Traders generally view U.S. Treasury bills as riskless, so a current quote from The Wall Street Journal or a similar source would be obtained for the interest rate. The trader would also know (or could count) the number of days to expiration exactly. Thus, the fraction of a year to expiration, $t$, could be calculated quickly.

The problem comes in determining the variance of the stock's return. The formula calls for the variance between the purchase date of October 4 and the expiration date. Unfortunately, this represents the future, so the correct value for variance is not available. Instead, traders frequently estimate variance from past data, just as we calculated variance in an earlier chapter. In addition, some traders may use intuition to adjust their estimate. For example, if anticipation of an upcoming event is likely to increase the volatility of the stock, the trader might adjust her estimate of variance upward to reflect this. (This problem was most severe right after the October 19, 1987 crash. The stock market was quite risky in the aftermath, so estimates using precrash data were too low.)

The preceding discussion was intended merely to mention the difficulties in variance estimation, not to present a solution. For our purposes, we assume that a trader has come up with an estimate of variance.

5. The variance of Private Equipment Co. has been estimated to be .09 per year.

Using these five parameters, we calculate the Black-Scholes value of the PEC call option in three steps:

* Step 1: Calculate $d_1$ and $d_2$. These values can be determined by a straightforward, albeit tedious insertion of our parameters into the basic formula. We have:

$$
\begin{aligned}
d_1 &= \left[ \ln{\left(\frac{S}{E}\right)} + (R + \sigma^2/2)t \right] / \sqrt{\sigma^2 t}\\
  &= \left[ \ln{\left(\frac{50}{49}\right)} + (.07 + .09/2) \times \frac{199}{365} \right] 
    / \sqrt{.09 \times \frac{199}{365}}\\
  &= [ .0202 + .0627]/.2215\\
  &= .3742\\\\
d_2 &= d_1 - \sqrt{\sigma^2 t}\\
  &= .1527
\end{aligned}
$$

* Step 2: Calculate $N(d_1)$ and $N(d_2)$. We can best understand the values $N(d_1)$ and $N(d_2)$ as the cumulative probability. We use a normal distribution with an expected value of 0 and a standard deviation of 1. This is called the __standardized normal distribution__. We mentioned in an earlier chapter that the probability that a drawing from this distribution will be between $-1$ and $+1$ (within one standard deviation of its mean, in other words) is $68.26$ percent.

Now let us ask a different question: What is the probability that a drawing from the standardized normal distribution will be _below_ a particular value? For example, the probability that a drawing will be below 0 is clearly 50 percent because the normal distribution is symmetric. Using statistical terminology, we say that the __cumulative probability__ of 0 is 50 percent. Statisticians also say that $N(0) = 50\%$.

### Using `R` to calculate normal distributions

We use the normal distribution function (Cumulative Distribution Function), `pnorm`,
to calculate $N(d_1)$ and $N(d_2)$.

```{r Example 22.4: Black-Scholes: finding normal distribution values}
d.1 <- .3742
d.2 <- .1527

pnorm(d.1)
pnorm(d.2)
```

Step 3: Calculate C.

$$
\begin{aligned}
C &= S \times N(d_1) - E e^{-Rt} \times N(d_2)\\
  &= \$50 \times N(d_1) - \$49 e^{-.07 \times (199/365)} \times N(d_2)\\
  &= (\$50 \times .6459) - (\$49 \times .9626 \times .5607)\\
  &= \$32.295 - \$26.447\\
  &= \$5.85
\end{aligned}
$$

---

## Stocks and bonds as options

The Popov Company has been awarded the concessions at next year's Olympic Games in Antarctica. Because the firm's principals live in Antarctica and because there is no other concession business on that continent, their enterprise will disband after the games. The firm has issued debt to help finance this venture. Interest and principal due on the debt next year will be \$800, at which time the debt will be paid off in full. The firm's cash flows next year are forcast as follows:

```{r example 22.5: stocks and bonds as options: table, echo=F}
m <- matrix(nrow=2, c(1000, -800, 850, -800, 700, -700, 550, -550))
csums <- colSums(m)
m <- rbind(m, csums)
rownames(m) <- c("Cash flow before interest and principal", "- interest and principal",
                 "Cash flow to stockholders")
colnames(m) <- c("Very Successful Games", "Moderately Successful Games", "Moderately Unsuccessful Games",
                 "Outright Failure")
knitr::kable(m, caption="Example 22.5: Popov's Cash Flow Schedule")
```

As can be seen, the principals forecast four equally likely scenarios. If either of the first two scenarios occurs, the bondholders will be paid in full. The extra cash flow goes to the stockholders. However, if either of the last two scenarios occurs, the bondholders will not be paid in full. Instead, they will receive the firm's entire cash flow, leaving the stockholders with nothgin.

---

## Options and capital budgeting

Exoff Oil Corporation is considering the purchase of an oil field in a remote part of Alaska. The seller has listed the property for \$10,000 and is eager to sell immediately. Initial drilling costs are \$500,000. Exoff anticipates that 10,000 barrels of oil can be extracted each year for many decades. Because the termination date is so far in the future and so hard to estimate, the firm vies the cash flow stream from the oil as a perpetuity. With oil prices at \$50 per barrel and extraction costs at \$46 a barrel, the firm anticipates a net margin of \$4 per barrel. Because oil prices are expected to rise at the inflation rate, the firm assumes that its cash flow per barrel will always be \$4 in real terms. The appropriate real discount rate is 10 percent. The firm has enough tax credits from bad years in the past that it will not need to pay taxes on any profits from the oil field. Should Exoff buy the property?

The NPV of the oil field to Exoff is

$$
-\$110,000 = -\$10,000 - \$500,000 + \frac{\$4 \times 10,000}{.10}
$$

According to this analysi, Exoff should not purchase the land.

Though this approach uses the standard capital budgeting techniques of this and other textbooks, its actually inappropriate for this situation. To see this, consider the analysis of Kirtley Thorton, a consultant to Exoff. He agrees that the price of oil is _expected_ to rise at the rate of inflation. However, he points out that the next year will be quite perilous for oil prices. On the one hand, OPEC is considering a long-term agreement that would raise oil prices to \$65 per barrel in real terms for many years in the future. On the other hand, National Motors recently indicated that cars using a mixture of sand and water for fuel are currently being tested. Thornton argues that oil will be priced at \$35 per barrel in real terms for many years should this developement prove successful. Full information about both these developments will be released in exactly one year.

Should oil prices rise to \$65 a barrel, the NPV of the project would be:

$$
\$1,390,000 = -\$10,000 - \$500,000 + \frac{\$65 - \$46) \times 10,000}{.10}
$$

However, should oil prices fall to \$35 a barrel, the NPV of the oil field will be even more negative than it is today.

Mr. Thornton makes two recommendations to Exoff's board. He argues that:

1. The land should be purchased.
2. The drilling decision should be delayed until information about both OPEC's new agreement and National Motors' new automobile is released.

Mr. Thornton explains his recommendations to the board by first assuming that the land has already been purchased. He argues that under this assumption, the drilling decision should be delayed. Second, he investigates his assumption that the land should have been purchased in the first place. This approach of examining the second decision (whether to drill) after assuming that the first decision (to buy the land) has been made was also used in our earlier presentation on decision trees. Let us now work through Mr. Thornton's analysis.

_Assume the land has already been purchased._ If the land has already been purchased, should drilling begin immediately? If drilling begins immediately, the nPV is -\$110,000: If the drilling decision is delayed until new information is released in a year, the optimal choice can be made at that time.

If oil prices drop to \$35 a barrel, Exoff should not drill. Instead, the firm should walk away from the project, losing nothing beyond its \$10,000 purchase price for the land. If oil prices rise to \$65, drilling should begin.

Mr. Thornton points out that by delaying, the firm will invest the \$500,000 of drilling costs only if oil prices rise. Thus, by delaying, the firm saves \$500,000 in the case where oil prices drop. Kirtley concludes that once the land is purchased, the drilling decision should be delayed[^1].

[^1]: Actually, there are three separate effects here. First, the firm avoids drilling costs in the case of low oil prices by delaying the decisin. This is the effect discussed by Mr. Thornton. Second, the present value of the \$500,000 payment is less when the decision is delayed, even if drilling eventually takes place. Third, the firm loses one year of cash inflows through delay.  
&nbsp;&nbsp;&nbsp;&nbsp;The first two effects support delaying the decision. The third effect supports immediate drilling. In this example, the first effect greatly outweighs the other two effects. Thus, Mr. Thornton avoided the second and third effects in his presentation.

_Should the land have been purchased in the first place?_ We now know that if the land has been purchased, it is optimal to defer the drilling decision until the relase of information. Given that we know this optimal decision concerning drilling, should the land be purchased in the first place? Without knowing the exact probability that oil prices will rise, Mr. Thornton is nevertheless confident that the land should be purchased. The NPV of the project at \$65 per barrel oil prices is \$1,390,000, whereas the cost of the land is only \$10,000. Mr. Thornton believes that an oil price rise is possible, though by no means probable. Even so, he argues that the high potential return is clearly worth the risk.

---

# Summary and Conclusions

1. The most familiar options are puts and calls. These options give the holder the right to sell or buy shares of common stock at a given exercise price. American options can be exercised at any time up to and including the expiration date. European options can be exercised only on the expiration date.

2. We showed that a strategy of buying a stock and buying a put is equivalent to a strategy of buying a call and buying a zero coupon bond. From this, the put-call parity relationship was established:

$$
\text{Value of stock} + \text{Value of put} - \text{Value of call} = \text{PV of exercise price}
$$

3. The value of an option depends on five factors:

> a. The price of the underlying asset.
> b. The exercise price.
> c. The expiration date.
> d. The variablility of the underlying asset.
> e. The interest rate on risk-free bonds.

The Black-Scholes model can determine the intrinsic price of an option from these five factors.

4. Much of corporate financial theory can be presented in terms of options. In this chapter, we pointed out that:

> a. Common stock can be represented as a call option on the firm.
> b. Stockholders enhance the value of their call by increasing the risk of their firm.
> c. Real projects have hidden options that enhance value.

---

# Concept Questions

## Options

What is a call option? A put option? Under what circumstances might you want to buy each? Which one has greater _potential_ profit? Why?

---

A call option gives the buyer the right (but not the obligation) to buy an asset (typically a stock) at a specified strike price before or at the expiration date. You would buy call option if you expect the price of the underlying asset to increase in the future.

A put option gives the buyer the right (but not the obligation) to sell an asset at a specified strike price before or at the expiration date. You would buy a put option if you expect the price of the underlying asset to decrease in the future.

A call opiton has theoretically unlimited profit potential, while a put option has a capped maximum profit.

---

## Options

Complete the following sentence for each of these investors:

> a. A buyer of call options.
> b. A buyer of put options.
> c. A seller (writer) of call options.
> d. A seller (writer) of put options.

"The (buyer/seller) of a (put/call) option (pays/receives) money for the (right/obligation) to (buy/sell) a specified asset at a fixed price for a fixed length of time."

---

a. The buyer of a call option pays money for the right to buy a specified asset at a fixed price for a fixed length of time.

b. The buyer of a put option pays money for the right to sell a specified asset at a fixed price for a fixed length of time.

c. A seller of a call option receives money for the obligation to sell a specified asset at a fixed price for a fixed length of time.

d. A seller of a put option receives money for the obligation to buy a specified asset at a fixed price for a fixed amount of time.

---

## American and European options

What is the difference between an American option and a European option?

---

An American option can be exercised at any time, including on the expriation date. A European option can only be exercised on expiration.

---

## Intrinsic value

What is the intrinsic value of a call option? Of a put option? How do we interpret these values?

---

The intrinsic value of a call option is the amount by which the stock price is above the strike price. If $S$ is the current stock price and $K$ is the strike price.

$$
\text{Intrinsic value (call)} = \text{max}(0, S - K)
$$

* If $S > K$, the call option has positive intrinsic value ("in the money").
* If $S \le K$, the intrinsic value is zero because the option would be worthless at expiration ("at the money" or "out of the money").

The intrinsic value of a put option is the amount by which the stock price is below the strike price.

$$
\text{Intrinsic value (put)} = \text{max}(0, K - S)
$$

* If $S < K$, the put option has positive intrinsic value ("in the money").
* If $S \ge K$, the intrinsic value is zero because the option would be worthless at expiration ("at the money" or "out of the money").

Intrinsic value represent the option's worth if exercised immediately. Options can have time value (meaning their total price may be higher than just the intrinsic value due to volatility and time until expiration). At expiration, an option's value is purely its intrins value---there is no remaining time value.

---

## Option pricing

You notice that shares of stock in the Patel Corporation are going for \$50 per share. Call options with an exercise price of \$35 per share are selling for \$10. What's wrong here? Describe how you can take advantage of this mispricing if the option expires today.

---

This situation presents an arbitrage opportunity due to the mispricing of the call option.

### Step 1: Identify the problem

The stock is trading at \$50 per share but the call option has a strike price of \$35 and is selling for \$10. The intrinsic value of the call option should be:

$$
\text{Intrinsic value} = S - K = 50 - 35 - 15
$$

Since options cannot be worth less than their intrinsic value, the option should be at least \$15, but it's priced at \$10, meaning it is undervalued by \$5.

### Step 2: Arbitrage strategy

Since the option expires today, the best strategy is to buy the call option and immediately exercise it. The profit per option will be $(50 - 35) - 10 = 5$.

This mispricing creaets a risk-free arbitrage because you're guaranteed a \$5 profit per option contract without any market risk. In reality, such opportunities rarely last because traders will quickly exploit them, causing prices to adjust.

---

## Options and stock risk

If the risk of a stock increases, what is likely to happen to the price of call options on the stock? To the price of put options? Why?

---

If the risk (volatility) of a stock increases, the prices of both call options and put options will generally increase.

### Impact on call option prices

A call option gives the holder the right (but not the obligation) to buy the stock at a fixed price (the strike price). When volatility increases:

* The potential upside gain increases because the stock price has a higher chance of rising significantly.
* Even though the stock price could also drop, call option holders are not obligated to exercise, so their potential loss is limited to the premium paid.
* Higher volatility makes the call option more valuable because there's a greater chance of the stock moving into profitable territory.

### Impact on put option prices

A put option gives the holder the right to sell the stock at a fixed price. When volatility increases:

* The downside potential increases, making puts more valuable because there's a higher chance of the stock price dropping below the strike price.
* Like call options, put options have limited downside risk (just the premium paid), but they benefit from large price swings downward.

### Overall effect

Since options derive their value in part from the probability of the stock moving favorably for the option holder, higher volatility increases the likelihood of extreme price swings, which raises the value of both calls and puts.

This relationship is reflected in option pricing models like the Black-Scholes model where volatility is a key input.

---

## Option risk

True or false: The unsystematic risk of a share of stock is irrelevant for valuing the stock because it can be diversified away; therefore, it is also irrelevant for valuing a call option on the stock. Explain.

---

False. While it is true that unsystematic risk (company-specific risk) of a stock is irrelevant for valuing the stock itself---because it can be diversified away in a well-diversified portfolio---it is not irrelevant for valuing a call option on the stock.

### Stock valuation vs. option valuation

The value of a stock is determined by its expected future cash flows, discounted at an appropriate rate based on its systematic risk (market risk). Since unsystematic risk can be eliminated through diversification, it does not affect the stock's required return or value. The value of an option, however, depends on the volatility of the stock price, which includes both systematic and unsystematic risk.

### Option pricing and volatility

Option prices (especially in models like Black-Scholes) are directly influenced by total volatility (which includes both systematic and unsystematic risk). Even if the unsystematic risk of a stock is diversifiable and irrelevant to its expected return, higher volatility increases the value of both call and put options because it raises the probability of large price swings, increasing the potential for profitable option exercises.

### Key takeaway

Although unsystematic risk doesn't matter for stock valuation, it does affect option valuation because options benefit from higher volatility. Thus, it is incorrect to conclude that unsystematic risk is irrelevant for valuing a call option just because it is irrelevant for valuing the stock.

---

## Option pricing

Suppose a certain stock currently sells for \$30 per share. If a put option and a call option are available with \$30 exercise prices, which do you think will sell for more? Explain.

---

If a put option and a call option both have an exercise price of \$30 (at-the-money options), the put option is likely to sell for more if all other factors (time to expiration, volatility, interest rates, etc.) are equal.

### Put-call parity and interest rates.

The put-call parity formula states:

$$
C - P = S - \frac{X}{(1 + r)^{t}}
$$

where:

* C = call option price
* P = put option price
* S = stock price
* X = exercise price
* r = risk-free interest rate
* t = time to expiration

Since S = X, the formula simplifies to 

$$
C - P = -\frac{X}{(1 + r)^{t}} + X
$$

Since $\frac{X}{(1 + r)^{t}} < X$, this means $P > C$ if interest rates are low or zero.

### Interest rate effect

When interest rates are low, the present value of $\frac{X}{(1 + r)^{t}}$ is close to $X$, making the put option more expensive. When interest rates are high, call options tend to be worth more because investors prefer to defer payments (buying shares later at a fixed price).

### Dividend effect

If the stock pays dividends, puts become more valuable because the stock price is expected to drop when dividends are payed.

### Conclusion

In most typical cases, especially when interest rates are low and the stock pays dividends, the put option will be slightly more expensive than the call option for an at-the-money option.

---

## Option price and interest rates

Suppose the interest rate on T-bills suddenly and unexpectedly rises. All other things being the same, what is the impact on call option values? On put option values?

---

If the interest rae on T-bills (or the risk-free rate) suddenly and unexpectedly rises, the impact on call and put option values can be understood using the Black-Scholes option pricing model and put-call parity.

### Effect on call option values

A rise in interest rates generally increases the value of a call option. The reason is higher interest rates reduce the present value of the exercise, or strike, price. Since buying a stock requires tying up capital, higher interest rates make delaying that purchase (via a call option) more beneficial).

### Effect on put optin values

A rise in interest rates generally decreases the value of a put option. The present value of the exercise price decreases, reducing the value of the guaranteed sale price in the future. A put option benefits from a high present value of the strike price, so when interest rates rise and this present value falls, the put becomes less valuable.

### Mathematical insight (Black-Scholes model)

The Black-Scholes formula for a European call option includes the risk-free rate $r$:

$$
C = S N(d_1) - X e^{-rt} N(d_2)
$$

The term $X e^{-rt}$ decreases when $r$ rises, making the call more valuable. Sicne put-call parity states:

$$
C - P = S - X e^{-rt}
$$

an increase in $r$ means $X e^{-rt}$ shrinks, making puts less valuable.

### Conclusion

Call options become more valuable when interest rates rise, while put option become less valuable.

---

## Contingent liabilities

When you take out an ordinary student loan, it is usually the case that whoever holds that loan is given a guarantee by the U.S. government, meaning that the government will make up any payments you skip. This is just one example of the many loan guarantees made by the U.S. government. Such guarantees don't show up in calculations of government spending or in official deficit figures. Why not? Should they show up?

---

Loan guarantees by the U.S. government do not show up in official government spending or deficit figures because they are contingent liabilities---the government is only required to make payments if borrowers default. Unlike direct government spending, which immediately affects the budget, loan guarantees are a potential obligation rather than an actual expenditure.

### Why don't they show up in deficit figures?

1. No immediate cash flow - A loan guarantee does not require the government to spend money immediately; it only comes into play if the borrower defaults.
2. Budget accounting rules - The U.S. government follows cash-based accounting, which means only actual cash outflows (not potential future liabilities) are counted in deficit calculations.
3. Uncertain costs - It's difficult to estimate when and how much the government will need to pay due to defaults, making it challenging to include in official spending numbers.

### Should they be included?

Many economists argue that yes, they should be reflected, at least as estimated liabilities, because:

* They represent a real financial risk to taxpayers.
* They create moral hazard---lenders may take more risks, knowing the government will cover losses.
* They can result in large, unexpected government payouts (e.g., the 2008 financial crisis, where many guarantees turned into actual costs).

A more transparent approach be to estimate the expected cost of guarantees and include that in budget deficit calculations, similar to how businesses account for potential liabilities.

---

## Options and expiration dates

What is the impact of lengthening the time to expiration on an option's value? Explain.

---

Lengthening the time to expiration generally increases the value of both call and put options, assuming all other factors remain constant.

### More time means more uncertainty (higher time value)

The longer the time until expiration, the greater the chance that the stock price will move favorably for the option oholder. For call options, a longer time horizon increases the chance that the stock price will rise above the strike price. For put options, it increases the chance that the stock price will fall below the strike price. More time mean more opportunity for volatility, which benefits option holders.

### Less time decay (slower theta)

Time decay (Theta) is the erosion of an option's value as expiration nears. Longer-term options lose value more gradually, whereas shot-term options experience rapid time decay. This makes longer-term options inhernetly more valuable than shorter-term options.

---

## Options and stock price volatility

What is the impact of an increase in the volatility of the underlying stock's return on an option's value? Explain.

---

An increase in the volatility of the underlying stock increases the value of both call and put options.

### Higher volatility means greater potential gains

Options give the holder the right, but not the obligation, to buy (calls) or sell (puts) the stock at a fixed strike price. More volatility increases the chance that the stock price will move significantly in either direction, making the option more valuable.

### Asymmetry of payoff (limited loss, unlimited gain)

If you own an option, your downside is limited to the premium paid, but your upside is unlimited (for calls) or significant (for puts). When volatility rises, the probability of large price swings increases, which enhances the potential profit of the option.

### Impact on time value

Higher volatility increase the time value component of an option's price. Even if the option is currently out-of-the-money, higher volatility means it is more likely to become profitable before expiration.

### Mathematical justification (Black-Scholes model)

The Black-Scholes option pricing model shows that the price of an option increases as the standard deviation ($\sigma$, volatility) of the stock's return increases. Since volatility is a key input, hihger $\sigma$ always leads to a higher option price, all else being equal.

### Key takeaway

If you expect increased volatility, options become more valuable because the greater probability of profitable price movements. This is why options tend to be expensive when markets are highly uncertain or turbulent.

---

## Insurance as an option

An insurance policy is considered analogous to an option. From the policyholder's point of view, what type of option is an insurance policy? Why?

---

An insurance policy is analogous to a put option from the policyholder's perspective.

### A put option provides downside protection

A put option gives the holder the right to sell an asset at a fixed price (strike price), protecting against a decline in value. Similarly, an insurance policy protects the policyholder against losses from events like accidents, illnesses, or property damage.

### The insurance premium is like an option premium

When buying a put option, you pay a premium to have the right to sell at the strike price. In insurance, you pay premiums to secure coverage in case of a negative event (a "claimable loss").

### The policyholder "Exercises" the insurance when a loss occurs

If the insured event happens (e.g., your house burns down), the insurance company pays out a claim---just like a put option pays off when the stock price falls below the strike price. If no loss occurs, just like an out-of-the-money put option expiring worthless, the policyholder gets no pay out but has still paid the premium.

### Key takeaway

An insurance policy functions like a put option on the insured asset because it provides financial protection against downside risk with the premium acting as the cost of securing that protection.

---

## Equity as a call option

It is said that the equityholders of a levered firm can be thought of as holding a call option on the firm's assets. Explain what is meant by this statement.

---

The statement that equityholders of a levered firm can be thought of as holding a call option on the firm's assets comes from the option pricing perspective of corporate finance, specifically the Merton model of capital structure.

### Equity as a call option on firm assets

A firm's assets can be thought of as the underlying asset. The firm's debt is like the strike price of a call option. Equityholders only receive value if the firm's assets exceed the value of the debt at maturity.

### How this works in a levered firm

If the firm is successful and its assets are worth more than the debt (D) at maturity, equityholders get the residual value.

$$
\text{Equity value} = \text{max}(A - D, 0)
$$

where A is the firm's asset value.

If the firm's assets are worth less than the debt, the firm defaults, and equityholders get nothing (since debt holders are paid first).

This is exactly how a call option works: If the asset value is above the strike price (debt), equityholders exercise their option and keep the excess value. If the asset value is below the strike price (debt), equityholders let the option expire worthless, losing their investment.

### Implication of the call option view

Equityholders prefer volatility, since option value sincrease with volatility, equityholders in a levered firm benefit from riskier projects that could increase firm value. Debt holders bear downside risk. If asset values fall below the debt level, debt holders take the hit in bankruptcy. The more debt a firm has, the more its equity resembles a highly leveraged call option.

### Key takeaway

Equity in a levered firm is structurally similar to a call option because shareholders only get paid if the firm's assets exceed its liabilities, just like a call option holder only profits if the underlying stock exceeds the strike price.

---

## Option valuation and NPV

You are the CEO of Titan Industries and have just been awarded a large number of employee stock options. The company has two mutually exclusive projects available. The first project has a large NPV and will reduce the total risk of the company. The second project has a small NPV and will increase the total risk of the company. You have decided to accept the first project when you remember your employee stock options. How might this affect your decision?

---

As the CEO of Titan Industries, your decision about which project to undertake is influenced by two competing incentives:

1. __Maximizing firm value__ (which favors choosing the project with the higher NPV).
2. __Maximizing the value of your employee stock options__ (Which may favor the project that increases risk).

### How emmployee stock options affect your decision

Employee stock options function like call options on the company's stock. Since the value of call options increases with risk (volatility), you might have an incentive to favor the riskier project, even though it has a lower NPV.

The reasons are:

* __Higher risk increases stock volatility__, which raises the potential upside of the stock price.
* __Since options have limited downside risk__ (you don't lose money if the stock price drops below the exercise price), increasing risk gives you a better chance of making a large gain without personally bearing the downside. The Black-Scholes model and other option valuation methods show that higher volatility leads to a higher option value.

### Conflict between firm value and personal gain

Without stock options, you would rationally choose the higher NPV project because it maximizes shareholder wealth. With stock options, you might be tempted to pick the riskier project to increase the value of your options, even though it creates less value for the company overall.

### Ethical and fiduciary considerations

As CEO, you have a fiduciary duty to act in the best interests of shareholders. If you choose the riskier, lower-NPV project for personal gain, it could be considered a conflict of interest. However, if shareholders understand this incentive misalignment, they may adjust your compensation (e.g., using restricted stock instead of options) to align your interests with theirs.

### Final decision

Even though your options might tempt you to pick the riskier project, the rational and ethical choice is to accept the higher NPV project---unless your personal gain outweighs potential reputational damage and long-term firm performance.

---

## Put-call parity

You find a put and a call with the same exercise price and maturity. What do you know about the relative prices of the put and call? Prove your answer and provide an intuitive explanation.

---

When you have a put and a call option with the same exercise price (strike price) and maturity, you can use put-call parity to understand the relationship between their prices.

Put-Call Parity states that for European options (which can only be exercised at expiration), the price of a call option (C) and the price of a put option (P) with the same strike price (K) and maturity (T) are related to the spot price (S) of the underlying asset and the risk-free interest rate (r) as in the following:

$$
C - P = S - Xe^{-rT}
$$

Where:

* C = call option price
* P = put option price
* S = spot price of the underlying asset
* X = Exercise (stike) price
* r = Risk-free interest rate
* T = Time to maturity

### Derivation

#### Replicating portfolio approach:

You can create a portfolio that consists of a long position in a call and a short position in a put that replicates the payoff of being long in the underlying asset (stock). The payoff of a call at expiration is $\text{max}(S_T - K,0)$, and the payoff of a put is $\text{max}(K - S_T, 0)$. If you take a long call and short a put, at maturity, the portfolio payoff will look like being long the stock, i.e., $S_T - K$ (since if the stock goes up, the call profits, and if the stock goes down, the put loses, but the combined effect is equivalent to being long the stock).

#### Equating the portfolio with stock:

A long position in the stock, at time 0, has a value of $, and it grows at the risk-free rate, so at maturity, it will be workth $Xe^{-rT}$, because the initial price grows at the rate $e^{rT}$ to match the strike price. Hence, the price of the call option minus the price of the put option should be equal to the present value of the stock, i.e., $S - Ke^{-rT}$.

### Intuitive explaination

#### Call option price (C):

The price of a call option increases with the spot price of the underlying asset, because if the underlying price increases, the call option becomes more valuable. It gives you the right to buy at the strike price, and the higher the underlying price, the more likely you are to profit from it.

#### Put option price (P):

Conversely, the price of a put option increases as the spot price decreases. A put gives you the right to sell at the strike price, and the lower the underlying price, the more likely you are to profit from it.

Thus, when the strike price and maturity are the same for both the put and the call, the prices of the two options are related through the put-call parity equation, meaning that you cannot have an arbitrage opportunity without violating the relationship.

### Conclusion

From the put-call parity equation, you can infer that if the spot price of the asset increases, the price of the call option will tend to increase relative to the put option, and vice versa. Therefore, the relative prices of the put and call are directly connected through the spot price, strike price, and time to maturity.

In summary:

* If the spot price $S$ increases, the call option price $C$ will increase relative to the put option price $P$, as the call becomes more valuable.

* If the spot price $S$ decreases, the put option price $P$ will increase relative to the call price $C$, as the put becomes more valuable.

---

## Put-call parity

A put and a call have the same maturity and strike price. If they have the same price, which one is in the money? Prove your answer and provide an intuitive explanation.

---

If a put and a call option have the same maturity and strike price and they have the same price, we can analyze their relationship using put-call parity and consider the payoff structures of both options to determine which one is in the money.

The put-call parity relationship for European options:

$$
C - P = S - Xe^{-rT}
$$

Where:

* C = call option price
* P = put option price
* S = spot price of the underlying asset
* X = Exercise (stike) price
* r = Risk-free interest rate
* T = Time to maturity

If $C = P$, this means the left-hand side of the equation $C - P = 0$, so:

$$
0 = S - X e^{-rT}
$$

Which simplifies to:

$$
S = X e^{-rT}
$$

This equation shows that if the call and put prices are equal, the spot price of the underlying asset must be equal to the present value of the strike price (adjusted for the risk-free interest rate).

### Intuitive explanation

A call option is in the money if the spot price $S$ is higher than the strike price $X$. This is because the holder of the call option can buy the underlying asset at the strike price, and if the market price (spot price) is higher, there is an immediate profit opportunity.

A put option is in the money if the spot price $S$ is lower than the strike price $X$. This is because the holder of the put opiton can sell the underlying asset at the strike price, and if the market price (spot price) is lower, there is an immediate profit opportunity.

For $C = P$ to hold, the spot price must be exactly equal to $X e^{-rT}$, the discounted strike price. This condition corresponds to the exact point where the underlying asset's price is neither too high nor too low to favor one option over the other. At this point, neither the call nor the put option is "in the money" because neither would provide a profit if exercised immediately (the call would break even if $S = K$, and the put would also break even at the same price). This situation occurs exactly when the spot price is at the strike price, adjusted for the time value of money.

If the spot price is exactly $X e^{-rT}$, then neither the call nor the put is "in the money," because both would have a payoff of zero at expiration. Therefore, neither of the options is in the money at this point. They are effectively at the money, meaning they both have no intrinsic value, only time value.

### Conclusion

If a put and a call have the same maturity, strike price, and price, neither option is in the money. They are both at the money at this particular point because the spot price equals the present value of the strike price, meaning theoptions have no intrinsic value at that moment.

---

## Put-call parity

One thing put-call parity tells us is that given any three of a stock, a call, a put, and a T-bill, the fourth can be synthesized or replicated using the other three. For example, how can we replicate a shre of stock using a call, a put, and a T-bill?

---

To replicate a shre of stock using a call option, a put option, and a T-bill (risk-free bond), we can use the put-call parity relationship. Let's break this down and see how to construct the replication.

### Put-call parity recap


$$
C - P = S - Xe^{-rT}
$$

Where:

* C = call option price
* P = put option price
* S = spot price of the underlying asset
* X = Exercise (stike) price
* r = Risk-free interest rate
* T = Time to maturity

This equation essentially tells us how the prices of a call, a put, and the stock are related. It says that the difference between the price of the call and the price of the put is equal to the difference between the spot price of the stock and the present value of the strike price.

### Replication strategy

To replicate a share of stock using a call option, a put option, and a T-bill, we want to create a portfolio that behaves identically to holding one share of the stock. Let's derive this portfolio step by step.

1. Start with put-call parity

Rearranging the put-call parity equation to isolate $S$ (the stock):

$$
S = C - P + X e^{-rT}
$$

This shows that the stock price can be replicated by buying a call option, selling a put option, and holding a position in a T-bill (which will earn $X e^{-rT}$ at maturity).

2. Interpret the components

* $C$: The price of the call option gives you the right to buy the stock at the strike price.
* $P$: The price of the put option gives you the right to sell the stock at the strike price.
* $X e^{-rT}$: The present value of the strike price of the options (discounted at the risk-free rate).

### Construct the portfolio

To replicate the stock $S$, you would do the following:

1. Buy one call option ($C$): This gives you the right to buy the stock at the strike price $K$.
2. Sell one put option ($P$): This obligates you to buy the stock at the strike price $X$ if the option is exercised. In effect, this gives you a position that will pay off similarly to holding the stock if the stock price falls below $X$.
3. Hold $X e^{-rT}$ in a T-bill: This is the present value of the strike price, and it will grow to $X$ at maturity, which aligns with the strike price of the options.

### Intuitive explanation

* __Call option ($C$)__: By buying the call, you're acquiring the right to buy the stock at the strike price $X$, which is beneficial when the stock price rises above $X$.
* __Put option ($P$)__: By selling the put, you're agreeing to buy the stock at the strike price $X$ if the stock falls below $X$. This is beneficial when the stock price falls below $X$, effectively allowing you to own the stock at the strike price.
* T-bill ($X e^{-rT}$): This investment is a way to ensure that you can afford to buy the stock at the strike price when either the call is exercised or the put is assigned, regardless of whether the stock price moves up or down.

Thus, this portfolio is effectively a synthetic long position in the stock, and the payoff will mirror the price movement of the stock. If the stock price goes up, the call option becomes more valuable, and if the stock price goes down, the put option becomes more valuable. The T-bill ensures that you can fulfill your obligations (buying the stock at the strike price) at maturity.

### Conclusion

By suing put-call parity, you can replicate a share of stock by: Buying a call option, selling a put option and holding a T-bill with a value of $X e^{-rt}$, where $X$ is the strike price.

---

# Questions and problems

## Two-state option pricing model

T-bills currently yield 4.8 percent. Stock in Nina Manufacturing is currently selling for \$63 per share. There is no possibility that the stock will be worth less than \$61 per share in one year.

a. What is the value of a call option with a \$60 exercise price? What is the intrinsic value?
b. What is the value of a call option with a \$50 exercise price? What is the intrinsic value?
c. What is the value of a put option with a \$60 exercise price? What is the intrinsic value?

---

Given information:

* Current stock price $S = 63$
* T-bills yield: $r = 0.048$
* The stock will not fall below \$61 per share in one year.
* The options expire in one year.

### Formula for the value of call and put options

To value the call and put options, we'll need to use the principle of no-arbitrage pricing. For European-syle optons, this can be done by discounting the expected future payoff of the options.

However, since we're provided with a limited amount of information, we can use the following approach:

* __Intrinsic value__: This is the immediate value of exercising the option.

For a call option, the intrinsic value is $\text{max}(S_T - K, 0)$, where $S_T$ is the price of the stock at maturity and $K$ is the strike price.

For a put option, the intrinsic value is $\text{max}(K - S_T, 0)$.

We'll use the upper bound for the stock price, since we know that the stock won't fall below \$61. The stock could either stay at \$63 or potentially fall to \$61 at maturity.

a. Call option with a \$60 Exercise price

* __Intrinsic value:__ For a call option, the intrinsic value is the amount by which the stock price exceeds the strike price, or zero if the stock price is below the strike price.

At maturity, if the stock price remains \$63, then the intrinsic value is $63 - 60 = 3$. If the stock price falls to $S = 61$, then the intrinsic value is \$1.

The intrinsic value of the call option is the value in the best case, so the intrinsic value is 3 (since the stock price is expected to be at least \$61, and the stock could be worth \$63 at maturity).

* __Value of the call option:__ The value of the call option can be approximated using the expected payoff discounted at the risk-free rate.

Since the stock price is at least \$61, the expected future price will be either \$63 or \$61. Assuming a 50-50 chance of each outcome:

$$
\begin{aligned}
\text{Expected payoff} &= \frac{1}{2} \times (63 - 60) + \frac{1}{2} \times (61 - 60)\\
  &= \frac{1}{2} \times 3 + \frac{1}{2} \times 1 = 2
\end{aligned}
$$

Now, discounting this payoff back to the present value using the T-bill yield:

$$
\begin{aligned}
\text{Value of call option} &= \frac{2}{(1.048)}\\
  &\approx 1.91
\end{aligned}
$$

So, the value of the call option is approximately \$1.91.

b. Call option with a \$50 exercise price

* __Intrinsic value:__ For a call option with a \$50 strike price, the intrinsic value is \$13 or \$11, depending if the stock price at maturity is either \$63 or \$61 respectively.

The value of the call option is thus:

$$
\begin{aligned}
\frac{.5 \times 13 + .5 \times 11}{1.048}\\
  &\approx 11.44
\end{aligned}
$$

c. Put option with a \$60 exercise price

The intrinsic value for a put option is the amount by which the strike price exceeds the stock price, or zero if the strike price is lower than the stock price. Either 63 or 61 is higher than the strike price of 60. Thus, the intrinsic value is zero. For the put option to have any value, it would need to fall below 60. Because of the scenario given above that is not possible.

---

## Understanding option quotes

Use the option quote information shown here to answer the questions that follow. The stock is currently selling for \$83.

```{r problem 2: understanding option quotes, echo=F}
symbol <- "RWJ"
expiration <- c("March", "April", "July", "October")
strike.price <- rep(80, 4)
calls.vol <- c(230, 170, 139, 60)
calls.last <- c(2.80, 6, 8.05, 10.20)
puts.vol <- c(160, 127, 43, 11)
puts.last <- c(.80, 1.40, 3.90, 3.65)

df <- data.frame(expiration, strike.price, calls.vol, calls.last, puts.vol, puts.last)

colnames(df) <- c("Expiration", "Strike Price", "Calls Volume", "Calles Last", "Puts Volume", "Puts Last")

knitr::kable(df, caption="Problem 2 Option quote")
```

a. Are the call options in the money? What is the intrinsic value of an RWJ Corp. call option?
b. Are the put options in the money? What is the intrinsic value of an RWJ Corp. put option?
c. Two of the options are clearly mispriced. Which ones? At a minimum, what should the mispriced options sell for? Explain how you could profit from the mispricing in each case.

---

a. The call options are in the money

$$
\text{Intrinsic value} = 83 - 80 = 3
$$

b. The put options are not in the money, they are worth 0.

c. To identify mispriced options, we compare the last price of each option with its minimum theoretical price.

A call option should be worth at least its intrinsic value of 3. The March call is mispriced. It is selling for \$2.80 which is less than the \$3 intrinsic value.

A put option should never be priced below its intrinsic value, but since all put options have an intrinsic value of 0, they should at least be worth their time value. In this case the October put is mispriced. It is selling for \$3.65, but the July put is selling for \$3.90. The October put should be at least \$3.90 because longer-term options should be more expensive than shorter-term ones.

### Arbitrage on the underpriced March call

Buy the March call for \$2.80, immediately exercise it by buying the stock at \$80, and sell the stock at \$83. The risk-free profit is \$0.20 per share.

### Arbitrage on the underpriced October put

Buy the October put at 3.65 and sell the July put at 3.90 (a risk-free arbitrage known as a "calendar spread"). The profit is $3.90 - 3.65 = 0.25$ per share.

---

## Calculating payoffs

Use the option quote information shown here to answer the questions that follow. The stock is currently selling for \$114.

```{r problem 3: calculating payoffs, echo=F}
symbol <- "Macrosoft"
expiration <- c("February", "March", "May", "August")
strike.price <- rep(110, 4)
calls.vol <- c(85, 61, 22, 3)
calls.last <- c(7.60, 8.80, 10.25, 13.05)
puts.vol <- c(40, 22, 11, 3)
puts.last <- c(.60, 1.55, 2.85, 4.70)

df <- data.frame(expiration, strike.price, calls.vol, calls.last, puts.vol, puts.last)

colnames(df) <- c("Expiration", "Strike Price", "Calls Volume", "Calles Last", "Puts Volume", "Puts Last")

knitr::kable(df, caption="Problem 3 Option quote")
```

a. Suppose you buy 10 contracts of the February 110 call option. How much will you pay, ignoring commissions?
b. In part (a), suppose that Macrosoft stock is selling for \$140 per share on the expiration date. How much is your options investment worth? What if the terminal stock price is \$125? Explain.
c. Suppose you buy 10 contracts of the August 110 put option. What is your maximum gain? On the expiration date, Macrosoft is selling for \$104 per share. How much is your options investment worth? What is your net gain?
d. In part (c), suppose you _sell_ 10 of the August 110 put contracts. What is your net gain or loss if Macrosoft is selling for \$103 at expiration? For \$132? What is the break-even price---that is, the terminal stock price that results in a zero profit?

---

### Cost of buying 10 February 110 call contracts.

* 10 contracts purchases 100 shares each contract for a total of 1,000 shares.
* The last price is \$7.60.
* __Total cost:__ $1,000 \times 7.60 = 7,600$

### Value of call options at expiration

If the stock price is \$140 on expiration then the investment is worth $(140 - 110) \times 1,000 = 30,000$.

If the stock price is \$125 on expiration then the investment is worth $(125 - 110) \times 1,000 = 15,000$.

### Maximum gain and value of 10 August 110 put contracts

The maximum gain of 10 August 110 put options occuris if the stock price goes to \$0. The total value is
$10 \times 100 \times 110 = 110,000$. The cost is $10 \times 100 \times 4.70 = 4,700$. The net gain is
$110,000 - 4,700 = 105,300$.

If Macrosoft is selling for \$104 on expiration, then the value is $(110 - 104) \times 1,000 = 6,000$.
The net gain is $6,000 - 4,700 = 1,300$.

### Selling 10 August 110 put contracts

If the stock price is \$103 on expiration then there is a loss of $(110 - 103) \times 1,000 = 7,000$. The net gain is $4,700 - 7,000 = -2,300$.

If the stock price is \$132 on expiration then the option is out of the money and is worth zero. You keep \$4,700.

Break-even occurs when net profit is zero. The loss comes from the put being in the money: $110 - S_T = 4.70$. Solving for $S_T = 110 - 4.70 = 105.30$.

---

## Two-state option pricing model

The price of Ervin Corp. stock will be either \$74 or \$96 at the end of the year. Call options are available with one year to expiration. T-bills currently yield 5 percent.

a. Suppose the current price of Ervin stock is \$80. What is the value of the call option if the exercise price is \$70 per share?
b. Suppose the exercise price is \$90 in part (a). What is the value of the call option now?

---

Part A:

If the price is \$74 at expiration, the future payoff is: $74 - 70 = 4$.

If the future price is \$96 at expiration, then the future payoff is: $96 - 70 = 26$.

The swing of the call is $26 - 4 = 22$.

The swing of the stock is $96 - 74 = 22$.

$$
\begin{aligned}
D &= \frac{22}{22} = 1
\end{aligned}
$$

The amount to borrow is:

Buying a stock brings us either \$96 or \$74 at expiration. This is \$70 more than the payout.

$$
\begin{aligned}
B &= \frac{70}{(1.05)}\\
  &= 66.67
\end{aligned}
$$

If the stock price is currently \$80:

$$
\begin{aligned}
V &= 80 \times 1 - 66.67\\
  &= 13.33
\end{aligned}
$$

Part B:

If the exercise price is \$90 (instead of \$70). Then:

future payouts: $96 - 90 = 6$ or $74 - 90 < 0$.

The swing of the call is: 6. The swing of the stock is 22.

$$
\begin{aligned}
D &= \frac{6}{22}\\
  &= 0.27
\end{aligned}
$$

Buying 0.27 of a share will bring us either $.27 \times 96 \approx 26$ or $.27 \times 74 \approx 20$.
 This means we need to borrow the present value of \$20.

$$
\begin{aligned}
B &= \frac{20}{(1.05)}\\
  &= 19.05
\end{aligned}
$$

$$
\begin{aligned}
V &= 80 \times .27 - 19.05\\
  &= 2.55
\end{aligned}
$$

Book value 2.60

---

## Two-state option pricing model

The price of Tara, Inc., stock will be either \$50 or \$70 at the end of the year. Call options are available with one year to expiration. T-bills currently yield 5 percent.

a. Suppose the current price of Tara stock is \$62. What is the value of the call option if the exercise price is \$35 per share?
b. Suppose the exercise price is \$60 in part (a). What is the value of the call option now?

---

### Exercise price of \$35

Buying a call:

Payouts: $70 - 35 = 35$ or $50 - 35 = 15$

Call swing: $35 - 15 = 20$
Stock swing:  $70 - 50 = 20$

Delta: 1

Buying a Stock:

$1 \times 70 = 70$ or $1 \times 50 = 50$.

Excess of payoffs is: $70 - 35 = 35$ or $50 - 15 = 35$.

Borrowing $\frac{35}{(1.05)} = 33.33$

$C = 62 \times 1 - 33.33 = 28.67$

### Exercise price of \$60

Buying a call:

Payouts: $70 - 60 = 10$ or $50 - 70 < 0$

Call swing: $10 - 0 = 10$
Stock swing: $70 - 50 = 20$

Delta: $\frac{10}{20} = \frac{1}{2}$

Buying a stock

$\frac{1}{2} \times 70 = 35$ or $\frac{1}{2} \times 50 = 25$

Excess of payoffs is $35 - 10 = 25$ or $25 - 0 = 25$

Borrowing $\frac{25}{(1.05)} = 23.81$

$C = 62 \times \frac{1}{2} - 23.81 = 7.19$

---

## Put-call parity

A stock is currently selling for \$38 per share. A call option with an exercise price of \$40 sells for \$3.80 and expires in three months. If the risk-free rate of interest is 2.6 percent per year, coumpounded continuously, what is the price of a put option with the same exercise price?

---

Put-Call parity

$$
S + P = C + \text{PV}(X)
$$

$$
\begin{aligned}
P &= C + \text{PV}(X) - S\\
  &= 3.80 + 40 e^{-.026 \times .25} - 38\\
  &= 5.54
\end{aligned}
$$

---

## Put-call parity

A put option that expires in six months with an exercise price of \$65 sells for \$4.89. The stock is currently priced at \$61, and the risk-free rate is 3.6 percent per year, compounded continuously. What is the price of a call option with the same exercise price?

---

$$
\begin{aligned}
C &= S + P - \text{PV}(X)\\
  &= 61 + 4.89 - 65 e^{-.036 \times .5}\\
  &= 2.05
\end{aligned}
$$

---

## Put-call parity

A put option and a call option with an exercise price of \$55 expire in two months and sell for \$2.65 and \$5.32, respectively. If the stock is currently priced at \$57.30, what is the annual continuously compounded rate of interest?

---

$$
\begin{aligned}
C &= S + P - X e^{-rT}\\\\
5.32 &= 57.30 + 2.65 - 55 e^{-.1667r}\\\\
e^{-.1667r} &= \frac{54.63}{55}\\
e^{-.1667r} &= .9933\\
-.1667T &= -.0068\\
r &= \frac{.0068}{.1667}\\
  &= .0405
\end{aligned}
$$

---

## Black-Scholes

What are the prices of a call option and a put option with the following characteristics?

* Stock price: \$57
* Exercise price: \$60
* Risk-free rate: .06 compounded continuously
* Maturity: 3 months
* Standard deviation ($\sigma$) = .54 (Variance = $\sigma^2$)

---

$$
\begin{aligned}
C &= S N(d_1) - E e^{-rT} N(d_2)\\\\
d_1 &= \frac{\ln(S/E) + (R + \sigma^2/2) T}{\sqrt{\sigma^2 T}}\\
d_2 &= d1 - \sqrt{\sigma^2 T}
\end{aligned}
$$

We will use `R` to calculate the put and call option prices:

```{r problem 10, Black-Scholes}
S <- 57
E <- 60
r <- .06
T <- .25
SD <- .54
V <- SD^2

(d.1 <- (log(S/E) + (r + V/2) * T)/(sqrt(V*T)))
(d.2 <- d.1 - sqrt(V*T))
(round(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2),2))
(round(P <- C + E*exp(-r*T) - S,2))
```

---

## Black-Scholes

What are the prices of a call option and a put option with the following characteristics?

* Stock price: \$93
* Exercise price: \$90
* Risk-free rate: .04
* Maturity: 5 months
* Standard deviation: .62

---

We will use `R` to calculate the put and call option prices:

```{r problem 11, Black-Scholes}
S <- 93
E <- 90
r <- .04
T <- 5/12
SD <- .62
V <- SD^2

(d.1 <- (log(S/E) + (r + V/2) * T)/(sqrt(V*T)))
(d.2 <- d.1 - sqrt(V*T))
(round(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2),2))
(round(P <- C + E*exp(-r*T) - S,2))
```

---

## Delta

What are the deltas of a call option and a put option with the following characteristics? What does the delta of the option tell you?

* Stock price: \$67
* Exercise price: \$70
* Risk-free rate: .05
* Maturity: 9 months
* Standard deviation: .49

---

```{r, problem 12, Deltas}
S <- 67
E <- 70
r <- .05
T <- 9/12
V <- .49^2

(delta.call <- pnorm(d.1 <- (log(S/E) + (r + V/2) * T)/(sqrt(V*T))))
(delta.put <- delta.call - 1)
```

---

## Black-Scholes and asset value

You own a lot in Key West, Florida, that is currently unused. Similar lots have recently sold for \$1.1 million. Over the past five years, the price of land in the area has increased 12 percent per year, with an annual standard deviation of 25 percent. A buyer has recently approached you and wants an option to buy the land in the next 12 months for \$1.25 million. The risk-free rate of interest is 5 percent per year, compounded continuously. How much should you charge for the option?

---

* Price: \$1.1 million
* Exercise price: \$1.25 million
* Risk-fre rate: .05
* Maturity: 12 months
* Standard deviaion: .25

---

```{r, problem 13, Black-Scholes and asset value, call}
G <- .12
E <- 1250000
r <- .05
T <- 1
V <- .25^2
(S <- 1100000*c(1,exp(G*T)))

(d.1 <- (log((S/E) + (.05 + V/2) * T)/sqrt(V*T)))
(d.2 <- d.1 - sqrt(V*T))

(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
```

---

## Black-Scholes and asset value

In the previous problem, suppose you wanted the option to sell the land to the buyer in one year. Assuming all the facts are the same, describe the transaction that would occur today. What is the price of the transaction today?

---

```{r, problem 14, Black-Scholes and asset value, put}
(P <- C - S + E * exp(-r*T))
```

---

## Time value of options

You are given the following information concerningoptions on a particular stock:

* Stock price: \$83
* Exercise price: \$80
* Risk-free rate: 0.06
* Maturity: 6 months
* Standard deviation: .53

a. What is the intrinsic value of the call option? Of the put option?
b. What is the time value of the call option? Of the put option?
c. Does the call or put have the larger time value component? Would you expect this to be true in general?

---

### Intrinsic value calculation

The intrinsic value of an option is the value it would have if exercised immediately. It is given by:

Call: $\text{max}(S - X, 0)$

Put: $\text{max}(X - S, 0)$

$$
\begin{aligned}
\text{IV}_{\text{call}} &= \text{max}(83 - 80, 0)\\
  &= 3\\\\
\text{IV}_{\text{put}} &= \text{max}(80 - 83, 0)\\
  &= 0
\end{aligned}
$$

### Time value calculation

The time value of an option is:

$$
\text{Time value} = \text{option price} - \text{intrinsic value}
$$

Since the option price isn't given, we calculate the total option price using Black_Scholes and then subtract the intrinsic value to get the time value.

```{r problem 14, time value of options}
S <- 83
E <- 80
r <- .06
T <- 6/12
SD <- .53
V <- SD^2

(d.1 <- (log(S/E) + (r + V/2) * T)/sqrt(V*T))
(d.2 <- d.1 - sqrt(V*T))
(C <- S * pnorm(d.1) - E * exp(-r * T) * pnorm(d.2))
(P <- C - S + E * exp(-r*T))

cat("Call option time value: ", C - 3, "\n")
cat("Put option time value: ", P - 0, "\n")
```

### Which option has the larger time value?

The call option has a larger time value \$11.81 than the put option \$9.44. This is expected because the stock price \$83 is above the strike price \$80, making the call in the money and the put out of the money. In general, the time value is larger for options that are at the money or slightly in the money because they have a higher probability of expiring profitably.

---

## Risk-neutral valuation

A stock is currently priced at \$73. The stock will either increase or decrease by 15 percent over the next year. There is a call option on the stock with a strike price of \$70 and one year until expiration. If the risk-free rate is 8 percent, what is the risk-neutral value of the call option?

---

* Stock price: \$73
* Stock movement range: $73 \times .15 = 10.95$
* Exercise price: \$70
* Maturity: 1 year
* Risk-free rate: .08

### Calculate the Delta

* Stock movement range: (83.95, 62.05)
* Call payoff range: (83.95 - 70, 0) = (13.95, 0)

$$
\begin{aligned}
D &= \frac{\text{Swing of the call}}{\text{Swing of the stock}}\\
  &= \frac{13.95}{21.90}\\
  &= .637
\end{aligned}
$$

### Determine the borrowing amount

* Buying $.637 \times 83.95 = 53.48$ Stock
* Buying $.637 \times 62.05 = 39.53$ Stock

Buying .637 share of stock brings us either \$53.48 or \$39.53 at expiration:

* $53.48 - 13.95 = 39.53$ or $39.53 - 0 = 39.53$

We need to borrow \$39.53 at the rate of .08:

$$
B = \frac{39.53}{(1.08)} = 36.60
$$

The value of the call:

$$
\begin{aligned}
C &= S \times D - B\\
  &= 73 \times .637 - 36.60\\
  &= 9.90
\end{aligned}
$$

The risk-neutral probability is:

$$
\begin{aligned}
p &= \frac{(1 + r) - d}{(u - d)}\\
  &= \frac{(1 + .08) - (1 - .15)}{(1 + .15) - (1 - .15)}\\
  &= \frac{1.08 - .85}{1.15 - .85}\\
  &= \frac{.07}{.30}\\
  &= .767
\end{aligned}
$$

The risk-neutral value:

$$
\begin{aligned}
C_{\text{risk-neutral}} &= \frac{(.767) \times 13.95 + (1 - .767) \times 0}{(1.08)}\\
  &= \frac{10.70}{1.08}\\
  &= 9.91
\end{aligned}
$$

### Final answer

* The risk-nuetral probability: 0.767
* The value of the call is: \$9.90
* The expected value of the option at expiration is: \$10.70
* The present value of the call option (Risk-nutral value): \$9.91

---

## Risk-nuetral valuation

In the previous problem, assume the risk-free rate is only 5 percent. What is the risk-neutral value of the option now? What happens to the risk-neutral probabilities of a stock price increase and a stock price decrease?

---

The risk-neutral probability and value is now:

$$
\begin{aligned}
p &= \frac{1.05 - .85}{.3}\\
  &= .667\\\\
C_{\text{risk-neutral}} &= \frac{.667 \times 13.95}{1.05}\\
  &= \frac{9.3}{1.05}\\
  &= 8.86
\end{aligned}
$$

### Effect on risk-neutral probabilities

The risk-neutral probability is givent by:

$$
p = \frac{(1 + r) - d}{u - d}
$$

Notice that the risk-neutral probability formula does not directly depend on $S$. Instead, $p$ is determined by the up factor $u$, down factor $d$, and risk-free rate $r$.

### What does change?

If $S$ increases, the absolute stock price levels at $S_u$ and $S_d$ also increase. However, the relative movements (percentages $u$ and $d$) remain the same, keeping the risk-neutral probability unchanged. The actual probability of an upward or downward move in the real world may change, but the risk-neutral probability remains fixed for a given $u$, $d$, and $r$.

### When does the risk-neutral probability Change?

If $u$ or $d$ changes (e.g., due to increased volatility), then $p$ will change. If $r$ changes, $p$ will also adjust.

---

## Black-Scholes

A call option matures in six months. The underlying stock price is \$75, and the stock's return has a standard deviation of 30 percent per year. The risk-free rate is 4 percent per year, compounded continuously. If the exercise price is \$0, what is the price of the call option?

---

* T: .5
* S: 75
* SD: .3
* r: .04
* E: 0

```{r problem 18: black-scholes}
T <- .5
S <- 75
V <- .3^2
r <- .04
E <- 0

(d.1 <- (log(S/E) + (r + V/2) * T)/sqrt(V*T))
(d.2 <- d.1 - sqrt(V*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
```

A call option with an exercise price of \$0 means the option holder automatically receives the stock at expiration. It is equivalent to owning the stock itself. The price of the option should therefore be equal to the current stock price.

---

## Black-Scholes

A call option has an exercise price of \$80 and matures in six months. The current stock price is \$84, and the risk-free rate is 5 percent per year, compounded continuously. What is the price of the call if the standard deviation of the stock is 0 percent per year?

---


* T: .5
* S: 84
* SD: 0
* r: .04
* E: 80

```{r problem 19: black-scholes}
T <- .5
S <- 84
V <- 0
r <- .04
E <- 80

(d.1 <- (log(S/E) + (r + V/2) * T)/sqrt(V*T))
(d.2 <- d.1 - sqrt(V*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
```

A call option with zero volatility ($\sigma = 0$) means the stock price will not change before expiration. The stock will stay at \$84 with certainty. The call option will expire in the money because the stock price is $S = 84$ is greater than $X = 80$. At expiration, the option's value will be $S - X = 84 - 80 = 4$. Since there is no uncertainty, the present value of this payoff is simply the discounted value.

---

## Black-Scholes

A stock is currently priced at \$35. A call option with an expiration of one year has an exercise price of \$50. The risk-free rate is 7 percent per year, compounded continuously, and the standard deviation of the stock's return is infinitely large. What is the price of the call option?

---

```{r problem 20: Black-Scholes}
S <- 35
T <- 1
E <- 50
r <- .07
V <- 1e8

(d.1 <- (log(S/E) + (r + V/2) * T)/sqrt(V*T))
(d.2 <- d.1 - sqrt(V*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
```

When the standard deviation is infinity, this means the stock price can take any value at expiration. The call option behaves like a 50-50 lottery. There is a 50-50 chance the option will be above or below the exercise price. Therefore, the time value component of the call option does not matter and the call option is equal to the stock price.

---

## Equity as an option

Sunburn Sunscreen has a zero coupon bond issue outstanding with a \$25,000 face value that matures in one year. The current market value of the firm's assets is \$26,300. The standard deviation of the return on the firm's assets is 38 percent per year, and the annual risk-free rate is 5 percent per year, compounded continuously. Based on the Black-Scholes model, what is the market value of the firm's equity and debt?

---


The firm's value:

$$
V = B + S = 25,000 + 26,300 = 51,300
$$

* S: 26,300
* B: 25,000
* T: 1
* SD: .38
* r: .05

The firm has zero-coupon debt, meaning it must pay $D = 25,000$ in one year. The firm's equity holders will receive any value of the firm's assets above this amount. If the firm's assets are below 25,000, equity holders get nothing, and debt holders take over.

This means:

* Equity is a call option on the firm's assets with strike price $D = 25,000$.
* Debt is equivalent to a risk-free bond minus a short put option.

Using the Black-Scholes call option pricing model, we can determine the equity value.

```{r problem 21: equity as an option}
S <- 26300
E <- 25000
T <- 1
V <- .38^2
r <- .05

(d.1 <- (log(S/E) + (r + V/2)*T)/sqrt(V*T))
(d.2 <- d.1 - sqrt(V*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
(P <- S - C)
```

---

## Equity as an option and NPV

Suppose the firm in the previous problem is considering two mutually exclusive investments. Project A has an NPV of \$1,200, and Project B has an NPV of \$1,600. As a result of taking Project A, the standard deviation of the return on the firm's assets will increase to 55 percent per year. If Project B is taken, the standard deviation will fall to 34 percent per year.

a. What is the value of the firm's equity and debt if Project A is undertaken? If Project B is undertaken?
b. Which project would the stockholders prefer? Can you reconcile your answer with the NPV rule?
c. Suppose the stockholders and bondholders are, in fact, the same group of investors. Would this affect your answer to (b)?
d. What does this problem suggest to you about stockholder incentives?

---

From the previous problem:

* Market value of assets: $V = 26,300$
* Face value of debt (strike price): $D = 25,000$
* Risk-free rate: $r = .05$
* Time to maturity: $T = 1$ year
* Initial volatility: $\sigma = .38$

Now, if the firm undertakes:

* __Project A:__ The NPV increases by \$1,200, and volatility increases to .55
* __Project B:__ The NPV increases by \$1,600, but volatility decreases to .34

The firm's new asset values will be:

* __Project A:__ $V_A = 26,300 + 1,200 = 27,500, \sigma_A = .55$
* __Project B:__ $V_B = 26,300 + 1,600 = 27,900, \sigma_B = .34$

### Compute equity and debt values

$$
\begin{aligned}
d_1 &= (ln{V/D} + (r + \sigma^2/2)T)/\sqrt{\sigma^2 T}\\
d_2 &= d_1 - \sqrt{\sigma^2 T}\\\\
C &= V N(d_1) - D e^{-rT} N(d_2)\\
D &= V - C
\end{aligned}
$$

```{r problem 22: equity as an option and npv: part a}
V.A <- 27500
SD.A <- .55
V.B <- 27900
SD.B <- .34
D <- 25000
T <- 1
r <- .05

d.1 <- function(V, D, r, s, T) (log(V/D) + (r + s^2/2)*T)/sqrt(s^2*T)
d.2 <- function(d1, s, T) d1 - sqrt(s^2*T)
C <- function(V, d1, D, r, T, d2) V * pnorm(d1) - D * exp(-r*T) * pnorm(d2)

(d.1.A <- d.1(V.A, D, r, SD.A, T))
(d.2.A <- d.2(d.1.A, SD.A, T))
(C.A <- C(V.A, d.1.A, D, r, T, d.2.A))
(D.A <- V.A - C.A)

(d.1.B <- d.1(V.B, D, r, SD.B, T))
(d.2.B <- d.2(d.1.B, SD.B, T))
(C.B <- C(V.B, d.1.B, D, r, T, d.2.B))
(D.B <- V.B - C.B)
```

Book answers:

* Project A: Equity = \$7,603.04, Debt = \$19,896.96
* Project B: Equity = \$5,921.30, Debt = \$21,978.70


### Stockholders preference

Stockholders own equity, which behaves like a call option. They prefer higher volatility ($\sigma$), because: Higher risk increases option value (equity value goes up). Debt remains fixed at $D = 25,000$, but equity holders capture upside potential. Project A increases the volatility to 55\% which makes the equity more valuable. Project B decreases volatility to 34\%, which benefits bondholders (less risk of default), but equity loses some of its value. Despite Project B having higher NPV, stockholders may prefer Project A due to its higher risk.

This shows a conflict between the NPV rule and stockholders incentives. The NPV rule suggests choosing the project with the higher dollar value increase (Project B), but stockholders may rationally prefer the riskier Project A.

### Impact if stockholders and bondholders are the same group

If stockholders and bondholders are the same group of investors, then the goal would be to maximize total firm value (assets) rather than just equity. In this case, investors would likely prefer Project B, since it maximizes firm value.

### Stockholders incentives

Stockholders have an incentive to take on riskier projects because they hold an option on the firm's assets. This is an agency problem since stockholders may choose riskier projects that benefit them at the expense of bondholders. Bondholders anticipate this and demand higher interest rates (credit spreads) r impose covenants to limit risk-taking.

### Conclusion

* If maximizing firm value: Project B is better.
* If maximizing stockholder value: Project A might be preferred due to higher equity value.
* If stockholders and bondholders are the same: Project B will be chosen.
* This problem illustrates the risk-shifting conflict between stockholders and bondholders in corporate finance.

---

## Equity as an option

Frostbite Thermalwear has a zero coupon bond issue outstanding with a face value of \$30,000 that matures in one year. The current market value of the firm's assets is \$36,400. The standard deviation of the return on the firm's assets is 53 percent per year, and the annual risk-free rate is 5 percent per year, compounded continuously. Based on the Black-Scholes model, what is the market value of the firm's equity and debt? What is the firm's continuously compounded cost of debt?

---

* Bond face value: \$30,000
* Market value of firm: \$36,400
* Standard deviation on the return on the firm's assets: .53
* Risk-free rate: .05

```{r problem 23: equity as an option}
E <- 30000
S <- 36400
SD <- .53
r <- .05
T <- 1

(d.1 <- (log(S/E) + (r + SD^2/2)*T)/sqrt(SD^2*T))
(d.2 <- d.1 - sqrt(SD^2*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
(P <- S - C)
```

### Continuously compounded cost of debt

$$
\begin{aligned}
P &= E e^{-yT}\\
\ln{P/E} &= -yT\\
y &= -\frac{\ln{P/E}}{T}\\
y &= -\frac{\ln{\frac{25,000}{30,000}}}{1}
  &= .1823
\end{aligned}
$$

---

## Mergers and equity as an option

Suppose Sunburn Sunscreen and Frostbite Thermalwear in the previous problems have decided to merge. Because the two companies have seasonal sales, the combined firm's return on assets will have a standard deviation of 29 percent per year.

a. What is the combined value of equity in the two existing companies? The value of debt?
b. What is the value of the new firm's equity? The value of debt?
c. What was the gain or loss for shareholders? For bondholders?
d. What happened to shareholder value here?

---

### Sunburn Sunscreen

* Market value of assets: \$26,300
* Face value of debt: \$25,000
* Debt maturity: 1 year
* Standard deviation of asset return: .38
* risk-free rate: .05

* Market value of equity: \$5,162.98
* Market value of debt: \$21,137.02

### Frostbite Thermalwear

* Market value of assets: \$36,400
* Face value of debt: \$30,000
* Debt maturity: 1 year
* Standard deviation of asset return: .53
* risk-free rate: .05

* Market value of assets: 11,399.73
* Market value of debt: 25,000.27

### Post merger outlook

* Combined assets: \$62,700
* Combined debt: \$55,000
* Debt maturity: 1 year
* Standard deviation: .29
* risk-free rate: .05

* Combined market value of assets: $5,162.98 + 11,399.73 = 16,562.71$
* Combined market value of debt: $21,137.02 + 25,000.27 = 46,137.29$

```{r problem 24: mergers and equity as an option}
S <- 62700
E <- 55000
T <- 1
SD <- .29
r <- .05

(d.1 <- (log(S/E) + (r + SD^2/2)*T)/sqrt(SD^2*T))
(d.2 <- d.1 - sqrt(SD^2*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
(P <- S - C)
```

The post merger gains:

* There is a combined loss in market value for equity $\$13,059.79 - \$16,562.71 = -\$3,502.92$
* The combined gain gain in market value for debt $\$49,640.21 - \$46,137.29 = 3,502.92$

As can be seen the value of the merger is bad for the stockholders. \$3,502.92 has been transfered from the stockholders to the bondholders.

---

## Equity as an option and NPV

A company has a single zero coupon bond outstanding that matures in 10 years with a face value of \$15 million. The current value of the company's assets is \$13.4 million, and the standard deviation of the return on the firm's assets is 39 percent per year. The risk-free rate is 6 percent per year, compounded continuously.

a. What is the current market value of the company's equity?
b. What is the current market value of the company's debt?
c. What is the company's continuously compounded cost of debt?
d. The company has a new project available. The project has an NPV of \$1,200,000. If the company undertakes the project, what will be the new market value of equity? Assume volatility is unchanged.
e. Assuming the company undertakes the new project and does not borrow any additional funds, what is the new continuousy compounded cost of debt? What is happening here?

---

* Face value of debt: \$15,000,000
* Time to maturity: 10 years
* Value of assets: \$13,400,000
* Standard deviation: .39
* risk-free rate: .06

```{r problem 25: equity as an option and NPV}
S <- 13400000
E <- 15000000
T <- 10
SD <- .39
r <- .06

(d.1 <- (log(S/E) + (r + SD^2/2)*T)/sqrt(SD^2*T))
(d.2 <- d.1 - sqrt(SD^2*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
(P <- S - C)
```

* Market value of equity: `r round(C/10^6, 6)` million
* Market value of the debt: `r round(P/10^6, 6)` million

The cost of debt:

$$
\begin{aligned}
D &= E \times e^{-yT}\\
\ln{D/E} &= -yT\\\\
y &= - \frac{\ln{D/E}}{T}\\
  &= - \frac{\ln{5,482,533/15,000,000}}{10}\\
  &= `r -(log(P/E)/T)`
\end{aligned}
$$

If the company takes on a new project with an NPV of \$1,200,000. The new market value of equity can be calculated.

```{r problem 25: equity as an option and NPV: adding a project}
(S <- S + 1200000)

(d.1 <- (log(S/E) + (r + SD^2/2)*T)/sqrt(SD^2*T))
(d.2 <- d.1 - sqrt(SD^2*T))
(C <- S * pnorm(d.1) - E * exp(-r*T) * pnorm(d.2))
(P <- S - C)
(y <- -log(P/E)/T)
```

* The new market value of the equity: `r round(C/10^6,6)` million.
* The new market value of the debt: `r round(P/10^6,6)` million.

---

## Two-state option pricing model

Ken is interested in buying a European call option written on Southeastern Airlines, Inc., a non-dividend-paying common stock, with a strike price of \$75 and one year until expiration. Currently, Southeastern's stock sells for \$78 per share. In one year Ken knows that Southeastern's stock will be trading at either \$93 per share or \$65 per share. Ken is able to borrow and lend at the risk-free EAR of 2.5 percent.

a. What should the call option sell for today?
b. If no options currently trade on the stock, is there a way to create a synthetic call option with identical payoffs to the call option just described? If there is, how would you do it?
c. How much does the synthetic call option cost? Is this greater than, less than, or equal to what the actual call option costs? Does this make sense?

---

* S: 78
* E: 75
* In one year the stock will be trading either: 93, or 65.
* risk-free EAR: .025
* T: 1 year

* Swing of call: $93 - 75 = 18$, $18 - 0 = 18$
* Swing of the stock: $93 - 65 = 28$

$$
\begin{aligned}
r_f &= \ln{(1 + r_{\text{EAR}} \times T)}\\
  &= \ln{(1.025 \times 1)}\\
  &= 0.025
D &= \frac{18}{28}\\
  &= .6429\\\\
\text{Buy .6429 of stock}\\
S_{93} &= .6429 \times 93\\
  &= 59.79\\
S_{65} &= .6429 \times 65\\
  &= 41.79\\\\
B &= \frac{41.79}{(1.025)}\\
  &= 40.77\\\\
C &= 78 \times .6429 - 40.77\\
  &= 9.38
\end{aligned}
$$

To create a synthetic option what you would do is buy .6429 share of the stock and borrow 40.77 at .025 EAR.

The cost of the synthetic call option should be the same as the actual call option. In an arbitrage-free market, the cost of the synthetic call option must equal the cost of the actual call option. If there were a difference, an arbitrageur could create a riskless profit by buying the cheaper option and selling the more expensive option.

Book answer:

a. \$9.38
b. .64; Borrow \$40.77
c. \$9.38

---

## Two-state option pricing model

Rob wishes to buy a European put option on BioLabs, Inc., a non-dividend-paying common stock, with a strike price of \$40 and six months until expiration. BioLabs' common stock is currently selling for \$30 per share, and Rob expects that the stock price will either rise to \$60 or fall to \$15 in six months. Rob can borrow and lend at the risk-free EAR of 5 percent.

a. What should the put option sell for today?
b. If no options currently trade on the stock, is there a way to create a synthetic put option with identical payoffs to the put option just described? If there is, how would you do it?
c. How much does the synthetic put option cost? Is this greater than, less than, or equal to what the actual put option costs? Does this make sense?

---

* S: 30
* E: 40
* T: 6/12 = .5
* In six months: 60 or 15
* r (EAR): .05 (we need to convert this to the real rate $r_f = \ln{(1 + r_{\text{EAR}} \times T)}$).

* Swing of call: $60 - 40 = 20$, $20 - 0 = 20$
* Swing of the stock: $60 - 15 = 45$

$$
\begin{aligned}
r_f &= \ln{(1 + r_{\text{EAR}} \times T)}\\
  & \ln{(1 + .05 \times .5)}\\
  &= \ln{(1.0125)}\\
  &= .025\\\\
D &= \frac{20}{45}\\
  &= .4444\\\\
\text{Buy .4444 of stock}\\
S_{60} &= .4444 \times 60\\
  &= 26.67\\
S_{15} &= .4444 \times 15\\
  &= 6.67\\\\
B &= \frac{6.67}{1.025}\\
  &= 6.59\\\\
C &= 30 \times .4444 - 6.59\\
  &= 6.51\\\\
P &= C - S + E \times e^{-.025*.5}\\
  &= 6.51 - 30 + 40 \times .9876\\
  &= 16.01
\end{aligned}
$$

---

## Two-state option pricing model

Maverick Manufacturing, Inc., must purchase gold in three months for use in its operations. Maverick's management has estimated that if the price of gold were to rise above \$1,530 per ounce, the firm would go bankrupt. The current price of gold is \$1,450 per ounce. The firm's chief financial officer believes that the price of gold will either rise to \$1,605 per ounce or fall to \$1,340 per ounce over the next three months. Management wishes to eliminate any risk of the firm going bankrupt. Maverick can borrow and lend at the risk-free EAR of 6.50 percent.

a. Should the company buy a call option or a put option on gold? To avoid bankruptcy, what strike price and time to expiration would the company like this option to have?
b. How much should such an option sell for in the open market?
c. If no options currently trade on gold, is there a way for the company to create a synthetic option with identical payoffs to the option just described? If there is, how would the firm do it?
d. How much does the synthetic option cost? Is this greater than less than, or equal to what the actual option costs? Does this make sense?

---

### Type of option

The company needs to protect itself against the price of gold rising. Therefore, it should buy a call option.

* S: 1,450
* E: 1,530 (strike price)
* up move (u): 1,605
* down move (d): 1,340
* r (EAR): .065
* T: .25

* Call option swing: $1,605 - 1,530 = 75$, $75 - 0 = 75$
* Price swing: $1,605 - 1,340 = 255$

$$
\begin{aligned}
r_f &= \ln{(1 + r_{\text{EAR}} \times T)}\\
  &= \ln{(1.0163)}\\
  &= 0.016\\\\
D &= \frac{75}{255}\\
  &= .2941\\\\
S_{\text{up}} &= .2941 \times 1,605\\
  &= 472.06\\
S_{\text{down}} &= .2941 \times 1,340\\
  &= 394.09\\
B &= \frac{394.09}{1.016}\\
  &= 370.04\\
C &= 1,450 \times .2941 - 370.04\\
  &= 56.41
\end{aligned}
$$

I would buy the put.
