---
title: "Financial Statements and Cash Flow"
author: "Ross"
date: "2024-04-26"
output: 
  pdf_document:
   toc: true 
---
# Summary and Conclusions

Besides introducing you to corporate accounting, the purpose of this chapter has been to teach you how to determine cash flow from the accounting statements of a typical company.

1. Cash flow is generated by the firm and paid to creditors and shareholders. It can be classified as:

  a. Cash flow from operations.
  b. Cash flow from changes in fixed assets.
  c. Cash flow from changes in working capital.

2. Calculations of cash flow are not difficult, but they require care and particular attention to detail in properly accounting for noncash expenses such as deprecation and deferred taxes. It is especially important that you do not confuse cash flow with changes in net working capital and net income.

# Concept Questions

1. __Liquidity__ True or false: All assets are liquid at some price. Explain.
2. __Accounting and Cash Flows__ Why might the revenue and cost figures shown on a standard income statement not represent the actual cash inflows and outflows that occurred during a period?
3. __Accounting Statement of Cash Flows__ Looking at the accounting statement of cash flows, what does the bottom line number mean? How useful is this number for analyzing a company?
4. __Cash Flows__ How do financial cash flows and the accounting statement of cash flows differ? Which is more useful for analyzing a company?
5. __Book Values versus Market Values__ Under standard accounting rules, it is possible for a company's liabilities to exceed its assets. When this occurs, the owners' equity is negative. Can this happen with market values? Why or why not?
6. __Cash Flow from Assets__ Why is it not necessarily bad for the cash flow from assets to be negative for a particular period?
7. __Operating Cash Flow__ Why is it not necessarily bad for the operating cash flow to be negative for a particular period?
8. __Net Working Capital and Capital Spending__ Could a company's change in net working capital be negative in a given year? (_Hint:_ Yes.) Explain how this might come about. What about net capital spending?
9. __Cash Flow to Stockholders and Creditors__ Could a company's cash flow to stockholders be negative in a given year? (_Hint:_ Yes.) Explain how this might come about. What about cash flow to creditors?
10. __Firm values__ Referring back to the Bank of America example at the beginning of the chapter, note that we suggested that Bank of America's stockholders probably didn't suffer as a result of the reported loss. What do you think was the basis for our conclusion?

# Questions and Problems

1. __Building a Balance Sheet__ Bishop, Inc., has current assets of \$5,700, net fixed assets of \$27,000, current liabilities of \$4,400, and long-term debt of \$12,900. What is the value of the shareholders' equity account for this firm? How much is net working capital?

$$
\begin{aligned}
\text{Assets} & = \text{Current Assets} + \text{Fixed Assets} = \$5,700 + \$27,000 = \$32,700 \\
\text{Liabilities} & = \text{Current Liabilities} - \text{Long-term liabilities} = \$4,400 + \$12,900 = \$17,300 \\
\text{Shareholders equity} & = \text{Assets} - \text{Liabilities} = \$32,700 - \$16,900 = \$15,400
\end{aligned}
$$

```{r}
current.assets <- 5700
net.fixed.assets <- 27000
current.liabilities <- 4400
long.term.debt <- 12900
assets <- current.assets + net.fixed.assets
liabilities <- current.liabilities + long.term.debt
se <- assets - liabilities
nwc <- current.assets - current.liabilities
cat("Shareholders' equity: $", se, "\n",
"Net working capital: $", nwc, "\n")
```

---

2. __Building an Income Statement__ Travis, Inc., has sales of \$387,000, costs of \$175,000, deprecation expense of \$40,000, interest expense of \$21,000, and a tax rate of 35%. What is the net income for the firm? Suppose the company paid out \$30,000 in cash dividends. What is the addition to retained earnings?

$$
\begin{aligned}
\text{EBIT} & = \text{Sales} - \text{Costs} - \text{Deprecation Expense} = 387 - 175 - 40 = 172 \\
\text{Gross Income} & = \text{EBIT} - \text{Interest Expense} = 172 - 21 = 151 \\
\text{Taxes Owed} & = \text{Gross Income} \times \text{Tax Rate} = 151 \times .35 = 52.85 \\
\text{Net Income} & = \text{Gross Income} - \text{Taxes Owed} = 151 - 52.85 = 98.15 \\
\text{Retained Earnings} & = \text{Net Income} - \text{Dividends} = 98.15 - 30 = 68.15
\end{aligned}
$$

```{r}
sales <- 387
costs <- 175
deprecation.expense <- 40
interest.expense <- 21
tax.rate <- .35
dividends <- 30
ebit <- sales - costs - deprecation.expense
pretax.income <- ebit - interest.expense 
taxes <- pretax.income*tax.rate
net.income <- pretax.income - taxes
retained.earnings <- net.income - dividends
cat("Sales:", sales, "\n",
    "Costs:", -costs, "\n",
"Depreciation:", deprecation.expense, "\n",
"Interest:", interest.expense, "\n",
"Gross:", pretax.income, "\n",
"Taxes:", -taxes, "\n",
"Net Income: ", net.income, "\n",
"Dividends Payed:", dividends, "\n",
"Retained Earnings (after dividends): ", retained.earnings, "\n")
```

---

3. __Market Values and Book Values__ Klingon Cruisers, Inc., purchased new cloaking machinery three years ago for \$9.5 million. The machinery can be sold to the Romulans today for \$6.5 million. Klingon's current balance sheet shows net fixed assets of \$5.2 million, current liabilities of \$2.4 million, and net working capital of \$800,000. If all the current assets were liquidated today, the company would receive \$2.6 million in cash. What is the book value of Klingon's assets today? What is the market value?

```{r echo=F}
cloaking.machine.market <- 6.5
net.fixed.assets <- 5.2
current.liabilities <- 2.4
net.working.capital <- .8
current.assets.market <- 2.6
```

Step 1: Calculate the book value of all assets.

```{r}
(current.assets <- net.working.capital + current.liabilities)
(assets.book <- current.assets + net.fixed.assets)
```

Step 2: Calculate the market value of all assets.

```{r}
(assets.market <- cloaking.machine.market + current.assets.market)
```

---

4. __Calculating Taxes__ The Locker Co. has \$273 thousand in taxable income. Using the rates from Table 2.3 in the chapter, calculate the company's income taxes. What is the average tax rate? What is the marginal tax rate?

```{r, echo=F}
table.2.3 <- data.frame(
  c(0, 50001, 75001, 100001, 335001, 10000001, 15000001, 18333334),
  c(50000, 75000, 100000, 335000, 10000000, 15000000, 18333333, 1E20),
  c(.15, .25, .34, .39, .34, .35, .38, .35)
)
colnames(table.2.3) <- c("Lower", "Upper", "Tax.Rate")
print(table.2.3)
income_tax <- function(income, brackets=table.2.3$Upper, rate=table.2.3$Tax.Rate){
  sum(diff(c(0, pmin(income, brackets))) * rate)
}
```

```{r}
taxable.income <- 273000
income.tax <- income_tax(taxable.income)
avg.rate <- round(income.tax/taxable.income*100)
marginal <- table.2.3$Tax.Rate[taxable.income < table.2.3$Upper][1]*100
cat("Average Tax Rate:", avg.rate, "%\n",
    "Marginal Tax Rate:", marginal, "%")
```

---

5. __Calculating OCF__ Ranney, Inc., has sales of \$18,700, costs of \$10,300, deprecation expense of \$1,900, and interest expense of \$1,250. If the tax rate is 40 percent, what is the operating cash flow, or OCF?

__Solution__

Operating Cash Flow:

$$
\begin{aligned}
\text{Gross Income} &= \text{sales} - \text{cost} - \text{deprecation} - \text{interest}
  = \$18,700 - \$10,300 - \$1,900 - \$1,250 = \$5,250 \\
\text{Taxes} &= \text{Gross Income} \times \text{Tax Rate} = \$5,250 \times .40 = \$2,100 \\
\text{Net Income} &= \text{Gross Income} - \text{Taxes} = \$5,250 - \$2,100 = \$3,150 \\
\text{Operating Cash Flow} &= \text{Net Income} + \text{deprecation} + \text{interest}
  = \$3,150 + \$1,900 + \$1,250 = \$6,300
\end{aligned}
$$

```{r}
sales <- 18700
costs <- 10300
deprecation <- 1900
interest <- 1250
tax <- .40

cat(" EBIT:", (ebit <- sum(c(sales,-costs, -deprecation))), "\n",
"Taxable income:", (gross.income <- sum(c(ebit, -interest))), "\n",
"Taxes:", (taxes <- gross.income*tax), "\n",
"Net Income:", (net.income <- gross.income - taxes), "\n",
"Operating Cash Flow (OCF):", (ocf <- sum(c(ebit, deprecation, -taxes))))
```

---

6. __Calculate Net Capital Spending__ Gordon Driving School's 2011 balance sheet showed net fixed assets of \$1.42 million, and the 2012 balance sheet showed net fixed assets of \$1.69 million. The company's 2012 income statement showed a deprecation expense of \$145,000. What was Gordon's net capital spending for 2012?

```{r}
fixed.assets.2011 <- 1.42
fixed.assets.2012 <- 1.69
deprecation.2012 <- .145

fixed.assets.2012 - fixed.assets.2011 + deprecation.2012
```

---

7. __Building a Balance Sheet__ The following table presents the long-term liabilities and stockholders' equity of Information Control Corp. one year ago:

```{r echo=F}
long.term.debt <- 65
preferred.stock <- 4
common.stock.1.par.value <- 15
accumulated.retained.earnings <- 135
capital.surplus <- 45
tb <- matrix(c(long.term.debt, preferred.stock, common.stock.1.par.value, accumulated.retained.earnings, capital.surplus), ncol=1)
rownames(tb) <- c("Long-term debt", "Preferred stock", "Common stock ($1 par value)",
                "Accumulated retained earnings", "Capital Surplus")
colnames(tb) <- c("")
print(tb)
```

During the past year, Information Control issued 10 million shares of new stock at a total price of \$58 million, and issued \$35 million in new long-term debt. The company generated \$9 million in net income and paid \$2 million in dividends. Construct the current balance sheet reflecting the changes that occurred at Information Control Corp. during the year.

```{r}
tb["Long-term debt", ] <- tb["Long-term debt", ] + 35
tb["Common stock ($1 par value)", ] <- tb["Common stock ($1 par value)", ] + 10
tb["Accumulated retained earnings", ] <- tb["Accumulated retained earnings", ] + 9 - 2
tb["Capital Surplus",] <- tb["Capital Surplus", ] + (58 - 10)
tb
```

---

8. __Cash Flow to Creditors__ The 2011 balance sheet of Anna's Tennis Shop, Inc., showed long-term debt of \$1.45 million, and the 2012 balance sheet showed long-term debt of \$1.52 million. The 2012 income statement showed an interest expense of \$127,000. What was the firm's cash flow to creditors during 2012?

```{r}
long.term.debt.2011 <- 1.45
long.term.debt.2012 <- 1.52
interest.payment <- .127
(cf.creditors <- interest.payment - (long.term.debt.2012 - long.term.debt.2011))
```

---

9. __Cash Flow to Stockholders__ The 2011 balance sheet of Anna's Tennis Shop, Inc., showed \$490,000 in the common stock account and \$3.4 million in the additional paid-in surplus account. The 2012 balance sheet showed \$525,000 and \$3.7 million in the same two accounts, respectively. If the company paid out \$275,000 in cash dividends during 2012, what was the cash flow to stockholders for the year?

```{r}
common.stock.2011 <- .490
surplus.2011 <- 3.4
common.stock.2012 <- .525
surplus.2012 <- 3.7
dividends.2012 <- .275
(cf.stockholders <- dividends.2012 -(common.stock.2012 + surplus.2012 - common.stock.2011 - surplus.2011))
```

---

10. __Calculating Cash Flows__ Given the information for Anna's Tennis Shop, Inc., in the previous two problems, suppose you also know that the firm's net capital spending for 2012 was \$945,000 and that the firm reduced its net working capital investment by \$87,000. What was the firm's 2012 operating cash flow, or OCF?

$$
\text{Cash Flow from Assets (CFFA)} = \text{Cash Flow to Creditors (CFC)} + \text{Cash Flow to Stockholders (CFS)}
$$

and

$$
\text{CFFA} = \text{Operating Cash Flow (OCF)} - \text{Net Capital Spending (NCS)}
  - \text{Change in Net Working Capital (CNWC)}
$$

$$
\text{OCF} = \text{CFFA} + \text{NCS} + \text{CNWC}
$$

then

$$
\text{OCF} = \text{CFC} + \text{CFS} + \text{NCS} + \text{CNWC}
$$


```{r}
ncs <- .945
nwci <- .087
(cfc <- cf.creditors)
(cfs <- cf.stockholders)
(cffa <- cfc + cfs)
(ocf <- cffa + ncs - nwci)
```

---

11. __Cash Flows__ Ritter Corporation's accountants prepared the following financial statements for year-end 2012:

  a. Explain the change in cash during 2012.
  b. Determine the change in net working capital in 2012.
  c. Determine the cash flow generated by the firm's assets during 2012.
  
```{r, echo=F}
income.statement <- matrix(c(750,565,90,95,75),ncol=1)
rownames(income.statement) <- c("Revenue", "Expenses", "Deprecation", "Net income", "Dividends")
balance.sheets <- matrix(c(65, 170, 390, 625, 125, 145, 355, 625, 55, 165, 370, 590, 115, 140, 335, 590), ncol=2)
rownames(balance.sheets) <- c("Cash", "Other current assets", "Net fixed assets",
                              "Total assets", "Accounts payable", "Long-term debt",
                              "Stockholders equity", "Total liabilities and equity")
colnames(balance.sheets) <- c("2012", "2011")
knitr::kable(income.statement, caption="Income Statement 2012")
knitr::kable(balance.sheets, caption="Balance Sheets December 31")
```

__Solution__

a. The cash balance has increased from \$55,000 in 2011 to \$65,000. This increase in cash could result from a combination of operating, investing, and financing activities, including cash generated from operations (net income), changes in working capital, and any new financing or debt payments.

b. The change in net working capital is Current Assets $-$ Current Liabilities. For 2012

$$
\text{NWC} = \text{(Cash + Other Current Assets)} - \text{Accounts Payable} \\
\text{NWC}_{2012} = (65 + 170) - 125 = 235 - 125 = 110 \\
\text{NWC}_{2011} = (55 + 165) - 115 = 220 - 115 = 105 \\
\text{Change in NWC} = \text{NWC}_{2012} - \text{NWC}_{2011} =  110 - 105 = 5
$$

c. The cash flow from assets:

$$
\text{CFFA} = \text{Operating Cash Flow} - \text{Net Capital Spending} - \text{Change in NWC}
$$

Calculate the Operating Cash Flow

$$
\text{OCF} = \text{New Income} + \text{Deprecation} = 95 + 90 = 185
$$

Calculate Net Capital Spending

$$
\text{NCS} = \text{Net Fixed Assets 2012} - \text{Net Fixed Assets 2011} + \text{Deprecation} = 390 - 370 + 90 = 110
$$

The Cash flow from assets

$$
\text{CFFA} = \text{OCF} - \text{NCS} - \text{Change in NWC} = 185 - 110 - 5 = 70
$$

---

12. __Financial Cash Flows__ The Stancil Corporation provided the following current information:

```{r echo=F}
tb <- matrix(c(17,4,21,1.9,14.5), ncol=1)
rownames(tb) <- c("Proceeds from long-term borrowing",
                  "Proceeds from the sale of common stock",
                  "Purchases of fixed assets",
                  "Purchases of inventories",
                  "Payment of dividends")
knitr::kable(tb)
```

Determine the cash flows from the firm and the cash flows to investors of the firm.

__Solution__

## Cash Flows:

### Inflows:

* Long term borrowing: \$17
* Proceeds from sale of stock: \$4

### Outflows:

* Purchase of fixed assets: \$21
* Purchase of inventories: \$14.5

Net cash flow from the firm is:

$$
\$17 + \$4 - \$21 - \$14.5 = - \$14.5
$$


$\$14,500 - \$4,000 = \$10,500$

---

13. __Building an Income Statement__
















