---
title: "Financial Statements and Cash Flow"
author: "Ross"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  pdf_document:
   toc: true 
---

# Summary

Besides introducing you to corporate accounting, the purpose of this chapter has been to teach you how to determine cash flow from the accounting statements of a typical company.

1. Cash flow is generated by the firm and paid to creditors and shareholders. It can be classified as:

>  a. Cash flow from operations.
>  b. Cash flow from changes in fixed assets.
>  c. Cash flow from changes in working capital.

2. Calculations of cash flow are not difficult, but they require care and particular attention to detail in properly accounting for non-cash expenses such as deprecation and deferred taxes. It is especially important that you do not confuse cash flow with changes in net working capital and net income.

# Concepts

##Liquidity

True or false: All assets are liquid at some price. Explain.

__True__. All assets are liquid at some price. However, the degree of liquidity and the ability to sell an asset varies. Assets are usually classified based on its liquidity. _Highly liquid assets_ are assets that can be converted to cash very quickly and with minimal loss to value. _Less liquid assets_ cannot be easily converted to cash without significant loss to value.

# Accounting and cash flows

Why might the revenue and cost figures shown on a standard income statement not represent the actual cash inflows and outflows that occurred during a period?

There are several reasons why a standard income statement does not represent the actual inflows and outflows that occur during the accounting period.

### Accrual accounting

Income statements are typically prepared using the _accrual basis of accounting_, which recognizes revenue when it is earned and expenses when they are incurred, regardless of when cash is actually received or paid.

### Non-cash expenses

Income statements include non-cash expenses such as _deprecation_, _deferred taxes_ and _amortization_. These items reduce reported profits but do not include any cash outflows.

### Timing differences

There may be timing differences between when cash is paid/received and when the corresponding revenue/expense is recognized.

### Financing activities excluded

Cash flows related to financing activities, such as borrowing money, issuing stock, or repaying debt, do not appear on the income statement but do impact actual cash flow.

### Capital expenditures omitted

The purchase of long-term assets, called _capital expenditures_, represents significant cash outflows but is not recorded as an expense on the income statement. Instead, the cost is allocated over time through depreciation.

### Changes in working capital

Changes in current assets and liabilities, such as accounts receivable, accounts payable, and inventory, affect cash flows but are not directly captured on the income statement.

### Summary

The income statement provides a measure of profitability based on accounting principles, but actual cash inflows and outflows are better captured in the _cash flow statement_, which reconciles net income with cash changes by adjusting for non-cash items, timing differences, and financing/capital activities.

# Accounting Statement of Cash Flows

Looking at the accounting statement of cash flows, what does the bottom line number mean? How useful is this number for analyzing a company?

The _bottom line number_ on the accounting statement of cash flows represents the _net change in cash_ during the reporting period. It is the difference between the company's beginning and ending cash balances and is derived by summing the cash flows from the three main activities:

1. __Operating Activities__: Cash generated or used in the company's core business operations.
2. __Investing Activities__: Cash spent on or received from the acquisition or disposal of long-term assets.
3. __Financing Activities__: Cash inflows and outflows from transactions with shareholders and creditors, such as issuing stock, borrowing, or repaying debt.

A positive net change in cash indicates that the company has increased its cash reserves during the period, while a negative net change in cash shows that the company has used more cash than it generated during the period.

The bottom line number has several advantages:

1. __Liquidity Insight__: It provides a snapshot of whether the company is generating or consuming cash, which is crucial for meeting short-term obligations.
2. __Cash Management__: Helps assess the company's ability to maintain sufficient cash reserves for operations, investments, and debt payments.
3. __Trend Analysis__: By comparing changes in cash over time, you can observe trends in cash generation and usage.

However, the bottom line number has some limitations:

1. __Not a measure of profitability__: The bottom line only shows cash movement, not profitability. A profitable company might have negative cash flows if it's investing heavily. On the opposite end, a low profitable company may have a high positive cash flow due to a one-time event, such as the sale of an asset.
2. __Lacks Detailed Insight__: A positive or negative number doesn't explain the cause. To understand it fully, you must analyze the cash flow from _operating_, _investing_, and _financing activities_ individually.
3. __Doesn't Address Sustainability__: A company could increase cash through unsustainable methods (e.g, excessive borrowing or asset sales).

### Summary

The bottom line number on the statement of cash flows is a v aluable indicator of a company's liquidity and financial flexibility, but it must be analyzed alongside other financial metrics and details from the cash flow statement to provide meaningful insights into the company's financial health and sustainability.


# Cash Flows

How do financial cash flows and the accounting statement of cash flows differ? Which is more useful for analyzing a company?

## Financial Cash Flows

Focus on the cash available to investors (shareholders and creditors).

$$
\begin{aligned}
\mathbf{Financial\ cash\ flow} &= \text{cash flow from assets} \\
  &= \text{operating cash flow} - \text{net capital spending}
  - \text{change in net working capital}
\end{aligned}
$$

- Cash generated from the company's operations (OCF).
- Reinvestment in long-term assets (capital spending).
- Changes in working capital requirements.
- Net cash flows available to distribute to creditors and shareholders.

The purpose of financial cash flows is to help assess the company's ability to generate cash to meet obligations and provide returns to investors.

Financial cash flows strip out non-cash items and accounting adjustments and provides a clear view of how much cash is available for debt service and shareholder returns.

## Account Statement of Cash Flows

Summarizing actual cash inflows and outflows from three activities: _operating_, _investing_, and _financing_.

1. __Operating Activities__: Cash from day-to-day business operations.
2. __Investing Activities__: Cash from buying/selling assets.
3. __Financing Activities__: Cash from debt and equity transactions.

- Tracks the company's cash position and liquidity.
- Matches the accounting framework (GAAP or IFRS).

The purpose of the accounting statement of cash flows is to ensure transparency in financial reporting and provides detailed insights into where cash is coming from and going to.

Accounting statement of cash flows offer a comprehensive and detailed breakdown of cash movements and useful for identifying specific sources of cash flow issues (e.g., excessive capital spending, operaitonal inefficiencies).

## Which is more useful for analyzing a company?

Financial cash flows are more useful for investors assessing the company's capacity to return cash and determining the sustainability of growth and profitability. The account statement of cash flows is more useful for assessing liquidity and operational health, and for understanding cash management practicies of the firm.

## Summary

The accounting statement of cash flows is more comprehensive for understanding short-term liquidity and detailed cash movements. However, financial cash flows provide a clearer picture of the company's ability to generate value for investors. Both are valuable tools, and their usefulness depends on the specific analysis being performed. Ideally, they should be used together for a well-rounded assessment.

# Book Values versus Market Values

Under standard accounting rules, it is possible for a company's liabilities to exceed its assets. When this occurs, the owners' equity is negative. Can this happen with market values? Why or why not?

Under standard accounting rules, liabilities can exceed assets, leading to negative owners' equity. However, this is unlikely to occur with market values.

## Accounting perspective

Accounting values for assets are recorded at historical cost minus deprecation, which may not reflect their current market value. If liabilities exceed the book value of assets, equity becomes negative.

## Market value perspective

Market value reflects investors' perception of the company's value based on future earnings potential, assets' current market value, and overall market sentiment.

The market value of equity is derived as:

$$
\mathbf{Market\ Value\ of\ Equity} = \text{Market Value of Assets}
  - \text{Market Value of Liabilities}
$$

Investors rarely assign a negative market value to equity because shareholders have limited liability; they are not personally responsible for debt beyond their initial investment. If liabilities exceed assets, the market value of equity may approach zero, but not go negative.

Therefore, there are three reasons why market values of firms do not go negative:

1. __Limited Liability__: Shareholders cannot owe more than the value of their equity investment. The worst-case scenario is that equity becomes worhtless.
2. __Future Expectations__: The market often values a company based on its potential future cash flows, even if it is currently in financial distress. This "hope value" can keep the equity market value above zero.
3. __Liquidation Value of Assets__: Even in bankruptcy, the company's assets may retain some liquidation value, reducing the gap between liabilities and assets.

# Cash Flow from Assets

Why is it not necessarily bad for the cash flow from assets to be negative for a particular period?

There are several reasons why cash flows from assets may be negative for a particular period and they do not indicate any real cause for concern:

1. A company may be heavily investing in capital expenditures (e.g., purchasing new equipment, expanding facilities, or acquiring assets) to support future growth.
2. Increasing net working capital can cause negative CFFA, such as when a retailer builds up inventory ahead of a major sale season.
3. One time costs can cause negative CFFA. These could be restructuring costs or costs of acquisitions.

Negative CFFA can be a problem if:

1. If negative CFFA occurs consistently without accompanying growth in revenue or profitability could indicate operational inefficiencies or excessive spending.
2. If the firm has weak cash reserves, leading to liquidity issues and distress.
3. If the firm made poor investment decisions that did not generate sufficient returns.

When assessing the implication of negative CFFA, consider:

1. The stage of the company. Is it a startup in a high growth environment. Then the negative CFFA could be because it is heavily reinvesting. Just make sure there are no liquidity issues.
2. A company may finance negative CFFA through equity or debt issuance is leveraging future cash flows. This can be sustainable if the returns on investment exceed the cost of capital.
3. Negative CFFA is acceptable if it leads to future cash flows.

# Operating Cash Flow

Why is it not necessarily bad for the operating cash flow to be negative for a particular period?

There are several reasons why negative operating cash flow (OCF) may not be concerning:

1. Investment for growth
2. Seasonal fluctuations of the business cycles may produce expected negative operating cash flows.
3. Strategic investments, such as in inventory or prepaying expenses may lead to temporary negative operating cash flow.
4. Broader economic or industry specific events, such as a recession, can lead to short-term negative operating cash flow.

If negative cash flows are consistent or if the company is in financial distress then there is a reason to be concerned.

# Net Working Capital and Capital Spending

Could a company's change in net working capital be negative in a given year? (_Hint:_ Yes.) Explain how this might come about. What about net capital spending?

## Net Working Capital

Net working capital is calculated as:

$$
\mathbf{Net\ Working\ Capital} = \text{Current Assets} - \text{Current Liabilities}
$$

A negative net working capital means that the current liabilities have increased more than the current assets (or its current assets decreased more than its current liabilities). This can occur because:

### Decreasing Current Assets

- By company collecting accounts receivable faster, reducing outstanding balances.
- Inventory levels are reduced, due to efficient inventory management or lower production. 
- The company uses cash reserves to pay off liabilities or fund operations.

### Increasing Current Liabilities

- The company delays paying accounts payable to suppliers, increasing its short-term obligations.
- The company takes on more short-term debt, such as notes payable or credit line, to finance operations or growth.

### Operations Strategy

- A company might deliberately reduce NWC to free up cash for investments or other uses. For example, by managing receivables or extending payables, the company improves cash flow.

## Net Capital Spending

Net capital spending can be calculated as:

$$
\mathbf{Net\ Capital\ Spending} = \text{Ending Net Fixed Assets}
  - \text{Beginning Net Fixed Assets} + \text{Deprecation}
$$

Negative net capital spending might occur when:

1. The company sells fixed assets (e.g., property, equipment) to generate cash, often during restructuring or downsizing.
2. The company chooses not to purchase significant new fixed assets, possibly due to a focus on efficiency, reduced growth plans, or financial constraints.
3. Fixed asset purchases are smaller than the deprecation expense for the period, resulting in a net reduction in the value of fixed assets.

## Summary

Negative NWC change is often due to increased liabilities or decreased current assets, which can be strategic to improve cash flow. Negative Net capital spending is often selling off more fixed assets than it is investing in, which may be part of a restructuring or a cash-raising efforts.

# Cash Flow to Stockholders and Creditors

Could a company's cash flow to stockholders be negative in a given year? (_Hint:_ Yes.) Explain how this might come about. What about cash flow to creditors?

## Cash flow to stockholders (CFS)

Cash flow to stockholders is calculated as:

$$
\mathbf{Cash\ flow\ to\ stockholders} = \text{dividends paid} - \text{net new equity raised}
$$

Negative cash flow to stockholders can occur because:

1. Issuance of new equity for growth
2. Reduction in dividends

## Cash flow to creditors (CFC)

Cash flow to creditors is calculated as:

$$
\mathbf{Cash\ flow\ to\ creditors} = \text{interest paid} - \text{net new borrowing}
$$

Negative cash flow to creditors could occur because:

1. Issuance of new debt to finance growth
2. refinancing or restructuring debt with new terms resulting in net borrowing that exceeds interest payments (replacing short-term debt with long-term bonds to reduce interest costs over time).
3. The firm has a low interest burden relative to the new debt that it takes on, especially during periods of low-interest rates.

## Key considerations

Negative cash flow to stockholders:

- May signal growth or strategic equity financing.
- Can also indicate reduced shareholder returns in the short term.

Negative cash flow to creditors:

- often reflects the company leveraging debt for investment or expansion.
- However, excessive borrowing migh indicate rising financial risk.

# Firm values

Referring back to the Bank of America example at the beginning of the chapter, note that we suggested that Bank of America's stockholders probably didn't suffer as a result of the reported loss. What do you think was the basis for our conclusion?

The basis for the conclusion that _Bank of America's stockholders probably didn't suffer as a result of the reported write-offs_ is rooted in the distinction between _accounting losses_ (such as write-offs) and _actual cash loses_. Here are the key points that explain why stockholders likely didn't experience significant losses:

### Write-offs are non-cash expenses

A _write-off_ is essentially an accounting adjustment that reduces the book value of an asset, typically when it is deemed uncollectible or impaired (e.g., bad loans or the declining value of assets). However, _write-offs_ are _non-cash_ transactions. They don't represent an actual outflow of cash. In the case of Bank of America's \$34 billion write-off for bad loands, the company wasn't actually paying out that \$34 billion in cash. Instead, they were acknowledging that the value of certain assets (loans) had declined.

### Write-offs clean up the balance sheet

While write-offs reduce the company's reported _net income_ and asset values, they also _clean up the balance sheet_ by removing impaired assets. This allows the company to more accurately reflect its true financial position going forward. As a result, these write-offs can make the company's financial statements more transparent, which could actually increase investor confidence.

### Write-offs can reflect conservative financial management

Large write-offs, especially during economic downturns, can be viewed as a _prudent_ measure by management to prepare for future losses. This reflects that the company is taking aciton to acknowledge and deal with problems early. Investors might view this as _responsible management_ rather than something to fear.

### Future earnings and cash flow are unaffected

__Cash flow__---which is the actual movement of money in and out of the company---is unaffected by write-offs. Even though write-offs might lead to reported losses, they do not impact the company's ability to generate cash in the future. If the bank continues to make loans and earn interest income, or if it finds new profitable opportunities, stockholders could still benefit in the long term.

### Market reaction may be more about the future

Investors typically care more about _future earnings potential_ than about past write-offs. The stock price of a company like Bank of America would likely reflect investors' expectations for future performance, rather than simply reacting to the past losses reported in the income statement.

### Broader economic and regulatory context

During period of economic stress, like the 2008--2009 financial crisis, large financial institutions often face widespread asset impairments. These write-offs, while large in nominal terms, are somewhat expected by investors who understand the risks in the banking industry. Additionally, government interventions (such as bailouts or stimulus packages) often provide a safety net, further cushioning stockholder losses.

<!--

**********************  Basic Questions ******************************

-->

# Basic Questions

## Building a balance sheet

Bishop, Inc., has current assets of \$5,700, net fixed assets of \$27,000, current liabilities of \$4,400, and long-term debt of \$12,900. What is the value of the shareholders' equity account for this firm? How much is net working capital?

### Solution

Here you use the _accounting definition_ of the balance sheet: Assets $\equiv$ Liabilities + Stockholders' equity

First assets by combining _current assets_ and _fixed assets_:

$$
\begin{aligned}
\mathbf{Assets} &= \text{Current Assets} + \text{Fixed Assets} \\
  &= 5,700 + 27,000 = 32,700
\end{aligned}
$$

Then combine liabilities by combining _current liabilities_ and _long-term liabilities_:

$$
\begin{aligned}
\mathbf{Liabilities} &= \text{Current Liabilities} - \text{Long-term liabilities} \\
  &= 4,400 + 12,900 = 17,300
\end{aligned}
$$

Finally reorder the equation to find the _shareholders equity_

$$
\begin{aligned}
\mathbf{Shareholders\ equity} &= \text{Assets} - \text{Liabilities} \\
  &= 32,700 - 16,900 = 15,400
\end{aligned}
$$

## Building an income statement

Travis, Inc., has sales of \$387,000, costs of \$175,000, deprecation expense of \$40,000, interest expense of \$21,000, and a tax rate of 35%. What is the net income for the firm? Suppose the company paid out \$30,000 in cash dividends. What is the addition to retained earnings?

### Solution

Building the _income statement_ we need to find the following:

$$
\begin{aligned}
\mathbf{Gross\ income} &= \text{sales} - \text{costs} \\ 
  &= 387,000 - 175,000 \\
  &= 212,000 \\ \\
\mathbf{EBIT} &= \text{Gross Income} - \text{Deprecation Expense} \\
  &= 387,000 - 175,000 - 40,000 \\
  &= 172,000 \\ \\
\mathbf{EBT} &= \text{EBIT} - \text{Interest Expense} \\
  &= 172,000 - 21,000 \\
  &= 151,000 \\ \\
\mathbf{Taxes\ owed} &= \text{EBT} \times \text{Tax Rate} \\
  &= 151,000 \times .35 \\
  &= 52,850 \\ \\
\mathbf{Net\ income} &= \text{EBT} - \text{Taxes Owed} \\
  &= 151,000 - 52,850 \\
  &= 98,150 \\ \\
\mathbf{Retained\ earnings} &= \text{Net Income} - \text{Dividends} \\
  &= 98,150 - 30,000 \\
  &= 68,150
\end{aligned}
$$

## Market values and book values

Klingon Cruisers, Inc., purchased new cloaking machinery three years ago for \$9.5 million. The machinery can be sold to the Romulans today for \$6.5 million. Klingon's current balance sheet shows net fixed assets of \$5.2 million, current liabilities of \$2.4 million, and net working capital of \$800,000. If all the current assets were liquidated today, the company would receive \$2.6 million in cash. What is the book value of Klingon's assets today? What is the market value?

### Solution

Klingon Cruisers, Inc. purchase of the cloaking machinery was three years ago so its deprecated value is already represented in the net fix assets of the current balance sheet. Calculating the book value of Klingon's assets and comparing it to the market value of those assets can be done by:

Step 1: Calculate the current assets of Klingon Cruisers Inc:

$$
\begin{aligned}
\mathbf{Current\ assets} &= \text{net working capital} + \text{current liabilities} \\
  &= 800,000 + 2,400,000 = 3,200,000 \\
\end{aligned}
$$

Step 2: Now calculate the book value of all assets:

$$
\begin{aligned}
\mathbf{Book\ value\ of\ assets} &= \text{current assets} + \text{net fixed assets} \\
  &= 3,200,000 + 5,200,000 = 8,400,000
\end{aligned}
$$

Step 3: Calculate the market value of all assets.

$$
\begin{aligned}
\mathbf{Market\ value\ of\ assets} &= \text{market value of cloaking machine} +
  \text{market value of all current assets} \\
  &= 6,500,000 + 2,600,000 = 9,100,000
\end{aligned}
$$

#### Solution

in `R`

```{r problem_3}
cloaking.machine.market <- 6.5
net.fixed.assets <- 5.2
current.liabilities <- 2.4
net.working.capital <- .8
current.assets.market <- 2.6
current.assets <- net.working.capital + current.liabilities
assets.book <- current.assets + net.fixed.assets
assets.market <- cloaking.machine.market + current.assets.market
```

```{r problem_3_output, echo=F}
table.p3 <- matrix(c(assets.book, assets.market), ncol=1)
rownames(table.p3) <- c("Assets Book Value", "Asset Market Value")
knitr::kable(table.p3, caption="Klingon Cruisers Inc. Market Value of Assets")
```

## Calculating Taxes

The Locker Co. has \$273 thousand in taxable income. Using the rates from the following Table, calculate the company's income taxes. What is the average tax rate? What is the marginal tax rate?

```{r corporate tax table, echo=F}
options(scipen=9999999)
table.tax.rates <- data.frame(
  c(0, 50001, 75001, 100001, 335001, 10000001, 15000001, 18333334),
  c(50000, 75000, 100000, 335000, 10000000, 15000000, 18333333, 1E20),
  c(.15, .25, .34, .39, .34, .35, .38, .35)
)
colnames(table.tax.rates) <- c("Lower", "Upper", "Tax.Rate")
knitr::kable(table.tax.rates, caption="Corporate Tax Rates")
```

### Solution

We use the table to sum up all the marginal tax rates:

$$
\begin{aligned}
50,000 \times 15\% &= \ 7,500 \\
75,000 - 50,000 \times 25\% &= \ 6,250 \\
100,000 - 75,000 \times 34\% &= \ 8,500 \\
273,000 - 100,000 \times 39\% &= \ 67,470 \\ \\
\mathbf{Average\ tax\ rate} &= 7,500 + 6,250 + 8,500 + 67,470 = 89,720
\end{aligned}
$$

#### Solution

in `R`

The following function will use the tax brackets from the tax rates table to calculate income taxes and find the marginal tax rate.

```{r income tax function}
income.tax <- function(income, brackets=table.tax.rates$Upper,
                       rate=table.tax.rates$Tax.Rate){
  sum(diff(c(0, pmin(income, brackets))) * rate)
}

marginal.tax.rate <- function(income, brackets=table.tax.rates$Upper,
                              rate=table.tax.rates$Tax.Rate){
  rate[income < brackets][1]*100
}
```

Using this function we can calculate the income tax, the average tax rate and the marginal tax rate.

```{r problem 4}
taxable.income <- 273000
income_tax <- income.tax(taxable.income)
avg.rate <- round(income_tax/taxable.income*100)
marginal <- marginal.tax.rate(taxable.income)
```

```{r problem 4 output, echo=F}
table.p4 <- matrix(c(avg.rate, marginal), ncol=1)
rownames(table.p4) <- c("Average Tax Rate", "Marginal Tax Rate")
knitr::kable(table.p4, caption="Average and marginal tax rates (percents)")
```

## Calculating OCF

Ranney, Inc., has sales of \$18,700, costs of \$10,300, deprecation expense of \$1,900, and interest expense of \$1,250. If the tax rate is 40 percent, what is the operating cash flow, or OCF?

### Solution

Operating Cash Flow:

$$
\begin{aligned}
\mathbf{Gross\ income} &= \text{sales} - \text{cost} \\
  &= 18,700 - 10,300 \\
  &= 8,400 \\ \\
\mathbf{EBIT} &= \text{gross income} - \text{deprecation} \\
  &= 8,400 - 1,900 \\
  &= 6,500 \\ \\
\mathbf{EBT} &= \text{EBIT} - \text{interest} \\
  &= 6,500 - 1,250 \\
  &= 5,250 \\ \\
\mathbf{Taxes} &= \text{EBT} \times \text{tax rate} \\
  &= 5,250 \times .40 \\
  &= 2,100 \\ \\
\mathbf{Net\ Income} &= \text{EBT} - \text{taxes} \\
  &= 5,250 - 2,100 \\
  &= 3,150 \\ \\
\mathbf{Operating\ cash\ flow} &= \text{net income} + \text{deprecation}
  + \text{interest} \\
  &= 3,150 + 1,900 + 1,250 \\
  &= 6,300
\end{aligned}
$$

## Calculate Net Capital Spending

Gordon Driving School's 2011 balance sheet showed net fixed assets of \$1.42 million, and the 2012 balance sheet showed net fixed assets of \$1.69 million. The company's 2012 income statement showed a deprecation expense of \$145,000. What was Gordon's net capital spending for 2012?

## Solution

Net capital spending is is also called _changes in fixed assets_. The net change
in fixed assets equals the acquisition of fixed assets minus the sales of fixed
assets. However, net capital spending can also be calculated.

$$
\begin{aligned}
\mathbf{Capital\ spending} &= \text{ending net fixed assets}
  - \text{beginning net fixed assets} + \text{deprecation} \\
  &= 1,690,000 - 1,420,000 + 145,000 \\
  &= 415,000
\end{aligned}
$$

## Building a Balance Sheet

The following table presents the long-term liabilities and stockholders' equity of Information Control Corp. one year ago:

```{r echo=F}
long.term.debt <- 65
preferred.stock <- 4
common.stock.1.par.value <- 15
accumulated.retained.earnings <- 135
capital.surplus <- 45
tb <- matrix(c(long.term.debt, preferred.stock, common.stock.1.par.value, accumulated.retained.earnings, capital.surplus), ncol=1)
rownames(tb) <- c("Long-term debt", "Preferred stock", "Common stock ($1 par value)",
                "Accumulated retained earnings", "Capital Surplus")
colnames(tb) <- c("")
knitr::kable(tb, caption="In millions of dollars")
```

During the past year, Information Control issued 10 million shares of new stock at a total price of \$58 million, and issued \$35 million in new long-term debt. The company generated \$9 million in net income and paid \$2 million in dividends. Construct the current balance sheet reflecting the changes that occurred at Information Control Corp. during the year.

### Solution

Updating the balance sheet:

$$
\begin{aligned}
\mathbf{Long-term\ debt} &= 65,000,000 + 35,000,000 \\
  &= 100,000,000 \\ \\
\mathbf{Preferred\ stock}\ \text{(no change)} &= 4,000,000 \\ \\
\mathbf{Common\ stock (1\ par\ value)} &= 15,000,000 + 10,000,000 \\
  &= 25,000,000\\ \\
\mathbf{Accumulated\ retained\ earnings} &= 135,000,000 + 9,000,000
  - 2,000,000 \\
  &= 142,000,000 \\ \\
\mathbf{Capital\ surplus} &= 45,000,000 + (58,000,000 - 10,000,000) \\
  &= 93,000,000
\end{aligned}
$$

## Cash Flow to Creditors

The 2011 balance sheet of Anna's Tennis Shop, Inc., showed long-term debt of \$1.45 million, and the 2012 balance sheet showed long-term debt of \$1.52 million. The 2012 income statement showed an interest expense of \$127,000. What was the firm's cash flow to creditors during 2012?

### Solution

$$
\begin{aligned}
\mathbf{Cash\ Flow\ to\ Creditors\ (CFC)} &= \text{interest expense}
  - (\text{ending long-term debt} - \text{beginning long-term debt}) \\
  &= 127,000 - (1,520,000 - 1,450,000) \\
  &= 57,000
\end{aligned}
$$

#### Solution

in `R`

```{r problem 8}
long.term.debt.2011 <- 1.45
long.term.debt.2012 <- 1.52
interest.payment <- .127
cf.creditors <- interest.payment - (long.term.debt.2012 - long.term.debt.2011)
```

```{r problem 8 output, echo=F}
output <- matrix(cf.creditors, ncol=1)
rownames(output) <- "Cash flow to creditors"
knitr::kable(output*1000, caption="In thousands of dollars")
```

## Cash Flow to Stockholders

The 2011 balance sheet of Anna's Tennis Shop, Inc., showed \$490,000 in the common stock account and \$3.4 million in the additional paid-in surplus account. The 2012 balance sheet showed \$525,000 and \$3.7 million in the same two accounts, respectively. If the company paid out \$275,000 in cash dividends during 2012, what was the cash flow to stockholders for the year?

### Solution

$$
\begin{aligned}
\mathbf{Cash\ Flow\ to\ Stockholders\ (CFS)} &= \text{dividends}
    - (\text{Ending Equity} - \text{Beginning Equity}) \\
  &= 275,000 - \left[(3,700,000 + 275,000)
    - (3,400,000 + 490,000)\right] \\
  &= - 60,000
\end{aligned}
$$

#### Solution

in `R`

```{r problem 9}
common.stock.2011 <- .490
surplus.2011 <- 3.4
common.stock.2012 <- .525
surplus.2012 <- 3.7
dividends.2012 <- .275
cf.stockholders <- dividends.2012 -
    (common.stock.2012 + surplus.2012 - common.stock.2011 - surplus.2011)
```

```{r problem 9 output, echo=F}
table.p9 <- matrix(cf.stockholders*1000, ncol=1)
rownames(table.p9) <- ("Cash flow to stockholders")
knitr::kable(table.p9, caption="In thousands of dollars")
```

## Calculating Cash Flows

Given the information for Anna's Tennis Shop, Inc., in the previous two problems, suppose you also know that the firm's net capital spending for 2012 was \$945,000 and that the firm reduced its net working capital investment by \$87,000. What was the firm's 2012 operating cash flow, or OCF?

### Solution

$$
\begin{aligned}
\mathbf{CFFA} &= \text{CFC} + \text{CFS} \\
  &= 57,000 + (- 60,000) \\
  &= -3,000 \\ \\
\mathbf{OCF} &= \text{CFFA} - \text{NCS} - \text{CNWC} \\
 &= (- 3,000) - 945,000 - (- 87,000) \\
 &= 855,000
\end{aligned}
$$

#### Solution

in `R`

```{r problem 10}
ncs <- .945
nwci <- -.087
cfc <- cf.creditors
cfs <- cf.stockholders
cffa <- cfc + cfs
ocf <- cffa + ncs + nwci
```

```{r problem 10 output, echo=F}
table.p10 <- matrix(ocf, ncol=1)
rownames(table.p10) <- c("Operating Cash Flow")
knitr::kable(table.p10)
```

<!--

**********************  Intermediate Questions  *************************

-->

# Intermediate Questions

## Cash Flows

Ritter Corporation's accountants prepared the following financial statements for year-end 2012:

  a. Explain the change in cash during 2012.
  b. Determine the change in net working capital in 2012.
  c. Determine the cash flow generated by the firm's assets during 2012.
  
```{r problem 11, echo=F}
income.statement <- matrix(c(750,565,90,95,75),ncol=1)
rownames(income.statement) <- c("Revenue", "Expenses", "Deprecation", "Net income", "Dividends")
balance.sheets <- matrix(c(65, 170, 390, 625, 125, 145, 355, 625, 55, 165, 370, 590, 115, 140, 335, 590), ncol=2)
rownames(balance.sheets) <- c("Cash", "Other current assets", "Net fixed assets",
                              "Total assets", "Accounts payable", "Long-term debt",
                              "Stockholders equity", "Total liabilities and equity")
colnames(balance.sheets) <- c("2012", "2011")
knitr::kable(income.statement, caption="Income Statement 2012")
knitr::kable(balance.sheets, caption="Balance Sheets December 31")
```


## Solution

a. The cash balance has increased from \$55 in 2011 to \$65. This increase in cash could result from a combination of operating, investing, and financing activities, including cash generated from operations (net income), changes in working capital, and any new financing or debt payments.

$$
\mathbf{Change\ in\ Cash} = 65 - 55 = 10
$$

b. The change in net working capital is current assets - current liabilities. For 2012

$$
\begin{aligned}
\mathbf{NWC} = \text{(cash + other current assets)} - \text{accounts payable} \\
\mathbf{NWC_{2012}} = (65 + 170) - 125 = 235 - 125 = 110 \\
\mathbf{NWC_{2011}} = (55 + 165) - 115 = 220 - 115 = 105 \\
\mathbf{Change in NWC} = \text{NWC}_{2012} - \text{NWC}_{2011} =  110 - 105 = 5
\end{aligned}
$$

c. The cash flow from assets:

$$
\begin{aligned}
\mathbf{CFFA} = \text{Operating Cash Flow} - \text{Net Capital Spending}
  - \text{Change in NWC} 
\end{aligned}
$$

Calculate the Operating Cash Flow

$$
\begin{aligned}
\mathbf{OCF} &= \text{New Income} + \text{Deprecation} \\
  &= 95 + 90 \\
  &= 185
\end{aligned}
$$

Calculate Net Capital Spending

$$
\begin{aligned}
\mathbf{NCS} &= \text{Net Fixed Assets 2012} - \text{Net Fixed Assets 2011}
  + \text{Deprecation} \\
  &= 390 - 370 + 90 \\
  &= 110
\end{aligned}
$$


The Cash flow from assets

$$
\begin{aligned}
\mathbf{CFFA} &= \text{OCF} - \text{NCS} - \text{Change in NWC} \\
  &= 185 - 110 - 5 \\
  &= 70
\end{aligned}
$$

## Financial Cash Flows

The Stancil Corporation provided the following current information:

```{r problem 12, echo=F}
tb <- matrix(c(17000,4000,21000,1900,14500), ncol=1)
rownames(tb) <- c("Proceeds from long-term borrowing",
                  "Proceeds from the sale of common stock",
                  "Purchases of fixed assets",
                  "Purchases of inventories",
                  "Payment of dividends")
knitr::kable(tb)
```

Determine the cash flows from the firm and the cash flows to investors of the firm.

### Solution

### Cash Flows from the firm

$$
\begin{aligned}
\mathbf{Cash\ flows\ from\ the\ firm} &= \text{Purchases of fixed assets}
  + \text{Purchases of invinetories} \\
  &= 21,000 + 1,900 \\
  &= 22,900
\end{aligned}
$$

### Cash Flows to investors

We consider creditors as investors.

$$
\begin{aligned}
\mathbf{Cash\ flows\ to\ investors}
  &= \text{Payment of dividends} \\
  &\ \ \ \ \ - \text{Proceeds from long term borrowing}  \\
  &\ \ \ \ \ - \text{Proceeds from the sale of common stock} \\
  &= 14,500 - 17000 - 4000 \\
  &= - 6,500
\end{aligned}
$$

## Building an Income Statement

During the year, the Senbet Discount Tire Company has gross sales of \$1.06 million. The firm's cost of goods sold and selling expenses were \$525,000 and \$215,000, respectively. Senbet also has notes payable of \$800,000. These notes carried an interest rate of 7 percent. Deprecation was \$130,000. Senbet's tax rate was 35 percent.

a. What was Senbet's net income?
b. What was Senbet's operating cash flow?

### Solution

a. Net Income

$$
\begin{aligned}
\mathbf{Gross\ income} &= \text{gross sales} - \text{cost of goods sold}
  - \text{selling expenses} \\
  &= 1,060,000 - 525,000 - 215,000 \\
  &= 320,000 \\ \\
\mathbf{EBIT} &= \text{gross income} - \text{deprecation} \\
  &= 320,000 - 130,000 \\
  &= 190,000 \\ \\
\mathbf{Interest\ payments} &= \text{notes payable} \times \text{interest rate} \\
  &= 800,000 \times .07 \\
  &= 56,000 \\ \\
\mathbf{EBT} &= \text{EBIT} - \text{interest payments} \\
  &= 190,000 - 56,000 \\
  &= 134,000 \\ \\
\mathbf{Taxes} &= \text{EBT} \times \text{tax rate} \\
  &= 134000 \times .35 \\
  &= 46,900 \\ \\
\mathbf{Net\ income} &= \text{EBT} - \text{taxes} \\
  &= 134,000 - 46,900 \\
  &= 87,100 \\ \\
\end{aligned}
$$

b. Operating Cash Flow

$$
\begin{aligned}
\mathbf{OCF} &= \text{EBIT} + \text{deprecation} - \text{taxes} \\
  &= 190,000 + 130,000 - 46,900 \\
  &= 273,100
\end{aligned}
$$

## Calculating Total Cash Flows

Schwert Corp. shows the following information on its 2012 income statement: sales = \$185,000; costs = \$98,000; other expenses = \$6,700; deprecation expense = \$16,500; interest expense = \$9,000; taxes = \$19,180; dividends = \$9,500. In addition, you're told that the firm issued \$7,550 in new equity during 2012 and redeemed \$7,100 in outstanding long-term debt.

a. What is the 2012 operating cash flow?
b. What is the 2012 cash flow to creditors?
c. What is the 2012 cash flow to stockholders?
d. If net fixed assets increased by \$26,100 during the year, what was the addition to net working capital (NWC)?

### Solution

a. Operating cash flow for 2012

$$
\begin{aligned}
\mathbf{Gross\ income} &= \text{sales} - \text{costs} - \text{other expenses} \\
  &= 185,000 - 98,000 - 6,700 \\
  &= 80,300 \\ \\
\mathbf{OCF} &= \text{gross income} - \text{taxes} \\
  &= 80,300 - 19,180 \\
  &= 61,120
\end{aligned}
$$

b. Cash flow to creditors (CFC)

$$
\begin{aligned}
\mathbf{CFC} &= \text{interest expense} - \text{new net borrowing} \\
  &= 9,000 - (-7,100) \\
  &= 16,100 \\ \\
\end{aligned}
$$

c. Cash flow to stockholders (CFS)

$$
\begin{aligned}
\mathbf{CFS} &= \text{dividends} - \text{net new equity raised} \\
 &= 9,500 - 7,550 \\
 &= 1,950
 \end{aligned}
$$

d. With net fixed assets increased to \$26,100 during the year, the addition to net working capital is:
 
$$
\begin{aligned}
\mathbf{CFFA} &= \text{CFC} + \text{CFS} \\
  &= 16,100 + 1,950 \\
  &= 18,050 \\ \\
\mathbf{Net\ capital\ spending} &= \text{increase in net fixed assets}
  + \text{deprecation} \\
  &= 26,100 + 16,500 \\
  &= 42,600 \\ \\
\mathbf{Change\ in\ NWC} &= \text{OCF} - \text{Net Capital Spending}
  - \text{CFFA} \\
  &= 61,120 - 42,600 - 18,050 \\
  &= 470
\end{aligned}
$$

## Using Income Statements

Given the following information for O'Hare Marine Co., calculate the depreciation expense: sales = \$41,000; costs = \$26,400; addition to retained earnings = \$4,900; dividends paid = \$1,570; interest expense = \$1,840; tax rate = 35 percent.

### Solution

$$
\begin{aligned}
\mathbf{Net\ Income} &= \text{addition to retained earnings}
  + \text{dividends paid} \\
  &= 4,900 + 1,570 \\
  &= 6,470 \\ \\
\mathbf{EBT} &= \frac{\text{net income}}{(1 - \text{tax rate})} \\
  &= \frac{6,470}{(1 - .35)} \\
  &= 9,954 \\ \\
\mathbf{EBIT} &= \text{EBT} + \text{interest paid} \\
  &= 9,954 + 1,840 \\
  &= 11,794 \\ \\
\mathbf{Gross\ income} &= \text{sales} - \text{costs} \\
  &= 41,000 - 26,400 \\
  &= 14,600 \\ \\
\mathbf{Deprecation} &=  \text{gross income} - \text{EBIT} \\
  &= 14,600 - 11,794 = 2,806
\end{aligned}
$$

## Residual Claims

Huang, Inc., is obligated to pay its creditors \$10,900 very soon.

a. What is the market value of the shareholders' equity if assets have a market value of \$12,400?
b. What if assets equal \$9,600?

### Solution

a. 

$$
\begin{aligned}
\mathbf{Shareholders'\ equity} &= \text{assets market value}
  - \text{credit obligations} \\
  &= 12,400 - 10,900 \\
  &= 1,500
\end{aligned}
$$

b.

$$
\begin{aligned}
\mathbf{Shareholders'\ equity} &= \text{assets market value}
  - \text{credit obligations} \\
  &= 9,600 - 10,900 \\
  &= - 1,300
\end{aligned}
$$

## Marginal versus Average Tax Rates

(Refer to the corporate tax rates table from problem 4) Corporation Growth \$86,000 in taxable income, and Corporation Income has \$8,600,000 in taxable income.

a. What is the tax bill for each firm?
b. Suppose both firms have identified a new project that will increase taxable income by \$10,000. How much in additional taxes will each firm pay? Why is this amount the same?

### Solution

To calculate the taxes you sum up the values of all the marginal tax brackets to the bracket the taxable income falls into. We will use the same function we created for problem 4.

a.

```{r problem 17a}
corp.growth <- income.tax(86000)
corp.income <- income.tax(8600000)
```

```{r problem 17a output, echo=F}
table.17a <- matrix(c(corp.growth, corp.income), ncol=1)
rownames(table.17a) <- c("Corporation Growth", "Corporation Income")
knitr::kable(table.17a)
```

b.

To find the marginal tax bracket we define a new function:

```{r problem 17b income.tax.bracket function}
income.tax.bracket <- function(income, brackets=table.tax.rates$Upper, rate=table.tax.rates$Tax.Rate){
  return(which(income <= brackets)[1])
}
```

We use the function to find the marginal tax bracket for both corporations with the next \$10,000 and the tax rate for each.

```{r problem17b calculation}
corp.growth.b <- income.tax.bracket(96000)
corp.income.b <- income.tax.bracket(8610000)
corp.growth.t <- table.tax.rates[corp.growth.b, c("Tax.Rate")]
corp.income.t <- table.tax.rates[corp.income.b, c("Tax.Rate")]
```

```{r problem17b output, echo=F}
table.p17 <- matrix(c(corp.growth.t, corp.income.t, corp.growth.b, corp.income.b), ncol=2)
rownames(table.p17) <- c("Corporation Growth", "Corporation Income")
colnames(table.p17) <- c("Tax Rate", "Tax Bracket")
knitr::kable(table.p17, caption="Marginal Tax Rates")
```

The third and fifth tax brackets tax the next \$10,000 at the same rate 34\%.

## Net Income and OCF

During 2012, Raines Umbrella Corp. had sales of \$630,000. Costs of goods sold, administrative and selling expenses, and deprecation expenses were \$470,000, \$95,000, and \$140,000, respectively. In addition, the company had an interest expense of \$70,000 and a tax rae of 35 percent. (ignore any tax loss carryback or carry forward provisions.)

a. What is Raines's net income for 2012?
b. What is its operating cash flow?
c. Explain your results in (a) and (b).

### Solution

a. EBT is negative so no taxes are owed. The net income is negative.

$$
\begin{aligned}
\mathbf{EBIT} &= \text{sales} - \text{costs of goods sold} - \text{deprecation} \\
  &= 630,000 - 470,000 - 95,000 - 140,000 = - 75,000 \\ \\
\mathbf{EBT} &= \text{EBIT} - \text{interest expense} \\
  &= - 75,000 - 70,000 = - 145,000 \\ \\
\mathbf{Taxes} &= 0 \\ \\
\mathbf{Net\ income} = - 145,000
\end{aligned}
$$

b. Deprecation is added back to EBIT

$$
\begin{aligned}
\mathbf{OCF} &= \text{EBIT} + \text{deprecation} - \text{taxes} \\
  &= - 75,000 + 140,000 - 0 \\
  &= 65,000
\end{aligned}
$$

c. Net income is negative (-\$145,000), which means the company was not profitable for the period. However, operating cash flow was positive (\$65,000). This is because deprecation is a non-cash expense. The company is potentially having problems raising enough cash to cover the cost of buying new fixted assets.

## Accounting values versus cash flows

In problem 18, suppose Raines Umbrella Corp. paid out \$34,000 in cash dividends. Is this possible? If spending on net fixed assets and net working capital was zero, and if no new stock was issued during the year, what was the change in the firm's long-term debt account?

### Solution

Yes, Raines Umbrella Corp. can pay out \$34,000 in cash dividends because operating cash flow is positive \$65,000. Since spending on net fixed assets, net working capital was zero, and no new stock was issued, the only other use of the operating cash flow is in long-term debt.

$$
\begin{aligned}
\mathbf{CFFA} &= \text{OCF} - \text{net capital spending} - \text{change in net working capital} \\
  &= 65,000 - 0 - 0 \\
  &= 65,000 \\ \\
\mathbf{CFC} &= \text{CFFA} + \text{CFS} \\
  &= 65,000 + (- 34,000) \\
  &= 36,000
\end{aligned}
$$

So, \$36,000 was used to pay off long-term debt.

## Calculating cash flows

Cusic Industries had the following operating results for 2012: sales = \$19,900; cost of goods sold = \$14,200; deprecation expense = \$2,700; interest expense = \$670; dividends paid = \$650. At the beginning of the year, net fixed assets were \$15,340, current assets were \$4,420, and current liabilities were \$2,470. At the end of the year, net fixed assets were \$16,770, current assets were \$5,135, and current liabilities were \$2,535. The tax rate for 2012 was 40 percent.

a. What is net income for 2012?
b. What is the operating cash flow for 2012?
c. What is the cash flow from assets from 2012? Is this possible? Explain.
d. If no new debt was issued during the year, what is the cash flow to creditors? What is the cash flow to stockholders? Explain and interpret the positive and negative signs of your answers in (a) through (d).

a. Calculating net income

$$
\begin{aligned}
\mathbf{EBT} &= \text{sales} - \text{costs} - \text{deprecation}
  - \text{interest} \\
  &= 19,000 - 14,200 - 2,700 - 670 \\
  &= 2,330 \\ \\
\mathbf{Taxes} &= \text{EBIT} \times \text{tax rate} \\
  &= 2,100 \times .4 = 932 \\ \\
\mathbf{Net\ Income} &= \text{EBIT} \times (1 - \text{tax rate}) \\
  &= 2,330 \times .6 \\
  &= 1,398
\end{aligned}
$$

b. The operating cash flow

$$
\begin{aligned}
\mathbf{EBIT} &= \text{sales} - \text{costs} - \text{deprecation} \\
  &= 19,900 - 14,200 - 2,700 \\
  &= 3,000 \\ \\
\mathbf{OCF} &= \text{EBIT} + \text{deprecation} - \text{taxes} \\
  &= 3,000 + 2,700 - 932 \\
  &= 4,768
\end{aligned}
$$

c. The cash flow from assets (CFFA)

$$
\begin{aligned}
\mathbf{CFFA} &= \text{OCF} + \text{net capital spending} + \text{change in net working capital}
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{Net\ capital\ spending} &= \text{net fixed assets beginning of year}
 - \text{net fixed assets end of year} \\
 &\ \ \ \ + \text{deprecation} \\
   &= 16,770 - 15,340 + 2,700 \\
   &= 4,130
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{Net\ working\ capital} &= \text{current assets} - \text{current liabilities} \\ \\
\mathbf{Change\ in\ net\ working\ capital} &= \text{net working capital beginning of year}
  - \text{net working capital end of year} \\
  &= (5,135 - 2,535) - (4,420 - 2,470) \\
  &= 650
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{CFFA} &= 3,960 - 1,430 - 650 \\
  &= - 12
\end{aligned}
$$

A negative CFFA indicates that the company invested more in fixed assets and working capital than generated in cash from operations. This is possible and can occur when the firm is in a growth phase, using cash for investments that could lead to future growth.

d. Cash flow to creditors and stockholders

$$
\begin{aligned}
\mathbf{CFC} &= \text{interest paid} - \text{net new borrowing} \\
  &= 670 - 0 \\
  &= 670
\end{aligned}
$$

$$
\begin{aligned}
\mathbf{CFS} &= \text{dividends paid} - \text{net new equity raised} \\
  &= 650 - 0 \\
  &= 650
\end{aligned}
$$

## Calculating cash flows

Consider the following abbreviated financial statements for Weston Enterprises:

a. What is owners' equity for 2011 and 2012?
b. What is the change in net working capital for 2012?
c. In 2012, Weston Enterprises purchased \$2,160 in new fixed assets. How much in fixed assets did Weston Enterprises sell? What is the cash flow from assets for the year? (The tax rate is 40 percent.)
d. During 2012, Weston Enterprises raised \$432 in new long-term debt. How much long-term debt must Weston Enterprises have paid off during the year? What is the cash flow to creditors?

```{r problem 21 balance sheet, echo=F}
 #Assets for balance sheet
table.p21.partial.assets <- matrix(c(936,4176,1015,4896), ncol=2)
rownames(table.p21.partial.assets) <- c("Current assets", "Net fixed assets")
colnames(table.p21.partial.assets) <- c("2011", "2012")
 #Liabilities for balance sheet
table.p21.partial.liabilities <- matrix(c(382,2160,416,2477), ncol=2)
rownames(table.p21.partial.liabilities) <- c("Current liabilities",
                                             "Long-term debt")
colnames(table.p21.partial.liabilities) <- c("2011", "2012")
table.p21.partial.balance.sheet <- list(
  table.p21.partial.assets,
  table.p21.partial.liabilities
)
knitr::kable(table.p21.partial.balance.sheet,
             caption="Weston Enterprises 2011 and 2012 parital balance sheet")
 #Income Statement 2012
table.p21.income.statement.2012 <- matrix(c(12380, 5776, 1150, 314), ncol=1)
rownames(table.p21.income.statement.2012) <- c("Sales", "Costs", "Deprecation",
                                               "Interest paid")
knitr::kable(table.p21.income.statement.2012,
             caption="Weston Enterprises 2012 Income Statement")
```

### Solution

a. Calculating the owners' equity for 2011 and 2012.

```{r problem 21 a}
total.assets.2011 <- sum(table.p21.partial.assets[,1])
total.liabilities.2011 <- sum(table.p21.partial.liabilities[,1])
total.assets.2012 <- sum(table.p21.partial.assets[,2])
total.liabilities.2012 <- sum(table.p21.partial.liabilities[,2])
share.holders.equity.2011 <- total.assets.2011 - total.liabilities.2011
share.holders.equity.2012 <- total.assets.2012 - total.liabilities.2012
share.holders.equity <- matrix(c(share.holders.equity.2011,
                                 share.holders.equity.2012),ncol=2)
colnames(share.holders.equity) <- c("2011", "2012")
knitr::kable(share.holders.equity, caption="Shareholders' equity")
```

b. Change in net working capital for 2012

```{r problem 21 b}
current.assets <- diff(table.p21.partial.assets[1,])
current.liabilities <- diff(table.p21.partial.liabilities[1,])
change.nwc <- as.numeric(diff(c(current.liabilities, current.assets)))
```

c. Fixed assets

```{r problem 21 c}
new.fixed.assets.2012 <- 2160
fixed.assets.2011 <- 4176
fixed.assets.2012 <- 4896
sold.fixed.assets.2012 <- new.fixed.assets.2012 + fixed.assets.2011 -
  fixed.assets.2012

gross.2012 <- 12380 - 5776
ebit.2012 <- gross.2012 - 1150
taxes.2012 <- ebit.2012 * .4
cffa.2012 <- 100
knitr::kable(cffa.2012, col.names = "CFFA")
```

## Financial Statements

```{r Ingersoll problems 22 and 23, echo=F}
table.ingersoll <- matrix( c(
  7835, 1125, 2696, 639, 525, 4109, 5439, 794, 13460, 34455, 4316, 9670, 956,
  8409, 1126, 3060, 534, 603, 5203, 6127, 746, 16050, 35277, 4185, 9938, 1051
), ncol=2)
colnames(table.ingersoll) <- c("2011", "2012")
rownames(table.ingersoll) <- c("Sales", "Deprecation", "Cost of goods sold",
                               "Other expenses", "Interest", "Cash",
                               "Account receivable", "Short-term notes payable",
                               "Long-term debt", "Net fixed assets",
                               "Accounts payable", "Inventory", "Dividends")
knitr::kable(table.ingersoll)
```

Draw up an income statement and balance sheet for the company for 2011 and 2012.

```{r problem 22 income statement, echo=F}
income.statement.headings <- c(
  "Sales",
  "Cost of goods sold",
  "Other expenses",
  "Deprecation",
  "EBIT",
  "Interest expense",
  "Pretax income",
  "Taxes",
  "Net income",
  "Retained earnings",
  "Dividends"
)
 2011
ingersoll.ebit.2011 <- sum(7835, -2696, -639, -1125)
ingersoll.pretax.income.2011 <- sum(ingersoll.ebit.2011, -525)
ingersoll.taxes.2011 <- income.tax(ingersoll.pretax.income.2011)
ingersoll.net.income.2011 <- sum(ingersoll.pretax.income.2011, -ingersoll.taxes.2011)
ingersoll.retained.earnings.2011 <- sum(ingersoll.net.income.2011, -956)
 2012
ingersoll.ebit.2012 <- sum(8409, -3060, -534, -1126)
ingersoll.pretax.income.2012 <- sum(ingersoll.ebit.2012, -603)
ingersoll.taxes.2012 <- income.tax(ingersoll.pretax.income.2012)
ingersoll.net.income.2012 <- sum(ingersoll.pretax.income.2012, -ingersoll.taxes.2012)
ingersoll.retained.earnings.2012 <- sum(ingersoll.net.income.2012, -1051)
ingersoll.income.statement <- matrix(
  c(
   # 2011
   7835,
   2696,
   639,
   1125,
   ingersoll.ebit.2011,
   525,
   ingersoll.pretax.income.2011,
   ingersoll.taxes.2011,
   ingersoll.net.income.2011,
   ingersoll.retained.earnings.2011,
   956,
   # 2012
   8409,
   3060,
   534,
   1126,
   ingersoll.ebit.2012,
   603,
   ingersoll.pretax.income.2012,
   ingersoll.taxes.2012,
   ingersoll.net.income.2012,
   ingersoll.retained.earnings.2012,
   1051
  ),
ncol=2)
rownames(ingersoll.income.statement) <- income.statement.headings
colnames(ingersoll.income.statement) <- c("2011", "2012")
knitr::kable(ingersoll.income.statement,
             caption="Ingersoll, Inc. Income Statement")
```

```{r problem 22 balance sheet, echo=F}
assets.headers <- c("Cash", "Accounts Receivable", "Inventory",
                    "Net Fixed Assets")
ingersoll.assets <- matrix(
  c(
    #2011
    4109,
    5439,
    9670,
    34455,
    # 2012
    5203,
    6127,
    9938,
    35277
  ),
  ncol = 2
)
rownames(ingersoll.assets) <- assets.headers
colnames(ingersoll.assets) <- c("2011", "2012")

liabilities.headers <- c("Short-term notes payable", "Accounts payable",
                         "Long-term debt")
ingersoll.liabilities <- matrix(
  c(
    # 2011
    794,
    4316,
    13460,
    # 2012
    746,
    4185,
    16050
  ),
  ncol=2
)
rownames(ingersoll.liabilities) <- liabilities.headers
colnames(ingersoll.liabilities) <- c("2011", "2012")
ingersoll.total.assets <- matrix(colSums(ingersoll.assets),ncol=2)
colnames(ingersoll.total.assets) <- c("2011", "2012")
rownames(ingersoll.total.assets) <- c("Total Assets")
knitr::kable(rbind(ingersoll.assets, ingersoll.total.assets),
             caption="Balance Sheet: Assets")
ingersoll.total.liabilities <- matrix(colSums(ingersoll.liabilities),ncol=2)
rownames(ingersoll.total.liabilities) <- c("Total Liabilities")
colnames(ingersoll.total.liabilities) <- c("2011", "2012")
knitr::kable(rbind(ingersoll.liabilities, ingersoll.total.liabilities),
             caption="Balance Sheet: Liabilities")
knitr::kable(sum(ingersoll.assets[,1], -ingersoll.liabilities[,1]),
             caption="Shareholders' equity", col.names="2011")
knitr::kable(sum(ingersoll.assets[,2], -ingersoll.liabilities[,2]),
             caption="Shareholders' equity", col.names="2012")
```

## Calculating Cash Flow

For 2012, calculate the cash flow from assets, cash flow to creditors, and cash flow to stockholders.

```{r problem 23, echo=F}
(ocf <- 3375 + 1125 - 427.5)
change.current.assets <- diff(c(sum(ingersoll.assets[1:3,1]), sum(ingersoll.assets[1:3,2])))
change.current.liabilities <- diff(c(sum(ingersoll.liabilities[1:2,1]), sum(ingersoll.liabilities[1:2,2])))
(change.nwc <- sum(change.current.assets, change.current.liabilities))
(ncs <- diff(c(34455, 35277)) + 1126)
cat("CFFA:", (cffa <- ocf - ncs - change.nwc), "\n")
net.new.borrowing <- diff(c(13460, 16050))
net.new.equity <- diff(c(35103, 35564))
cat("CFC:", (cfc <- 603 - net.new.borrowing), "\n")
cat("CFS:", (cfs <-956))
```





<!--

**********************  Challenge Questions  *************************

-->

# Challenge

## Cash Flows

You are researching Time Manufacturing and have found the following accounting statement of cash flows for the most recent year. You also know that the company paid \$98 million in current taxes and had an interest expense of \$48 million. Use the accounting statement of cash flows to construct the financial statement of cash flows.

```{r problem 24 time manufacturing statement of cash flows, echo=F, message=F}
require(kableExtra)
 # net income, deprecation, deferred taxes
operations <- c(173, 94, 19)
 # changes in assets and liabilities: accounts receivable, inventories,
 ## accounts payable, accrued expenses, other
changes.in.assets.and.liabilities <- c(-18, 22, 17, -9, 3)
 # acquisition of fixed assets, sale of fixed assets
investing.activities <- c(-215, 23)
 # retirement of long-term debt, proceeds from long-term debt sales,
 ## changes in notes payable, dividends, repurchase of stock,
 ## proceeds from new stock issue
financing.activities <- c(-162, 116, 6, -86, -13, 44)
statement.of.cash.flows <- matrix(c(
  operations,
  changes.in.assets.and.liabilities,
  investing.activities,
  financing.activities), ncol=1)
rownames(statement.of.cash.flows) <- c(
  "Net income", "Deprecation", "Deferred taxes", "Changes in Accounts receivable",
  "Changes in Inventories", "Changes in Accounts payable",
  "Changes in Accrued expenses", "Changes in Other expenses",
  "Acquisition of fixed assets", "Sale of fixed assets",
  "Retirement of long-term debt", "Proceeds from long-term debt sales",
  "Change in notes payable", "Dividends", "Repurchase of stock",
  "Proceeds from new stock issue"
)
knitr::kable(statement.of.cash.flows,
             booktabs=T,
             caption="Time Manufacturing Inc.: Statement of Cash Flows") %>%
  pack_rows( index=c(
    "Operations" = 8,
    "Investing activities" = 2,
    "Financing activities" = 6
  )) %>%
  kable_styling(latex_options = "striped")
```

### Solution

1. Calculate Operating Cash Flow.

knowing net income (\$173), taxes (\$117), and interest (\$48), we observe:

$$
\begin{aligned}
\mathbf{EBIT} &= \text{net income} + \text{taxes} + \text{interest} \\
  &= 173 + (98 + 19) + 48 \\
  &= 338 \\ \\
\mathbf{OCF} &= \text{EBIT} + \text{deprecation} - \text{current taxes} \\
  &= 338 + 94 - 98 \\
  &= 334 \\ \\
\end{aligned}
$$

2. Calculate the Net capital spending and the net working capital.

$$
\begin{aligned}
\mathbf{Net\ capital\ spending} &= \text{sale of fixed assets} - \text{acquisition of fixed assets} \\
  &= 23 - 215 \\
  &= 192 \\ \\
\mathbf{Change\ in\ net\ working\ capital} &=
  \text{Changes in assets and liabilities} \\
  &= (-18) + 22 + 17 + (-9) + 3 \\
  &= 15 \\ \\
\end{aligned}
$$

3. Finally calculate the cash flow from assets, cash flow to creditors, and cash flow to stockholders.

$$
\begin{aligned}
\mathbf{CFFA} &= \text{OCF} - \text{net capital spending}
    - \text{change in net working capital} \\
  &= 334 - 192 - 15 \\
  &= 127 \\ \\
\mathbf{CFC} &= \text{interest expense} - \text{net new borrowing} \\
  &= 48 - (116 - 162) \\
  &= 94 \\ \\
\mathbf{CFS} &= \text{dividends paid} - \text{net new equity} \\
  &= 86 - (44 - 13) \\
  &= 28
\end{aligned}
$$

## Net Fixed Assets and Deprecation

On the balance sheet, the net fixed assets (NFA) account is equal to the gross fixed assets (FA) account, which records the acquisition cost of fixed assets, minus the accumulated deprecation (AD) account, which records the total deprecation taken by the firm against its fixed assets. Using the fact that NFA = FA - AD, show that the expression given in the chapter for net capital spending,
$\text{NFA}_{\text{end}} - \text{NFA}_{\text{beg}} + D$ (where D is the depreciation expense during the year), is equivalent to $\text{FA}_{\text{end}} - \text{FA}_{\text{beg}}$.

### Solution

Given:

$$
\begin{aligned}
\mathbf{NFA} = \text{FA} - \text{AD}
\end{aligned}
$$

Then
$$
\begin{aligned}
\mathbf{FA} = \text{NFA} + \text{AD}
\end{aligned}
$$
The accumulated depreciation is the sum of the depreciations each year up until the current year:

$$
\begin{aligned}
\mathbf{AD} = \sum_{\text{year}=0}^{\text{current year}}{D_{\text{year}}}
\end{aligned}
$$

Therefore:

$$
\begin{aligned}
\text{NFA}_{\text{end}} - \text{NAF}_{\text{beg}} + D &= 
    \text{FA}_{\text{end}} - \text{FA}_{\text{beg}} \\
\Delta \text{NFA} + D &= \Delta \text{FA} \\ \\
\text{NFA} + D &= \text{FA}
\end{aligned}
$$

## Tax Rates

Refer to the corporate marginal tax rate information in Table 2.3.

a. Why do you think the marginal tax rate jumps up from 34 percent to 39 percent at a taxable income of \$100,001, and then falls back to a 34 percent marginal rate at a taxable income of \$335,001?

b. Compute the average tax rate for a corporation with exactly \$335,001 in taxable income. Does this confirm your explanation in part (a)? What is the average tax rate for a corporation with exactly \$18,333,334? Is the same thing happening here?

c. The 39 percent and 38 percent tax rates both represent what is called a tax "bubble." Suppose the government wanted to lower the upper threshold of the 39 percent marginal tax bracket from \$335,000 to \$200,000. What would the new 39 percent bubble rate have to be?

### Solution













