---
title: "Financial Statements Analysis and Financial Models"
author: "Ross"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
  pdf_document:
   toc: true 
---

# Chapter summary

This chapter focuses on working with information contained in financial statements Specifically, we studied standardized financial statements, ratio analysis, and long-term financial planning.

1. We explained that differences in firm size make it difficult to compare financial statements, and we discussed how to form common-size statements to make comparisons easier and more meaningful.

2. Evaluating ratios of accounting numbers is another way of comparing financial statment information. We defined a number of the most commonly used ratios, and we discussed the famous DuPont identity.

3. We showed how pro forma financial statements can be generated and used to plan for future financing needs.

After you have studied this chapter, we hope that you have some perspective on the uses and abuses of financial statement information. You should also find that your vocabulary of business and financial terms have grown substantially.

# Concepts

## Financial ratio analysis

A financial ratio by itself tells us little about a company because financial ratios vary a great deal across industries. There are two basic methods for analyzing financial ratios for a company: Time trend analysis and peer group analysis. In time trend analysis, you find the ratios for the company over some period, say five years, and examine how each ratio has changed over this period. In peer group analysis, you compare a company's financial ratios to those of its peers. Why might each of these analysis methods be useful? What does each tell you about the company's financial health?

### Solution

#### Time trend analysis

is useful because it provides insights into a company's performance over time. By examining financial ratios across a historical period (e.g., five years), you can identify trends, such as:

* __Improvement or deterioration__: For instance, a steadily improving profit margin suggests better cost control or increasing revenue efficiency, whereas a declining ratio might signal inefficiencies or problems.

* __Seasonal patterns__: Some businesses are cyclical, and observing trends over time can help understand seasonal impacts.

* __Long-term strategy effects__: Changes in ratios might reflect the outcomes of strategic decisions, such as entering new markets, adjusting pricing strategies, or adopting cost-cutting measures.

What time trend analysis reveals about hte financial health of the firm is:

* __Sustainability__: Whether improvements or declines are part of a sustainable trend.

* __Early warning signs__: Sudden deviations in ratios (e.g., liquidity or leverage ratios) could indicate upcoming financial challenges.

* __Growth trajectory__: Trends in profitability and efficiency ratios can show whether the company is scaling effectively.

#### Peer group analysis

involves comparing a company's financial ratios with those of similar firms in the same industry. It provides context for understanding performance by showing whether a company is operating competitively.

Peer group analysis will reveal:

1. A company's performance with respect to its peers.
2. The company's market position with respect to its peers.
3. Shows a firm's competitive advantage and weaknesses with its peers.

Both methods are complementary. Time trend analysis provides insights into how the company is evolving internally. Peer group analysis situates the company's performance within the competitive landscape.

Using both methods together offers a holistic view of a company's financial health, allowing for better-informed decisions by combining internal trends with external comparisons.

## Industry-specific ratios

So-called "same-store sales" are a very important measure for companies as diverse as McDonald's and Sears. As the name suggests, examining same-store sales means comparing revenues from the same stores or restaurants at two different points in time. Why might companies focus on same-store sales rather than total sales?

### Solution

Comapnies often focus on __same-store sales__ rather than total sales because same-store sales provide a more accurate and meaningful measure of a company's __organic growth and performance__ over time, excluding the effects of new store openings or closures.

#### Measures organic growth

__Same-store sales__ focus solely on stores that have been operational for a certain period (usually more than a year), eliminating the impact of new store openings, expansions, or closures. This metric highlights whether existing stores are attracting more customers, selling more products, or increasing revenues through better pricing strategies. For example: A 10% increase in total sales could be misleading if it's driven entirely by new stores, while same-store sales growth indicates improvements in the performance of existing locations.

#### Excludes non-comparable factors

Total sales can be influenced by non-operational factors such as: new store openings or acquisitions, which can inflate growth artificially; store closures, which might mask declines in performance. Same-store sales provide a "like-for-like" comparison, removing these distortions.

#### Indicates customer loyalty and market trends

Growth in same-store sales often reflects higher customer retention, increased traffic, or greater spending per customer at established locations. Declines in same-store sales could signal market saturation, poor customer satisfaction, or increased competition.

#### Helps evaluate management effectiveness

By focusing on same-store sales, management can assess the effectiveness of marketing campaigns, operational improvements, and product offerings within the existing store base. It's a key indicator of how well the company is utilizing its current resources to drive growth.

## Sales forecast

Why do you think most long-term financial planning begins with sales forecasts? Put differently, why are future sales the key input?


### Solution

Most long-term financial planning begins with __sales forecasts__ because sales are the primary driver of a company's operations and financial performance. Future sales serve as the foundation for estimating many other critical financial elements, making them the key input for planning.

Sales drive revenues and cash flows. Sales are the top line of the income statement, and all other financial metrics---such as costs, profits, and cash flows---stem from this figure. Future sales projections determine expected revenue, which is critical for estimating future earnings and cash availability.

Companies use sales forecasts to plan for necessary investments in assets, such as production capacity, equipment, or inventory. For instance: A forecast of rising sales might prompt a company to expand facilities or hire more staff. Declining sales forecasts could signal the need to reduce investment and preserve cash.

Sales growth often requires financing to support operations, including working capital, inventory, and capital expenditures. A robust sales forecast helps determine how much financing the company will need and when, allowing for better capital planning and debt management.

Sales forecasts align operational strategies, such as marketing campaigns, product launches, or market expansion plans, with realistic revenue expectations. For instance, if projected sales growth is aggressive, the company might allocate more resources to marketing or innovation.

Sales forecasts are a key metric for communicating future performance expectations to stakeholders, including investors, lenders, and employees. They form the basis for convincing stakeholders of the company's growth potential and ability to meet obligations.

## Sustainable growth

In the chapter, we used Rosengarten Corporation to demonstrate how to calculate EFN. The ROE for Rosengarten is about 7.3 percent, and the plowback ratio is about 67 percent. If you calculate the sustainable growth rate for Rosengarten, you will find it is only 5.14 percent. In our calculation for EFN, we used a growth rate of 25 percent. Is this possible? (_Hint_: Yes, How?)

### Solution

Yes, it is possible for a company like Rosengarten Corporation to have a __growth rate__ significantly higher than its __sustainable growth rate__ (SGR), such as 25% compared to 5.14%. However, achieving such a high growth rate requires the company to __seek external financing__.

#### Sustainable growth rate defined

The __sustainable growth rate (SGR)__ is the maximum growth rate a company can achieve using only internally generated funds (retained earnings) without needing external financing. To calculate SGR:

$$
\mathbf{SGR} = \text{ROE} \times \text{Plowback Ratio}
$$

For Rosengarten:

$$
\mathbf{SGR} = 7.3\% \times 67\% = 5.14\%
$$

#### Growth beyond SGR

If the company wants to grow faster than its SGR (e.g., at 25%), it must rely on __external financing__ to fund the additional growth. This external financing could come from:

* __Debt__: Borrowing funds to invest in growth opportunities.

* __Equity__: Issuing new shares to raise captial.

* __Hybrid instruments__: Convertible debt or other forms of capital raising.

For example, to support a 25% growth rate: Rosengarten might need to significantly increase its debt levels (which increases the financial leverage). Alternatively, it could issue new equity, although this might dilute existing shareholders' ownership.

#### Why external financing is needed

At a growth rate of 25%, the company's __assets__ will need to grow to support the higher revenue. Since the company's retained earnings (from the 67% plowback ratio) can only fund growth up to the SGR, the company must cover the gap with external financing.

* __Required Asset Growth__: To support 25% revenue growth, assets typically need to grow by a similar percentage.

* __EFN (External Financing Needed)__: This represents the funds required to bridge the gap between the internally generated funds and the total funds needed for the desired growth.

#### Practical implications

* __Financial Risks__: Relying on external financing increases financial risks, especially if the company takes on excessive debt. High leverage can strain cash flows due to interest payments and principal repayments.

* __Market Perception__: Issuing equity might dilute ownership and potentially impact stock prices negatively.

* __Strategic Trade-offs__: The compaany must balance the benefits of high growth with the risks of relying on external funding.

#### Conclusion

While Rosengarten sustainable growth rate is only 5.14%, it can achieve a 25% growth rate by raising external funds. This approach is feasible but requires careful management of financial risks and strategic decisions about the type and amount of external financing to pursue.

## EFN and growth rate

Broslofski Co. maintains a positive retention ratio and keeps its debt-equity ratio constant every year. When sales grow by 20 percent, the firm has a negative projected EFN. What does this tell you about the firm's sustainable growth rate? Do you know, with certainty, if the internal growth rate is greater than or less than 20 percent? Why? WHat happens to the projected EFN if the retention ratio is increased? What if the retention ratio is decreased? What if the retention ratio is zero?

### Solution

This situation with Broslofski Co. provides valuable insights into its __sustainable growth rate (SGR)__, __internal growth rate (IGR)__, and how these relate to projected external financing needs (EFN). Let's analyze each part:

#### What does a negative projected EFN tell you about the firm's sustainable growth rate?

A __negative EFN__ means that the company is generating more internal funds than it needs to support its growth at the projected 20% sales increase. This suggests that the firm's sustainable growth rate (SGR) is __grater than 20%__ because it can fund this growth internally while maintaining its desired capital structure.

#### Is the internal growth rate greater than or less than 20 percent?

The __internal growth rate (IGR)__ is the maximum growth rate the firm can achieve using only retained earnings, without borrowing additional funds or issuing equity. Since the company maintains a constant debt-equity ratio and has a __negative EFN__, it implies that the firm is not reliant on external financing at the 20% growth rate. Therefore, the __IGR must be less than 20%__ because the firm's ability to grow solely using internal funds (IGR) is inherently lower than its ability to grow with the help of leverage (SGR).

#### What happens to the projected EFN if the retention ratio is increased?

The increased retention ratio means the firm reinvests a larger portion of its earnings back into the business, reducing dividends. This increases the amount of internally generated funds available for growth, reducing EFN further. If EFN was negative, it would become more negative, meaning the firm generates even more surplus funds than needed fo rht eprojected growth.

#### What happens to the projected EFN if the retention ratio is decreased?

The decreased retntion ratio means the firm retains less of its earnings and distributes more as dividends. This reduces the amount of internal funds available for reinvestment, potentially increasing EFN. If the EFN was negative, it might move closer to zero, indicating reduced surplus funds. If EFN becomes positive, the firm would need external financing to support its growth.

#### What if the retention ratio is zero?

A zero retention ratio means the firm does not reinvest any of its earnings; all profits are distributed as dividends. In this case: the firm is entirely reliant on external financing (debt or equity) to fund growth; the EFN will likely become positive, as there are no internal funds to support growth; and the firm's IGR and SGR will also drop to zero, as both require retained earnings for reinvestment.

#### Summary of the relationships

1. __SGR > 20%__: The firm's ability to grow sustainably exceeds 20%, indicated by the negative EFN.

2. __IGR < 20%__: The IGR is less than 20% because it represents growth achievable without any external leverage.

3. __Retention Ratio impact__: an increase means EFN becomes more negative (more surplus funds); a decrease means EFN moves closer to zero or becomes positive (less surplus funds); and if zero, EFN becomes positive and no internally generated funds are available to support growth.

This analysis highlights how retention policy and internal vs. external financing decisions influence a firm's ability to sustain and project growth.

## Common-size financials

One tool of financial analysis is common-size financial statements. Why do you think common-size income statements and balance sheets are used? Note that the accounting statement of cash flows is not converted into a common-size statement. Why do you think this is?

### Solution

#### Simplifies comparison across companies

Because each line item is expressed as a percentage of total sales for income statements and percentage of total assets for balance sheets. This helps to compare companies of different size, industry, or geographic location because they eliminate the impact of scale.

#### Highlights financial trends

By converting financial statements into percentages, common-size analysis helps identify trends over time, such as changes in cost structures, profitability, or asset composition. For instance: a rising cost of goods sold (COGS) as a percentage of sales might indicate worsening cost efficiency; or Increasing current liabilities as a percentage of total assets could signal growing liquidity risks.

#### Aids in benchmarking

Common-size statements facilitate comparisons against industry averages or peer companies, helping analysts assess a company's competitive position. For example, a company with a lower operating expense ratio than peers might have superior cost control.

#### The cash flow statement is not common-sized because

1. __Lack of logical Base__ The cash flow statement tracks the sources and uses of cash across three categories: operating, investing, and financing activities. There is no single, consistent "base" to use for expressing cash flow items as percentages. Unlike sales in the income statement or total assets in the balance sheet, the cash flow statement doesn't have a comparable anchor figure.

2. __Focus on absolute values__ The cash flow statement's primary purpose is to show actual cash inflows and outflows, which are absolute values. Understanding whether the company generates enough cash to cover its obligations or fund growth is more important than relative percentages.

3.__Varied nature of cash flows__ The cash flow statement includes diverse line items that don't naturally align with a common denominator: operating cash flow links to profitability and working capital; investing cash flow reflects capital expenditures and assets purchases or sales; financing cash flow deals with borrowing, repaying debt, and equity transactions. These activities are not proportional to a single metric like sales or total assets.

#### Conclusion

Common-size financial statements are invaluable for standardizing financial data to facilitate analysis and comparison, particularly for the income statement and balance sheet. However, the cash flow statement is best analyzed using absolute figures, as it focuses on cash management rather than proportional relationships.

## Asset utilization and EFN

One of the implicit assumptions we made in calculating the external funds needed was that the company was operating at full capacity. If the company is operating at less than full capacity, how will this affect the external funds needed?

### Solution

If a company is operating at less than full capacity, it has the potential to increase production without immediately investing in additional fixed assets. This can significantly affect the calculation of external funds needed (EFN):

#### Reduced need for fixed asset investment

At less than full capacity, the company can increase sales and production without requiring proportional increases in fixed assets (e.g., machinery, equipment, facilities). Since EFN calculations typically assume proportional growth in all asset categories, this assumption would overestimate the need for fixed asset investment in such a scenario. This would result in lower EFN because the company doesn't need to spend as much on capital expenditures.

#### Implications for asset turnover

Operating below capacity implies that existing fixed assets are underutilized. As production ramps up to meet increased sales, the fixed asset turnover ratio improves (sales per unit of fixed assets), further reducing the need for additional investments.

#### Working capital requirements still increase

Even if fixed asset investments are unnecessary, growth in sales will still require additional current assets (e.g., inventory and accounts receivable) and might increase current liabilities (e.g., accounts payable). However, these changes are typically smaller than those involving fixed assets.

#### Adjusted EFN

If the company adjusts for underutilized capacity, then EFN will be calculated based only on increases in current assets and current liabilities, excluding or reducing the need for new fixed assets. This will result in a lower EFN compared to the assumption of full capacity utilization.

#### Example

Assume sales grow by 20%, and the company is operating at 80% capacity. The company can handle the 20% growth without additional fixed asset investment because it only needs to utilize the remaining 20% capacity. EFN calculations based on proportional growth would have overestimated the required increase in fixed assets, inflating EFN unnecessarily.

#### Key takeaways

When operating below full capacity, EFN decreases because fixed asset investments are reduced or unnecessary. Thus, the calculation must be adjusted to reflect the reality of underutilized capacity. However, as the company approaches full capacity, future growth will eventually require investment in fixed assets, at which point EFN will increase.

Understanding capacity utilization ensures more accurate financial planning and avoids overstating the need for external funding.

## Comparing ROE and ROA

Both ROA and ROE measure profitability. Which one is more useful for comparing two companies? Why?

### Solution

Both return on assets (ROA) and return on equity (ROE) measure profitability, but which one is more useful for comparing two companies depends on the context and what you're trying to analyze.

#### ROA (return on assets)

Measures how efficiently a company uses its total assets to generate profit. ROA also accounts for both equity and debt financing since assets are funded by both.

$$
\mathbf{ROA} = \frac{\text{net income}}{\text{total assets}}
$$

You would use ROA to compare two companies when:

##### Capital structure differs

ROA provides a more comparable measure of operational efficiency because it ignores the effects of leverage. For example, if one company is heavily debt-financed and the other relies mostly on equity, ROA gives a better sense of how well each uses its total assets to generate profits.

##### Assessing asset efficiency

ROA evaluates how effectively the companies deploy their assets, regardless of how they are financed.

#### ROE (return on equity)

Focuses on the return generated on the shareholders' investment and reflects the impact of leverage (debt) on profitability.

$$
\mathbf{ROE} = \frac{\text{net income}}{\text{shareholder's equity}}
$$

You would use ROE to compare two companies when:

##### Capital structure is similar

If two companies have similar levels of debt, ROE is more relevant because it focuses on returns to shareholders.

##### Evaluating shareholder returns

ROE directly shows how much profit shareholders earn per dollar of equity invested, which is key for equity investors.

#### Key considerations

##### Impact of leverage

High leverage increases ROE because debt financing boosts returns on equity (if the cost of debt is lower than the return on assets). However, this can distort comparisons between companies with different debt levels. ROA is unaffected by leverage, making it a more neutral measure.

##### Industry differences

In asset-heavy industries (e.g., manufacturing), ROA is particularly relevant for assessing operational efficiency. In industries where leverage is common (e.g., banking), ROE might be might indicative of performance.

#### Conclusion

##### ROA

is more useful for comparing companies with different capital structures because it measures asset efficiency independently of financing decisions.

##### ROE

is more relevant when comparing companies with similar debt levels or when focusing on returns from a shoareholder's perspective.

Ultimately, understanding both metrics provides a more comprehensive view of profitability, but the choice depends on what aspect of performance you wish to compare.

## Ratio analysis

Consider the ratio EBITD/Assets. What does this ratio tell us? Why might it be more useful than ROA in comparing two companies?

### Solution

The ratio EBITD/Assets (Earnings Before Interest, Taxes, and Deprecation divided by Toatl Assets) provides insight into a company's ability to generate operating cash flow relative to its total assets.

#### EBITD/Assets viewpoint

##### Operational efficiency

It measures how efficiently a company uses its assets to generate operating cash flow, excluding non-cash charges (deprecation and amortization) and before the effects of financing and taxes.

##### Cash flow focus

By using EBITD instead of net income, the ratio focuses on cash-based profitability rather than accounting profitability, which can be affected by non-cash expenses and financing decisions.

##### Asset Utilization

Like ROA, this ratio indicates how effectively a company's total assets are utilized, but with an emphasis on cash-generating ability.

#### EBITD/Assets may be more useful in the following scenarios

##### Removes the effect of financing decisions

ROA uses net income in the numerator, which is influenced by interest expenses (financing structure), while EBITD/Assets excludes interest, making it more neutral for comparisons between companies with different levels of debt.

##### Excludes non-cash accounting effects

Deprecation and amortization are non-cash expenses that vary based on accounting methods and asset age. EBITD excludes these charges, providing a better measure of operational cash flow and enabling a more apples-to-apples comparison of companies with different asset bases or deprecation policies.

##### Focuses on oprational performance

EBITD reflects the compnay's core operational performance before taxes and financing costs. This makes it particularly useful for comparing companies in capital-intensive industries where deprecation and amortization can significantly distort net income.

##### Better for cross-industry or cross-region comparisons

ROA can be affected by tax rates or regional accounting practices, which vary across industries or countries.  EBITD/Assets neutralizes these differences, focusing on operational cash flow generation relative to assets.

#### Example comparison

Company A has high leverage, uses debt financing extensively which results in high interest expenses. Company B has low leverage, primarily equit-financed, with minimal interest expenses. EBITD/Assets allows for a direct comparison of operational efficiency without being skewed by financing differences, whereas ROA might make Company B appear more profitable soley due to its financing structure.

#### Situations where EBITD/Assets excels

##### Capital-intensive industries

In industries like manufacturing or utilities, EBITD/Assets highlights operational efficiency despite large deprecation charges.

##### High-leverage companies

Useful for comparing companies with varying levels of debt, as it excludes interest effects.

##### Startups or Growth Companies

Young companies with large upfron capital investments but limited net income can still show strong EBITD/Assets.

### Conclusion

EBITD/Assets is particularly useful when comparing companies with different financing structures, deprecation policies, or tax environments. It provides a clearer view of a company's ability to generate operating cash flow from its assets, making more neutral and informative in specific contexts than ROA.

## Return on investment

A ratio that is becoming more widely used is return on investment. Return on investment is calculated as net income divided by long-term liabilities plus equity. What do you think return on investment is intended to measure? What is the relationship between return on investment and return on assets?

### Solution

Return on investment (ROI) is calculated as

$$
\mathbf{ROI} = \frac{\text{net income}}{\text{long-term liabilitieis} + \text{equity}}
$$

This ratio is intended to measure the profitability of a company relative to its total capital base, including bot debt (long-term liabilities) and equity.

#### ROI measures

##### Profitablitility of total capital

ROI indicates how effectively a company is generating net income from the resources provided by both creditors (long-term debt) and owners (equity). It evaluates the return on all long-term financing rather than just equity, making it a broader profitability measure.

##### Financial leverage impact

Since long-term liabilities are included in the denominator, ROI accounts for the impact of leverage on profitability, though it does not distinguish between returns to debt and equity holders.

##### Capital efficiency

A higher ROI means the company is better at converting its long-term financing into net income.

#### Relationship between ROI and ROA

Return on assets (ROA) is calculated as:

$$
\mathbf{ROA} = \frac{\text{net income}}{\text{total assets}}
$$

ROA uses total assets as the base, capturing the entire resource base employed by the company. ROI uses long-term liabilities + equity, which represents the company's permanent capital structure (ignoring current liabilities and short-term financing). ROI can be thought of as a subset of ROA, focusing specifically on returns from capital funded through long-term financing. For a compnay with minimal current liabilities, ROI and ROA will be similar. However, if the company has significant current liabilities, ROI will typically be higher than ROA because the denominator (long-term liabilities + equity) is smaller than total assets.

---

Use the following information to answer the next five questions: A small business called the Grandmother Calendar Company began selling personalized photo calendar kits. The kits were a hit, and sales soon sharply exceeded forecasts. The rush of orders created a huge backlog, so the company leased more space and expanded capacity, but it still could not keep up with demand. Equipment failed from overuse and quality suffered. Working capital was drained to expand production, and, at the same time, payments from customers were often delayed until the product was hsipped. Unable to deliver on orders, the company became so strapped for cash that employee paychecks began to bounce. Finally, out of cash, the company ceased operations entirely three years later.

## Product sales

Do you think the company would have suffered the same fate if its product had been less popular? Why or Why not?

### Solution

While high demand is generally positive, in this case, the company's inability to manage the operational and financial challenges of rapid growth led to its failure. If its product had been less popular, the company might have grown at a manageable pace, avoiding the operational and financial crises that ultimately led to its closure.

## Cash flow

The Grandmother Calendar Company clearly had a cash flow problem. In the context of the cash flow analysis we developed in Chapter 2, what was the impact of customers not paying until orders were shipped?

### Solution

Customers delaying payments until orders were shipped created a significant cash flow mismatch for the Grandmother Calendar Company. This issue compounded the financial stress caused by high demand and operational challenges, ultimately leading to a liquidity crisis. Managing cash flow by aligning inflows and outflows is crucual for sustaining operations, particularly for growing businesses like this one.

## Corporate borrowing

If the firm was so successful at selling, why wouldn't a bank or some other lender step in and provide it with the cash it needed to continue?

### Solution

Despite strong sales, the Grandmother Calendar Company faced a combination of operational inefficiencies, cash flow challenges, and poor financial management, which likely made it appear too risky for banks or lenders. To secure funding, the company would have needed to deomonstrate not only strong sales but also a clear plan for addressing its liquidity issues, stabilizing operations, and ensuring repayment ability. Without such assurances, even successful sales alone would not have been enough to secure additional financing.

## Cash flow

Which was the biggest culprit here: Too  many orders, too little cash, or too little production capacity?

### Solution

While too many orders and too little production capacity were significant challenges, too little cash was the biggest cu;prit. With better cash flow management (e.g., requiring deposits, better aligning inflows and outflows, or securing short-term financing), the company could have scaled its production capacity and fulfilled orders. Ultimately, the cash shortage turned what could have been a success story into a failure.

## Cash flow

What are some actions a small company like The Grandmother Calendar Company can take (besides expansion of capacity) if it finds itself in a situation in which growth in sales outstrips production?

### Solution

1. Manage demand strategically: adjust pricing, limit order intake, introduce pre-orders.
2. Outsource or partner: contract manufacturing or collaborate with competitors
3. Optimize operations: improve production efficiency or focus on core products
4. Enhance cash flow management: request customer deposits, offer discounts for early payments, negotiate supplier terms.
5. Prioritize customer satisfaction: communicate with customers, offer premium options.
6. Leverage technology: adopt inventory and order management software, forecast demand.
7. Use short-term financing: bridge loans or credit lines, factoring accounts receivable.
8. Explore alternative fulfillment models: Licensing or franchising, digital products or services.

# Basic Questions

## DuPont identity

If Jares, Inc., has an equity multiplier of 1.55, total asset turnover of 1.75, and a profit margin of 4.3 percent, what is its ROE?

### Solution

$$
\begin{aligned}
\mathbf{ROE} &= \text{profit margin} \times \text{asset turnover}
  \times \text{equity multiplier} \\
  &= 4.3 \times 1.75 \times 1.55 = 11.66
\end{aligned}
$$

---

## Equity multiplier and return on equity

Nuber Company has a debt-equity ratio of .80. Return on assets is 9.7 percent, and total equity is \$735,000. What is the equity multiplier? Return on equity? Net income?

$$
\begin{aligned}
\mathbf{Equity\ multiplier} &= 1 + \text{debt-equity ratio} \\
  &= 1 + .80 \\
  &= 1.80 \\ \\
\mathbf{ROE} &= \text{ROA} \times \text{equity multiplier} \\
  &= 9.7\% \times 1.8 \\
  &= 17.46\% \\ \\
\mathbf{Net\ income} &= \text{ROA} \times \text{total assets} \\
  &= \text{ROA} \times (\text{equity multiplier} \times \text{total equity}) \\
  &= .097 \times (1.80 \times 735,000) \\
  &= 128,331
\end{aligned}
$$

---

## EFN

The most recent financial statements for Martin, Inc., are shown here:

```{r problem 4, echo=F}
p4.income.statement <- matrix(c(37300, 25800, 11500, 3910, 7590), ncol=1)
row.names(p4.income.statement) <- c("Sales", "Costs", "taxable income",
                                "Taxes (34%)", "Net income")
p4.balance.sheet <- matrix(c(127000, 30500, 96500), ncol=1)
row.names(p4.balance.sheet) <- c("Assets", "Debt", "Equity")

knitr::kable(p4.income.statement, caption="Income Statement")
knitr::kable(p4.balance.sheet, caption="Balance Sheet")
```

Assets and costs are proportional to sales. Debt and equity are not. A dividend of \$2,500 was paid, and Martin wishes to maintain a constant payout ratio. Next year's sales are projected to be \$42,300. What external financing is needed?

### Solution

Since Assets and costs are proportional to projected sales (\$42,300) for the income statement we get:

$$
\begin{aligned}
\mathbf{Payout\ ratio} &= \frac{\text{dividends}}{\text{net income}} \\ \\
  &= \frac{2,500}{7,590} \approx 0.3294 \\ \\
\mathbf{Growth\ rate} &= \frac{\text{projected sales} - \text{current sales}}
  {\text{current sales}} \\ \\
  &= \frac{42,300 - 37,300}{37,300} = 0.1343 \\ \\
\mathbf{Projected\ assets} &= \text{current assets} \times (1 + \text{growth rate}) \\
  &= 127,000 \times 1.1343 \\
  &= 144,646 \\ \\
\mathbf{Projected\ net\ income} &= \text{current net income} \times (1 + \text{growth rate}) \\
  &= 7,590 \times 1.1343 \\
  &= 8,608.20 \\ \\
\mathbf{Addition\ to\ retained\ earnings} &= \text{net income} \times \text{retention raio} \\
  &= 8,608.20 \times 0.6706 \\
  &= 5,774.31 \\ \\
\mathbf{Projected\ equity} &= \text{current equity} + \text{addition to retained earnings} \\
  &= 96,500 + 5,774.31 \\
  &= 102,274.31 \\ \\
\mathbf{EFN} &= \text{projected assets} - (\text{current debt} + \text{projected equity}) \\
  &= 144,646.10 - (30,500 + 102,274.31) \\
  &= 11,871.79
\end{aligned}
$$

```{r solution 4}
pct.inc.sales <- 42300/37300
payout.ratio <- 2500/7590
assets.proforma <- 127000 * pct.inc.sales
netincome.proforma <- 7590 * pct.inc.sales
retained.earnings.proforma <- netincome.proforma*(1-payout.ratio)
equity.proforma <- 96500 + retained.earnings.proforma
efn <- assets.proforma - (30500 + equity.proforma)
print(efn)
```

---

## Sales and growth

The most recent financial statements for Fontenot Co. are shown here:

```{r problem 5, echo=F}
p5.is <- matrix(c(54000, 39300, 14700, 4998, 9702), ncol=1)
row.names(p5.is) <- c("Sales", "Costs", "Taxable income", "Taxes (34%)", "Net income")
p5.bs <- matrix(c(31000, 118000, 149000, 68000, 81000, 149000), ncol = 1)
row.names(p5.bs) <- c("Current assets", "Fixed assets", "Total Assets", "Long-term debt", "Equity", "Debt and Equity")
knitr::kable(p5.is, caption="Income Statement")
knitr::kable(p5.bs, caption="Balance Sheet")
```

Assets and costs are proportional to sales. The company maintains a constant 30 percent dividend payout ratio and a constant debt-equity ratio. What is the maximum increase in sales that can be sustained assuming no new equity is issued?

### Solution

$$
\begin{aligned}
\mathbf{Retention\ ratio} &= 1 - \text{dividend payout ratio} \\
  &= 1 - 0.30 \\
  &= 0.70 \\ \\
\mathbf{DER} &= \frac{\text{Long-term Debt}}{\text{Equity}} \\
  &= \frac{68,000}{81,000} \\
  &= 0.8395 \\ \\
\mathbf{ROE} &= \frac{\text{net income}}{\text{equity}} \\
  &= \frac{9,702}{81,000} \\
  &= 0.1198 \\ \\
\mathbf{SGR} &= \text{ROE} \times \text{retention ratio} \\
  &= 0.1198 \times 0.70 \\
  &= 0.0839 \\ \\
\mathbf{Maximum\ increase\ in\ sales} &= \text{current sales} \times \text{SGR} \\
  &= 54,000 \times 0.0839 \\
  &= 4,531 \\ \\
\mathbf{Projected\ sales} &= \text{current sales} + \text{maximum increase in sales} \\
  &= 54,000 + 4,531 \\
  &= 58,531
\end{aligned}
$$

\$4,531 is the largest sales growth Fontenot Co. can sustain without issuing new equity while maintaining its current policies. The total projected sales is \$58,531.

```{r solution 5}
p5.retention.ratio <- 0.70
p5.debt.to.equity.ratio <- 68000/81000
p5.roe <- 9702/81000
p5.sgr <- p5.roe * p5.retention.ratio
cat("Maximum increase in sales:", p5.max.inc.sales <- 54000 * p5.sgr)
cat("Projected Total Sales:", p5.projected.sales <- 54000 + p5.max.inc.sales)
```

---

## Sustainable growth

If the Layla Corp. has a 13 percent ROE and a 20 percent payout ratio, what is its sustainable growth rate.

### Solution

$$
\begin{aligned}
\mathbf{SGR} &= \text{ROE} \times \text{retention ratio} = \text{ROE} 
  + (1 - \text{payout ratio}) \\
  &= .13 \times (1 - .20) \\
  &= 0.104
\end{aligned}
$$

---

## Sustainable growth

Assuming the following ratios are constant, what is the sustainable growth rate?

* Total asset turnover = 2.20
* Profit margin = 7.4%
* Equity multiplier = 1.40
* Payout ratio = 40%

$$
\begin{aligned}
\mathbf{ROA} &= \text{profit margin} \times \text{total asset turnover} \\
  &= .074 \times 2.20 \\
  &= .1628 \\ \\
\mathbf{ROE} &= \text{ROA} \times \text{equity multiplier} \\
  &= .1628 \times 1.40 \\
  &= .2279 \\ \\
\mathbf{SGR} &= \text{ROE} \times \text{retention ratio} = \text{ROE} +
  (1 - \text{payout ratio}) \\
  &= .2279 + (1 - .4) \\
  &= .1367
\end{aligned}
$$

---

## Calculating EFN

The most recent financial statements for Bradley, Inc., are shown here (assuming no income taxes):

```{r problem 8, echo=F}
p8.is <- matrix(c(6500, 5320, 1180), ncol=1)
row.names(p8.is) <- c("Sales", "Costs", "Net income")
p8.bs <- matrix(c(17400, 17400, 8400, 9000, 17400), ncol=1)
row.names(p8.bs) <- c("Assets", "Total Assets", "Debt", "Equity", "Debt and Equity")
knitr::kable(p8.is, caption="Income Statement")
knitr::kable(p8.bs, caption="Balance Sheet")
```

Assets and costs are proportional to sales. Debt and equity are not. No dividends are paid. Next year's sales are projected to be \$7,280. What is the external financing needed?

### Solution

$$
\begin{aligned}
\mathbf{Growth\ rate} &= \frac{\text{projected sales} - \text{current sales}}
  {\text{current sales}} \\
  &= \frac{7,280 - 6,500}{6,500} \\
  &= 0.12 \\ \\
\mathbf{Projected\ assets} &= \text{current assets} \times (1 + \text{growth rate}) \\
  &= 17,400 \times 1.12 \\
  &= 19,488 \\ \\
\mathbf{Projected\ net\ income} &= \text{current net income} \times (1 + \text{growth rate}) \\
  &= 1,180 \times 1.12 \\
  &= 1,321.60 \\ \\
\mathbf{Projected\ equity} &= \text{current equity} + \text{Net income} \\
  &= 9,00 + 1,321.60 \\
  &= 10,321.60 \\ \\
\mathbf{EFN} &= \text{projected assets} - (\text{current debt} + \text{projected equity}) \\
  &= 19,488 - (8,400 + 10,321.60) \\
  &= 766.40
\end{aligned}
$$

---

## External funds needed

Cheryl Colby, CFO of Charming Florist Ltd., has created the firm's pro forma balance sheet for the next fiscal year. Sales are projected to grow by 10 percent to \$420 million. Current assets, fixed assets, and short-term debt are 20 percent, 75 percent, and 15 percent of sales, respectively. Charming Florist pays out 30 percent of its net income in dividends. The company currently has \$120 million of long-term debt and \$48 million in common stock par value. The profit margin is 9 percent.

a. Construct the current balance sheet for the firm using the projected sales figure.
b. Based on Ms. Colby's sales growth forecast, how much does Charming Florist need in external funds for the upcoming fiscal year?
c. Construct the firm's pro forma balance sheet for the next fiscal year and confirm the external funds needed that you calculated in part (b).

### Solution

```{r problem 9, echo=F}
p9.proj.sales <- 420
p9.bs <- matrix(c(
  420*c(.2, .75),
  sum(420*c(.2, .75)),
  420*.15,
  120,
  sum(420*c(.2, .75), -420*.15, -120)),
  ncol=1)
row.names(p9.bs) <- c("current assets", "fixed assets", "total assets",
                      "short-term debt", "long-term debt", "equity")
(p9.net.income <- .09 * p9.proj.sales)
(p9.retained.earnings <- p9.net.income * (1 - 0.30))
(p9.proj.equity <- as.numeric(p9.bs[6,1]) + p9.retained.earnings)
(p9.proj.total.assets <- as.numeric(p9.bs[3,1]) * 1.1)
(p9.proj.st.debt <- .15 * 420)
(p9.proj.total.liabilities <- p9.proj.st.debt + 120)
(p9.efn <- p9.proj.total.assets - (p9.proj.total.liabilities + p9.proj.equity))
```

---

## Sustainable growth rate

The Steiben company has an ROE of 13.1 percent and a payout ratio of 40 percent.

a. What is the company's sustainable growth rate?
b. Can the company's actual growth rate be different from its sustainable growth rate? Why or why not?
c. How can the company increase its sustainable growth rate?

### Solution

$$
\begin{aligned}
\mathbf{SGR} &= \text{ROE} \times (1 - \text{payout ratio}) \\
  &= .131 \times .6 \\
  &= .0786
\end{aligned}
$$

b. Yes, a company growth rate can be different from its sustainable growth rate. If it is below the SGR, then the company is operating at less than full capacity. If the growth rate is above the sustainable growth rate, it would need external funds to sustain the growth rate.

c. It can increase its sustainable growth rate by improving its capital structure.

# Intermediate Questions

## Return on equity

Firm A and Firm B have debt-total asset ratios of 35 percent and 55 percent and returns on total assets of 9 percent and 7 percent, respectively. Which firm has a greater return on equity?

### Solution

#### Definitions

The standard leverage ratios

$$
\begin{aligned}
\mathbf{Total\ assets} &= \text{total debt} + \text{total equity} \\ \\
\mathbf{Debt\ ratio} &= \text{D/A} = \frac{\text{total assets} - \text{total equity}}{\text{total assets}} \\ \\
\mathbf{D/E} &= \frac{\text{total debt}}{\text{total equity}} 
  = \frac{1}{1 - \text{D/A}} \\ \\
\mathbf{Equity\ multiplier} &= 1 + \frac{\text{total assets}}{\text{total equity}} \\ \\
\mathbf{ROE} &= \text{ROA} \times \text{EM}
\end{aligned}
$$

Firm A:

$$
\begin{aligned}
\mathbf{EM}_{A} = 1 + \frac{1}{1 - .35} = 1.5385 \\ \\
\mathbf{ROE}_{A} = 0.09 \times 1.5385 = 0.1385
\end{aligned}
$$

Firm B:

$$
\begin{aligned}
\mathbf{EM}_{B} = 1 + \frac{1}{1 - .55} = 2.2222 \\ \\
\mathbf{ROE}_{B} = 0.07 \times 2.2222 = .1556
\end{aligned}
$$


### Conclusion

Although firm B has lower ROA, it has a greater return on equity because it uses more leverage (higher debt-total asset ratio).

---

## Ratios and Foreign Companies

Prince Albert Canning PLC had a net loss of 37,543 pounds on sales of 345,182 pounds. What was the company's profit margin? Does the fact that these figures are quoted in a foreign currency make any difference? Why? In dollars, sales were \$559,725. What was the net loss in dollars?

### Solution

$$
\begin{aligned}
\mathbf{Profit\ margin} &= \frac{\text{net income}}{\text{sales}} \\
  &= \frac{37,543}{345,182} \\
  &= - 0.1088
\end{aligned}
$$

The profit margin would be the same in any currency because it is a ratio.

If sales were 559,725 in dollars. Then the exchange rate between dollars and pounds is

$$
\frac{559,725}{345,182} = 1.62
$$

The net loss in dollars is

$$
(37,543) \times 1.62 \approx (60,820)
$$

---

## External Funds Needed

The Optical Scam Company has forecast a 15 percent sales growth rate for next year. The current financial statements are shown here:

```{r problem 13, echo=F}
p13.is <- matrix(c(30400000, 26720000, 3680000,
                   1288000, 2392000, 956800, 1435200),
                 ncol=1)
row.names(p13.is) <- c("Sales", "Costs", "Taxable income", "Taxes",
                       "Net income", "Dividends",
                       "Addition to retained earnings")

p13.bs.assets <- matrix(c(7200000, 17600000, 24800000), ncol=1)
row.names(p13.bs.assets) <- c("Current assets", "Fixed assets",
                              "Total assets")
p13.bs.liabilities.and.equity <- matrix(c(6400000, 4800000,
                                          3200000, 10400000,
                                          13600000, 24800000), ncol=1)
row.names(p13.bs.liabilities.and.equity) <- c("Short-term debt",
        "Long-term debt", "Common stock", "Accumulated retained earnings",
        "Total equity", "Total liabilities and equity")
p13.bs <- list(p13.bs.assets, p13.bs.liabilities.and.equity)
```

```{r problem 13 tables, echo=F}
knitr::kable(p13.is, caption="Income Statement")
knitr::kable(p13.bs, caption="Balance Sheet")
```

a. Using the equation, calculate the external funds needed for next year.
b. Construct the firm's pro forma balance sheet for next year and confirm the external funds needed that you calculated in part (a).
c. Calculate the sustainable growth rate for the company.
d. Can Optical Scam eliminate the need for external funds by changing its dividend policy? What other options are available to the company to met its growth objectives?

### Solution

a. The equation for EFN

$$
\mathbf{EFN} = \frac{\text{assets}}{\text{sales}} \times \Delta \text{sales}
  - \frac{\text{spontaneous liabilities}}{\text{sales}} \times \Delta \text{sales}
  - \text{profit margin} \times \text{projected sales} \times
  (1 - \text{dividend payout ratio})
$$


```{r solution 13a sales incease}
p13.sales <- as.numeric(p13.is[1,])
p13.sales.inc <- p13.sales * .15
(p13.asset.growth <- as.numeric(p13.bs.assets[3,]) / p13.sales * p13.sales.inc)
(p13.spont.liabilities.growth <- as.numeric(p13.bs.assets[1,]) / p13.sales * p13.sales.inc)
(p13.profit.margin <- as.numeric(p13.is[5,]) / p13.sales)
(p13.projected.sales <- p13.sales + p13.sales.inc)
(p13.dividend.payout <- as.numeric(p13.is[6,]) / as.numeric(p13.is[5,]))
(p13.equity.addition <- p13.profit.margin * p13.projected.sales * (1 - p13.dividend.payout))
(p13.efn <- p13.asset.growth - p13.spont.liabilities.growth - p13.equity.addition)
```

b. The Pro forma balance sheet

```{r solution 13b}
(p13.bs.assets.pro.forma <- p13.bs.assets * 1.15)
(p13.st.liabilities <- p13.bs.liabilities.and.equity[1,] * 1.15)
(p13.retained.earnings <- p13.bs.liabilities.and.equity[4,] +
  p13.is[7,] * 1.15)
(p13.efn)
```

c. Sustainable growth

$$
\mathbf{Sustainable\ growth\ rate} = \frac{\text{ROE} \times b}
  {1 - \text{ROE} \times b}
$$

$$
\begin{aligned}
\mathbf{ROE} &= \frac{\text{net income}}{\text{total equity}} \\
  &= \frac{2,392,000}{13,600,000} \\
  &= 0.176 \\ \\
\mathbf{b} &= 1 - .4 = .6 \\ \\
\mathbf{SGR} &= \frac{0.176 \times .6}{1 - 0.176 \times .6} \\
  &= .1056
\end{aligned}
$$

d. Options to external funds needed:

* Reducing the payout ratio would increase retained earnings, reducing EFN.
* Increasing profit margin to generate more internal funds.
* Reduce asset growth (e.g., improve efficiency or delay capital expenditures).
* Increase spontaneous liabilities (e.g., negotiate better payment terms with suppliers).
* Issue new equity or take on additional long-term debt.

---

## Days' sales in receivables

A company has net income of \$265,000, a profit margin of 9.3 percent, and an accounts receivable balance of \$145,300. Assuming 80 percent of sales are on credit, what is the company's days' sales in receivables?

### Solution

$$
\begin{aligned}
\mathbf{Profit\ margin} &= \frac{\text{net income}}{\text{sales}} \\ \\
\mathbf{Sales} &= \frac{\text{net income}}{\text{profit margin}} \\
  &= \frac{265,000}{0.093} \\
  &= 2,849,462 \\ \\
\mathbf{Sales\ on\ credit} &= \text{sales} \times 0.8 \\
  &= 2,279,570 \\ \\
\mathbf{Receivables\ turnover} &= \frac{\text{sales on credit}}{\text{accounts receivable}} \\
  &= \frac{2,279,570}{145,300} \\
  &= 15.7 \\ \\
\mathbf{Days'\ sales\ in\ receivables} &= \frac{365\ \text{days}}{\text{receivables turnover}} \\
  &= \frac{365}{15.7} \\
  &= 23.2
\end{aligned}
$$

## Ratios and fixed assets

The Le Bleu Company has a ratio of long-term debt to total assets of .35 and a current ratio of 1.25. Current liabilities are \$950, sales are \$5,780, profit margin is 9.4 percent, and ROE is 18.2 percent. What is the amount of the firm's net fixed assets?

### Solution

$$
\begin{aligned}
\mathbf{Net\ income} &= \text{profit margin} \times \text{sales} \\
  &= .094 \times 5,780 \\
  &= 543.32 \\ \\
\mathbf{Equity} &= \frac{\text{net income}}{\text{ROE}} \\
  &= \frac{543.32}{.182} \\
  &= 2985.28 \\ \\
\mathbf{Assets} &= \text{liabilities} + \text{Equity} \\
\mathbf{Assets} &= \frac{\text{long-term debt}}{\text{assets}} \times \text{assets} +
  \text{current liabilities} + \text{equity} \\ \\
.65 \times \text{assets} &= \text{current liabilities} + \text{equity} \\
\text{assets} &= \frac{3935}{.65} \\
  &= 6053 \\ \\
\mathbf{Fixed\ assets} &= \text{assets} - \text{current assets} \\
  &= 6053 - (1.25)(950) \\
  &= 4865
\end{aligned}
$$

---

## Calculating the cash coverage ratio

Titan Inc.'s net income for the most recent year was \$8,320. The tax rate was 34 percent. The firm paid \$1,940 in total interest expense and deducted \$2,730 in depreciation expense. What was Titan's cash coverage ratio for the year?

### Solution

$$
\begin{aligned}
\mathbf{taxes} &= \frac{\text{net income}}{1 - \text{tax rate}}
  \times \text{tax rate} \\
  &= \frac{8,320}{1 - .34} \times .34 \\
  &= 4,286 \\ \\
\mathbf{EBITDA} &= \text{net income} + \text{interest} + \text{taxes}
  + \text{deprecation} \\
  &= 8,320 + 1,940 + 4286 + 2,730 \\
  &= 17,276 \\ \\
\mathbf{Cash\ Coverage\ Ratio} &= \frac{\text{EBITDA}}{\text{interest}} \\
  &= \frac{17,276}{1,940} \\
  &= 8.9
\end{aligned}
$$

---

## DuPont identity

The DuPont identity presented in the chapter is commonly referred to as the three-factor DuPont identity. Another common way that the DuPont identity is expressed is the five-factor model, which is:

$$
\mathbf{ROE} = \frac{\text{net income}}{\text{EBT}} \times \frac{\text{EBT}}{\text{EBIT}}
  \times \frac{\text{EBIT}}{\text{sales}} \times \frac{\text{sales}}{\text{total assets}}
  \times \frac{\text{total assets}}{\text{equity}}
$$

Derive the five-factor DuPont identity (EBT is earnings before tax, but after interest). What does each term measure?

### Solution

The five-factor DuPont identity breaks out the profit margin into three different factors.

$$
\begin{aligned}
\mathbf{Tax\ burden} &= \frac{\text{net income}}{\text{EBT}} \\ \\
\mathbf{Interest\ burden} &= \frac{\text{EBT}}{\text{EBIT}} \\ \\
\mathbf{Operating\ profit\ margin} &= \frac{\text{EBIT}}{\text{sales}} \\ \\
\mathbf{Total\ asset\ turnover} &= \frac{\text{sales}}{\text{total assets}} \\ \\
\mathbf{Profit\ margin} &= \frac{\text{total assets}}{\text{equity}}
\end{aligned}
$$

---

## Common-size and common-base year financial statements

In addition to common-size financial statements, common-base year financial statements are often used. Common-based year financial statements are constructed by dividing the current year account value by the base year account value. Thus, the result shows the growth rate in the account. Using the following financial statements, construct the common-size balance sheet and common-base year balance sheet for the company. Use 2011 as the base year.

```{r problem 18 balance sheet, echo=F}
p18.bs.assets <- matrix(c(8014, 20453, 36822, 65289, 215370, 280659,
                          9954, 22937, 41797, 74688, 243340, 318028), ncol=2)
row.names(p18.bs.assets) <- c("Cash", "Accounts receivable", "Inventory", "Total",
                              "Net plant and equipment", "Total assets")
colnames(p18.bs.assets) <- c("2011", "2012")
p18.bs.liabilities.equity <- matrix(c(40898, 17464, 58362, 24000, 38000, 160297,
                                      198297, 280659, 45884, 17035, 62919, 31000,
                                      39200, 184909, 224109, 318028), ncol=2)
row.names(p18.bs.liabilities.equity) <- c("Accounts payable", "Notes payable",
                                       "Total", "Long-term debt",
                                       "Common stock and paid-in surplus",
                                       "Retained earnings", "Total",
                                       "Total liabilities and owners' equity")
colnames(p18.bs.liabilities.equity) <- c("2011", "2012")
knitr::kable(list(p18.bs.assets, p18.bs.liabilities.equity),
             caption="Jarrow Corporation 2011 and 2012 Balance Sheets")
total.assets <- p18.bs.assets["Total assets",]
```

### Solution

```{r solution 18, echo=F}
base.year.assets <- p18.bs.assets[,1]
common.base.assets.2012 <- p18.bs.assets[,2] / base.year.assets
base.year.liabilities.equity <- p18.bs.liabilities.equity[,1]
common.base.liabilities.equity.2012 <- p18.bs.liabilities.equity[,2] /
  base.year.liabilities.equity
knitr::kable(list(common.base.assets.2012, common.base.liabilities.equity.2012),
             caption="Common Base 2012")
```

---

Use the following information for Problems 19, 20, and 22:

The discussion of EFN in the chapter implicitly assumed that the company was operating at full capacity. Often, this is not the case. For example, assume that Rosengarten was operating at 90 percent capacity. Full-capacity sales would be \$1,000/.90 = \$1,111. The balance sheet shows \$1,800 in fixed assets. The capital intensity ratio for the company is

$$
\begin{aligned}
\mathbf{Capital\ intensity\ ratio} &= \frac{\text{fixed assets}}{\text{full-capacity sales}} \\
 &= \frac{1,800}{1,111} \\
 &= 1.62
\end{aligned}
$$

This means that Rosengarten needs \$1.62 in fixed assets for every dollar in sales when it reaches full capacity. At the projected sales level of \$1,250, it needs $\$1,250 \times 1.62 = \$2,025$ in fixed assets, which is \$225 lower than our projection of \$2,250 in fixed assets. So, EFN is only \$565 - 225 = \$340.

## Full-capacity sales


Thorpe Mfg., Inc., is currently operating at only 90 percent of fixed asset capacity. Current sales are \$725,000. How much can sales increase before any new fixed assets are needed?

### Solution

```{r problem 19}
725000/.9 - 725000
```

---

## Fixed assets and capacity usage

For the company in the previous problem, suppose fixed assets are \$690,000 and sales are projected to grow to \$830,000. How much in new fixed assets are required to support this growth in sales?

### Solution

```{r problem 20}
sales.full.capacity <- 725000/.9
asset.to.sales <- 690000 / sales.full.capacity
fixed.assets.required <- asset.to.sales * 830000
(net.fixed.assets.required <- fixed.assets.required - 690000)
```

---

## Calculating EFN

The most recent financial statements for Moose Tours, Inc., appear below. Sales for 2012 are projected to grow by 20 percent. Interest expense will remain constant; the tax rate and the dividend payout rate will also remain constant. Costs, other expenses, current assets, fixed assets, and accounts payable increase spontaneously with sales. If the firm is operating at full capacity and no new debt or equity is issued, what external financing is needed to support the 20 percent growth rate in sales?

```{r problem 21, echo=F}
p21.is <- matrix(c(836100, 650700,17100, 168300, 12600, 155700, 54495,
                   101205, 30300, 70905), ncol=1)
row.names(p21.is) <- c("Sales", "Costs", "Other expenses",
                       "Earnings before interest and taxes", "Interest expense",
                       "Taxable income", "Taxes", "Net income", "Dividends",
                       "Addition to retained earnings")
p21.bs.assets <- matrix(c(24035, 38665, 82555, 145255, 392350, 537605), ncol=1)
row.names(p21.bs.assets) <- c("Cash", "Accounts receivable", "Inventory", "Total",
                              "Fixed assets", "Total Assets")
p21.bs.liabilieies.equity <- matrix(c(64600, 16150, 80750, 150000, 130000,
                                      176855, 306855, 537605), ncol=1)
row.names(p21.bs.liabilieies.equity) <- c("Accounts payable", "Notes payable",
                                          "Total", "Long-term debt",
                                          "Common stock and paid-in surplus",
                                          "Retained earnings", "Total",
                                          "Total liabilities and owners' equity")
knitr::kable(p21.is, caption="2011 Income Statement")
knitr::kable(list(p21.bs.assets, p21.bs.liabilieies.equity),
             caption="31 Dec 2011 Balance Sheet")
```

### Solution

```{r solution 21a}

```

```{r solution 21}
(p21.sales <- 836100)
(p21.growth.rate <- .2)
(p21.change.sales <- p21.sales * p21.growth.rate)
(p21.sales.proj <- 836100 * 1.2)
(p21.assets <- 537605)
(p21.accounts.payable <- 64600)
(p21.notes.payable <- 16150)
(p21.spontaneous.liabilities <- 64600 + 16150)
(p21.net.income <- 101205)
(p21.dividends <- 30300)
(p21.dividend.payout.ratio <- p21.dividends / p21.net.income)
(p21.profit.margin <- p21.net.income / p21.sales)
(p21.assets.to.sales <- p21.assets / p21.sales)
(p21.spontaneous.liabilities.to.sales <- p21.spontaneous.liabilities / p21.sales)
(p21.retained.earnings <- p21.profit.margin * p21.sales.proj * (1 - p21.dividend.payout.ratio))
(p21.efn <- p21.assets.to.sales * p21.change.sales - p21.spontaneous.liabilities.to.sales * p21.change.sales - p21.retained.earnings)
```

---

## Capacity usage and growth

In the previous problem, suppose the firm was operating at only 80 percent capacity in 2011. What is EFN now?

### Solution

```{r problem 22}
(p22.full.capacity.sales <- 836100 / .8)
(p22.change.sales <- p21.change.sales)
(p22.proportional.assets <- 145255 / 836100)
(p22.sales.proj <- p21.sales.proj)
(p22.projected.total.assets <- p22.proportional.assets * p22.sales.proj + 392350)
(p22.assets.to.sales <- p22.projected.total.assets / p22.sales.proj)
(p22.efn <- p22.assets.to.sales * p22.change.sales - p21.spontaneous.liabilities.to.sales *
    p22.change.sales - p21.retained.earnings)
```

---

## Calculating EFN

In problem 21, suppose the firm wishes to  keep its debt-equity ratio constant. What is the EFN now?

### Solution

```{r problem 23}
p23.debt.equity <- 80750 / 306855
```

---





