--- 
title: "Financial Statements Analysis and Financial Models"
author: "Ross, Westerfield, Jaffe"
header-includes:
  - \usepackage{booktabs}
toc: true 
output: pdf_document
pdf_engine: xelatex
---

```{r setup, echo=FALSE, includes=FALSE, message=FALSE}
require(kableExtra)
require(dplyr)
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
pretty <- function(x) format(x, big.mark=",")
```

# Chapter introduction

The price of a share of common stock in green tractor manufacturer John Deere
closed at about \$88 on May 13, 2011. At that price, John Deere had a
price-earnings (PE) ratio of 17. That is, investors were willing to pay \$17 for
every dollar in income earned by John Deere. At the same time, investors were
willing to pay \$9, \$13, and \$27 for each dollar earned by Ford, Coca-Cola,
and Google, respectively. At the other extreme were Bank of America and United
States Steel. Both had negative earnings for the previous year, but Bank of
America was priced at about \$12 per share and United States Steel at about \$45
per share. Because they had negative earnings, their PE ratios would have been
negative, so they were not reported. At the same time, the typical stock in the
S&P 500 index of large company stocks was trading at a PE of about 14, or about
14 times earnings, as they say on Wall Street.

Price--earnings comparisons are examples of the use of financial ratios. As we
will see in this chapter, there are a wide variety of financial ratios, all
designed to summarize specific aspects of a firm's financial position. In
addition to discussing how to analyze financial statements and compute financial
ratios, we will have quite a bit to say about who uses this information and why.

---

# List of topics

1. Financial statement analysis

   * Standardizing statements
   * Common-size balance sheets
   * Common-size income statements

2. Ratio analysis

   * Short-term solvency or liquidity measures
   * Long-term solvency measures
   * Asset management or turnover measures
   * Profitability measures
   * Market value measures

3. The DuPont Identity

   * A closer look at ROE
   * Problems with financial statement analysis

4. Financial models

   * A simple planning model
   * The percentage of sales approach

5. External financing and growth

   * EFN and growth
   * Financial policy and growth
   * A note about sustainable growth rate calculations

6. Some caveats regarding financial planning models

---

# Notes

## Financial statements analysis

### Standardizing statements

Comparing company's financial statements is done by standarizing their
statements to common-size statements.

### Common-size balance sheets

We need a reference. We can use percentage of assets.

### Common-size income statements

Common measures of earnigns.

#### Net Income

The so-called bottom line, defined as total revenue minus total expenses. Net
income reflects differences in a firm's capital structure and taxes as well as
operating income. Interest expense and taxes are subtracted from operating
income in computing net income. Shareholders look closely at net income because
dividend payout and retained earings are closely linked to net income.

#### EPS

Net income divided by the number of shares outstanding. It expresses net income
on a per share basis.

#### EBIT

Earnings before interest expense and taxes. EBIT is usually called "income from
operations" on the income statement and is income before unusual items,
discontinued operations or extrodinary items. To calculate EBIT, operating
expenses are subtracted from total operations revenues. Analysts like BIT
because it abstracts from differences in earnings from a firm's capital
structure (interest expense) and taxes.

#### EBITDA

Earnings before interest expenses, taxes, depreciation, and amortization refers
to a noncash expense similar to depreciation except it applies to an intangible
asset (such as a patent), rather than a tangible asset (such as a machine). The
word amortization here does not refer to the payment of debt. Analysts like to
use EBITDA because it adds back two noncash items (depreciation and
amortization) to EBIT and thus is a better measure of before-tax operating cash
flow.

## Ratio Analysis

### Short-term solvency or liquidity measures

#### Current ratio

$$
\text{Current ratio} = \frac{\text{Current assets}}{\text{Current liabilties}}
$$

#### Quick (or Acid-Test) Ratio

$$
\text{Quick ratio} = \frac{\text{current assets} - \text{inventory}}
  {\text{current liabilities}}
$$

#### Cash ratio

$$
\text{Cash ratio} = \frac{\text{cash}}{\text{current liabilities}}
$$

### Long-term solvency measures

#### Total debt ratio

$$
\text{Total debt ratio} = \frac{\text{total assets} - \text{total equity}}
  {\text{total assets}}
$$

#### Debt-equity ratio

$$
\text{Debt-equity ratio} = \frac{\text{total debt}}{\text{total equity}}
$$

#### Equity multiplier

$$
\text{Equity multiplier} = \frac{\text{total assets}}{\text{total equity}}
$$

#### Times interest earned

$$
\text{Times interest earned ratio} = \frac{\text{EBIT}}{\text{interest}}
$$

##### Cash coverage

$$
\text{Cash coverage ratio} = \frac{\text{EBIT}
  + (\text{Depreciation and amortization})}{\text{interest}}
$$

### Asset managment and turnover measures

#### Inventory turnover and days' sales in inventory

$$
\text{inventory turnover} = \frac{\text{cost of goods sold}}
  {\text{inventory}}
$$

$$
\text{Days' sales in inventory} = \frac{\text{365 days}}
  {\text{inventory turnover}}
$$

#### Receivables turnover and days' sales in receivables

$$
\text{Receivables turnover} = \frac{\text{sales}}{\text{accounts receivable}}
$$

$$
\text{Days' sales in receivables} = \frac{\text{365 days}}
  {\text{receivables turnover}}
$$

#### Total asset turnover

$$
\text{Total asset turnover} = \frac{\text{sales}}{\text{total assets}}
$$

### Profitability measures

#### Profit margin

$$
\text{Profit margin} = \frac{\text{net income}}{\text{sales}}
$$

#### EBITDA margin

$$
\text{EBITDA margin} = \frac{\text{EBITDA}}{\text{sales}}
$$

#### Return on assets

$$
\text{Return on assets} = \frac{\text{net income}}
  {\text{total assets}}
$$

#### Return on equity

$$
\text{Return on equity} = \frac{\text{net income}}
  {\text{total equity}}
$$

### Market value measures

#### Price-earnings ratio

$$
\text{PE ratio} = \frac{\text{price per share}}{\text{earnings per share}}
$$

#### Market-to-book ratio

$$
\text{Market-to-book ratio} = \frac{\text{market value per share}}
  {\text{book value per share}}
$$

#### Market capitalization

$$
\text{Market capitalization} = \text{price per share}
  \times \text{shares outstanding}
$$

#### Enterprise value

$$
\text{Enterprise value} = \text{market capitalization}
  + \text{market value of interest bearing debt}
$$

#### Enterprise value multiples

$$
\text{Enterprise value multiples} = \frac{\text{enterprise value}}
  {\text{EBITDA}}
$$

## The DuPont identity

The difference between ROA and ROE reflects the firm's use of debt financing.
Higher ROE compared to ROA means heavy debt financing.

$$
\begin{aligned}
\text{Return on equity} &= \frac{\text{net income}}{\text{total equity}}\\\\
  &= \frac{\text{net income}}{\text{total equity}}
  \times \frac{\text{assets}}{\text{assets}}\\\\
  &= \frac{\text{net income}}{\text{assets}}
  \times \frac{\text{assets}}{\text{total equity}}\\\\
  &= \text{ROA} \times \text{equity multiplier}\\\\
  &= \text{ROA} \times (1 + \text{debt-equity ratio})
\end{aligned}
$$

The Dupont Identity:

$$
\begin{aligned}
\text{ROE} &= \underbrace{\frac{\text{net income}}{\text{sales}}
  \times \frac{\text{sales}}{\text{assets}}}_{\text{Return on assets}}
  \times \frac{\text{assets}}{\text{total equity}}\\
  &= \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier}
\end{aligned}\tag{3.21}
$$

We partitioned return on assets (ROA) into two components: profit margin and
total asset turnover.

The DuPont identity tells us that ROE is affected by three things:

1. Operating efficiency (as measured by profit margin).
2. Asset use efficiency (as measured by total asset turnover).
3. Financial leverage (as measured by the equity multiplier).


Decomposing ROE into these components is a good start at financial analysis and
will give information about the current state of the business.

### Problems with financial statement analysis

Ther is no financial analysis theory that tells you which ratios are more
important.

Difficulty in assessing firms exist because of many factors such as:

1.  Some firms are conglamerates of multiple businesses making analysis
    difficult.

2. Globalization spreads competition out into different countries, using
   different currencies, operating in different regulatory and political
   environments. All these factors make comparisons difficult.

3. Different accounting procedures can make comparisons difficult. For instance,
   differences in inventory accounting can make financial analysis very
   difficult.

## Financial models

### A simple financial planning model

Simple example to demonstrate financial planning and estimation. We will use the
percentage of sales approach.

All variables will be tied to the rate of growth of sales.

The income statement and the balance sheet has been simplified.

```{r financial models; computerfield income statement, echo=F}
computerfield <- new.env()
computerfield$is <- matrix(c(1000, 800, 200), ncol=1)
rownames(computerfield$is) <- c("Sales", "Costs", "Net income")

knitr::kable(pretty(computerfield$is), align = "lr",
             caption="Computerfield corporation, income statement") %>%
    add_indent(c(3))
```

```{r financial models; computerfield balance sheet, echo=F}
computerfield$bs <- matrix(c(500, 500, 250, 250, 500), ncol=1)
rownames(computerfield$bs) <- c("Assets", "Total Assets", "Debt", "Equity",
                                "Total Debt and Equity")

knitr::kable(pretty(computerfield$bs),
      align = "lr", caption="Computerfield Corporation balance sheet") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    pack_rows("Assets", 1, 2, bold=T, background="#f2f2f2") %>%
    pack_rows("Debt and Equity", 3, 5, bold=T, background="#D3E4CD") %>%
    add_indent(c(2, 5))
```

If sales increase by \%20, rising from \$1,000 to \$1,200, then the pro forma
income statement would be:

```{r financial models; computerfield pro forma income statement, echo=F}
computerfield$is.proforma <- computerfield$is
computerfield$is.proforma <- computerfield$is * 1.2
knitr::kable(pretty(computerfield$is.proforma), align="lr",
    caption="Computerfield corporation, pro forma income statement") %>%
  add_indent(c(3))
```

Using the assumption that all variables will grow by 20 percent we will be able
to construct the pro forma balance sheet:

Similar to the income statement, if sales grow \%20 then the pro forma balance
sheet

```{r financial models; computerfield pro forma balance sheet, echo=F}
computerfield$bs.proforma <- computerfield$bs
computerfield$bs.proforma <- computerfield$bs * 1.2
rownames(computerfield$bs.proforma) <-
    c("Assets (+100)", "Total Assets (+100)", "Debt (+50)",
      "Equity (+50)", "Total Debt and Equity (+100)")
knitr::kable(computerfield$bs.proforma, align = "lr",
      caption="Computerfield Corporation pro forma balance sheet") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    pack_rows("Assets", 1, 2, bold=T, background="#f2f2f2") %>%
    pack_rows("Debt and Equity", 3, 5, bold=T, background="#D3E4CD") %>%
    add_indent(c(2, 5))
```

Now that we have constructed the two pro forma financial statements we need to
reconcile the two. How can net income be equal to \$240 and equity increase by
only \$50? The answer is that Computerfield must have paid out the difference.
$\$240 - 50 = \$190$ could be paid out as a cash dividend. The cash dividend is
called a _plug variable_.

If Computerfield did not pay out the \$190 dividend then that means that
retained earnings would have grown the full \$240, making equity grow to $250 +
240 = \$490$, and so debt must be retired to keep total assets equal to \$600.

Computerfield would have to retire \$490 in debt making the new debt balance: 
$\$600 - 490 = \$110$. This new pro forma balance sheet would look like the
following:

```{r financial models; computerfield alternate pro forma balance sheet, echo=F}
computerfield$bs.proforma.alt <- computerfield$bs
computerfield$bs.proforma.alt <- matrix(c(600, 600, 110, 490, 600), ncol=1)
rownames(computerfield$bs.proforma.alt) <-
    c("Assets (+100)", "Total Assets (+100)", "Debt (-140)",
      "Equity (+240)", "Total Debt and Equity (+100)")
knitr::kable(pretty(computerfield$bs.proforma.alt), align = "lr",
      caption="Computerfield Corporation alternate pro forma balance sheet") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    pack_rows("Assets", 1, 2, bold=T, background="#f2f2f2") %>%
    pack_rows("Debt and Equity", 3, 5, bold=T, background="#D3E4CD") %>%
    add_indent(c(2, 5))
```

In this case, the _plug variable_ used to balance projected total assets and
total liabilities will be debt.

### The percentage of sales approach

The previous example used the percentage of sales approach in which every item
increased at the same rate as sales. This may be reasonable for some accounts
but for others, such as long-term borrowing, it probably is not. Long-term
borrowing is set by management, and it does not necessarily relate directly to
the level of sales.

Here we will separate hte income statement and balance sheet accounts into two
groups, those that vary directly with sales and those that are independent.

#### The income statement

Below is the recent income statement for the Rosengarten Corporation. The income
statement has been simplified, combining costs, depreciation, and interest in a
single cost figure.

Rosengarten is projecting a 25 percent increase in sales for the coming year.

```{r financial models; rosengarten corporation; income statement, echo=F}
rosengarten <- new.env()
rosengarten$growth.rate <- .25
rosengarten$is <- matrix(c(1000, 800, 200, 68, 132, 44, 88), ncol=1)
rownames(rosengarten$is) <- c("Sales", "Costs", "Taxable income", "Taxes (34%)",
                              "Net income", "Dividends",
                              "Addition to retained earnings")
knitr::kable(pretty(rosengarten$is),
      col.names=c("Item", "Amount ($)"), align="lr",
      caption="Rosengarten Corporation income statement") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    add_indent(c(5))
```

Rosengarten has projected a 25 percent increase in sales for the coming year, so
we are anticipating sales of $\$1,000 \times 1.25 = \$1,250$. To generate the
pro forma income statement we will assume that total costs will continue to run
at $\$800/1,000 = 80$ percent of sales. With this assumption, Rosengarten's pro
forma income statement is as shown:

```{r financial models; rosengarten corp; pro forma income statement, echo=F}
rosengarten$is.proforma <- matrix(
    rosengarten$is[1:5,] * (1 + rosengarten$growth.rate), ncol=1)
rownames(rosengarten$is.proforma) <-
    c("Sales (projected)", "Costs", "Taxable income",
      "Taxes (34%)", "Net income")
knitr::kable(pretty(rosengarten$is.proforma), align = "lr",
     caption="Rosengarten Corporation pro forma income statement") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   add_indent(c(5))
```

To assume that costs are a constant percentage of sales is to assume that the
profit margin is constant.

$$
\begin{aligned}
\text{Profit margin} = \frac{\text{net income}}{\text{sales}}
  = \frac{132}{1,000} = \frac{165}{1,250} = .132
\end{aligned}
$$

Next we project the dividend payment. This is up to management. Assume
Rosengarten has a policy of paying out a constant fraction of net income in the
form of a cash dividend. The _dividend payout ratio_ was:

```{r financial models; rosengarten dividend ratio, echo=F}
rosengarten$dividend.ratio <- .3333
```

$$
\text{Dividend payout ratio} = \frac{\text{cash dividends}}{\text{net income}}
  = \frac{88}{132} = 33\ \tfrac{1}{3}\%
$$

The ratio of addition to retained earnings to net income (also called the
_retention ratio_ or the _plowbak ratio_):

$$
\text{retention ratio} = \frac{\text{Addition to retained earnings}}
  {\text{net income}} = \$88/132 = 66\ \tfrac{2}{3}\%
$$

Therefore, we predict the following for Dividends and the addition to retained
earnings:

$$
\begin{aligned}
\text{Projected dividends paid to shareholders} &= \$165 \times 1/3 = &\$ &\ 55\\
\text{Projected addition to retained earnings} &= \$165 \times 2/3 = & &110\\
& &\$ &165
\end{aligned}
$$

#### The balance sheet

We assume that some of the items on the balance sheet vary directly with sales
and other items do not. For those items that vary with sales, we express each as
a percentage of sales for the year just completed. For items that do not vary
with sales we use `NA`.

As an example, on the asset side, we assume that inventory is 60 percent of
sales. That means for every \$1 increase in sales, inventory will increase
\$.60. The ratio of total assets to sales for the year is
$\$3,000/1,000 = 3$, or 300 percent. This ratio of total assets to sales is
called the _capital intensity ratio_. This ratio tells us the amount of assets
we need to generate \$1 in sales; the higher the ratio is, the more capital
intensive the firm is. This ratio is just the reciprical of the total asset
turnover.

```{r financial models; rosengarten corp; balance sheet, echo=F}
rosengarten$bs <- matrix(c(160, 440, 600, 1200, 1800, 3000, 300, 100,
                           400, 800, 800, 1000, 1800, 3000), ncol=1)
rownames(rosengarten$bs) <- c("Cash", "Accounts receivable", "Inventory",
                              "Total current assets", "Net plant and equipment",
                              "Total assets", "Accounts payable",
                              "Notes payable", "Total current liabilities",
                              "Long-term debt",
                              "Common stock and paid-in surplus",
                              "Retained earnings", "Total equity",
                              "Total liabilities and owners' equity")
rosengarten$bs.pct.of.sales <- c(.16, .44, .60, 1.2, 1.8, 3, .3,
                                 NA, NA, NA, NA, NA, NA, NA)
rosengarten$bs <- cbind(rosengarten$bs, rosengarten$bs.pct.of.sales)

knitr::kable(pretty(rosengarten$bs),
     col.names=c("", "Amount ($)", "% of sales"), align="lrr",
     caption="Rosengarten Corporation balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Current assets", 1, 4, bold=T, background="#f2f2f2") %>%
   pack_rows("Fixed assets", 5, 5, bold=T, background="#D3E4CD") %>%
   pack_rows("Current liabilities", 7, 9, bold=T, background="#FAD4D4") %>%
   pack_rows("Owners' equity", 11, 13, bold=T, background="#D9E4EC") %>%
   row_spec(5, extra_latex_after = "\\addlinespace") %>%
   row_spec(6, extra_latex_after = "\\addlinespace") %>%
   row_spec(13, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(4, 9, 13))
```

On the liability side of the balance sheet, we show accounts payable varying
with sales. This is because we expect more orders with suppliers as sales
volumes grow. Notes payable represents short term borrowing and will not vary
unless we take specific actions to change the amount. This also is the case for
long term debt and common stock and paid-in surplus. Retained earnings will vary
with sales, but it won't be a simple percentage of sales. Instead, we will
explicitly calculate the change in retained earnings based on our projected net
income and dividends.

We can now construct a partial pro forma balance sheet for Rosengarten. We use
the percentage of sales (25%) to calculate the change in each item that we want
to vary by sales. All of the items on the asset side of the balance sheet. On
the liabilities and owners' equity side, only accounts payable will vary
directly with sales. The items that we don't vary directly with sales are the
items that management will control. Notice that the change in retained earnings
is equal to the \$110 addition to retained earnings calculated in the income
statement.

```{r financial models; rosengarten corp; pro forma balance sheet, echo=F}
tmp <- matrix(rosengarten$bs[,1], ncol=1)
rownames(tmp) <- rownames(rosengarten$bs)
rosengarten$bs.proforma <- tmp
rosengarten$bs.proforma[1:7] <- tmp * (1 + rosengarten$growth.rate)
rosengarten$bs.proforma["Total current liabilities",] <-
    rosengarten$bs.proforma["Accounts payable",] +
    rosengarten$bs.proforma["Notes payable",]
rosengarten$bs.proforma["Retained earnings",] <-
    tmp["Retained earnings",] +
    rosengarten$is.proforma["Net income",] * (1 - rosengarten$dividend.ratio)
rosengarten$bs.proforma["Total equity",] <-
    tmp["Total equity",] +
    rosengarten$is.proforma["Net income",] * (1 - rosengarten$dividend.ratio)
rosengarten$bs.proforma["Total liabilities and owners' equity",] <-
    rosengarten$bs.proforma["Total current liabilities",] +
    rosengarten$bs.proforma["Long-term debt",] +
    rosengarten$bs.proforma["Total equity",]
rosengarten$bs.proforma.efn <-
    rosengarten$bs.proforma["Total assets",] -
    rosengarten$bs.proforma["Total liabilities and owners' equity",]
rosengarten$bs.proforma.delta <- c(rosengarten$bs.proforma - rosengarten$bs[,1],
                                   rosengarten$bs.proforma.efn)
rosengarten$bs.proforma <- rbind(rosengarten$bs.proforma,
                                 rosengarten$bs.proforma.efn)
rosengarten$bs.proforma <- cbind(rosengarten$bs.proforma,
                                 rosengarten$bs.proforma.delta)
rownames(rosengarten$bs.proforma) <- c(rownames(rosengarten$bs),
                                       "External financing needed")
knitr::kable(pretty(round(rosengarten$bs.proforma)),
     col.names=c("Item", "Amount ($)", "Change from current year"), align="lrr",
     caption="Rosengarten Corporation pro forma balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Current assets", 1, 4, bold=T, background="#f2f2f2") %>%
   pack_rows("Fixed assets", 5, 5, bold=T, background="#D3E4CD") %>%
   pack_rows("Current liabilities", 7, 9, bold=T, background="#FAD4D4") %>%
   pack_rows("Owners' equity", 11, 13, bold=T, background="#D9E4EC") %>%
   row_spec(5, extra_latex_after = "\\addlinespace") %>%
   row_spec(6, extra_latex_after = "\\addlinespace") %>%
   row_spec(13, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(4, 9, 13))
```

Inspecting the pro forma balance sheet, notice that assets are projected to
increase by \$750. However, without additional financing, liabilities and equity
will increase by only \$185, leaving a shortfall of
\$750 - 185 = \$`r #rosengarten$bs.proforma[15,2]`. We label this amount as
_external financing needed_ (EFN) in the balance sheet. We could also calculate
EFN directly:

$$
\begin{aligned}
\text{EFN} = &\frac{\text{assets}}{\text{sales}} \times \Delta \text{sales}
  - \frac{\text{spontaneous liabilities}}{\text{sales}}
    \times \Delta \text{sales} - \text{PM}\\
  & \times \text{projected sales} \times (1 - d)
\end{aligned}\tag{3.23}
$$

In this expression, "$\Delta$sales" is the projected change in sales (in
dollars); "Spontaneous liabilities" are the liabilities that naturally move up
and down with sales (here its accounts payable); and PM and d are the profit
margin and dividend payout ratios. For our example:

$$
\text{EFN} = \frac{\$3,000}{1,000} \times \$250 - \frac{\$300}{1,000}
  \times \$250 - .132 \times \$1,250 \times \left(1 - \frac{1}{3} \right)
  = \$565
$$

#### A particular scenario

The current projection of 25 percent sales growth will require \$565 in new
financing. Now management will need to develop a strategy to raise the funds for
the growth. There are three possible sources of new funding: short-term
borrowing, long-term borrowing, and new equity. Management can choose some
combination of any of these.

For our example, suppose management has a policy of not issuing new equity.
Management would also like to keep total net working capital unchanged. In this
scenario, with current assets increasing by \$300 and accounts payable only
increasing by \$75, management could borrow \$300 - \$75 = \$225 in short-term
notes payable. This would leave \$565 - 225 = \$340 coming from long-term
borrowing. Below is the updated pro forma balance sheet.

```{r financial models; rosengarten corp; scenario, echo=F}
# remove the efn row
tmp <- matrix(rosengarten$bs.proforma[-nrow(rosengarten$bs.proforma),1], ncol=1)
rownames(tmp) <- rownames(rosengarten$bs.proforma)[-15]
rosengarten$bs.scenario <- round(tmp)
rosengarten$bs.scenario["Notes payable",] <- 
    rosengarten$bs.scenario["Notes payable",] +
    225
rosengarten$bs.scenario["Total current liabilities",] <-
    rosengarten$bs.scenario["Total current liabilities",] +
    225
rosengarten$bs.scenario["Long-term debt",] <-
    rosengarten$bs.scenario["Long-term debt",] +
    340
rosengarten$bs.scenario["Total liabilities and owners' equity",] <-
    rosengarten$bs.scenario["Total current liabilities",] +
    rosengarten$bs.scenario["Long-term debt",] +
    rosengarten$bs.scenario["Total equity",]
rosengarten$bs.scenario.delta <-
    round(rosengarten$bs.scenario - rosengarten$bs[,1])
rosengarten$bs.scenario <- cbind(rosengarten$bs.scenario,
                                 rosengarten$bs.scenario.delta)
knitr::kable(pretty(rosengarten$bs.scenario),
     col.names=c("Item", "Amount ($)", "Change from current year"), align="lrr",
     caption="Rosengarten Corporation pro forma balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Current assets", 1, 4, bold=T, background="#f2f2f2") %>%
   pack_rows("Fixed assets", 5, 5, bold=T, background="#D3E4CD") %>%
   pack_rows("Current liabilities", 7, 9, bold=T, background="#FAD4D4") %>%
   pack_rows("Owners' equity", 11, 13, bold=T, background="#D9E4EC") %>%
   row_spec(5, extra_latex_after = "\\addlinespace") %>%
   row_spec(6, extra_latex_after = "\\addlinespace") %>%
   row_spec(13, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(4, 9, 13))
```

In this scenario, we used a combination of short- and long-term debt as the
plug. However, this is just one possible strategy; not necessarily the best one.

## External financing and growth

External financing and growth are related. The higher the rate of growth in
sales or assets, the greater the need for external financing.

### EFN and growth

We will use the following simplified income statement and balance sheet.

```{r external financing and growth; hoffman company, echo=F}
hoffman <- new.env()
hoffman$is <- matrix(c(500, 400, 100, 34, 66, 22, 44), ncol=1)
hoffman$is.accounts <- c("Sales", "Costs", "Taxable income", "Taxes (34%)",
                         "Net income", "Dividend",
                         "Additions to retained earnings")
rownames(hoffman$is) <- hoffman$is.accounts
knitr::kable(pretty(hoffman$is),
             caption="Hoffman Company income statement") %>%
    add_indent(c(6,7))

hoffman$bs <- data.frame(
  Category = c("Current assets", "Net fixed assets", "Total assets",
               "Total debt", "Owners' equity",
               "Total liabilities and owners' equity"),
  Amount = c(200, 300, 500, 250, 250, 500),
  pctSales = c(.4, .6, 1, NA, NA, NA)
)

knitr::kable(pretty(hoffman$bs),
     col.names=c("Item", "Amount ($)", "% of sales"), align="lrr",
     caption="Hoffman Corporation balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Assets", 1, 3, bold=T, background="#f2f2f2") %>%
   pack_rows("Liabilities and Owners' Equity", 4, 6, bold=T,
             background="#D3E4CD") %>%
   row_spec(3, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(3,6))
```

Notice that short- and long-term liabilities have been combined into a single
_Total debt_ figure. We are assuming that none of the current liabilities vary
spontaneously with sales. This assumption isn't as restrictive as it sounds. If
any current liabilities (such as accounts payable) vary with sales, we can
assume that any such accounts have been netted out in current assets.

If Hoffman Company is forecasting next year's sales level at \$600, a \$100
increase, then the percentage increase in sales is \$100/500 = 20 percent.

Using the percentage of sales approach we create the pro forma financial
statements:

```{r external financing and growth; hoffman company pro forma, echo=F}
hoffman$is.proforma <- hoffman$is * 1.2
knitr::kable(pretty(hoffman$is),
             caption="Hoffman Company pro forma income statement")

hoffman$bs.proforma <- data.frame(
  Category = c(hoffman$bs$Category),
  #          Assets all vary with sales;    Total Debt unchanged
  Amount = c(hoffman$bs$Amount[1:3] * 1.2, hoffman$bs$Amount[4],
  #           Add addition to retained earnings to owners' equity
              hoffman$bs$Amount[5] + hoffman$is.proforma[7, 1],
  #           Do the same with total liabilities and owners' equity
              hoffman$bs$Amount[6] + hoffman$is.proforma[7, 1]),
  pctSales = hoffman$bs$pctSales
)
efn <- hoffman$bs.proforma[3,2] - hoffman$bs.proforma[6,2]
hoffman$bs.proforma <- rbind(hoffman$bs.proforma,
  c("External financing needed", efn, NA))

knitr::kable(pretty(hoffman$bs.proforma),
     col.names=c("Item", "Amount ($)", "% of sales"), align="lrr",
     caption="Hoffman Corporation pro forma balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Assets", 1, 3, bold=T, background="#f2f2f2") %>%
   pack_rows("Liabilities and Owners' Equity", 4, 6, bold=T,
             background="#D3E4CD") %>%
   row_spec(c(3,6), extra_latex_after = "\\addlinespace") %>%
   add_indent(c(3,6))
```

At 20 percent growth, Hoffman will need \$100 in new assets. The projected
addition to retained earnings is \$52.8, so the external financing needed, EFN,
is \$100 - 52.8 = \$47.2.


Below shows EFN for several different growth rates. The projected addition to
retained earnings and the projected debt-equity ratio for each scenario are also
given. In determining the debt-equity ratios, we assumed that any needed funds
were borrowed, and we also assumed any surplus funds were used to pay off debt.
Thus, for the zero growth case the debt falls by \$44, from \$250 to \$206.

```{r external financing and growth; growth and projected EFN, echo=F}
efn <- new.env()
efn$sales.growth <- c(0, .05, .1, .15, .20, .25)
efn$assets.increase <- 500 * efn$sales.growth
efn$addition.retained.earnings <- 44 + (44 * efn$sales.growth)
efn$efn <- (500 + efn$assets.increase) - (500 + efn$addition.retained.earnings)
# We assume that all needed funds are borrowed
efn$de.ratio <- round((250 + efn$efn) /
                      (250 + efn$addition.retained.earnings),2)
efn$growth.table <- data.frame(
  Sales.Growth = efn$sales.growth,
  Assets.Increase = efn$assets.increase,
  Addition.Retained.Earnings = efn$addition.retained.earnings,
  External.Financing.Needed = efn$efn,
  DE.Ratio = efn$de.ratio
)
efn$gt.cols <- c("Projected Sales Growth", "Increase in Assets Required",
                 "Addition to Retained Earnings",
                 "External Financing Needed, EFN",
                 "Projected Debt-Equity Ratio")
knitr::kable(efn$growth.table, col.names=efn$gt.cols)
```

### Financial policy and growth

```{r external financing and growth; asset vs growth, echo=F}
plot(efn$sales.growth*100, efn$assets.increase, type="l",
     xlab="Projected growth in sales (%)",
     ylab="Asset needs and retained earnings ($)",
     col="lightblue",
     main="Growth and related financing needed for the Hoffman Company")
lines(efn$sales.growth*100, efn$addition.retained.earnings,
      col="#FF6666")
```

#### The internal growth rate

The maximum growth rate that can be achieved with no external financing of any
kind. In the figure above, the internal growth rate is the point where the
increase in assets required line (blue), and the projected addition to retained
earnings line (red) intersect. This is defined:

$$
\text{Internal growth rate} = \frac{\text{ROA} \times b}
  {1 - \text{ROA} \times b}\tag{3.24}
$$

ROA is the return on assets and $b$ is the plowback ratio.

For the Hoffman Company, net income was \$66 and total assets were \$500. ROA is
$66/500 = .132$ or 13.2\%. Of the \$66 net income, \$44 was retained, so the
plowback ratio, $b$, is $44/66 = 2/3$.

$$
\text{Internal growth rate} = \frac{.132 \times (2/3)}{1 - .132 \times (2/3)}
  = .0965
$$

Thus, the Hoffman Company can expand at a maximum rate of 9.65\% per year
without external financing.

#### The sustainable growth rate

The next growth rate is the maximum growth rate a firm can achieve with no
external _equity_ financing while it maintains a constant debt-equity ratio.

$$
\text{Sustainable growth rate} = \frac{\text{ROE} \times b}
  {1 - \text{ROE} \times b}\tag{3.25}
$$

This is identical to the internal growth rate except we use ROE instead of ROA.

For the Hoffman Company, net income was \$66 and total equity was \$250; ROE is
thus $66/250 =.264$ or 26.4\%. The plowback ratio, $b$ is still 2/3.

$$
\text{Sustainable growth rate} = \frac{.264 \times (2/3)}{1 = .264 \times (2/3)}
  = .2136
$$

Thus, the Hoffman Company can expand at a maximum rate of 21.36 percent per year
without external equity financing.

#### Determinants of growth

Remember that ROE can be decomposed into its various components using the DuPont
identity. These factors that are important in determining ROE are also important
factors for determining growth.

$$
\text{ROE} = \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier}
$$

The firm's ability to sustain growth depends explicitly on the following four
factors:

1. _Profit margin_: An increase in profit margin will increase the firm's
   ability to generate funds internally and thereby increase its sustainable
   growth.

2. _Dividend policy_: A decrease in the percentage of net income paid out as
   dividends will increase the retention ratio. This increases internally
   generated equity and thus increases sustainable growth.

3. _Financial policy_: An increase in the debt-equity ratio increases the firm's
   financial leverage. Because this makes additional debt financing available,
   it increases the sustainable growth rate.

4. _Total asset turnover_: An increase in the firm's total asset turnover
   increases the sales generated for each dollar in assets. This decreases the
   firm's need for new assets as sales growth and thereby increases the
   sustainable growth rate. Notice that increasing total asset turnover is the
   same thing as decreasing capital intensity (because they are recipricals).

Given values for all four of these, there is only one growth rate that can be
achieved.

__NOTE:__

> __If a firm does not wish to sell new equity and its profit margin, dividend
> policy, financial policy, and total asset turnover (or capital intensity) are
> all fixed, then there is only one possible growth rate.__

### A note about sustainable growth rate calculations

Very commonly, the sustainable growth rate is calculated using just the
numerator in our expression, ROE $\times b$. This causes some confusion. The
issue has to do with how ROE is computed. Recall that ROE is calculated as net
income divided by total equity. If total equity istaken from an ending balance
sheet, then our formula is the right one. However, if total equity is from the
beginning of the period, then the simpler formula is the correct one.

In principle, you'll get exactly the same sustainable growth rate regardless of
which way you calculate it (as long as you match up the ROE calculation with the
right formula). In reality, you may see some differences because of
accounting-related complications.

If you use the average of beginning and ending equity, you will need yet another
formula.

This also applies to the internal growth rate as well.

## Some caveats regarding financial planning models

Financial planning models do not always ask the right questions. A primary
reason is that they tend to rely on accounting relationships and not financial
relationships. In particular, the three basic elements of firm value tend to get
left out---namely, cash flow size, risk, and timing.

Because of this, financial planning models sometimes do not produce output that
gives the user many meaningful clues about what strategies will lead to
increases in value. Instead, they divert the user's attention to questions
concerning the association of, say, the debt-equity ratio and firm growth.

Financial planning is an iterative process. Plans are created, examined, and
modified over and over. The final plan will be a result negotiated between all
the different parities to the process. In fact, long-term financial planning in
most corporations relies on what might be called the Procrustes approach.
Upper-level management has a goal in mind, and it is up to the planning staff to
rework and ultimately deliver a feasible plan that meets that goal.

---

# Examples

## Example 3.1: Current events

Suppose a firm were to pay off some of its suppliers and short-term creditors.
What would happen to the current ratio? Suppose a firm buys some inventory. What
happens in this case? What happens if a firm sells some merchandise?

The first case is a trick question. What happens is that the current ratio moves
away from 1. If it is greater than 1, it will get bigger, but if it is less than
1, it will get smaller. To see this, suppose the firm has \$4 in current assets
and \$2 in current liabilities for a current ratio of 2. If we use \$1 in cash
to reduce current liabilities, the new current ratio is
$(\$4 - 1)/(\$2 - 1) = 3$. If we reverse the original situation to \$2 in
current assets and \$4 in current liabilities, the change will cause the current
ratio to fall to 1/3 from 1/2.

The second case is not quite as tricky. Nothing happens to the current ratio
because cash goes down while inventory goes up---total current assets are
unaffected.

In the third case, the current ratio would usually rise because inventory is
normally shown at cost and the sale would normally be at something greater than
cost (the difference is the markup). The increase in either cash or receivables
is therefore greater than the decrease in inventory. This increases current
assets, and the current ratio rises.

## Example 3.2: Payables turnover

Here is a variation on the receivables collection period. How long, on average,
does it take for Prufrock Corporation to _pay_ its bills? To answer, we need to
calculate the accounts payable turnover rate using cost of goods sold. We will
assume that Prufrock purchases everything on credit.

The cost of goods sold is \$1,344, and accounts payable are \$344. The turnover
is therefore \$1,344/\$344 = 3.9 times. So, payables turned over about every
365/3.9 = 94 days. On average, then, Prufrock takes 94 days to pay. As a
potential creditor, we might take note of this fact.

## Example 3.3: More turnover

Suppose you find that a particular company generates \$.40 in annual sales for
every dollar in total assets. How often does this company turn over its total
assets?

The total asset turnover here is .40 times per year. It takes 1/.40 = 2.5 years
to turn assets over completely.

## Example 3.4: Financial ratios

Consider the following 2010 data for Lowe's Companies and Home Depot (billions
except for price per share):

```{r example 3.4, echo=F}
ex3.4 <- new.env()

ex3.4$lowes <- c(48.8, 3.2, 2.0, 0.7, 1.6, 6.5, 33.7, 25, 1.3, 18.1)
ex3.4$home.depot <- c(68.0, 5.8, 3.3, 0.5, 1.6, 8.7, 40.1, 37, 1.6, 18.9)

ex3.4$rows <- c("Sales", "EBIT", "Net income", "Cash", "Depreciation",
                "Interest bearing debt", "Total assets", "Price per share",
                "Shares outstanding", "Shareholder equity")
ex3.4$col <- c("Lowe's Company, Inc.", "The Home Depot, Inc.")

ex3.4$tbl <- matrix(c(ex3.4$lowes, ex3.4$home.depot), ncol=2)
rownames(ex3.4$tbl) <- ex3.4$rows
colnames(ex3.4$tbl) <- ex3.4$col
knitr::kable(ex3.4$tbl, align="cc")

ex3.4$equity.multiplier <- round(ex3.4$tbl["Total assets",] /
    ex3.4$tbl["Shareholder equity",],1)
ex3.4$total.asset.turnover <- round(ex3.4$tbl["Sales",] /
    ex3.4$tbl["Total assets",],1)
ex3.4$profit.margin <- round(ex3.4$tbl["Net income",] /
    ex3.4$tbl["Sales",],3)
ex3.4$roe <- round(ex3.4$tbl["Net income",] /
    ex3.4$tbl["Shareholder equity",],3)
ex3.4$market.capitalization <- round(ex3.4$tbl["Price per share",] *
    ex3.4$tbl["Shares outstanding",],1)
ex3.4$enterprise.value <- round(ex3.4$market.capitalization +
    ex3.4$tbl["Interest bearing debt",] - ex3.4$tbl["Cash",],1)
ex3.4$eps <- round(ex3.4$tbl["Net income",] / 
    ex3.4$tbl["Shares outstanding",],2)
ex3.4$pe.multiple <- round(ex3.4$tbl["Price per share",] / ex3.4$eps,1)
ex3.4$ebitda <- round(ex3.4$tbl["EBIT",] + ex3.4$tbl["Depreciation",],1)
ex3.4$ev.multiple <- round(ex3.4$enterprise.value / ex3.4$ebitda,1)
```

#### 1.

Determine the profit margin, ROE, market capitalization, enterprise value, PE
multiple, and EV multiple for both Lowe's and Home Depot.

$$
\begin{aligned}
\ \ & & \text{Lowe's Company, Inc.} & & \text{The Home Depot, Inc.}\\\\
\text{Equity multiplier} & & `r ex3.4$tbl["Total assets", 1]`
  / `r ex3.4$tbl["Shareholder equity", 1]` = `r ex3.4$equity.multiplier[1]`
  & & `r ex3.4$tbl["Total assets", 2]`
  / `r ex3.4$tbl["Shareholder equity", 2]` = `r ex3.4$equity.multiplier[2]`\\
\text{Total asset turnover} & & `r ex3.4$tbl["Sales",1]`
  / `r ex3.4$tbl["Total assets",1]` = `r ex3.4$total.asset.turnover[1]`
  & & `r ex3.4$tbl["Sales",2]`
  / `r ex3.4$tbl["Total assets",2]` = `r ex3.4$total.asset.turnover[2]`\\
\text{Profit margin} & & `r ex3.4$tbl["Net income",1]`
  / `r ex3.4$tbl["Sales",1]` = `r ex3.4$profit.margin[1]*100`\%
  & & `r ex3.4$tbl["Net income",2]`
  / `r ex3.4$tbl["Sales",2]` = `r ex3.4$profit.margin[2]*100`\%\\
\text{ROE} & & `r ex3.4$tbl["Net income",1]`
  /`r ex3.4$tbl["Shareholder equity",1]` = `r ex3.4$roe[1]*100`\%
  & & `r ex3.4$tbl["Net income",2]`
  /`r ex3.4$tbl["Shareholder equity",2]` = `r ex3.4$roe[2]*100`\%\\
\text{Market capitalization} & & `r ex3.4$tbl["Price per share",1]`
  \times `r ex3.4$tbl["Shares outstanding",1]` = `r ex3.4$market.capitalization[1]`
  & & `r ex3.4$tbl["Price per share",2]`
  \times `r ex3.4$tbl["Shares outstanding",2]` = `r ex3.4$market.capitalization[2]`\\
\text{Enterprise value} & & `r ex3.4$market.capitalization[1]`
  + `r ex3.4$tbl["Interest bearing debt",1]` - `r ex3.4$tbl["Cash",1]`
  = `r ex3.4$enterprise.value[1]`
  & & `r ex3.4$market.capitalization[2]`
  + `r ex3.4$tbl["Interest bearing debt",2]` - `r ex3.4$tbl["Cash",2]`
  = `r ex3.4$enterprise.value[2]`\\
\text{PE multiple} & & `r ex3.4$tbl["Price per share",1]`
  / `r ex3.4$eps[1]` = `r ex3.4$pe.multiple[1]`
  & & `r ex3.4$tbl["Price per share",2]`
  / `r ex3.4$eps[2]` = `r ex3.4$pe.multiple[2]`\\
\text{EBITDA} & & `r ex3.4$tbl["EBIT",1]` + `r ex3.4$tbl["Depreciation",1]`
  = `r ex3.4$ebitda[1]`
  & & `r ex3.4$tbl["EBIT",2]` + `r ex3.4$tbl["Depreciation",2]`
  = `r ex3.4$ebitda[2]`\\
\text{EV multiple} & & `r ex3.4$enterprise.value[1]` / `r ex3.4$ebitda[1]`
  = `r ex3.4$ev.multiple[1]`
  & & `r ex3.4$enterprise.value[2]` / `r ex3.4$ebitda[2]`
  = `r ex3.4$ev.multiple[2]`
\end{aligned}
$$

#### 2.

How would you describe these two companies from a financial point of view?
Overall, they are similarly situated. In 2010, Home Depot had a higher ROE
(partially because of a higher total asset turnover and a higher profit margin),
but Lowe's had slightly lower PE and EV multiples. Both companies' PE multiples
were somewhat above the general market, indicating possible future growth
prospects.

## Example 3.5: Sustainable growth

Suppose Hoffman grows at exactly the sustainable growth rate of 21.36 percent.
What will the pro forma statements look like?

At a 21.36 percent growth rate, sales will rise from \$500 to \$606.8. The pro
forma income statement will look like this:

```{r example 3.5; income statement, echo=F}
ex3.5 <- new.env()
ex3.5$is.proforma <- matrix(c(606.8, 485.4, 121.4, 41.3, 80.1, 26.7, 53.4),
                            ncol=1)
ex3.5$is.rows <- c("Sales (Projected)", "Costs (80% of sales)",
                   "Taxable income", "Taxes (34%)", "Net income",
                   "Dividends", "Addition to retained earnings")
rownames(ex3.5$is.proforma) <- ex3.5$is.rows
knitr::kable(ex3.5$is.proforma,
    caption="Hoffman Company pro forma income statement") %>%
  add_indent(c(6,7))
```

We construct the balance sheet just as we did before. Notice, in this case that
owners' equity will rise from \$250 to \$303.4 because the addition to retained
earnings is \$53.4.

```{r example 3.5; balance sheet, echo=F}
ex3.5$bs.cols <- c("$", "% of sales")
ex3.5$bs.assets <- matrix(c(242.7, 364.1, 606.8, .4, .6, 1), ncol=2)
ex3.5$bs.assets.rows <- c("Current assets", "Net fixed assets", "Total assets")
rownames(ex3.5$bs.assets) <- ex3.5$bs.assets.rows
colnames(ex3.5$bs.assets) <- ex3.5$bs.cols
ex3.5$bs.lande <- matrix(c(250.0, 303.4, 553.4, 53.4, NA, NA, NA, NA), ncol=2)
ex3.5$bs.lande.rows <- c("Total debt", "Owners' equity",
  "Total liabilities and owners' equity", "External financing needed")
rownames(ex3.5$bs.lande) <- ex3.5$bs.lande.rows
colnames(ex3.5$bs.lande) <- ex3.5$bs.cols
ex3.5$bs.display <- list(ex3.5$bs.assets, ex3.5$bs.lande)
knitr::kable(ex3.5$bs.display,
    caption="Hoffman Company pro forma balance sheet", align="cc") %>%
  add_indent(c(3)) %>%
  kable_styling(latex_options = "HOLD_position")
```

As illustrated, EFN is \$53.4. If Hoffman borrows this amount, then total debt
will rise to \$303.4 and the debt-equity ratio will be exactly 1.0, which
verifies our earlier calculation. At any other growth rate, something would
have to change.

## Example 3.6: Profit margins and sustainable growth

The Sandar Co. has a debt-equity ratio of .5, a profit margin of 3 percent, a
dividend payout ratio of 40 percent, and a capital intensity ratio of 1. What is
its sustainable growth rate? If Sandar desired a 10 percent sustainable growth
rate and planned to achieve this goal by improving profit margins, what would
you think?

ROE is .03 $\times$ 1 $\times$ 1.5 = 4.5 percent. The retention ratio is
1 - .40 = .60. Sustainable growth is thus
.045(.60)/[1 - .045(.60)] = 2.77 percent.

For the company to achieve a 10 percent growth rate, the profit margin will have
to rise. To see this, assume that sustainable growth is equal to 10 percent and
then solve for profit margin, PM:

$$
\begin{aligned}
.10 = \text{PM}(1.5)(.6)/[1 - \text{PM}(1.5)(.6)]\\
\text{PM} = .1/.99 = 10.1\%
\end{aligned}
$$

For the plan to succeed, the necessary increase in profit margin is
substaintial, from 3 percent to about 10 percent. This may not be feasible.

---

# Chapter summary

This chapter focuses on working with information contained in financial
statements Specifically, we studied standardized financial statements, ratio
analysis, and long-term financial planning.

1. We explained that differences in firm size make it difficult to compare
   financial statements, and we discussed how to form common-size statements to
   make comparisons easier and more meaningful.

2. Evaluating ratios of accounting numbers is another way of comparing financial
   statment information. We defined a number of the most commonly used ratios,
   and we discussed the famous DuPont identity.

3. We showed how pro forma financial statements can be generated and used to
   plan for future financing needs.

After you have studied this chapter, we hope that you have some perspective on
the uses and abuses of financial statement information. You should also find
that your vocabulary of business and financial terms have grown substantially.

---

# Concepts

## 1. Financial ratio analysis

A financial ratio by itself tells us little about a company because financial
ratios vary a great deal across industries. There are two basic methods for
analyzing financial ratios for a company: Time trend analysis and peer group
analysis. In time trend analysis, you find the ratios for the company over some
period, say five years, and examine how each ratio has changed over this period.
In peer group analysis, you compare a company's financial ratios to those of its
peers. Why might each of these analysis methods be useful? What does each tell
you about the company's financial health?

---

#### Time trend analysis

Time trend analysis is useful because it provides insights into a company's
performance over time. By examining financial ratios across a historical period
(e.g., five years), you can identify trends, such as:

* __Improvement or deterioration__: For instance, a steadily improving profit
  margin suggests better cost control or increasing revenue efficiency, whereas
  a declining ratio might signal inefficiencies or problems.

* __Seasonal patterns__: Some businesses are cyclical, and observing trends over
  time can help understand seasonal impacts.

* __Long-term strategy effects__: Changes in ratios might reflect the outcomes
  of strategic decisions, such as entering new markets, adjusting pricing
  strategies, or adopting cost-cutting measures.

What time trend analysis reveals about the financial health of the firm is:

* __Sustainability__: Whether improvements or declines are part of a sustainable
  trend.

* __Early warning signs__: Sudden deviations in ratios (e.g., liquidity or
  leverage ratios) could indicate upcoming financial challenges.

* __Growth trajectory__: Trends in profitability and efficiency ratios can show
  whether the company is scaling effectively.

#### Peer group analysis

Peer group analysis involves comparing a company's financial ratios with those
of similar firms in the same industry. It provides context for understanding
performance by showing whether a company is operating competitively.

Peer group analysis will reveal:

1. A company's performance with respect to its peers.
2. The company's market position with respect to its peers.
3. Shows a firm's competitive advantage and weaknesses with its peers.

Both methods are complementary. Time trend analysis provides insights into how
the company is evolving internally. Peer group analysis situates the company's
performance within the competitive landscape.

Using both methods together offers a holistic view of a company's financial
health, allowing for better-informed decisions by combining internal trends with
external comparisons.

---

## 2. Industry-specific ratios

So-called "same-store sales" are a very important measure for companies as
diverse as McDonald's and Sears. As the name suggests, examining same-store
sales means comparing revenues from the same stores or restaurants at two
different points in time. Why might companies focus on same-store sales rather
than total sales?

---

Companies often focus on __same-store sales__ rather than total sales because
same-store sales provide a more accurate and meaningful measure of a company's
__organic growth and performance__ over time, excluding the effects of new store
openings or closures. Same-store sales can also give the company insight into
the operating performance of each of its locations.

#### Measures organic growth

__Same-store sales__ focus solely on stores that have been operational for a
certain period (usually more than a year), eliminating the impact of new store
openings, expansions, or closures. This metric highlights whether existing
stores are attracting more customers, selling more products, or increasing
revenues through better pricing strategies. For example: A 10% increase in total
sales could be misleading if it's driven entirely by new stores, while
same-store sales growth indicates improvements in the performance of existing
locations.

#### Excludes non-comparable factors

Total sales can be influenced by non-operational factors such as: new store
openings or acquisitions, which can inflate growth artificially; store closures,
which might mask declines in performance. Same-store sales provide a
"like-for-like" comparison, removing these distortions.

#### Indicates customer loyalty and market trends

Growth in same-store sales often reflects higher customer retention, increased
traffic, or greater spending per customer at established locations. Declines in
same-store sales could signal market saturation, poor customer satisfaction, or
increased competition.

#### Helps evaluate management effectiveness

By focusing on same-store sales, management can assess the effectiveness of
marketing campaigns, operational improvements, and product offerings within the
existing store base. It's a key indicator of how well the company is utilizing
its current resources to drive growth.

---

## 3. Sales forecast

Why do you think most long-term financial planning begins with sales forecasts?
Put differently, why are future sales the key input?

---

Most long-term financial planning begins with __sales forecasts__ because sales
are the primary driver of a company's operations and financial performance.
Future sales serve as the foundation for estimating many other critical
financial elements, making them the key input for planning.

Sales drive revenues and cash flows. Sales are the top line of the income
statement, and all other financial metrics---such as costs, profits, and cash
flows---stem from this figure. Future sales projections determine expected
revenue, which is critical for estimating future earnings and cash availability.

Companies use sales forecasts to plan for necessary investments in assets, such
as production capacity, equipment, or inventory. For instance: A forecast of
rising sales might prompt a company to expand facilities or hire more staff.
Declining sales forecasts could signal the need to reduce investment and
preserve cash.

Sales growth often requires financing to support operations, including working
capital, inventory, and capital expenditures. A robust sales forecast helps
determine how much financing the company will need and when, allowing for better
capital planning and debt management.

Sales forecasts align operational strategies, such as marketing campaigns,
product launches, or market expansion plans, with realistic revenue
expectations. For instance, if projected sales growth is aggressive, the company
might allocate more resources to marketing or innovation.

Sales forecasts are a key metric for communicating future performance
expectations to stakeholders, including investors, lenders, and employees. They
form the basis for convincing stakeholders of the company's growth potential and
ability to meet obligations.

---

## 4. Sustainable growth

In the chapter, we used Rosengarten Corporation to demonstrate how to calculate
EFN. The ROE for Rosengarten is about 7.3 percent, and the plowback ratio is
about 67 percent. If you calculate the sustainable growth rate for Rosengarten,
you will find it is only 5.14 percent. In our calculation for EFN, we used a
growth rate of 25 percent. Is this possible? (_Hint_: Yes, How?)

---

Yes, it is possible for a company like Rosengarten Corporation to have a
__growth rate__ significantly higher than its __sustainable growth rate__ (SGR),
such as 25% compared to 5.14%. However, achieving such a high growth rate
requires the company to __seek external financing__.

#### Sustainable growth rate defined

The __sustainable growth rate (SGR)__ is the maximum growth rate a company can
achieve using only internally generated funds (retained earnings) without
needing external financing. To calculate SGR:

$$
\mathbf{SGR} = \text{ROE} \times \text{Plowback Ratio}
$$

For Rosengarten:

$$
\mathbf{SGR} = 7.3\% \times 67\% = 5.14\%
$$

#### Growth beyond SGR

If the company wants to grow faster than its SGR (e.g., at 25%), it must rely on
__external financing__ to fund the additional growth. This external financing
could come from:

* __Debt__: Borrowing funds to invest in growth opportunities.

* __Equity__: Issuing new shares to raise captial.

* __Hybrid instruments__: Convertible debt or other forms of capital raising.

For example, to support a 25% growth rate: Rosengarten might need to
significantly increase its debt levels (which increases the financial leverage).
Alternatively, it could issue new equity, although this might dilute existing
shareholders' ownership.

#### Why external financing is needed

At a growth rate of 25%, the company's __assets__ will need to grow to support
the higher revenue. Since the company's retained earnings (from the 67% plowback
ratio) can only fund growth up to the SGR, the company must cover the gap with
external financing.

* __Required Asset Growth__: To support 25% revenue growth, assets typically
  need to grow by a similar percentage.

* __EFN (External Financing Needed)__: This represents the funds required to
  bridge the gap between the internally generated funds and the total funds
  needed for the desired growth.

#### Practical implications

* __Financial Risks__: Relying on external financing increases financial risks,
  especially if the company takes on excessive debt. High leverage can strain
  cash flows due to interest payments and principal repayments.

* __Market Perception__: Issuing equity might dilute ownership and potentially
  impact stock prices negatively.

* __Strategic Trade-offs__: The compaany must balance the benefits of high
  growth with the risks of relying on external funding.

#### Conclusion

While Rosengarten sustainable growth rate is only 5.14%, it can achieve a 25%
growth rate by raising external funds. This approach is feasible but requires
careful management of financial risks and strategic decisions about the type and
amount of external financing to pursue.

---

## 5. EFN and growth rate

Broslofski Co. maintains a positive retention ratio and keeps its debt-equity
ratio constant every year. When sales grow by 20 percent, the firm has a
negative projected EFN. What does this tell you about the firm's sustainable
growth rate? Do you know, with certainty, if the internal growth rate is greater
than or less than 20 percent? Why? WHat happens to the projected EFN if the
retention ratio is increased? What if the retention ratio is decreased? What if
the retention ratio is zero?

---

This situation with Broslofski Co. provides valuable insights into its
__sustainable growth rate (SGR)__, __internal growth rate (IGR)__, and how these
relate to projected external financing needs (EFN). Let's analyze each part:

#### What does a negative projected EFN tell you about the firm's sustainable growth rate?

A __negative EFN__ means that the company is generating more internal funds than
it needs to support its growth at the projected 20% sales increase. This
suggests that the firm's sustainable growth rate (SGR) is __grater than 20%__
because it can fund this growth internally while maintaining its desired capital
structure.

#### Is the internal growth rate greater than or less than 20 percent?

The __internal growth rate (IGR)__ is the maximum growth rate the firm can
achieve using only retained earnings, without borrowing additional funds or
issuing equity. Since the company maintains a constant debt-equity ratio and has
a __negative EFN__, it implies that the firm is not reliant on external
financing at the 20% growth rate. Therefore, the __IGR must be less than 20%__
because the firm's ability to grow solely using internal funds (IGR) is
inherently lower than its ability to grow with the help of leverage (SGR).

#### What happens to the projected EFN if the retention ratio is increased?

The increased retention ratio means the firm reinvests a larger portion of its
earnings back into the business, reducing dividends. This increases the amount
of internally generated funds available for growth, reducing EFN further. If EFN
was negative, it would become more negative, meaning the firm generates even
more surplus funds than needed fo rht eprojected growth.

#### What happens to the projected EFN if the retention ratio is decreased?

The decreased retntion ratio means the firm retains less of its earnings and
distributes more as dividends. This reduces the amount of internal funds
available for reinvestment, potentially increasing EFN. If the EFN was negative,
it might move closer to zero, indicating reduced surplus funds. If EFN becomes
positive, the firm would need external financing to support its growth.

#### What if the retention ratio is zero?

A zero retention ratio means the firm does not reinvest any of its earnings; all
profits are distributed as dividends. In this case: the firm is entirely reliant
on external financing (debt or equity) to fund growth; the EFN will likely
become positive, as there are no internal funds to support growth; and the
firm's IGR and SGR will also drop to zero, as both require retained earnings for
reinvestment.

#### Summary of the relationships

1. __SGR > 20%__: The firm's ability to grow sustainably exceeds 20%, indicated
   by the negative EFN.

2. __IGR < 20%__: The IGR is less than 20% because it represents growth
   achievable without any external leverage.

3. __Retention Ratio impact__: an increase means EFN becomes more negative (more
   surplus funds); a decrease means EFN moves closer to zero or becomes positive
   (less surplus funds); and if zero, EFN becomes positive and no internally
   generated funds are available to support growth.

This analysis highlights how retention policy and internal vs. external
financing decisions influence a firm's ability to sustain and project growth.

---

## 6. Common-size financials

One tool of financial analysis is common-size financial statements. Why do you
think common-size income statements and balance sheets are used? Note that the
accounting statement of cash flows is not converted into a common-size
statement. Why do you think this is?

---

#### Simplifies comparison across companies

Because each line item is expressed as a percentage of total sales for income
statements and percentage of total assets for balance sheets. This helps to
compare companies of different size, industry, or geographic location because
they eliminate the impact of scale.

#### Highlights financial trends

By converting financial statements into percentages, common-size analysis helps
identify trends over time, such as changes in cost structures, profitability, or
asset composition. For instance: a rising cost of goods sold (COGS) as a
percentage of sales might indicate worsening cost efficiency; or Increasing
current liabilities as a percentage of total assets could signal growing
liquidity risks.

#### Aids in benchmarking

Common-size statements facilitate comparisons against industry averages or peer
companies, helping analysts assess a company's competitive position. For
example, a company with a lower operating expense ratio than peers might have
superior cost control.

#### The cash flow statement is not common-sized because

1. __Lack of logical Base__ The cash flow statement tracks the sources and uses
   of cash across three categories: operating, investing, and financing
   activities. There is no single, consistent "base" to use for expressing cash
   flow items as percentages. Unlike sales in the income statement or total
   assets in the balance sheet, the cash flow statement doesn't have a
   comparable anchor figure.

2. __Focus on absolute values__ The cash flow statement's primary purpose is to
   show actual cash inflows and outflows, which are absolute values.
   Understanding whether the company generates enough cash to cover its
   obligations or fund growth is more important than relative percentages.

3.__Varied nature of cash flows__ The cash flow statement includes diverse line
items that don't naturally align with a common denominator: operating cash flow
links to profitability and working capital; investing cash flow reflects capital
expenditures and assets purchases or sales; financing cash flow deals with
borrowing, repaying debt, and equity transactions. These activities are not
proportional to a single metric like sales or total assets.

#### Conclusion

Common-size financial statements are invaluable for standardizing financial data
to facilitate analysis and comparison, particularly for the income statement and
balance sheet. However, the cash flow statement is best analyzed using absolute
figures, as it focuses on cash management rather than proportional
relationships.

---

## 7. Asset utilization and EFN

One of the implicit assumptions we made in calculating the external funds needed
was that the company was operating at full capacity. If the company is operating
at less than full capacity, how will this affect the external funds needed?

---

If a company is operating at less than full capacity, it has the potential to
increase production without immediately investing in additional fixed assets.
This can significantly affect the calculation of external funds needed (EFN):

#### Reduced need for fixed asset investment

At less than full capacity, the company can increase sales and production
without requiring proportional increases in fixed assets (e.g., machinery,
equipment, facilities). Since EFN calculations typically assume proportional
growth in all asset categories, this assumption would overestimate the need for
fixed asset investment in such a scenario. This would result in lower EFN
because the company doesn't need to spend as much on capital expenditures.

#### Implications for asset turnover

Operating below capacity implies that existing fixed assets are underutilized.
As production ramps up to meet increased sales, the fixed asset turnover ratio
improves (sales per unit of fixed assets), further reducing the need for
additional investments.

#### Working capital requirements still increase

Even if fixed asset investments are unnecessary, growth in sales will still
require additional current assets (e.g., inventory and accounts receivable) and
might increase current liabilities (e.g., accounts payable). However, these
changes are typically smaller than those involving fixed assets.

#### Adjusted EFN

If the company adjusts for underutilized capacity, then EFN will be calculated
based only on increases in current assets and current liabilities, excluding or
reducing the need for new fixed assets. This will result in a lower EFN compared
to the assumption of full capacity utilization.

#### Example

Assume sales grow by 20%, and the company is operating at 80% capacity. The
company can handle the 20% growth without additional fixed asset investment
because it only needs to utilize the remaining 20% capacity. EFN calculations
based on proportional growth would have overestimated the required increase in
fixed assets, inflating EFN unnecessarily.

#### Key takeaways

When operating below full capacity, EFN decreases because fixed asset
investments are reduced or unnecessary. Thus, the calculation must be adjusted
to reflect the reality of underutilized capacity. However, as the company
approaches full capacity, future growth will eventually require investment in
fixed assets, at which point EFN will increase.

Understanding capacity utilization ensures more accurate financial planning and
avoids overstating the need for external funding.

---

## 8. Comparing ROE and ROA

Both ROA and ROE measure profitability. Which one is more useful for comparing
two companies? Why?

---

Both return on assets (ROA) and return on equity (ROE) measure profitability,
but which one is more useful for comparing two companies depends on the context
and what you're trying to analyze.

#### ROA (return on assets)

Measures how efficiently a company uses its total assets to generate profit. ROA
also accounts for both equity and debt financing since assets are funded by
both.

$$
\mathbf{ROA} = \frac{\text{net income}}{\text{total assets}}
$$

You would use ROA to compare two companies when:

##### Capital structure differs

ROA provides a more comparable measure of operational efficiency because it
ignores the effects of leverage. For example, if one company is heavily
debt-financed and the other relies mostly on equity, ROA gives a better sense of
how well each uses its total assets to generate profits.

##### Assessing asset efficiency

ROA evaluates how effectively the companies deploy their assets, regardless of
how they are financed.

#### ROE (return on equity)

Focuses on the return generated on the shareholders' investment and reflects the
impact of leverage (debt) on profitability.

$$
\mathbf{ROE} = \frac{\text{net income}}{\text{shareholder's equity}}
$$

You would use ROE to compare two companies when:

##### Capital structure is similar

If two companies have similar levels of debt, ROE is more relevant because it
focuses on returns to shareholders.

##### Evaluating shareholder returns

ROE directly shows how much profit shareholders earn per dollar of equity
invested, which is key for equity investors.

#### Key considerations

##### Impact of leverage

High leverage increases ROE because debt financing boosts returns on equity (if
the cost of debt is lower than the return on assets). However, this can distort
comparisons between companies with different debt levels. ROA is unaffected by
leverage, making it a more neutral measure.

##### Industry differences

In asset-heavy industries (e.g., manufacturing), ROA is particularly relevant
for assessing operational efficiency. In industries where leverage is common
(e.g., banking), ROE might be might indicative of performance.

#### Conclusion

##### ROA

is more useful for comparing companies with different capital structures because
it measures asset efficiency independently of financing decisions.

##### ROE

is more relevant when comparing companies with similar debt levels or when
focusing on returns from a shoareholder's perspective.

Ultimately, understanding both metrics provides a more comprehensive view of
profitability, but the choice depends on what aspect of performance you wish to
compare.

---

## 9. Ratio analysis

Consider the ratio EBITD/Assets. What does this ratio tell us? Why might it be
more useful than ROA in comparing two companies?

---

The ratio EBITD/Assets (Earnings Before Interest, Taxes, and Deprecation divided
by Toatl Assets) provides insight into a company's ability to generate operating
cash flow relative to its total assets.

#### EBITD/Assets viewpoint

##### Operational efficiency

It measures how efficiently a company uses its assets to generate operating cash
flow, excluding non-cash charges (deprecation and amortization) and before the
effects of financing and taxes.

##### Cash flow focus

By using EBITD instead of net income, the ratio focuses on cash-based
profitability rather than accounting profitability, which can be affected by
non-cash expenses and financing decisions.

##### Asset Utilization

Like ROA, this ratio indicates how effectively a company's total assets are
utilized, but with an emphasis on cash-generating ability.

#### EBITD/Assets may be more useful in the following scenarios

##### Removes the effect of financing decisions

ROA uses net income in the numerator, which is influenced by interest expenses
(financing structure), while EBITD/Assets excludes interest, making it more
neutral for comparisons between companies with different levels of debt.

##### Excludes non-cash accounting effects

Deprecation and amortization are non-cash expenses that vary based on accounting
methods and asset age. EBITD excludes these charges, providing a better measure
of operational cash flow and enabling a more apples-to-apples comparison of
companies with different asset bases or deprecation policies.

##### Focuses on oprational performance

EBITD reflects the compnay's core operational performance before taxes and
financing costs. This makes it particularly useful for comparing companies in
capital-intensive industries where deprecation and amortization can
significantly distort net income.

##### Better for cross-industry or cross-region comparisons

ROA can be affected by tax rates or regional accounting practices, which vary
across industries or countries.  EBITD/Assets neutralizes these differences,
focusing on operational cash flow generation relative to assets.

#### Example comparison

Company A has high leverage, uses debt financing extensively which results in
high interest expenses. Company B has low leverage, primarily equit-financed,
with minimal interest expenses. EBITD/Assets allows for a direct comparison of
operational efficiency without being skewed by financing differences, whereas
ROA might make Company B appear more profitable soley due to its financing
structure.

#### Situations where EBITD/Assets excels

##### Capital-intensive industries

In industries like manufacturing or utilities, EBITD/Assets highlights
operational efficiency despite large deprecation charges.

##### High-leverage companies

Useful for comparing companies with varying levels of debt, as it excludes
interest effects.

##### Startups or Growth Companies

Young companies with large upfron capital investments but limited net income can
still show strong EBITD/Assets.

### Conclusion

EBITD/Assets is particularly useful when comparing companies with different
financing structures, deprecation policies, or tax environments. It provides a
clearer view of a company's ability to generate operating cash flow from its
assets, making more neutral and informative in specific contexts than ROA.

---

## 10. Return on investment

A ratio that is becoming more widely used is return on investment. Return on
investment is calculated as net income divided by long-term liabilities plus
equity. What do you think return on investment is intended to measure? What is
the relationship between return on investment and return on assets?

---

Return on investment (ROI) is calculated as

$$
\mathbf{ROI} = \frac{\text{net income}}{\text{long-term liabilitieis} + \text{equity}}
$$

This ratio is intended to measure the profitability of a company relative to its
total capital base, including bot debt (long-term liabilities) and equity.

#### ROI measures

##### Profitablitility of total capital

ROI indicates how effectively a company is generating net income from the
resources provided by both creditors (long-term debt) and owners (equity). It
evaluates the return on all long-term financing rather than just equity, making
it a broader profitability measure.

##### Financial leverage impact

Since long-term liabilities are included in the denominator, ROI accounts for
the impact of leverage on profitability, though it does not distinguish between
returns to debt and equity holders.

##### Capital efficiency

A higher ROI means the company is better at converting its long-term financing
into net income.

#### Relationship between ROI and ROA

Return on assets (ROA) is calculated as:

$$
\mathbf{ROA} = \frac{\text{net income}}{\text{total assets}}
$$

ROA uses total assets as the base, capturing the entire resource base employed
by the company. ROI uses long-term liabilities + equity, which represents the
company's permanent capital structure (ignoring current liabilities and
short-term financing). ROI can be thought of as a subset of ROA, focusing
specifically on returns from capital funded through long-term financing. For a
compnay with minimal current liabilities, ROI and ROA will be similar. However,
if the company has significant current liabilities, ROI will typically be higher
than ROA because the denominator (long-term liabilities + equity) is smaller
than total assets.

---

Use the following information to answer the next five questions: A small
business called the Grandmother Calendar Company began selling personalized
photo calendar kits. The kits were a hit, and sales soon sharply exceeded
forecasts. The rush of orders created a huge backlog, so the company leased more
space and expanded capacity, but it still could not keep up with demand.
Equipment failed from overuse and quality suffered. Working capital was drained
to expand production, and, at the same time, payments from customers were often
delayed until the product was shipped. Unable to deliver on orders, the company
became so strapped for cash that employee paychecks began to bounce. Finally,
out of cash, the company ceased operations entirely three years later.

---

## 11. Product sales

Do you think the company would have suffered the same fate if its product had
been less popular? Why or Why not?

---

While high demand is generally positive, in this case, the company's inability
to manage the operational and financial challenges of rapid growth led to its
failure. If its product had been less popular, the company might have grown at a
manageable pace, avoiding the operational and financial crises that ultimately
led to its closure.

---

## 12. Cash flow

The Grandmother Calendar Company clearly had a cash flow problem. In the context
of the cash flow analysis we developed in Chapter 2, what was the impact of
customers not paying until orders were shipped?

---

Customers delaying payments until orders were shipped created a significant cash
flow mismatch for the Grandmother Calendar Company. This issue compounded the
financial stress caused by high demand and operational challenges, ultimately
leading to a liquidity crisis. Managing cash flow by aligning inflows and
outflows is crucial for sustaining operations, particularly for growing
businesses like this one.

---

## 13. Corporate borrowing

If the firm was so successful at selling, why wouldn't a bank or some other
lender step in and provide it with the cash it needed to continue?

---

Despite strong sales, the Grandmother Calendar Company faced a combination of
operational inefficiencies, cash flow challenges, and poor financial management,
which likely made it appear too risky for banks or lenders. To secure funding,
the company would have needed to deomonstrate not only strong sales but also a
clear plan for addressing its liquidity issues, stabilizing operations, and
ensuring repayment ability. Without such assurances, successful sales alone
would not have been enough to secure additional financing.

---

## 14. Cash flow

Which was the biggest culprit here: Too  many orders, too little cash, or too
little production capacity?

---

While too many orders and too little production capacity were significant
challenges, too little cash was the biggest culprit. With better cash flow
management (e.g., requiring deposits, better aligning inflows and outflows, or
securing short-term financing), the company could have scaled its production
capacity and fulfilled orders. Ultimately, the cash shortage turned what could
have been a success story into a failure.

---

## 15. Cash flow

What are some actions a small company like The Grandmother Calendar Company can
take (besides expansion of capacity) if it finds itself in a situation in which
growth in sales outstrips production?

---

1. Manage demand strategically: adjust pricing, limit order intake, introduce
   pre-orders.

2. Outsource or partner: contract manufacturing or collaborate with competitors.

3. Optimize operations: improve production efficiency or focus on core products.

4. Enhance cash flow management: request customer deposits, offer discounts for
   early payments, negotiate supplier terms.

5. Prioritize customer satisfaction: communicate with customers, offer premium
   options.

6. Leverage technology: adopt inventory and order management software, forecast
   demand.

7. Use short-term financing: bridge loans or credit lines, factoring accounts
   receivable.

8. Explore alternative fulfillment models: Licensing or franchising, digital
   products or services.

---

# Questions and Problems

## 1. DuPont identity

If Jares, Inc., has an equity multiplier of 1.55, total asset turnover of 1.75,
and a profit margin of 4.3 percent, what is its ROE?

---

```{r problem 1, echo=F}
p1 <- new.env()
p1$equity.multiplier <- 1.55
p1$total.asset.turnover <- 1.75
p1$profit.margin <- .043
p1$roe <- p1$profit.margin * p1$total.asset.turnover * p1$equity.multiplier
```

$$
\begin{aligned}
\mathbf{ROE} &= \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier} \\
  &= `r p1$profit.margin` \times `r p1$total.asset.turnover`
  \times `r p1$equity.multiplier`\\
  &= `r round(p1$roe, 4)`
\end{aligned}
$$

---

## 2. Equity multiplier and return on equity

Nuber Company has a debt-equity ratio of .80. Return on assets is 9.7 percent,
and total equity is \$735,000. What is the equity multiplier? Return on equity?
Net income?

```{r problem 2, echo=F}
p2 <- new.env()
p2$de.ratio <- .8
p2$roa <- .097
p2$total.equity <- 735000
p2$equity.multiplier <- 1 + p2$de.ratio
p2$roe <- p2$roa * p2$equity.multiplier
p2$net.income <- p2$roe * p2$total.equity
```

$$
\begin{aligned}
\mathbf{Equity\ multiplier} &= 1 + \text{debt-equity ratio}\\
  &= 1 + `r p2$de.ratio`\\
  &= `r p2$equity.multiplier`\\\\
\mathbf{ROE} &= \text{ROA} \times \text{equity multiplier}\\
  &= `r p2$roa` \times `r p2$equity.multiplier`\\
  &= `r p2$roe`\\\\
\mathbf{Net\ income} &= \text{ROE} \times \text{total equity}\\
  &= `r p2$roe` \times `r p2$total.equity`\\
  &= `r pretty(p2$net.income)`\\\\
\end{aligned}
$$

---

## 3. Using the DuPont Identity

Y3K, Inc., has sales of \$2,700, total assets of \$1,310, and a debt-equity ratio
of 1.20. If its return on equity is 15 percent, what is its net income?

---

```{r problem 3, echo=F} 
p3 <- new.env()
p3$assets <- 1310
p3$de.ratio <- 1.2
p3$roe <- .15
p3$net.income <- (p3$roe * p3$assets)/(1 + p3$de.ratio)
```

Using the DuPont Identity we know:

$$
\text{ROE} = \frac{\text{net income}}{\text{assets}}
  \times (1 + \text{debt-equity ratio})
$$

Therefore, we can find `net income`:

$$
\begin{aligned}
\text{net income} &= \frac{\text{ROE} \times \text{assets}}
  {(1 + \text{debt-equity ratio})}\\
  &= \frac{`r p3$roe` \times `r pretty(p3$assets)`}
  {(1 + `r p3$de.ratio`)}\\
  &= `r round(p3$net.income, 2)`
\end{aligned}
$$

---

## 4. EFN

The most recent financial statements for Martin, Inc., are shown here:

```{r problem 4 financial statements, echo=F}
p4 <- new.env()

p4$is <- data.frame(
    Category = c("Sales", "Costs", "taxable income",
                 "Taxes (34%)", "Net income"),
    Amount = c(37300, 25800, 11500, 3910, 7590)
)

p4$bs <- data.frame(
    Category =  c("Assets", "Debt", "Equity"),
    Amount = c(127000, 30500, 96500)
)

knitr::kable(p4$is, caption="Income Statement")
knitr::kable(p4$bs, caption="Balance Sheet")

p4$dividend <- 2500
p4$projected.sales <- 42300
p4$payout.ratio <- p4$dividend / p4$is$Amount[5]
p4$growth.rate <- (p4$projected.sales - p4$is$Amount[1]) / p4$is$Amount[1]
p4$projected.assets <- p4$bs$Amount[1] * (1 + p4$growth.rate)
p4$projected.net.income <- p4$is$Amount[5] * (1 + p4$growth.rate)
p4$addition.to.retained.earnings <- p4$projected.net.income *
                                     (1 - p4$payout.ratio)
p4$projected.equity <- p4$bs$Amount[3] + p4$addition.to.retained.earnings
p4$efn <- p4$projected.assets - (p4$bs$Amount[2] + p4$projected.equity)
```

Assets and costs are proportional to sales. Debt and equity are not. A dividend
of \$2,500 was paid, and Martin wishes to maintain a constant payout ratio. Next
year's sales are projected to be \$42,300. What external financing is needed?

---

Since Assets and costs are proportional to projected sales (\$42,300) for the
income statement we get:

$$
\begin{aligned}
\mathbf{Payout\ ratio} &= \frac{\text{dividends}}{\text{net income}}\\
  &= \frac{`r pretty(p4$dividend)`}
          {`r pretty(p4$is$Amount[5])`}\\
  &\approx `r pretty(round(p4$payout.ratio, 4))`\\\\
\mathbf{Growth\ rate} &= \frac{\text{projected sales} - \text{current sales}}
  {\text{current sales}}\\
  &= \frac{`r pretty(p4$projected.sales)` - `r pretty(p4$is$Amount[1])`}
          {`r pretty(p4$is$Amount[1])`}\\
  &= `r round(p4$growth.rate, 4)`\\\\
\mathbf{Projected\ assets} &= \text{current assets}
  \times (1 + \text{growth rate})\\
  &= `r pretty(p4$bs$Amount[1])` \times `r 1 + p4$growth.rate`\\
  &= `r pretty(p4$projected.assets)`\\\\
\mathbf{Projected\ net\ income} &= \text{current net income}
  \times (1 + \text{growth rate})\\
  &= `r pretty(p4$is$Amount[5])` \times `r 1 + round(p4$growth.rate, 4)`\\
  &= `r pretty(round(p4$projected.net.income, 2))`\\\\
\mathbf{Addition\ to\ retained\ earnings} &= \text{net income}
  \times \text{retention raio}\\
  &= `r pretty(round(p4$projected.net.income, 2))`
  \times (1 - `r round(p4$payout.ratio, 4)`)\\
  &= `r pretty(round(p4$addition.to.retained.earnings, 2))`\\\\
\mathbf{Projected\ equity} &= \text{current equity}
  + \text{addition to retained earnings}\\
  &= `r pretty(p4$bs$Amount[3])` +
     `r pretty(round(p4$addition.to.retained.earnings))`\\
  &= `r pretty(round(p4$projected.equity, 2))`\\\\
\mathbf{EFN} &= \text{projected assets} - (\text{current debt}
  + \text{projected equity})\\
  &= `r pretty(p4$projected.assets)` - (`r pretty(p4$bs$Amount[2])`
  + `r pretty(p4$projected.equity)`)\\
  &= `r pretty(p4$efn)`
\end{aligned}
$$

```{r solution 4, echo=F}
pct.inc.sales <- 42300/37300
payout.ratio <- 2500/7590
assets.proforma <- 127000 * pct.inc.sales
netincome.proforma <- 7590 * pct.inc.sales
retained.earnings.proforma <- netincome.proforma*(1-payout.ratio)
equity.proforma <- 96500 + retained.earnings.proforma
efn <- assets.proforma - (30500 + equity.proforma)
```

---

## 5. Sales and growth

The most recent financial statements for Fontenot Co. are shown here:

```{r problem 5, echo=F}
p5 <- new.env()
p5$is <- matrix(c(54000, 39300, 14700, 4998, 9702), ncol=1)
row.names(p5$is) <- c("Sales", "Costs", "Taxable income",
                      "Taxes (34%)", "Net income")
p5$bs <- matrix(c(31000, 118000, 149000, 68000, 81000, 149000), ncol = 1)
row.names(p5$bs) <- c("Current assets", "Fixed assets", "Total Assets",
                      "Long-term debt", "Equity", "Debt and Equity")
knitr::kable(p5$is, caption="Income Statement")
knitr::kable(p5$bs, caption="Balance Sheet")
```

Assets and costs are proportional to sales. The company maintains a constant 30
percent dividend payout ratio and a constant debt-equity ratio. What is the
maximum increase in sales that can be sustained assuming no new equity is
issued?

---

```{r problem 5; work, echo=F}
p5$payout.ratio <- .3
p5$de.ratio <- as.numeric(p5$bs["Long-term debt",]) /
               as.numeric(p5$bs["Equity",])
p5$roe <- as.numeric(p5$is["Net income",]) /
          as.numeric(p5$bs["Equity",])
p5$plow.back <- 1 - p5$payout.ratio
p5$sgr <- (p5$roe * p5$plow.back) / (1 - p5$roe * p5$plow.back)
```

$$
\begin{aligned}
\mathbf{Retention\ ratio} &= 1 - \text{dividend payout ratio}\\
  &= 1 - \\
  &= 0.70\\\\
\mathbf{DER} &= \frac{\text{Long-term Debt}}{\text{Equity}} \\
  &= \frac{68,000}{81,000} \\
  &= 0.8395 \\ \\
\mathbf{ROE} &= \frac{\text{net income}}{\text{equity}} \\
  &= \frac{9,702}{81,000} \\
  &= 0.1198 \\ \\
\mathbf{SGR} &= \text{ROE} \times \text{retention ratio} \\
  &= 0.1198 \times 0.70 \\
  &= 0.0839 \\ \\
\mathbf{Maximum\ increase\ in\ sales} &= \text{current sales} \times \text{SGR} \\
  &= 54,000 \times 0.0839 \\
  &= 4,531 \\ \\
\mathbf{Projected\ sales} &= \text{current sales} + \text{maximum increase in sales} \\
  &= 54,000 + 4,531 \\
  &= 58,531
\end{aligned}
$$

\$4,531 is the largest sales growth Fontenot Co. can sustain without issuing new
equity while maintaining its current policies. The total projected sales is
\$58,531.

```{r solution 5}
p5$retention.ratio <- 0.70
p5$debt.to.equity.ratio <- 68000/81000
p5$roe <- 9702/81000
p5$sgr <- p5$roe * p5$retention.ratio
cat("Maximum increase in sales:", p5$max.inc.sales <- 54000 * p5$sgr)
cat("Projected Total Sales:", p5$projected.sales <- 54000 + p5$max.inc.sales)
```

---

## 6. Sustainable growth

If the Layla Corp. has a 13 percent ROE and a 20 percent payout ratio, what is
its sustainable growth rate.

---

```{r problem 6, echo=F}
p6 <- new.env()
p6$roe <- .13
p6$payout.ratio <- .2
p6$sgr <- (p6$roe * (1 - p6$payout.ratio)) /
          (1 - p6$roe * (1 - p6$payout.ratio))
```

$$
\begin{aligned}
\mathbf{SGR} &= \frac{\text{ROE} \times \text{retention ratio}}
     {1 - \text{ROE} \times \text{retention ratio}}\\
  &= \frac{`r p6$roe` \times (1 - `r p6$payout.ratio`)}
          {1 - `r p6$roe` \times (1 - `r p6$payout.ratio`)}\\
  &= `r round(p6$sgr, 4)`
\end{aligned}
$$

---

## 7. Sustainable growth

Assuming the following ratios are constant, what is the sustainable growth rate?

* Total asset turnover = 2.20
* Profit margin = 7.4%
* Equity multiplier = 1.40
* Payout ratio = 40%

---

```{r problem 7, echo=F}
p7 <- new.env()
p7$total.asset.turnover <- 2.2
p7$profit.margin <- .074
p7$equity.multiplier <- 1.4
p7$payout.ratio <- .4
p7$roa <- p7$profit.margin * p7$total.asset.turnover
p7$roe <- p7$roa * p7$equity.multiplier
p7$sgr <- p7$roe * (1 - p7$payout.ratio)
```

$$
\begin{aligned}
\mathbf{ROA} &= \text{profit margin} \times \text{total asset turnover} \\
  &= `r p7$profit.margin` \times `r p7$total.asset.turnover`\\
  &= `r p7$roa`\\\\
\mathbf{ROE} &= \text{ROA} \times \text{equity multiplier} \\
  &= `r p7$roa` \times `r p7$equity.multiplier`\\
  &= `r round(p7$roe, 4)`\\\\
\mathbf{Sustainable\ Growth\ Rate} &= \text{ROE} \times \text{retention ratio}\\
  &= \text{ROE} + (1 - \text{payout ratio}) \\
  &= `r p7$roe` + (1 - `r p7$payout.ratio`)\\
  &= `r round(p7$sgr, 4)`\\\\
\end{aligned}
$$

---

## 8. Calculating EFN

The most recent financial statements for Bradley, Inc., are shown here (assuming no income taxes):

```{r problem 8, echo=F}
p8 <- new.env()
p8$is <- matrix(c(6500, 5320, 1180), ncol=1)
row.names(p8$is) <- c("Sales", "Costs", "Net income")
p8$bs <- matrix(c(17400, 17400, 8400, 9000, 17400), ncol=1)
row.names(p8$bs) <- c("Assets", "Total Assets", "Debt", "Equity", "Debt and Equity")
knitr::kable(p8$is, caption="Income Statement")
knitr::kable(p8$bs, caption="Balance Sheet")
```

Assets and costs are proportional to sales. Debt and equity are not. No dividends are paid. Next year's sales are projected to be \$7,280. What is the external financing needed?

---

```{r problem 8; work, echo=F}
p8$projected.sales <- 7280
p8$growth <- p8$projected.sales / as.numeric(p8$is["Sales",])
p8$projected.assets <- as.numeric(p8$bs["Assets",]) * p8$growth
p8$projected.net.income <- as.numeric(p8$is["Net income",]) * p8$growth
p8$projected.equity <- as.numeric(p8$bs["Equity",]) + p8$projected.net.income
p8$efn <- p8$projected.assets - (as.numeric(p8$bs["Debt",])
                              + p8$projected.equity)
```

$$
\begin{aligned}
\mathbf{Growth\ rate} &= \frac{\text{projected sales} - \text{sales}}
  {\text{sales}}\\
  &= \frac{`r pretty(p8$projected.sales)`
  - `r pretty(as.numeric(p8$is["Sales",]))`}
          {`r pretty(as.numeric(p8$is["Sales",]))`}\\
  &= `r p8$growth - 1`\\
\mathbf{Projected\ assets} &= \text{assets} \times (1 + \text{growth rate})\\
  &= `r pretty(as.numeric(p8$bs["Assets",]))` \times `r p8$growth`\\
  &= `r pretty(p8$projected.assets)`\\\\
\mathbf{Projected\ net\ income} &= \text{net income}
  \times (1 + \text{growth rate})\\
  &= `r pretty(as.numeric(p8$is["Net income",]))` \times `r p8$growth`\\
  &= `r pretty(p8$projected.net.income)`\\\\
\mathbf{Projected\ equity} &= \text{equity} + \text{projected net income}\\
  &= `r pretty(as.numeric(p8$bs["Equity",]))` +
     `r pretty(as.numeric(p8$projected.net.income))`\\
  &= `r pretty(p8$projected.equity)`\\\\
\mathbf{EFN} &= \text{projected assets} - (\text{debt} + \text{projected equity})\\
  &= `r pretty(p8$projected.assets)` - (`r pretty(as.numeric(p8$bs["Debt",]))`
  +  `r pretty(p8$projected.equity)`)\\
  &= `r pretty(p8$efn)`
\end{aligned}
$$

---

## 9. External funds needed

Cheryl Colby, CFO of Charming Florist Ltd., has created the firm's pro forma
balance sheet for the next fiscal year. Sales are projected to grow by 10
percent to \$420 million. Current assets, fixed assets, and short-term debt are
20 percent, 75 percent, and 15 percent of sales, respectively. Charming Florist
pays out 30 percent of its net income in dividends. The company currently has
\$120 million of long-term debt and \$48 million in common stock par value. The
profit margin is 9 percent.

a. Construct the current balance sheet for the firm using the projected sales
   figure.

b. Based on Ms. Colby's sales growth forecast, how much does Charming Florist
   need in external funds for the upcoming fiscal year?

c. Construct the firm's pro forma balance sheet for the next fiscal year and
   confirm the external funds needed that you calculated in part (b).

---

```{r problem 9, echo=F}
options(digits=5)
p9 <- new.env()

# Given
p9$projected.sales <- 420
p9$sales.growth <- .1
p9$current.assets.pct.of.sales <- .2
p9$fixed.assets.pct.of.sales <- .75
p9$short.term.debt.pct.of.sales <- .15
p9$long.term.debt <- 120
p9$dividend.payout <- .3
p9$equity.par.value <- 48
p9$profit.margin <- .09

p9$sales <- p9$projected.sales / (1 + p9$sales.growth)

# Current variables
p9$current <- matrix(p9$sales * c(
    p9$current.assets.pct.of.sales,
    p9$fixed.assets.pct.of.sales,
    p9$short.term.debt.pct.of.sales), ncol=1)
rownames(p9$current) <- c(
    "Current assets",
    "Fixed assets",
    "Short-term debt")
p9$current.total.assets <- sum(p9$current[1:2,])
p9$current.total.liabilities <- as.numeric(p9$current[3,]) + p9$long.term.debt
p9$current.net.income <- p9$profit.margin * p9$sales
p9$current.retained.earnings <- p9$current.net.income *
  (1 - p9$dividend.payout)
p9$current.equity <- p9$current.total.assets - p9$current.total.liabilities

# Projected variables
p9$projected <- matrix(p9$projected.sales * c(
    p9$current.assets.pct.of.sales,
    p9$fixed.assets.pct.of.sales,
    p9$short.term.debt.pct.of.sales), ncol=1)
rownames(p9$projected) <- c(
    "Current assets",
    "Fixed assets",
    "Short-term debt")
p9$projected.total.assets <- sum(p9$projected[1:2,])
p9$projected.total.liabilities <- p9$projected[3,] + p9$long.term.debt
p9$projected.net.income <- p9$profit.margin * p9$projected.sales
p9$projected.retained.earnings <- p9$projected.net.income *
  (1 - p9$dividend.payout)
p9$projected.equity <- p9$current.equity + p9$projected.retained.earnings

p9$total.available.financing <- p9$projected["Short-term debt",] +
    p9$long.term.debt + p9$projected.equity

p9$efn <- p9$projected.total.assets - p9$total.available.financing
```

### A. Constructing the current balance sheet based on current sales.

#### NOTE:

All values are in millions.

$$
\begin{aligned}
\text{Current sales} &= \frac{\text{Projected sales}}{1 + \text{sales growth}}\\
  &= \frac{`r pretty(p9$projected.sales)`}{`r 1 + p9$sales.growth`}\\
  &= `r pretty(p9$sales)`\\\\
\mathbf{Assets}\\
\text{Current assets} &= \text{current sales}
  \times \text{current assets as percent of sales}\\
  &= `r pretty(p9$sales)` \times `r p9$current.assets.pct.of.sales`\\
  &= `r pretty(p9$current["Current assets",])`\\\\
\text{Fixed assets} &= \text{current sales}
  \times \text{fixed assets as percent of sales}\\
  &= `r pretty(p9$sales)` \times `r p9$fixed.assets.pct.of.sales`\\
  &= `r pretty(p9$current["Fixed assets",])`\\\\
\text{Total assets} &= \text{current assets} + \text{fixed assets}\\
  &= `r pretty(p9$current["Current assets",])`
  + `r pretty(p9$current["Fixed assets",])`\\
  &= `r pretty(p9$current.total.assets)`\\\\
\mathbf{Liabilities}\\
\text{Short-term debt} &= \text{current sales}
  \times \text{short-term assets as percent of sales}\\
  &= `r pretty(p9$sales)` \times `r pretty(p9$short.term.debt.pct.of.sales)`\\
  &= `r pretty(p9$current["Short-term debt",])`\\\\
\text{Long-term debt} &= `r pretty(p9$long.term.debt)`\\\\
\text{Total debt} &= \text{short-term debt} + \text{long-term debt}\\
  &= `r pretty(p9$current["Short-term debt",])` + `r pretty(p9$long.term.debt)`\\
  &= `r pretty(p9$current.total.liabilities)`\\\\
\mathbf{Equity}\\
\text{Equity} &= \text{Total assets} - \text{Total liabilities}\\
  &= `r pretty(p9$current.total.assets)`
  - `r pretty(p9$current.total.liabilities)`\\
  &= `r pretty(p9$current.equity)`
\end{aligned}
$$

### B. External financing is needed

#### Step 1: Projected sales = \$`r pretty(p9$projected.sales)`

$$
\begin{aligned}
\text{Projected net income} &= \text{profit margin}
  \times \text{projected sales}\\
  &= `r p9$profit.margin` \times `r p9$projected.sales`\\
  &= `r p9$projected.net.income`\\\\
\text{Projected retained earnings} &= \text{retention ratio}
  \times \text{projected net income}\\
  &= `r (1 - p9$dividend.payout)` \times `r p9$projected.net.income`\\
  &= `r p9$projected.retained.earnings`
\end{aligned}
$$

#### Step 2: Project assets

$$
\begin{aligned}
\text{Projected current assets} &= \text{current assets percent of sales}
  \times \text{projected sales}\\
  &= `r p9$current.assets.pct.of.sales` \times `r p9$projected.sales`\\
  &= `r p9$projected["Current assets",]`\\\\
\text{Projected fixed assets} &= \text{fixed assets percent of sales}
  \times \text{projected sales}\\
  &= `r p9$fixed.assets.pct.of.sales` \times `r p9$projected.sales`\\
  &= `r p9$projected["Fixed assets",]`\\\\
\text{Projected total Assets} &= \text{projected current assets}
  + \text{projected fixed assets}\\
  &= `r p9$projected["Current assets",]`
  + `r p9$projected["Fixed assets",]`\\
  &= `r p9$projected.total.assets`
\end{aligned}
$$

#### Step 3: Project spontaneous liabilities

$$
\begin{aligned}
\text{Projected short-term debt} &= \text{short-term debt percent of sales}
  \times \text{projected sales}\\
  &= `r p9$short.term.debt.pct.of.sales` \times `r p9$projected.sales`\\
  &= `r p9$projected["Short-term debt",]`
\end{aligned}
$$

#### Step 4: Add existing long-term debt and equity

$$
\begin{aligned}
\text{Long-term debt} &= `r p9$long.term.debt`\\\\
\text{Projected equity} &= \text{equity from current balance sheet}
  + \text{projected retained earnings}\\
  &= `r p9$current.equity` + `r p9$projected.retained.earnings`\\
  &= `r p9$projected.equity`
\end{aligned}
$$

#### Step 5: Total available financing

$$
\begin{aligned}
\text{Projected liabilties} + \text{projected equity} &=
  \text{projected short-term debt}\\
  &\ \ \ \ + \text{long-term debt}\\
  &\ \ \ \ + \text{projected equity}\\
  &= `r p9$projected["Short-term debt",]`+ `r p9$long.term.debt`
  +  `r p9$projected.equity`\\
  &= `r p9$total.available.financing`
\end{aligned}
$$

#### Step 6: EFN = Total assets - Total financing

$$
\begin{aligned}
\text{EFN} &= \text{projected total assets} - \text{total financing}\\
  &= `r p9$projected.total.assets` - `r p9$total.available.financing`\\
  &= `r p9$efn`
\end{aligned}
$$

### C. Pro forma balance sheet

$$
\begin{aligned}
\mathbf{Assets}\\
\text{Current assets} &= \text{current sales}
  \times \text{current assets as percent of sales}\\
  &= `r pretty(p9$projected.sales)` \times `r p9$current.assets.pct.of.sales`\\
  &= `r pretty(p9$projected["Current assets",])`\\\\
\text{Fixed assets} &= \text{current sales}
  \times \text{fixed assets as percent of sales}\\
  &= `r pretty(p9$projected.sales)` \times `r p9$fixed.assets.pct.of.sales`\\
  &= `r pretty(p9$projected["Fixed assets",])`\\\\
\text{Total assets} &= \text{current assets} + \text{fixed assets}\\
  &= `r pretty(p9$projected["Current assets",])`
  + `r pretty(p9$projected["Fixed assets",])`\\
  &= `r pretty(p9$projected.total.assets)`\\\\
\mathbf{Liabilities}\\
\text{Short-term debt} &= \text{current sales}
  \times \text{short-term assets as percent of sales}\\
  &= `r pretty(p9$projected.sales)`
  \times `r pretty(p9$short.term.debt.pct.of.sales)`\\
  &= `r pretty(p9$projected["Short-term debt",])`\\\\
\text{Long-term debt} &= `r pretty(p9$long.term.debt)`\\\\
\text{Total debt} &= \text{short-term debt} + \text{long-term debt}\\
  &= `r pretty(p9$projected["Short-term debt",])`
  + `r pretty(p9$long.term.debt)`\\
  &= `r pretty(p9$projected.total.liabilities)`\\\\
\mathbf{Equity}\\
\text{Equity} &= \text{Current equity}
  - \text{Projected retained earnings}\\
  &= `r pretty(p9$current.equity)`
  + `r pretty(p9$projected.retained.earnings)`\\
  &= `r pretty(p9$projected.equity)`
\end{aligned}
$$

---

## 10. Sustainable growth rate

The Steiben company has an ROE of 13.1 percent and a payout ratio of 40 percent.

a. What is the company's sustainable growth rate?

b. Can the company's actual growth rate be different from its sustainable growth
   rate? Why or why not?

c. How can the company increase its sustainable growth rate?

---

```{r problem 10; sustainable growth rate, echo=F}
p10 <- new.env()
p10$roe <- .131
p10$payout.ratio <- .4
p10$sgr <- (p10$roe * (1 - p10$payout.ratio)) /
           (1 - p10$roe * (1 - p10$payout.ratio))
```

$$
\begin{aligned}
\mathbf{SGR} &= \frac{\text{ROE} \times (1 - \text{payout ratio})}
     {1 - \text{ROE} \times (1 - \text{payout ratio})}\\
  &= \frac{`r p10$roe` \times (1 - `r p10$payout.ratio`)}
     {1 - `r p10$roe` \times (1 - `r p10$payout.ratio`)}\\
  &= `r p10$sgr`
\end{aligned}
$$

b. Yes, a company growth rate can be different from its sustainable growth rate.
If it is below the SGR, then the company is operating at less than full
capacity. If the growth rate is above the sustainable growth rate, it would need
external funds to sustain the growth rate.

c. The sustainable growth rate is determined by the following factors based on
   the ROE and the payout ratio.

$$
\text{ROE} = \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier}
$$

The following four factors affect the sustainable growth rate:

#### Profit margin

An increase in profit margin will increase the firm's ability to generate funds
internally and thereby increase its sustainable growth.

#### Dividend policy

A decrease in the payout ratio will increase the retention ratio. This increases
internally generated equity and thus increases sustainable growth.

#### Financial policy

An increase in the debt-equity ratio increases the firm's financial leverage.
Because this makes additional debt financing available, it increases the
sustainable growth rate.

#### Total asset turnover

An increase in the firm's total asset turnover increases the sales generated for
each dollar in assets. This decreases the firm's need for new assets as sales
grow and thereby increases the sustainable growth rate. Notice that increasing
total asset turnover is the same thing as decreasing capital intensity.

---

## 11. Return on equity

Firm A and Firm B have debt-total asset ratios of 35 percent and 55 percent and
returns on total assets of 9 percent and 7 percent, respectively. Which firm has
a greater return on equity?

---

```{r problem 11; return on equity, echo=F}
p11 <- new.env()
p11$total.debt.ratio <- c(.35, .55)
p11$roa <- c(.09, .07)
p11$de.ratio <- (1 - p11$total.debt.ratio)
p11$equity.multiplier <- 1 + p11$de.ratio
p11$roe <- p11$roa * p11$equity.multiplier
```

The debt-total asset ratio is also called the total debt ratio.

$$
\begin{aligned}
\mathbf{Total\ debt\ ratio}\\
\text{Firm A} &= `r p11$total.debt.ratio[1]`\\
\text{Firm B} &= `r p11$total.debt.ratio[2]`\\\\
\mathbf{ROA}\\
\text{Firm A} &= `r p11$roa[1]`\\
\text{Firm B} &= `r p11$roa[2]`\\\\
\mathbf{DE\ ratio} &= (1 - \text{Total debt ratio})\\
\text{Firm A} &= `r p11$de.ratio[1]`\\
\text{Firm B} &= `r p11$de.ratio[2]`\\\\
\mathbf{Equity\ multiplier} &= 1 + \text{DE ratio}\\
\text{Firm A} &= `r p11$equity.multiplier[1]`\\
\text{Firm B} &= `r p11$equity.multiplier[2]`\\\\
\mathbf{ROE} &= \text{ROA} \times \text{Equity multiplier}\\
\text{Firm A} &= `r p11$roe[1]`\\
\text{Firm B} &= `r p11$roe[2]`
\end{aligned}
$$

### Conclusion

Although firm B has lower ROA, it has a greater return on equity because it uses more leverage (higher debt-total asset ratio).

---

## 12. Ratios and Foreign Companies

Prince Albert Canning PLC had a net loss of 37,543 pounds on sales of 345,182
pounds. What was the company's profit margin? Does the fact that these figures
are quoted in a foreign currency make any difference? Why? In dollars, sales
were \$559,725. What was the net loss in dollars?

---

```{r problem 12; ratios and foreign companies, echo=F}
p12 <- new.env()
p12$net.loss.pd <- 37543
p12$sales.pd <- 345182
p12$sales.us <- 559725
p12$profit.margin <- - p12$net.loss.pd / p12$sales.pd
p12$pd.us <- p12$sales.us / p12$sales.pd
p12$net.loss.us <- p12$net.loss.pd * p12$pd.us
```

$$
\begin{aligned}
\mathbf{Profit\ margin} &= \frac{\text{net income}}{\text{sales}}\\
  &= \frac{`r pretty(p12$net.loss.pd)`}{`r pretty(p12$sales.pd)`}\\
  &= `r round(p12$profit.margin, 4)`
\end{aligned}
$$

The profit margin would be the same in any currency because it is a ratio.

If sales were 559,725 in dollars. Then the exchange rate between dollars and
pounds is

$$
\begin{aligned}
\text{exchange rate} &= \frac{\text{sales USD}}{\text{sales PD}}\\
  &= \frac{`r pretty(p12$sales.us)`}{`r pretty(p12$sales.pd)`}\\
  &= `r round(p12$pd.us, 2)`
\end{aligned}
$$

The net loss in dollars is

$$
\begin{aligned}
\text{Net loss USD} &= \text{net loss PD} \times \text{exchange rate}\\
  &= `r pretty(p12$net.loss.pd)` \times `r p12$pd.us`\\
  &= `r pretty(p12$net.loss.us)`
\end{aligned}
$$

---

## 13. External Funds Needed

The Optical Scam Company has forecast a 15 percent sales growth rate for next
year. The current financial statements are shown here:

```{r problem 13; income statement, echo=F}
p13 <- new.env()
p13$is <- matrix(c(30400000, 26720000, 3680000, 1288000,
                   2392000, 956800, 1435200),
                 ncol=1)
row.names(p13$is) <- c("Sales", "Costs", "Taxable income", "Taxes",
                       "Net income", "Dividends",
                       "Addition to retained earnings")

knitr::kable(pretty(p13$is), caption="Income Statement") %>%
    kable_styling(latex_options = "HOLD_position")
```

```{r problem 13; balance sheet, echo=F}
p13$bs.assets <- c(7200000, 17600000)
p13$bs.liabilities.and.equity <- c(6400000, 4800000, 3200000, 10400000)
p13$bs.total.assets <- sum(p13$bs.assets)
p13$bs.total.liabilities <- sum(p13$bs.liabilities.and.equity[1:2])
p13$bs.total.equity <- sum(p13$bs.liabilities.and.equity[3:4])
p13$bs.total.liabilities.and.equity <- sum(p13$bs.liabilities.and.equity)
p13$bs <- matrix(c(p13$bs.assets, p13$bs.total.assets,
                   p13$bs.liabilities.and.equity, p13$bs.total.equity,
                   p13$bs.total.liabilities.and.equity), ncol=1)
rownames(p13$bs) <- c("Current assets", "Fixed assets", "Total assets",
                      "Short-term debt", "Long-term debt", "Common stock",
                      "Accumulated retained earnings", "Total equity",
                      "Total liabilities and equity")
knitr::kable(pretty(p13$bs), caption="Balance Sheet") %>%
    pack_rows("Assets", 1, 2) %>%
    pack_rows("Liabilities", 4, 5) %>%
    pack_rows("Stockholders' equity", 6, 8) %>%
    add_indent(8) %>%
    row_spec(c(2, 3, 7, 8),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")
```

a. Using the equation, calculate the external funds needed for next year.

b. Construct the firm's pro forma balance sheet for next year and confirm the
   external funds needed that you calculated in part (a).

c. Calculate the sustainable growth rate for the company.

d. Can Optical Scam eliminate the need for external funds by changing its
   dividend policy? What other options are available to the company to met its
   growth objectives?

---

### A. The equation for EFN

```{r solution 13a sales incease, echo=F}
p13$pct.sales.inc <- .15
p13$sales <- as.numeric(p13$is["Sales",])
p13$delta.sales <- p13$sales * p13$pct.sales.inc
p13$projected.sales <- p13$sales + p13$delta.sales
p13$short.term.debt <- as.numeric(p13$bs["Short-term debt",])
p13$capital.intensity.ratio <- as.numeric(p13$bs.total.assets) / p13$sales
p13$profit.margin <- as.numeric(p13$is["Net income",]) / p13$sales
p13$dividend.payout.ratio <- as.numeric(p13$is["Dividends",]) /
                              as.numeric(p13$is["Net income",])
p13$retention.ratio <- 1 - p13$dividend.payout.ratio
p13$efn  <- p13$capital.intensity.ratio * p13$delta.sales -
            p13$short.term.debt / p13$sales * p13$delta.sales -
            p13$profit.margin * p13$projected.sales * p13$retention.ratio
```

$$
\begin{aligned}
\Delta \text{Sales} &= \text{Sales}
  \times \text{precent increase in projected sales}\\
  &= `r pretty(p13$sales)` \times `r p13$pct.sales.inc`\\
  &= `r pretty(p13$delta.sales)`\\\\
\text{Capital intensity ratio} &= \frac{\text{total assets}}{\text{sales}}\\
  &= \frac{`r pretty(p13$bs.total.assets)`}{`r pretty(p13$sales)`}\\
  &= `r p13$capital.intensity.ratio`\\\\
\text{Profit margin} &= \frac{\text{net income}}{\text{sales}}\\
  &= \frac{`r pretty(p13$is["Net income",])`}{`r pretty(p13$sales)`}\\
  &= `r p13$profit.margin`\\\\
\text{Projected Sales} &= \text{Sales} + \Delta \text{Sales}\\
  &= `r pretty(p13$sales)` + `r pretty(p13$delta.sales)`\\
  &= `r pretty(p13$projected.sales)`\\\\
\text{Retention ratio} &= 1 - `r p13$dividend.payout.ratio`\\
  &= `r p13$retention.ratio`\\\\
\text{EFN} &= \text{capital intensity ratio} \times \Delta \text{sales}\\
  &\ \ \ \ - \frac{\text{spontaneous liabilities}}{\text{sales}}
  \times \Delta \text{sales}\\
  &\ \ \ \ - \text{profit margin} \times \text{projected sales}
  \times \text{retention ratio}\\
  &= `r p13$capital.intensity.ratio` \times `r p13$delta.sales`\\
  &\ \ \ \ - \frac{`r p13$short.term.debt`}{`r p13$sales`}
  \times `r p13$delta.sales`\\
  &\ \ \ \ - `r p13$profit.margin` \times `r p13$projected.sales`
  \times `r p13$retention.ratio`\\
  &= `r pretty(p13$efn)`
\end{aligned}
$$

### B. The Pro forma balance sheet

```{r solution 13b, echo=F}
p13$bs.assets.pro.forma <- p13$bs.assets * 1.15
p13$bs.total.assets.pro.forma <- sum(p13$bs.assets.pro.forma)
p13$bs.short.term.debt.pro.forma <- as.numeric(p13$bs["Short-term debt",]) * 1.15
p13$bs.total.liabilities.pro.forma <- p13$bs.short.term.debt.pro.forma +
                                       as.numeric(p13$bs["Long-term debt",])
p13$bs.accumulated.retained.earnings.pro.forma <-
    as.numeric(p13$bs["Accumulated retained earnings",]) +
    as.numeric(p13$is["Addition to retained earnings",]) * 1.15
p13$bs.total.equity.pro.forma <- as.numeric(p13$bs["Common stock",]) +
                              p13$bs.accumulated.retained.earnings.pro.forma
p13$bs.total.liabilities.and.equity.pro.forma <-
    p13$bs.total.liabilities.pro.forma +
    p13$bs.total.equity.pro.forma
p13$bs.pro.forma <- matrix(c(p13$bs.assets.pro.forma,
                             p13$bs.total.assets.pro.forma,
                             p13$bs.short.term.debt.pro.forma,
                             p13$bs["Long-term debt",],
                             p13$bs["Common stock",],
                             p13$bs.accumulated.retained.earnings.pro.forma,
                             p13$bs.total.equity.pro.forma,
                             p13$bs.total.liabilities.and.equity.pro.forma,
                             p13$efn),
                           ncol=1)
rownames(p13$bs.pro.forma) <- c("Current assets", "Fixed assets",
                                "Total assets", "Short-term debt",
                                "Long-term debt", "Common stock",
                                "Accumulated retained earnings", "Total equity",
                                "Total liabilities and equity",
                                "External financing needed")
knitr::kable(pretty(p13$bs.pro.forma),
             caption="Pro Forma Balance Sheet") %>%
    pack_rows("Assets", 1, 2) %>%
    pack_rows("Liabilities", 4, 5) %>%
    pack_rows("Stockholders' equity", 6, 8) %>%
    add_indent(8) %>%
    row_spec(c(2, 3, 7, 8),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")
```

### C. Sustainable growth

```{r problem 13c, echo=F}
p13$roe <- as.numeric(p13$is["Net income",]/p13$bs["Total equity",])
p13$sustainable.growth.rate <- (p13$roe * p13$retention.ratio) /
                                (1 - p13$roe * p13$retention.ratio)
```

$$
\begin{aligned}
\text{ROE} &= \frac{\text{net income}}{\text{total equity}}\\
  &= \frac{`r pretty(p13$is["Net income",])`}
          {`r pretty(p13$bs["Total assets",])`}\\
  &= `r round(p13$roe, 3)`\\\\
\text{Sustainable growth rate} &= \frac{\text{ROE} \times b}
                                       {1 - \text{ROE} \times b}\\
  &= \frac{`r p13$roe` \times `r p13$retention.ratio`}
          {1 - `r p13$roe` \times `r p13$retention.ratio`}\\
  &= `r p13$sustainable.growth.rate`
\end{aligned}
$$

### D. Options to external funds needed

* Reducing the payout ratio would increase retained earnings, reducing EFN.
* Increasing profit margin to generate more internal funds.
* Reduce asset growth (e.g., improve efficiency or delay capital expenditures).
* Increase spontaneous liabilities (e.g., negotiate better payment terms with suppliers).
* Issue new equity or take on additional long-term debt.

---

## 14. Days' sales in receivables

A company has net income of \$265,000, a profit margin of 9.3 percent, and an accounts receivable balance of \$145,300. Assuming 80 percent of sales are on credit, what is the company's days' sales in receivables?

---

```{r problem 14; days sales in receivables, echo=F}
p14 <- new.env()
p14$net.income <- 265000
p14$profit.margin <- .093
p14$accounts.receivable <- 145300
p14$pct.sales.on.credit <- .8
p14$sales <- p14$net.income / p14$profit.margin
p14$sales.on.credit <- p14$sales * p14$pct.sales.on.credit
p14$receivables.turnover <- round(p14$sales.on.credit /
                                  p14$accounts.receivable, 1)
p14$days.sales.in.receivables <- round(365 / p14$receivables.turnover, 1)
```

$$
\begin{aligned}
\text{Sales} &= \frac{\text{net income}}{\text{profit margin}}\\
  &= \frac{`r pretty(p14$net.income)`}{`r p14$profit.margin`}\\
  &= `r pretty(p14$sales)`\\\\
\text{Sales on credit} &= \text{sales}
  \times \text{percent of sales on credit}\\
  &= `r pretty(p14$sales)` \times `r p14$pct.sales.on.credit`\\
  &= `r pretty(p14$sales.on.credit)`\\\\
\text{Receivables turnover} &= \frac{\text{sales on credit}}
                                    {\text{accounts receivable}}\\
  &= \frac{`r pretty(p14$sales.on.credit)`}
          {`r pretty(p14$accounts.receivable)`}\\
  &= `r p14$receivables.turnover`\\
\text{Days' sales in receivables} &= \frac{365\ \text{days}}
                                          {\text{receivables turnover}}\\
  &= \frac{365}{`r round(p14$receivables.turnover, 1)`}\\
  &= `r p14$days.sales.in.receivables`\\\\
\end{aligned}
$$

## 15. Ratios and fixed assets

The Le Bleu Company has a ratio of long-term debt to total assets of .35 and a
current ratio of 1.25. Current liabilities are \$950, sales are \$5,780, profit
margin is 9.4 percent, and ROE is 18.2 percent. What is the amount of the firm's
net fixed assets?

---

```{r problem 15 ratios and fixed assets, echo=F}
p15 <- new.env()
p15$lt.debt.asset.ratio <- .35
p15$current.ratio <- 1.25
p15$current.liabilities <- 950
p15$sales <- 5780
p15$profit.margin <- .094
p15$roe <- .182
p15$current.assets <- p15$current.ratio * p15$current.liabilities
p15$net.income <- p15$profit.margin * p15$sales
p15$equity <- round(p15$net.income / p15$roe, 2)
p15$total.assets <- round((p15$current.liabilities + p15$equity) /
                          (1 - p15$lt.debt.asset.ratio), 2)
p15$fixed.assets <- p15$total.assets - p15$current.assets
p15$fixed.assets2 <- p15$current.liabilities / (1 - p15$lt.debt.asset.ratio) +
                     p15$equity / (1 - p15$lt.debt.asset.ratio) -
                     p15$current.assets
```

$$
\begin{aligned}
\text{current assets} &= \text{current ratio}
  \times \text{current liabilities}\\
  &= `r p15$current.ratio` \times `r p15$current.liabilities`\\
  &= `r pretty(p15$current.assets)`\\\\
\text{Net income} &= \text{profit margin} \times \text{sales}\\
  &= `r p15$profit.margin` \times `r p15$sales`\\
  &= `r p15$net.income`\\\\
\text{Equity} &= \frac{\text{net income}}{\text{ROE}}\\
  &= \frac{`r p15$net.income`}{`r p15$roe`}\\
  &= `r pretty(p15$equity)`\\\\
\end{aligned}
$$

$$
\begin{aligned}
\text{Assets} &= \text{liabilities} + \text{Equity}\\
  &= \frac{\text{long-term debt}}{\text{assets}} \times \text{assets} +
  \text{current liabilities} + \text{equity} \\ \\
\end{aligned}
$$

$$
\begin{aligned}
\text{assets}
  \times \left(1 - \frac{\text{long-term debt}}{\text{assets}} \right) &=
  \text{current liabilities} + \text{equity}\\
\end{aligned}
$$

$$
\begin{aligned}
\text{assets} &= \frac{\text{current liabilities}
  + \text{equity}}{(1 - \frac{\text{long-term debt}}{\text{assets}})}\\
\text{assets} &= \frac{`r p15$current.liabilities` + `r pretty(p15$equity)`}
  {`r 1 - p15$lt.debt.asset.ratio`}\\
  &= `r pretty(p15$total.assets)`\\\\
\end{aligned}
$$

$$
\begin{aligned}
\text{Fixed\ assets} &= \text{assets} - \text{current assets}\\
  &= `r pretty(p15$total.assets)` - `r pretty(p15$current.assets)`\\
  &= `r pretty(p15$fixed.assets)`
\end{aligned}
$$

---

## 16. Calculating the cash coverage ratio

Titan Inc.'s net income for the most recent year was \$8,320. The tax rate was
34 percent. The firm paid \$1,940 in total interest expense and deducted \$2,730
in depreciation expense. What was Titan's cash coverage ratio for the year?

---

```{r problem 16; calculating the cash coverage ratio, echo=F}
p16 <- new.env()
p16$net.income <- 8320
p16$tax.rate <- .34
p16$interest <- 1940
p16$depreciation <- 2730
p16$taxable.income <- round(p16$net.income / (1 - p16$tax.rate), 2)
p16$ebitda <- p16$taxable.income + p16$interest + p16$depreciation
p16$cash.coverage.ratio <- round(p16$ebitda / p16$interest, 2)
```

$$
\begin{aligned}
\text{Taxable income} &= \frac{\text{net income}}{(1 - \text{tax rate})}\\
  &= \frac{`r pretty(p16$net.income)`}{(1 - `r p16$tax.rate`)}\\
  &= `r pretty(p16$taxable.income)`\\\\
\text{EBITDA} &= \text{taxable income} + \text{interest} + \text{depreciation}\\
  &= `r pretty(p16$taxable.income)` + `r pretty(p16$interest)`
  +  `r pretty(p16$depreciation)`\\
  &= `r pretty(p16$ebitda)`\\\\
\text{Cash coverage ratio} &= \frac{\text{EBITDA}}{\text{interest}}\\
  &= \frac{`r pretty(p16$ebitda)`}{`r pretty(p16$interest)`}\\
  &= `r p16$cash.coverage.ratio`
\end{aligned}
$$

---

## 17. DuPont identity

The DuPont identity presented in the chapter is commonly referred to as the
three-factor DuPont identity. Another common way that the DuPont identity is
expressed is the five-factor model, which is:

$$
\text{ROE} = \frac{\text{net income}}{\text{EBT}} \times \frac{\text{EBT}}{\text{EBIT}}
  \times \frac{\text{EBIT}}{\text{sales}} \times \frac{\text{sales}}{\text{total assets}}
  \times \frac{\text{total assets}}{\text{equity}}
$$

Derive the five-factor DuPont identity (EBT is earnings before tax, but after interest). What does each term measure?

---

Deriving the five-factor DuPont identity from the three-factor DuPont identity:

$$
\begin{aligned}
\text{ROE} &= \frac{\text{net income}}{\text{sales}}
  \cdot \frac{\text{sales}}{\text{total assets}}
  \cdot \frac{\text{total assets}}{\text{equity}}\\\\
\text{ROE} &= \frac{\text{net income}}{\text{EBIT}}
  \cdot \frac{\text{EBIT}}{\text{sales}}
  \cdot \frac{\text{sales}}{\text{total assets}}
  \cdot \frac{\text{total assets}}{\text{equity}}\\\\
\text{ROE} &= \frac{\text{net income}}{\text{EBT}}
  \cdot \frac{\text{EBT}}{\text{EBIT}}
  \cdot \frac{\text{EBIT}}{\text{sales}}
  \cdot \frac{\text{sales}}{\text{total assets}}
  \cdot \frac{\text{total assets}}{\text{equity}}\\\\
\end{aligned}
$$

The five-factor DuPont identity breaks out the profit margin into three different factors.

$$
\begin{aligned}
\mathbf{Tax\ burden} &= \frac{\text{net income}}{\text{EBT}}\\\\
\mathbf{Interest\ burden} &= \frac{\text{EBT}}{\text{EBIT}}\\\\
\mathbf{Operating\ profit\ margin} &= \frac{\text{EBIT}}{\text{sales}}\\\\
\end{aligned}
$$

The last two factors are the same as in the three factor DuPont identity: Total
asset turnover and equity multiplier.

$$
\begin{aligned}
\mathbf{Total\ asset\ turnover} &= \frac{\text{sales}}{\text{total assets}}\\\\
\mathbf{Equity\ multiplier} &= \frac{\text{total assets}}{\text{equity}}
\end{aligned}
$$

---

## 18. Common-size and common-base year financial statements

In addition to common-size financial statements, common-base year financial
statements are often used. Common-based year financial statements are
constructed by dividing the current year account value by the base year account
value. Thus, the result shows the growth rate in the account. Using the
following financial statements, construct the common-size balance sheet and
common-base year balance sheet for the company. Use 2011 as the base year.

```{r problem 18 balance sheet, echo=F}
p18 <- new.env()
p18$bs.assets <- matrix(c(8014, 20453, 36822, 65289, 215370, 280659,
                          9954, 22937, 41797, 74688, 243340, 318028), ncol=2)

p18$account.names <- c("Cash", "Accounts receivable", "Inventory",
                       "Total current assets", "Net plant and equipment",
                       "Total assets", "Accounts payable", "Notes payable",
                       "Total current liabilities", "Long-term debt",
                       "Common stock and paid-in surplus", "Retained earnings",
                       "Total owners' equity",
                       "Total liabilities and owners' equity")
p18$years <- c("2011", "2012")

p18$bs.liabilities.equity <- matrix(c(40898, 17464, 58362, 24000, 38000, 160297,
                                      198297, 280659, 45884, 17035, 62919, 31000,
                                      39200, 184909, 224109, 318028), ncol=2)
p18$bs <- rbind(p18$bs.assets, p18$bs.liabilities.equity)
colnames(p18$bs) <- p18$years
rownames(p18$bs) <- p18$account.names
knitr::kable(p18$bs,
             caption="Jarrow Corporation 2011 and 2012 Balance Sheets") %>%
    pack_rows("Assets", 1, 6) %>%
    pack_rows("Current assets", 1, 4) %>%
    pack_rows("Fixed assets", 5, 5) %>%
    pack_rows("Liabilities and Owners' Equity", 7, 14) %>%
    pack_rows("Current liabilities", 7, 9) %>%
    pack_rows("Owners' equity", 11, 13) %>%
    row_spec(c(3, 4, 5, 6, 8, 9, 11, 12, 13, 14),
             extra_latex_after = "\\addlinespace") %>%
    add_indent(c(4, 9, 13)) %>%
    kable_styling(latex_options = "HOLD_position")
```

---

To construct the common-base year balance sheet you simply divide each of
the comparison years by the base year:

```{r problem 18; common-base balance sheet, echo=F}
p18$tmp <- round(p18$bs / p18$bs[,1], 2)
p18$bs.common.base <- matrix(p18$tmp[,2], ncol=1)
rownames(p18$bs.common.base) <- p18$account.names
colnames(p18$bs.common.base) <- "2011 -> 2012 Change"
p18$caption="Jarrow Corporation Common-base balance sheet base year 2011"
knitr::kable(p18$bs.common.base, caption=p18$caption) %>%
    pack_rows("Assets", 1, 6) %>%
    pack_rows("Current assets", 1, 4) %>%
    pack_rows("Fixed assets", 5, 5) %>%
    pack_rows("Liabilities and Owners' Equity", 7, 14) %>%
    pack_rows("Current liabilities", 7, 9) %>%
    pack_rows("Owners' equity", 11, 13) %>%
    row_spec(c(3, 4, 5, 6, 8, 9, 11, 12, 13, 14),
             extra_latex_after = "\\addlinespace") %>%
    add_indent(c(4, 9, 13)) %>%
    kable_styling(latex_options = "HOLD_position")
```

To construct the common-size year balance sheet we will use percentage of total
assets:

```{r problem 18; common-size balance sheet, echo=F}
p18$bs.common.size <- p18$bs / as.numeric(p18$bs["Total assets",])
p18$pct.change <- as.numeric(p18$bs.common.size[,2] -
                             p18$bs.common.size[,1])
p18$bs.common.size <- cbind(p18$bs.common.size, p18$pct.change)
rownames(p18$bs.common.size) <- p18$account.names
colnames(p18$bs.common.size) <- c(p18$years, "Change %")
p18$caption="Jarrow Corporation Common-size balance sheet base year 2011"
knitr::kable(p18$bs.common.size, caption=p18$caption) %>%
    pack_rows("Assets", 1, 6) %>%
    pack_rows("Current assets", 1, 4) %>%
    pack_rows("Fixed assets", 5, 5) %>%
    pack_rows("Liabilities and Owners' Equity", 7, 14) %>%
    pack_rows("Current liabilities", 7, 9) %>%
    pack_rows("Owners' equity", 11, 13) %>%
    row_spec(c(3, 4, 5, 6, 8, 9, 11, 12, 13, 14),
             extra_latex_after = "\\addlinespace") %>%
    add_indent(c(4, 9, 13)) %>%
    kable_styling(latex_options = "HOLD_position")
```

---

Use the following information for Problems 19, 20, and 22:

The discussion of EFN in the chapter implicitly assumed that the company was
operating at full capacity. Often, this is not the case. For example, assume
that Rosengarten was operating at 90 percent capacity. Full-capacity sales would
be \$1,000/.90 = \$1,111. The balance sheet shows \$1,800 in fixed assets. The
capital intensity ratio for the company is

$$
\begin{aligned}
\mathbf{Capital\ intensity\ ratio} &= \frac{\text{fixed assets}}
  {\text{full-capacity sales}}\\
  &= \frac{1,800}{1,111}\\
  &= 1.62
\end{aligned}
$$

This means that Rosengarten needs \$1.62 in fixed assets for every dollar in
sales when it reaches full capacity. At the projected sales level of \$1,250, it
needs $\$1,250 \times 1.62 = \$2,025$ in fixed assets, which is \$225 lower than
our projection of \$2,250 in fixed assets. So, EFN is only \$565 - 225 = \$340.

## 19. Full-capacity sales

Thorpe Mfg., Inc., is currently operating at only 90 percent of fixed asset
capacity. Current sales are \$725,000. How much can sales increase before any
new fixed assets are needed?

---

```{r problem 19; full-capacity sales, echo=F}
p19 <- new.env()
p19$current.sales <- 725000
p19$operating.capacity <- .9
p19$sales.full.capacity <- p19$current.sales / p19$operating.capacity
p19$sales.increase <- p19$sales.full.capacity - p19$current.sales
```

$$
\begin{aligned}
\text{Full-capacity sales} &= \frac{\text{Current sales}}{\text{pct sales}}\\
  &= \frac{`r pretty(p19$current.sales)`}{`r p19$operating.capacity`}\\
  &= `r pretty(p19$sales.increase)`
\end{aligned}
$$

Thorpe's full capacity sales is \$`r pretty(p19$sales.full.capacity)`. Therefore
with current sales as \$`r pretty(p19$current.sales)`, Thrope can increase sales
by \$`r pretty(p19$sales.increase)` before it would need new fixed assets.

---

## 20. Fixed assets and capacity usage

For the company in the previous problem, suppose fixed assets are \$690,000 and
sales are projected to grow to \$830,000. How much in new fixed assets are
required to support this growth in sales?

---

```{r problem 20; fixed assets and capacity usage, echo=F}
p20 <- new.env()
p20$fixed.assets <- 690000
p20$projected.sales <- 830000
p20$capital.intensity.ratio <-
    round(p20$fixed.assets / p19$sales.full.capacity, 3)
p20$projected.fixed.assets <- p20$projected.sales * p20$capital.intensity.ratio
p20$new.fixed.assets <- p20$projected.fixed.assets - p20$fixed.assets
```

The Thorpe's capital intensity ratio is:

$$
\begin{aligned}
\text{CIR} &= \frac{\text{fixed assets}}{\text{full-capacity sales}}\\
  &= \frac{`r pretty(p20$fixed.assets)`}{`r pretty(p19$sales.full.capacity)`}\\
  &= `r p20$capital.intensity.ratio`
\end{aligned}
$$

With a  projected sales growth of \$830,000, Thrope will need:

$$
\begin{aligned}
\text{New fixed assets} &= ( \text{projected sales} \times \text{CIR} )
  - \text{fixed assets}\\
  &= `r pretty(p20$projected.sales)`
  \times `r pretty(p20$capital.intensity.ratio)`
  - `r pretty(p20$fixed.assets)`\\
  &= `r pretty(p20$new.fixed.assets)`
\end{aligned}
$$

---

## 21. Calculating EFN

The most recent financial statements for Moose Tours, Inc., appear below. Sales
for 2012 are projected to grow by 20 percent. Interest expense will remain
constant; the tax rate and the dividend payout rate will also remain constant.
Costs, other expenses, current assets, fixed assets, and accounts payable
increase spontaneously with sales. If the firm is operating at full capacity and
no new debt or equity is issued, what external financing is needed to support
the 20 percent growth rate in sales?

```{r problem 21; income statement, echo=F}
p21 <- new.env()

p21$growth.rate <- .2
p21$is <- matrix(c(
    p21$sales <- 836100,
    p21$costs <- 650700,
    p21$other.expenses <- 17100,
    p21$ebit <- 168300,
    p21$interest <- 12600,
    p21$taxable.income <- 155700,
    p21$taxes <- 54495,
    p21$net.income <- p21$taxable.income - p21$taxes,
    p21$dividends <- 30300,
    p21$atre <- 70905), ncol=1)
row.names(p21$is) <- c("Sales", "Costs", "Other expenses",
    "Earnings before interest and taxes", "Interest expense", "Taxable income",
    "Taxes", "Net income", "Dividends", "Addition to retained earnings")

knitr::kable(pretty(p21$is), caption="Income Statement") %>%
    add_indent(c(9, 10)) %>%
    kable_styling(latex_options = "HOLD_position")
```

```{r problem 21; balance sheet, echo=F}
p21$bs.assets <- matrix(c(
    p21$cash <- 24035,
    p21$accounts.receivable <- 38665,
    p21$inventory <- 82555,
    p21$total.current.assets<- p21$cash + p21$accounts.receivable +
        p21$inventory,
    p21$fixed.assets <- 392350,
    p21$total.assets <- p21$total.current.assets + p21$fixed.assets), ncol=1)
row.names(p21$bs.assets) <- c("Cash", "Accounts receivable", "Inventory",
                              "Total current assets", "Fixed assets",
                              "Total assets")

p21$bs.liabilities <- matrix(c(
    p21$accounts.payable <- 64600,
    p21$notes.payable <- 16150,
    p21$total.current.liabilities <- p21$accounts.payable + p21$notes.payable,
    p21$long.term.debt <- 150000,
    p21$total.liabilities <- p21$total.current.liabilities +
                             p21$long.term.debt), ncol=1)
row.names(p21$bs.liabilities) <- c("Accounts payable", "Notes payable",
    "Total current liabilities", "Long-term debt", "Total liabilities")

p21$bs.equity <- matrix(c(
    p21$common.stock <- 130000,
    p21$retained.earnings <- 176855,
    p21$total.equity <- p21$common.stock + p21$retained.earnings), ncol=1)
row.names(p21$bs.equity) <- c("Common stock and paid-in surplus",
                              "Retained earnings", "Total equity")

p21$bs.total.liabilities.and.equity <- matrix(c(
    p21$total.liabilities + p21$total.equity), ncol=1)
row.names(p21$bs.total.liabilities.and.equity) <-
    c("Total liabilities and owners' equity")

p21$bs <- rbind(p21$bs.assets, p21$bs.liabilities,
                p21$bs.equity, p21$bs.total.liabilities.and.equity)

knitr::kable(pretty(p21$bs),
             caption="Balance Sheet") %>%
    pack_rows("Assets", 1, 5) %>%
    pack_rows("Liabilities", 7, 11) %>%
    pack_rows("Owners' equity", 12, 14) %>%
    add_indent(c(4, 9, 11, 14)) %>%
    row_spec(c(4, 11, 14),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")
```

### Building pro forma income statement

With a projected sales growth of 20\%, the pro forma income statement is:

```{r problem 21; pro forma income statement, echo=F}
p21$dividend.payout.ratio <- as.numeric(p21$is["Dividends",] /
                                        p21$is["Net income",])
p21$tax.rate <- as.numeric(p21$is["Taxes",] / p21$is["Taxable income",])

p21$is.proforma <- p21$is
p21$is.proforma[c("Sales", "Costs", "Other expenses",
                  "Earnings before interest and taxes"),] <- p21$is[c(
                  "Sales", "Costs", "Other expenses",
                  "Earnings before interest and taxes"),] *
                  (1 + p21$growth.rate)

p21$taxable.income <- p21$is.proforma["Earnings before interest and taxes",] -
    p21$is.proforma["Interest expense",]

p21$taxes <- p21$taxable.income * p21$tax.rate
p21$net.income <- p21$taxable.income - p21$taxes
p21$dividends <- p21$net.income * p21$dividend.payout.ratio
p21$atre <- p21$net.income * (1 - p21$dividend.payout.ratio)
p21$is.proforma["Taxable income",] <- p21$taxable.income
p21$is.proforma["Taxes",] <- p21$taxes
p21$is.proforma["Net income",] <- p21$net.income
p21$is.proforma["Dividends",] <- p21$dividends
p21$is.proforma["Addition to retained earnings",] <- p21$atre
knitr::kable(pretty(p21$is.proforma), caption="Pro Forma Income Statement") %>%
    add_indent(c(9, 10)) %>%
    kable_styling(latex_options = "HOLD_position")
```

### Building pro forma balance sheet

```{r problem 21; pro forma balance sheet, echo=F}
p21$bs.proforma <- p21$bs
p21$bs.proforma[1:7,] <- p21$bs.proforma[1:7,] * (1 + p21$growth.rate)

p21$bs.proforma["Total current liabilities",] <-
    p21$bs.proforma["Accounts payable",] + p21$bs.proforma["Notes payable",]

p21$bs.proforma["Total liabilities",] <- p21$bs.proforma["Long-term debt",] +
    p21$bs.proforma["Total current liabilities",]

p21$bs.proforma["Retained earnings",] <- p21$bs["Retained earnings",] +
    p21$is.proforma["Addition to retained earnings",]

p21$bs.proforma["Total equity",] <-
    p21$bs.proforma["Common stock and paid-in surplus",] +
    p21$bs.proforma["Retained earnings",]

p21$bs.proforma["Total liabilities and owners' equity",] <-
    p21$bs.proforma["Total liabilities",] + p21$bs.proforma["Total equity",]

knitr::kable(pretty(p21$bs.proforma),
             caption="Pro Forma Balance Sheet") %>%
    pack_rows("Assets", 1, 5) %>%
    pack_rows("Liabilities", 7, 11) %>%
    pack_rows("Owners' equity", 12, 14) %>%
    add_indent(c(4, 9, 11, 14)) %>%
    row_spec(c(4, 11, 14),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")

p21$test <- as.numeric(p21$bs.proforma["Total assets",] -
    p21$bs.proforma["Total liabilities and owners' equity",])
```

```{r problem 21; efn, echo=F}
p21$delta.sales <- as.numeric(p21$is["Sales",] * p21$growth.rate) 
p21$cir <- p21$bs["Total assets",] / p21$is["Sales",]
p21$sls <- p21$bs["Accounts payable",] / p21$is["Sales",]
p21$profit.margin <- p21$is["Net income",] / p21$is["Sales",]
p21$retention.ratio <- 1 - p21$dividend.payout.ratio
p21$efn <- as.numeric(p21$cir * p21$delta.sales - p21$sls * p21$delta.sales -
    p21$profit.margin * p21$is.proforma["Sales",] * p21$retention.ratio)
```

Finding the external financing needed to support 20\% growth:

$$
\begin{aligned}
\Delta \text{sales} &= \text{current sales} \times \text{growth rate}\\
  &= `r pretty(p21$is["Sales",])` \times `r p21$growth.rate`\\
  &= `r pretty(p21$delta.sales)`\\\\
\text{Projected sales} &= \text{Current sales} + \text{Change in sales}\\
  &= `r pretty(p21$is["Sales",])` + `r pretty(p21$delta.sales)`\\
  &= `r pretty(p21$is.proforma["Sales",])`\\\\
\text{Capital intensity ratio} &= \frac{\text{total assets}}{\text{sales}}\\
  &= \frac{`r pretty(p21$bs["Total assets",])`}
          {`r pretty(p21$is["Sales",])`}\\
  &= `r p21$cir`\\\\
\text{Spontaneous liabilities to sales} &= \frac{\text{accounts payable}}
  {\text{sales}}\\
  &= \frac{`r pretty(p21$bs["Accounts payable",])`}
  {`r pretty(p21$is["Sales",])`}\\
  &= `r pretty(p21$sls)`\\\\
\text{Dividend payout ratio} &= \frac{\text{dividends}}{\text{net income}}\\
  &= \frac{`r pretty(p21$is["Dividends",])`}{`r pretty(p21$is["Net income",])`}\\
  &= `r pretty(p21$dividend.payout.ratio)`\\\\
\text{Retention ratio} &= 1 - \text{dividend payout ratio}\\
  &= 1 - `r p21$dividend.payout.ratio`\\
  &= `r p21$retention.ratio`\\\\
\text{Profit margin} &= \frac{\text{net income}}{\text{sales}}\\
  &= \frac{`r pretty(p21$is["Net income",])`}{`r pretty(p21$is["Sales",])`}\\
  &= `r p21$profit.margin`\\\\
\text{EFN} &= \text{capital intensity ratio} \times \text{change in sales}\\
  &\ \ \ - \text{spontaneous liabilities to sales} \times \text{change in sales}\\
  &\ \ \ - \text{profit margin} \times \text{projected sales}
  \times \text{retention ratio}\\
  &= `r p21$cir` \times `r pretty(p21$delta.sales)`\\
  &\ \ \ - `r p21$sls` \times `r pretty(p21$delta.sales)`\\
  &\ \ \ - `r p21$profit.margin` \times `r pretty(p21$is.proforma["Sales",])`
  \times `r p21$retention.ratio`\\
  &= `r pretty(p21$efn)`
\end{aligned}
$$

---

## 22. Capacity usage and growth

In the previous problem, suppose the firm was operating at only 80 percent capacity in 2011. What is EFN now?

---

If Moose Tours, Inc. was operating at only 80\% of capacity in 2011, then the
firm has excess capacity in its fixed assets and does not need to invest in new
fixed assets until it reaches full capacity.

```{r problem 22; capacity usage and growth, echo=F}
p22 <- new.env()
p22$capacity.usage <- .8
p22$sales.full.capacity <- as.numeric(p21$is["Sales",]) / .8
p22$cir <- as.numeric(p21$bs.assets["Fixed assets",] /
                       p22$sales.full.capacity)
p22$projected.fixed.assets <- as.numeric(p21$is.proforma["Sales",] * p22$cir)
p22$previous.fixed.assets <- as.numeric(p21$bs.assets["Fixed assets",] *
                              (1 + p21$growth.rate))
p22$diff <- p22$previous.fixed.assets - p22$projected.fixed.assets
p22$efn <- p21$efn - p22$diff
```

Key information from the previous problem:

* Sales: \$`r pretty(p21$is["Sales",])`
* Fixed Assets: \$`r pretty(p21$bs["Fixed assets",])`
* Capacity usage: 80\%
* Projected Sales: \$`r pretty(p21$is.proforma["Sales",])`
  (`r p21$growth.rate*100`\% growth)
* Addition to retained earnings:
  \$`r pretty(p21$is["Addition to retained earnings",])`
* Increase in A/P (spontaneous liabilities):
  \$`r pretty(p21$bs["Accounts payable",])`
* Cash, A/R, Inventory scale with sales.

Full capacity sales:

$$
\begin{aligned}
\text{Full capacity sales} &= \frac{`r pretty(p21$is["Sales",])`}
  {`r p22$capacity.usage`}
  &= `r pretty(p22$sales.full.capacity)`
\end{aligned}
$$

From the balance sheet, our fixed assets are
\$`r pretty(p21$bs["Fixed assets",])`. The capital intensity ratio:

$$
\begin{aligned}
\text{Capital intensity ratio} &= \text{fixed assets}
  / \text{full-capacity sales}\\
  &= `r pretty(p21$bs["Fixed assets",])` / `r pretty(p22$sales.full.capacity)`\\
  &= `r pretty(p22$cir)`
\end{aligned}
$$

This means that, for every dollar in sales when full capacity is reached,
`r p22$cir` of fixed assets will be needed. At projected sales level of
\$`r pretty(p21$is.proforma["Sales",])`, Moose Tours will needs
\$`r pretty(p22$projected.fixed.assets)` in fixed assets,
which is \$`r pretty(p22$diff)` lower then the previous projection. Therefore we
need \$`r pretty(p22$efn)` in external funding.

---

## 23. Calculating EFN

In problem 21, suppose the firm wishes to  keep its debt-equity ratio constant. What is the EFN now?

---

```{r problem 23; pro forma income statement, echo=F}
p23 <- new.env()
p23$is.proforma <- p21$is
p23$is.proforma[1:4,] <- p21$is.proforma[1:4,] * (1 + p21$growth.rate)
p23$is.proforma["Taxable income",] <-
    p23$is.proforma["Earnings before interest and taxes",] -
    p23$is.proforma["Interest expense",]
p23$is.proforma["Taxes",] <- p23$is.proforma["Taxable income",] *
    p21$tax.rate
p23$is.proforma["Net income",] <- p23$is.proforma["Taxable income",] -
    p23$is.proforma["Taxes",]
p23$is.proforma["Dividends",] <- p23$is.proforma["Net income",] *
    p21$dividend.payout.ratio
p23$is.proforma["Addition to retained earnings",] <-
    p23$is.proforma["Net income",] * (1 - p21$dividend.payout.ratio)
knitr::kable(pretty(p23$is.proforma), caption="Pro Forma Income Statement") %>%
    add_indent(c(9, 10)) %>%
    kable_styling(latex_options = "HOLD_position")
```

```{r problem 23; pro forma balance sheet, echo=F}
p23$bs.proforma <- p21$bs
p23$bs.proforma[1:6,] <- p23$bs.proforma[1:6,] * (1 + p21$growth.rate)
p23$bs.proforma["Accounts payable",] * (1 + p21$growth.rate)
p23$bs.proforma["Notes payable",] <-
    p23$bs.proforma["Notes payable",] +
    19669
p23$bs.proforma["Total current liabilities",] <-
    p23$bs.proforma["Accounts payable",] +
    p23$bs.proforma["Notes payable",]
p23$bs.proforma["Long-term debt",] <-
    p23$bs.proforma["Long-term debt",] +
    59008
p23$bs.proforma["Total liabilities",] <-
    p23$bs.proforma["Total current liabilities",] +
    p23$bs.proforma["Long-term debt",]
p23$bs.proforma["Retained earnings",] <-
    p23$bs.proforma["Retained earnings",] +
    p23$is.proforma["Addition to retained earnings",]
p23$bs.proforma["Total equity",] <-
    p23$bs.proforma["Total equity",] +
    p23$is.proforma["Addition to retained earnings",]
p23$bs.proforma["Total liabilities and owners' equity",] <-
    p23$bs.proforma["Total liabilities",] +
    p23$bs.proforma["Total equity",]
pretty(p23$bs.proforma)
p23$bs.proforma["Total liabilities",] / p23$bs.proforma["Total equity",]
p21$bs["Total liabilities",] / p21$bs["Total equity",]
pretty(p23$bs.proforma["Total assets",] -
       p23$bs.proforma["Total liabilities and owners' equity",])
```

---

```{r}
ls()
```
