---
title: "Stock valuation"
author: "Joseph Bennett"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

1. A stock can be valued by discounting its dividends. We mention three types of situations:

a. The case of zero growth of dividends.
b. The case of constant growth of dividends.
c. The case of differential growth.

2. An estimate of the growth rate of dividends is needed for the dividend discount model. A useful estimate of the growth rate is:

$$
g = \text{retention ratio} \times \text{return on retained earnings (ROE)}
$$

As long as the firm holds its ratio of dividends to earning constant, $g$ represents the growth rate of both dividends and earnings.

3. The price of a share of stock can be viewed as the sum of its price under the assumption that the firm is a "cash cow" plus the per-share value of the firm's growth opportunities. A company is termed a cash cow if it pays out all of its earnings as dividends.

We write the value of a share as:

$$
\frac{\text{EPS}}{R} + \text{NPVGO}
$$

4. Negative NPV projects lower the value of the firm. That is, projects with rates of return below the discount rate lower firm value. Nevertheless, both the earnings and dividends of a firm will grow as long as the firm's projects have positive rates of return.

5. From accounting, we know that earnings are divided into two parts: dividends and retained earnings. Most firms continually retain earnings in order to create future dividends. One should not discount earnings to obtain price per share since part of earnings must be reinvested. Only dividends reach the stockholders and only they should be discounted to obtain share price.

6. Some analysts value stock via multiples, such as the price-earnings ratio. However, we caution that one must apply the same multiple only to similar companies.

7. We suggest that a firm's price-earnings ratio is a function of three factors:

a. The per-share amount of the firm's valuable growth opportunities.
b. The risk of the stock.
c. The type of accounting method used by the firm.

8. The two biggest stock markets in the United States are the NYSE and the NASDAQ. We discussed the organization and operation of these two markets, and we saw how stock price information is reported.

# Concepts

## Stock valuation

Why does the value of a stock depend on dividends?

The value of a stock fundamentally depends on dividends because dividends represent the cash flows that shareholders receive as a return on their investment. The value of any financial asset is generally derived from the present value of its expected future cash flows. In the case of stocks, these cash flows primarily come from dividends.

1. Intrinsic value and dividends

A common method for valuing stocks is the Dividend Discount Model, which calculates the present value of all expected future dividends. The formula is:

$$
\begin{aligned}
P_0 &= \sum^{\infty}_{t=1}{\frac{D_t}{(1+r)^{t}}}\\
\text{Where:}\\ \\
P_0\ \text{is the current stock price.}\\
D_t\ \text{is the dividend at time } t\text{.}\\
r\ \text{is the discount rate (investor's required rate of return).}\\
\end{aligned}
$$

This model explicitly ties the stock price to dividends, showing that the stock's value is based on the expected stream of future dividends.

2. Dividends as a measure of profitability

Dividends are often seen as a tangible measure of a company's profitability and financial health. Companies that pay regular dividends signal to investors that they are generating consistent earnings and have confidence in their future performance.

3. No Dividends

Even for companies that do not pay dividends currently (e.g., growth companies), their value is still tied to the expectation that they will eventually generate dividends or distribute cash to shareholders (e.g., through share buybacks). Investors often expect the company to reinvest profits now to generate higher returns in the future, leading to higher dividends or payouts down the line.

4. Alternative returns

Stock value is also influenced by capital gains (price appreciation). However, capital gains are indirectly tied to dividends because the market value of a stock reflects investors' expectations about future cash flows, which ultimately include dividends or other distributions.

5. Reinvestment of earnings

For companies that retain earnings instead of paying dividends, the retained earnings ar expected to be reinvested to grow the business. This growth should in theory, increase the company's future ability to pay dividends, enhancing the stock's value.

Dividends are a direct return to shareholders and a key componenet of the cash flows used to determine the intrinsic value of a stock. Even if a company doesn't currently pay dividends, its stock value reflects the expectation of future payouts or reinvested earnings that lead to higher future value.

---

## Stock valuation

A substantial percentage of the companies listed on the NYSE and the NSADAQ don't pay dividends, but investors are nonetheless willing to buy shares in them. How is this possible given your answer to the previous question?

As stated in section 4 of the previous answer, the companies that do not pay dividends currently do have an expectation that eventually they will generate dividends. These companies are usually high growth companies that are constantly reinvesting their profits now to generate higher returns in the future.

---

## Dividend policy

Referring to the previous questions, under what circumstances might a company choose not to pay dividends?

A company might choose not to pay dividends under several circumstances, usually tied to its financial strategy, growth objectives, or market conditions.

1. Growth objectives

Companies in their growth phase often choose to reinvest earnings into expanding operations, developing new products, or entering new markets rather than paying dividends. Retained earnings can fund projects with higher potential returns than the returns shareholders might achieve by reinvesting dividends themselves. Also, many startups and technology companies prioritize reinvestment to fuel rapid growth, innovation, and market capture.

2. Financial constraints

If a company has low or negative earnings, it may not have enough cash flow to pay dividends. Companies with significant debt may prioritize paying down liabilities or meeting interest payments before distributing dividends. Even profitable companies might face liquidity constraints that prevent them from distributing dividends.

3. Shareholder preferences

Some shareholders prefer capital gains over dividends because gains are often taxed at a lower rate or deferred until shares are sold. By retaining earnings, the company can increase its value, leading to higher stock prices. Companies with a long-term investor base may focus on reinvesting earnings to drive sustainable growth rather than paying out dividends.

4. Strategic considerations

A company may retain earingns to fund acquisitions or large-scale expansions rather than distributing them as dividends. During periods of economic uncertainty or downturns, companies may conserve cash to maintain flexibility and stability.

5. Market norms and industry practices

In some industries (e.g., technology or biotech), it's common for companies to avoid paying dividends, as investors expect returns through capital appreciation instead. Companies in mature industries are more likely to pay dividends, while those in emerging sectors often retain earnings.

6. Tax efficiency

Dividends are typically taxed as income for shareholders, while capital gains are taxed differently. Companies may avoid paying dividends to help shareholders minimize tax burdens.

7. Signal to investors

By retaining earnings, a company might signal confidence in its growth prospects, suggesting that it expects to generate higher returns through reinvestment. Also, once a company establishes a dividend policy, cutting or suspending dividends can negatively impact investor confidence. Some companies avoid initiating dividends to maintain flexibility.

---

## Dividend growth model

Under what two assumptions can we use the dividend growth model presented in the chapter to determine the value of a share of stock? Comment on the reasonableness of these assumptions.

The dividend growth model can be expressed:

$$
P_0 = \frac{\text{Div}}{(R - g)}
$$

The two assumptions that the dividend discount model can be used under:

1. The return on reinvestment of future retained earnings is equal to the firm's past ROE.

2. The future retention ratio is equal to the past retention ratio.

These assumptions relate to how future growth ($g$) is determined and the reasonableness of using past metrics to project future growth.

For the first assumption, we can determine that it is reasonable when: the firm's operating environment and profitability are stable over time; the firm consistently invests in projects with similar risk-adjusted returns; the industry is mature, and competitive dynamics do not significantly erode profitability; or when the firm has demonstrated strong, consistent historical performance in reinvesting retained earnings effectively.

The first assumption is not reasonable when: the firm's profitability is declining due to increasing competition or market saturation; the firm is in a highly volatile or rapidly changing industry (e.g., technology), where past performance may not predict future returns; the firm is entering new markets or undertaking projects with significantly different risk-return profiles; or when economic downturns, regulatory changes, or disruptive innovations alter the firm's ability to generate consistent returns.

The second assumption is reasonable when: the firm's dividend policy is stable, and management does not plan significant changes to the payout ratio; the firm's growth strategy and capital structure remain consistent; the firm has steady earnings growth and no major capital expenditure requirements that would necessitate changes in retained earnings.

The second assumption is not reasonable when: the firm plans to shift its dividend policy (e.g., from reinvestment to returning more capital to shareholders in the form of dividends or buybacks); earnings volatility causes fluctuations in the retention ratio; the firm faces new growth opportunities or competitive pressures requiring higher or lower retention of earnings; a significant change in financial strategy occurs, such as debt reduction or acquisitions, which affects how earnings are allocated.

---

## Common versus preferred stock

Suppose a company has a preferred stock issue and a common stock issue. Both have just paid a \$2 dividend. Which do you think will have a higher price, a share of the preferred or a share of the common?

---

## Dividend growth model

Based on the dividend growth model, what are the two components of the total return on a share of stock? Which do you think is typically larger?

The dividend growth model is expressed as:

$$
\begin{aligned}
P_0 &= \frac{\text{Div}_1}{R - g} \\
\text{Where: }\\
P_0 &=\ \text{the present value of the common stock investment}\\
\text{Div}_1 &=\ \text{the dividend on the stock at the end of the first period}\\
R &= \text{discount rate}\\
g &= \text{dividend growth rate}
\end{aligned}
$$

Rearranging the equation so that we solve for the discount rate $R$:

$$
\begin{aligned}
R - g &= \frac{\text{Div}_1}{P_0} + g\\ \\
R &= \frac{\text{Div}_1}{P_0} + g
\end{aligned}
$$

The two components of this equation is $\frac{\text{Div}_1}{P_0}$, the __dividend yield__, and $g$ the dividend growth rate. This rate is also the growth rate of the stock price and is called the __capital gains yield__.

Which of the two components are larger depends on the type of company: high growth companies will have a larger capital gains yield because they will be reinvesting most of their earnings to drive growth. Mature company's in stable markets will have a higher dividend yield because they prioritize returning cash back to investors. Mature companies that return all earning as dividends are called _cash cows_.

---

## Growth rate

In the context of the dividend growth model, is it true that the growth rate in dividends and the growth rate in the price of the stock are identical?

Yes, according to the dividend growth model, the growth rate in dividends and the growth rate in the price of the stock are the same. They are called the __capital gains yield__.

---

## Price--earnings ratio

What are the three factors that determine a company's price--earnings ratio?

---

##  Corporate ethics

Is it unfair or unethical for corporations to create classes of stock with unequal voting rights?

---

## Stock valuation

Evaluate the following statement: Managers should not focus on the current stock value because doing so will lead to an overemphasis on short-term profits at the expense of long-term profits.

# Problems

## Stock values

The Starr Co. just paid a dividend of \$2.15 per share on its stock. The dividends are expected to grow at a constant rate of 5 percent per year, indefinitely. If investors require a return of 11 percent on the stock, what is the current price? What will the price be in three years? In 15 years?


$$
\begin{aligned}
\text{Div}_1 &= \text{Div}_0 \times (1 + .05)\\
  &= 2.2575\\ \\
P_0 &= \frac{\text{Div}}{R - g}\\ \\
P_0 &= \frac{2.2575}{.11 - .05} \\
 &= \frac{2.15}{.06} \\
 &= 37.625
\end{aligned}
$$

The price in three years from now is dependent on the dividend expected in four years:

$$
\begin{aligned}
\text{Div}_4 &= \text{Div}_0 \times (1 + g)^4\\
  &= 2.15 \times (1.05)^4\\
  &= 2.613\\ \\
\text{P}_3 &= \frac{\text{Div}_4}{R - g}\\
  &= \frac{2.613}{.06}\\
  &= 43.55
\end{aligned}
$$

The price in 15 years:

$$
\begin{aligned}
\text{Div}_{16} &= \text{Div}_0 \times (1 + g)^{16}\\
  &= 2.15 \times (1.05)^{16}\\
  &= 4.693\\ \\
P_{15} &= \frac{\text{Div}_{16}}{R - g}\\
  &= \frac{4.693}{.06}\\
  &= 78.217
\end{aligned}
$$

---

## Stock values

The next dividend payment by ECY, Inc., will be \$3.20 per share. The dividends are anticipated to maintain a growth rate of 6 percent, forever. If ECY stock currently sells for \$63.50 per share, what is the required return?

$$
\begin{aligned}
R &= \frac{\text{Div}}{P_0} + g\\
  &= \frac{3.20}{63.50} + .06\\
  &= .11 = 11\%
\end{aligned}
$$

---

## Stock values

For the company in the previous problem, what is the dividend yield? What is the expected capital gains yield?

$$
\begin{aligned}
R &= \underbrace{\frac{\text{Div}}{P_0}}_{\text{dividend yield}} + \underbrace{g}_{\text{capital gains yield}}\\
R &= 5\% + 6\%
\end{aligned}
$$

The dividend yield is 5\% and the capital gains yield is 6\%

---

## Stock values

White Wedding Corporation will pay a \$2.65 per share dividend next year. The company pledges to increase its dividend by 4.75 percent per year, indefinitely. If the company has a dividend yield of 4.3 percent, what is the required return on the company's stock?

$$
\begin{aligned}
R &= \text{dividend yield} + \text{capital gains yield}\\
  &= .043 + .0475 \\
  &= .0905 = 9.05\%
\end{aligned}
$$

---

## Stock valuation

Siblings, Inc., is expected to maintain a constant 6.4 percent growth rate in its dividends, indefinitely. If the company has a dividend yield of 4.3 percent, what is the required return on the company's stock?

$$
\begin{aligned}
R &= \text{dividend yield} + \text{capital gains yield}\\
  &= .043 + .064\\
  &= .107 = 10.7\%
\end{aligned}
$$

---

## Stock valuation

Suppose you know that a company's stock currently sells for \$72 per share and the required return on the stock is 11.5 percent. You also know that the total return on the stock is evenly divided between a capital gains yield and a dividend yield. If it's the company's policy to always maintain a constant growth rate in its dividends, what is the current dividend per share?

Since the total return on the stock is evenly divided between capital gains and dividends:

$$
\begin{aligned}
R &= 2 \times g\\
g &= 5.75\%\\ \\
\end{aligned}
$$

The dividend yield is defined:

$$
\begin{aligned}
\text{dividend yield} &= \frac{\text{Div}_1}{P_0}\\ \\
\text{Div}_1 &= \text{Div}_0 \cdot (1 + g)\ \ \ \text{(next year's dividend)}\\ \\
\text{Div}_0 &= \frac{\text{dividend yield} \cdot P_0}{1 + g}\\
  &= \frac{.0575 \cdot 72}{1 + .0575}\\
  &= 3.91
\end{aligned}
$$

---

## Valuing preferred stock

Ayden, Inc., has an issue of preferred stock outstanding that pays a \$5.90 dividend every year, in perpetuity. If this issue currently sells for \$87 per share, what is the required return?

Preferred stock is almost like a bond, it pays out a fixed dividend. We can treat the dividend as a perpetuity:

$$
\begin{aligned}
P_0 &= \frac{\text{Div}}{R}\\ \\
R &= \frac{\text{Div}}{P_0}\\
  &= \frac{5.90}{87}\\
  &= .0678 = 6.78\%
\end{aligned}
$$

---

## Growth rate

The newspaper reported last week that Bennington Enterprises earned \$34 million this year. The report also stated that the firm's return on equity is 16 percent. Bennington retains 80 percent of its earnings. What is the firm's earnings growth rate? What will next year's earnings be?

$$
\begin{aligned}
g &= \text{retention ratio} \times \text{return on retained earnings (ROE)}\\
  &= .8 \times .16\\
  &= .128 = 12.8\%\\ \\
\text{Next years earnings} &= \text{earnings this year} \times (1 + g)\\
  &= 34 \times (1 + .128)\\
  &= 38.352
\end{aligned}
$$

---

## Stock valuation and PE

The Germinating Flower Co. has earnings of \$1.75 per share. The benchmark PE for the company is 18. What stock price would you consider appropriate? What if the benchmark PE were 21?

$$
\begin{aligned}
\text{PE} &= \frac{P}{E}\\
P &= \text{PE} \cdot E\\
  &= 18 \cdot 1.75\\
  &= 31.50
\end{aligned}
$$

---

## Stock valuation

Universal Laser, Inc., just paid a dividend of \$3.10 on its stock. The growth rate in dividends is expected to be a constant 6 percent per year indefinitely. Investors require a 15 percent return on the stock for the first three years, a 13 percent return for the next three years, and then an 11 percent return thereafter. What is the current share price for the stock?

Calculate the dividends for the six years:

$$
\begin{aligned}
\text{Div}_1 &= \text{Div}_{0} \times (1 + g)\\
  &= 3.10 \times 1.06\\
  &= 3.286\\ \\
\text{Div}_2 &= 3.10 \times (1.06)^2 = 3.483\\
\text{Div}_3 &= 3.10 \times (1.06)^3 = 3.692\\
\text{Div}_4 &= 3.10 \times (1.06)^4 = 3.914\\
\text{Div}_5 &= 3.10 \times (1.06)^5 = 4.148\\
\text{Div}_6 &= 3.10 \times (1.06)^6 = 4.397
\end{aligned}
$$

After six years the dividends grow perpetually at $g=6$%:

$$
\begin{aligned}
D_7 &= 4.397 \cdot (1.06) = 4.661\\
P_6 &= \frac{D_7}{R_3 - g}\\
  &= \frac{4.661}{.11 - .06}\\
  &= 93.22\\ \\
\text{PV} &= \frac{93.22}{(1.11)^{6}}\\
  &= 49.84
\end{aligned}
$$

$$
\begin{aligned}
\text{PV} &= \frac{3.286}{1.15} + \frac{3.483}{(1.15)^{2}} + \frac{3.692}{(1.15)^3}\\
  &= 7.919
\end{aligned}
$$

$$
\begin{aligned}
\text{PV} &= \frac{3.913}{(1.13)^4} + \frac{4.148}{(1.13)^5} + \frac{4.397}{(1.13)^6}\\
  &= 6.763
\end{aligned}
$$

$$
P_0 = 7.92 + 6.76 + 49.84 = 64.52
$$

---

## Nonconstant growth

Metallica Bearings, Inc., is a young start-up company. No dividends will be paid on the stock over the next nine years, because the firm needs to plow back its earnings to fuel growth. The company will pay a \$15 per share dividend in 10 years and will increase the dividend by 5.5 percent per year thereafter. If the required return on this stock is 13 percent, what is the current share price?

We will calculate the present value of the growth perpetuity and then discount it back to the present (nine years).

$$
\text{PV}_{+9} = \frac{15}{.13 - .055} = \frac{15}{.075} = 200
$$

Now we discount back

$$
PV = \frac{200}{(1.13)^{9}} = 66.58
$$

---

## Nonconstant dividends

Bucksnort, Inc., has an odd dividend policy. The company has just paid a dividend of \$12 per share and has announced that it will increase the dividend by \$3 per share for each of the next five years, and then never pay another dividend. If you require a return of 12 percent on the company's stock, how much will you pay for a share today?

The dividend payments will be:

$$
\text{Div}_{0} = 15\\
\text{Div}_{1} = 18\\
\text{Div}_{2} = 21\\
\text{Div}_{3} = 24\\
\text{Div}_{4} = 27\\
$$

The price of the stock is:

$$
P_{0} = \frac{15}{(1.12)} + \frac{18}{(1.12)^2} + \frac{21}{(1.12)^3}
  + \frac{24}{(1.12)^4} + \frac{27}{(1.12)^5} = 73.26
$$

---

## Nonconstant dividends

Lohn Corporation is expected to pay the following dividends over the next four years: \$10, \$7, \$6, and \$2.75. Afterwards, the company pledges to maintain a constant 5 percent growth rate in dividends forever. If the required return on the stock is 13 percent, what is the current share price?

The present value of the first four years is:

$$
\text{PV}_{\text{4yrs}} = \frac{10}{(1.13)} + \frac{7}{(1.13)^2} + \frac{6}{(1.13)^3} + \frac{2.75}{(1.13)^4} = 20.18
$$

Now the present value of the growth perpetuity, discounted back four years:

$$
\begin{aligned}
P_4 &= 2.75 \times (1.055) = 2.90
\text{PV}_{\text{perpituity}} &= \frac{2.75}{.13 - .05} = \frac{2.90}{.08} = 36.25\\ \\
\text{PV}_{\text{discounted}} &= \frac{36.25}{(1.13)^4} = 22.23
\end{aligned}
$$

Now we sum the two:

$$
20.18 + 22.23 = 42.41
$$

---

## Differential growth

Hughes Co. is growing quickly. Dividends are expected to grow at a rate of 20 percent for the next three years, with the growth rate falling off to a constant 5 percent thereafter. If the required return is 12 percent and the company just paid a \$2.80 dividend, what is the current share price?

We have a growth rate of 20 percent for the first three years. Then we have a perpetuity payment that grows at 5 percent.

$$
\begin{aligned}
D_1 &= 2.80 \times 1.2 = 3.36\\
D_2 &= 3.36 \times 1.2 = 4.03\\
D_3 &= 4.03 \times 1.2 = 4.84\\
D_4 &= 4.84 \times 1.05 = 5.08\\
\text{PV} &= \frac{3.36}{(1.12)} + \frac{4.03}{(1.12)^2} + \frac{4.84}{(1.12)^3}
  + \frac{\left(\frac{5.08}{.07}\right)}{(1.12)^3} = 61.13
\end{aligned}
$$

---

## Differential growth

Janicek Corp. is experiencing rapid growth. Dividends are expected to grow at 30 percent per year during the next three years, 18 percent over the following year, and then 8 percent per year indefinitely. The required return on this stock is 11 percent, and the stock currently sells for \$65 per share. What is the projected dividend for the coming year?

$$
\begin{aligned}
P_0 &= \frac{D_1}{1 + R} + \frac{D_2}{(1 + R)^2} + \frac{D_3}{(1 + R)^3}
  + \frac{D_4}{(1 + R)^4} + \frac{P_4}{(1 + R)^4}\\ \\
D_2 &= D_1 \cdot (1 + g) = D_1 \cdot 1.3 &= 1.3 \cdot D_1\\
D_3 &= D_1 \cdot (1 + g)^2 = D_1 \cdot (1.3)^2 &= 1.69 \cdot D_1\\
D_4 &= D_1 \cdot (1 + g) = D_1 \cdot (1.3)^2 \cdot (1.18) &= 1.99 \cdot D_1\\
D_5 &= D_1 \cdot (1 + g) = D_1 \cdot (1.3)^2 \cdot (1.18) \cdot (1.08) &= 2.15 \cdot D_1\\ \\
P_4 &= \frac{D_5}{R - g} = \frac{2.15}{.03} \cdot D_1 &= 71.67 \cdot D1\\
\text{Plugging in the values we get:}\\
P_0 &= \left( \underbrace{\frac{1.3}{1.11} + \frac{1.69}{(1.11)^2} + \frac{1.99}{(1.11)^3} + \frac{2.15}{(1.11)^4}}_{\text{initial growth}} + \underbrace{\frac{71.67}{(1.11)^4}}_{\text{growth perpetuity}}\right) \cdot D_1 \\ \\
D_1 &= \frac{65}{52.62} = 1.24
\end{aligned}
$$

---

## Negative growth

Antiques R Us is a mature manufacturing firm. The company just paid a dividend of \$9, but management expects to reduce the payout by 4 percent per year, indefinitely. If you require an 11 percent return on this stock, what will you pay for a share today?

$$
\begin{aligned}
D_1  &= 9 \cdot (1 - .04) = 8.64\\
P_0 &= \frac{D_1}{(.11 + .04)} = 57.60
\end{aligned}
$$

---

## Finding the dividend

Mau Corporation stock currently sells for \$58.32 per share. The market requires a return of 11.5 percent on the firm's stock. If the company maintains a constant 5 percent growth rate in dividends, what was the most recent dividend per share paid on the stock?

$$
\begin{aligned}
P_0 &= 58.32\\
P_0 &= \frac{D_1}{(R - g)}\\
D_1 &= P_0 \cdot (R - g) = 58.32 \cdot (.115 - .05) = 3.79\\
D_0 &= \frac{D_1}{(1 + g)} = \frac{3.79}{1.05} = 3.61\\ \\
\end{aligned}
$$

---

## Valuing preferred stock

Fifth National Bank just issued some new preferred stock. The issue will pay an annual dividend of \$8 in perpetuity, beginning five years from now. If the market requires a return of 5.6 percent on this investment, how much does a share of preferred stock cost today?

$$
\begin{aligned}
P_0 &= \frac{\frac{D}{R}}{(1 + R)^5}\\
  &= \frac{142.86}{(1.056)^5}\\
  &= 108.79
\end{aligned}
$$

---

## Using stock quotes

You have found the following stock quote for RJW Enterprises, Inc., in the financial pages of today's newspaper. What is the annual dividend? What was the closing price for this stock that appeared in _yesterday's_ paper? If the company currently has 25 million shares of stock outstanding, what was net income for the most recent four quarters?

```{r problem 20, echo=F}
p20.d <- c("-1.1", "RJW Enterp.", "RJW", "1.9", "23", "26.18", "-.13")
p20.h <- c("YTD %Chg", "Stock", "SYM", "YLD", "PE", "Last", "Net Chg")
p20.df <- data.frame()
p20.df <- rbind(p20.df, p20.d)
colnames(p20.df) <- p20.h
knitr::kable(p20.df)
```

$$
\begin{aligned}
\text{Dividend} &= \frac{\text{YLD}}{100} \cdot \text{Last}\\
 &= .019 \cdot 26.18\\
 &= .497\\ \\
\text{Yesterdays Closing} &= \text{Last} - \text{Net Chg}\\
 &= 26.18 - (-.13)\\
 &= 26.31\\ \\
\end{aligned}
$$

$$
\begin{aligned}
\text{P/E ratio} &= \frac{\text{stock price}}
  {\text{earnings per share (EPS)}}\\ \\
\text{EPS} &= \frac{\text{stock price}}{\text{P/E ratio}}\\
 &= \frac{26.18}{23}\\
 &= 1.13826\\ \\
\text{Net income} &= \text{EPS} \times \text{shares outstanding}\\
 &= 1.14 \times 25,000,000\\
 &= 28,500,000
\end{aligned}
$$

---

## Taxes and stock price

You own \$100,000 worth of Smart Money stock. One year from now, you will receive a dividend of \$2.25 per share. You will receive a \$2.40 dividend two years from now. You will sell the stock for \$65 per share three years from now. Dividends are taxed at the rate of 28 percent. Assume there is no capital gains tax. The aftertax required rate of return is 15 percent. How many shares of stock do you own?

$$
\begin{aligned}
P_0 &= \left(\frac{2.25}{(1.15)} + \frac{2.40}{(1.15)^2}\right)(1 - .28)
  + \frac{65}{(1.15)^3}\\
  &= 45.45\\ \\
\text{number of shares} &= \frac{100,000}{45.45} = 2200\ \text{shares}
\end{aligned}
$$

---

## Nonconstant growth and quarterly dividends

Pasqually Mineral Water, Inc., will pay a quarterly dividend per share of \$.80 at the end of each of the next 12 quarters. Thereafter, the dividend will grow at a quarterly rate of 1 percent, forever. The appropriate rate of return on the stock is 10 percent, compounded quarterly. What is the current stock price?

The quarterly rate is:

$$
\begin{aligned}
R &= \frac{10}{4} = 2.5\%\\
\end{aligned}
$$

---

For the first 12 quarters:

$$
\begin{aligned}
\text{PVIA} &= \frac{1 - (1 + R)^T}{R}\\
  &= \frac{1 - (1.025)^{12}}{.025}\\
  &= 10.25776\\
P_{\text{12qtrs}} &= .8 \cdot 10.25776\\
  &= 8.21\\
D_1 &= .8 \cdot (1.01)\\
  &= .808\\
P_{\text{perpetuity}} &= \frac{\frac{.808}{.025 - .01}}{(1.025)^{12}}\\
  &= 40.05\\
P_{\text{total}} &= 8.21 + 40.05 = 48.26
\end{aligned}
$$

---

## Finding the dividend

Briley, Inc., is expected to pay equal dividends at the end of each of the next two years. Thereafter, the dividend will grow at a constant annual rate of 4 percent, forever. The current stock price is \$45. What is next year's dividend payment if the required rate of return is 11 percent.

$$
\begin{aligned}
P_0 &= 45\\
P_0 &= \frac{D_1}{(1.11)} + \frac{D_1}{(1.11)^2} + \frac{P_2}{(1.11)^2}\\
P_2 &= \frac{D_1 \cdot (1.04)}{.11 - .04}\\
  &= 14.85714 \cdot D_1\\
P_0 &= D_1 \cdot \left(\frac{1}{(1.11)} + \frac{1}{(1.11)^2}
  + \frac{14.85714}{(1.11)^2}\right)\\
  &= 13.77091\\
D_1 &= \frac{45}{13.77091}\\
  &= 3.27
\end{aligned}
$$

---

## Finding the required return

Juggernaut Satellite Corporation earned \$18 million for the fiscal year ending yeasterday. The firm also paid out 30 percent of its earnings as dividends yesterday. The firm will continue to pay out 30 percent of its earnings as annual, end-of-year dividends. The remaining 70 percent of earnings is retained by the company for use in projects. The company has 2 million shares of common stock outstanding. The current stock price is \$93. The historical return on equity (ROE) of 13 percent is expected to continue in the future. What is the required rate of return on the stock?

$$
\begin{aligned}
P_0 &= 93\\
\text{payout} &= .3\\
\text{ROE} &= .13\\
g &= \text{retention ratio} \times \text{ROE}\\
  &= .7 \times .13\\
  &= .091\\
D &= \text{payout} \cdot \frac{\text{earnings}}{\text{stock shares}}
  \cdot \text{ROE}\\
  &= .3 \cdot \frac{18}{2} \cdot 1.091\\
  &= 2.9457\\
R &= \frac{D}{P_0} + g\\
  &= \frac{2.9457}{93} + .091\\
  &= .1227 = 12.27\%
\end{aligned}
$$

---

## Dividend growth

Four years ago, Bling Diamond, Inc., paid a dividend of \$1.35 per share. Bling paid a dividend of \$1.77 per share yesterday. Dividends will grow over the next five years at the same rate they grew over the last four years. Thereafter dividends will grow at 5 percent per year. What will Bling Diamond's cash dividend be in seven years?

$$
\begin{aligned}
D_{\text{-4yrs}} &= D_1 \cdot (1 + g)^4\\
\left(\frac{D_{\text{-4yrs}}}{D_1}\right)^{1/4} &= 1 + g\\
g &= \left(\frac{D_\text{-4yrs}}{D_1}\right)^{1/4} - 1\\
  &= .070\\ \\
D_0 &= 1.77\\
D_5 &= 1.77 \cdot (1.07)^5\\
    &= 2.4825\\
D_7 &= 2.4825 \cdot (1.05)^2\\
    &= 2.74
\end{aligned}
$$

---

## Price-earnings ratio

Consider Pacific Energy Company and U.S. Bluechips, Inc., both of which reported earnings of \$950,000. Without new projects, both firms will continue to generate earnings of \$950,000 in perpetuity. Assume that all earnings are paid as dividends and that both firms require a return of 12 percent.

a. What is the current PE ratio for each company?
b. Pacific Energy Company has a new project that will generate additional earnings of \$100,000 each year in perpetuity. Calculate the new PE ratio of the company.
c. U.S. Bluechips has a new project that will increase earnings by \$200,000 in perpetuity. Calculate the new PE ratio of the firm.

$$
\begin{aligned}
P_0 &= \frac{D}{R}\\
  &= \frac{950000}{.12}\\
  &= 7,916,667\\
\text{PE} &=  \frac{P_0}{E}\\
  &= \frac{7,916,667}{950,000}\\
  &= 8.33\\
\end{aligned}
$$

$$
\begin{aligned}
\Delta P &= \frac{100,000}{.12}\\
  &= 833,333.33\\
P &= 7,916,667 + 833,333.33\\
  &= 8,750,000\\
\text{PE} &= \frac{8,750,000}{1,050,000}\\
  &= 8.33
\end{aligned}
$$


$$
\begin{aligned}
\Delta P &= \frac{200,000}{.12}\\
  &= 1,666,667\\
P &= 7,916,667 + 1,666,667\\
  &= 9,583,334\\
\text{PE} &= \frac{9,583,334}{1,150,000}\\
  &= 8.33
\end{aligned}
$$

---

## Growth opportunities

The Stambaugh Corporation currently has earnings per share of \$9.40. The company has no growth and pays out all earnings as dividends. It has a new project that will require an investment of \$1.95 per share in one year. The project is only a two-year project, and it will increase earnings in the two years following the investment by \$2.75 and \$3.05, respectively. Investors require a return of 12 percent on Stambaugh stock.

a. What is the value per share of the company's stock assuming the firm does not undertake the investment opportunity?
b. If the company does undertake the investment, what is the value per share now?
c. Again, assume the company undertakes the investment. What will the price per share be four years from today?

$$
\begin{aligned}
\text{EPS} &= \text{D} = 9.40\\
P_0 &= \frac{9.40}{.12}\\
  &= 78.33\\ \\
P_0 &= \frac{12.15}{1.12} + \frac{15.20}{(1.12)^2} + \frac{\frac{9.40}{.12}}{(1.12)^2}\\
  &= 85.41\\
P_4 &= P_0 = 78.33
\end{aligned}
$$

---

## Growth opportunities

Rite Bite Enterprises sells toothpicks. Gross revenues last year were \$7.5 million, and total costs were \$3.4 million. Rite Bite has 1 million shares of common stock outstanding. Gross revenues and costs are expected to grow at 5 percent per year. Rite Bite pays no income taxes. All earnings are paid out as dividends.

a. If the appropriate discount rate is 13 percent and all cash flows are received at year's end, what is the price per share of Rite Bite stock?
b. Rite Bite has decided to produce toothbrushes. The project requires an immediate outlay of \$17 million. In one year, another outlay of \$6 million will be needed. The year after that, earnings will increase by \$4.2 million. That profit level will be maintained in perpetuity. What effect will undertaking this project have on price per share of the stock?

$$
\begin{aligned}
\text{Revenues} &= R = 7.5 - 3.4 = 4.1\ \text{million}\\
D &= R \cdot (1 + g)\\
  &= 4.1 \cdot 1.13\\
  &= 4.633\\
P_0 &= \frac{D}{R}\\
  &= \frac{4.633}{.13}\\
  &= 35.64\\ \\
\text{Investment} &= I_{0} = 17 + \frac{6}{(1.13)}\\
  &= 22.31\ \text{million}\\ \\
\text{NPV}_{\text{toothbrushes}} &= \frac{\frac{4.2}{.13}}{(1.13)^2} - 22.31\\
  &= 25.30 - 22.31\\
  &= 2.99
\end{aligned}
$$

---

## Growth opportunities

California Real Estate, Inc.,  expects to earn \$71 million per year in perpetuity if it does not undertake any new projects. The firm has an opportunity to invest \$16 million toady and \$5 million in one year in real estate. The new investment will generate annual earnings of \$11 million in perpetuity, beginning two years from today. The firm has 15 million shares of common stock outstanding, and the required rate of return on the stock is 12 percent. Land investments are not depreciable. Ignore taxes.

a. What is the price of a share of stock if the firm does not undertake the new investment?
b. What is the value of the investment?
c. What is the per-share stock price if the firm undertakes the investment?

$$
\begin{aligned}
P_0 &= \frac{\frac{71}{.12}}{15}\\
  &= 39.44\\
\text{PV}_{\text{investment}} &= \frac{\frac{11}{.12}}{(1.12)^2}
  - 16 - \frac{5}{(1.12)}\\
  &= 52.61\ \text{million}\\\\
P_{\text{with investment}} &= 39.44 + \frac{52.61}{15}\\
  &= 42.95
\end{aligned}
$$

---

## Stock valuation and PE

Ramsay Corp. currently has an EPS of \$2.35, and the benchmark PE for the company is 21. Earnings are expected to grow at 7 percent per year.

a. What is your estimate of the current stock price?
b. What is the target stock price in one year?
c. Assuming the company pays no dividends, what is the implied return on the company's stock over the next year? What does this tell you about the implicit stock return using PE valuation?

$$
\begin{aligned}
P_0 &= \text{PE} \cdot \text{EPS}\\
    &= 2.35 \cdot 21\\
    &= 49.35\\\\
P_1 &= \text{PE} \cdot (1 + g) \cdot \text{EPS}\\
    &= 2.35 \cdot 1.07 \cdot 21\\
    &= 52.80\\\\
\text{Implied return} &= \frac{\text{target price}
  - \text{current price}}{\text{current price}}\\
  &= \frac{52.80 - 49.35}{49.35}\\
  &\approx .07 = 7\%
\end{aligned}
$$

The implied stock return of 7 percent matches the earnings growth rate. This tells us that, under the assumption of a constant PE ratio and no dividends, the return on stock is entirely driven by earnings growth. This highlights the relationship between the PE ratio, earnings growth, and stock returns in valuation.

---

## Stock valuation and EV

FFDP Corp. has yearly sales of \$28 million and costs of \$12 million. The company's balance sheet shows debt of \$54 million and cash of \$18 million. There are 950,000 shares outstanding and the industry EV/EBITDA multiple is 7.5. What is the company's enterprise value? What is the stock price per share?

$$
\begin{aligned}
\text{EBITDA} &= 28 - 12 = 16\ \text{million}\\\\
\text{EV} &\approx \text{industry multiple} \cdot \text{EBITDA}\\
  &= 7.5 \cdot 16 = 120\ \text{million}\\\\
P &= \frac{120+18-54}{.95} = 88.42
\end{aligned}
$$

---

## Stock valuation and cash flows

Eberhart Manufacturing has projected sales of \$145 million next year. Costs are expected to be \$81 million and net investment is expected to be \$15 million. Each of these values is expected to grow at 14 percent the following year, with the growth rate declining by 2 percent per year until the growth rate reaches 6 percent, where it is expected to remain indefinitely. There are 5.5 million shares of stock outstanding and investors require a return of 13 percent return on the company's stock. The corporate tax rate is 40 percent.

a. What is your estimate of the current stock price?
b. Suppose instead that you estimate the terminal value of the company using a PE multiple. The industry PE multiple is 11. What is your new estimate of the company's stock price?

We will solve this with `R`:

```{r problem 31: stock valuation and cash flows}
options(scipen=99)
# all in millions
sales <- 145
costs <- 81
net.investment <- 15
shares <- 5.5
# 
(g <- c(0, .14 - c(0, .02, .04, .06, .08)))
g.p <- .06
r <- .13
tax <- .4
#
(ebitda <- cumprod(1+g)*(sales-costs))
(net.income <- ebitda * (1 - tax))
(retained.earnings <- net.investment * cumprod(1+g))
(payout <- net.income - retained.earnings)
(price <- sum(payout/(1+r)^(1:6) + payout[6] / (1+r)^6))/shares

## Using PE
pe <- 11
eps <- net.income / shares
(price.pe <- pe * eps)
```




