---
title: Chapter 26 Short-Term Finance and Planning
subtitle: Text
output:
  pdf_document:
    toc: true
    includes:
      in_headers: preamble.tex
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F)
require(tidyverse)
pretty <- function(x, n=2) format(round(x, n), big.mark=',')
```

The term _net working capital_ is associated with short-term financial decision
making. Short-term financial management is often called _working capital
management_.

There is no universally accepted definition of short-term finance. The most
important difference between short-term and long-term finance is in the timing
of cash flows. Short-term financial decisions typically involve cash inflows and
outflows that occur within a year. Long-term financial decisions are involve
decisions that span multiple years.

Typical short-term finance questions:

1. What is a reasonable level of cash to keep on hand (in a bank) to pay bills?

2. How much should the firm borrow in the short term?

3. How much credit should be extended to customers?

# 26.1 Tracing Cash and Net Working Capital

Short-term finance involves funding short-term operating activities of the firm.
This usually means their impact on cash and net working capital.

_Current assets_ are cash and other assets that are expected to be converted to
cash within the year. Current assets are presented on the balance sheet in order
of their accounting liquidity--the ease with which they can be converted to cash
and the time it takes to convert them.

The four most important items found in the current asset section of the balance
sheet are:

1. cash and cash equivalents

2. marketable securities

3. accounts receivable

4. inventories.

Short-term debt is called _current liabilities_. Current liabilities are
obligations that are expected to be paid off within one year (or within the
operating period). The three most important items in current liabilities are:

1. accounts payable

2. expenses payable

3. notes payable.

We focus on changes in cash. We restate the definition of net working capital

\begin{equation}
\text{Net working capital} = (\text{Cash} + \text{other current assets})
  - \text{current liabilities}\tag{26.2}\label{eq:26.2}
\end{equation}

Using the basic balance sheet identity we can define cash:

\begin{equation}
\text{Cash} = \text{long-term debt} + \text{equity}
  + \text{current liabilities}
  - \text{current assets other than cash}
  - \text{fixed assets}
  \tag{26.3}\label{eq:26.3}
\end{equation}

We can see that some activities increase cash and some activities decrease cash

\textbf{Activities that increase cash}

* Increasing long-term debt (borrowing over the long term)

* Increasing equity (selling some stock)

* Increasing current liabilities (getting a 90-day loan)

* Decreasing current assets other than cash (selling some inventory for cash)

* Decreasing fixed assets (selling some property)

\textbf{Activities that decrease cash}

* Decreasing long-term debt (paying off a long-term debt)

* Decreasing equity (repurchasing some stock)

* Decreasing current liabilties (paying off a 90-day loan)

* Increasing current assets other than cash (buying some inventory for cash)

* Increasing fixed assets (buying some property)

Activites that increase cash are called _sources of cash_, and activities that
decrease cash are called _uses of cash_.

#### Example 26.1 Sources and Uses

Accounts payable are what we oew our suppliers. This is a short-term debt. If
accounts payable rises by \$100, we have effectively borrowed the money, which
is a _source_ of cash. Receivables are what our customers owe to us, so an
increase of \$100 in accounts receivable means that we have loaned the moeny;
this is a _use_ of cash.

# 26.2 The Operating Cycle and the Cash Cycle

## Defining the operating and cash cycles

We will start with a simple example. On Day 0, we purchase \$1,000 worth of
inventory on credit. We pay the bill 30 days later, and, after 30 more days,
someone buys the \$1,000 in inventory for \$1,400. Our buyer does not actually
pay for another 45 days. The following table summarizes the events.

```{r example table page 807}
tbl = tibble(
  Day = c(0, 30, 60, 105),
  Activity = c("Acquire inventory",
               "Pay for inventory",
               "Sell inventory on credit",
               "Collect on sale"),
  `Cash Effect` = c(NA, -1000, NA, 1400)
)
tbl.display = tbl
tbl.display$`Cash Effect` = pretty(tbl.display$`Cash Effect`)
knitr::kable(tbl.display)
```

#### The Operating Cycle

The operating cycle consists of the time we acquire some inventory to the time
we collect the cash, in our example it is 105 days. Therefore, the operating
cycle is defined as the time it takes to acquire inventory, sell it, and collect
for it. There are two components to this cycle. First, the time it takes to
acquire and sell the inventory (60 days in our example) is called the inventory
period. Second, is the time it takes to collect on the sale (45 days), called
the accounts receivable period.

Therefore, the operating cycle can be defined:

\begin{equation}
\text{Operating cycle} = \text{Inventory period}
  + \text{Accounts receivable period}\tag{26.4}\label{eq:26.4}
\end{equation}

The operating cycle describes how a product moves through the current asset
accounts. A product begins as inventory when it is bought, it is converted to a
receivable when it is sold, and it is finally converted to  cash when we collect
from the sale.

#### The Cash Cycle

The cash flows and other events are not synchronized. In our example, we don't
actually pay for the inventory until 30 days after acquiring it. This period is
called the accounts payable period. Next we spend cash on Day 30, but we don't
collect until Day 105. We have to finance the \$1,000 for 105 - 30 = 75 days.
This period is called the cash cycle.

The cash cycle is the number of days that pass before we collect othe cash from
a sale. This is measured from when we actually pay for the inventory. Based on
our definitions, the cash cylce is the difference between the operating cycle
and the accounts payable period:

\begin{equation}
\text{Cash cycle} = \text{Operating cycle} - \text{Accounts payable period}
  \tag{26.5}\label{eq:26.5}
\end{equation}

The gap between short-term inflows and outflows can be filled either by
borrowing or by holding a liquidity reserve in the form of cash or marketable
securities. Alternatively, the gap can be shortened by changing the inventory,
receivable, and payable periods. These are all managerial options.

## The Operating cycle and the firm's organization chart

The following shows a list of manager titles and their relationship with
short-term financial decision making:

| Title of Manager | Duties related to short-term financial management | Assets/Liabilities Influenced |
|:---|:---|:---|
| Cash manager | Collection, concentration, disbursement; short-term investments; short-term borrowing; banking relations | Cash, marketable securities, short-term loans |
||
| Credit manager | Monitoring and control of accounts receivable; credit policy descisions | Accounts receivable |
||
| Marketing manager | Credit policy decisions | Accounts receivable |
||
| Purchasing manager | Decisions on purchases, suppliers; may negotiate payment terms | Inventory, accounts payable |
||
| Production manager | Setting of production schedules and materials requirements | Inventory, accounts payable |
||
| Payables manager | Decisions on payment policies and on whether to take discounts | Accounts payable |
||
| Controller | Accounting information on cash flows; reconciliation of accounts payable; application of payments to accounts receivable | Accounts receivable, accoutns payable |
||

Examining the table we can see that there are a lot of different financial and
nonfinancial managers who influence short-term finances. There is potential for
conflict between these managers.

## Calculating the operating and cash cycles

We can sue some of the ratios stated in chapter 3 here.

### The operating cycle

We can calculate the inventory period.

\[
\text{Inventory turnover} = \frac{\text{Cost of goods sold}}
  {\text{Average inventory}} = \frac{\$8.2\ \text{million}}{2.5\ \text{million}}
  = 3.28\ \text{times}
\]

This example shows that we bought and sold our inventory 3.28 times during the
year.

\[
\text{Inventory period} = \frac{365\ \text{days}}{\text{Inventory turnover}}
  = \frac{365}{3.28} = 111.3\ \text{days}
\]

So, the inventory period is about 111 days. On average, in other words,
inventory sat for about 111 days before it was sold.

Similarly, receivables averaged \$1.8 million, and sales were \$11.5 million.
Assuming that all sales were credit sales, the receivables turnover is:

\[
\text{Receivables turnover} = \frac{\text{Credit sales}}
  {\text{Average accounts receivable}} = \frac{11.5\ \text{million}}
  {1.8\ \text{million}} = 6.4\ \text{times}
\]

If we turn over our receivables 6.4 times a year, then the receivables period
is:

\[
\text{Receivables period} = \frac{365\ \text{days}}{\text{Receivables turnover}}
  = \frac{365}{6.4} = 57\ \text{days}
\]

The receivables period is also called the _days' sales in receivables_, or the
_average collection period_. Whatever it is called, it tells us that our
customers took an average of 57 days to pay.

The operating cycle is the sum of the inventory and receivables periods:

\[
\text{Operating cycle} = \text{Inventory period}
  + \text{Accounts receivable period} = 111\ \text{days} + 57\ \text{days}
  = 168\ \text{days}
\]

This tells us that, on average, 168 days elapse between the time we acquire
inventory and, having sold it, collect for the sale.

### The cash cycle

We now calculate the payables period.

\[
\text{Payables turnover} = \frac{\text{Cost of goods sold}}
  {\text{Average payables}} = \frac{\$8.2\ \text{million}}{.875\ \text{million}}
  = 9.4\ \text{times}
\]

The payables period is:

\[
\text{Payables period} = \frac{365\ \text{days}}{\text{Payables turnover}}
  = \frac{365}{9.4} = 39\ \text{days}
\]

Thus, we took an average of 39 days to pay our bills.

Finally, the cash cycle is the difference between the operating cycle and the
payables period:

\[
\text{Cash cycle} = \text{Operating cycle} - \text{Accounts payable period}
  = 168\ \text{days} - 39\ \text{days} = 129\ \text{days}
\]

So, on average, there is a 129-day delay between the time we pay for merchandise
and the time we collect on the sale.

#### Example 26.2 The operating and cash cycles

You have collected the following information for the Slowpay Company:

| Item | Beginning | Ending |
|:---|---:|---:|
| Inventory | \$5,000 | \$7,000 |
| Accounts receivable | \ 1,600 | \ 2,400 |
| Accounts payable | \ 2,700 | \ 4,800 |

Credit sales for the year just ended were \$50,000, and cost of goods sold was
\$30,000. How long does it take Slowpay to collect on its receivables? How long
does merchandise stay around before it is sold? How long does Slowpay take to
pay its bills?

We can first calculate the three turnover ratios:

\[
\begin{aligned}
\text{Inventory turnover} &= \$30,000/6,000 &= 5\ \ \text{times}\\
\text{Receivables turnover} &= \$50,000/2,000 &= 25\ \text{times}\\
\text{Payables turnover} &= \$30,000/3,750 &= 8\ \ \text{times}
\end{aligned}
\]

We use these to get the various periods:

\[
\begin{aligned}
\text{Inventory period} &= 365/5 &= 73\ \ \ \text{days}\\
\text{Receivables period} &= 365/25 &= 14.6\ \text{days}\\
\text{Payables period} &= 365/8 &= 8\ \ \ \ \text{days}
\end{aligned}
\]

All told, Slowpay collects on a sale in 14.6 days, inventory sits around for 73
days, and bills get paid after about 46 days. The operating cycle here is the
sum of the inventory and receivables periods: 73 + 14.6 = 87.6 days. The cash
cycle is the difference between the operating cycle and the payables period: 876
- 45.6 = 42 days.

## Interpreting the cash cycle

The cash cycle depends on the inventory, receivables, and payables periods. The
cash cycle increases as the inventory and receivables periods get longer. It
decreases if the company is able to defer payment of payables and thereby
lengthen the payables period.

The link between a firm's cash cycle and its profitability can be seen by
recalling that one of the basic determinants of profitability and growth for a
firm is its total asset turnover, which is defined as Sales/Total assets. The
higher this ratio is, the greater are the firm's accounting return on assets,
ROA, and return on equity, ROE. Thus, all other things being the same, the
shorter the cash cycle is, the lower is the firm's investment in inventories and
receivables. As a result, the firm's total assets are lower, and total turnover
is higher.

## A look at operating and cash cycles


