--- 
title: "Financial Statements Analysis and Financial Models"
subtitle: "Text"
author: "Ross, Westerfield, Jaffe"
header-includes:
  - \usepackage{booktabs}
toc: true 
output: pdf_document
pdf_engine: xelatex
---

```{r setup, echo=FALSE, includes=FALSE, message=FALSE}
require(kableExtra)
require(dplyr)
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
pretty <- function(x) format(x, big.mark=",")
```

# Chapter introduction

The price of a share of common stock in green tractor manufacturer John Deere
closed at about \$88 on May 13, 2011. At that price, John Deere had a
price-earnings (PE) ratio of 17. That is, investors were willing to pay \$17 for
every dollar in income earned by John Deere. At the same time, investors were
willing to pay \$9, \$13, and \$27 for each dollar earned by Ford, Coca-Cola,
and Google, respectively. At the other extreme were Bank of America and United
States Steel. Both had negative earnings for the previous year, but Bank of
America was priced at about \$12 per share and United States Steel at about \$45
per share. Because they had negative earnings, their PE ratios would have been
negative, so they were not reported. At the same time, the typical stock in the
S&P 500 index of large company stocks was trading at a PE of about 14, or about
14 times earnings, as they say on Wall Street.

Price--earnings comparisons are examples of the use of financial ratios. As we
will see in this chapter, there are a wide variety of financial ratios, all
designed to summarize specific aspects of a firm's financial position. In
addition to discussing how to analyze financial statements and compute financial
ratios, we will have quite a bit to say about who uses this information and why.

---

# List of topics

1. Financial statement analysis

   * Standardizing statements
   * Common-size balance sheets
   * Common-size income statements

2. Ratio analysis

   * Short-term solvency or liquidity measures
   * Long-term solvency measures
   * Asset management or turnover measures
   * Profitability measures
   * Market value measures

3. The DuPont Identity

   * A closer look at ROE
   * Problems with financial statement analysis

4. Financial models

   * A simple planning model
   * The percentage of sales approach

5. External financing and growth

   * EFN and growth
   * Financial policy and growth
   * A note about sustainable growth rate calculations

6. Some caveats regarding financial planning models

---

# Notes

## Financial statements analysis

### Standardizing statements

Comparing company's financial statements is done by standarizing their
statements to common-size statements.

### Common-size balance sheets

We need a reference. We can use percentage of assets.

### Common-size income statements

Common measures of earnigns.

#### Net Income

The so-called bottom line, defined as total revenue minus total expenses. Net
income reflects differences in a firm's capital structure and taxes as well as
operating income. Interest expense and taxes are subtracted from operating
income in computing net income. Shareholders look closely at net income because
dividend payout and retained earings are closely linked to net income.

#### EPS

Net income divided by the number of shares outstanding. It expresses net income
on a per share basis.

#### EBIT

Earnings before interest expense and taxes. EBIT is usually called "income from
operations" on the income statement and is income before unusual items,
discontinued operations or extrodinary items. To calculate EBIT, operating
expenses are subtracted from total operations revenues. Analysts like BIT
because it abstracts from differences in earnings from a firm's capital
structure (interest expense) and taxes.

#### EBITDA

Earnings before interest expenses, taxes, depreciation, and amortization refers
to a noncash expense similar to depreciation except it applies to an intangible
asset (such as a patent), rather than a tangible asset (such as a machine). The
word amortization here does not refer to the payment of debt. Analysts like to
use EBITDA because it adds back two noncash items (depreciation and
amortization) to EBIT and thus is a better measure of before-tax operating cash
flow.

## Ratio Analysis

### Short-term solvency or liquidity measures

#### Current ratio

$$
\text{Current ratio} = \frac{\text{Current assets}}{\text{Current liabilties}}
$$

#### Quick (or Acid-Test) Ratio

$$
\text{Quick ratio} = \frac{\text{current assets} - \text{inventory}}
  {\text{current liabilities}}
$$

#### Cash ratio

$$
\text{Cash ratio} = \frac{\text{cash}}{\text{current liabilities}}
$$

### Long-term solvency measures

#### Total debt ratio

$$
\text{Total debt ratio} = \frac{\text{total assets} - \text{total equity}}
  {\text{total assets}}
$$

#### Debt-equity ratio

$$
\text{Debt-equity ratio} = \frac{\text{total debt}}{\text{total equity}}
$$

#### Equity multiplier

$$
\text{Equity multiplier} = \frac{\text{total assets}}{\text{total equity}}
$$

#### Times interest earned

$$
\text{Times interest earned ratio} = \frac{\text{EBIT}}{\text{interest}}
$$

##### Cash coverage

$$
\text{Cash coverage ratio} = \frac{\text{EBIT}
  + (\text{Depreciation and amortization})}{\text{interest}}
$$

### Asset managment and turnover measures

#### Inventory turnover and days' sales in inventory

$$
\text{inventory turnover} = \frac{\text{cost of goods sold}}
  {\text{inventory}}
$$

$$
\text{Days' sales in inventory} = \frac{\text{365 days}}
  {\text{inventory turnover}}
$$

#### Receivables turnover and days' sales in receivables

$$
\text{Receivables turnover} = \frac{\text{sales}}{\text{accounts receivable}}
$$

$$
\text{Days' sales in receivables} = \frac{\text{365 days}}
  {\text{receivables turnover}}
$$

#### Total asset turnover

$$
\text{Total asset turnover} = \frac{\text{sales}}{\text{total assets}}
$$

### Profitability measures

#### Profit margin

$$
\text{Profit margin} = \frac{\text{net income}}{\text{sales}}
$$

#### EBITDA margin

$$
\text{EBITDA margin} = \frac{\text{EBITDA}}{\text{sales}}
$$

#### Return on assets

$$
\text{Return on assets} = \frac{\text{net income}}
  {\text{total assets}}
$$

#### Return on equity

$$
\text{Return on equity} = \frac{\text{net income}}
  {\text{total equity}}
$$

### Market value measures

#### Price-earnings ratio

$$
\text{PE ratio} = \frac{\text{price per share}}{\text{earnings per share}}
$$

#### Market-to-book ratio

$$
\text{Market-to-book ratio} = \frac{\text{market value per share}}
  {\text{book value per share}}
$$

#### Market capitalization

$$
\text{Market capitalization} = \text{price per share}
  \times \text{shares outstanding}
$$

#### Enterprise value

$$
\text{Enterprise value} = \text{market capitalization}
  + \text{market value of interest bearing debt}
$$

#### Enterprise value multiples

$$
\text{Enterprise value multiples} = \frac{\text{enterprise value}}
  {\text{EBITDA}}
$$

## The DuPont identity

The difference between ROA and ROE reflects the firm's use of debt financing.
Higher ROE compared to ROA means heavy debt financing.

$$
\begin{aligned}
\text{Return on equity} &= \frac{\text{net income}}{\text{total equity}}\\\\
  &= \frac{\text{net income}}{\text{total equity}}
  \times \frac{\text{assets}}{\text{assets}}\\\\
  &= \frac{\text{net income}}{\text{assets}}
  \times \frac{\text{assets}}{\text{total equity}}\\\\
  &= \text{ROA} \times \text{equity multiplier}\\\\
  &= \text{ROA} \times (1 + \text{debt-equity ratio})
\end{aligned}
$$

The Dupont Identity:

$$
\begin{aligned}
\text{ROE} &= \underbrace{\frac{\text{net income}}{\text{sales}}
  \times \frac{\text{sales}}{\text{assets}}}_{\text{Return on assets}}
  \times \frac{\text{assets}}{\text{total equity}}\\
  &= \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier}
\end{aligned}\tag{3.21}
$$

We partitioned return on assets (ROA) into two components: profit margin and
total asset turnover.

The DuPont identity tells us that ROE is affected by three things:

1. Operating efficiency (as measured by profit margin).
2. Asset use efficiency (as measured by total asset turnover).
3. Financial leverage (as measured by the equity multiplier).


Decomposing ROE into these components is a good start at financial analysis and
will give information about the current state of the business.

### Problems with financial statement analysis

Ther is no financial analysis theory that tells you which ratios are more
important.

Difficulty in assessing firms exist because of many factors such as:

1.  Some firms are conglamerates of multiple businesses making analysis
    difficult.

2. Globalization spreads competition out into different countries, using
   different currencies, operating in different regulatory and political
   environments. All these factors make comparisons difficult.

3. Different accounting procedures can make comparisons difficult. For instance,
   differences in inventory accounting can make financial analysis very
   difficult.

## Financial models

### A simple financial planning model

Simple example to demonstrate financial planning and estimation. We will use the
percentage of sales approach.

All variables will be tied to the rate of growth of sales.

The income statement and the balance sheet has been simplified.

```{r financial models; computerfield income statement, echo=F}
computerfield <- new.env()
computerfield$is <- matrix(c(1000, 800, 200), ncol=1)
rownames(computerfield$is) <- c("Sales", "Costs", "Net income")

knitr::kable(pretty(computerfield$is), align = "lr",
             caption="Computerfield corporation, income statement") %>%
    add_indent(c(3))
```

```{r financial models; computerfield balance sheet, echo=F}
computerfield$bs <- matrix(c(500, 500, 250, 250, 500), ncol=1)
rownames(computerfield$bs) <- c("Assets", "Total Assets", "Debt", "Equity",
                                "Total Debt and Equity")

knitr::kable(pretty(computerfield$bs),
      align = "lr", caption="Computerfield Corporation balance sheet") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    pack_rows("Assets", 1, 2, bold=T, background="#f2f2f2") %>%
    pack_rows("Debt and Equity", 3, 5, bold=T, background="#D3E4CD") %>%
    add_indent(c(2, 5))
```

If sales increase by \%20, rising from \$1,000 to \$1,200, then the pro forma
income statement would be:

```{r financial models; computerfield pro forma income statement, echo=F}
computerfield$is.proforma <- computerfield$is
computerfield$is.proforma <- computerfield$is * 1.2
knitr::kable(pretty(computerfield$is.proforma), align="lr",
    caption="Computerfield corporation, pro forma income statement") %>%
  add_indent(c(3))
```

Using the assumption that all variables will grow by 20 percent we will be able
to construct the pro forma balance sheet:

Similar to the income statement, if sales grow \%20 then the pro forma balance
sheet

```{r financial models; computerfield pro forma balance sheet, echo=F}
computerfield$bs.proforma <- computerfield$bs
computerfield$bs.proforma <- computerfield$bs * 1.2
rownames(computerfield$bs.proforma) <-
    c("Assets (+100)", "Total Assets (+100)", "Debt (+50)",
      "Equity (+50)", "Total Debt and Equity (+100)")
knitr::kable(computerfield$bs.proforma, align = "lr",
      caption="Computerfield Corporation pro forma balance sheet") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    pack_rows("Assets", 1, 2, bold=T, background="#f2f2f2") %>%
    pack_rows("Debt and Equity", 3, 5, bold=T, background="#D3E4CD") %>%
    add_indent(c(2, 5))
```

Now that we have constructed the two pro forma financial statements we need to
reconcile the two. How can net income be equal to \$240 and equity increase by
only \$50? The answer is that Computerfield must have paid out the difference.
$\$240 - 50 = \$190$ could be paid out as a cash dividend. The cash dividend is
called a _plug variable_.

If Computerfield did not pay out the \$190 dividend then that means that
retained earnings would have grown the full \$240, making equity grow to $250 +
240 = \$490$, and so debt must be retired to keep total assets equal to \$600.

Computerfield would have to retire \$490 in debt making the new debt balance: 
$\$600 - 490 = \$110$. This new pro forma balance sheet would look like the
following:

```{r financial models; computerfield alternate pro forma balance sheet, echo=F}
computerfield$bs.proforma.alt <- computerfield$bs
computerfield$bs.proforma.alt <- matrix(c(600, 600, 110, 490, 600), ncol=1)
rownames(computerfield$bs.proforma.alt) <-
    c("Assets (+100)", "Total Assets (+100)", "Debt (-140)",
      "Equity (+240)", "Total Debt and Equity (+100)")
knitr::kable(pretty(computerfield$bs.proforma.alt), align = "lr",
      caption="Computerfield Corporation alternate pro forma balance sheet") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    pack_rows("Assets", 1, 2, bold=T, background="#f2f2f2") %>%
    pack_rows("Debt and Equity", 3, 5, bold=T, background="#D3E4CD") %>%
    add_indent(c(2, 5))
```

In this case, the _plug variable_ used to balance projected total assets and
total liabilities will be debt.

### The percentage of sales approach

The previous example used the percentage of sales approach in which every item
increased at the same rate as sales. This may be reasonable for some accounts
but for others, such as long-term borrowing, it probably is not. Long-term
borrowing is set by management, and it does not necessarily relate directly to
the level of sales.

Here we will separate hte income statement and balance sheet accounts into two
groups, those that vary directly with sales and those that are independent.

#### The income statement

Below is the recent income statement for the Rosengarten Corporation. The income
statement has been simplified, combining costs, depreciation, and interest in a
single cost figure.

Rosengarten is projecting a 25 percent increase in sales for the coming year.

```{r financial models; rosengarten corporation; income statement, echo=F}
rosengarten <- new.env()
rosengarten$growth.rate <- .25
rosengarten$is <- matrix(c(1000, 800, 200, 68, 132, 44, 88), ncol=1)
rownames(rosengarten$is) <- c("Sales", "Costs", "Taxable income", "Taxes (34%)",
                              "Net income", "Dividends",
                              "Addition to retained earnings")
knitr::kable(pretty(rosengarten$is),
      col.names=c("Item", "Amount ($)"), align="lr",
      caption="Rosengarten Corporation income statement") %>%
    kable_styling(bootstrap_options="striped", full_width=F) %>%
    add_indent(c(5))
```

Rosengarten has projected a 25 percent increase in sales for the coming year, so
we are anticipating sales of $\$1,000 \times 1.25 = \$1,250$. To generate the
pro forma income statement we will assume that total costs will continue to run
at $\$800/1,000 = 80$ percent of sales. With this assumption, Rosengarten's pro
forma income statement is as shown:

```{r financial models; rosengarten corp; pro forma income statement, echo=F}
rosengarten$is.proforma <- matrix(
    rosengarten$is[1:5,] * (1 + rosengarten$growth.rate), ncol=1)
rownames(rosengarten$is.proforma) <-
    c("Sales (projected)", "Costs", "Taxable income",
      "Taxes (34%)", "Net income")
knitr::kable(pretty(rosengarten$is.proforma), align = "lr",
     caption="Rosengarten Corporation pro forma income statement") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   add_indent(c(5))
```

To assume that costs are a constant percentage of sales is to assume that the
profit margin is constant.

$$
\begin{aligned}
\text{Profit margin} = \frac{\text{net income}}{\text{sales}}
  = \frac{132}{1,000} = \frac{165}{1,250} = .132
\end{aligned}
$$

Next we project the dividend payment. This is up to management. Assume
Rosengarten has a policy of paying out a constant fraction of net income in the
form of a cash dividend. The _dividend payout ratio_ was:

```{r financial models; rosengarten dividend ratio, echo=F}
rosengarten$dividend.ratio <- .3333
```

$$
\text{Dividend payout ratio} = \frac{\text{cash dividends}}{\text{net income}}
  = \frac{88}{132} = 33\ \tfrac{1}{3}\%
$$

The ratio of addition to retained earnings to net income (also called the
_retention ratio_ or the _plowbak ratio_):

$$
\text{retention ratio} = \frac{\text{Addition to retained earnings}}
  {\text{net income}} = \$88/132 = 66\ \tfrac{2}{3}\%
$$

Therefore, we predict the following for Dividends and the addition to retained
earnings:

$$
\begin{aligned}
\text{Projected dividends paid to shareholders} &= \$165 \times 1/3 = &\$ &\ 55\\
\text{Projected addition to retained earnings} &= \$165 \times 2/3 = & &110\\
& &\$ &165
\end{aligned}
$$

#### The balance sheet

We assume that some of the items on the balance sheet vary directly with sales
and other items do not. For those items that vary with sales, we express each as
a percentage of sales for the year just completed. For items that do not vary
with sales we use `NA`.

As an example, on the asset side, we assume that inventory is 60 percent of
sales. That means for every \$1 increase in sales, inventory will increase
\$.60. The ratio of total assets to sales for the year is
$\$3,000/1,000 = 3$, or 300 percent. This ratio of total assets to sales is
called the _capital intensity ratio_. This ratio tells us the amount of assets
we need to generate \$1 in sales; the higher the ratio is, the more capital
intensive the firm is. This ratio is just the reciprical of the total asset
turnover.

```{r financial models; rosengarten corp; balance sheet, echo=F}
rosengarten$bs <- matrix(c(160, 440, 600, 1200, 1800, 3000, 300, 100,
                           400, 800, 800, 1000, 1800, 3000), ncol=1)
rownames(rosengarten$bs) <- c("Cash", "Accounts receivable", "Inventory",
                              "Total current assets", "Net plant and equipment",
                              "Total assets", "Accounts payable",
                              "Notes payable", "Total current liabilities",
                              "Long-term debt",
                              "Common stock and paid-in surplus",
                              "Retained earnings", "Total equity",
                              "Total liabilities and owners' equity")
rosengarten$bs.pct.of.sales <- c(.16, .44, .60, 1.2, 1.8, 3, .3,
                                 NA, NA, NA, NA, NA, NA, NA)
rosengarten$bs <- cbind(rosengarten$bs, rosengarten$bs.pct.of.sales)

knitr::kable(pretty(rosengarten$bs),
     col.names=c("", "Amount ($)", "% of sales"), align="lrr",
     caption="Rosengarten Corporation balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Current assets", 1, 4, bold=T, background="#f2f2f2") %>%
   pack_rows("Fixed assets", 5, 5, bold=T, background="#D3E4CD") %>%
   pack_rows("Current liabilities", 7, 9, bold=T, background="#FAD4D4") %>%
   pack_rows("Owners' equity", 11, 13, bold=T, background="#D9E4EC") %>%
   row_spec(5, extra_latex_after = "\\addlinespace") %>%
   row_spec(6, extra_latex_after = "\\addlinespace") %>%
   row_spec(13, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(4, 9, 13))
```

On the liability side of the balance sheet, we show accounts payable varying
with sales. This is because we expect more orders with suppliers as sales
volumes grow. Notes payable represents short term borrowing and will not vary
unless we take specific actions to change the amount. This also is the case for
long term debt and common stock and paid-in surplus. Retained earnings will vary
with sales, but it won't be a simple percentage of sales. Instead, we will
explicitly calculate the change in retained earnings based on our projected net
income and dividends.

We can now construct a partial pro forma balance sheet for Rosengarten. We use
the percentage of sales (25%) to calculate the change in each item that we want
to vary by sales. All of the items on the asset side of the balance sheet. On
the liabilities and owners' equity side, only accounts payable will vary
directly with sales. The items that we don't vary directly with sales are the
items that management will control. Notice that the change in retained earnings
is equal to the \$110 addition to retained earnings calculated in the income
statement.

```{r financial models; rosengarten corp; pro forma balance sheet, echo=F}
tmp <- matrix(rosengarten$bs[,1], ncol=1)
rownames(tmp) <- rownames(rosengarten$bs)
rosengarten$bs.proforma <- tmp
rosengarten$bs.proforma[1:7] <- tmp * (1 + rosengarten$growth.rate)
rosengarten$bs.proforma["Total current liabilities",] <-
    rosengarten$bs.proforma["Accounts payable",] +
    rosengarten$bs.proforma["Notes payable",]
rosengarten$bs.proforma["Retained earnings",] <-
    tmp["Retained earnings",] +
    rosengarten$is.proforma["Net income",] * (1 - rosengarten$dividend.ratio)
rosengarten$bs.proforma["Total equity",] <-
    tmp["Total equity",] +
    rosengarten$is.proforma["Net income",] * (1 - rosengarten$dividend.ratio)
rosengarten$bs.proforma["Total liabilities and owners' equity",] <-
    rosengarten$bs.proforma["Total current liabilities",] +
    rosengarten$bs.proforma["Long-term debt",] +
    rosengarten$bs.proforma["Total equity",]
rosengarten$bs.proforma.efn <-
    rosengarten$bs.proforma["Total assets",] -
    rosengarten$bs.proforma["Total liabilities and owners' equity",]
rosengarten$bs.proforma.delta <- c(rosengarten$bs.proforma - rosengarten$bs[,1],
                                   rosengarten$bs.proforma.efn)
rosengarten$bs.proforma <- rbind(rosengarten$bs.proforma,
                                 rosengarten$bs.proforma.efn)
rosengarten$bs.proforma <- cbind(rosengarten$bs.proforma,
                                 rosengarten$bs.proforma.delta)
rownames(rosengarten$bs.proforma) <- c(rownames(rosengarten$bs),
                                       "External financing needed")
knitr::kable(pretty(round(rosengarten$bs.proforma)),
     col.names=c("Item", "Amount ($)", "Change from current year"), align="lrr",
     caption="Rosengarten Corporation pro forma balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Current assets", 1, 4, bold=T, background="#f2f2f2") %>%
   pack_rows("Fixed assets", 5, 5, bold=T, background="#D3E4CD") %>%
   pack_rows("Current liabilities", 7, 9, bold=T, background="#FAD4D4") %>%
   pack_rows("Owners' equity", 11, 13, bold=T, background="#D9E4EC") %>%
   row_spec(5, extra_latex_after = "\\addlinespace") %>%
   row_spec(6, extra_latex_after = "\\addlinespace") %>%
   row_spec(13, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(4, 9, 13))
```

Inspecting the pro forma balance sheet, notice that assets are projected to
increase by \$750. However, without additional financing, liabilities and equity
will increase by only \$185, leaving a shortfall of
\$750 - 185 = \$`r #rosengarten$bs.proforma[15,2]`. We label this amount as
_external financing needed_ (EFN) in the balance sheet. We could also calculate
EFN directly:

$$
\begin{aligned}
\text{EFN} = &\frac{\text{assets}}{\text{sales}} \times \Delta \text{sales}
  - \frac{\text{spontaneous liabilities}}{\text{sales}}
    \times \Delta \text{sales} - \text{PM}\\
  & \times \text{projected sales} \times (1 - d)
\end{aligned}\tag{3.23}
$$

In this expression, "$\Delta$sales" is the projected change in sales (in
dollars); "Spontaneous liabilities" are the liabilities that naturally move up
and down with sales (here its accounts payable); and PM and d are the profit
margin and dividend payout ratios. For our example:

$$
\text{EFN} = \frac{\$3,000}{1,000} \times \$250 - \frac{\$300}{1,000}
  \times \$250 - .132 \times \$1,250 \times \left(1 - \frac{1}{3} \right)
  = \$565
$$

#### A particular scenario

The current projection of 25 percent sales growth will require \$565 in new
financing. Now management will need to develop a strategy to raise the funds for
the growth. There are three possible sources of new funding: short-term
borrowing, long-term borrowing, and new equity. Management can choose some
combination of any of these.

For our example, suppose management has a policy of not issuing new equity.
Management would also like to keep total net working capital unchanged. In this
scenario, with current assets increasing by \$300 and accounts payable only
increasing by \$75, management could borrow \$300 - \$75 = \$225 in short-term
notes payable. This would leave \$565 - 225 = \$340 coming from long-term
borrowing. Below is the updated pro forma balance sheet.

```{r financial models; rosengarten corp; scenario, echo=F}
tmp <- matrix(rosengarten$bs.proforma[-nrow(rosengarten$bs.proforma),1], ncol=1)
rownames(tmp) <- rownames(rosengarten$bs.proforma)[-15]
rosengarten$bs.scenario <- round(tmp)
rosengarten$bs.scenario["Notes payable",] <- 
    rosengarten$bs.scenario["Notes payable",] +
    225
rosengarten$bs.scenario["Total current liabilities",] <-
    rosengarten$bs.scenario["Total current liabilities",] +
    225
rosengarten$bs.scenario["Long-term debt",] <-
    rosengarten$bs.scenario["Long-term debt",] +
    340
rosengarten$bs.scenario["Total liabilities and owners' equity",] <-
    rosengarten$bs.scenario["Total current liabilities",] +
    rosengarten$bs.scenario["Long-term debt",] +
    rosengarten$bs.scenario["Total equity",]
rosengarten$bs.scenario.delta <-
    round(rosengarten$bs.scenario - rosengarten$bs[,1])
rosengarten$bs.scenario <- cbind(rosengarten$bs.scenario,
                                 rosengarten$bs.scenario.delta)
knitr::kable(pretty(rosengarten$bs.scenario),
     col.names=c("Item", "Amount ($)", "Change from current year"), align="lrr",
     caption="Rosengarten Corporation pro forma balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Current assets", 1, 4, bold=T, background="#f2f2f2") %>%
   pack_rows("Fixed assets", 5, 5, bold=T, background="#D3E4CD") %>%
   pack_rows("Current liabilities", 7, 9, bold=T, background="#FAD4D4") %>%
   pack_rows("Owners' equity", 11, 13, bold=T, background="#D9E4EC") %>%
   row_spec(5, extra_latex_after = "\\addlinespace") %>%
   row_spec(6, extra_latex_after = "\\addlinespace") %>%
   row_spec(13, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(4, 9, 13))
```

In this scenario, we used a combination of short- and long-term debt as the
plug. However, this is just one possible strategy; not necessarily the best one.

## External financing and growth

External financing and growth are related. The higher the rate of growth in
sales or assets, the greater the need for external financing.

### EFN and growth

We will use the following simplified income statement and balance sheet.

```{r external financing and growth; hoffman company, echo=F}
hoffman <- new.env()
hoffman$is <- matrix(c(500, 400, 100, 34, 66, 22, 44), ncol=1)
hoffman$is.accounts <- c("Sales", "Costs", "Taxable income", "Taxes (34%)",
                         "Net income", "Dividend",
                         "Additions to retained earnings")
rownames(hoffman$is) <- hoffman$is.accounts
knitr::kable(pretty(hoffman$is),
             caption="Hoffman Company income statement") %>%
    add_indent(c(6,7))

hoffman$bs <- data.frame(
  Category = c("Current assets", "Net fixed assets", "Total assets",
               "Total debt", "Owners' equity",
               "Total liabilities and owners' equity"),
  Amount = c(200, 300, 500, 250, 250, 500),
  pctSales = c(.4, .6, 1, NA, NA, NA)
)

knitr::kable(pretty(hoffman$bs),
     col.names=c("Item", "Amount ($)", "% of sales"), align="lrr",
     caption="Hoffman Corporation balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Assets", 1, 3, bold=T, background="#f2f2f2") %>%
   pack_rows("Liabilities and Owners' Equity", 4, 6, bold=T,
             background="#D3E4CD") %>%
   row_spec(3, extra_latex_after = "\\addlinespace") %>%
   add_indent(c(3,6))
```

Notice that short- and long-term liabilities have been combined into a single
_Total debt_ figure. We are assuming that none of the current liabilities vary
spontaneously with sales. This assumption isn't as restrictive as it sounds. If
any current liabilities (such as accounts payable) vary with sales, we can
assume that any such accounts have been netted out in current assets.

If Hoffman Company is forecasting next year's sales level at \$600, a \$100
increase, then the percentage increase in sales is \$100/500 = 20 percent.

Using the percentage of sales approach we create the pro forma financial
statements:

```{r external financing and growth; hoffman company pro forma, echo=F}
hoffman$is.proforma <- hoffman$is * 1.2
knitr::kable(pretty(hoffman$is),
             caption="Hoffman Company pro forma income statement")

hoffman$bs.proforma <- data.frame(
  Category = c(hoffman$bs$Category),
  #          Assets all vary with sales;    Total Debt unchanged
  Amount = c(hoffman$bs$Amount[1:3] * 1.2, hoffman$bs$Amount[4],
  #           Add addition to retained earnings to owners' equity
              hoffman$bs$Amount[5] + hoffman$is.proforma[7, 1],
  #           Do the same with total liabilities and owners' equity
              hoffman$bs$Amount[6] + hoffman$is.proforma[7, 1]),
  pctSales = hoffman$bs$pctSales
)
efn <- hoffman$bs.proforma[3,2] - hoffman$bs.proforma[6,2]
hoffman$bs.proforma <- rbind(hoffman$bs.proforma,
  c("External financing needed", efn, NA))

knitr::kable(pretty(hoffman$bs.proforma),
     col.names=c("Item", "Amount ($)", "% of sales"), align="lrr",
     caption="Hoffman Corporation pro forma balance sheet") %>%
   kable_styling(bootstrap_options="striped", full_width=F) %>%
   pack_rows("Assets", 1, 3, bold=T, background="#f2f2f2") %>%
   pack_rows("Liabilities and Owners' Equity", 4, 6, bold=T,
             background="#D3E4CD") %>%
   row_spec(c(3,6), extra_latex_after = "\\addlinespace") %>%
   add_indent(c(3,6))
```

At 20 percent growth, Hoffman will need \$100 in new assets. The projected
addition to retained earnings is \$52.8, so the external financing needed, EFN,
is \$100 - 52.8 = \$47.2.


Below shows EFN for several different growth rates. The projected addition to
retained earnings and the projected debt-equity ratio for each scenario are also
given. In determining the debt-equity ratios, we assumed that any needed funds
were borrowed, and we also assumed any surplus funds were used to pay off debt.
Thus, for the zero growth case the debt falls by \$44, from \$250 to \$206.

```{r external financing and growth; growth and projected EFN, echo=F}
efn <- new.env()
efn$sales.growth <- c(0, .05, .1, .15, .20, .25)
efn$assets.increase <- 500 * efn$sales.growth
efn$addition.retained.earnings <- 44 + (44 * efn$sales.growth)
efn$efn <- (500 + efn$assets.increase) - (500 + efn$addition.retained.earnings)
# We assume that all needed funds are borrowed
efn$de.ratio <- round((250 + efn$efn) /
                      (250 + efn$addition.retained.earnings),2)
efn$growth.table <- data.frame(
  Sales.Growth = efn$sales.growth,
  Assets.Increase = efn$assets.increase,
  Addition.Retained.Earnings = efn$addition.retained.earnings,
  External.Financing.Needed = efn$efn,
  DE.Ratio = efn$de.ratio
)
efn$gt.cols <- c("Projected Sales Growth", "Increase in Assets Required",
                 "Addition to Retained Earnings",
                 "External Financing Needed, EFN",
                 "Projected Debt-Equity Ratio")
knitr::kable(efn$growth.table, col.names=efn$gt.cols)
```

### Financial policy and growth

```{r external financing and growth; asset vs growth, echo=F}
plot(efn$sales.growth*100, efn$assets.increase, type="l",
     xlab="Projected growth in sales (%)",
     ylab="Asset needs and retained earnings ($)",
     col="lightblue",
     main="Growth and related financing needed for the Hoffman Company")
lines(efn$sales.growth*100, efn$addition.retained.earnings,
      col="#FF6666")
```

#### The internal growth rate

The maximum growth rate that can be achieved with no external financing of any
kind. In the figure above, the internal growth rate is the point where the
increase in assets required line (blue), and the projected addition to retained
earnings line (red) intersect. This is defined:

$$
\text{Internal growth rate} = \frac{\text{ROA} \times b}
  {1 - \text{ROA} \times b}\tag{3.24}
$$

ROA is the return on assets and $b$ is the plowback ratio.

For the Hoffman Company, net income was \$66 and total assets were \$500. ROA is
$66/500 = .132$ or 13.2\%. Of the \$66 net income, \$44 was retained, so the
plowback ratio, $b$, is $44/66 = 2/3$.

$$
\text{Internal growth rate} = \frac{.132 \times (2/3)}{1 - .132 \times (2/3)}
  = .0965
$$

Thus, the Hoffman Company can expand at a maximum rate of 9.65\% per year
without external financing.

#### The sustainable growth rate

The next growth rate is the maximum growth rate a firm can achieve with no
external _equity_ financing while it maintains a constant debt-equity ratio.

$$
\text{Sustainable growth rate} = \frac{\text{ROE} \times b}
  {1 - \text{ROE} \times b}\tag{3.25}
$$

This is identical to the internal growth rate except we use ROE instead of ROA.

For the Hoffman Company, net income was \$66 and total equity was \$250; ROE is
thus $66/250 =.264$ or 26.4\%. The plowback ratio, $b$ is still 2/3.

$$
\text{Sustainable growth rate} = \frac{.264 \times (2/3)}{1 = .264 \times (2/3)}
  = .2136
$$

Thus, the Hoffman Company can expand at a maximum rate of 21.36 percent per year
without external equity financing.

#### Determinants of growth

Remember that ROE can be decomposed into its various components using the DuPont
identity. These factors that are important in determining ROE are also important
factors for determining growth.

$$
\text{ROE} = \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier}
$$

The firm's ability to sustain growth depends explicitly on the following four
factors:

1. _Profit margin_: An increase in profit margin will increase the firm's
   ability to generate funds internally and thereby increase its sustainable
   growth.

2. _Dividend policy_: A decrease in the percentage of net income paid out as
   dividends will increase the retention ratio. This increases internally
   generated equity and thus increases sustainable growth.

3. _Financial policy_: An increase in the debt-equity ratio increases the firm's
   financial leverage. Because this makes additional debt financing available,
   it increases the sustainable growth rate.

4. _Total asset turnover_: An increase in the firm's total asset turnover
   increases the sales generated for each dollar in assets. This decreases the
   firm's need for new assets as sales growth and thereby increases the
   sustainable growth rate. Notice that increasing total asset turnover is the
   same thing as decreasing capital intensity (because they are recipricals).

Given values for all four of these, there is only one growth rate that can be
achieved.

__NOTE:__

> __If a firm does not wish to sell new equity and its profit margin, dividend
> policy, financial policy, and total asset turnover (or capital intensity) are
> all fixed, then there is only one possible growth rate.__

### A note about sustainable growth rate calculations

Very commonly, the sustainable growth rate is calculated using just the
numerator in our expression, ROE $\times b$. This causes some confusion. The
issue has to do with how ROE is computed. Recall that ROE is calculated as net
income divided by total equity. If total equity istaken from an ending balance
sheet, then our formula is the right one. However, if total equity is from the
beginning of the period, then the simpler formula is the correct one.

In principle, you'll get exactly the same sustainable growth rate regardless of
which way you calculate it (as long as you match up the ROE calculation with the
right formula). In reality, you may see some differences because of
accounting-related complications.

If you use the average of beginning and ending equity, you will need yet another
formula.

This also applies to the internal growth rate as well.

## Some caveats regarding financial planning models

Financial planning models do not always ask the right questions. A primary
reason is that they tend to rely on accounting relationships and not financial
relationships. In particular, the three basic elements of firm value tend to get
left out---namely, cash flow size, risk, and timing.

Because of this, financial planning models sometimes do not produce output that
gives the user many meaningful clues about what strategies will lead to
increases in value. Instead, they divert the user's attention to questions
concerning the association of, say, the debt-equity ratio and firm growth.

Financial planning is an iterative process. Plans are created, examined, and
modified over and over. The final plan will be a result negotiated between all
the different parities to the process. In fact, long-term financial planning in
most corporations relies on what might be called the Procrustes approach.
Upper-level management has a goal in mind, and it is up to the planning staff to
rework and ultimately deliver a feasible plan that meets that goal.

---

# Examples

## Example 3.1: Current events

Suppose a firm were to pay off some of its suppliers and short-term creditors.
What would happen to the current ratio? Suppose a firm buys some inventory. What
happens in this case? What happens if a firm sells some merchandise?

The first case is a trick question. What happens is that the current ratio moves
away from 1. If it is greater than 1, it will get bigger, but if it is less than
1, it will get smaller. To see this, suppose the firm has \$4 in current assets
and \$2 in current liabilities for a current ratio of 2. If we use \$1 in cash
to reduce current liabilities, the new current ratio is
$(\$4 - 1)/(\$2 - 1) = 3$. If we reverse the original situation to \$2 in
current assets and \$4 in current liabilities, the change will cause the current
ratio to fall to 1/3 from 1/2.

The second case is not quite as tricky. Nothing happens to the current ratio
because cash goes down while inventory goes up---total current assets are
unaffected.

In the third case, the current ratio would usually rise because inventory is
normally shown at cost and the sale would normally be at something greater than
cost (the difference is the markup). The increase in either cash or receivables
is therefore greater than the decrease in inventory. This increases current
assets, and the current ratio rises.

## Example 3.2: Payables turnover

Here is a variation on the receivables collection period. How long, on average,
does it take for Prufrock Corporation to _pay_ its bills? To answer, we need to
calculate the accounts payable turnover rate using cost of goods sold. We will
assume that Prufrock purchases everything on credit.

The cost of goods sold is \$1,344, and accounts payable are \$344. The turnover
is therefore \$1,344/\$344 = 3.9 times. So, payables turned over about every
365/3.9 = 94 days. On average, then, Prufrock takes 94 days to pay. As a
potential creditor, we might take note of this fact.

## Example 3.3: More turnover

Suppose you find that a particular company generates \$.40 in annual sales for
every dollar in total assets. How often does this company turn over its total
assets?

The total asset turnover here is .40 times per year. It takes 1/.40 = 2.5 years
to turn assets over completely.

## Example 3.4: Financial ratios

Consider the following 2010 data for Lowe's Companies and Home Depot (billions
except for price per share):

```{r example 3.4, echo=F}
ex3.4 <- new.env()

ex3.4$lowes <- c(48.8, 3.2, 2.0, 0.7, 1.6, 6.5, 33.7, 25, 1.3, 18.1)
ex3.4$home.depot <- c(68.0, 5.8, 3.3, 0.5, 1.6, 8.7, 40.1, 37, 1.6, 18.9)

ex3.4$rows <- c("Sales", "EBIT", "Net income", "Cash", "Depreciation",
                "Interest bearing debt", "Total assets", "Price per share",
                "Shares outstanding", "Shareholder equity")
ex3.4$col <- c("Lowe's Company, Inc.", "The Home Depot, Inc.")

ex3.4$tbl <- matrix(c(ex3.4$lowes, ex3.4$home.depot), ncol=2)
rownames(ex3.4$tbl) <- ex3.4$rows
colnames(ex3.4$tbl) <- ex3.4$col
knitr::kable(ex3.4$tbl, align="cc")

ex3.4$equity.multiplier <- round(ex3.4$tbl["Total assets",] /
    ex3.4$tbl["Shareholder equity",],1)
ex3.4$total.asset.turnover <- round(ex3.4$tbl["Sales",] /
    ex3.4$tbl["Total assets",],1)
ex3.4$profit.margin <- round(ex3.4$tbl["Net income",] /
    ex3.4$tbl["Sales",],3)
ex3.4$roe <- round(ex3.4$tbl["Net income",] /
    ex3.4$tbl["Shareholder equity",],3)
ex3.4$market.capitalization <- round(ex3.4$tbl["Price per share",] *
    ex3.4$tbl["Shares outstanding",],1)
ex3.4$enterprise.value <- round(ex3.4$market.capitalization +
    ex3.4$tbl["Interest bearing debt",] - ex3.4$tbl["Cash",],1)
ex3.4$eps <- round(ex3.4$tbl["Net income",] / 
    ex3.4$tbl["Shares outstanding",],2)
ex3.4$pe.multiple <- round(ex3.4$tbl["Price per share",] / ex3.4$eps,1)
ex3.4$ebitda <- round(ex3.4$tbl["EBIT",] + ex3.4$tbl["Depreciation",],1)
ex3.4$ev.multiple <- round(ex3.4$enterprise.value / ex3.4$ebitda,1)
```

__1.__ Determine the profit margin, ROE, market capitalization, enterprise
   value, PE multiple, and EV multiple for both Lowe's and Home Depot.

$$
\begin{aligned}
\ \ & & \text{Lowe's Company, Inc.} & & \text{The Home Depot, Inc.}\\\\
\text{Equity multiplier} & & `r ex3.4$tbl["Total assets", 1]`
  / `r ex3.4$tbl["Shareholder equity", 1]` = `r ex3.4$equity.multiplier[1]`
  & & `r ex3.4$tbl["Total assets", 2]`
  / `r ex3.4$tbl["Shareholder equity", 2]` = `r ex3.4$equity.multiplier[2]`\\
\text{Total asset turnover} & & `r ex3.4$tbl["Sales",1]`
  / `r ex3.4$tbl["Total assets",1]` = `r ex3.4$total.asset.turnover[1]`
  & & `r ex3.4$tbl["Sales",2]`
  / `r ex3.4$tbl["Total assets",2]` = `r ex3.4$total.asset.turnover[2]`\\
\text{Profit margin} & & `r ex3.4$tbl["Net income",1]`
  / `r ex3.4$tbl["Sales",1]` = `r ex3.4$profit.margin[1]*100`\%
  & & `r ex3.4$tbl["Net income",2]`
  / `r ex3.4$tbl["Sales",2]` = `r ex3.4$profit.margin[2]*100`\%\\
\text{ROE} & & `r ex3.4$tbl["Net income",1]`
  /`r ex3.4$tbl["Shareholder equity",1]` = `r ex3.4$roe[1]*100`\%
  & & `r ex3.4$tbl["Net income",2]`
  /`r ex3.4$tbl["Shareholder equity",2]` = `r ex3.4$roe[2]*100`\%\\
\text{Market capitalization} & & `r ex3.4$tbl["Price per share",1]`
  \times `r ex3.4$tbl["Shares outstanding",1]` = `r ex3.4$market.capitalization[1]`
  & & `r ex3.4$tbl["Price per share",2]`
  \times `r ex3.4$tbl["Shares outstanding",2]` = `r ex3.4$market.capitalization[2]`\\
\text{Enterprise value} & & `r ex3.4$market.capitalization[1]`
  + `r ex3.4$tbl["Interest bearing debt",1]` - `r ex3.4$tbl["Cash",1]`
  = `r ex3.4$enterprise.value[1]`
  & & `r ex3.4$market.capitalization[2]`
  + `r ex3.4$tbl["Interest bearing debt",2]` - `r ex3.4$tbl["Cash",2]`
  = `r ex3.4$enterprise.value[2]`\\
\text{PE multiple} & & `r ex3.4$tbl["Price per share",1]`
  / `r ex3.4$eps[1]` = `r ex3.4$pe.multiple[1]`
  & & `r ex3.4$tbl["Price per share",2]`
  / `r ex3.4$eps[2]` = `r ex3.4$pe.multiple[2]`\\
\text{EBITDA} & & `r ex3.4$tbl["EBIT",1]` + `r ex3.4$tbl["Depreciation",1]`
  = `r ex3.4$ebitda[1]`
  & & `r ex3.4$tbl["EBIT",2]` + `r ex3.4$tbl["Depreciation",2]`
  = `r ex3.4$ebitda[2]`\\
\text{EV multiple} & & `r ex3.4$enterprise.value[1]` / `r ex3.4$ebitda[1]`
  = `r ex3.4$ev.multiple[1]`
  & & `r ex3.4$enterprise.value[2]` / `r ex3.4$ebitda[2]`
  = `r ex3.4$ev.multiple[2]`
\end{aligned}
$$

__2.__ How would you describe these two companies from a financial point of
view? Overall, they are similarly situated. In 2010, Home Depot had a higher ROE
(partially because of a higher total asset turnover and a higher profit margin),
but Lowe's had slightly lower PE and EV multiples. Both companies' PE multiples
were somewhat above the general market, indicating possible future growth
prospects.

## Example 3.5: Sustainable growth

Suppose Hoffman grows at exactly the sustainable growth rate of 21.36 percent.
What will the pro forma statements look like?

At a 21.36 percent growth rate, sales will rise from \$500 to \$606.8. The pro
forma income statement will look like this:

```{r example 3.5; income statement, echo=F}
ex3.5 <- new.env()
ex3.5$is.proforma <- matrix(c(606.8, 485.4, 121.4, 41.3, 80.1, 26.7, 53.4),
                            ncol=1)
ex3.5$is.rows <- c("Sales (Projected)", "Costs (80% of sales)",
                   "Taxable income", "Taxes (34%)", "Net income",
                   "Dividends", "Addition to retained earnings")
rownames(ex3.5$is.proforma) <- ex3.5$is.rows
knitr::kable(ex3.5$is.proforma,
    caption="Hoffman Company pro forma income statement") %>%
  add_indent(c(6,7))
```

We construct the balance sheet just as we did before. Notice, in this case that
owners' equity will rise from \$250 to \$303.4 because the addition to retained
earnings is \$53.4.

```{r example 3.5; balance sheet, echo=F}
ex3.5$bs.cols <- c("$", "% of sales")
ex3.5$bs.assets <- matrix(c(242.7, 364.1, 606.8, .4, .6, 1), ncol=2)
ex3.5$bs.assets.rows <- c("Current assets", "Net fixed assets", "Total assets")
rownames(ex3.5$bs.assets) <- ex3.5$bs.assets.rows
colnames(ex3.5$bs.assets) <- ex3.5$bs.cols
ex3.5$bs.lande <- matrix(c(250.0, 303.4, 553.4, 53.4, NA, NA, NA, NA), ncol=2)
ex3.5$bs.lande.rows <- c("Total debt", "Owners' equity",
  "Total liabilities and owners' equity", "External financing needed")
rownames(ex3.5$bs.lande) <- ex3.5$bs.lande.rows
colnames(ex3.5$bs.lande) <- ex3.5$bs.cols
ex3.5$bs.display <- list(ex3.5$bs.assets, ex3.5$bs.lande)
knitr::kable(ex3.5$bs.display,
    caption="Hoffman Company pro forma balance sheet", align="cc") %>%
  add_indent(c(3)) %>%
  kable_styling(latex_options = "HOLD_position")
```

As illustrated, EFN is \$53.4. If Hoffman borrows this amount, then total debt
will rise to \$303.4 and the debt-equity ratio will be exactly 1.0, which
verifies our earlier calculation. At any other growth rate, something would
have to change.

## Example 3.6: Profit margins and sustainable growth

The Sandar Co. has a debt-equity ratio of .5, a profit margin of 3 percent, a
dividend payout ratio of 40 percent, and a capital intensity ratio of 1. What is
its sustainable growth rate? If Sandar desired a 10 percent sustainable growth
rate and planned to achieve this goal by improving profit margins, what would
you think?

ROE is .03 $\times$ 1 $\times$ 1.5 = 4.5 percent. The retention ratio is
1 - .40 = .60. Sustainable growth is thus
.045(.60)/[1 - .045(.60)] = 2.77 percent.

For the company to achieve a 10 percent growth rate, the profit margin will have
to rise. To see this, assume that sustainable growth is equal to 10 percent and
then solve for profit margin, PM:

$$
\begin{aligned}
.10 = \text{PM}(1.5)(.6)/[1 - \text{PM}(1.5)(.6)]\\
\text{PM} = .1/.99 = 10.1\%
\end{aligned}
$$

For the plan to succeed, the necessary increase in profit margin is
substaintial, from 3 percent to about 10 percent. This may not be feasible.

---

# Chapter summary

This chapter focuses on working with information contained in financial
statements Specifically, we studied standardized financial statements, ratio
analysis, and long-term financial planning.

1. We explained that differences in firm size make it difficult to compare
   financial statements, and we discussed how to form common-size statements to
   make comparisons easier and more meaningful.

2. Evaluating ratios of accounting numbers is another way of comparing financial
   statment information. We defined a number of the most commonly used ratios,
   and we discussed the famous DuPont identity.

3. We showed how pro forma financial statements can be generated and used to
   plan for future financing needs.

After you have studied this chapter, we hope that you have some perspective on
the uses and abuses of financial statement information. You should also find
that your vocabulary of business and financial terms have grown substantially.

