--- 
title: "Financial Statements Analysis and Financial Models"
author: "Ross, Westerfield, Jaffe"
header-includes:
  - \usepackage{booktabs}
toc: true 
output: pdf_document
pdf_engine: xelatex
---

```{r setup, echo=FALSE, includes=FALSE, message=FALSE}
require(kableExtra)
require(dplyr)
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
pretty <- function(x) format(x, big.mark=",")
```

# 1. Financial ratio analysis

A financial ratio by itself tells us little about a company because financial
ratios vary a great deal across industries. There are two basic methods for
analyzing financial ratios for a company: Time trend analysis and peer group
analysis. In time trend analysis, you find the ratios for the company over some
period, say five years, and examine how each ratio has changed over this period.
In peer group analysis, you compare a company's financial ratios to those of its
peers. Why might each of these analysis methods be useful? What does each tell
you about the company's financial health?

---

### Time trend analysis

Time trend analysis is useful because it provides insights into a company's
performance over time. By examining financial ratios across a historical period
(e.g., five years), you can identify trends, such as:

* __Improvement or deterioration__: For instance, a steadily improving profit
  margin suggests better cost control or increasing revenue efficiency, whereas
  a declining ratio might signal inefficiencies or problems.

* __Seasonal patterns__: Some businesses are cyclical, and observing trends over
  time can help understand seasonal impacts.

* __Long-term strategy effects__: Changes in ratios might reflect the outcomes
  of strategic decisions, such as entering new markets, adjusting pricing
  strategies, or adopting cost-cutting measures.

What time trend analysis reveals about the financial health of the firm is:

* __Sustainability__: Whether improvements or declines are part of a sustainable
  trend.

* __Early warning signs__: Sudden deviations in ratios (e.g., liquidity or
  leverage ratios) could indicate upcoming financial challenges.

* __Growth trajectory__: Trends in profitability and efficiency ratios can show
  whether the company is scaling effectively.

### Peer group analysis

Peer group analysis involves comparing a company's financial ratios with those
of similar firms in the same industry. It provides context for understanding
performance by showing whether a company is operating competitively.

Peer group analysis will reveal:

1. A company's performance with respect to its peers.
2. The company's market position with respect to its peers.
3. Shows a firm's competitive advantage and weaknesses with its peers.

Both methods are complementary. Time trend analysis provides insights into how
the company is evolving internally. Peer group analysis situates the company's
performance within the competitive landscape.

Using both methods together offers a holistic view of a company's financial
health, allowing for better-informed decisions by combining internal trends with
external comparisons.

---

# 2. Industry-specific ratios

So-called "same-store sales" are a very important measure for companies as
diverse as McDonald's and Sears. As the name suggests, examining same-store
sales means comparing revenues from the same stores or restaurants at two
different points in time. Why might companies focus on same-store sales rather
than total sales?

---

Companies often focus on __same-store sales__ rather than total sales because
same-store sales provide a more accurate and meaningful measure of a company's
__organic growth and performance__ over time, excluding the effects of new store
openings or closures. Same-store sales can also give the company insight into
the operating performance of each of its locations.

### Measures organic growth

__Same-store sales__ focus solely on stores that have been operational for a
certain period (usually more than a year), eliminating the impact of new store
openings, expansions, or closures. This metric highlights whether existing
stores are attracting more customers, selling more products, or increasing
revenues through better pricing strategies. For example: A 10% increase in total
sales could be misleading if it's driven entirely by new stores, while
same-store sales growth indicates improvements in the performance of existing
locations.

### Excludes non-comparable factors

Total sales can be influenced by non-operational factors such as: new store
openings or acquisitions, which can inflate growth artificially; store closures,
which might mask declines in performance. Same-store sales provide a
"like-for-like" comparison, removing these distortions.

### Indicates customer loyalty and market trends

Growth in same-store sales often reflects higher customer retention, increased
traffic, or greater spending per customer at established locations. Declines in
same-store sales could signal market saturation, poor customer satisfaction, or
increased competition.

### Helps evaluate management effectiveness

By focusing on same-store sales, management can assess the effectiveness of
marketing campaigns, operational improvements, and product offerings within the
existing store base. It's a key indicator of how well the company is utilizing
its current resources to drive growth.

---

# 3. Sales forecast

Why do you think most long-term financial planning begins with sales forecasts?
Put differently, why are future sales the key input?

---

Most long-term financial planning begins with __sales forecasts__ because sales
are the primary driver of a company's operations and financial performance.
Future sales serve as the foundation for estimating many other critical
financial elements, making them the key input for planning.

Sales drive revenues and cash flows. Sales are the top line of the income
statement, and all other financial metrics---such as costs, profits, and cash
flows---stem from this figure. Future sales projections determine expected
revenue, which is critical for estimating future earnings and cash availability.

Companies use sales forecasts to plan for necessary investments in assets, such
as production capacity, equipment, or inventory. For instance: A forecast of
rising sales might prompt a company to expand facilities or hire more staff.
Declining sales forecasts could signal the need to reduce investment and
preserve cash.

Sales growth often requires financing to support operations, including working
capital, inventory, and capital expenditures. A robust sales forecast helps
determine how much financing the company will need and when, allowing for better
capital planning and debt management.

Sales forecasts align operational strategies, such as marketing campaigns,
product launches, or market expansion plans, with realistic revenue
expectations. For instance, if projected sales growth is aggressive, the company
might allocate more resources to marketing or innovation.

Sales forecasts are a key metric for communicating future performance
expectations to stakeholders, including investors, lenders, and employees. They
form the basis for convincing stakeholders of the company's growth potential and
ability to meet obligations.

---

# 4. Sustainable growth

In the chapter, we used Rosengarten Corporation to demonstrate how to calculate
EFN. The ROE for Rosengarten is about 7.3 percent, and the plowback ratio is
about 67 percent. If you calculate the sustainable growth rate for Rosengarten,
you will find it is only 5.14 percent. In our calculation for EFN, we used a
growth rate of 25 percent. Is this possible? (_Hint_: Yes, How?)

---

Yes, it is possible for a company like Rosengarten Corporation to have a
__growth rate__ significantly higher than its __sustainable growth rate__ (SGR),
such as 25% compared to 5.14%. However, achieving such a high growth rate
requires the company to __seek external financing__.

### Sustainable growth rate defined

The __sustainable growth rate (SGR)__ is the maximum growth rate a company can
achieve using only internally generated funds (retained earnings) without
needing external financing. To calculate SGR:

$$
\mathbf{SGR} = \text{ROE} \times \text{Plowback Ratio}
$$

For Rosengarten:

$$
\mathbf{SGR} = 7.3\% \times 67\% = 5.14\%
$$

### Growth beyond SGR

If the company wants to grow faster than its SGR (e.g., at 25%), it must rely on
__external financing__ to fund the additional growth. This external financing
could come from:

* __Debt__: Borrowing funds to invest in growth opportunities.

* __Equity__: Issuing new shares to raise captial.

* __Hybrid instruments__: Convertible debt or other forms of capital raising.

For example, to support a 25% growth rate: Rosengarten might need to
significantly increase its debt levels (which increases the financial leverage).
Alternatively, it could issue new equity, although this might dilute existing
shareholders' ownership.

### Why external financing is needed

At a growth rate of 25%, the company's __assets__ will need to grow to support
the higher revenue. Since the company's retained earnings (from the 67% plowback
ratio) can only fund growth up to the SGR, the company must cover the gap with
external financing.

* __Required Asset Growth__: To support 25% revenue growth, assets typically
  need to grow by a similar percentage.

* __EFN (External Financing Needed)__: This represents the funds required to
  bridge the gap between the internally generated funds and the total funds
  needed for the desired growth.

### Practical implications

* __Financial Risks__: Relying on external financing increases financial risks,
  especially if the company takes on excessive debt. High leverage can strain
  cash flows due to interest payments and principal repayments.

* __Market Perception__: Issuing equity might dilute ownership and potentially
  impact stock prices negatively.

* __Strategic Trade-offs__: The compaany must balance the benefits of high
  growth with the risks of relying on external funding.

### Conclusion

While Rosengarten sustainable growth rate is only 5.14%, it can achieve a 25%
growth rate by raising external funds. This approach is feasible but requires
careful management of financial risks and strategic decisions about the type and
amount of external financing to pursue.

---

# 5. EFN and growth rate

Broslofski Co. maintains a positive retention ratio and keeps its debt-equity
ratio constant every year. When sales grow by 20 percent, the firm has a
negative projected EFN. What does this tell you about the firm's sustainable
growth rate? Do you know, with certainty, if the internal growth rate is greater
than or less than 20 percent? Why? WHat happens to the projected EFN if the
retention ratio is increased? What if the retention ratio is decreased? What if
the retention ratio is zero?

---

This situation with Broslofski Co. provides valuable insights into its
__sustainable growth rate (SGR)__, __internal growth rate (IGR)__, and how these
relate to projected external financing needs (EFN). Let's analyze each part:

### What does a negative projected EFN tell you about the firm's sustainable growth rate?

A __negative EFN__ means that the company is generating more internal funds than
it needs to support its growth at the projected 20% sales increase. This
suggests that the firm's sustainable growth rate (SGR) is __grater than 20%__
because it can fund this growth internally while maintaining its desired capital
structure.

### Is the internal growth rate greater than or less than 20 percent?

The __internal growth rate (IGR)__ is the maximum growth rate the firm can
achieve using only retained earnings, without borrowing additional funds or
issuing equity. Since the company maintains a constant debt-equity ratio and has
a __negative EFN__, it implies that the firm is not reliant on external
financing at the 20% growth rate. Therefore, the __IGR must be less than 20%__
because the firm's ability to grow solely using internal funds (IGR) is
inherently lower than its ability to grow with the help of leverage (SGR).

### What happens to the projected EFN if the retention ratio is increased?

The increased retention ratio means the firm reinvests a larger portion of its
earnings back into the business, reducing dividends. This increases the amount
of internally generated funds available for growth, reducing EFN further. If EFN
was negative, it would become more negative, meaning the firm generates even
more surplus funds than needed fo rht eprojected growth.

### What happens to the projected EFN if the retention ratio is decreased?

The decreased retntion ratio means the firm retains less of its earnings and
distributes more as dividends. This reduces the amount of internal funds
available for reinvestment, potentially increasing EFN. If the EFN was negative,
it might move closer to zero, indicating reduced surplus funds. If EFN becomes
positive, the firm would need external financing to support its growth.

### What if the retention ratio is zero?

A zero retention ratio means the firm does not reinvest any of its earnings; all
profits are distributed as dividends. In this case: the firm is entirely reliant
on external financing (debt or equity) to fund growth; the EFN will likely
become positive, as there are no internal funds to support growth; and the
firm's IGR and SGR will also drop to zero, as both require retained earnings for
reinvestment.

### Summary of the relationships

1. __SGR > 20%__: The firm's ability to grow sustainably exceeds 20%, indicated
   by the negative EFN.

2. __IGR < 20%__: The IGR is less than 20% because it represents growth
   achievable without any external leverage.

3. __Retention Ratio impact__: an increase means EFN becomes more negative (more
   surplus funds); a decrease means EFN moves closer to zero or becomes positive
   (less surplus funds); and if zero, EFN becomes positive and no internally
   generated funds are available to support growth.

This analysis highlights how retention policy and internal vs. external
financing decisions influence a firm's ability to sustain and project growth.

---

# 6. Common-size financials

One tool of financial analysis is common-size financial statements. Why do you
think common-size income statements and balance sheets are used? Note that the
accounting statement of cash flows is not converted into a common-size
statement. Why do you think this is?

---

### Simplifies comparison across companies

Because each line item is expressed as a percentage of total sales for income
statements and percentage of total assets for balance sheets. This helps to
compare companies of different size, industry, or geographic location because
they eliminate the impact of scale.

### Highlights financial trends

By converting financial statements into percentages, common-size analysis helps
identify trends over time, such as changes in cost structures, profitability, or
asset composition. For instance: a rising cost of goods sold (COGS) as a
percentage of sales might indicate worsening cost efficiency; or Increasing
current liabilities as a percentage of total assets could signal growing
liquidity risks.

### Aids in benchmarking

Common-size statements facilitate comparisons against industry averages or peer
companies, helping analysts assess a company's competitive position. For
example, a company with a lower operating expense ratio than peers might have
superior cost control.

### The cash flow statement is not common-sized because

1. __Lack of logical Base__ The cash flow statement tracks the sources and uses
   of cash across three categories: operating, investing, and financing
   activities. There is no single, consistent "base" to use for expressing cash
   flow items as percentages. Unlike sales in the income statement or total
   assets in the balance sheet, the cash flow statement doesn't have a
   comparable anchor figure.

2. __Focus on absolute values__ The cash flow statement's primary purpose is to
   show actual cash inflows and outflows, which are absolute values.
   Understanding whether the company generates enough cash to cover its
   obligations or fund growth is more important than relative percentages.

3.__Varied nature of cash flows__ The cash flow statement includes diverse line
items that don't naturally align with a common denominator: operating cash flow
links to profitability and working capital; investing cash flow reflects capital
expenditures and assets purchases or sales; financing cash flow deals with
borrowing, repaying debt, and equity transactions. These activities are not
proportional to a single metric like sales or total assets.

### Conclusion

Common-size financial statements are invaluable for standardizing financial data
to facilitate analysis and comparison, particularly for the income statement and
balance sheet. However, the cash flow statement is best analyzed using absolute
figures, as it focuses on cash management rather than proportional
relationships.

---

# 7. Asset utilization and EFN

One of the implicit assumptions we made in calculating the external funds needed
was that the company was operating at full capacity. If the company is operating
at less than full capacity, how will this affect the external funds needed?

---

If a company is operating at less than full capacity, it has the potential to
increase production without immediately investing in additional fixed assets.
This can significantly affect the calculation of external funds needed (EFN):

### Reduced need for fixed asset investment

At less than full capacity, the company can increase sales and production
without requiring proportional increases in fixed assets (e.g., machinery,
equipment, facilities). Since EFN calculations typically assume proportional
growth in all asset categories, this assumption would overestimate the need for
fixed asset investment in such a scenario. This would result in lower EFN
because the company doesn't need to spend as much on capital expenditures.

### Implications for asset turnover

Operating below capacity implies that existing fixed assets are underutilized.
As production ramps up to meet increased sales, the fixed asset turnover ratio
improves (sales per unit of fixed assets), further reducing the need for
additional investments.

### Working capital requirements still increase

Even if fixed asset investments are unnecessary, growth in sales will still
require additional current assets (e.g., inventory and accounts receivable) and
might increase current liabilities (e.g., accounts payable). However, these
changes are typically smaller than those involving fixed assets.

### Adjusted EFN

If the company adjusts for underutilized capacity, then EFN will be calculated
based only on increases in current assets and current liabilities, excluding or
reducing the need for new fixed assets. This will result in a lower EFN compared
to the assumption of full capacity utilization.

### Example

Assume sales grow by 20%, and the company is operating at 80% capacity. The
company can handle the 20% growth without additional fixed asset investment
because it only needs to utilize the remaining 20% capacity. EFN calculations
based on proportional growth would have overestimated the required increase in
fixed assets, inflating EFN unnecessarily.

### Key takeaways

When operating below full capacity, EFN decreases because fixed asset
investments are reduced or unnecessary. Thus, the calculation must be adjusted
to reflect the reality of underutilized capacity. However, as the company
approaches full capacity, future growth will eventually require investment in
fixed assets, at which point EFN will increase.

Understanding capacity utilization ensures more accurate financial planning and
avoids overstating the need for external funding.

---

# 8. Comparing ROE and ROA

Both ROA and ROE measure profitability. Which one is more useful for comparing
two companies? Why?

---

Both return on assets (ROA) and return on equity (ROE) measure profitability,
but which one is more useful for comparing two companies depends on the context
and what you're trying to analyze.

### ROA (return on assets)

Measures how efficiently a company uses its total assets to generate profit. ROA
also accounts for both equity and debt financing since assets are funded by
both.

$$
\mathbf{ROA} = \frac{\text{net income}}{\text{total assets}}
$$

You would use ROA to compare two companies when:

#### Capital structure differs

ROA provides a more comparable measure of operational efficiency because it
ignores the effects of leverage. For example, if one company is heavily
debt-financed and the other relies mostly on equity, ROA gives a better sense of
how well each uses its total assets to generate profits.

#### Assessing asset efficiency

ROA evaluates how effectively the companies deploy their assets, regardless of
how they are financed.

### ROE (return on equity)

Focuses on the return generated on the shareholders' investment and reflects the
impact of leverage (debt) on profitability.

$$
\mathbf{ROE} = \frac{\text{net income}}{\text{shareholder's equity}}
$$

You would use ROE to compare two companies when:

#### Capital structure is similar

If two companies have similar levels of debt, ROE is more relevant because it
focuses on returns to shareholders.

#### Evaluating shareholder returns

ROE directly shows how much profit shareholders earn per dollar of equity
invested, which is key for equity investors.

### Key considerations

#### Impact of leverage

High leverage increases ROE because debt financing boosts returns on equity (if
the cost of debt is lower than the return on assets). However, this can distort
comparisons between companies with different debt levels. ROA is unaffected by
leverage, making it a more neutral measure.

#### Industry differences

In asset-heavy industries (e.g., manufacturing), ROA is particularly relevant
for assessing operational efficiency. In industries where leverage is common
(e.g., banking), ROE might be might indicative of performance.

### Conclusion

#### ROA

is more useful for comparing companies with different capital structures because
it measures asset efficiency independently of financing decisions.

#### ROE

is more relevant when comparing companies with similar debt levels or when
focusing on returns from a shoareholder's perspective.

Ultimately, understanding both metrics provides a more comprehensive view of
profitability, but the choice depends on what aspect of performance you wish to
compare.

---

# 9. Ratio analysis

Consider the ratio EBITD/Assets. What does this ratio tell us? Why might it be
more useful than ROA in comparing two companies?

---

The ratio EBITD/Assets (Earnings Before Interest, Taxes, and Deprecation divided
by Toatl Assets) provides insight into a company's ability to generate operating
cash flow relative to its total assets.

### EBITD/Assets viewpoint

#### Operational efficiency

It measures how efficiently a company uses its assets to generate operating cash
flow, excluding non-cash charges (deprecation and amortization) and before the
effects of financing and taxes.

#### Cash flow focus

By using EBITD instead of net income, the ratio focuses on cash-based
profitability rather than accounting profitability, which can be affected by
non-cash expenses and financing decisions.

#### Asset Utilization

Like ROA, this ratio indicates how effectively a company's total assets are
utilized, but with an emphasis on cash-generating ability.

### EBITD/Assets may be more useful in the following scenarios

#### Removes the effect of financing decisions

ROA uses net income in the numerator, which is influenced by interest expenses
(financing structure), while EBITD/Assets excludes interest, making it more
neutral for comparisons between companies with different levels of debt.

#### Excludes non-cash accounting effects

Deprecation and amortization are non-cash expenses that vary based on accounting
methods and asset age. EBITD excludes these charges, providing a better measure
of operational cash flow and enabling a more apples-to-apples comparison of
companies with different asset bases or deprecation policies.

#### Focuses on oprational performance

EBITD reflects the compnay's core operational performance before taxes and
financing costs. This makes it particularly useful for comparing companies in
capital-intensive industries where deprecation and amortization can
significantly distort net income.

#### Better for cross-industry or cross-region comparisons

ROA can be affected by tax rates or regional accounting practices, which vary
across industries or countries.  EBITD/Assets neutralizes these differences,
focusing on operational cash flow generation relative to assets.

### Example comparison

Company A has high leverage, uses debt financing extensively which results in
high interest expenses. Company B has low leverage, primarily equit-financed,
with minimal interest expenses. EBITD/Assets allows for a direct comparison of
operational efficiency without being skewed by financing differences, whereas
ROA might make Company B appear more profitable soley due to its financing
structure.

### Situations where EBITD/Assets excels

#### Capital-intensive industries

In industries like manufacturing or utilities, EBITD/Assets highlights
operational efficiency despite large deprecation charges.

#### High-leverage companies

Useful for comparing companies with varying levels of debt, as it excludes
interest effects.

#### Startups or Growth Companies

Young companies with large upfron capital investments but limited net income can
still show strong EBITD/Assets.

## Conclusion

EBITD/Assets is particularly useful when comparing companies with different
financing structures, deprecation policies, or tax environments. It provides a
clearer view of a company's ability to generate operating cash flow from its
assets, making more neutral and informative in specific contexts than ROA.

---

# 10. Return on investment

A ratio that is becoming more widely used is return on investment. Return on
investment is calculated as net income divided by long-term liabilities plus
equity. What do you think return on investment is intended to measure? What is
the relationship between return on investment and return on assets?

---

Return on investment (ROI) is calculated as

$$
\mathbf{ROI} = \frac{\text{net income}}{\text{long-term liabilitieis} + \text{equity}}
$$

This ratio is intended to measure the profitability of a company relative to its
total capital base, including bot debt (long-term liabilities) and equity.

### ROI measures

#### Profitablitility of total capital

ROI indicates how effectively a company is generating net income from the
resources provided by both creditors (long-term debt) and owners (equity). It
evaluates the return on all long-term financing rather than just equity, making
it a broader profitability measure.

#### Financial leverage impact

Since long-term liabilities are included in the denominator, ROI accounts for
the impact of leverage on profitability, though it does not distinguish between
returns to debt and equity holders.

#### Capital efficiency

A higher ROI means the company is better at converting its long-term financing
into net income.

### Relationship between ROI and ROA

Return on assets (ROA) is calculated as:

$$
\mathbf{ROA} = \frac{\text{net income}}{\text{total assets}}
$$

ROA uses total assets as the base, capturing the entire resource base employed
by the company. ROI uses long-term liabilities + equity, which represents the
company's permanent capital structure (ignoring current liabilities and
short-term financing). ROI can be thought of as a subset of ROA, focusing
specifically on returns from capital funded through long-term financing. For a
compnay with minimal current liabilities, ROI and ROA will be similar. However,
if the company has significant current liabilities, ROI will typically be higher
than ROA because the denominator (long-term liabilities + equity) is smaller
than total assets.

---

Use the following information to answer the next five questions: A small
business called the Grandmother Calendar Company began selling personalized
photo calendar kits. The kits were a hit, and sales soon sharply exceeded
forecasts. The rush of orders created a huge backlog, so the company leased more
space and expanded capacity, but it still could not keep up with demand.
Equipment failed from overuse and quality suffered. Working capital was drained
to expand production, and, at the same time, payments from customers were often
delayed until the product was shipped. Unable to deliver on orders, the company
became so strapped for cash that employee paychecks began to bounce. Finally,
out of cash, the company ceased operations entirely three years later.

---

# 11. Product sales

Do you think the company would have suffered the same fate if its product had
been less popular? Why or Why not?

---

While high demand is generally positive, in this case, the company's inability
to manage the operational and financial challenges of rapid growth led to its
failure. If its product had been less popular, the company might have grown at a
manageable pace, avoiding the operational and financial crises that ultimately
led to its closure.

---

# 12. Cash flow

The Grandmother Calendar Company clearly had a cash flow problem. In the context
of the cash flow analysis we developed in Chapter 2, what was the impact of
customers not paying until orders were shipped?

---

Customers delaying payments until orders were shipped created a significant cash
flow mismatch for the Grandmother Calendar Company. This issue compounded the
financial stress caused by high demand and operational challenges, ultimately
leading to a liquidity crisis. Managing cash flow by aligning inflows and
outflows is crucial for sustaining operations, particularly for growing
businesses like this one.

---

# 13. Corporate borrowing

If the firm was so successful at selling, why wouldn't a bank or some other
lender step in and provide it with the cash it needed to continue?

---

Despite strong sales, the Grandmother Calendar Company faced a combination of
operational inefficiencies, cash flow challenges, and poor financial management,
which likely made it appear too risky for banks or lenders. To secure funding,
the company would have needed to deomonstrate not only strong sales but also a
clear plan for addressing its liquidity issues, stabilizing operations, and
ensuring repayment ability. Without such assurances, successful sales alone
would not have been enough to secure additional financing.

---

# 14. Cash flow

Which was the biggest culprit here: Too  many orders, too little cash, or too
little production capacity?

---

While too many orders and too little production capacity were significant
challenges, too little cash was the biggest culprit. With better cash flow
management (e.g., requiring deposits, better aligning inflows and outflows, or
securing short-term financing), the company could have scaled its production
capacity and fulfilled orders. Ultimately, the cash shortage turned what could
have been a success story into a failure.

---

# 15. Cash flow

What are some actions a small company like The Grandmother Calendar Company can
take (besides expansion of capacity) if it finds itself in a situation in which
growth in sales outstrips production?

---

1. Manage demand strategically: adjust pricing, limit order intake, introduce
   pre-orders.

2. Outsource or partner: contract manufacturing or collaborate with competitors.

3. Optimize operations: improve production efficiency or focus on core products.

4. Enhance cash flow management: request customer deposits, offer discounts for
   early payments, negotiate supplier terms.

5. Prioritize customer satisfaction: communicate with customers, offer premium
   options.

6. Leverage technology: adopt inventory and order management software, forecast
   demand.

7. Use short-term financing: bridge loans or credit lines, factoring accounts
   receivable.

8. Explore alternative fulfillment models: Licensing or franchising, digital
   products or services.

---

# Questions and Problems

# 1. DuPont identity

If Jares, Inc., has an equity multiplier of 1.55, total asset turnover of 1.75,
and a profit margin of 4.3 percent, what is its ROE?

---

```{r problem 1, echo=F}
p1 <- new.env()
p1$equity.multiplier <- 1.55
p1$total.asset.turnover <- 1.75
p1$profit.margin <- .043
p1$roe <- p1$profit.margin * p1$total.asset.turnover * p1$equity.multiplier
```

$$
\begin{aligned}
\mathbf{ROE} &= \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier} \\
  &= `r p1$profit.margin` \times `r p1$total.asset.turnover`
  \times `r p1$equity.multiplier`\\
  &= `r round(p1$roe, 4)`
\end{aligned}
$$

---

# 2. Equity multiplier and return on equity

Nuber Company has a debt-equity ratio of .80. Return on assets is 9.7 percent,
and total equity is \$735,000. What is the equity multiplier? Return on equity?
Net income?

```{r problem 2, echo=F}
p2 <- new.env()
p2$de.ratio <- .8
p2$roa <- .097
p2$total.equity <- 735000
p2$equity.multiplier <- 1 + p2$de.ratio
p2$roe <- p2$roa * p2$equity.multiplier
p2$net.income <- p2$roe * p2$total.equity
```

$$
\begin{aligned}
\mathbf{Equity\ multiplier} &= 1 + \text{debt-equity ratio}\\
  &= 1 + `r p2$de.ratio`\\
  &= `r p2$equity.multiplier`\\\\
\mathbf{ROE} &= \text{ROA} \times \text{equity multiplier}\\
  &= `r p2$roa` \times `r p2$equity.multiplier`\\
  &= `r p2$roe`\\\\
\mathbf{Net\ income} &= \text{ROE} \times \text{total equity}\\
  &= `r p2$roe` \times `r p2$total.equity`\\
  &= `r pretty(p2$net.income)`\\\\
\end{aligned}
$$

---

# 3. Using the DuPont Identity

Y3K, Inc., has sales of \$2,700, total assets of \$1,310, and a debt-equity ratio
of 1.20. If its return on equity is 15 percent, what is its net income?

---

```{r problem 3, echo=F} 
p3 <- new.env()
p3$assets <- 1310
p3$de.ratio <- 1.2
p3$roe <- .15
p3$net.income <- (p3$roe * p3$assets)/(1 + p3$de.ratio)
```

Using the DuPont Identity we know:

$$
\text{ROE} = \frac{\text{net income}}{\text{assets}}
  \times (1 + \text{debt-equity ratio})
$$

Therefore, we can find `net income`:

$$
\begin{aligned}
\text{net income} &= \frac{\text{ROE} \times \text{assets}}
  {(1 + \text{debt-equity ratio})}\\
  &= \frac{`r p3$roe` \times `r pretty(p3$assets)`}
  {(1 + `r p3$de.ratio`)}\\
  &= `r round(p3$net.income, 2)`
\end{aligned}
$$

---

# 4. EFN

The most recent financial statements for Martin, Inc., are shown here:

```{r problem 4 financial statements, echo=F}
p4 <- new.env()

p4$is <- data.frame(
    Category = c("Sales", "Costs", "taxable income",
                 "Taxes (34%)", "Net income"),
    Amount = c(37300, 25800, 11500, 3910, 7590)
)

p4$bs <- data.frame(
    Category =  c("Assets", "Debt", "Equity"),
    Amount = c(127000, 30500, 96500)
)

knitr::kable(p4$is, caption="Income Statement")
knitr::kable(p4$bs, caption="Balance Sheet")

p4$dividend <- 2500
p4$projected.sales <- 42300
p4$payout.ratio <- p4$dividend / p4$is$Amount[5]
p4$growth.rate <- (p4$projected.sales - p4$is$Amount[1]) / p4$is$Amount[1]
p4$projected.assets <- p4$bs$Amount[1] * (1 + p4$growth.rate)
p4$projected.net.income <- p4$is$Amount[5] * (1 + p4$growth.rate)
p4$addition.to.retained.earnings <- p4$projected.net.income *
                                     (1 - p4$payout.ratio)
p4$projected.equity <- p4$bs$Amount[3] + p4$addition.to.retained.earnings
p4$efn <- p4$projected.assets - (p4$bs$Amount[2] + p4$projected.equity)
```

Assets and costs are proportional to sales. Debt and equity are not. A dividend
of \$2,500 was paid, and Martin wishes to maintain a constant payout ratio. Next
year's sales are projected to be \$42,300. What external financing is needed?

---

Since Assets and costs are proportional to projected sales (\$42,300) for the
income statement we get:

$$
\begin{aligned}
\mathbf{Payout\ ratio} &= \frac{\text{dividends}}{\text{net income}}\\
  &= \frac{`r pretty(p4$dividend)`}
          {`r pretty(p4$is$Amount[5])`}\\
  &\approx `r pretty(round(p4$payout.ratio, 4))`\\\\
\mathbf{Growth\ rate} &= \frac{\text{projected sales} - \text{current sales}}
  {\text{current sales}}\\
  &= \frac{`r pretty(p4$projected.sales)` - `r pretty(p4$is$Amount[1])`}
          {`r pretty(p4$is$Amount[1])`}\\
  &= `r round(p4$growth.rate, 4)`\\\\
\mathbf{Projected\ assets} &= \text{current assets}
  \times (1 + \text{growth rate})\\
  &= `r pretty(p4$bs$Amount[1])` \times `r 1 + p4$growth.rate`\\
  &= `r pretty(p4$projected.assets)`\\\\
\mathbf{Projected\ net\ income} &= \text{current net income}
  \times (1 + \text{growth rate})\\
  &= `r pretty(p4$is$Amount[5])` \times `r 1 + round(p4$growth.rate, 4)`\\
  &= `r pretty(round(p4$projected.net.income, 2))`\\\\
\mathbf{Addition\ to\ retained\ earnings} &= \text{net income}
  \times \text{retention raio}\\
  &= `r pretty(round(p4$projected.net.income, 2))`
  \times (1 - `r round(p4$payout.ratio, 4)`)\\
  &= `r pretty(round(p4$addition.to.retained.earnings, 2))`\\\\
\mathbf{Projected\ equity} &= \text{current equity}
  + \text{addition to retained earnings}\\
  &= `r pretty(p4$bs$Amount[3])` +
     `r pretty(round(p4$addition.to.retained.earnings))`\\
  &= `r pretty(round(p4$projected.equity, 2))`\\\\
\mathbf{EFN} &= \text{projected assets} - (\text{current debt}
  + \text{projected equity})\\
  &= `r pretty(p4$projected.assets)` - (`r pretty(p4$bs$Amount[2])`
  + `r pretty(p4$projected.equity)`)\\
  &= `r pretty(p4$efn)`
\end{aligned}
$$

```{r solution 4, echo=F}
pct.inc.sales <- 42300/37300
payout.ratio <- 2500/7590
assets.proforma <- 127000 * pct.inc.sales
netincome.proforma <- 7590 * pct.inc.sales
retained.earnings.proforma <- netincome.proforma*(1-payout.ratio)
equity.proforma <- 96500 + retained.earnings.proforma
efn <- assets.proforma - (30500 + equity.proforma)
```

---

# 5. Sales and growth

The most recent financial statements for Fontenot Co. are shown here:

```{r problem 5, echo=F}
p5 <- new.env()
p5$is <- matrix(c(54000, 39300, 14700, 4998, 9702), ncol=1)
row.names(p5$is) <- c("Sales", "Costs", "Taxable income",
                      "Taxes (34%)", "Net income")
p5$bs <- matrix(c(31000, 118000, 149000, 68000, 81000, 149000), ncol = 1)
row.names(p5$bs) <- c("Current assets", "Fixed assets", "Total Assets",
                      "Long-term debt", "Equity", "Debt and Equity")
knitr::kable(p5$is, caption="Income Statement")
knitr::kable(p5$bs, caption="Balance Sheet")
```

Assets and costs are proportional to sales. The company maintains a constant 30
percent dividend payout ratio and a constant debt-equity ratio. What is the
maximum increase in sales that can be sustained assuming no new equity is
issued?

---

```{r problem 5; work, echo=F}
p5$payout.ratio <- .3
p5$de.ratio <- as.numeric(p5$bs["Long-term debt",]) /
               as.numeric(p5$bs["Equity",])
p5$roe <- as.numeric(p5$is["Net income",]) /
          as.numeric(p5$bs["Equity",])
p5$plow.back <- 1 - p5$payout.ratio
p5$sgr <- (p5$roe * p5$plow.back) / (1 - p5$roe * p5$plow.back)
```

$$
\begin{aligned}
\mathbf{Retention\ ratio} &= 1 - \text{dividend payout ratio}\\
  &= 1 - \\
  &= 0.70\\\\
\mathbf{DER} &= \frac{\text{Long-term Debt}}{\text{Equity}} \\
  &= \frac{68,000}{81,000} \\
  &= 0.8395 \\ \\
\mathbf{ROE} &= \frac{\text{net income}}{\text{equity}} \\
  &= \frac{9,702}{81,000} \\
  &= 0.1198 \\ \\
\mathbf{SGR} &= \text{ROE} \times \text{retention ratio} \\
  &= 0.1198 \times 0.70 \\
  &= 0.0839 \\ \\
\mathbf{Maximum\ increase\ in\ sales} &= \text{current sales} \times \text{SGR} \\
  &= 54,000 \times 0.0839 \\
  &= 4,531 \\ \\
\mathbf{Projected\ sales} &= \text{current sales} + \text{maximum increase in sales} \\
  &= 54,000 + 4,531 \\
  &= 58,531
\end{aligned}
$$

\$4,531 is the largest sales growth Fontenot Co. can sustain without issuing new
equity while maintaining its current policies. The total projected sales is
\$58,531.

```{r solution 5}
p5$retention.ratio <- 0.70
p5$debt.to.equity.ratio <- 68000/81000
p5$roe <- 9702/81000
p5$sgr <- p5$roe * p5$retention.ratio
cat("Maximum increase in sales:", p5$max.inc.sales <- 54000 * p5$sgr)
cat("Projected Total Sales:", p5$projected.sales <- 54000 + p5$max.inc.sales)
```

---

# 6. Sustainable growth

If the Layla Corp. has a 13 percent ROE and a 20 percent payout ratio, what is
its sustainable growth rate.

---

```{r problem 6, echo=F}
p6 <- new.env()
p6$roe <- .13
p6$payout.ratio <- .2
p6$sgr <- (p6$roe * (1 - p6$payout.ratio)) /
          (1 - p6$roe * (1 - p6$payout.ratio))
```

$$
\begin{aligned}
\mathbf{SGR} &= \frac{\text{ROE} \times \text{retention ratio}}
     {1 - \text{ROE} \times \text{retention ratio}}\\
  &= \frac{`r p6$roe` \times (1 - `r p6$payout.ratio`)}
          {1 - `r p6$roe` \times (1 - `r p6$payout.ratio`)}\\
  &= `r round(p6$sgr, 4)`
\end{aligned}
$$

---

# 7. Sustainable growth

Assuming the following ratios are constant, what is the sustainable growth rate?

* Total asset turnover = 2.20
* Profit margin = 7.4%
* Equity multiplier = 1.40
* Payout ratio = 40%

---

```{r problem 7, echo=F}
p7 <- new.env()
p7$total.asset.turnover <- 2.2
p7$profit.margin <- .074
p7$equity.multiplier <- 1.4
p7$payout.ratio <- .4
p7$roa <- p7$profit.margin * p7$total.asset.turnover
p7$roe <- p7$roa * p7$equity.multiplier
p7$sgr <- p7$roe * (1 - p7$payout.ratio)
```

$$
\begin{aligned}
\mathbf{ROA} &= \text{profit margin} \times \text{total asset turnover} \\
  &= `r p7$profit.margin` \times `r p7$total.asset.turnover`\\
  &= `r p7$roa`\\\\
\mathbf{ROE} &= \text{ROA} \times \text{equity multiplier} \\
  &= `r p7$roa` \times `r p7$equity.multiplier`\\
  &= `r round(p7$roe, 4)`\\\\
\mathbf{Sustainable\ Growth\ Rate} &= \text{ROE} \times \text{retention ratio}\\
  &= \text{ROE} + (1 - \text{payout ratio}) \\
  &= `r p7$roe` + (1 - `r p7$payout.ratio`)\\
  &= `r round(p7$sgr, 4)`\\\\
\end{aligned}
$$

---

# 8. Calculating EFN

The most recent financial statements for Bradley, Inc., are shown here (assuming no income taxes):

```{r problem 8, echo=F}
p8 <- new.env()
p8$is <- matrix(c(6500, 5320, 1180), ncol=1)
row.names(p8$is) <- c("Sales", "Costs", "Net income")
p8$bs <- matrix(c(17400, 17400, 8400, 9000, 17400), ncol=1)
row.names(p8$bs) <- c("Assets", "Total Assets", "Debt", "Equity", "Debt and Equity")
knitr::kable(p8$is, caption="Income Statement")
knitr::kable(p8$bs, caption="Balance Sheet")
```

Assets and costs are proportional to sales. Debt and equity are not. No dividends are paid. Next year's sales are projected to be \$7,280. What is the external financing needed?

---

```{r problem 8; work, echo=F}
p8$projected.sales <- 7280
p8$growth <- p8$projected.sales / as.numeric(p8$is["Sales",])
p8$projected.assets <- as.numeric(p8$bs["Assets",]) * p8$growth
p8$projected.net.income <- as.numeric(p8$is["Net income",]) * p8$growth
p8$projected.equity <- as.numeric(p8$bs["Equity",]) + p8$projected.net.income
p8$efn <- p8$projected.assets - (as.numeric(p8$bs["Debt",])
                              + p8$projected.equity)
```

$$
\begin{aligned}
\mathbf{Growth\ rate} &= \frac{\text{projected sales} - \text{sales}}
  {\text{sales}}\\
  &= \frac{`r pretty(p8$projected.sales)`
  - `r pretty(as.numeric(p8$is["Sales",]))`}
          {`r pretty(as.numeric(p8$is["Sales",]))`}\\
  &= `r p8$growth - 1`\\
\mathbf{Projected\ assets} &= \text{assets} \times (1 + \text{growth rate})\\
  &= `r pretty(as.numeric(p8$bs["Assets",]))` \times `r p8$growth`\\
  &= `r pretty(p8$projected.assets)`\\\\
\mathbf{Projected\ net\ income} &= \text{net income}
  \times (1 + \text{growth rate})\\
  &= `r pretty(as.numeric(p8$is["Net income",]))` \times `r p8$growth`\\
  &= `r pretty(p8$projected.net.income)`\\\\
\mathbf{Projected\ equity} &= \text{equity} + \text{projected net income}\\
  &= `r pretty(as.numeric(p8$bs["Equity",]))` +
     `r pretty(as.numeric(p8$projected.net.income))`\\
  &= `r pretty(p8$projected.equity)`\\\\
\mathbf{EFN} &= \text{projected assets} - (\text{debt} + \text{projected equity})\\
  &= `r pretty(p8$projected.assets)` - (`r pretty(as.numeric(p8$bs["Debt",]))`
  +  `r pretty(p8$projected.equity)`)\\
  &= `r pretty(p8$efn)`
\end{aligned}
$$

---

# 9. External funds needed

Cheryl Colby, CFO of Charming Florist Ltd., has created the firm's pro forma
balance sheet for the next fiscal year. Sales are projected to grow by 10
percent to \$420 million. Current assets, fixed assets, and short-term debt are
20 percent, 75 percent, and 15 percent of sales, respectively. Charming Florist
pays out 30 percent of its net income in dividends. The company currently has
\$120 million of long-term debt and \$48 million in common stock par value. The
profit margin is 9 percent.

a. Construct the current balance sheet for the firm using the projected sales
   figure.

b. Based on Ms. Colby's sales growth forecast, how much does Charming Florist
   need in external funds for the upcoming fiscal year?

c. Construct the firm's pro forma balance sheet for the next fiscal year and
   confirm the external funds needed that you calculated in part (b).

---

```{r problem 9, echo=F}
options(digits=5)
p9 <- new.env()

p9$projected.sales <- 420
p9$sales.growth <- .1
p9$current.assets.pct.of.sales <- .2
p9$fixed.assets.pct.of.sales <- .75
p9$short.term.debt.pct.of.sales <- .15
p9$long.term.debt <- 120
p9$dividend.payout <- .3
p9$equity.par.value <- 48
p9$profit.margin <- .09

p9$sales <- p9$projected.sales / (1 + p9$sales.growth)

# Current variables
p9$current <- matrix(p9$sales * c(
    p9$current.assets.pct.of.sales,
    p9$fixed.assets.pct.of.sales,
    p9$short.term.debt.pct.of.sales), ncol=1)
rownames(p9$current) <- c(
    "Current assets",
    "Fixed assets",
    "Short-term debt")
p9$current.total.assets <- sum(p9$current[1:2,])
p9$current.total.liabilities <- as.numeric(p9$current[3,]) + p9$long.term.debt
p9$current.net.income <- p9$profit.margin * p9$sales
p9$current.retained.earnings <- p9$current.net.income *
  (1 - p9$dividend.payout)
p9$current.equity <- p9$current.total.assets - p9$current.total.liabilities

# Projected variables
p9$projected <- matrix(p9$projected.sales * c(
    p9$current.assets.pct.of.sales,
    p9$fixed.assets.pct.of.sales,
    p9$short.term.debt.pct.of.sales), ncol=1)
rownames(p9$projected) <- c(
    "Current assets",
    "Fixed assets",
    "Short-term debt")
p9$projected.total.assets <- sum(p9$projected[1:2,])
p9$projected.total.liabilities <- p9$projected[3,] + p9$long.term.debt
p9$projected.net.income <- p9$profit.margin * p9$projected.sales
p9$projected.retained.earnings <- p9$projected.net.income *
  (1 - p9$dividend.payout)
p9$projected.equity <- p9$current.equity + p9$projected.retained.earnings

p9$total.available.financing <- p9$projected["Short-term debt",] +
    p9$long.term.debt + p9$projected.equity

p9$efn <- p9$projected.total.assets - p9$total.available.financing
```

## A. Constructing the current balance sheet based on current sales.

### NOTE:

All values are in millions.

$$
\begin{aligned}
\text{Current sales} &= \frac{\text{Projected sales}}{1 + \text{sales growth}}\\
  &= \frac{`r pretty(p9$projected.sales)`}{`r 1 + p9$sales.growth`}\\
  &= `r pretty(p9$sales)`\\\\
\mathbf{Assets}\\
\text{Current assets} &= \text{current sales}
  \times \text{current assets as percent of sales}\\
  &= `r pretty(p9$sales)` \times `r p9$current.assets.pct.of.sales`\\
  &= `r pretty(p9$current["Current assets",])`\\\\
\text{Fixed assets} &= \text{current sales}
  \times \text{fixed assets as percent of sales}\\
  &= `r pretty(p9$sales)` \times `r p9$fixed.assets.pct.of.sales`\\
  &= `r pretty(p9$current["Fixed assets",])`\\\\
\text{Total assets} &= \text{current assets} + \text{fixed assets}\\
  &= `r pretty(p9$current["Current assets",])`
  + `r pretty(p9$current["Fixed assets",])`\\
  &= `r pretty(p9$current.total.assets)`\\\\
\mathbf{Liabilities}\\
\text{Short-term debt} &= \text{current sales}
  \times \text{short-term assets as percent of sales}\\
  &= `r pretty(p9$sales)` \times `r pretty(p9$short.term.debt.pct.of.sales)`\\
  &= `r pretty(p9$current["Short-term debt",])`\\\\
\text{Long-term debt} &= `r pretty(p9$long.term.debt)`\\\\
\text{Total debt} &= \text{short-term debt} + \text{long-term debt}\\
  &= `r pretty(p9$current["Short-term debt",])` + `r pretty(p9$long.term.debt)`\\
  &= `r pretty(p9$current.total.liabilities)`\\\\
\mathbf{Equity}\\
\text{Equity} &= \text{Total assets} - \text{Total liabilities}\\
  &= `r pretty(p9$current.total.assets)`
  - `r pretty(p9$current.total.liabilities)`\\
  &= `r pretty(p9$current.equity)`
\end{aligned}
$$

## B. External financing is needed

### Step 1: Projected sales = \$`r pretty(p9$projected.sales)`

$$
\begin{aligned}
\text{Projected net income} &= \text{profit margin}
  \times \text{projected sales}\\
  &= `r p9$profit.margin` \times `r p9$projected.sales`\\
  &= `r p9$projected.net.income`\\\\
\text{Projected retained earnings} &= \text{retention ratio}
  \times \text{projected net income}\\
  &= `r (1 - p9$dividend.payout)` \times `r p9$projected.net.income`\\
  &= `r p9$projected.retained.earnings`
\end{aligned}
$$

### Step 2: Project assets

$$
\begin{aligned}
\text{Projected current assets} &= \text{current assets percent of sales}
  \times \text{projected sales}\\
  &= `r p9$current.assets.pct.of.sales` \times `r p9$projected.sales`\\
  &= `r p9$projected["Current assets",]`\\\\
\text{Projected fixed assets} &= \text{fixed assets percent of sales}
  \times \text{projected sales}\\
  &= `r p9$fixed.assets.pct.of.sales` \times `r p9$projected.sales`\\
  &= `r p9$projected["Fixed assets",]`\\\\
\text{Projected total Assets} &= \text{projected current assets}
  + \text{projected fixed assets}\\
  &= `r p9$projected["Current assets",]`
  + `r p9$projected["Fixed assets",]`\\
  &= `r p9$projected.total.assets`
\end{aligned}
$$

### Step 3: Project spontaneous liabilities

$$
\begin{aligned}
\text{Projected short-term debt} &= \text{short-term debt percent of sales}
  \times \text{projected sales}\\
  &= `r p9$short.term.debt.pct.of.sales` \times `r p9$projected.sales`\\
  &= `r p9$projected["Short-term debt",]`
\end{aligned}
$$

### Step 4: Add existing long-term debt and equity

$$
\begin{aligned}
\text{Long-term debt} &= `r p9$long.term.debt`\\\\
\text{Projected equity} &= \text{equity from current balance sheet}
  + \text{projected retained earnings}\\
  &= `r p9$current.equity` + `r p9$projected.retained.earnings`\\
  &= `r p9$projected.equity`
\end{aligned}
$$

### Step 5: Total available financing

$$
\begin{aligned}
\text{Projected liabilties} + \text{projected equity} &=
  \text{projected short-term debt}\\
  &\ \ \ \ + \text{long-term debt}\\
  &\ \ \ \ + \text{projected equity}\\
  &= `r p9$projected["Short-term debt",]`+ `r p9$long.term.debt`
  +  `r p9$projected.equity`\\
  &= `r p9$total.available.financing`
\end{aligned}
$$

### Step 6: EFN = Total assets - Total financing

$$
\begin{aligned}
\text{EFN} &= \text{projected total assets} - \text{total financing}\\
  &= `r p9$projected.total.assets` - `r p9$total.available.financing`\\
  &= `r p9$efn`
\end{aligned}
$$

## C. Pro forma balance sheet

$$
\begin{aligned}
\mathbf{Assets}\\
\text{Current assets} &= \text{current sales}
  \times \text{current assets as percent of sales}\\
  &= `r pretty(p9$projected.sales)` \times `r p9$current.assets.pct.of.sales`\\
  &= `r pretty(p9$projected["Current assets",])`\\\\
\text{Fixed assets} &= \text{current sales}
  \times \text{fixed assets as percent of sales}\\
  &= `r pretty(p9$projected.sales)` \times `r p9$fixed.assets.pct.of.sales`\\
  &= `r pretty(p9$projected["Fixed assets",])`\\\\
\text{Total assets} &= \text{current assets} + \text{fixed assets}\\
  &= `r pretty(p9$projected["Current assets",])`
  + `r pretty(p9$projected["Fixed assets",])`\\
  &= `r pretty(p9$projected.total.assets)`\\\\
\mathbf{Liabilities}\\
\text{Short-term debt} &= \text{current sales}
  \times \text{short-term assets as percent of sales}\\
  &= `r pretty(p9$projected.sales)`
  \times `r pretty(p9$short.term.debt.pct.of.sales)`\\
  &= `r pretty(p9$projected["Short-term debt",])`\\\\
\text{Long-term debt} &= `r pretty(p9$long.term.debt)`\\\\
\text{Total debt} &= \text{short-term debt} + \text{long-term debt}\\
  &= `r pretty(p9$projected["Short-term debt",])`
  + `r pretty(p9$long.term.debt)`\\
  &= `r pretty(p9$projected.total.liabilities)`\\\\
\mathbf{Equity}\\
\text{Equity} &= \text{Current equity}
  - \text{Projected retained earnings}\\
  &= `r pretty(p9$current.equity)`
  + `r pretty(p9$projected.retained.earnings)`\\
  &= `r pretty(p9$projected.equity)`
\end{aligned}
$$

---

# 10. Sustainable growth rate

The Steiben company has an ROE of 13.1 percent and a payout ratio of 40 percent.

a. What is the company's sustainable growth rate?

b. Can the company's actual growth rate be different from its sustainable growth
   rate? Why or why not?

c. How can the company increase its sustainable growth rate?

---

```{r problem 10; sustainable growth rate, echo=F}
p10 <- new.env()
p10$roe <- .131
p10$payout.ratio <- .4
p10$sgr <- (p10$roe * (1 - p10$payout.ratio)) /
           (1 - p10$roe * (1 - p10$payout.ratio))
```

$$
\begin{aligned}
\mathbf{SGR} &= \frac{\text{ROE} \times (1 - \text{payout ratio})}
     {1 - \text{ROE} \times (1 - \text{payout ratio})}\\
  &= \frac{`r p10$roe` \times (1 - `r p10$payout.ratio`)}
     {1 - `r p10$roe` \times (1 - `r p10$payout.ratio`)}\\
  &= `r p10$sgr`
\end{aligned}
$$

b. Yes, a company growth rate can be different from its sustainable growth rate.
If it is below the SGR, then the company is operating at less than full
capacity. If the growth rate is above the sustainable growth rate, it would need
external funds to sustain the growth rate.

c. The sustainable growth rate is determined by the following factors based on
   the ROE and the payout ratio.

$$
\text{ROE} = \text{profit margin} \times \text{total asset turnover}
  \times \text{equity multiplier}
$$

The following four factors affect the sustainable growth rate:

### Profit margin

An increase in profit margin will increase the firm's ability to generate funds
internally and thereby increase its sustainable growth.

### Dividend policy

A decrease in the payout ratio will increase the retention ratio. This increases
internally generated equity and thus increases sustainable growth.

### Financial policy

An increase in the debt-equity ratio increases the firm's financial leverage.
Because this makes additional debt financing available, it increases the
sustainable growth rate.

### Total asset turnover

An increase in the firm's total asset turnover increases the sales generated for
each dollar in assets. This decreases the firm's need for new assets as sales
grow and thereby increases the sustainable growth rate. Notice that increasing
total asset turnover is the same thing as decreasing capital intensity.

---

# 11. Return on equity

Firm A and Firm B have debt-total asset ratios of 35 percent and 55 percent and
returns on total assets of 9 percent and 7 percent, respectively. Which firm has
a greater return on equity?

---

```{r problem 11; return on equity, echo=F}
p11 <- new.env()
p11$total.debt.ratio <- c(.35, .55)
p11$roa <- c(.09, .07)
p11$de.ratio <- (1 - p11$total.debt.ratio)
p11$equity.multiplier <- 1 + p11$de.ratio
p11$roe <- p11$roa * p11$equity.multiplier
```

The debt-total asset ratio is also called the total debt ratio.

$$
\begin{aligned}
\mathbf{Total\ debt\ ratio}\\
\text{Firm A} &= `r p11$total.debt.ratio[1]`\\
\text{Firm B} &= `r p11$total.debt.ratio[2]`\\\\
\mathbf{ROA}\\
\text{Firm A} &= `r p11$roa[1]`\\
\text{Firm B} &= `r p11$roa[2]`\\\\
\mathbf{DE\ ratio} &= (1 - \text{Total debt ratio})\\
\text{Firm A} &= `r p11$de.ratio[1]`\\
\text{Firm B} &= `r p11$de.ratio[2]`\\\\
\mathbf{Equity\ multiplier} &= 1 + \text{DE ratio}\\
\text{Firm A} &= `r p11$equity.multiplier[1]`\\
\text{Firm B} &= `r p11$equity.multiplier[2]`\\\\
\mathbf{ROE} &= \text{ROA} \times \text{Equity multiplier}\\
\text{Firm A} &= `r p11$roe[1]`\\
\text{Firm B} &= `r p11$roe[2]`
\end{aligned}
$$

## Conclusion

Although firm B has lower ROA, it has a greater return on equity because it uses more leverage (higher debt-total asset ratio).

---

# 12. Ratios and Foreign Companies

Prince Albert Canning PLC had a net loss of 37,543 pounds on sales of 345,182
pounds. What was the company's profit margin? Does the fact that these figures
are quoted in a foreign currency make any difference? Why? In dollars, sales
were \$559,725. What was the net loss in dollars?

---

```{r problem 12; ratios and foreign companies, echo=F}
p12 <- new.env()
p12$net.loss.pd <- 37543
p12$sales.pd <- 345182
p12$sales.us <- 559725
p12$profit.margin <- - p12$net.loss.pd / p12$sales.pd
p12$pd.us <- p12$sales.us / p12$sales.pd
p12$net.loss.us <- p12$net.loss.pd * p12$pd.us
```

$$
\begin{aligned}
\mathbf{Profit\ margin} &= \frac{\text{net income}}{\text{sales}}\\
  &= \frac{`r pretty(p12$net.loss.pd)`}{`r pretty(p12$sales.pd)`}\\
  &= `r round(p12$profit.margin, 4)`
\end{aligned}
$$

The profit margin would be the same in any currency because it is a ratio.

If sales were 559,725 in dollars. Then the exchange rate between dollars and
pounds is

$$
\begin{aligned}
\text{exchange rate} &= \frac{\text{sales USD}}{\text{sales PD}}\\
  &= \frac{`r pretty(p12$sales.us)`}{`r pretty(p12$sales.pd)`}\\
  &= `r round(p12$pd.us, 2)`
\end{aligned}
$$

The net loss in dollars is

$$
\begin{aligned}
\text{Net loss USD} &= \text{net loss PD} \times \text{exchange rate}\\
  &= `r pretty(p12$net.loss.pd)` \times `r p12$pd.us`\\
  &= `r pretty(p12$net.loss.us)`
\end{aligned}
$$

---

# 13. External Funds Needed

The Optical Scam Company has forecast a 15 percent sales growth rate for next
year. The current financial statements are shown here:

```{r problem 13; income statement, echo=F}
p13 <- new.env()
p13$is <- matrix(c(30400000, 26720000, 3680000, 1288000,
                   2392000, 956800, 1435200),
                 ncol=1)
row.names(p13$is) <- c("Sales", "Costs", "Taxable income", "Taxes",
                       "Net income", "Dividends",
                       "Addition to retained earnings")

knitr::kable(pretty(p13$is), caption="Income Statement") %>%
    kable_styling(latex_options = "HOLD_position")
```

```{r problem 13; balance sheet, echo=F}
p13$bs.assets <- c(7200000, 17600000)
p13$bs.liabilities.and.equity <- c(6400000, 4800000, 3200000, 10400000)
p13$bs.total.assets <- sum(p13$bs.assets)
p13$bs.total.liabilities <- sum(p13$bs.liabilities.and.equity[1:2])
p13$bs.total.equity <- sum(p13$bs.liabilities.and.equity[3:4])
p13$bs.total.liabilities.and.equity <- sum(p13$bs.liabilities.and.equity)
p13$bs <- matrix(c(p13$bs.assets, p13$bs.total.assets,
                   p13$bs.liabilities.and.equity, p13$bs.total.equity,
                   p13$bs.total.liabilities.and.equity), ncol=1)
rownames(p13$bs) <- c("Current assets", "Fixed assets", "Total assets",
                      "Short-term debt", "Long-term debt", "Common stock",
                      "Accumulated retained earnings", "Total equity",
                      "Total liabilities and equity")
knitr::kable(pretty(p13$bs), caption="Balance Sheet") %>%
    pack_rows("Assets", 1, 2) %>%
    pack_rows("Liabilities", 4, 5) %>%
    pack_rows("Stockholders' equity", 6, 8) %>%
    add_indent(8) %>%
    row_spec(c(2, 3, 7, 8),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")
```

a. Using the equation, calculate the external funds needed for next year.

b. Construct the firm's pro forma balance sheet for next year and confirm the
   external funds needed that you calculated in part (a).

c. Calculate the sustainable growth rate for the company.

d. Can Optical Scam eliminate the need for external funds by changing its
   dividend policy? What other options are available to the company to met its
   growth objectives?

---

## A. The equation for EFN

```{r solution 13a sales incease, echo=F}
p13$pct.sales.inc <- .15
p13$sales <- as.numeric(p13$is["Sales",])
p13$delta.sales <- p13$sales * p13$pct.sales.inc
p13$projected.sales <- p13$sales + p13$delta.sales
p13$short.term.debt <- as.numeric(p13$bs["Short-term debt",])
p13$capital.intensity.ratio <- as.numeric(p13$bs.total.assets) / p13$sales
p13$profit.margin <- as.numeric(p13$is["Net income",]) / p13$sales
p13$dividend.payout.ratio <- as.numeric(p13$is["Dividends",]) /
                              as.numeric(p13$is["Net income",])
p13$retention.ratio <- 1 - p13$dividend.payout.ratio
p13$efn  <- p13$capital.intensity.ratio * p13$delta.sales -
            p13$short.term.debt / p13$sales * p13$delta.sales -
            p13$profit.margin * p13$projected.sales * p13$retention.ratio
```

$$
\begin{aligned}
\Delta \text{Sales} &= \text{Sales}
  \times \text{precent increase in projected sales}\\
  &= `r pretty(p13$sales)` \times `r p13$pct.sales.inc`\\
  &= `r pretty(p13$delta.sales)`\\\\
\text{Capital intensity ratio} &= \frac{\text{total assets}}{\text{sales}}\\
  &= \frac{`r pretty(p13$bs.total.assets)`}{`r pretty(p13$sales)`}\\
  &= `r p13$capital.intensity.ratio`\\\\
\text{Profit margin} &= \frac{\text{net income}}{\text{sales}}\\
  &= \frac{`r pretty(p13$is["Net income",])`}{`r pretty(p13$sales)`}\\
  &= `r p13$profit.margin`\\\\
\text{Projected Sales} &= \text{Sales} + \Delta \text{Sales}\\
  &= `r pretty(p13$sales)` + `r pretty(p13$delta.sales)`\\
  &= `r pretty(p13$projected.sales)`\\\\
\text{Retention ratio} &= 1 - `r p13$dividend.payout.ratio`\\
  &= `r p13$retention.ratio`\\\\
\text{EFN} &= \text{capital intensity ratio} \times \Delta \text{sales}\\
  &\ \ \ \ - \frac{\text{spontaneous liabilities}}{\text{sales}}
  \times \Delta \text{sales}\\
  &\ \ \ \ - \text{profit margin} \times \text{projected sales}
  \times \text{retention ratio}\\
  &= `r p13$capital.intensity.ratio` \times `r p13$delta.sales`\\
  &\ \ \ \ - \frac{`r p13$short.term.debt`}{`r p13$sales`}
  \times `r p13$delta.sales`\\
  &\ \ \ \ - `r p13$profit.margin` \times `r p13$projected.sales`
  \times `r p13$retention.ratio`\\
  &= `r pretty(p13$efn)`
\end{aligned}
$$

## B. The Pro forma balance sheet

```{r solution 13b, echo=F}
p13$bs.assets.pro.forma <- p13$bs.assets * 1.15
p13$bs.total.assets.pro.forma <- sum(p13$bs.assets.pro.forma)
p13$bs.short.term.debt.pro.forma <- as.numeric(p13$bs["Short-term debt",]) * 1.15
p13$bs.total.liabilities.pro.forma <- p13$bs.short.term.debt.pro.forma +
                                       as.numeric(p13$bs["Long-term debt",])
p13$bs.accumulated.retained.earnings.pro.forma <-
    as.numeric(p13$bs["Accumulated retained earnings",]) +
    as.numeric(p13$is["Addition to retained earnings",]) * 1.15
p13$bs.total.equity.pro.forma <- as.numeric(p13$bs["Common stock",]) +
                              p13$bs.accumulated.retained.earnings.pro.forma
p13$bs.total.liabilities.and.equity.pro.forma <-
    p13$bs.total.liabilities.pro.forma +
    p13$bs.total.equity.pro.forma
p13$bs.pro.forma <- matrix(c(p13$bs.assets.pro.forma,
                             p13$bs.total.assets.pro.forma,
                             p13$bs.short.term.debt.pro.forma,
                             p13$bs["Long-term debt",],
                             p13$bs["Common stock",],
                             p13$bs.accumulated.retained.earnings.pro.forma,
                             p13$bs.total.equity.pro.forma,
                             p13$bs.total.liabilities.and.equity.pro.forma,
                             p13$efn),
                           ncol=1)
rownames(p13$bs.pro.forma) <- c("Current assets", "Fixed assets",
                                "Total assets", "Short-term debt",
                                "Long-term debt", "Common stock",
                                "Accumulated retained earnings", "Total equity",
                                "Total liabilities and equity",
                                "External financing needed")
knitr::kable(pretty(p13$bs.pro.forma),
             caption="Pro Forma Balance Sheet") %>%
    pack_rows("Assets", 1, 2) %>%
    pack_rows("Liabilities", 4, 5) %>%
    pack_rows("Stockholders' equity", 6, 8) %>%
    add_indent(8) %>%
    row_spec(c(2, 3, 7, 8),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")
```

## C. Sustainable growth

```{r problem 13c, echo=F}
p13$roe <- as.numeric(p13$is["Net income",]/p13$bs["Total equity",])
p13$sustainable.growth.rate <- (p13$roe * p13$retention.ratio) /
                                (1 - p13$roe * p13$retention.ratio)
```

$$
\begin{aligned}
\text{ROE} &= \frac{\text{net income}}{\text{total equity}}\\
  &= \frac{`r pretty(p13$is["Net income",])`}
          {`r pretty(p13$bs["Total assets",])`}\\
  &= `r round(p13$roe, 3)`\\\\
\text{Sustainable growth rate} &= \frac{\text{ROE} \times b}
                                       {1 - \text{ROE} \times b}\\
  &= \frac{`r p13$roe` \times `r p13$retention.ratio`}
          {1 - `r p13$roe` \times `r p13$retention.ratio`}\\
  &= `r p13$sustainable.growth.rate`
\end{aligned}
$$

## D. Options to external funds needed

* Reducing the payout ratio would increase retained earnings, reducing EFN.
* Increasing profit margin to generate more internal funds.
* Reduce asset growth (e.g., improve efficiency or delay capital expenditures).
* Increase spontaneous liabilities (e.g., negotiate better payment terms with suppliers).
* Issue new equity or take on additional long-term debt.

---

# 14. Days' sales in receivables

A company has net income of \$265,000, a profit margin of 9.3 percent, and an accounts receivable balance of \$145,300. Assuming 80 percent of sales are on credit, what is the company's days' sales in receivables?

---

```{r problem 14; days sales in receivables, echo=F}
p14 <- new.env()
p14$net.income <- 265000
p14$profit.margin <- .093
p14$accounts.receivable <- 145300
p14$pct.sales.on.credit <- .8
p14$sales <- p14$net.income / p14$profit.margin
p14$sales.on.credit <- p14$sales * p14$pct.sales.on.credit
p14$receivables.turnover <- round(p14$sales.on.credit /
                                  p14$accounts.receivable, 1)
p14$days.sales.in.receivables <- round(365 / p14$receivables.turnover, 1)
```

$$
\begin{aligned}
\text{Sales} &= \frac{\text{net income}}{\text{profit margin}}\\
  &= \frac{`r pretty(p14$net.income)`}{`r p14$profit.margin`}\\
  &= `r pretty(p14$sales)`\\\\
\text{Sales on credit} &= \text{sales}
  \times \text{percent of sales on credit}\\
  &= `r pretty(p14$sales)` \times `r p14$pct.sales.on.credit`\\
  &= `r pretty(p14$sales.on.credit)`\\\\
\text{Receivables turnover} &= \frac{\text{sales on credit}}
                                    {\text{accounts receivable}}\\
  &= \frac{`r pretty(p14$sales.on.credit)`}
          {`r pretty(p14$accounts.receivable)`}\\
  &= `r p14$receivables.turnover`\\
\text{Days' sales in receivables} &= \frac{365\ \text{days}}
                                          {\text{receivables turnover}}\\
  &= \frac{365}{`r round(p14$receivables.turnover, 1)`}\\
  &= `r p14$days.sales.in.receivables`\\\\
\end{aligned}
$$

# 15. Ratios and fixed assets

The Le Bleu Company has a ratio of long-term debt to total assets of .35 and a
current ratio of 1.25. Current liabilities are \$950, sales are \$5,780, profit
margin is 9.4 percent, and ROE is 18.2 percent. What is the amount of the firm's
net fixed assets?

---

```{r problem 15 ratios and fixed assets, echo=F}
p15 <- new.env()
p15$lt.debt.asset.ratio <- .35
p15$current.ratio <- 1.25
p15$current.liabilities <- 950
p15$sales <- 5780
p15$profit.margin <- .094
p15$roe <- .182
p15$current.assets <- p15$current.ratio * p15$current.liabilities
p15$net.income <- p15$profit.margin * p15$sales
p15$equity <- round(p15$net.income / p15$roe, 2)
p15$total.assets <- round((p15$current.liabilities + p15$equity) /
                          (1 - p15$lt.debt.asset.ratio), 2)
p15$fixed.assets <- p15$total.assets - p15$current.assets
p15$fixed.assets2 <- p15$current.liabilities / (1 - p15$lt.debt.asset.ratio) +
                     p15$equity / (1 - p15$lt.debt.asset.ratio) -
                     p15$current.assets
```

$$
\begin{aligned}
\text{current assets} &= \text{current ratio}
  \times \text{current liabilities}\\
  &= `r p15$current.ratio` \times `r p15$current.liabilities`\\
  &= `r pretty(p15$current.assets)`\\\\
\text{Net income} &= \text{profit margin} \times \text{sales}\\
  &= `r p15$profit.margin` \times `r p15$sales`\\
  &= `r p15$net.income`\\\\
\text{Equity} &= \frac{\text{net income}}{\text{ROE}}\\
  &= \frac{`r p15$net.income`}{`r p15$roe`}\\
  &= `r pretty(p15$equity)`\\\\
\end{aligned}
$$

$$
\begin{aligned}
\text{Assets} &= \text{liabilities} + \text{Equity}\\
  &= \frac{\text{long-term debt}}{\text{assets}} \times \text{assets} +
  \text{current liabilities} + \text{equity} \\ \\
\end{aligned}
$$

$$
\begin{aligned}
\text{assets}
  \times \left(1 - \frac{\text{long-term debt}}{\text{assets}} \right) &=
  \text{current liabilities} + \text{equity}\\
\end{aligned}
$$

$$
\begin{aligned}
\text{assets} &= \frac{\text{current liabilities}
  + \text{equity}}{(1 - \frac{\text{long-term debt}}{\text{assets}})}\\
\text{assets} &= \frac{`r p15$current.liabilities` + `r pretty(p15$equity)`}
  {`r 1 - p15$lt.debt.asset.ratio`}\\
  &= `r pretty(p15$total.assets)`\\\\
\end{aligned}
$$

$$
\begin{aligned}
\text{Fixed\ assets} &= \text{assets} - \text{current assets}\\
  &= `r pretty(p15$total.assets)` - `r pretty(p15$current.assets)`\\
  &= `r pretty(p15$fixed.assets)`
\end{aligned}
$$

---

# 16. Calculating the cash coverage ratio

Titan Inc.'s net income for the most recent year was \$8,320. The tax rate was
34 percent. The firm paid \$1,940 in total interest expense and deducted \$2,730
in depreciation expense. What was Titan's cash coverage ratio for the year?

---

```{r problem 16; calculating the cash coverage ratio, echo=F}
p16 <- new.env()
p16$net.income <- 8320
p16$tax.rate <- .34
p16$interest <- 1940
p16$depreciation <- 2730
p16$taxable.income <- round(p16$net.income / (1 - p16$tax.rate), 2)
p16$ebitda <- p16$taxable.income + p16$interest + p16$depreciation
p16$cash.coverage.ratio <- round(p16$ebitda / p16$interest, 2)
```

$$
\begin{aligned}
\text{Taxable income} &= \frac{\text{net income}}{(1 - \text{tax rate})}\\
  &= \frac{`r pretty(p16$net.income)`}{(1 - `r p16$tax.rate`)}\\
  &= `r pretty(p16$taxable.income)`\\\\
\text{EBITDA} &= \text{taxable income} + \text{interest} + \text{depreciation}\\
  &= `r pretty(p16$taxable.income)` + `r pretty(p16$interest)`
  +  `r pretty(p16$depreciation)`\\
  &= `r pretty(p16$ebitda)`\\\\
\text{Cash coverage ratio} &= \frac{\text{EBITDA}}{\text{interest}}\\
  &= \frac{`r pretty(p16$ebitda)`}{`r pretty(p16$interest)`}\\
  &= `r p16$cash.coverage.ratio`
\end{aligned}
$$

---

# 17. DuPont identity

The DuPont identity presented in the chapter is commonly referred to as the
three-factor DuPont identity. Another common way that the DuPont identity is
expressed is the five-factor model, which is:

$$
\text{ROE} = \frac{\text{net income}}{\text{EBT}} \times \frac{\text{EBT}}{\text{EBIT}}
  \times \frac{\text{EBIT}}{\text{sales}} \times \frac{\text{sales}}{\text{total assets}}
  \times \frac{\text{total assets}}{\text{equity}}
$$

Derive the five-factor DuPont identity (EBT is earnings before tax, but after interest). What does each term measure?

---

Deriving the five-factor DuPont identity from the three-factor DuPont identity:

$$
\begin{aligned}
\text{ROE} &= \frac{\text{net income}}{\text{sales}}
  \cdot \frac{\text{sales}}{\text{total assets}}
  \cdot \frac{\text{total assets}}{\text{equity}}\\\\
\text{ROE} &= \frac{\text{net income}}{\text{EBIT}}
  \cdot \frac{\text{EBIT}}{\text{sales}}
  \cdot \frac{\text{sales}}{\text{total assets}}
  \cdot \frac{\text{total assets}}{\text{equity}}\\\\
\text{ROE} &= \frac{\text{net income}}{\text{EBT}}
  \cdot \frac{\text{EBT}}{\text{EBIT}}
  \cdot \frac{\text{EBIT}}{\text{sales}}
  \cdot \frac{\text{sales}}{\text{total assets}}
  \cdot \frac{\text{total assets}}{\text{equity}}\\\\
\end{aligned}
$$

The five-factor DuPont identity breaks out the profit margin into three different factors.

$$
\begin{aligned}
\mathbf{Tax\ burden} &= \frac{\text{net income}}{\text{EBT}}\\\\
\mathbf{Interest\ burden} &= \frac{\text{EBT}}{\text{EBIT}}\\\\
\mathbf{Operating\ profit\ margin} &= \frac{\text{EBIT}}{\text{sales}}\\\\
\end{aligned}
$$

The last two factors are the same as in the three factor DuPont identity: Total
asset turnover and equity multiplier.

$$
\begin{aligned}
\mathbf{Total\ asset\ turnover} &= \frac{\text{sales}}{\text{total assets}}\\\\
\mathbf{Equity\ multiplier} &= \frac{\text{total assets}}{\text{equity}}
\end{aligned}
$$

---

# 18. Common-size and common-base year financial statements

In addition to common-size financial statements, common-base year financial
statements are often used. Common-based year financial statements are
constructed by dividing the current year account value by the base year account
value. Thus, the result shows the growth rate in the account. Using the
following financial statements, construct the common-size balance sheet and
common-base year balance sheet for the company. Use 2011 as the base year.

```{r problem 18 balance sheet, echo=F}
p18 <- new.env()
p18$bs.assets <- matrix(c(8014, 20453, 36822, 65289, 215370, 280659,
                          9954, 22937, 41797, 74688, 243340, 318028), ncol=2)

p18$account.names <- c("Cash", "Accounts receivable", "Inventory",
                       "Total current assets", "Net plant and equipment",
                       "Total assets", "Accounts payable", "Notes payable",
                       "Total current liabilities", "Long-term debt",
                       "Common stock and paid-in surplus", "Retained earnings",
                       "Total owners' equity",
                       "Total liabilities and owners' equity")
p18$years <- c("2011", "2012")

p18$bs.liabilities.equity <- matrix(c(40898, 17464, 58362, 24000, 38000, 160297,
                                      198297, 280659, 45884, 17035, 62919, 31000,
                                      39200, 184909, 224109, 318028), ncol=2)
p18$bs <- rbind(p18$bs.assets, p18$bs.liabilities.equity)
colnames(p18$bs) <- p18$years
rownames(p18$bs) <- p18$account.names
knitr::kable(p18$bs,
             caption="Jarrow Corporation 2011 and 2012 Balance Sheets") %>%
    pack_rows("Assets", 1, 6) %>%
    pack_rows("Current assets", 1, 4) %>%
    pack_rows("Fixed assets", 5, 5) %>%
    pack_rows("Liabilities and Owners' Equity", 7, 14) %>%
    pack_rows("Current liabilities", 7, 9) %>%
    pack_rows("Owners' equity", 11, 13) %>%
    row_spec(c(3, 4, 5, 6, 8, 9, 11, 12, 13, 14),
             extra_latex_after = "\\addlinespace") %>%
    add_indent(c(4, 9, 13)) %>%
    kable_styling(latex_options = "HOLD_position")
```

---

To construct the common-base year balance sheet you simply divide each of
the comparison years by the base year:

```{r problem 18; common-base balance sheet, echo=F}
p18$tmp <- round(p18$bs / p18$bs[,1], 2)
p18$bs.common.base <- matrix(p18$tmp[,2], ncol=1)
rownames(p18$bs.common.base) <- p18$account.names
colnames(p18$bs.common.base) <- "2011 -> 2012 Change"
p18$caption="Jarrow Corporation Common-base balance sheet base year 2011"
knitr::kable(p18$bs.common.base, caption=p18$caption) %>%
    pack_rows("Assets", 1, 6) %>%
    pack_rows("Current assets", 1, 4) %>%
    pack_rows("Fixed assets", 5, 5) %>%
    pack_rows("Liabilities and Owners' Equity", 7, 14) %>%
    pack_rows("Current liabilities", 7, 9) %>%
    pack_rows("Owners' equity", 11, 13) %>%
    row_spec(c(3, 4, 5, 6, 8, 9, 11, 12, 13, 14),
             extra_latex_after = "\\addlinespace") %>%
    add_indent(c(4, 9, 13)) %>%
    kable_styling(latex_options = "HOLD_position")
```

To construct the common-size year balance sheet we will use percentage of total
assets:

```{r problem 18; common-size balance sheet, echo=F}
p18$bs.common.size <- p18$bs / as.numeric(p18$bs["Total assets",])
p18$pct.change <- as.numeric(p18$bs.common.size[,2] -
                             p18$bs.common.size[,1])
p18$bs.common.size <- cbind(p18$bs.common.size, p18$pct.change)
rownames(p18$bs.common.size) <- p18$account.names
colnames(p18$bs.common.size) <- c(p18$years, "Change %")
p18$caption="Jarrow Corporation Common-size balance sheet base year 2011"
knitr::kable(p18$bs.common.size, caption=p18$caption) %>%
    pack_rows("Assets", 1, 6) %>%
    pack_rows("Current assets", 1, 4) %>%
    pack_rows("Fixed assets", 5, 5) %>%
    pack_rows("Liabilities and Owners' Equity", 7, 14) %>%
    pack_rows("Current liabilities", 7, 9) %>%
    pack_rows("Owners' equity", 11, 13) %>%
    row_spec(c(3, 4, 5, 6, 8, 9, 11, 12, 13, 14),
             extra_latex_after = "\\addlinespace") %>%
    add_indent(c(4, 9, 13)) %>%
    kable_styling(latex_options = "HOLD_position")
```

---

Use the following information for Problems 19, 20, and 22:

The discussion of EFN in the chapter implicitly assumed that the company was
operating at full capacity. Often, this is not the case. For example, assume
that Rosengarten was operating at 90 percent capacity. Full-capacity sales would
be \$1,000/.90 = \$1,111. The balance sheet shows \$1,800 in fixed assets. The
capital intensity ratio for the company is

$$
\begin{aligned}
\mathbf{Capital\ intensity\ ratio} &= \frac{\text{fixed assets}}
  {\text{full-capacity sales}}\\
  &= \frac{1,800}{1,111}\\
  &= 1.62
\end{aligned}
$$

This means that Rosengarten needs \$1.62 in fixed assets for every dollar in
sales when it reaches full capacity. At the projected sales level of \$1,250, it
needs $\$1,250 \times 1.62 = \$2,025$ in fixed assets, which is \$225 lower than
our projection of \$2,250 in fixed assets. So, EFN is only \$565 - 225 = \$340.

# 19. Full-capacity sales

Thorpe Mfg., Inc., is currently operating at only 90 percent of fixed asset
capacity. Current sales are \$725,000. How much can sales increase before any
new fixed assets are needed?

---

```{r problem 19; full-capacity sales, echo=F}
p19 <- new.env()
p19$current.sales <- 725000
p19$operating.capacity <- .9
p19$sales.full.capacity <- p19$current.sales / p19$operating.capacity
p19$sales.increase <- p19$sales.full.capacity - p19$current.sales
```

$$
\begin{aligned}
\text{Full-capacity sales} &= \frac{\text{Current sales}}{\text{pct sales}}\\
  &= \frac{`r pretty(p19$current.sales)`}{`r p19$operating.capacity`}\\
  &= `r pretty(p19$sales.increase)`
\end{aligned}
$$

Thorpe's full capacity sales is \$`r pretty(p19$sales.full.capacity)`. Therefore
with current sales as \$`r pretty(p19$current.sales)`, Thrope can increase sales
by \$`r pretty(p19$sales.increase)` before it would need new fixed assets.

---

# 20. Fixed assets and capacity usage

For the company in the previous problem, suppose fixed assets are \$690,000 and
sales are projected to grow to \$830,000. How much in new fixed assets are
required to support this growth in sales?

---

```{r problem 20; fixed assets and capacity usage, echo=F}
p20 <- new.env()
p20$fixed.assets <- 690000
p20$projected.sales <- 830000
p20$capital.intensity.ratio <-
    round(p20$fixed.assets / p19$sales.full.capacity, 3)
p20$projected.fixed.assets <- p20$projected.sales * p20$capital.intensity.ratio
p20$new.fixed.assets <- p20$projected.fixed.assets - p20$fixed.assets
```

The Thorpe's capital intensity ratio is:

$$
\begin{aligned}
\text{CIR} &= \frac{\text{fixed assets}}{\text{full-capacity sales}}\\
  &= \frac{`r pretty(p20$fixed.assets)`}{`r pretty(p19$sales.full.capacity)`}\\
  &= `r p20$capital.intensity.ratio`
\end{aligned}
$$

With a  projected sales growth of \$830,000, Thrope will need:

$$
\begin{aligned}
\text{New fixed assets} &= ( \text{projected sales} \times \text{CIR} )
  - \text{fixed assets}\\
  &= `r pretty(p20$projected.sales)`
  \times `r pretty(p20$capital.intensity.ratio)`
  - `r pretty(p20$fixed.assets)`\\
  &= `r pretty(p20$new.fixed.assets)`
\end{aligned}
$$

---

# 21. Calculating EFN

The most recent financial statements for Moose Tours, Inc., appear below. Sales
for 2012 are projected to grow by 20 percent. Interest expense will remain
constant; the tax rate and the dividend payout rate will also remain constant.
Costs, other expenses, current assets, fixed assets, and accounts payable
increase spontaneously with sales. If the firm is operating at full capacity and
no new debt or equity is issued, what external financing is needed to support
the 20 percent growth rate in sales?

```{r problem 21; income statement, echo=F}
p21 <- new.env()

p21$growth.rate <- .2
p21$is <- matrix(c(
    p21$sales <- 836100,
    p21$costs <- 650700,
    p21$other.expenses <- 17100,
    p21$ebit <- 168300,
    p21$interest <- 12600,
    p21$taxable.income <- 155700,
    p21$taxes <- 54495,
    p21$net.income <- p21$taxable.income - p21$taxes,
    p21$dividends <- 30300,
    p21$atre <- 70905), ncol=1)
row.names(p21$is) <- c("Sales", "Costs", "Other expenses",
    "Earnings before interest and taxes", "Interest expense", "Taxable income",
    "Taxes", "Net income", "Dividends", "Addition to retained earnings")

knitr::kable(pretty(p21$is), caption="Income Statement") %>%
    add_indent(c(9, 10)) %>%
    kable_styling(latex_options = "HOLD_position")
```

```{r problem 21; balance sheet, echo=F}
p21$bs.assets <- matrix(c(
    p21$cash <- 24035,
    p21$accounts.receivable <- 38665,
    p21$inventory <- 82555,
    p21$total.current.assets<- p21$cash + p21$accounts.receivable +
        p21$inventory,
    p21$fixed.assets <- 392350,
    p21$total.assets <- p21$total.current.assets + p21$fixed.assets), ncol=1)
row.names(p21$bs.assets) <- c("Cash", "Accounts receivable", "Inventory",
                              "Total current assets", "Fixed assets",
                              "Total assets")

p21$bs.liabilities <- matrix(c(
    p21$accounts.payable <- 64600,
    p21$notes.payable <- 16150,
    p21$total.current.liabilities <- p21$accounts.payable + p21$notes.payable,
    p21$long.term.debt <- 150000,
    p21$total.liabilities <- p21$total.current.liabilities +
                             p21$long.term.debt), ncol=1)
row.names(p21$bs.liabilities) <- c("Accounts payable", "Notes payable",
    "Total current liabilities", "Long-term debt", "Total liabilities")

p21$bs.equity <- matrix(c(
    p21$common.stock <- 130000,
    p21$retained.earnings <- 176855,
    p21$total.equity <- p21$common.stock + p21$retained.earnings), ncol=1)
row.names(p21$bs.equity) <- c("Common stock and paid-in surplus",
                              "Retained earnings", "Total equity")

p21$bs.total.liabilities.and.equity <- matrix(c(
    p21$total.liabilities + p21$total.equity), ncol=1)
row.names(p21$bs.total.liabilities.and.equity) <-
    c("Total liabilities and owners' equity")

p21$bs <- rbind(p21$bs.assets, p21$bs.liabilities,
                p21$bs.equity, p21$bs.total.liabilities.and.equity)

knitr::kable(pretty(p21$bs),
             caption="Balance Sheet") %>%
    pack_rows("Assets", 1, 5) %>%
    pack_rows("Liabilities", 7, 11) %>%
    pack_rows("Owners' equity", 12, 14) %>%
    add_indent(c(4, 9, 11, 14)) %>%
    row_spec(c(4, 11, 14),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")
```

## Building pro forma income statement

With a projected sales growth of 20\%, the pro forma income statement is:

```{r problem 21; pro forma income statement, echo=F}
p21$dividend.payout.ratio <- as.numeric(p21$is["Dividends",] /
                                        p21$is["Net income",])
p21$tax.rate <- as.numeric(p21$is["Taxes",] / p21$is["Taxable income",])

p21$is.proforma <- p21$is
p21$is.proforma[c("Sales", "Costs", "Other expenses",
                  "Earnings before interest and taxes"),] <- p21$is[c(
                  "Sales", "Costs", "Other expenses",
                  "Earnings before interest and taxes"),] *
                  (1 + p21$growth.rate)

p21$taxable.income <- p21$is.proforma["Earnings before interest and taxes",] -
    p21$is.proforma["Interest expense",]

p21$taxes <- p21$taxable.income * p21$tax.rate
p21$net.income <- p21$taxable.income - p21$taxes
p21$dividends <- p21$net.income * p21$dividend.payout.ratio
p21$atre <- p21$net.income * (1 - p21$dividend.payout.ratio)
p21$is.proforma["Taxable income",] <- p21$taxable.income
p21$is.proforma["Taxes",] <- p21$taxes
p21$is.proforma["Net income",] <- p21$net.income
p21$is.proforma["Dividends",] <- p21$dividends
p21$is.proforma["Addition to retained earnings",] <- p21$atre
knitr::kable(pretty(p21$is.proforma),
             caption="Pro Forma Income Statement") %>%
    add_indent(c(9, 10)) %>%
    kable_styling(latex_options = "HOLD_position")
```

## Building pro forma balance sheet

```{r problem 21; pro forma balance sheet, echo=F}
p21$bs.proforma <- p21$bs
p21$bs.proforma[1:7,] <- p21$bs.proforma[1:7,] * (1 + p21$growth.rate)

p21$bs.proforma["Total current liabilities",] <-
    p21$bs.proforma["Accounts payable",] + p21$bs.proforma["Notes payable",]

p21$bs.proforma["Total liabilities",] <- p21$bs.proforma["Long-term debt",] +
    p21$bs.proforma["Total current liabilities",]

p21$bs.proforma["Retained earnings",] <- p21$bs["Retained earnings",] +
    p21$is.proforma["Addition to retained earnings",]

p21$bs.proforma["Total equity",] <-
    p21$bs.proforma["Common stock and paid-in surplus",] +
    p21$bs.proforma["Retained earnings",]

p21$bs.proforma["Total liabilities and owners' equity",] <-
    p21$bs.proforma["Total liabilities",] + p21$bs.proforma["Total equity",]

p21$bs.diff <- p21$bs.proforma - p21$bs
p21$bs.pf.knitr <- cbind(p21$bs.proforma, p21$bs.diff)
knitr::kable(pretty(p21$bs.pf.knitr),
             col.names=c("Next year", "Change from current year"),
             caption="Pro Forma Balance Sheet") %>%
    pack_rows("Assets", 1, 5) %>%
    pack_rows("Liabilities", 7, 11) %>%
    pack_rows("Owners' equity", 12, 14) %>%
    add_indent(c(4, 9, 11, 14)) %>%
    row_spec(c(4, 11, 14),
             extra_latex_after = "\\addlinespace") %>%
    kable_styling(latex_options = "HOLD_position")

p21$test <- as.numeric(p21$bs.proforma["Total assets",] -
    p21$bs.proforma["Total liabilities and owners' equity",])
```

```{r problem 21; efn, echo=F}
p21$delta.sales <- as.numeric(p21$is["Sales",] * p21$growth.rate) 
p21$cir <- p21$bs["Total assets",] / p21$is["Sales",]
p21$sls <- p21$bs["Accounts payable",] / p21$is["Sales",]
p21$profit.margin <- p21$is["Net income",] / p21$is["Sales",]
p21$retention.ratio <- 1 - p21$dividend.payout.ratio
p21$efn <- as.numeric(p21$cir * p21$delta.sales - p21$sls * p21$delta.sales -
    p21$profit.margin * p21$is.proforma["Sales",] * p21$retention.ratio)
```

Finding the external financing needed to support 20\% growth:

$$
\begin{aligned}
\Delta \text{sales} &= \text{current sales} \times \text{growth rate}\\
  &= `r pretty(p21$is["Sales",])` \times `r p21$growth.rate`\\
  &= `r pretty(p21$delta.sales)`\\\\
\text{Projected sales} &= \text{Current sales} + \text{Change in sales}\\
  &= `r pretty(p21$is["Sales",])` + `r pretty(p21$delta.sales)`\\
  &= `r pretty(p21$is.proforma["Sales",])`\\\\
\text{Capital intensity ratio} &= \frac{\text{total assets}}{\text{sales}}\\
  &= \frac{`r pretty(p21$bs["Total assets",])`}
          {`r pretty(p21$is["Sales",])`}\\
  &= `r p21$cir`\\\\
\text{Spontaneous liabilities to sales} &= \frac{\text{accounts payable}}
  {\text{sales}}\\
  &= \frac{`r pretty(p21$bs["Accounts payable",])`}
  {`r pretty(p21$is["Sales",])`}\\
  &= `r pretty(p21$sls)`\\\\
\text{Dividend payout ratio} &= \frac{\text{dividends}}{\text{net income}}\\
  &= \frac{`r pretty(p21$is["Dividends",])`}{`r pretty(p21$is["Net income",])`}\\
  &= `r pretty(p21$dividend.payout.ratio)`\\\\
\text{Retention ratio} &= 1 - \text{dividend payout ratio}\\
  &= 1 - `r p21$dividend.payout.ratio`\\
  &= `r p21$retention.ratio`\\\\
\text{Profit margin} &= \frac{\text{net income}}{\text{sales}}\\
  &= \frac{`r pretty(p21$is["Net income",])`}{`r pretty(p21$is["Sales",])`}\\
  &= `r p21$profit.margin`\\\\
\text{EFN} &= \text{capital intensity ratio} \times \text{change in sales}\\
  &\ \ \ - \text{spontaneous liabilities to sales} \times \text{change in sales}\\
  &\ \ \ - \text{profit margin} \times \text{projected sales}
  \times \text{retention ratio}\\
  &= `r p21$cir` \times `r pretty(p21$delta.sales)`\\
  &\ \ \ - `r p21$sls` \times `r pretty(p21$delta.sales)`\\
  &\ \ \ - `r p21$profit.margin` \times `r pretty(p21$is.proforma["Sales",])`
  \times `r p21$retention.ratio`\\
  &= `r pretty(p21$efn)`
\end{aligned}
$$

---

# 22. Capacity usage and growth

In the previous problem, suppose the firm was operating at only 80 percent capacity in 2011. What is EFN now?

---

If Moose Tours, Inc. was operating at only 80\% of capacity in 2011, then the
firm has excess capacity in its fixed assets and does not need to invest in new
fixed assets until it reaches full capacity.

```{r problem 22; capacity usage and growth, echo=F}
p22 <- new.env()
p22$capacity.usage <- .8
p22$sales.full.capacity <- as.numeric(p21$is["Sales",]) / .8
p22$cir <- as.numeric(p21$bs.assets["Fixed assets",] /
                       p22$sales.full.capacity)
p22$projected.fixed.assets <- as.numeric(p21$is.proforma["Sales",] * p22$cir)
p22$previous.fixed.assets <- as.numeric(p21$bs.assets["Fixed assets",] *
                              (1 + p21$growth.rate))
p22$diff <- p22$previous.fixed.assets - p22$projected.fixed.assets
p22$efn <- p21$efn - p22$diff
```

Key information from the previous problem:

* Sales: \$`r pretty(p21$is["Sales",])`
* Fixed Assets: \$`r pretty(p21$bs["Fixed assets",])`
* Capacity usage: 80\%
* Projected Sales: \$`r pretty(p21$is.proforma["Sales",])`
  (`r p21$growth.rate*100`\% growth)
* Addition to retained earnings:
  \$`r pretty(p21$is["Addition to retained earnings",])`
* Increase in A/P (spontaneous liabilities):
  \$`r pretty(p21$bs["Accounts payable",])`
* Cash, A/R, Inventory scale with sales.

Full capacity sales:

$$
\begin{aligned}
\text{Full capacity sales} &= \frac{`r pretty(p21$is["Sales",])`}
  {`r p22$capacity.usage`}
  &= `r pretty(p22$sales.full.capacity)`
\end{aligned}
$$

From the balance sheet, our fixed assets are
\$`r pretty(p21$bs["Fixed assets",])`. The capital intensity ratio:

$$
\begin{aligned}
\text{Capital intensity ratio} &= \text{fixed assets}
  / \text{full-capacity sales}\\
  &= `r pretty(p21$bs["Fixed assets",])` / `r pretty(p22$sales.full.capacity)`\\
  &= `r pretty(p22$cir)`
\end{aligned}
$$

This means that, for every dollar in sales when full capacity is reached,
`r p22$cir` of fixed assets will be needed. At projected sales level of
\$`r pretty(p21$is.proforma["Sales",])`, Moose Tours will needs
\$`r pretty(p22$projected.fixed.assets)` in fixed assets,
which is \$`r pretty(p22$diff)` lower then the previous projection. Therefore we
need \$`r pretty(p22$efn)` in external funding.

---

# 23. Calculating EFN

In problem 21, suppose the firm wishes to  keep its debt-equity ratio constant. What is the EFN now?

---

```{r problem 23; pro forma income statement, echo=F}
p23 <- new.env()
p23$is.proforma <- p21$is
p23$is.proforma[1:4,] <- p21$is.proforma[1:4,] * (1 + p21$growth.rate)
p23$is.proforma["Taxable income",] <-
    p23$is.proforma["Earnings before interest and taxes",] -
    p23$is.proforma["Interest expense",]
p23$is.proforma["Taxes",] <- p23$is.proforma["Taxable income",] *
    p21$tax.rate
p23$is.proforma["Net income",] <- p23$is.proforma["Taxable income",] -
    p23$is.proforma["Taxes",]
p23$is.proforma["Dividends",] <- p23$is.proforma["Net income",] *
    p21$dividend.payout.ratio
p23$is.proforma["Addition to retained earnings",] <-
    p23$is.proforma["Net income",] * (1 - p21$dividend.payout.ratio)
knitr::kable(pretty(p23$is.proforma), caption="Pro Forma Income Statement") %>%
    add_indent(c(9, 10)) %>%
    kable_styling(latex_options = "HOLD_position")
```

```{r problem 23; pro forma balance sheet, echo=F}
p23$bs.proforma <- p21$bs
p23$bs.proforma[1:6,] <- p23$bs.proforma[1:6,] * (1 + p21$growth.rate)
p23$bs.proforma["Accounts payable",] * (1 + p21$growth.rate)
p23$bs.proforma["Notes payable",] <-
    p23$bs.proforma["Notes payable",] +
    19669
p23$bs.proforma["Total current liabilities",] <-
    p23$bs.proforma["Accounts payable",] +
    p23$bs.proforma["Notes payable",]
p23$bs.proforma["Long-term debt",] <-
    p23$bs.proforma["Long-term debt",] +
    59008
p23$bs.proforma["Total liabilities",] <-
    p23$bs.proforma["Total current liabilities",] +
    p23$bs.proforma["Long-term debt",]
p23$bs.proforma["Retained earnings",] <-
    p23$bs.proforma["Retained earnings",] +
    p23$is.proforma["Addition to retained earnings",]
p23$bs.proforma["Total equity",] <-
    p23$bs.proforma["Total equity",] +
    p23$is.proforma["Addition to retained earnings",]
p23$bs.proforma["Total liabilities and owners' equity",] <-
    p23$bs.proforma["Total liabilities",] +
    p23$bs.proforma["Total equity",]
pretty(p23$bs.proforma)
p23$bs.proforma["Total liabilities",] / p23$bs.proforma["Total equity",]
p21$bs["Total liabilities",] / p21$bs["Total equity",]
pretty(p23$bs.proforma["Total assets",] -
       p23$bs.proforma["Total liabilities and owners' equity",])
```

---

```{r}
ls()
```
