---
title: "Financial Statements and Cash Flow"
subtitle: "Text"
author: "Ross, Westerfield, Jaffe"
header-includes:
  - \usepackage{booktabs}
output: 
  pdf_document:
   latex_engine: xelatex
   toc: true 
---

```{r setup, include=FALSE}
require(kableExtra)
require(dplyr)
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
pretty <- function(x) format(x, big.mark=",")
```

# Chapter introduction

A write-off by a company frequently means that the value of the company's assets
has declined. For example, in 2010, Bank of America announced a total of \$34
billion in write-offs for bad loans. In addition, the company announced a \$10.5
billion write-off because recent legislation has reduced the value of its credit
card business. The write-offs made by Bank of America during 2010 followed
write-offs of \$33.7 billion in 2009 and \$16.2 billion in 2008. Of course, Bank
of America was not alone. Moody's Investor Services estimated that banks had
written off \$476 billion during 2009 and 2010, and it expected bank write-offs
of \$286 billion during 2011.

So did stockholders in these banks lose hundreds of billions of dollars (or
more) because of the write-offs? The answer is probably not. Understanding why
ultimately leads us to the main subject of this chapter: that all-important
substance known as cash flow.

---

# List of topics

1. The balance sheet

   * Liquidity
   * Debt versus equity
   * Value versus cost

2. The income statement

   * Generally Accepted Accounting Principles
   * Noncash items
   * Time and costs

3. Taxes

   * Corporate tax rates
   * Average versus marginal tax rates

4. New working capital

5. Financial cash flow

6. The accounting statement of cash flows

   * Cash flow from operating activities
   * Cash flow from investing activities
   * Cash flow from financing activities

7. Cash flow managment

---

# Notes

## The balance sheet

The accounting definition of the balance sheet:

$$
\text{Assets} \equiv \text{liabilities} + \text{Stockholders' equity}
$$

The balance sheet has two sides: On the left are the _assets_ and on the right
are the _liabilities_ and _stockholders' equity_. The stockholders' equity is
_defined_ to be the difference between the assets and the liabilities of the
firm.

In principle, equity is what the stockholders would have remaining after the
firm discharged its obligations.

Assets in the balance sheet are listed in order by the length of time it
normally would take an ongoing firm to convert them into cash.

The liabilities and stockholders' equity side reflects the types and proportions
of financing, which depend on management's choice of capital structure, as
between debt and equity and between current debt and long-term debt.

There are three concerns of a balance sheet:

1. Liquidity
2. Debt versus equity
3. Value versus cost

### Liquidity

_Liquidity_ refers to the ease and quickness with which assets can be converted
to cash (without significant loss in value). _Current assets_ are the most
liquid and include cash and assets that will be turned into cash within a year
from the date of the balance sheet.

_Accounts receivable_ are amounts not yet collected from customers for goods or
services sold to them (after adjustment for potential bad debts).

_Inventory_ is composed of raw materials to be used in production, work in
process, and finished goods.

_Fixed assets_ are the least liquid kind of assets.

### Debt versus equity

_Liabilities_ are obligations of the firm that require a payout of cash within a
stipulated period. Liabilities are debts and are frequently associated with
nominally fixed cash burdens, called _debt service_, that put the firm in
default of a contract if they are not paid. _Stockholders' equity_ is a claim
against the firm's assets that is residual and not fixed. In general terms, when
the firm borrows, it gives the bondholders first claim on the firm's cash flow.

$$
\text{Assets} - \text{Liabilities} \equiv \text{Stockholders' equity}
$$

The accounting value of stockholders' equity increases when retained earnings
are added. This occurs when the firm retains part of its earnings instead of
paying them out as dividends.

### Value versus cost

The accounting value of a firm's assets is frequently referred to as the
_carrying value_ or the _book value_ of the assets.

Under __generally accepted accounting principles (GAAP)__, audited financial
statements of firms in the United States carry the assets at cost.

_Market value_ is the price at which willing buyers and sellers would trade the
assets. It would be only a coincidence if accounting value and market value were
the same.

With the increasing globalization of business, there has been a growing need to
make accounting standards more comparable across countries. In recent years,
U.S. accounting standards have become more closely tied to International
Financial Reporting Standards (IFRS). In particular, the Financial Accounting
Standards Board, which is in charge of GAAP policies, and the International
Accounting Standards Board, in charge if IFRS polices, have been working towards
convergence of policies by June 2011. A decision about full adoption of IFRS by
U.S.-based companies is expected to follow afterwards.

See [__Market value versus book value__](#market-value-versus-book-value) for an
example of the difference between market value and book value of a corporation.

## The income statement

The income statement measures performance over a specific period---say a year.
The accounting definition of income is:

$$
\text{Revenue} - \text{Expenses} \equiv \text{Income}
$$

Income statement sections:

1. The operations section reports the firm's revenues and expenses from
   principal operations. Earnings before interest and taxes (EBIT) is very
   important. It summarizes earnings before taxes and financing costs.
2. The nonoperating section of the income statement includes all financing
   costs, such as interest expense.
3. Taxes levied on income.
4. The last section of the income statement is the bottom line the __net
   income__. Net income is frequently expressed per share of common stock---that
   is earnings per share.

### Generally Accepted Accounting Principles

*  Income is recognized when an exchange of goods occur, but not necessarily
  when income is received.

*  The unrealized appreciation from owning property will not be recognized as
  income.

* The matching principle of GAAP dictates that revenues be matched with
  expenses. Thus, income is reported when it is earned, or accrued, even though
  no cash flow has necessarily occurred.

### Noncash items

There are several noncash items that appear on the income statement.

* __Depreciation__ reflects the accountant's estimate of the cost of equipment
  used up in the production process.

* __Deferred taxes__ result from differences between accounting income and true
  taxable income (such as when accounting income uses accelerated depreciation
  expense procedures for tax purposes, but use straight-line depreciation for
  reporting purposes). The current tax portion is the actual tax paid. Deferred
  taxes is not. However, in theory, if taxable income is less than accounting
  income in the current year, it will be more than accounting income later on.
  Therefore, the taxes will have to be paid later on. The balance sheet will
  show deferred taxes as a liability.

### Time and costs

* There are two distinct time frames: long run and short run.

* The _short run_ is the period in which certain equipment, resources, and
  commitments of a firm are fixed; but the time is long enough for the firm to
  vary its output by using more labor and raw materials. Therefore, in the short
  run some costs are fixed and some costs (such as labor and materials) are
  variable.

* In the _long run_, all costs are variable.

* Financial accountants do not distinguish between variable costs and fixed
  costs. Instead, accounting costs usually fit into a classification that
  distinguishes product costs from period costs.

* __Product costs__ are the total production costs incurred during a
  period---raw materials, direct labor, and manufacturing overhead---and are
  reported on the income statement as costs of goods sold. Both variable and
  fixed costs are reported in product costs.

* __Period costs__ are costs that are allocated to a time period; they
  are called _selling_, _general_, and _administrative expenses_. An example of
  a period cost is the company president's salary.

## Taxes

Taxes can be one of the largest cash outflows a firm experiences. Taxes can be
viewed through two lenses: The average tax rate and the marginal tax rate. The
tax schedule is broken down into different income categories. As a firm moves up
the income schedule, the income in that bracket is taxed differently than the
income in the previous brackets (the tax rate does not necessarily increase as
you go up the tax schedule).

The average tax rate is the summation of income times the rate for each bracket.

The marginal tax rate is the tax rate on the next dollar earned.

The following table is an example of a corporate tax schedule:

\label{table:taxrates}
```{r corporate tax table, echo=F}
corporate.tax.rate.table <- data.frame(
  c(0, 50001, 75001, 100001, 335001, 10000001, 15000001, 18333334),
  c(50000, 75000, 100000, 335000, 10000000, 15000000, 18333333, Inf),
  c(.15, .25, .34, .39, .34, .35, .38, .35)
)
colnames(corporate.tax.rate.table) <- c("Lower", "Upper", "Tax.Rate")
knitr::kable(pretty(corporate.tax.rate.table), booktabs=T)
```

An example analysis of taxes can be found in
[__Deep in the heart of taxes__](#deep-in-the-heart-of-taxes)

## Net working capital

__NOTE:__ The change in net working capital is usually positive in a growing
firm.

Net working capital is current assets minus current liabilities:

$$
\text{Net working capital} = \text{Current assets}
  - \text{Current liabilities}
$$

* When net working capital is positive it means that the cash that will become
  available in the next 12 months will be greater that the cash that must be
  paid out.

* In addition to investing in fixed assets (i.e., capital spending), a firm can
  invest in net working capital. This is called the __change in net working
  capital__. The change in net working capital is usually positive in a growing
  firm.

## Financial cash flow

* The most important item that can be extracted from financial statements is the
  actual cash flow of the firm.

* The _statement of cash flows_ is an accounting statement that helps explain
  the change in accounting cash and equivalents.

* In finance, the value of a firm is its ability to generate financial cash
  flow.

* Cash flow is not the same as net working capital. Increasing inventory
  requires using cash. Because both inventories and cash are current assets,
  this does not affect net working capital. In this case, an increase in
  inventory is associated with decreasing cash flow.

* Just as the value of a firm's assets is always equal to the combined value of
  the liabilities and the value of the equity, the cash flows received from the
  firm's assets (that is, its operating activities), $\text{CF}(A)$ must equal
  the cash flows to the firm's creditors $\text{CF}(B)$, and equity investors
  $\text{CF}(S)$.

$$
\begin{aligned}
\text{CF}(A) &\equiv \text{CF}(B) + \text{CF}(S)\\\\
\text{CF}(A) &= \text{Operating cash flow} - \text{Capital spending}
  - \text{Increase in net working capital}\\
\text{CF}(B) &= \text{Interest} + \text{Debt retirement} - \text{New debt}\\
\text{CF}(S) &= \text{Dividends} + \text{Equity repurchase} - \text{New equity}
\end{aligned}
$$

### Cash flow from operations

* Operating cash flow is the cash flow generated by business activities,
  including sales of goods and services. Operating cash flow reflex tax
  payments, but not financing, capital spending, or changes in net working
  capital.

$$
\text{Operating cash flow} = \text{EBIT} + \text{Depreciation} - \text{Taxes}
$$

```{r calculating operating cash flow, echo=F}
m <- c(219, 90, -71)
df <- data.frame(c(m, sum(m)))
colnames(df) <- c("$ in millions")
rownames(df) <- c("Earnings before interest and taxes",
                  "Depreciation",
                  "Current taxes",
                  "Operating cash flow")
knitr::kable(df, caption="Operating cash flow")
```

* Changes in fixed assets are the result of capital spending

$$
\text{Capital spending} = \text{Acquisition of fixed assets}
  - \text{Sale of fixed assets}
$$

```{r calculating changes in fixed assets, echo=F}
m <- c(198, -25)
df <- data.frame(c(m, sum(m)))
colnames(df) <- c("$ in millions")
rownames(df) <- c("Acquisition of fixed assets",
                  "Sales of fixed assets",
                  "Capital spending")
knitr::kable(df, caption="Capital spending")
```

* Capital spending can also be calculated as:

$$
\text{Capital spending} = \text{Ending net fixed assets}
 - \text{Beginning net fixed assets} + \text{Depreciation}
$$

* Cash flows are also used to make investments in net working capital.

```{r calculating additions to net working capital, echo=F}
m <- c(23)
df <- data.frame(m)
rownames(df) <- c("Additions to net working capital")
knitr::kable(df,
             caption="Additions to net woring capital",
             col.names=c("$ in millions"))
```

* Total cash flows generated by the firm's assets are then equal to:

```{r calculating total cash flow of the firm, echo=F}
m <- c(238, -173, -23)
df <- data.frame(c(m, sum(m)))
colnames(df) <- c("$ in millions")
rownames(df) <- c("Operating cash flow",
                  "Capital spending",
                  "Additions to net working capital",
                  "Total cash flow of the firm")
knitr::kable(df, caption="Cash flow of the firm")
```

* The total outgoing cash flow of the firm can be separated into cash flow paid
  to creditors and cash flow paid to stockholders.

* Creditors are paid an amount generally referred to as a _debt service_. Debt
  service is interest payments plus repayments of principal (that is, retirement
  of debt). An important source of cash flow is the sale of new debt.

```{r calculating cash flow paid to creditors, echo=F}
m <- c(49, 73)
m.1 <- c(m, sum(m))
m.2 <- c(m.1, -86)
df <- data.frame(c(m.2, sum(m.2)))
colnames(df) <- c("$ in millions")
rownames(df) <- c("Interest",
                  "Retirement of debt",
                  "Debt service",
                  "Proceeds from long-term debt sales",
                  "Total")
knitr::kable(df, caption="Cash flow paid to creditors")
```

* Cash flow paid to creditors can also be calculated:

$$
\text{Cash flow paid to creditors} = \text{interest paid}
  - \text{net borrowing}
$$

* Cash flow of the firm is also paid to the stockholders in the form of
  dividends and stock repurchases.

```{r calculating cash flow paid to stockholders, echo=F}
m <- c(43, 6)
m.1 <- c(m, sum(m))
m.2 <- c(m.1, -43)
df <- data.frame(c(m.2, sum(m.2)))
colnames(df) <- c("$ in millions")
rownames(df) <- c("Dividends",
                  "Repurchase of stock",
                  "Cash to stockholders",
                  "Proceeds from new stock issue",
                  "Total")
knitr::kable(df, caption="Cash flow to stockholders")
```

* In general, cash flow to stockholders is determined:

$$
\text{Cash flow to stockholders} = \text{Dividends paid}
  - \text{Net new equity raised}
$$

### Observatations of cash flows

From our discussion of cash flows several important observations can be drawn:

#### Operating cash flow

defined as earnings before interest plus depreciation minus taxes, measures the
cash generated from operations not counting capital spending or working capital
requirements. It is usually positive; a firm is in trouble if operating cash
flow is negative for a long time because the firm is not generating enough cash
to pay operating costs.

#### Total cash flow of the firm

includes adjustments for capital spending and additions to net working capital.
It will frequently be negative. When a firm is growing rapidly, spending on
inventory and fixed assets can be higher than operating cash flow.

#### Net income is not cash flow.

When determining the economic and financial condition of a firm, cash flow is
more revealing.

## The accounting statement of cash flows

The _statement of cash flows_ is an accounting statement that is useful in
understanding financial cash flow.

The first step in determining the change in cash is to figure out cash flow from
operating activities. This is the cash flow that results from the firm's normal
activities in producing and selling goods and services. The second step is to
make an adjustment for cash flow from investing activities. The final step is to
make an adjustment for cash flow from financing activities. Financing activiites
are the net payments to creditors and owners (excluding interest expense) made
during the year.

### Cash flow from operating activities

To calculate the cash flow from operating activities we start with net income
and add back all noncash expenses and adjust for changes in current assets and
liabilities (other than cash and notes payable). The result is cash flow from
operating activities. Notes payable will be included in the financing activities
section.

```{r calculating cash flow from operating activities, echo=F}
net.income <- c(86, 90, 13)
changes.in.assets.and.liabilities <- c(-24, 11, 16, 18, -8)
cash.flow.from.operating.activities <- sum(net.income,
                                           changes.in.assets.and.liabilities)
cols <- c("$ in millions")
ni.rows <- c("Net income",
             "Depreciation",
             "Deferred taxes")
ciaal.rows <- c("Accounts receivable",
                "Inventories",
                "Accounts payable",
                "Accrued expense",
                "Other")
ni.df <- data.frame(net.income)
colnames(ni.df) <- cols
rownames(ni.df) <- ni.rows
ciaal.df <- data.frame(changes.in.assets.and.liabilities)
colnames(ciaal.df) <- cols
rownames(ciaal.df) <- ciaal.rows
cffoa.df <- data.frame(cash.flow.from.operating.activities)
colnames(cffoa.df) <- cols
rownames(cffoa.df) <- c("Cash flow from operating activities")

knitr::kable(ni.df, caption="Net Income")
knitr::kable(ciaal.df, caption="Change in assets and liabilities")
knitr::kable(cffoa.df, caption="Cash Flow from Operating Acitivites")
```

## Cash flow from investing activities

Cash flow from investing activities involves changes in capital assets:
acquisition of fixed assets and sales fo fixed assets (i.e., net capital
expenditures).

```{r calculating cash flow from investing activities, echo=F}
m <- c(-198, 25)
cffia.df <- data.frame(c(m, sum(m)))
colnames(cffia.df) <- "$ in millions"
rownames(cffia.df) <- c("Acquisition of fixed assets",
                        "Sales of fixed assets",
                        "Cash flow from investing activities")
knitr::kable(df, caption="Cash Flow from Investing Activities")
```

## Cash flow from financing activities

```{r calculating cash flow from financing activities, echo=F}
m <- c(-73, 86, -3, -43, -6, 43)
cfffa.df <- data.frame(c(m, sum(m)))
colnames(cfffa.df) <- "$ in millions"
rownames(cfffa.df) <- c("Retirement of long-term debt",
                        "Proceeds from long-term debt sales",
                        "Change in notes payable",
                        "Dividends",
                        "Repurchase of stock",
                        "Proceeds from new stock issue",
                        "Cash flow from financing activities")
knitr::kable(cfffa.df, caption="Cash flow from financing activities")
```

The statement of cash flows is the addition of cash flows from operations, cash
flows from investing activities, and cash flows from financing activities.

```{r statement of cash flows, echo=F}
cic.df <- data.frame(c(33))
colnames(cic.df) <- "$ in millions"
rownames(cic.df) <- "Change in cash (on the balance sheet"
socf.df <- rbind(cffoa.df, cffia.df, cfffa.df, cic.df)
knitr::kable(socf.df, caption="Statement of Cash Flows")
```

---

# Chapter examples

## Market value versus book value

The Cooney Corporation has fixed assets with a book value of \$700 and an
appraised market value of about \$1,000. Net working capital is \$400 on the
books, but approximately \$600 would be realized if all the current accounts
were liquidated. Cooney has \$500 in long-term debt, both book value and market
value. What is the book value of the equity? What is the market value?

We can construct two simplified balance sheets, one in accounting (book value)
terms and one in economic (market value) terms:

*Cooney Corporation Balance Sheets---Market value versus book value*

\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
Assets
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ 
Liabilities and Shareholders' equity

| . | Book | Market | . | Book | Market |
|:---:|:---:|:---:|:---:|:---:|:---:|
| Net working capital | \$\ 400 | \$\ 600 | Long-term debt | \$\ 500 | \$\ 500 |
| Net fixed assets | \ \ 700 | \ 1,000 | Shareholders' equity | \ \ 600 | \ 1,100 |
| . | --- --- | --- --- | . | --- --- | --- --- |
| . | \$1,100 | \$1,600 | . | \$1,100 | \$1,600 |

In this example, shareholders' equity is actually worth almost twice as much as
what is shown on the books. The distinction between book and market values is
important precisely because book vlaues can be so different from market values.

---

## Deep in the heart of taxes

Algeron, Inc., has a taxable income of \$85,00. What is its tax bill? What is
its average tax rate? Its marginal tax rate?

From the [Tax Rate Table](#table:taxrates), we see taht the tax rate appled to the first \$50,000
is 15 percent; the rate applied to the next \$25,000 is 25 percent; and the rate
applied after that up to \$100,000 is 34 percent. So Algernon must pay
$.15 \times \$50,000 + .25 \times 25,000 + .34 \times (85,000 - 75,000)$
$= \$17,150$. The average tax rate is thus $\$17,150/85,000 = 20.18\%$. The
marginal rate is 34 percent because Algernon's taxes would rise by 34 cents if
it had another dollar in taxable income.

---

# Summary and conclusions

Besides introducing you to corporate accounting, the purpose of this chapter
has been to teach you how to determine cash flow from the accounting statements
of a typical company.

1. Cash flow is generated by the firm and paid to creditors and shareholders.
   It can be classified as:

>  a. Cash flow from operations.
>  b. Cash flow from changes in fixed assets.
>  c. Cash flow from changes in working capital.

2. Calculations of cash flow are not difficult, but they require care and
   particular attention to detail in properly accounting for non-cash expenses
   such as deprecation and deferred taxes. It is especially important that you
   do not confuse cash flow with changes in net working capital and net income.

3. We showed how pro forma financial statements can be generated and used to
   plan for future financing needs.
