---
title: Chapter 25 Derivatives and hedging risk
subtitle: Questions
output:
  pdf_document:
    includes:
      in_headers: preamble.tex
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F)
pretty <- function(x, n=2) format(round(x, n), big.mark=',')
require(tidyverse)
```

# 1. Futures Quotes

Refer to Table 25.2 in the text to answer this question. Suppose you purchase a
March 2012 cocoa futures contract on November 22, 2011, at the last price of the
day. What will your profit or loss be if cocoa prices turn out to be \$2,431 per
metric ton at expiration?

We are buying a futures contract at the price of \$2,313. We will profit only if
the price of cocoa increases on expriation in March 2012. Since the spot price
is \$2,431 in March 2012, we will have profited \(\$2,414 - \$2,4312 = \$17\)
per metric ton. The total profit is \(17 \times 10 \times 1 = \$170\).


# 2. Futures Quotes

Refer to Table 25.2 in the text to answer this question. Suppose you sell five
March 2012 silver futures contracts on November 22, 2011, at the last price of
the day. What will your profit or loss be if silver prices turn out to be
\$31.39 per ounce at expiration? What if silver prices are \$30.86 per ounce at
expiration.

We are selling five contracts. We will profit only if the price of cocoa
decreases on expiration in March 2012.

The settle price on November 22, 2011 is \$31.187 per troy ounce.

If the price is \$31.39 the __loss per troy once__ is \(31.187 - 31.39 =
(\$.203)\).  The loss is \(.203 \times 5,000 \times 5 = (\$5,075)\).

If price is \$30.86 the __gain per troy ounce__ is \(31.187 - 30.86 = \$.327\). The gain is \(.327 \times 5,000 \times 5 = \$8,175\).

# 3. Put and Call Payoffs

Suppose a financial manager buys call options on 50,000 barrels of oil with an
exercise price of \$95 per barrel. She simultaneously sells a put option on
50,000 barrels of oil with the same exercise price of \$95 per barrel. Consider
her gains and losses if oil prices are \$90, \$92, \$95, \$98, and \$100. What
do you notice about the payoff profile?

```{r problem 3}
p3   = new.env()
p3$E = 95
p3$Q = 50000
p3$S = c(90, 92, 95, 98, 100)
p3$C = pmax(p3$S - p3$E, 0)
p3$P = -pmax(p3$E - p3$S, 0)
```

## Synthetic long futures position

By buying a call and selling a put with the same strike price (\$95) and the
same notional size (50,000 barrels), the financial manager creates a position
that behaves almost identically to being long 50,000 barrels of crude oil
futures (or a forward contract) at exactly \$95 per barrel.

```{r problem 3; table}
tbl = tibble(
  `Oil Price` = p3$S,
  `Call payoff` = p3$C,
  `Put payoff` = p3$P,
  `Net` = p3$C + p3$P,
  `Total payoff` = Net * p3$Q
)
knitr::kable(tbl)
```

## What you should notice about the payoff profile

1. It is a straight line with slope +1. For every \$1 the oil price moves (up or
   down), the strategy gains or loses exactly \$1 per barrel exactly \$50,000
   per \$1 move in oil.

2. It is virtually identical to being long an oil futures contract at \$%95. The
   only minor difference is the net premium paid or received for the two
   options.

3. Unlimited upside, unlimited downside. Exactly like a naked long futures
   position. There is no cap and no floor.

4. Zero payoff at exactly the strike (\$95)

## Conclusion

The combination of long call + short put at the same strike creates a synthetic
long futures contract. The payoff profile is linear, delta \(\approx\) +1.00,
gamma \(\approx\) 0, and the position profits or loses dollar-for-dollar with
the price of oil--exactly what the financial manager wanted if she is bullish on
oil or simply wnants futures-like exposure without posting traditional futures
margin.
